### 2013-07-24 schemeworkshop13.tex
Add todo list. Some drafting

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
%\usepackage{url}
%\usepackage[style=alphabetic, sorting=nyt]{biblatex}
%\addbibresource{bibliography.bib}

\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\usepackage{amsmath}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Subtitle Text, if any}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

Typed Clojure is a gradual type system for Clojure, based on the
static portion of Typed Racket.
This paper gives an overview of Typed Clojure, concentrating on extensions
that could interest Typed Racket users.

% what are the goals of Typed Clojure?
% where do they differ from Typed Racket? why?

% Things todo
% - small calculus based on previous TR formalisation
%   that demos the ``flow'' environment & not/intersection type
% - discussion and examples for flow filters
% - discussion on the not/difference type, real code where it helps
% - higher-rank polymorphism, show what it enables (eg. monads)
%   - also show more complicated example, making an extensible
%     type for clojure.core/conj, and why it's not straightforward
% - discussion on limitations of local type inference
% - demonstration of how we check multimethods
% - dealing with array covariance
% - note on how we avoid null-pointer exceptions
% - discussion on why Typed Clojure only uses the ``static'' part
%   of Typed Racket
% - the equality filter problem (and draft solution?), a proposition
%   that knows about binding aliasing. Sounds straightforward?

\section{Using negative filters}

Occurrence typing plays an important role in Typed Racket and Typed Clojure.
By maintaining a \emph{proposition environment} of propositions relating types to
bindings, we can update bindings with more accurate types as programs progress.
It follows that there is some correspondence between propositions and types,
characterised by the \emph{update} function, which takes a type and a proposition
and returns a type which updates the input type using the proposition.

There is a straightforward relationship between ``positive'' propositions and types.
For example 
{\tt (update Number (is Integer 0))}
updates Number by Integer, which is Integer, because Integer <: Number.

The relationship between ``negative'' propositions and types is not always obvious.
A common proposition in Typed Clojure is (! (U nil false) a): the proposition that
local binding ``a'' is \emph{not} of type (U nil false).
This problem is most visible in expressions like {\tt (filter identity coll)}, where
``identity'' has a ``then'' proposition that has negative information: (! (U nil false) 0),
which reads, the 0th argument of identity does not contain (U nil false).

\section{Extending occurrence typing}

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \expe{} &::=& (\expe{}\ \expe{}) \alt \abs{\x a}{\expe{}}
                &\mbox{Expressions}
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
\end{altgrammar}
$$
\end{figure*}

%$$
%\begin{tdisplay}{Evaluation Contexts}
%  \begin{altgrammar}
%    \E{} &::=& [ ] % application rules
%              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
%              % map rules
%              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
%              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
%              &\mbox{Evaluation Contexts}
%  \end{altgrammar}
%\end{tdisplay}
%$$

%\appendix
%\section{Appendix Title}
%
%This is the text of the appendix, if you need one.

%\acks
%
%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

%\bibliographystyle{abbrvnat}
%
%% The bibliography should be embedded for final submission.
%
%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}

\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2013-07-25 schemeworkshop13.tex
Draft bivariant arrays

@@
 
-\section{Introduction}
+\section{Typed Clojure}
 
@@
 static portion of Typed Racket.
-This paper gives an overview of Typed Clojure, concentrating on extensions
-that could interest Typed Racket users.
 
@@
 %   that knows about binding aliasing. Sounds straightforward?
+
+\section{Higher-rank polymorphism}
+
+Typed Clojure supports f-bounded polymorphism and higher-rank
+types.
 
@@
 
+\section{Java interoperability}
+
+As Clojure supports Java interoperability, Typed Clojure has some
+integration with Java's type system. Calls to Java methods from
+Clojure can be type checked by Typed Clojure.
+
+We currently do not support Java generics
+
+\subsection{Arrays}
+
+Supporting statically sound interactions with Java arrays is a goal
+of Typed Clojure. This is complicated by Java's decision to make
+arrays covariant in their argument, a well documented source of static
+unsoundness.
+
+Our basic approach is to make our array types \emph{bivariant}. Array types
+look like \lstinline|(Array2 w r)| and
+are reminicent of functions or pipes: having a contravariant parameter for input (writing)
+and a covariant parameter for output (reading).
+
+Most commonly, an array type is invariant in its parameter; it can
+write and read input of the same type.
+We can get the same effect by setting our input and output
+parameters to the same type. For example, \lstinline|(Array2 Number Number)|
+(or equivalently, \lstinline|(Array Number)|)
+in Typed Clojure is similar to invariant array types of \lstinline|Number| in languages like Scala.
+
+The biggest gain in using a separate input parameter is the ability
+to specify \emph{read-only} arrays. Crucially, our type system features an
+explicit bottom type \lstinline|Nothing|, allowing a read-only \lstinline|Number| array
+to be of type \lstinline|(Array2 Nothing Number)|.
+
+To realise why defining read-only arrays are useful, we need to examine
+what makes array covariance unsound in Java.
+Array covariance lets us lie about the type of an array so the consumer
+of an array cannot tell the actual type of the array when examining a type
+signature.
+
+\begin{verbatim}
+...
+public static Number[] getNumberArray() {
+  Number[] n = new Integer[10];
+  return n;
+}
+...
+\end{verbatim}
+
+To the casual consumer \emph{getNumberArray} returns an array that can both
+read and write \lstinline|Number|s. However it is clear from the implementation
+that attempting to write say a \lstinline|Double| to this array will result
+in a runtime error.
+
+\begin{verbatim}
+...
+Number[] myArray = getNumberArray();
+myArray[0] = 1.1;
+/* Exception in thread "main" java.lang.ArrayStoreException: java.lang.Double */
+...
+\end{verbatim}
+
+Notice that this is a runtime error, and Java's type system has not helped
+statically prevent it.
+This could cause a similar issue for other statically-typed languages offering
+interoperability with Java.
+
+To prevent these sorts of runtime exceptions in Typed Clojure, we declare
+all arrays from unknown sources to be \emph{read-only}. Put differently,
+the only way to define a writeable array is to create it in type-checked Clojure
+code.
+
+\begin{lstlisting}
+(let [n (CovariantArray/getNumberArray)]
+  (aset n 0 1.1))
+
+; Polymorphic static method clojure.lang.RT/aset could not be 
+; applied to arguments:
+; Domains: 
+;         (Array2 i o) clojure.core.typed/AnyInteger i
+; 
+; Arguments:
+;         (Array2 Nothing java.lang.Number) int (Value 1.1)
+; 
+; with expected type:
+;         Any
+\end{lstlisting}
+
+The type inferred for the local \lstinline|n| is (Array2 Nothing Number)
+which tells the type system: it is never safe to write to this array, but
+it is safe to assume \lstinline|Number|s can be read from this array.
+
+To emphasise, this is a static type error. Errors like this help Clojure programmers
+use foreign Java libraries more correctly.
+
+\begin{verbatim}
+Note that Java libraries are often large and complex and programmers will probably
+enjoy the extra help from the type system.
+\end{verbatim}
+
 %\appendix

### 2013-07-25 schemeworkshop13.tex
preprint
Add dealing with null

@@
 
-\documentclass{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
+%\titlebanner{banner above paper title}        % These are ignored unless
+%\preprintfooter{short description of paper}   % 'preprint' option specified.
 
-\title{Title Text}
-\subtitle{Subtitle Text, if any}
+\title{Typed Clojure}
+%\subtitle{Subtitle Text, if any}
 
@@
 look like \lstinline|(Array2 w r)| and
-are reminicent of functions or pipes: having a contravariant parameter for input (writing)
+are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
 and a covariant parameter for output (reading).
@@
 
-The type inferred for the local \lstinline|n| is (Array2 Nothing Number)
+The type inferred for the local \lstinline|n| is \lstinline|(Array2 Nothing Number)|
 which tells the type system: it is never safe to write to this array, but
@@
 \end{verbatim}
+
+\subsection{Handling null}
+
+Probably the most common pain point in Java programming is dealing with \emph{null}.
+It is crucial that Typed Clojure deals with \emph{null} intelligently so that
+all potentially erroneous interactions with \emph{null} in typed code are caught at compile time.
+We also want to represent \emph{null} in a way that is natural to Clojure programmers
+and approximates their mental model when reasoning about Clojure code.
+
+Firstly, we separate the concepts of \emph{null} and reference types at the type level.
+This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
+We provide a singleton type \lstinline|nil|, which contains just the value \lstinline|nil|,
+Clojure's equivalent of \emph{null}.
+
+Armed with general unions and the type \lstinline|nil|, 
+we can make Typed Clojure's reference types \emph{non-nullable}.
+A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
+in Typed Clojure.
+
+Typed Racket already provides what is needed to write such types; indeed this is a repurposing
+of a common Typed Racket idiom of approximating an Option or Maybe type by
+making a union that includes \emph{\#f}, Racket's only false value.
+However, like handling covariant arrays, the subtlety lies in our interactions with foreign
+Java code.
+
+A set of conversion rules define how to convert a Java type signature into a null-safe
+Clojure version. These rules are designed to be conservative and can be explicitly overridden
+by programmers. Method parameters are non-nullable, while return types are nullable.
+Field types are nullable. Constructors parameters are non-nullable, and return types
+are also non-nullable: it is a Java invariant that all constructors return an instance
+of the Class they are constructing.
 

### 2013-07-26 schemeworkshop13.tex
Work on judgements

@@
 
-\documentclass[preprint]{sigplanconf}
+\documentclass[preprint,10pt]{sigplanconf}
 
@@
 \title{Typed Clojure}
-%\subtitle{Subtitle Text, if any}
+\subtitle{A Practical Type System for Clojure based on Typed Racket}
 
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
+\authorinfo{Ambrose Bonnaire-Sergeant}
+           {}
+           {abonnairesergeant@gmail.com}
+\authorinfo{Rowan Davies}
+           {University of Western Australia}
+           {rowan@csse.uwa.edu.au}
 
@@
 \begin{abstract}
-This is the text of the abstract.
+Typed Racket has proved to be a valuable starting point for
+a gradual type system for Clojure.
+Building a similar type system for a new language gives the
+designer some flexibility to repurpose and extend features.
+This paper gives an overview of Typed Clojure, concentrating
+on the extensions and differences from Typed Racket. We also
+show where Typed Racket's features were particularly useful
+for type checking non-trivial Clojure idioms.
 \end{abstract}
@@
 \keywords
-keyword1, keyword2
+gradual types, Clojure
 
@@
 %   that knows about binding aliasing. Sounds straightforward?
+
+\section{Contributions}
+
+The main result of this paper is to demonstrate that Typed Racket's design
+and implementation is applicable to a practical type system for Clojure
+by implementing Typed Clojure.
 
@@
 
+\section{Multimethods}
+
+Clojure provides multimethods which can dispatch on the result of an
+arbitrary function, the \emph{dispatch function}. \emph{Methods} are
+added to the multimethod
+which are associated with a \emph{dispatch value}, which is compared to the
+result of apply the dispatch function to the argument passed to the
+multimethod. If this comparison unambiguously dispatches to one method,
+that method is then called with the current arguments.
+
+Occurrence typing, in particular when extended with paths,
+for checking method definitions for arbitrary dispatch functions.
+
+\begin{lstlisting}
+(ns example.mm
+  (:require 
+    [clojure.core.typed :refer [ann]]))
+
+(ann my-mm [Any -> Number])
+(defmulti my-mm class)
+
+(defmethod my-mm Float [f] (inc f))
+\end{lstlisting}
+
+\begin{lstlisting}
+(ann clojure.core/class
+  [Any -> (U nil Class)
+   :object {:id 0, :path [Class]}])
+\end{lstlisting}
+
+\begin{verbatim}
+TODO
+- example of normal MM usage
+- explain descrepency btwn TC syntax and formal
+  TR calculs syntax. eg. :object/:filter vs over arrow
+\end{verbatim}
+
 \section{Extending occurrence typing}
@@
 \begin{altgrammar}
-  \expd{}, \expe{} &::=& (\expe{}\ \expe{}) \alt \abs{\x a}{\expe{}}
-                &\mbox{Expressions}
+  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
+                      \alt \true{} \alt \false{} \alt \nil{}
+                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
+                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
+                      \alt {\doexp {\e{}} {\e{}}}
+                      \alt {\fnexp {\x{}} {\t{}} {\e{}}}
+                &\mbox{Expressions} \\ \\
+  \s{}, \t{}    &::=& (\class{}\ \overrightarrow{\t{}})
+                      \alt (\ma{\Value\ \v})
+                      \alt (\cup\ \overrightarrow{\t{}})
+                      \alt (\cap\ \overrightarrow{\t{}})
+                      \alt (\Not\ \t{})
+                      \alt (\All\ [\overrightarrow{\x{}}] \t{})
+                      \alt (\TFn\ [\overrightarrow{\x{}}] \t{})
+                &\mbox{Types}
 %  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
@@
 $$
+\caption{Syntax of Expressions, Types}
+\end{figure*}
+
+\begin{figure*}
+\begin{mathpar}
+\infer [T-Var]
+{ \inpropenv {\propenv{}} {\isprop {\x} {\tau}}}
+{ \judgement {\propenv{}} {{\hastype {\x{}} {\t{}}}} {{\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}}
+             {\emptyobject{}} {\topprop{}}}
+
+\infer [T-True]
+{}
+{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} {\topprop{}}}
+
+\infer [T-False]
+{}
+{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}
+
+\infer [T-Nil]
+{}
+{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}
+
+\infer [T-Do]
+{ \judgement {\propenv{}} {{\hastype {\e1} {\t1}}} {{\filterset {\prop{1+}} {\prop2}}} {\object{1}} {{\prop{f1}}}
+\\
+  \judgement {\propenv{}, \prop{f1}} {{\hastype {\e2} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object{}} {\prop{f2}}
+  }
+{ \judgement {\propenv{}} {{\hastype {\doexp {\e1} {\e2}} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object} 
+             {{\andprop {\prop{f1}} {\prop{f2}}}}}
+
+\infer [T-If]
+
+%\infer [T-App]
+
+\end{mathpar}
+\caption{Typing Rules}
 \end{figure*}
@@
 myArray[0] = 1.1;
-/* Exception in thread "main" java.lang.ArrayStoreException: java.lang.Double */
+/* Exception in thread "main" 
+   java.lang.ArrayStoreException: 
+   java.lang.Double */
 ...
@@
 \begin{verbatim}
-Note that Java libraries are often large and complex and programmers will probably
+Note that Java libraries are often large 
+and complex and programmers will probably
 enjoy the extra help from the type system.
@@
 Armed with general unions and the type \lstinline|nil|, 
-we can make Typed Clojure's reference types \emph{non-nullable}.
+we make Typed Clojure's reference types \emph{non-nullable}.
 A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
@@
 by programmers. Method parameters are non-nullable, while return types are nullable.
-Field types are nullable. Constructors parameters are non-nullable, and return types
+Field types are nullable. Constructor parameters are non-nullable, and return types
 are also non-nullable: it is a Java invariant that all constructors return an instance
-of the Class they are constructing.
+of the class they are constructing.
+
+\begin{verbatim}
+TODO 
+- write conversion rules for 
+  Java type -> Clojure type
+\end{verbatim}
 

### 2013-07-29 schemeworkshop13.tex
More semantics, start Java conversions rules
more judgements, metafunctions

@@
 \begin{abstract}
-Typed Racket has proved to be a valuable starting point for
-a gradual type system for Clojure.
+Typed Racket is a valuable starting point for
+a gradual type system that targets Clojure.
 Building a similar type system for a new language gives the
@@
 
-\section{Higher-rank polymorphism}
+\section{A Practical Type System for Clojure}
+
+\subsection{Higher-rank polymorphism}
 
@@
 
-\section{Using negative filters}
+\subsection{Using negative filters}
 
@@
 
-\section{Multimethods}
+\subsection{Multimethods}
 
@@
 
-\section{Extending occurrence typing}
+\subsection{Extending occurrence typing}
 
@@
                       \alt \true{} \alt \false{} \alt \nil{}
+                      \alt \class{}
                       \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
                       \alt {\ifexp {\e{}} {\e{}} {\e{}}}
-                      \alt {\doexp {\e{}} {\e{}}}
-                      \alt {\fnexp {\x{}} {\t{}} {\e{}}}
+                      \alt
+                      {\doexp {\e{}} {\e{}}}
+                      \\ &\alt& 
+                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
+                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                 &\mbox{Expressions} \\ \\
-  \s{}, \t{}    &::=& (\class{}\ \overrightarrow{\t{}})
-                      \alt (\ma{\Value\ \v})
-                      \alt (\cup\ \overrightarrow{\t{}})
+  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
+                      \alt {\Value \v{}}
+                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                       \alt (\cap\ \overrightarrow{\t{}})
                       \alt (\Not\ \t{})
-                      \alt (\All\ [\overrightarrow{\x{}}] \t{})
-                      \alt (\TFn\ [\overrightarrow{\x{}}] \t{})
-                &\mbox{Types}
+                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
+                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
+                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
+                                   {\t{}}
+                                   {\filterset {\prop{}} {\prop{}}}
+                                   {\object{}}
+                                   {\prop{}}}
+                      \\ &\alt&
+                      {\ArrayTwo {\t{}} {\t{}}}
+                &\mbox{Types} \\ \\
+  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+                      \alt {\impprop {\prop{}} {\prop{}}}
+                      \alt {\andprop {\prop{}} {\prop{}}}
+                      \alt {\orprop {\prop{}} {\prop{}}}
+                      \alt \topprop{}
+                      \alt \botprop{}
+                &\mbox{Propositions} \\
+  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
+                      \alt \emptyobject{}
+                &\mbox{Objects} \\
+  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
+                &\mbox{Paths} \\
+  \pesyntax{}   &::=& \classpe{}
+                &\mbox{Path Elements} \\ \\
+                % TODO objects with path
 %  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
 %              &\mbox{Constants}
+  \Gamma       &::=& \overrightarrow{\prop{}}
+               &\mbox{Environments}
 \end{altgrammar}
 $$
-\caption{Syntax of Expressions, Types}
+\caption{Syntax of Terms, Types, Propositions, and Objects}
 \end{figure*}
@@
 \infer [T-Var]
-{ \inpropenv {\propenv{}} {\isprop {\x} {\tau}}}
-{ \judgement {\propenv{}} {{\hastype {\x{}} {\t{}}}} {{\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}}
+{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
+{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
              {\emptyobject{}} {\topprop{}}}
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}
+{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}
+{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}
 
 \infer [T-Do]
-{ \judgement {\propenv{}} {{\hastype {\e1} {\t1}}} {{\filterset {\prop{1+}} {\prop2}}} {\object{1}} {{\prop{f1}}}
-\\
-  \judgement {\propenv{}, \prop{f1}} {{\hastype {\e2} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object{}} {\prop{f2}}
+{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
+             {\flowprop {\prop{1}}}
+\\\\
+  \judgement {\propenv{}, {\flowprop {\prop{1}}}} {\hastype {\e2} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} 
+             {\object{}} {\flowprop {\prop{2}}}
   }
-{ \judgement {\propenv{}} {{\hastype {\doexp {\e1} {\e2}} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object} 
-             {{\andprop {\prop{f1}} {\prop{f2}}}}}
+{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e2}} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} {\object} 
+             {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}}
 
 \infer [T-If]
+{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+             {\object{1}} {\flowprop {\prop{1}}} 
+             \\\\
+  \judgement {\propenv{}} {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
+             {\object{}} {\flowprop {\prop{2}}}
+  \\\\
+  \judgement {\propenv{}} {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
+             {\object{}} {\flowprop {\prop{3}}}
+             }
+{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
+            % do we need to include prop1+/- here?
+             {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
+                         {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
+             {\object{}} {\orprop {\flowprop {\prop{2}}}
+                                  {\flowprop {\prop{3}}}} }
 
-%\infer [T-App]
+\infer [T-Let]
+{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+             {\object{1}} {\flowprop {\prop{1}}}
+  \\\\
+  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
+              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
+              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
+             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object{}} {\flowprop {\prop{2}}}
+             }
+{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
+             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                          {\object{1}}
+                          {\x{}}}
+             {\replacefor {\object{}} 
+                          {\object{1}}
+                          {\x{}}}
+             {\replacefor {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}
+                          {\object{1}}
+                          {\x{}}}}
+
+\infer [T-Throw]
+{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object{}} {\flowprop {\prop{}}}}
+{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
+             {\emptyobject{}} {\botprop} }
+
+\infer [T-Fn]
+{}
+{ TODO }
+
+\infer [T-New]
+{ }
+{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
+             {\filterset {\topprop{}} {\topprop{}}}
+             {\emptyobject}
+             {\topprop{}}}
 
@@
 
-\section{Java interoperability}
+\begin{figure*}
+$$
+\begin{altgrammar}
+{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
+\\
+{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
+    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
+            {\javatotc {\class{2}} {\true{}}}
+            {\filterset {\topprop{}} {\topprop{}}}
+            {\emptyobject{}}
+            {\topprop{}}}
+\\
+{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
+    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
+            {\javatotc {\class{2}} {\false{}}}
+            {\filterset {\topprop{}} {\topprop{}}}
+            {\emptyobject{}}
+            {\topprop{}}}
+\\\\
+
+{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
+\\
+{\javatotc {\class{}} {\false{}}} &=& \class{}
+\\
+{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \true{}}}
+\\
+{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \false{}}
+
+\end{altgrammar}
+$$
+\caption{Converting Java types to Typed Clojure types}
+\end{figure*}
+
+\subsection{Java interoperability}
 
@@
 
-\subsection{Arrays}
+\subsubsection{Arrays}
 

### 2013-07-29 schemeworkshop13.tex
more stuff

@@
 %\usepackage{url}
-%\usepackage[style=alphabetic, sorting=nyt]{biblatex}
-%\addbibresource{bibliography.bib}
+%\usepackage{biblatex}
+\include{bibliography.bib}
 
@@
 \usepackage{amsmath}
+
+\newcommand\smallsection[1]{\bf \emph{#1}}
 
@@
 \title{Typed Clojure}
-\subtitle{A Practical Type System for Clojure based on Typed Racket}
+\subtitle{A Practical Type System for Clojure, based on Typed Racket}
 
 \authorinfo{Ambrose Bonnaire-Sergeant}
-           {}
+           {University of Western Australia}
            {abonnairesergeant@gmail.com}
@@
 % Things todo
-% - small calculus based on previous TR formalisation
+% - DONE small calculus based on previous TR formalisation
 %   that demos the ``flow'' environment & not/intersection type
@@
 \section{A Practical Type System for Clojure}
+
+\subsection{Based on Typed Racket}
+
+The design, implementation and theory developed for the Typed Racket project
+has proven to be a valuable base for a type system for Clojure.
+
+{\smallsection {Occurrence typing}}
+Tobin-Hochstadt and Felleisen~\cite{TF08,TF10} developed \emph{occurrence
+typing}, which enables Typed Racket to infer better types at branches.
+Typed Clojure uses occurrence typing in a similar way to Typed Racket.
+We extend and apply occurrence typing in several interesting ways discussed
+in Section [?].
+
+{\smallsection {Practical Variable-Arity Polymorphism}}
+Functions with non-trivial variable parameters are common in Racket.
+For example, Racket provides \emph{map} which takes a function and a
+variable number of collections and applies the function simultaneously
+to each element of the provided collections, returning a list of results.
+Strickland, Tobin-Hochstadt and Felleisen%~\cite{STF09}
+developed a practical system that handles advanced variable parameters
+which can handle applications of functions like \emph{map}.
+
+Clojure has a similar emphasis on variable-arity functions. In some ways,
+Clojure's core library encourages even more complicated variable-parameter schemes.
+The \emph{assoc} core function, for example, takes three parameters and
+then a quantity of variable parameters that is a multiple of two.
+This is beyond what Typed Racket can currently handle.
+
+Functions like \emph{map} are common in Clojure, so we provide an implementation
+of variable-arity polymorphism which has similar capabilities as Typed Racket's
+implementation.
+
+{\smallsection {Local Type Inference}}
+
+% TODO references
+% why is TR such a good base?
+% - immutability
+% - common lisp ancestry
+% differences?
+% - Clojure is built on JVM
+% - interop with JVM
+% - Clojure's idiomatic primitives are different
+% - multimethods + protocols
+% - less sophisticated macro system
+%  - not an issue
+%  - implementation difference, AST walking vs syntax walking
 
@@
 - explain descrepency btwn TC syntax and formal
-  TR calculs syntax. eg. :object/:filter vs over arrow
+  TR calculs syntax. eg. :object/:filter 
+    vs over arrow
 \end{verbatim}
@@
 { \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
-             {\emptyobject{}} {\topprop{}}}
+             {\x{}} {\topprop{}}}
 
@@
 arrays covariant in their argument, a well documented source of static
-unsoundness.
+unsoundness. Bracha~\cite{Bra98} summarises Java's approach to maintaining
+soundness at runtime, which involves all array writes being checked by
+runtime assertions.
+
+This approach fits Java's type system, but we can do better in a more powerful
+type system like Typed Clojure. Our goal is to catch all type-incorrect array
+writes at compile time so the type system can do more to help Clojure programmers
+use arrays, especially those being passed from foreign Java code.
 
 Our basic approach is to make our array types \emph{bivariant}. Array types
-look like \lstinline|(Array2 w r)| and
+look like {\ArrayTwo {\t{w}} {\t{r}}} and
 are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
 and a covariant parameter for output (reading).
+This type can write type {\t{w}} and read type {\t{r}}.
 
@@
 We can get the same effect by setting our input and output
-parameters to the same type. For example, \lstinline|(Array2 Number Number)|
-(or equivalently, \lstinline|(Array Number)|)
-in Typed Clojure is similar to invariant array types of \lstinline|Number| in languages like Scala.
+parameters to the same type. For example, {\ArrayTwo {\Number} {\Number}}
+(or equivalently, {\Array {\Number}}
+in Typed Clojure is similar to invariant array types of \Number in languages like Scala.
 
@@
 to specify \emph{read-only} arrays. Crucially, our type system features an
-explicit bottom type \lstinline|Nothing|, allowing a read-only \lstinline|Number| array
+explicit bottom type \lstinline|Nothing|, enabling a read-only \lstinline|Number| array
 to be of type \lstinline|(Array2 Nothing Number)|.
@@
 what makes array covariance unsound in Java.
-Array covariance lets us lie about the type of an array so the consumer
+Array covariance about the type of an array so the consumer
 of an array cannot tell the actual type of the array when examining a type
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 ...
@@
 ...
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-%\bibliographystyle{abbrvnat}
-%
-%% The bibliography should be embedded for final submission.
-%
+\bibliographystyle{abbrvnat}
+
+% The bibliography should be embedded for final submission.
+
+\bibliography{bibliography}
+
 %\begin{thebibliography}{}

### 2013-07-30 schemeworkshop13.tex
intro

@@
 
+We present a practical type system for Clojure
+
 The main result of this paper is to demonstrate that Typed Racket's design
@@
 
-\section{A Practical Type System for Clojure}
+\section{Overview of Typed Clojure}
 
-\subsection{Based on Typed Racket}
+Typed Clojure is a gradual type system for Clojure, designed to
 
-The design, implementation and theory developed for the Typed Racket project
-has proven to be a valuable base for a type system for Clojure.
+{\smallsection {Based on Typed Racket}}
+Initially, the similarities between untyped Racket and Clojure and Typed Racket's 
+ability to type check Racket code led us to investigate a similar type system for Clojure.
+After two years of development, the solid basis of Typed Racket 
+helps us type check many Clojure idioms without significant differences
+in implementation or theory. We found that extending Typed Clojure to check
+those idioms that have no obvious Racket equivalent did not significantly alter the structure
+of the type system.
 
 {\smallsection {Occurrence typing}}
-Tobin-Hochstadt and Felleisen~\cite{TF08,TF10} developed \emph{occurrence
-typing}, which enables Typed Racket to infer better types at branches.
-Typed Clojure uses occurrence typing in a similar way to Typed Racket.
-We extend and apply occurrence typing in several interesting ways discussed
-in Section [?].
+\citet{TF08,TF10} developed \emph{occurrence typing}, which helps improve types at branches.
+Typed Clojure uses occurrence typing in a similar way to Typed Racket, with
+some extensions (discussed in in Section [?]). %FIXME
 
@@
 to each element of the provided collections, returning a list of results.
-Strickland, Tobin-Hochstadt and Felleisen%~\cite{STF09}
-developed a practical system that handles advanced variable parameters
+\citet*{STF09} developed a practical system that handles advanced variable parameters
 which can handle applications of functions like \emph{map}.
@@
 then a quantity of variable parameters that is a multiple of two.
-This is beyond what Typed Racket can currently handle.
+This is beyond what Typed Racket (and Typed Clojure) can currently handle.
 
@@
 {\smallsection {Local Type Inference}}
+We use Pierce and Turner's Local Type Inference~\cite{PT00} to infer some polymorphic
+applications. Our implementation is based on Typed Racket's, which has extensions
+to support applications of polymorphic variable-arity functions like \emph{map}.
+
+{\smallsection {Unions and intersections}}
+Like Typed Racket, we include union and ordered intersection types. Unions define
+a least-upper-bound for a set of types. For example, we can express a type that is
+either \Number or \Symbol by including them in a union: {\Union {\Number} {\Symbol}}
+
+Ordered intersections (described further by \citet{SA+12})
+are used for overloading function types. We can express a function that takes
+a \Number and returns a \Symbol, and vice-versa with an ordered intersection function type:
+
+\begin{lstlisting}[label=lst:ordered]
+(Fn [Number -> Symbol]
+    [Symbol -> Number])
+\end{lstlisting}
+
+As our intersections are \emph{ordered}, we can express fine invariants in the
+case where arity parameter types overlap. Similar to a pattern match, earlier arities 
+are tried first, and the first arity to match ``wins''.
+
+For example, applying an \lstinline|Integer| argument to a function of type
+
+\begin{lstlisting}[label=lst:ordered]
+(Fn [Integer -> Integer]
+    [Number -> Number])
+\end{lstlisting}
+
+returns an \lstinline|Integer|. Reversing the arities however gives
+type \lstinline|Number|, because the arity taking a \lstinline|Number|
+always matches first.
+
+{\smallsection {Hosted on the Java Virtual Machine}}
+Clojure is built to run on the Java Virtual Machine (JVM),
+offering good interoperability with existing Java code.
+Typed Clojure helps programmers correctly call Java code
+by integrating with Java's type system.
+
+We give Java arrays and Java's \emph{null} special treatment
+when involved with interoperability. Arrays are treated as \emph{read-only}
+when sourced from Java methods, discussed in Section \ref{sec:arrays}.
+We are explicit, and conservative by default, in the positions where
+Java's \emph{null} can be passed, discussed in Section \ref{sec:null}.
 
@@
 \subsubsection{Arrays}
+\label{sec:arrays}
 
@@
 parameters to the same type. For example, {\ArrayTwo {\Number} {\Number}}
-(or equivalently, {\Array {\Number}}
+(or equivalently, {\Array {\Number}})
 in Typed Clojure is similar to invariant array types of \Number in languages like Scala.
@@
 what makes array covariance unsound in Java.
+\begin{verbatim}
+FIXME
 Array covariance about the type of an array so the consumer
@@
 signature.
+\end{verbatim}
 
@@
 
-To emphasise, this is a static type error. Errors like this help Clojure programmers
+To emphasise, Typed Clojure throws a static type error. Errors like this help Clojure programmers
 use foreign Java libraries more correctly.
@@
 \subsection{Handling null}
+\label{sec:null}
 

### 2014-08-28 schemeworkshop13.tex
add more type rules

@@
 \title{Typed Clojure}
-\subtitle{A Practical Type System for Clojure, based on Typed Racket}
+\subtitle{An Optional Type System for Clojure, based on Typed Racket}
 
@@
            {University of Western Australia}
-           {rowan@csse.uwa.edu.au}
+           {rowan.davies@uwa.edu.au}
 
@@
 \begin{abstract}
+Clojure is a dynamically typed language hosted on the Java 
+Virtual Machine.
 Typed Racket is a valuable starting point for
@@
 
-\section{Typed Clojure}
+\section{Clojure with static typing}
+
+% current situation
+
+Programming is an error prone activity, especially in the presence of
+multiple languages working together. Over the past decade, it has become increasingly
+frequent to extend existing dynamically typed languages with optional type systems.
+The guarantees that a sound type system provides can be the difference between
+choosing a language that supports a type system or one that does not.
+
+% presenting clojure
+
+Clojure is a dynamically typed language built to run on the Java Virtual Machine (JVM)
+which has recently gained popularity as an alternative JVM language.
+It offers the flexibility of a Lisp dialect, including Common Lisp-style macros,
+emphasising a functional style via a standard library of immutable datastructures. 
+Clojure has good interoperability with existing Java code, allowing programmers to
+use existing Java libaries without leaving Clojure.
+
+%problem?
+
+Clojure
 
@@
 
+% list the delta from Typed Racket
+
 \section{Overview of Typed Clojure}
 
-Typed Clojure is a gradual type system for Clojure, designed to
+Typed Clojure is a gradual type system for Clojure. It is designed
+to type check normal Clojure code by adding annotations. It is implemented
+as a library, and can be seamlessly included in any Clojure project; no
+separate compiler or language is needed.
 
@@
 
-\begin{lstlisting}[label=lst:ordered]
+\begin{lstlisting}
 (Fn [Integer -> Integer]
@@
 
+\section{A Formal Model of Typed Clojure}
+
+We describe our type system in a similar way to Typed Racket (\citet{TF10}).
+The main judgement 
+{\judgement {\Gamma} {\hastype {\e{}} {\t{}}} 
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+            {\object{}} {\flowprop {\prop{}}}}
+says expression \e{} is of type \t{} in environment $\Gamma$, with 
+then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}},
+object \object{} and flow proposition {\flowprop {\prop{}}}.
+
 \subsection{Higher-rank polymorphism}
@@
 types.
+% why include them?
+% examples, involving generic type operators for monads & conduits
+% useful for generic programming, similar to Hashell type classes
 
@@
                                    {\filterset {\prop{}} {\prop{}}}
-                                   {\object{}}
-                                   {\prop{}}}
+                                   {\object{}}}
                       \\ &\alt&
                       {\ArrayTwo {\t{}} {\t{}}}
+                      \alt {\HMapc {TODO}}
+                      \alt {\HMapp {TODO} {TODO}}
+
                 &\mbox{Types} \\ \\
@@
                 &\mbox{Paths} \\
-  \pesyntax{}   &::=& \classpe{}
+  \pesyntax{}   &::=& \classpe{} \alt \keype{k}
                 &\mbox{Path Elements} \\ \\
@@
 { \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
-{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
-             {\x{}} {\topprop{}}}
+{ \judgementfour {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
+             {\x{}}}
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} {\topprop{}}}
+{ \judgementfour {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}
+{ \judgementfour {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}
+{ \judgementfour {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
 
 \infer [T-Do]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
-             {\flowprop {\prop{1}}}
+{ \judgementfour {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
 \\\\
-  \judgement {\propenv{}, {\flowprop {\prop{1}}}} {\hastype {\e2} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} 
-             {\object{}} {\flowprop {\prop{2}}}
+  \judgementfour {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
+                 {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} 
+                 {\object{}}
   }
-{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e2}} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} {\object} 
-             {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}}
+{ \judgementfour {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
+             }
 
 \infer [T-If]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}} {\flowprop {\prop{1}}} 
+{ \judgementfour {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+                 {\object{1}} 
              \\\\
-  \judgement {\propenv{}} {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
-             {\object{}} {\flowprop {\prop{2}}}
+  \judgementfour {\propenv{}, {\thenprop {\prop{1}}}}
+                 {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
+                 {\object{}}
   \\\\
-  \judgement {\propenv{}} {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
-             {\object{}} {\flowprop {\prop{3}}}
+  \judgementfour {\propenv{}, {\elseprop {\prop{1}}}}
+                 {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
+                 {\object{}}
              }
-{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
-            % do we need to include prop1+/- here?
+{ \judgementfour {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
                  {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
                              {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
-             {\object{}} {\orprop {\flowprop {\prop{2}}}
-                                  {\flowprop {\prop{3}}}} }
+                 {\object{}} }
 
 \infer [T-Let]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}} {\flowprop {\prop{1}}}
+{ \judgementfour {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+             {\object{1}}
   \\\\
-  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
+  \judgementfour {\propenv{}, {\isprop {\s{}} {\x{}}}, 
               {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
@@
              {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} {\flowprop {\prop{2}}}
+             {\object{}} 
              }
-{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
+{ \judgementfour {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
              {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
@@
                           {\x{}}}
-             {\replacefor {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}
-                          {\object{1}}
-                          {\x{}}}}
+             }
 
 \infer [T-Throw]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} {\flowprop {\prop{}}}}
-{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
-             {\emptyobject{}} {\botprop} }
+{ \judgementfour {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                 {\object{}}
+}
+{ \judgementfour {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
+             {\emptyobject{}} }
 
-\infer [T-Fn]
+\infer [T-App]
+{ \judgementfour {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
+                                                       {\t{}}
+                                                       {\filterset {\thenprop {\prop{f}}}
+                                                                   {\elseprop {\prop{f}}}}
+                                                       {\object{f}}}}
+                {\filterset {\thenprop {\prop{}}}
+                            {\elseprop {\prop{}}}}
+                {\object{}}
+  \\\\
+  \judgementfour {\propenv{}}
+                 {\hastype {\ep{}} {\s{}}}
+                 {\filterset {\thenprop {\propp{}}}
+                             {\elseprop {\propp{}}}}
+                 {\objectp{}} 
+}
+{ \judgementfour {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
+                                        {\replacefor {\t{}}
+                                                     {\objectp{}}
+                                                     {\x{}}}}
+                 {\replacefor {\filterset {\thenprop {\prop{f}}}
+                                          {\elseprop {\prop{f}}}}
+                              {\objectp{}}
+                              {\x{}}}
+                 {\replacefor {\object{f}}
+                              {\objectp{}}
+                              {\x{}}}
+}
+
+\infer [T-Abs]
+{}
+{ TODO }
+
+\infer [T-Try]
 {}
@@
 \infer [T-New]
+{ TODO }
+{ \judgementfour {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
+             {\filterset {\topprop{}} {\botprop{}}}
+             {\emptyobject}
+             }
+
+\infer [T-Field]
+{ TODO }
 { }
-{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
-             {\filterset {\topprop{}} {\topprop{}}}
-             {\emptyobject}
-             {\topprop{}}}
+
+\infer [T-Method]
+{ TODO }
+{ }
+
+\infer [T-KwInvoke]
+{}
+{ TODO }
+
+\infer [T-HMapInvoke]
+{}
+{ TODO }
+
+\infer [T-HMapFull]
+{}
+{ TODO }
+
+\infer [T-HMapAbsent]
+{}
+{ TODO }
+
+\infer [T-HMapBoth]
+{}
+{ TODO }
+
+\infer [T-Defmulti]
+{}
+{ TODO }
+
+\infer [T-Defmethod]
+{}
+{ TODO }
+
+\infer [T-InstanceOf]
+{}
+{ TODO }
+
+\infer [T-Count]
+{}
+{ TODO }
+
+\infer [T-Nth]
+{}
+{ TODO }
+
+\infer [T-Class]
+{}
+{ TODO }
+
+\infer [T-Subsume]
+{}
+{ TODO }
 
@@
             {\emptyobject{}}
-            {\topprop{}}}
+            }
 \\
@@
             {\emptyobject{}}
-            {\topprop{}}}
+            }
 \\\\
@@
 \\
-{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \true{}}}
+{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}}
 \\
-{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \false{}}
+{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}
 
@@
 We currently do not support Java generics
+% because generics don't "exist" on the JVM
+% clojure also doesn't use them
 

### 2014-08-31 schemeworkshop13.tex
more typing judgements

@@
 \authorinfo{Ambrose Bonnaire-Sergeant}
-           {University of Western Australia}
-           {abonnairesergeant@gmail.com}
+           {Indiana University}
+           {abonnair@indiana.edu}
 \authorinfo{Rowan Davies}
@@
            {rowan.davies@uwa.edu.au}
+\authorinfo{Sam Tobin-Hochstadt}
+           {Indiana University}
+           {samth@indiana.edu}
 
@@
               {\elseprop {\prop{}}}}
-            {\object{}} {\flowprop {\prop{}}}}
+            {\object{}}}
 says expression \e{} is of type \t{} in environment $\Gamma$, with 
-then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}},
-object \object{} and flow proposition {\flowprop {\prop{}}}.
+then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
+and object \object{}.
 
@@
                       {\ArrayTwo {\t{}} {\t{}}}
-                      \alt {\HMapc {TODO}}
-                      \alt {\HMapp {TODO} {TODO}}
+                      \alt {\HMapc {\mandatory}}
+                      \alt {\HMapp {\mandatory} {\absent}}
 
@@
 \begin{mathpar}
-\infer [T-Var]
+\infer [T-Local]
 { \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
-{ \judgementfour {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
+{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
              {\x{}}}
@@
 {}
-{ \judgementfour {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }
+{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }
 
@@
 {}
-{ \judgementfour {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
+{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
 
@@
 {}
-{ \judgementfour {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
+{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
 
 \infer [T-Do]
-{ \judgementfour {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
+{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
 \\\\
-  \judgementfour {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
+  \judgement {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
                  {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} 
@@
   }
-{ \judgementfour {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
+{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
              }
@@
 \infer [T-If]
-{ \judgementfour {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
                  {\object{1}} 
              \\\\
-  \judgementfour {\propenv{}, {\thenprop {\prop{1}}}}
+  \judgement {\propenv{}, {\thenprop {\prop{1}}}}
                  {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
@@
   \\\\
-  \judgementfour {\propenv{}, {\elseprop {\prop{1}}}}
+  \judgement {\propenv{}, {\elseprop {\prop{1}}}}
                  {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
@@
              }
-{ \judgementfour {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
+{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
                  {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
@@
 \infer [T-Let]
-{ \judgementfour {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
              {\object{1}}
   \\\\
-  \judgementfour {\propenv{}, {\isprop {\s{}} {\x{}}}, 
+  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
               {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
@@
              }
-{ \judgementfour {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
+{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
              {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
@@
 \infer [T-Throw]
-{ \judgementfour {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
                  {\object{}}
 }
-{ \judgementfour {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
+{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop{}} {\botprop{}}}
              {\emptyobject{}} }
@@
 \infer [T-App]
-{ \judgementfour {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
+{ \judgement {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
                                                        {\t{}}
@@
   \\\\
-  \judgementfour {\propenv{}}
+  \judgement {\propenv{}}
                  {\hastype {\ep{}} {\s{}}}
@@
 }
-{ \judgementfour {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
+{ \judgement {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
                                         {\replacefor {\t{}}
@@
 \infer [T-Try]
-{}
-{ TODO }
+{ \judgement {\propenv{}} {\hastype {\e{1}} {\t{}}}
+                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                 {\object{}}
+\\\\
+  \judgement {\propenv{}, {\isprop {\x{}} {\class{1}}}}
+             {\hastype {\e{2}} {\t{}}}
+                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                 {\object{}}
+\\
+  \judgement {\propenv{}}
+             {\hastype {\e{3}} {\t{3}}}
+                 {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
+                 {\object{3}}
+}
+{ \judgement {\propenv{}} {\hastype {\tryexp {\e{1}} {\class{1}} {\x{}} {\e{2}} {\e{3}}} {\t{}}}
+                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                 {\object{}}
+                 }
 
 \infer [T-New]
-{ TODO }
-{ \judgementfour {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
+{}
+{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\class{}}}
              {\filterset {\topprop{}} {\botprop{}}}
@@
 
+\infer [T-KeywordMap]
+{ \judgement {\propenv{}} {\hastype {\overrightarrow{{\e{v}}}} {\overrightarrow {{\t{v}}}}}
+             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object}}
+{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{k_{v}}\ {\e{v}}}\}} {\HMapc {\overrightarrow{{k_{v}}\ {\t{v}}}}}}
+             {\filterset {\topprop{}} {\botprop{}}}
+             {\emptyobject}}
+
+\infer [T-Map]
+{ 
+\judgement {\propenv{}} {\hastype {\overrightarrow{{\e{k}}}} {\overrightarrow {{\t{k}}}}}
+             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object}
+\\
+\judgement {\propenv{}} {\hastype {\overrightarrow{{\e{v}}}} {\overrightarrow {{\t{v}}}}}
+             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object}}
+{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{\e{k}}\ {\e{v}}}\}}
+                                    {\IPersistentMap {\Unionsplice {\overrightarrow{{t_{k}}}}}
+                                                     {\Unionsplice {\overrightarrow{{t_{v}}}}}}}
+             {\filterset {\topprop{}} {\botprop{}}}
+             {\emptyobject}}
+
 \infer [T-HMapFull]
 {}
-{ TODO }
+{ }
 
@@
 \infer [T-Defmulti]
-{}
-{ TODO }
+{ \inpropenv {\propenv{}} {\isprop {\x{}} 
+                                   {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{m}}
+                                    {\filterset {\thenprop {\prop{m}}}
+                                                {\elseprop {\prop{m}}}}
+                                    {\object{m}}}}
+\\\\
+  \judgement {\propenv{}} {\hastype {\e{d}} {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{r}}
+                                                    {\filterset {\thenprop {\prop{d}}}
+                                                                {\elseprop {\prop{d}}}}
+                                                    {\object{d}}}}
+             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object{}}
+\\\\
+{\adddispatchfn
+{\x{}}
+{\e{d}}}
+}
+{ \judgement {\propenv{}} {\hastype {\defmultiexp {\x{}} {\e{d}}} {\MultiFn{}}}
+             {\filterset {\topprop{}} {\botprop{}}}
+             {\emptyobject}
+}
 
 \infer [T-Defmethod]
-{}
-{ TODO }
+{ \inpropenv {\propenv{}} {\isprop {\x{m}}
+                                   {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{m}}
+                                    {\filterset {\thenprop {\prop{m}}}
+                                                {\elseprop {\prop{m}}}}
+                                    {\object{m}}}}
+\\
+\getdispatchfn {\x{m}} {\e{d}}
+\\\\
+\judgement {\propenv{}, {\overrightarrow {\isprop {\x{p}} {\t{p}}}}}
+           {\hastype {\appexp {\e{d}} {\overrightarrow {\x{p}}}}
+                                  {\t{d}}}
+           {\filterset {\thenprop {\prop{d}}}
+                       {\elseprop {\prop{d}}}}
+           {\object{d}}
+\\\\
+\judgement {\propenv{}, {\thenprop {\prop{d}}}}
+           {\hastype {\e{b}} {\t{m}}}
+           {\filterset {\thenprop {\prop{m}}}
+                       {\elseprop {\prop{m}}}}
+           {\object{m}}
+}
+{ \judgement {\propenv{}} {\hastype {\defmethodexp {\x{m}} {\e{v}} {\overrightarrow{\x{p}}}
+                                                   {\e{b}}}
+                                    {\MultiFn{}}}
+             {\filterset {\topprop{}} {\botprop{}}}
+             {\emptyobject}
+}
 
 \infer [T-InstanceOf]
-{}
-{ TODO }
+{ \judgement {\propenv{}} 
+             {\hastype {\e{}} {\t{}}}
+             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object}
+}
+{ \judgement {\propenv{}} {\hastype {\instanceapp {\class{}} {\e{}}}
+                                    {\Boolean}}
+             {\filterset {\isprop  {\class{}} {\object{}}}
+                         {\notprop {\class{}} {\object{}}}}
+             {\emptyobject}
+}
 
-\infer [T-Count]
-{}
-{ TODO }
-
-\infer [T-Nth]
-{}
-{ TODO }
-
-\infer [T-Class]
-{}
-{ TODO }
+\infer [T-Isa]
+{ \isacompare {\e{1}} {\e{2}}
+             {\filterset {\thenprop {\prop{}}}
+                         {\elseprop {\prop{}}}}
+}
+{ \judgement {\propenv{}} {\hastype {\isaapp {\e{1}} {\e{2}}}
+                                    {\Boolean{}}}
+             {\filterset {\thenprop {\prop{}}}
+                         {\elseprop {\prop{}}}}
+             {\emptyobject}}
 
 \infer [T-Subsume]
-{}
-{ TODO }
+{ \judgement {\propenv{}} {\hastype {\e{}} {\t{}}}
+             {\filterset {\thenprop {\prop{}}}
+                         {\elseprop {\prop{}}}}
+             {\object}
+\\\\
+\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
+\\
+\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+\\\\
+\inpropenv {} {\issubtype {\t{}} {\tp{}}}
+\\
+\inpropenv {} {\issubtype {\object{}} {\objectp{}}}
+}
+{ \judgement {\propenv{}} {\hastype {\e{}} {\tp{}}}
+             {\filterset {\thenprop {\propp{}}}
+                         {\elseprop {\propp{}}}}
+             {\objectp}}
 
@@
 
-{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
+{\javatotc {\pint{}} {b}} &=& \Integer{}
 \\
-{\javatotc {\class{}} {\false{}}} &=& \class{}
+{\javatotc {\Void{}} {b}} &=& \Nil
+\\
+{\javatotc {\pvoid{}} {b}} &=& \Nil
+\\
+{\javatotc {\pchar{}} {b}} &=& \Character{}
+\\
+{\javatotc {\pdouble{}} {b}} &=& \Double{}
+\\
+{\javatotc {\JavaArray {\pint{}}} {b}} &=& {\Union {\Nil} {\ArrayP {\pint{}}}}
+\\
+{\javatotc {\JavaArray {\pdouble{}}} {b}} &=& {\Union {\Nil} {\ArrayP {\pdouble{}}}}
 \\
@@
 {\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}
+\\
+{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
+\\
+{\javatotc {\class{}} {\false{}}} &=& \class{}
 
@@
 This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
-We provide a singleton type \lstinline|nil|, which contains just the value \lstinline|nil|,
+We provide a singleton type \Nil{}, which contains just the value \lstinline|nil|,
 Clojure's equivalent of \emph{null}.

### 2014-09-30 icfp15-typedclojure.tex
rename to icfp
type hints

--- 2014-08-31 schemeworkshop13.tex
@@
 \begin{altgrammar}
-  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
+  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} \alt {\abs {\x a} {\e{}}}
                       \alt \true{} \alt \false{} \alt \nil{}
@@
                       \alt {\ifexp {\e{}} {\e{}} {\e{}}}
+                      \\ &\alt& 
+                      {\doexp {\e{}} {\e{}}}
                       \alt
-                      {\doexp {\e{}} {\e{}}}
-                      \\ &\alt& 
                       {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
                       \alt {\letexp {\x{}} {\e{}} {\e{}}}
+                      \alt {\fieldexp {fld} {\e{}}}
+                      \alt {\methodexp {fld} {\e{}} {\overrightarrow{\e{}}}}
+                      \alt {\newexp {\e{}} {\overrightarrow{\e{}}}}
                 &\mbox{Expressions} \\ \\
   \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
+                      \alt \class\ 
+                      \alt \protocol\ 
+                      \alt {\Instance {\class}}
+                      \alt {\Satisfies {\protocol}}
                       \alt {\Value \v{}}
                       \alt {\Unionsplice {\overrightarrow{\t{}}}}
-                      \alt (\cap\ \overrightarrow{\t{}})
-                      \alt (\Not\ \t{})
-                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
-                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
-                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
+                      \alt {\In {\t{}\ \overrightarrow{\t{}}}}
+                      \\ &\alt&
+                      %\alt (\Not\ \t{})
+                      {\All {\overrightarrow{\x{}}} {\t{}}}
+                      \alt
+                      {\TFn {\overrightarrow{\x{}}} \t{}}
+                      \alt
+                      {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
                                    {\t{}}
@@
                                    {\object{}}}
-                      \\ &\alt&
+                      \alt
                       {\ArrayTwo {\t{}} {\t{}}}
@@
                 &\mbox{Paths} \\
-  \pesyntax{}   &::=& \classpe{} \alt \keype{k}
+  \pesyntax{}   &::=& \classpe{} \alt \keype{k} \alt \countpe{}
                 &\mbox{Path Elements} \\ \\
@@
 \infer [T-Abs]
-{}
-{ TODO }
+{ \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}}
+             {\filterset {\thenprop {\prop{}}}
+                         {\elseprop {\prop{}}}}
+             {\object{}}}
+{ \judgement {\propenv{}} {\hastype {\e{}} {\ArrowOne {\x{}} {\s{}}
+                                                      {\t{}}
+                                                      {\filterset {\thenprop {\prop{}}}
+                                                                  {\elseprop {\prop{}}}}
+                                                      {\object{}}}}
+                {\filterset {\topprop}
+                            {\botprop}}
+                {\object{}}}
 
@@
 \infer [T-New]
-{}
+{ \ctorparams {\overrightarrow {\e{i}}} {\overrightarrow {\t{i}}}
+  \\
+  {\overrightarrow{
+  \judgement {\propenv{}} {\hastype {\e{i}} {\t{i}}} 
+             {\filterset {\prop{}} {\prop{}}}
+         {\object{}}}}^{i}
+           }
 { \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\class{}}}
@@
 \infer [T-Field]
-{ TODO }
-{ }
+{ \judgement {\propenv{}} {\hastype {\e{t}} {\t{t}}}
+             {\filterset {\prop{}} {\prop{}}}
+             {\object{}}
+           \\
+         \fieldtype {\t{t}} {fld} {\t{}}}
+{ \judgement {\propenv{}} {\hastype {\fieldexp {fld} {\e{t}}} {\t{}}}
+             {\filterset {\topprop{}} {\topprop}}
+             {\emptyobject{}}}
 
 \infer [T-Method]
-{ TODO }
-{ }
+{ \judgement {\propenv{}} {\hastype {\e{t}} {\t{t}}}
+             {\filterset {\prop{}} {\prop{}}}
+             {\object{}}
+             \\
+             {\overrightarrow{
+ \judgement {\propenv{}} {\hastype {\e{a}} {\t{a}}}
+             {\filterset {\prop{}} {\prop{}}}
+             {\object{}}
+           }}^{a}
+             \\
+             \methodtype {\t{t}} {\overrightarrow{\t{a}}} {\t{}}
+           }
+{ \judgement {\propenv{}} {\hastype {\methodexp {mth} {\e{t}} {\overrightarrow{\e{a}}}}
+                                    {\t{}}}
+             {\filterset {\topprop{}} {\topprop{}}}
+             {\emptyobject{}}}
 
@@
 
-\infer [T-HMapInvoke]
-{}
-{ TODO }
+%{ \judgement {\propenv{}} {\hastype {\e{m}} {\t{m}}}
+%             {\filterset {\prop{}} {\prop{}}}
+%           {\object{m}}}
+%             \\
+%  \judgement {\propenv{}} {\hastype {\e{k}} {\t{k}}}
+%             {\filterset {\prop{}} {\prop{}}}
+%             {\object{}}
+             %\\\\
+%  \judgement {\propenv{}} {\hastype {\e{d}} {\t{d}}}
+%             {\filterset {\prop{}} {\prop{}}}
+%             {\object{}}
+
+             %\\
+             %\t{k} %= {\Value {k}}
+         %}
+% invoke complete HMap with present key
+\infer [T-HMapCInvokePresent]
+{ \judgement {\propenv{}} {\hastype {\e{m}} {\t{m}}}
+             {\filterset {\prop{}} {\prop{}}}
+           {\object{m}}
+         \\
+  \judgement {\propenv{}} {\hastype {\e{k}} {\t{k}}}
+             {\filterset {\prop{}} {\prop{}}}
+             {\object{}}
+             \\\\
+  \judgement {\propenv{}} {\hastype {\e{d}} {\t{d}}}
+             {\filterset {\prop{}} {\prop{}}}
+             {\object{}}
+             \\
+             {\t{m}} = {\HMapc {\mandatory{}}}
+             \\
+             {\t{k}} = {\Value {k}}
+             \\
+             {(k\ \t{})} \in {\mandatory{}}
+       }
+{ \judgement {\propenv{}} {\hastype {\appexp {\e{m}} {\e{k}\ \e{d}}}
+                                    {\t{}}}
+             {\filterset {\topprop{}} {\topprop{}}}
+             {\path {\keype{k}} {\object{m}}}
+                                  }
 
@@
              {\object}}
-{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{k_{v}}\ {\e{v}}}\}} {\HMapc {\overrightarrow{{k_{v}}\ {\t{v}}}}}}
+{ \judgement {\propenv{}} {\hastype {\curlymap {\overrightarrow{{k_{v}}\ {\e{v}}}}}
+  {\HMapc {\mandatoryset {{\overrightarrow{({k_{v}}\ {\t{v}})}}}}}}
              {\filterset {\topprop{}} {\botprop{}}}
@@
 
-\infer [T-HMapFull]
-{}
-{ }
-
-\infer [T-HMapAbsent]
-{}
-{ TODO }
-
-\infer [T-HMapBoth]
-{}
-{ TODO }
-
 \infer [T-Defmulti]
@@
 
+\end{mathpar}
+\caption{Typing Rules 1}
+\end{figure*}
+
+\begin{figure*}
+\begin{mathpar}
+
 \infer [T-Defmethod]
@@
 
+\infer [T-Subsume]
+{ \judgement {\propenv{}} {\hastype {\e{}} {\t{}}}
+             {\filterset {\thenprop {\prop{}}}
+                         {\elseprop {\prop{}}}}
+             {\object}
+\\\\
+\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
+\\
+\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+\\\\
+\inpropenv {} {\issubtype {\t{}} {\tp{}}}
+\\
+\inpropenv {} {\issubtype {\object{}} {\objectp{}}}
+}
+{ \judgement {\propenv{}} {\hastype {\e{}} {\tp{}}}
+             {\filterset {\thenprop {\propp{}}}
+                         {\elseprop {\propp{}}}}
+             {\objectp}}
+
 \infer [T-InstanceOf]
@@
                                     {\Boolean}}
-             {\filterset {\isprop  {\class{}} {\object{}}}
-                         {\notprop {\class{}} {\object{}}}}
+             {\filterset {\isprop  {{\Instance {\class{}}}} {\object{}}}
+                         {\notprop {{\Instance {\class{}}}} {\object{}}}}
              {\emptyobject}
@@
 
+\infer [T-Satisfies]
+{ \judgement {\propenv{}} 
+             {\hastype {\e{}} {\t{}}}
+             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+             {\object}
+             \\
+             {\tp{}} = {\Satisfies {\protocol{}}}
+}
+{ \judgement {\propenv{}} {\hastype {\satisfiesapp {\protocol{}} {\e{}}}
+                                    {\Boolean}}
+            {\filterset {\isprop  {\tp{}} {\object{}}}
+                        {\notprop {\tp{}} {\object{}}}}
+             {\emptyobject}
+}
 \infer [T-Isa]
@@
 
-\infer [T-Subsume]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\t{}}}
+\end{mathpar}
+\caption{Typing rules 2}
+\end{figure*}
+
+\begin{figure*}
+\begin{mathpar}
+
+\infer [SO-Refl]
+{}
+{ \inpropenv {} {\issubtype {\object{}} {\object{}}}}
+
+\infer [SO-Top]
+{}
+{ \inpropenv {} {\issubtype {\object{}} {\emptyobject{}}}}
+
+\infer [S-Refl]
+{}
+{ \inpropenv {} {\issubtype {\t{}} {\t{}}}}
+
+\infer [S-Top]
+{}
+{ \inpropenv {} {\issubtype {\t{}} {\Top{}}}}
+
+% FIXME fix overrightarrow caption
+\infer [S-IntersectSuper]
+{ {\overrightarrow {\inpropenv {} {\issubtype {\t{}} {\s{i}}}}}^{i}}
+{ \inpropenv {} {\issubtype {\t{}} {\In{{\overrightarrow{\s{}}}^{i}}}}}
+
+\infer [S-IntersectSub]
+{ \exists i.\ {\inpropenv {} {\issubtype {\t{i}} {\s{}}}}}
+{ \inpropenv {} {\issubtype {\In{{\overrightarrow{\t{}}}^{i}}} {\s{}}}}
+
+\infer [S-UnionSuper]
+{ \exists i.\ {\inpropenv {} {\issubtype {\t{}} {\s{i}}}}}
+{ \inpropenv {} {\issubtype {\t{}} {\Unionsplice{{\overrightarrow{\s{}}}^{i}}}}}
+
+% FIXME fix overrightarrow caption
+\infer [S-UnionSub]
+{ {\overrightarrow {\inpropenv {} {\issubtype {\t{i}} {\s{}}}}}^{i}}
+{ \inpropenv {} {\issubtype {\Unionsplice{\overrightarrow{\t{}}^{i}}} {\s{}}}}
+
+\infer [S-Fun]
+{ \inpropenv {} {\issubtype {\sp{}} {\s{}}}
+  \\
+  \inpropenv {} {\issubtype {\t{}} {\tp{}}}
+  \\\\
+  \inpropenv {\thenprop {\prop{}}} {\thenprop {\propp{}}}
+  \\
+  \inpropenv {\elseprop {\prop{}}} {\elseprop {\propp{}}}
+  \\
+  \inpropenv {\object{}} {\objectp{}}
+}
+{ \inpropenv {}
+  {\issubtype
+    {\ArrowOne {\x{}} {\s{}}
+              {\t{}}
               {\filterset {\thenprop {\prop{}}}
                           {\elseprop {\prop{}}}}
-             {\object}
-\\\\
-\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
-\\
-\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-\\\\
-\inpropenv {} {\issubtype {\t{}} {\tp{}}}
-\\
-\inpropenv {} {\issubtype {\object{}} {\objectp{}}}
-}
-{ \judgement {\propenv{}} {\hastype {\e{}} {\tp{}}}
+              {\object{}}}
+    {\ArrowOne {\xp{}} {\sp{}}
+              {\tp{}}
               {\filterset {\thenprop {\propp{}}}
                           {\elseprop {\propp{}}}}
-             {\objectp}}
+              {\objectp{}}}}}
+
+\infer [S-Instance]
+{ \class{},\ \classp{}\ not\ arrays
+\\ \assignable {\class{}} {\classp{}}}
+{ \inpropenv {} {\issubtype {\Instance {\class{}}} {\Instance {\classp{}}}} }
 
 \end{mathpar}
-\caption{Typing Rules}
+\caption{Subtyping rules}
 \end{figure*}
@@
 
-We currently do not support Java generics
-% because generics don't "exist" on the JVM
-% clojure also doesn't use them
+We do not support Java Generics for several reasons. Firstly
+Clojure programmers already use \emph{type hints} to resolve
+Java methods which does not interact with generics. Secondly
+it would complicate the type system to need to consider advanced
+Generics features like existentials and f-bounded polymorphism
+which has complicated other type systems like Scala's.
+Finally Java interop is rare in Clojure code, so the extra
+work needed would not be worth the effort compared to the number
+of new bugs core.typed catches.
+
+Thanks to type erasure, only unparameterised classes, primitive
+types and array classes live at runtime. This simplifies 
+inference for \emph{tools.analyzer}, which uses type hints
+to resolve ambiguities in Java interop like overloaded methods.
+
+\subsection{Clojure type hints}
+
+Clojure comes with a small type system that helps improve
+performance in Clojure code: type hints. They are used to help
+the compiler resolve Java methods and prevent unboxing in
+certain situations. Tags are propagated bidirectionally
+and the programmer is free to add them where they like.
 

### 2014-10-04 icfp15-typedclojure.tex
add 7 place relation and some color

@@
 \usepackage{clj-grammar}
+\usepackage{xcolor}
 %\usepackage{url}
@@
 The main judgement 
-{\judgement {\Gamma} {\hastype {\e{}} {\t{}}} 
+{\judgementsevcol {\mmenv}
+                 {\taenv}
+                 {\propenv}
+               {\hastype {\e{}} {\t{}}}
   {\filterset {\thenprop {\prop{}}}
               {\elseprop {\prop{}}}}
-            {\object{}}}
-says expression \e{} is of type \t{} in environment $\Gamma$, with 
-then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
-and object \object{}.
+  {\object{}}
+  {\mmenvp}}
+says expression \e{} is of type \t{} in the 
+multimethod environment $\mmenv$, the \emph{tools.analyzer} environment
+$\taenv'$
+and proposition environment $\propenv$, with 
+then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}},
+object \object{} and new multimethod environment
+${\mmenvp}$.
 
@@
                       \alt {\fieldexp {fld} {\e{}}}
-                      \alt {\methodexp {fld} {\e{}} {\overrightarrow{\e{}}}}
-                      \alt {\newexp {\e{}} {\overrightarrow{\e{}}}}
+                      \alt {\methodexp {mth} {\e{}} {\overrightarrow{\e{}}}}
+                      \alt {\newexp {\class{}} {\overrightarrow{\e{}}}}
                 &\mbox{Expressions} \\ \\
@@
 { \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
-{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
-             {\x{}}}
+{ \judgementfillcol {\propenv{}} 
+                   {\hastype {\x{}} {\t{}}}
+                   {\filterset {{\notprop {\falsydiff} {\x{}}}} {{\isprop {\falsydiff} {\x{}}}}}
+                   {\x{}}
+                   }
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }
+{ \judgementfillcol {\propenv{}}
+      {{\hastype {\true{}} {\True{}}}}
+      {{\filterset {\topprop{}} {\botprop{}}}}
+      {\emptyobject{}}
+    }
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
+{ \judgementfillcol {\propenv{}}
+      {{\hastype {\false{}} {\False{}}}}
+      {\filterset {\botprop{}} {\topprop{}}}
+    {\emptyobject{}}
+   }
 
@@
 {}
-{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
+{ \judgementfillcol {\propenv{}} 
+      {{\hastype {\nil{}} {\Nil{}}}} 
+      {\filterset {\botprop{}} {\topprop{}}}
+      {\emptyobject{}} 
+    }
 
 \infer [T-Do]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
-\\\\
-  \judgement {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
+\\\\
+  \judgementfillcol {\propenv{}, {\trdiff {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}}
                  {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} 
@@
   }
-{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
+{ \judgementfillcol {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
              }
@@
 \infer [T-If]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
                  {\object{1}} 
              \\\\
-  \judgement {\propenv{}, {\thenprop {\prop{1}}}}
+  \judgementfillcol {\propenv{}, {\thenprop {\prop{1}}}}
                  {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
@@
   \\\\
-  \judgement {\propenv{}, {\elseprop {\prop{1}}}}
+  \judgementfillcol {\propenv{}, {\elseprop {\prop{1}}}}
                  {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
@@
              }
-{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
+{ \judgementfillcol {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
                  {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
@@
 \infer [T-Let]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
              {\object{1}}
   \\\\
-  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
-              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
-              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
+  \judgementfillcol {\propenv{}, {\isprop {\s{}} {\x{}}}, 
+                    {\impprop {\notprop {\falsydiff} {\x{}}} {\thenprop {\prop{1}}}}, 
+                    {\impprop {\isprop {\falsydiff} {\x{}}} {\elseprop {\prop{1}}}}}
              {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
@@
              }
-{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
              {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
@@
 \infer [T-Throw]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-}
-{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop{}} {\botprop{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                 {\object{}}
+}
+{ \judgementfillcol {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\trdiff {\filterset {\botprop{}} {\botprop{}}}}
              {\emptyobject{}} }
@@
 \infer [T-App]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
                                                        {\t{}}
@@
   \\\\
-  \judgement {\propenv{}}
+  \judgementfillcol {\propenv{}}
                  {\hastype {\ep{}} {\s{}}}
@@
 }
-{ \judgement {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
                                         {\replacefor {\t{}}
@@
 \infer [T-Abs]
-{ \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}}
+{ \judgementfillcol {\propenv{}, {\isprop {\s{}} {\x{}}}}
+            {\hastype {\e{}} {\t{}}}
              {\filterset {\thenprop {\prop{}}}
@@
              {\object{}}}
-{ \judgement {\propenv{}} {\hastype {\e{}} {\ArrowOne {\x{}} {\s{}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{}} {\ArrowOne {\x{}} {\s{}}
                                                       {\t{}}
@@
 \infer [T-Try]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\t{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{1}} {\t{}}}
                  {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
@@
 \\\\
-  \judgement {\propenv{}, {\isprop {\x{}} {\class{1}}}}
+  \judgementfillcol {\propenv{}, {\isprop {\x{}} {\class{1}}}}
              {\hastype {\e{2}} {\t{}}}
@@
 \\
-  \judgement {\propenv{}}
+  \judgementfillcol {\propenv{}}
              {\hastype {\e{3}} {\t{3}}}
@@
 }
-{ \judgement {\propenv{}} {\hastype {\tryexp {\e{1}} {\class{1}} {\x{}} {\e{2}} {\e{3}}} {\t{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\tryexp {\e{1}} {\class{1}} {\x{}} {\e{2}} {\e{3}}} {\t{}}}
                  {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
@@
   {\overrightarrow{
-  \judgement {\propenv{}} {\hastype {\e{i}} {\t{i}}} 
+  \judgementfillcol {\propenv{}} {\hastype {\e{i}} {\t{i}}} 
              {\filterset {\prop{}} {\prop{}}}
@@
            }
-{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\class{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\class{}}}
              {\filterset {\topprop{}} {\botprop{}}}
@@
 
+\end{mathpar}
+\caption{Typing Rules 1}
+\end{figure*}
+
+\begin{figure*}
+\begin{mathpar}
+
 \infer [T-Defmulti]
@@
 
-\end{mathpar}
-\caption{Typing Rules 1}
-\end{figure*}
-
-\begin{figure*}
-\begin{mathpar}
-
 \infer [T-Defmethod]
@@
 \infer [T-Subsume]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\t{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{}} {\t{}}}
              {\filterset {\thenprop {\prop{}}}
@@
 }
-{ \judgement {\propenv{}} {\hastype {\e{}} {\tp{}}}
+{ \judgementfillcol {\propenv{}} {\hastype {\e{}} {\tp{}}}
              {\filterset {\thenprop {\propp{}}}

### 2014-10-10 formal-model.tex
link up to ta calculus

--- 2014-08-31 schemeworkshop13.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint,10pt]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{mmm}
-\usepackage{mathpartir}
-\usepackage{clj-grammar}
-%\usepackage{url}
-%\usepackage{biblatex}
-\include{bibliography.bib}
-
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
-
-\usepackage{amsmath}
-
-\newcommand\smallsection[1]{\bf \emph{#1}}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-%\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Typed Clojure}
-\subtitle{An Optional Type System for Clojure, based on Typed Racket}
-
-\authorinfo{Ambrose Bonnaire-Sergeant}
-           {Indiana University}
-           {abonnair@indiana.edu}
-\authorinfo{Rowan Davies}
-           {University of Western Australia}
-           {rowan.davies@uwa.edu.au}
-\authorinfo{Sam Tobin-Hochstadt}
-           {Indiana University}
-           {samth@indiana.edu}
-
-\maketitle
-
-\begin{abstract}
-Clojure is a dynamically typed language hosted on the Java 
-Virtual Machine.
-Typed Racket is a valuable starting point for
-a gradual type system that targets Clojure.
-Building a similar type system for a new language gives the
-designer some flexibility to repurpose and extend features.
-This paper gives an overview of Typed Clojure, concentrating
-on the extensions and differences from Typed Racket. We also
-show where Typed Racket's features were particularly useful
-for type checking non-trivial Clojure idioms.
-\end{abstract}
-
-\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-\keywords
-gradual types, Clojure
-
-\section{Clojure with static typing}
-
-% current situation
-
-Programming is an error prone activity, especially in the presence of
-multiple languages working together. Over the past decade, it has become increasingly
-frequent to extend existing dynamically typed languages with optional type systems.
-The guarantees that a sound type system provides can be the difference between
-choosing a language that supports a type system or one that does not.
-
-% presenting clojure
-
-Clojure is a dynamically typed language built to run on the Java Virtual Machine (JVM)
-which has recently gained popularity as an alternative JVM language.
-It offers the flexibility of a Lisp dialect, including Common Lisp-style macros,
-emphasising a functional style via a standard library of immutable datastructures. 
-Clojure has good interoperability with existing Java code, allowing programmers to
-use existing Java libaries without leaving Clojure.
-
-%problem?
-
-Clojure
-
-Typed Clojure is a gradual type system for Clojure, based on the
-static portion of Typed Racket.
-
-% what are the goals of Typed Clojure?
-% where do they differ from Typed Racket? why?
-
-% Things todo
-% - DONE small calculus based on previous TR formalisation
-%   that demos the ``flow'' environment & not/intersection type
-% - discussion and examples for flow filters
-% - discussion on the not/difference type, real code where it helps
-% - higher-rank polymorphism, show what it enables (eg. monads)
-%   - also show more complicated example, making an extensible
-%     type for clojure.core/conj, and why it's not straightforward
-% - discussion on limitations of local type inference
-% - demonstration of how we check multimethods
-% - dealing with array covariance
-% - note on how we avoid null-pointer exceptions
-% - discussion on why Typed Clojure only uses the ``static'' part
-%   of Typed Racket
-% - the equality filter problem (and draft solution?), a proposition
-%   that knows about binding aliasing. Sounds straightforward?
-
-\section{Contributions}
-
-We present a practical type system for Clojure
-
-The main result of this paper is to demonstrate that Typed Racket's design
-and implementation is applicable to a practical type system for Clojure
-by implementing Typed Clojure.
-
-% list the delta from Typed Racket
-
-\section{Overview of Typed Clojure}
-
-Typed Clojure is a gradual type system for Clojure. It is designed
-to type check normal Clojure code by adding annotations. It is implemented
-as a library, and can be seamlessly included in any Clojure project; no
-separate compiler or language is needed.
-
-{\smallsection {Based on Typed Racket}}
-Initially, the similarities between untyped Racket and Clojure and Typed Racket's 
-ability to type check Racket code led us to investigate a similar type system for Clojure.
-After two years of development, the solid basis of Typed Racket 
-helps us type check many Clojure idioms without significant differences
-in implementation or theory. We found that extending Typed Clojure to check
-those idioms that have no obvious Racket equivalent did not significantly alter the structure
-of the type system.
-
-{\smallsection {Occurrence typing}}
-\citet{TF08,TF10} developed \emph{occurrence typing}, which helps improve types at branches.
-Typed Clojure uses occurrence typing in a similar way to Typed Racket, with
-some extensions (discussed in in Section [?]). %FIXME
-
-{\smallsection {Practical Variable-Arity Polymorphism}}
-Functions with non-trivial variable parameters are common in Racket.
-For example, Racket provides \emph{map} which takes a function and a
-variable number of collections and applies the function simultaneously
-to each element of the provided collections, returning a list of results.
-\citet*{STF09} developed a practical system that handles advanced variable parameters
-which can handle applications of functions like \emph{map}.
-
-Clojure has a similar emphasis on variable-arity functions. In some ways,
-Clojure's core library encourages even more complicated variable-parameter schemes.
-The \emph{assoc} core function, for example, takes three parameters and
-then a quantity of variable parameters that is a multiple of two.
-This is beyond what Typed Racket (and Typed Clojure) can currently handle.
-
-Functions like \emph{map} are common in Clojure, so we provide an implementation
-of variable-arity polymorphism which has similar capabilities as Typed Racket's
-implementation.
-
-{\smallsection {Local Type Inference}}
-We use Pierce and Turner's Local Type Inference~\cite{PT00} to infer some polymorphic
-applications. Our implementation is based on Typed Racket's, which has extensions
-to support applications of polymorphic variable-arity functions like \emph{map}.
-
-{\smallsection {Unions and intersections}}
-Like Typed Racket, we include union and ordered intersection types. Unions define
-a least-upper-bound for a set of types. For example, we can express a type that is
-either \Number or \Symbol by including them in a union: {\Union {\Number} {\Symbol}}
-
-Ordered intersections (described further by \citet{SA+12})
-are used for overloading function types. We can express a function that takes
-a \Number and returns a \Symbol, and vice-versa with an ordered intersection function type:
-
-\begin{lstlisting}[label=lst:ordered]
-(Fn [Number -> Symbol]
-    [Symbol -> Number])
-\end{lstlisting}
-
-As our intersections are \emph{ordered}, we can express fine invariants in the
-case where arity parameter types overlap. Similar to a pattern match, earlier arities 
-are tried first, and the first arity to match ``wins''.
-
-For example, applying an \lstinline|Integer| argument to a function of type
-
-\begin{lstlisting}
-(Fn [Integer -> Integer]
-    [Number -> Number])
-\end{lstlisting}
-
-returns an \lstinline|Integer|. Reversing the arities however gives
-type \lstinline|Number|, because the arity taking a \lstinline|Number|
-always matches first.
-
-{\smallsection {Hosted on the Java Virtual Machine}}
-Clojure is built to run on the Java Virtual Machine (JVM),
-offering good interoperability with existing Java code.
-Typed Clojure helps programmers correctly call Java code
-by integrating with Java's type system.
-
-We give Java arrays and Java's \emph{null} special treatment
-when involved with interoperability. Arrays are treated as \emph{read-only}
-when sourced from Java methods, discussed in Section \ref{sec:arrays}.
-We are explicit, and conservative by default, in the positions where
-Java's \emph{null} can be passed, discussed in Section \ref{sec:null}.
-
-% TODO references
-% why is TR such a good base?
-% - immutability
-% - common lisp ancestry
-% differences?
-% - Clojure is built on JVM
-% - interop with JVM
-% - Clojure's idiomatic primitives are different
-% - multimethods + protocols
-% - less sophisticated macro system
-%  - not an issue
-%  - implementation difference, AST walking vs syntax walking
-
 \section{A Formal Model of Typed Clojure}
 
-We describe our type system in a similar way to Typed Racket (\citet{TF10}).
+We describe our type system in a similar way to Typed Racket (\citet{TF10}),
+with differences highlighted in $\trdiff{\text{blue}}$.
 The main judgement
-{\judgement {\Gamma} {\hastype {\e{}} {\t{}}} 
+
+$
+{\judgementsevcol {\mmenv}
+                 {\taenv}
+                 {\propenv}
+               {\hastype {\e{}} {\t{}}}
   {\filterset {\thenprop {\prop{}}}
               {\elseprop {\prop{}}}}
-            {\object{}}}
-says expression \e{} is of type \t{} in environment $\Gamma$, with 
-then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
-and object \object{}.
+  {\object{}}
+  {\mmenvp{}}}
+$
+
+says expression \e{} is of type \t{} in the 
+multimethod environment $\trdiff{\mmenv{}}$, the \emph{tools.analyzer} environment
+$\trdiff{\taenv{}}$
+and proposition environment $\propenv{}$, with 
+then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}},
+object \object{} and new multimethod environment
+$\trdiff{\mmenvp{}}$.
 
@@
 
-\begin{figure*}
-$$
-\begin{altgrammar}
-  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
-                      \alt \true{} \alt \false{} \alt \nil{}
-                      \alt \class{}
-                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
-                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
-                      \alt
-                      {\doexp {\e{}} {\e{}}}
-                      \\ &\alt& 
-                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
-                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
-                &\mbox{Expressions} \\ \\
-  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
-                      \alt {\Value \v{}}
-                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
-                      \alt (\cap\ \overrightarrow{\t{}})
-                      \alt (\Not\ \t{})
-                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
-                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
-                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
-                                   {\t{}}
-                                   {\filterset {\prop{}} {\prop{}}}
-                                   {\object{}}}
-                      \\ &\alt&
-                      {\ArrayTwo {\t{}} {\t{}}}
-                      \alt {\HMapc {\mandatory}}
-                      \alt {\HMapp {\mandatory} {\absent}}
-
-                &\mbox{Types} \\ \\
-  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-                      \alt {\impprop {\prop{}} {\prop{}}}
-                      \alt {\andprop {\prop{}} {\prop{}}}
-                      \alt {\orprop {\prop{}} {\prop{}}}
-                      \alt \topprop{}
-                      \alt \botprop{}
-                &\mbox{Propositions} \\
-  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
-                      \alt \emptyobject{}
-                &\mbox{Objects} \\
-  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
-                &\mbox{Paths} \\
-  \pesyntax{}   &::=& \classpe{} \alt \keype{k}
-                &\mbox{Path Elements} \\ \\
-                % TODO objects with path
-%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
-%              &\mbox{Constants}
-  \Gamma       &::=& \overrightarrow{\prop{}}
-               &\mbox{Environments}
-\end{altgrammar}
-$$
-\caption{Syntax of Terms, Types, Propositions, and Objects}
-\end{figure*}
-
-\begin{figure*}
-\begin{mathpar}
-\infer [T-Local]
-{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
-{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
-             {\x{}}}
-
-\infer [T-True]
-{}
-{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }
-
-\infer [T-False]
-{}
-{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
-
-\infer [T-Nil]
-{}
-{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
-
-\infer [T-Do]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
-\\\\
-  \judgement {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
-                 {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} 
-                 {\object{}}
-  }
-{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
-             }
-
-\infer [T-If]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-                 {\object{1}} 
-             \\\\
-  \judgement {\propenv{}, {\thenprop {\prop{1}}}}
-                 {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
-                 {\object{}}
-  \\\\
-  \judgement {\propenv{}, {\elseprop {\prop{1}}}}
-                 {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
-                 {\object{}}
-             }
-{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
-                 {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
-                             {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
-                 {\object{}} }
-
-\infer [T-Let]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}}
-  \\\\
-  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
-              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
-              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
-             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} 
-             }
-{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
-             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                          {\object{1}}
-                          {\x{}}}
-             {\replacefor {\object{}} 
-                          {\object{1}}
-                          {\x{}}}
-             }
-
-\infer [T-Throw]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-}
-{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop{}} {\botprop{}}}
-             {\emptyobject{}} }
-
-\infer [T-App]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
-                                                       {\t{}}
-                                                       {\filterset {\thenprop {\prop{f}}}
-                                                                   {\elseprop {\prop{f}}}}
-                                                       {\object{f}}}}
-                {\filterset {\thenprop {\prop{}}}
-                            {\elseprop {\prop{}}}}
-                {\object{}}
-  \\\\
-  \judgement {\propenv{}}
-                 {\hastype {\ep{}} {\s{}}}
-                 {\filterset {\thenprop {\propp{}}}
-                             {\elseprop {\propp{}}}}
-                 {\objectp{}} 
-}
-{ \judgement {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
-                                        {\replacefor {\t{}}
-                                                     {\objectp{}}
-                                                     {\x{}}}}
-                 {\replacefor {\filterset {\thenprop {\prop{f}}}
-                                          {\elseprop {\prop{f}}}}
-                              {\objectp{}}
-                              {\x{}}}
-                 {\replacefor {\object{f}}
-                              {\objectp{}}
-                              {\x{}}}
-}
-
-\infer [T-Abs]
-{}
-{ TODO }
-
-\infer [T-Try]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\t{}}}
-                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-\\\\
-  \judgement {\propenv{}, {\isprop {\x{}} {\class{1}}}}
-             {\hastype {\e{2}} {\t{}}}
-                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-\\
-  \judgement {\propenv{}}
-             {\hastype {\e{3}} {\t{3}}}
-                 {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
-                 {\object{3}}
-}
-{ \judgement {\propenv{}} {\hastype {\tryexp {\e{1}} {\class{1}} {\x{}} {\e{2}} {\e{3}}} {\t{}}}
-                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-                 }
-
-\infer [T-New]
-{}
-{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\class{}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}
-             }
-
-\infer [T-Field]
-{ TODO }
-{ }
-
-\infer [T-Method]
-{ TODO }
-{ }
-
-\infer [T-KwInvoke]
-{}
-{ TODO }
-
-\infer [T-HMapInvoke]
-{}
-{ TODO }
-
-\infer [T-KeywordMap]
-{ \judgement {\propenv{}} {\hastype {\overrightarrow{{\e{v}}}} {\overrightarrow {{\t{v}}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}}
-{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{k_{v}}\ {\e{v}}}\}} {\HMapc {\overrightarrow{{k_{v}}\ {\t{v}}}}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}}
-
-\infer [T-Map]
-{ 
-\judgement {\propenv{}} {\hastype {\overrightarrow{{\e{k}}}} {\overrightarrow {{\t{k}}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}
-\\
-\judgement {\propenv{}} {\hastype {\overrightarrow{{\e{v}}}} {\overrightarrow {{\t{v}}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}}
-{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{\e{k}}\ {\e{v}}}\}}
-                                    {\IPersistentMap {\Unionsplice {\overrightarrow{{t_{k}}}}}
-                                                     {\Unionsplice {\overrightarrow{{t_{v}}}}}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}}
-
-\infer [T-HMapFull]
-{}
-{ }
-
-\infer [T-HMapAbsent]
-{}
-{ TODO }
-
-\infer [T-HMapBoth]
-{}
-{ TODO }
-
-\infer [T-Defmulti]
-{ \inpropenv {\propenv{}} {\isprop {\x{}} 
-                                   {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{m}}
-                                    {\filterset {\thenprop {\prop{m}}}
-                                                {\elseprop {\prop{m}}}}
-                                    {\object{m}}}}
-\\\\
-  \judgement {\propenv{}} {\hastype {\e{d}} {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{r}}
-                                                    {\filterset {\thenprop {\prop{d}}}
-                                                                {\elseprop {\prop{d}}}}
-                                                    {\object{d}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}}
-\\\\
-{\adddispatchfn
-{\x{}}
-{\e{d}}}
-}
-{ \judgement {\propenv{}} {\hastype {\defmultiexp {\x{}} {\e{d}}} {\MultiFn{}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}
-}
-
-\infer [T-Defmethod]
-{ \inpropenv {\propenv{}} {\isprop {\x{m}}
-                                   {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{m}}
-                                    {\filterset {\thenprop {\prop{m}}}
-                                                {\elseprop {\prop{m}}}}
-                                    {\object{m}}}}
-\\
-\getdispatchfn {\x{m}} {\e{d}}
-\\\\
-\judgement {\propenv{}, {\overrightarrow {\isprop {\x{p}} {\t{p}}}}}
-           {\hastype {\appexp {\e{d}} {\overrightarrow {\x{p}}}}
-                                  {\t{d}}}
-           {\filterset {\thenprop {\prop{d}}}
-                       {\elseprop {\prop{d}}}}
-           {\object{d}}
-\\\\
-\judgement {\propenv{}, {\thenprop {\prop{d}}}}
-           {\hastype {\e{b}} {\t{m}}}
-           {\filterset {\thenprop {\prop{m}}}
-                       {\elseprop {\prop{m}}}}
-           {\object{m}}
-}
-{ \judgement {\propenv{}} {\hastype {\defmethodexp {\x{m}} {\e{v}} {\overrightarrow{\x{p}}}
-                                                   {\e{b}}}
-                                    {\MultiFn{}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}
-}
-
-\infer [T-InstanceOf]
-{ \judgement {\propenv{}} 
-             {\hastype {\e{}} {\t{}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}
-}
-{ \judgement {\propenv{}} {\hastype {\instanceapp {\class{}} {\e{}}}
-                                    {\Boolean}}
-             {\filterset {\isprop  {\class{}} {\object{}}}
-                         {\notprop {\class{}} {\object{}}}}
-             {\emptyobject}
-}
-
-\infer [T-Isa]
-{ \isacompare {\e{1}} {\e{2}}
-             {\filterset {\thenprop {\prop{}}}
-                         {\elseprop {\prop{}}}}
-}
-{ \judgement {\propenv{}} {\hastype {\isaapp {\e{1}} {\e{2}}}
-                                    {\Boolean{}}}
-             {\filterset {\thenprop {\prop{}}}
-                         {\elseprop {\prop{}}}}
-             {\emptyobject}}
-
-\infer [T-Subsume]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\t{}}}
-             {\filterset {\thenprop {\prop{}}}
-                         {\elseprop {\prop{}}}}
-             {\object}
-\\\\
-\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
-\\
-\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-\\\\
-\inpropenv {} {\issubtype {\t{}} {\tp{}}}
-\\
-\inpropenv {} {\issubtype {\object{}} {\objectp{}}}
-}
-{ \judgement {\propenv{}} {\hastype {\e{}} {\tp{}}}
-             {\filterset {\thenprop {\propp{}}}
-                         {\elseprop {\propp{}}}}
-             {\objectp}}
-
-\end{mathpar}
-\caption{Typing Rules}
-\end{figure*}
-
-%$$
-%\begin{tdisplay}{Evaluation Contexts}
-%  \begin{altgrammar}
-%    \E{} &::=& [ ] % application rules
-%              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
-%              % map rules
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
-%              &\mbox{Evaluation Contexts}
-%  \end{altgrammar}
-%\end{tdisplay}
-%$$
-
-\begin{figure*}
-$$
-\begin{altgrammar}
-{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
-\\
-{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
-    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
-            {\javatotc {\class{2}} {\true{}}}
-            {\filterset {\topprop{}} {\topprop{}}}
-            {\emptyobject{}}
-            }
-\\
-{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
-    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
-            {\javatotc {\class{2}} {\false{}}}
-            {\filterset {\topprop{}} {\topprop{}}}
-            {\emptyobject{}}
-            }
-\\\\
-
-{\javatotc {\pint{}} {b}} &=& \Integer{}
-\\
-{\javatotc {\Void{}} {b}} &=& \Nil
-\\
-{\javatotc {\pvoid{}} {b}} &=& \Nil
-\\
-{\javatotc {\pchar{}} {b}} &=& \Character{}
-\\
-{\javatotc {\pdouble{}} {b}} &=& \Double{}
-\\
-{\javatotc {\JavaArray {\pint{}}} {b}} &=& {\Union {\Nil} {\ArrayP {\pint{}}}}
-\\
-{\javatotc {\JavaArray {\pdouble{}}} {b}} &=& {\Union {\Nil} {\ArrayP {\pdouble{}}}}
-\\
-{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}}
-\\
-{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}
-\\
-{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
-\\
-{\javatotc {\class{}} {\false{}}} &=& \class{}
-
-\end{altgrammar}
-$$
-\caption{Converting Java types to Typed Clojure types}
-\end{figure*}
-
 \subsection{Java interoperability}
@@
 \end{verbatim}
-
-%\appendix
-%\section{Appendix Title}
-%
-%This is the text of the appendix, if you need one.
-
-%\acks
-%
-%Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{bibliography}
-
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-10-10 icfp15-typedclojure.tex
add tools.analyzer rules
add ta-field
link up to ta calculus

--- 2014-08-31 schemeworkshop13.tex
@@
 
-\section{Contributions}
+\input{contributions}
 
-We present a practical type system for Clojure
+\input{overview}
 
-The main result of this paper is to demonstrate that Typed Racket's design
-and implementation is applicable to a practical type system for Clojure
-by implementing Typed Clojure.
+\input{formal-model}
 
-% list the delta from Typed Racket
+\input{syntax-figure}
 
-\section{Overview of Typed Clojure}
+\input{typing-rules}
 
-Typed Clojure is a gradual type system for Clojure. It is designed
-to type check normal Clojure code by adding annotations. It is implemented
-as a library, and can be seamlessly included in any Clojure project; no
-separate compiler or language is needed.
+\input{tools-analyzer-figure}
 
-{\smallsection {Based on Typed Racket}}
-Initially, the similarities between untyped Racket and Clojure and Typed Racket's 
-ability to type check Racket code led us to investigate a similar type system for Clojure.
-After two years of development, the solid basis of Typed Racket 
-helps us type check many Clojure idioms without significant differences
-in implementation or theory. We found that extending Typed Clojure to check
-those idioms that have no obvious Racket equivalent did not significantly alter the structure
-of the type system.
-
-{\smallsection {Occurrence typing}}
-\citet{TF08,TF10} developed \emph{occurrence typing}, which helps improve types at branches.
-Typed Clojure uses occurrence typing in a similar way to Typed Racket, with
-some extensions (discussed in in Section [?]). %FIXME
-
-{\smallsection {Practical Variable-Arity Polymorphism}}
-Functions with non-trivial variable parameters are common in Racket.
-For example, Racket provides \emph{map} which takes a function and a
-variable number of collections and applies the function simultaneously
-to each element of the provided collections, returning a list of results.
-\citet*{STF09} developed a practical system that handles advanced variable parameters
-which can handle applications of functions like \emph{map}.
-
-Clojure has a similar emphasis on variable-arity functions. In some ways,
-Clojure's core library encourages even more complicated variable-parameter schemes.
-The \emph{assoc} core function, for example, takes three parameters and
-then a quantity of variable parameters that is a multiple of two.
-This is beyond what Typed Racket (and Typed Clojure) can currently handle.
-
-Functions like \emph{map} are common in Clojure, so we provide an implementation
-of variable-arity polymorphism which has similar capabilities as Typed Racket's
-implementation.
-
-{\smallsection {Local Type Inference}}
-We use Pierce and Turner's Local Type Inference~\cite{PT00} to infer some polymorphic
-applications. Our implementation is based on Typed Racket's, which has extensions
-to support applications of polymorphic variable-arity functions like \emph{map}.
-
-{\smallsection {Unions and intersections}}
-Like Typed Racket, we include union and ordered intersection types. Unions define
-a least-upper-bound for a set of types. For example, we can express a type that is
-either \Number or \Symbol by including them in a union: {\Union {\Number} {\Symbol}}
-
-Ordered intersections (described further by \citet{SA+12})
-are used for overloading function types. We can express a function that takes
-a \Number and returns a \Symbol, and vice-versa with an ordered intersection function type:
-
-\begin{lstlisting}[label=lst:ordered]
-(Fn [Number -> Symbol]
-    [Symbol -> Number])
-\end{lstlisting}
-
-As our intersections are \emph{ordered}, we can express fine invariants in the
-case where arity parameter types overlap. Similar to a pattern match, earlier arities 
-are tried first, and the first arity to match ``wins''.
-
-For example, applying an \lstinline|Integer| argument to a function of type
-
-\begin{lstlisting}
-(Fn [Integer -> Integer]
-    [Number -> Number])
-\end{lstlisting}
-
-returns an \lstinline|Integer|. Reversing the arities however gives
-type \lstinline|Number|, because the arity taking a \lstinline|Number|
-always matches first.
-
-{\smallsection {Hosted on the Java Virtual Machine}}
-Clojure is built to run on the Java Virtual Machine (JVM),
-offering good interoperability with existing Java code.
-Typed Clojure helps programmers correctly call Java code
-by integrating with Java's type system.
-
-We give Java arrays and Java's \emph{null} special treatment
-when involved with interoperability. Arrays are treated as \emph{read-only}
-when sourced from Java methods, discussed in Section \ref{sec:arrays}.
-We are explicit, and conservative by default, in the positions where
-Java's \emph{null} can be passed, discussed in Section \ref{sec:null}.
-
-% TODO references
-% why is TR such a good base?
-% - immutability
-% - common lisp ancestry
-% differences?
-% - Clojure is built on JVM
-% - interop with JVM
-% - Clojure's idiomatic primitives are different
-% - multimethods + protocols
-% - less sophisticated macro system
-%  - not an issue
-%  - implementation difference, AST walking vs syntax walking
-
-\section{A Formal Model of Typed Clojure}
-
-We describe our type system in a similar way to Typed Racket (\citet{TF10}).
-The main judgement 
-{\judgement {\Gamma} {\hastype {\e{}} {\t{}}} 
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-            {\object{}}}
-says expression \e{} is of type \t{} in environment $\Gamma$, with 
-then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
-and object \object{}.
-
-\subsection{Higher-rank polymorphism}
-
-Typed Clojure supports f-bounded polymorphism and higher-rank
-types.
-% why include them?
-% examples, involving generic type operators for monads & conduits
-% useful for generic programming, similar to Hashell type classes
-
-\subsection{Using negative filters}
-
-Occurrence typing plays an important role in Typed Racket and Typed Clojure.
-By maintaining a \emph{proposition environment} of propositions relating types to
-bindings, we can update bindings with more accurate types as programs progress.
-It follows that there is some correspondence between propositions and types,
-characterised by the \emph{update} function, which takes a type and a proposition
-and returns a type which updates the input type using the proposition.
-
-There is a straightforward relationship between ``positive'' propositions and types.
-For example 
-{\tt (update Number (is Integer 0))}
-updates Number by Integer, which is Integer, because Integer <: Number.
-
-The relationship between ``negative'' propositions and types is not always obvious.
-A common proposition in Typed Clojure is (! (U nil false) a): the proposition that
-local binding ``a'' is \emph{not} of type (U nil false).
-This problem is most visible in expressions like {\tt (filter identity coll)}, where
-``identity'' has a ``then'' proposition that has negative information: (! (U nil false) 0),
-which reads, the 0th argument of identity does not contain (U nil false).
-
-\subsection{Multimethods}
-
-Clojure provides multimethods which can dispatch on the result of an
-arbitrary function, the \emph{dispatch function}. \emph{Methods} are
-added to the multimethod
-which are associated with a \emph{dispatch value}, which is compared to the
-result of apply the dispatch function to the argument passed to the
-multimethod. If this comparison unambiguously dispatches to one method,
-that method is then called with the current arguments.
-
-Occurrence typing, in particular when extended with paths,
-for checking method definitions for arbitrary dispatch functions.
-
-\begin{lstlisting}
-(ns example.mm
-  (:require 
-    [clojure.core.typed :refer [ann]]))
-
-(ann my-mm [Any -> Number])
-(defmulti my-mm class)
-
-(defmethod my-mm Float [f] (inc f))
-\end{lstlisting}
-
-\begin{lstlisting}
-(ann clojure.core/class
-  [Any -> (U nil Class)
-   :object {:id 0, :path [Class]}])
-\end{lstlisting}
-
-\begin{verbatim}
-TODO
-- example of normal MM usage
-- explain descrepency btwn TC syntax and formal
-  TR calculs syntax. eg. :object/:filter 
-    vs over arrow
-\end{verbatim}
-
-\subsection{Extending occurrence typing}
-
-\begin{figure*}
-$$
-\begin{altgrammar}
-  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
-                      \alt \true{} \alt \false{} \alt \nil{}
-                      \alt \class{}
-                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
-                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
-                      \alt
-                      {\doexp {\e{}} {\e{}}}
-                      \\ &\alt& 
-                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
-                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
-                &\mbox{Expressions} \\ \\
-  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
-                      \alt {\Value \v{}}
-                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
-                      \alt (\cap\ \overrightarrow{\t{}})
-                      \alt (\Not\ \t{})
-                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
-                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
-                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
-                                   {\t{}}
-                                   {\filterset {\prop{}} {\prop{}}}
-                                   {\object{}}}
-                      \\ &\alt&
-                      {\ArrayTwo {\t{}} {\t{}}}
-                      \alt {\HMapc {\mandatory}}
-                      \alt {\HMapp {\mandatory} {\absent}}
-
-                &\mbox{Types} \\ \\
-  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-                      \alt {\impprop {\prop{}} {\prop{}}}
-                      \alt {\andprop {\prop{}} {\prop{}}}
-                      \alt {\orprop {\prop{}} {\prop{}}}
-                      \alt \topprop{}
-                      \alt \botprop{}
-                &\mbox{Propositions} \\
-  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
-                      \alt \emptyobject{}
-                &\mbox{Objects} \\
-  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
-                &\mbox{Paths} \\
-  \pesyntax{}   &::=& \classpe{} \alt \keype{k}
-                &\mbox{Path Elements} \\ \\
-                % TODO objects with path
-%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
-%              &\mbox{Constants}
-  \Gamma       &::=& \overrightarrow{\prop{}}
-               &\mbox{Environments}
-\end{altgrammar}
-$$
-\caption{Syntax of Terms, Types, Propositions, and Objects}
-\end{figure*}
-
-\begin{figure*}
-\begin{mathpar}
-\infer [T-Local]
-{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
-{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
-             {\x{}}}
-
-\infer [T-True]
-{}
-{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }
-
-\infer [T-False]
-{}
-{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
-
-\infer [T-Nil]
-{}
-{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }
-
-\infer [T-Do]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
-\\\\
-  \judgement {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
-                 {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} 
-                 {\object{}}
-  }
-{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
-             }
-
-\infer [T-If]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-                 {\object{1}} 
-             \\\\
-  \judgement {\propenv{}, {\thenprop {\prop{1}}}}
-                 {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
-                 {\object{}}
-  \\\\
-  \judgement {\propenv{}, {\elseprop {\prop{1}}}}
-                 {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
-                 {\object{}}
-             }
-{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
-                 {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
-                             {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
-                 {\object{}} }
-
-\infer [T-Let]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}}
-  \\\\
-  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
-              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
-              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
-             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} 
-             }
-{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
-             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                          {\object{1}}
-                          {\x{}}}
-             {\replacefor {\object{}} 
-                          {\object{1}}
-                          {\x{}}}
-             }
-
-\infer [T-Throw]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-}
-{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop{}} {\botprop{}}}
-             {\emptyobject{}} }
-
-\infer [T-App]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
-                                                       {\t{}}
-                                                       {\filterset {\thenprop {\prop{f}}}
-                                                                   {\elseprop {\prop{f}}}}
-                                                       {\object{f}}}}
-                {\filterset {\thenprop {\prop{}}}
-                            {\elseprop {\prop{}}}}
-                {\object{}}
-  \\\\
-  \judgement {\propenv{}}
-                 {\hastype {\ep{}} {\s{}}}
-                 {\filterset {\thenprop {\propp{}}}
-                             {\elseprop {\propp{}}}}
-                 {\objectp{}} 
-}
-{ \judgement {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
-                                        {\replacefor {\t{}}
-                                                     {\objectp{}}
-                                                     {\x{}}}}
-                 {\replacefor {\filterset {\thenprop {\prop{f}}}
-                                          {\elseprop {\prop{f}}}}
-                              {\objectp{}}
-                              {\x{}}}
-                 {\replacefor {\object{f}}
-                              {\objectp{}}
-                              {\x{}}}
-}
-
-\infer [T-Abs]
-{}
-{ TODO }
-
-\infer [T-Try]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\t{}}}
-                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-\\\\
-  \judgement {\propenv{}, {\isprop {\x{}} {\class{1}}}}
-             {\hastype {\e{2}} {\t{}}}
-                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-\\
-  \judgement {\propenv{}}
-             {\hastype {\e{3}} {\t{3}}}
-                 {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
-                 {\object{3}}
-}
-{ \judgement {\propenv{}} {\hastype {\tryexp {\e{1}} {\class{1}} {\x{}} {\e{2}} {\e{3}}} {\t{}}}
-                 {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                 {\object{}}
-                 }
-
-\infer [T-New]
-{}
-{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\class{}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}
-             }
-
-\infer [T-Field]
-{ TODO }
-{ }
-
-\infer [T-Method]
-{ TODO }
-{ }
-
-\infer [T-KwInvoke]
-{}
-{ TODO }
-
-\infer [T-HMapInvoke]
-{}
-{ TODO }
-
-\infer [T-KeywordMap]
-{ \judgement {\propenv{}} {\hastype {\overrightarrow{{\e{v}}}} {\overrightarrow {{\t{v}}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}}
-{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{k_{v}}\ {\e{v}}}\}} {\HMapc {\overrightarrow{{k_{v}}\ {\t{v}}}}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}}
-
-\infer [T-Map]
-{ 
-\judgement {\propenv{}} {\hastype {\overrightarrow{{\e{k}}}} {\overrightarrow {{\t{k}}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}
-\\
-\judgement {\propenv{}} {\hastype {\overrightarrow{{\e{v}}}} {\overrightarrow {{\t{v}}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}}
-{ \judgement {\propenv{}} {\hastype {\{\overrightarrow{{\e{k}}\ {\e{v}}}\}}
-                                    {\IPersistentMap {\Unionsplice {\overrightarrow{{t_{k}}}}}
-                                                     {\Unionsplice {\overrightarrow{{t_{v}}}}}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}}
-
-\infer [T-HMapFull]
-{}
-{ }
-
-\infer [T-HMapAbsent]
-{}
-{ TODO }
-
-\infer [T-HMapBoth]
-{}
-{ TODO }
-
-\infer [T-Defmulti]
-{ \inpropenv {\propenv{}} {\isprop {\x{}} 
-                                   {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{m}}
-                                    {\filterset {\thenprop {\prop{m}}}
-                                                {\elseprop {\prop{m}}}}
-                                    {\object{m}}}}
-\\\\
-  \judgement {\propenv{}} {\hastype {\e{d}} {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{r}}
-                                                    {\filterset {\thenprop {\prop{d}}}
-                                                                {\elseprop {\prop{d}}}}
-                                                    {\object{d}}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}}
-\\\\
-{\adddispatchfn
-{\x{}}
-{\e{d}}}
-}
-{ \judgement {\propenv{}} {\hastype {\defmultiexp {\x{}} {\e{d}}} {\MultiFn{}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}
-}
-
-\infer [T-Defmethod]
-{ \inpropenv {\propenv{}} {\isprop {\x{m}}
-                                   {\Arrow {\overrightarrow {\hastype {\x{p}} {\t{p}}}} {\t{m}}
-                                    {\filterset {\thenprop {\prop{m}}}
-                                                {\elseprop {\prop{m}}}}
-                                    {\object{m}}}}
-\\
-\getdispatchfn {\x{m}} {\e{d}}
-\\\\
-\judgement {\propenv{}, {\overrightarrow {\isprop {\x{p}} {\t{p}}}}}
-           {\hastype {\appexp {\e{d}} {\overrightarrow {\x{p}}}}
-                                  {\t{d}}}
-           {\filterset {\thenprop {\prop{d}}}
-                       {\elseprop {\prop{d}}}}
-           {\object{d}}
-\\\\
-\judgement {\propenv{}, {\thenprop {\prop{d}}}}
-           {\hastype {\e{b}} {\t{m}}}
-           {\filterset {\thenprop {\prop{m}}}
-                       {\elseprop {\prop{m}}}}
-           {\object{m}}
-}
-{ \judgement {\propenv{}} {\hastype {\defmethodexp {\x{m}} {\e{v}} {\overrightarrow{\x{p}}}
-                                                   {\e{b}}}
-                                    {\MultiFn{}}}
-             {\filterset {\topprop{}} {\botprop{}}}
-             {\emptyobject}
-}
-
-\infer [T-InstanceOf]
-{ \judgement {\propenv{}} 
-             {\hastype {\e{}} {\t{}}}
-             {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object}
-}
-{ \judgement {\propenv{}} {\hastype {\instanceapp {\class{}} {\e{}}}
-                                    {\Boolean}}
-             {\filterset {\isprop  {\class{}} {\object{}}}
-                         {\notprop {\class{}} {\object{}}}}
-             {\emptyobject}
-}
-
-\infer [T-Isa]
-{ \isacompare {\e{1}} {\e{2}}
-             {\filterset {\thenprop {\prop{}}}
-                         {\elseprop {\prop{}}}}
-}
-{ \judgement {\propenv{}} {\hastype {\isaapp {\e{1}} {\e{2}}}
-                                    {\Boolean{}}}
-             {\filterset {\thenprop {\prop{}}}
-                         {\elseprop {\prop{}}}}
-             {\emptyobject}}
-
-\infer [T-Subsume]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\t{}}}
-             {\filterset {\thenprop {\prop{}}}
-                         {\elseprop {\prop{}}}}
-             {\object}
-\\\\
-\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
-\\
-\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-\\\\
-\inpropenv {} {\issubtype {\t{}} {\tp{}}}
-\\
-\inpropenv {} {\issubtype {\object{}} {\objectp{}}}
-}
-{ \judgement {\propenv{}} {\hastype {\e{}} {\tp{}}}
-             {\filterset {\thenprop {\propp{}}}
-                         {\elseprop {\propp{}}}}
-             {\objectp}}
-
-\end{mathpar}
-\caption{Typing Rules}
-\end{figure*}
+\input{subtyping}
 
@@
 
-\begin{figure*}
-$$
-\begin{altgrammar}
-{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
-\\
-{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
-    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
-            {\javatotc {\class{2}} {\true{}}}
-            {\filterset {\topprop{}} {\topprop{}}}
-            {\emptyobject{}}
-            }
-\\
-{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
-    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
-            {\javatotc {\class{2}} {\false{}}}
-            {\filterset {\topprop{}} {\topprop{}}}
-            {\emptyobject{}}
-            }
-\\\\
-
-{\javatotc {\pint{}} {b}} &=& \Integer{}
-\\
-{\javatotc {\Void{}} {b}} &=& \Nil
-\\
-{\javatotc {\pvoid{}} {b}} &=& \Nil
-\\
-{\javatotc {\pchar{}} {b}} &=& \Character{}
-\\
-{\javatotc {\pdouble{}} {b}} &=& \Double{}
-\\
-{\javatotc {\JavaArray {\pint{}}} {b}} &=& {\Union {\Nil} {\ArrayP {\pint{}}}}
-\\
-{\javatotc {\JavaArray {\pdouble{}}} {b}} &=& {\Union {\Nil} {\ArrayP {\pdouble{}}}}
-\\
-{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}}
-\\
-{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}
-\\
-{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
-\\
-{\javatotc {\class{}} {\false{}}} &=& \class{}
-
-\end{altgrammar}
-$$
-\caption{Converting Java types to Typed Clojure types}
-\end{figure*}
-
-\subsection{Java interoperability}
-
-As Clojure supports Java interoperability, Typed Clojure has some
-integration with Java's type system. Calls to Java methods from
-Clojure can be type checked by Typed Clojure.
-
-We currently do not support Java generics
-% because generics don't "exist" on the JVM
-% clojure also doesn't use them
-
-\subsubsection{Arrays}
-\label{sec:arrays}
-
-Supporting statically sound interactions with Java arrays is a goal
-of Typed Clojure. This is complicated by Java's decision to make
-arrays covariant in their argument, a well documented source of static
-unsoundness. Bracha~\cite{Bra98} summarises Java's approach to maintaining
-soundness at runtime, which involves all array writes being checked by
-runtime assertions.
-
-This approach fits Java's type system, but we can do better in a more powerful
-type system like Typed Clojure. Our goal is to catch all type-incorrect array
-writes at compile time so the type system can do more to help Clojure programmers
-use arrays, especially those being passed from foreign Java code.
-
-Our basic approach is to make our array types \emph{bivariant}. Array types
-look like {\ArrayTwo {\t{w}} {\t{r}}} and
-are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
-and a covariant parameter for output (reading).
-This type can write type {\t{w}} and read type {\t{r}}.
-
-Most commonly, an array type is invariant in its parameter; it can
-write and read input of the same type.
-We can get the same effect by setting our input and output
-parameters to the same type. For example, {\ArrayTwo {\Number} {\Number}}
-(or equivalently, {\Array {\Number}})
-in Typed Clojure is similar to invariant array types of \Number in languages like Scala.
-
-The biggest gain in using a separate input parameter is the ability
-to specify \emph{read-only} arrays. Crucially, our type system features an
-explicit bottom type \lstinline|Nothing|, enabling a read-only \lstinline|Number| array
-to be of type \lstinline|(Array2 Nothing Number)|.
-
-To realise why defining read-only arrays are useful, we need to examine
-what makes array covariance unsound in Java.
-\begin{verbatim}
-FIXME
-Array covariance about the type of an array so the consumer
-of an array cannot tell the actual type of the array when examining a type
-signature.
-\end{verbatim}
-
-\begin{lstlisting}
-...
-public static Number[] getNumberArray() {
-  Number[] n = new Integer[10];
-  return n;
-}
-...
-\end{lstlisting}
-
-To the casual consumer \emph{getNumberArray} returns an array that can both
-read and write \lstinline|Number|s. However it is clear from the implementation
-that attempting to write say a \lstinline|Double| to this array will result
-in a runtime error.
-
-\begin{verbatim}
-...
-Number[] myArray = getNumberArray();
-myArray[0] = 1.1;
-/* Exception in thread "main" 
-   java.lang.ArrayStoreException: 
-   java.lang.Double */
-...
-\end{verbatim}
-
-Notice that this is a runtime error, and Java's type system has not helped
-statically prevent it.
-This could cause a similar issue for other statically-typed languages offering
-interoperability with Java.
-
-To prevent these sorts of runtime exceptions in Typed Clojure, we declare
-all arrays from unknown sources to be \emph{read-only}. Put differently,
-the only way to define a writeable array is to create it in type-checked Clojure
-code.
-
-\begin{lstlisting}
-(let [n (CovariantArray/getNumberArray)]
-  (aset n 0 1.1))
-
-; Polymorphic static method clojure.lang.RT/aset could not be 
-; applied to arguments:
-; Domains: 
-;         (Array2 i o) clojure.core.typed/AnyInteger i
-; 
-; Arguments:
-;         (Array2 Nothing java.lang.Number) int (Value 1.1)
-; 
-; with expected type:
-;         Any
-\end{lstlisting}
-
-The type inferred for the local \lstinline|n| is \lstinline|(Array2 Nothing Number)|
-which tells the type system: it is never safe to write to this array, but
-it is safe to assume \lstinline|Number|s can be read from this array.
-
-To emphasise, Typed Clojure throws a static type error. Errors like this help Clojure programmers
-use foreign Java libraries more correctly.
-
-\begin{verbatim}
-Note that Java libraries are often large 
-and complex and programmers will probably
-enjoy the extra help from the type system.
-\end{verbatim}
-
-\subsection{Handling null}
-\label{sec:null}
-
-Probably the most common pain point in Java programming is dealing with \emph{null}.
-It is crucial that Typed Clojure deals with \emph{null} intelligently so that
-all potentially erroneous interactions with \emph{null} in typed code are caught at compile time.
-We also want to represent \emph{null} in a way that is natural to Clojure programmers
-and approximates their mental model when reasoning about Clojure code.
-
-Firstly, we separate the concepts of \emph{null} and reference types at the type level.
-This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
-We provide a singleton type \Nil{}, which contains just the value \lstinline|nil|,
-Clojure's equivalent of \emph{null}.
-
-Armed with general unions and the type \lstinline|nil|, 
-we make Typed Clojure's reference types \emph{non-nullable}.
-A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
-in Typed Clojure.
-
-Typed Racket already provides what is needed to write such types; indeed this is a repurposing
-of a common Typed Racket idiom of approximating an Option or Maybe type by
-making a union that includes \emph{\#f}, Racket's only false value.
-However, like handling covariant arrays, the subtlety lies in our interactions with foreign
-Java code.
-
-A set of conversion rules define how to convert a Java type signature into a null-safe
-Clojure version. These rules are designed to be conservative and can be explicitly overridden
-by programmers. Method parameters are non-nullable, while return types are nullable.
-Field types are nullable. Constructor parameters are non-nullable, and return types
-are also non-nullable: it is a Java invariant that all constructors return an instance
-of the class they are constructing.
-
-\begin{verbatim}
-TODO 
-- write conversion rules for 
-  Java type -> Clojure type
-\end{verbatim}
+\input{convert-types-figure}
 

### 2014-10-10 overview.tex
link up to ta calculus

--- 2013-07-30 schemeworkshop13.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint,10pt]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{mmm}
-\usepackage{mathpartir}
-\usepackage{clj-grammar}
-%\usepackage{url}
-%\usepackage{biblatex}
-\include{bibliography.bib}
-
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
-
-\usepackage{amsmath}
-
-\newcommand\smallsection[1]{\bf \emph{#1}}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-%\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Typed Clojure}
-\subtitle{A Practical Type System for Clojure, based on Typed Racket}
-
-\authorinfo{Ambrose Bonnaire-Sergeant}
-           {University of Western Australia}
-           {abonnairesergeant@gmail.com}
-\authorinfo{Rowan Davies}
-           {University of Western Australia}
-           {rowan@csse.uwa.edu.au}
-
-\maketitle
-
-\begin{abstract}
-Typed Racket is a valuable starting point for
-a gradual type system that targets Clojure.
-Building a similar type system for a new language gives the
-designer some flexibility to repurpose and extend features.
-This paper gives an overview of Typed Clojure, concentrating
-on the extensions and differences from Typed Racket. We also
-show where Typed Racket's features were particularly useful
-for type checking non-trivial Clojure idioms.
-\end{abstract}
-
-\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-\keywords
-gradual types, Clojure
-
-\section{Typed Clojure}
-
-Typed Clojure is a gradual type system for Clojure, based on the
-static portion of Typed Racket.
-
-% what are the goals of Typed Clojure?
-% where do they differ from Typed Racket? why?
-
-% Things todo
-% - DONE small calculus based on previous TR formalisation
-%   that demos the ``flow'' environment & not/intersection type
-% - discussion and examples for flow filters
-% - discussion on the not/difference type, real code where it helps
-% - higher-rank polymorphism, show what it enables (eg. monads)
-%   - also show more complicated example, making an extensible
-%     type for clojure.core/conj, and why it's not straightforward
-% - discussion on limitations of local type inference
-% - demonstration of how we check multimethods
-% - dealing with array covariance
-% - note on how we avoid null-pointer exceptions
-% - discussion on why Typed Clojure only uses the ``static'' part
-%   of Typed Racket
-% - the equality filter problem (and draft solution?), a proposition
-%   that knows about binding aliasing. Sounds straightforward?
-
-\section{Contributions}
-
-We present a practical type system for Clojure
-
-The main result of this paper is to demonstrate that Typed Racket's design
-and implementation is applicable to a practical type system for Clojure
-by implementing Typed Clojure.
-
 \section{Overview of Typed Clojure}
 
-Typed Clojure is a gradual type system for Clojure, designed to
+Typed Clojure is a gradual type system for Clojure. It is designed
+to type check normal Clojure code by adding annotations. It is implemented
+as a library, and can be seamlessly included in any Clojure project; no
+separate compiler or language is needed.
 
@@
 %  - implementation difference, AST walking vs syntax walking
-
-\subsection{Higher-rank polymorphism}
-
-Typed Clojure supports f-bounded polymorphism and higher-rank
-types.
-
-\subsection{Using negative filters}
-
-Occurrence typing plays an important role in Typed Racket and Typed Clojure.
-By maintaining a \emph{proposition environment} of propositions relating types to
-bindings, we can update bindings with more accurate types as programs progress.
-It follows that there is some correspondence between propositions and types,
-characterised by the \emph{update} function, which takes a type and a proposition
-and returns a type which updates the input type using the proposition.
-
-There is a straightforward relationship between ``positive'' propositions and types.
-For example 
-{\tt (update Number (is Integer 0))}
-updates Number by Integer, which is Integer, because Integer <: Number.
-
-The relationship between ``negative'' propositions and types is not always obvious.
-A common proposition in Typed Clojure is (! (U nil false) a): the proposition that
-local binding ``a'' is \emph{not} of type (U nil false).
-This problem is most visible in expressions like {\tt (filter identity coll)}, where
-``identity'' has a ``then'' proposition that has negative information: (! (U nil false) 0),
-which reads, the 0th argument of identity does not contain (U nil false).
-
-\subsection{Multimethods}
-
-Clojure provides multimethods which can dispatch on the result of an
-arbitrary function, the \emph{dispatch function}. \emph{Methods} are
-added to the multimethod
-which are associated with a \emph{dispatch value}, which is compared to the
-result of apply the dispatch function to the argument passed to the
-multimethod. If this comparison unambiguously dispatches to one method,
-that method is then called with the current arguments.
-
-Occurrence typing, in particular when extended with paths,
-for checking method definitions for arbitrary dispatch functions.
-
-\begin{lstlisting}
-(ns example.mm
-  (:require 
-    [clojure.core.typed :refer [ann]]))
-
-(ann my-mm [Any -> Number])
-(defmulti my-mm class)
-
-(defmethod my-mm Float [f] (inc f))
-\end{lstlisting}
-
-\begin{lstlisting}
-(ann clojure.core/class
-  [Any -> (U nil Class)
-   :object {:id 0, :path [Class]}])
-\end{lstlisting}
-
-\begin{verbatim}
-TODO
-- example of normal MM usage
-- explain descrepency btwn TC syntax and formal
-  TR calculs syntax. eg. :object/:filter 
-    vs over arrow
-\end{verbatim}
-
-\subsection{Extending occurrence typing}
-
-\begin{figure*}
-$$
-\begin{altgrammar}
-  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
-                      \alt \true{} \alt \false{} \alt \nil{}
-                      \alt \class{}
-                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
-                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
-                      \alt
-                      {\doexp {\e{}} {\e{}}}
-                      \\ &\alt& 
-                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
-                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
-                &\mbox{Expressions} \\ \\
-  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
-                      \alt {\Value \v{}}
-                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
-                      \alt (\cap\ \overrightarrow{\t{}})
-                      \alt (\Not\ \t{})
-                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
-                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
-                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
-                                   {\t{}}
-                                   {\filterset {\prop{}} {\prop{}}}
-                                   {\object{}}
-                                   {\prop{}}}
-                      \\ &\alt&
-                      {\ArrayTwo {\t{}} {\t{}}}
-                &\mbox{Types} \\ \\
-  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-                      \alt {\impprop {\prop{}} {\prop{}}}
-                      \alt {\andprop {\prop{}} {\prop{}}}
-                      \alt {\orprop {\prop{}} {\prop{}}}
-                      \alt \topprop{}
-                      \alt \botprop{}
-                &\mbox{Propositions} \\
-  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
-                      \alt \emptyobject{}
-                &\mbox{Objects} \\
-  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
-                &\mbox{Paths} \\
-  \pesyntax{}   &::=& \classpe{}
-                &\mbox{Path Elements} \\ \\
-                % TODO objects with path
-%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
-%              &\mbox{Constants}
-  \Gamma       &::=& \overrightarrow{\prop{}}
-               &\mbox{Environments}
-\end{altgrammar}
-$$
-\caption{Syntax of Terms, Types, Propositions, and Objects}
-\end{figure*}
-
-\begin{figure*}
-\begin{mathpar}
-\infer [T-Var]
-{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
-{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
-             {\x{}} {\topprop{}}}
-
-\infer [T-True]
-{}
-{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} {\topprop{}}}
-
-\infer [T-False]
-{}
-{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}
-
-\infer [T-Nil]
-{}
-{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}
-
-\infer [T-Do]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
-             {\flowprop {\prop{1}}}
-\\\\
-  \judgement {\propenv{}, {\flowprop {\prop{1}}}} {\hastype {\e2} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} 
-             {\object{}} {\flowprop {\prop{2}}}
-  }
-{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e2}} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} {\object} 
-             {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}}
-
-\infer [T-If]
-{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}} {\flowprop {\prop{1}}} 
-             \\\\
-  \judgement {\propenv{}} {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
-             {\object{}} {\flowprop {\prop{2}}}
-  \\\\
-  \judgement {\propenv{}} {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
-             {\object{}} {\flowprop {\prop{3}}}
-             }
-{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
-            % do we need to include prop1+/- here?
-             {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
-                         {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
-             {\object{}} {\orprop {\flowprop {\prop{2}}}
-                                  {\flowprop {\prop{3}}}} }
-
-\infer [T-Let]
-{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}} {\flowprop {\prop{1}}}
-  \\\\
-  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
-              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
-              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
-             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} {\flowprop {\prop{2}}}
-             }
-{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
-             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                          {\object{1}}
-                          {\x{}}}
-             {\replacefor {\object{}} 
-                          {\object{1}}
-                          {\x{}}}
-             {\replacefor {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}
-                          {\object{1}}
-                          {\x{}}}}
-
-\infer [T-Throw]
-{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} {\flowprop {\prop{}}}}
-{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
-             {\emptyobject{}} {\botprop} }
-
-\infer [T-Fn]
-{}
-{ TODO }
-
-\infer [T-New]
-{ }
-{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
-             {\filterset {\topprop{}} {\topprop{}}}
-             {\emptyobject}
-             {\topprop{}}}
-
-\end{mathpar}
-\caption{Typing Rules}
-\end{figure*}
-
-%$$
-%\begin{tdisplay}{Evaluation Contexts}
-%  \begin{altgrammar}
-%    \E{} &::=& [ ] % application rules
-%              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
-%              % map rules
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
-%              &\mbox{Evaluation Contexts}
-%  \end{altgrammar}
-%\end{tdisplay}
-%$$
-
-\begin{figure*}
-$$
-\begin{altgrammar}
-{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
-\\
-{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
-    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
-            {\javatotc {\class{2}} {\true{}}}
-            {\filterset {\topprop{}} {\topprop{}}}
-            {\emptyobject{}}
-            {\topprop{}}}
-\\
-{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
-    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
-            {\javatotc {\class{2}} {\false{}}}
-            {\filterset {\topprop{}} {\topprop{}}}
-            {\emptyobject{}}
-            {\topprop{}}}
-\\\\
-
-{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
-\\
-{\javatotc {\class{}} {\false{}}} &=& \class{}
-\\
-{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \true{}}}
-\\
-{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \false{}}
-
-\end{altgrammar}
-$$
-\caption{Converting Java types to Typed Clojure types}
-\end{figure*}
-
-\subsection{Java interoperability}
-
-As Clojure supports Java interoperability, Typed Clojure has some
-integration with Java's type system. Calls to Java methods from
-Clojure can be type checked by Typed Clojure.
-
-We currently do not support Java generics
-
-\subsubsection{Arrays}
-\label{sec:arrays}
-
-Supporting statically sound interactions with Java arrays is a goal
-of Typed Clojure. This is complicated by Java's decision to make
-arrays covariant in their argument, a well documented source of static
-unsoundness. Bracha~\cite{Bra98} summarises Java's approach to maintaining
-soundness at runtime, which involves all array writes being checked by
-runtime assertions.
-
-This approach fits Java's type system, but we can do better in a more powerful
-type system like Typed Clojure. Our goal is to catch all type-incorrect array
-writes at compile time so the type system can do more to help Clojure programmers
-use arrays, especially those being passed from foreign Java code.
-
-Our basic approach is to make our array types \emph{bivariant}. Array types
-look like {\ArrayTwo {\t{w}} {\t{r}}} and
-are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
-and a covariant parameter for output (reading).
-This type can write type {\t{w}} and read type {\t{r}}.
-
-Most commonly, an array type is invariant in its parameter; it can
-write and read input of the same type.
-We can get the same effect by setting our input and output
-parameters to the same type. For example, {\ArrayTwo {\Number} {\Number}}
-(or equivalently, {\Array {\Number}})
-in Typed Clojure is similar to invariant array types of \Number in languages like Scala.
-
-The biggest gain in using a separate input parameter is the ability
-to specify \emph{read-only} arrays. Crucially, our type system features an
-explicit bottom type \lstinline|Nothing|, enabling a read-only \lstinline|Number| array
-to be of type \lstinline|(Array2 Nothing Number)|.
-
-To realise why defining read-only arrays are useful, we need to examine
-what makes array covariance unsound in Java.
-\begin{verbatim}
-FIXME
-Array covariance about the type of an array so the consumer
-of an array cannot tell the actual type of the array when examining a type
-signature.
-\end{verbatim}
-
-\begin{lstlisting}
-...
-public static Number[] getNumberArray() {
-  Number[] n = new Integer[10];
-  return n;
-}
-...
-\end{lstlisting}
-
-To the casual consumer \emph{getNumberArray} returns an array that can both
-read and write \lstinline|Number|s. However it is clear from the implementation
-that attempting to write say a \lstinline|Double| to this array will result
-in a runtime error.
-
-\begin{verbatim}
-...
-Number[] myArray = getNumberArray();
-myArray[0] = 1.1;
-/* Exception in thread "main" 
-   java.lang.ArrayStoreException: 
-   java.lang.Double */
-...
-\end{verbatim}
-
-Notice that this is a runtime error, and Java's type system has not helped
-statically prevent it.
-This could cause a similar issue for other statically-typed languages offering
-interoperability with Java.
-
-To prevent these sorts of runtime exceptions in Typed Clojure, we declare
-all arrays from unknown sources to be \emph{read-only}. Put differently,
-the only way to define a writeable array is to create it in type-checked Clojure
-code.
-
-\begin{lstlisting}
-(let [n (CovariantArray/getNumberArray)]
-  (aset n 0 1.1))
-
-; Polymorphic static method clojure.lang.RT/aset could not be 
-; applied to arguments:
-; Domains: 
-;         (Array2 i o) clojure.core.typed/AnyInteger i
-; 
-; Arguments:
-;         (Array2 Nothing java.lang.Number) int (Value 1.1)
-; 
-; with expected type:
-;         Any
-\end{lstlisting}
-
-The type inferred for the local \lstinline|n| is \lstinline|(Array2 Nothing Number)|
-which tells the type system: it is never safe to write to this array, but
-it is safe to assume \lstinline|Number|s can be read from this array.
-
-To emphasise, Typed Clojure throws a static type error. Errors like this help Clojure programmers
-use foreign Java libraries more correctly.
-
-\begin{verbatim}
-Note that Java libraries are often large 
-and complex and programmers will probably
-enjoy the extra help from the type system.
-\end{verbatim}
-
-\subsection{Handling null}
-\label{sec:null}
-
-Probably the most common pain point in Java programming is dealing with \emph{null}.
-It is crucial that Typed Clojure deals with \emph{null} intelligently so that
-all potentially erroneous interactions with \emph{null} in typed code are caught at compile time.
-We also want to represent \emph{null} in a way that is natural to Clojure programmers
-and approximates their mental model when reasoning about Clojure code.
-
-Firstly, we separate the concepts of \emph{null} and reference types at the type level.
-This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
-We provide a singleton type \lstinline|nil|, which contains just the value \lstinline|nil|,
-Clojure's equivalent of \emph{null}.
-
-Armed with general unions and the type \lstinline|nil|, 
-we make Typed Clojure's reference types \emph{non-nullable}.
-A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
-in Typed Clojure.
-
-Typed Racket already provides what is needed to write such types; indeed this is a repurposing
-of a common Typed Racket idiom of approximating an Option or Maybe type by
-making a union that includes \emph{\#f}, Racket's only false value.
-However, like handling covariant arrays, the subtlety lies in our interactions with foreign
-Java code.
-
-A set of conversion rules define how to convert a Java type signature into a null-safe
-Clojure version. These rules are designed to be conservative and can be explicitly overridden
-by programmers. Method parameters are non-nullable, while return types are nullable.
-Field types are nullable. Constructor parameters are non-nullable, and return types
-are also non-nullable: it is a Java invariant that all constructors return an instance
-of the class they are constructing.
-
-\begin{verbatim}
-TODO 
-- write conversion rules for 
-  Java type -> Clojure type
-\end{verbatim}
-
-%\appendix
-%\section{Appendix Title}
-%
-%This is the text of the appendix, if you need one.
-
-%\acks
-%
-%Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{bibliography}
-
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-11-21 formal-model.tex
fixes after meeting

@@
 says expression \e{} is of type \t{} in the 
-multimethod environment $\trdiff{\mmenv{}}$, the \emph{tools.analyzer} environment
-$\trdiff{\taenv{}}$
-and proposition environment $\propenv{}$, with 
-then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}},
-object \object{} and new multimethod environment
-$\trdiff{\mmenvp{}}$.
+proposition environment $\propenv{}$, with 
+then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
+and object \object{}.
 

### 2015-02-04 icfp15-typedclojure.tex
break out reflection calculus
more judgements
start operational semantics
start isa figure
multimethod update
change HMap app to get
add constants and other stuff
clean up syntax figure
proof
proof cases
B-Get proof
support `get` on unions in GetHMap and update

@@
 
-\documentclass[preprint,10pt]{sigplanconf}
+\documentclass[preprint,9pt]{sigplanconf}
 
@@
 \include{bibliography.bib}
+\usepackage{framed}
 
@@
 
+\usepackage{amsthm}
+
+\newtheorem{assumption}{Assumption}
+\newtheorem{lemma}{Lemma}
+
+\theoremstyle{remark}
+\newtheorem*{case}{Case}
+
+\theoremstyle{remark}
+\newtheorem*{subcase}{Subcase}
+
 \begin{document}
@@
 
-\title{Typed Clojure}
-\subtitle{An Optional Type System for Clojure, based on Typed Racket}
+\title{Practical Optional Types for Clojure}
 
@@
 
+\begin{lstlisting}
+(defn grandparent [^File f]
+  (when f
+    (let [f (. f (getParent))]
+      (when f
+        (. (new File f) (getParent))))))
+\end{lstlisting}
+%(some-> f (.getParent) (File.) (.getParent))
+
+\begin{lstlisting}
+(ann parent [(U nil File) -> (U nil String)])
+(defn parent [^File f]
+  (when f
+    (.getParent f)))
+\end{lstlisting}
+
+\begin{lstlisting}
+(defalias Expr
+  (U '{:op :if :test Expr :then Expr :else Expr}
+     '{:op :const :val Any} ...))
+(defmulti tree-str (fn [e :- Expr] (:op e)))
+(defmethod tree-str :if [{:keys [test then else]}] 
+  (str (tree-str test) (tree-str then) (tree-str else)))
+(defmethod tree-str :const [{:keys [val]}] 
+  (str val))
+\end{lstlisting}
+
 % what are the goals of Typed Clojure?
@@
 
+\input{proof}
+
 \input{syntax-figure}
@@
 \input{convert-types-figure}
+\input{constant}
+
+\input{isa-figure}
+\input{get-method}
+
+\input{operational-semantics}
+\input{update}
 

### 2015-02-06 icfp15-typedclojure.tex
fix fieldexp formatting
update T-GetHMap proof

@@
     (let [f (. f (getParent))]
-      (when f
-        (. (new File f) (getParent))))))
+      (assert f "Expected parent string, got nil")
+      (. (new File f) (getParent)))))
 \end{lstlisting}
@@
      '{:op :const :val Any} ...))
-(defmulti tree-str (fn [e :- Expr] (:op e)))
+(defmulti tree-str (fn [e :- Expr] (get :op e)))
 (defmethod tree-str :if [{:keys [test then else]}] 

### 2015-02-07 icfp15-typedclojure.tex
try to type-set the parent example

--- 2015-02-04 icfp15-typedclojure.tex
@@
 \begin{lstlisting}
-(defn grandparent [^File f]
+(ann parent [(U nil File) -> (U nil File)])
+(defn parent [^File f]
   (when f
-    (let [f (. f (getParent))]
-      (when f
-        (. (new File f) (getParent))))))
+    (let [s (. f (getParent))]
+      (assert s "Expected parent string, got nil")
+      (new File s))))
 \end{lstlisting}
+
+$$
+\abs{\x{}}{\Union{\Nil}{\File}}
+{\letexp {\hinted{\File{}}{f}} {f}
+  {\ifexp {f} 
+    {\letexp {s} {\methodstaticexp {\File{}} {\String{}} {\String{}} {getParent} {f}}
+      {\doexp {\ifexp {s}
+                {\nil{}}
+                {\errorval{input}}}
+              {\newstaticexp {\String{}} {\File{}} {\File{}} {s}}}}
+    {\nil{}}
+}}
+              $$
+
 %(some-> f (.getParent) (File.) (.getParent))
 
-\begin{lstlisting}
-(ann parent [(U nil File) -> (U nil String)])
-(defn parent [^File f]
-  (when f
-    (.getParent f)))
-\end{lstlisting}
+%\begin{lstlisting}
+%(ann parent [(U nil File) -> (U nil String)])
+%(defn parent [^File f]
+%  (when f
+%    (.getParent f)))
+%\end{lstlisting}
 
@@
      '{:op :const :val Any} ...))
-(defmulti tree-str (fn [e :- Expr] (:op e)))
+(ann tree-str [Expr -> String])
+(defmulti tree-str (fn [e :- Expr] (get :op e)))
 (defmethod tree-str :if [{:keys [test then else]}] 

### 2015-02-18 formal-model.tex
B-Get proof
support `get` on unions in GetHMap and update
new structure, see small.tex

@@
 $
-{\judgementsevcol {\mmenv}
-                 {\taenv}
-                 {\propenv}
+{\judgement   {\propenv}
               {\hastype {\e{}} {\t{}}}
@@
               {\elseprop {\prop{}}}}
-  {\object{}}
-  {\mmenvp{}}}
+  {\object{}}}
 $
@@
 and object \object{}.
-
-\subsection{Higher-rank polymorphism}
-
-Typed Clojure supports f-bounded polymorphism and higher-rank
-types.
-% why include them?
-% examples, involving generic type operators for monads & conduits
-% useful for generic programming, similar to Hashell type classes
-
-\subsection{Using negative filters}
-
-Occurrence typing plays an important role in Typed Racket and Typed Clojure.
-By maintaining a \emph{proposition environment} of propositions relating types to
-bindings, we can update bindings with more accurate types as programs progress.
-It follows that there is some correspondence between propositions and types,
-characterised by the \emph{update} function, which takes a type and a proposition
-and returns a type which updates the input type using the proposition.
-
-There is a straightforward relationship between ``positive'' propositions and types.
-For example 
-{\tt (update Number (is Integer 0))}
-updates Number by Integer, which is Integer, because Integer <: Number.
-
-The relationship between ``negative'' propositions and types is not always obvious.
-A common proposition in Typed Clojure is (! (U nil false) a): the proposition that
-local binding ``a'' is \emph{not} of type (U nil false).
-This problem is most visible in expressions like {\tt (filter identity coll)}, where
-``identity'' has a ``then'' proposition that has negative information: (! (U nil false) 0),
-which reads, the 0th argument of identity does not contain (U nil false).
 
@@
 
-\subsubsection{Arrays}
-\label{sec:arrays}
-
-Supporting statically sound interactions with Java arrays is a goal
-of Typed Clojure. This is complicated by Java's decision to make
-arrays covariant in their argument, a well documented source of static
-unsoundness. Bracha~\cite{Bra98} summarises Java's approach to maintaining
-soundness at runtime, which involves all array writes being checked by
-runtime assertions.
-
-This approach fits Java's type system, but we can do better in a more powerful
-type system like Typed Clojure. Our goal is to catch all type-incorrect array
-writes at compile time so the type system can do more to help Clojure programmers
-use arrays, especially those being passed from foreign Java code.
-
-Our basic approach is to make our array types \emph{bivariant}. Array types
-look like {\ArrayTwo {\t{w}} {\t{r}}} and
-are reminiscent of functions or pipes: having a contravariant parameter for input (writing)
-and a covariant parameter for output (reading).
-This type can write type {\t{w}} and read type {\t{r}}.
-
-Most commonly, an array type is invariant in its parameter; it can
-write and read input of the same type.
-We can get the same effect by setting our input and output
-parameters to the same type. For example, {\ArrayTwo {\Number} {\Number}}
-(or equivalently, {\Array {\Number}})
-in Typed Clojure is similar to invariant array types of \Number in languages like Scala.
-
-The biggest gain in using a separate input parameter is the ability
-to specify \emph{read-only} arrays. Crucially, our type system features an
-explicit bottom type \lstinline|Nothing|, enabling a read-only \lstinline|Number| array
-to be of type \lstinline|(Array2 Nothing Number)|.
-
-To realise why defining read-only arrays are useful, we need to examine
-what makes array covariance unsound in Java.
-\begin{verbatim}
-FIXME
-Array covariance about the type of an array so the consumer
-of an array cannot tell the actual type of the array when examining a type
-signature.
-\end{verbatim}
-
-\begin{lstlisting}
-...
-public static Number[] getNumberArray() {
-  Number[] n = new Integer[10];
-  return n;
-}
-...
-\end{lstlisting}
-
-To the casual consumer \emph{getNumberArray} returns an array that can both
-read and write \lstinline|Number|s. However it is clear from the implementation
-that attempting to write say a \lstinline|Double| to this array will result
-in a runtime error.
-
-\begin{verbatim}
-...
-Number[] myArray = getNumberArray();
-myArray[0] = 1.1;
-/* Exception in thread "main" 
-   java.lang.ArrayStoreException: 
-   java.lang.Double */
-...
-\end{verbatim}
-
-Notice that this is a runtime error, and Java's type system has not helped
-statically prevent it.
-This could cause a similar issue for other statically-typed languages offering
-interoperability with Java.
-
-To prevent these sorts of runtime exceptions in Typed Clojure, we declare
-all arrays from unknown sources to be \emph{read-only}. Put differently,
-the only way to define a writeable array is to create it in type-checked Clojure
-code.
-
-\begin{lstlisting}
-(let [n (CovariantArray/getNumberArray)]
-  (aset n 0 1.1))
-
-; Polymorphic static method clojure.lang.RT/aset could not be 
-; applied to arguments:
-; Domains: 
-;         (Array2 i o) clojure.core.typed/AnyInteger i
-; 
-; Arguments:
-;         (Array2 Nothing java.lang.Number) int (Value 1.1)
-; 
-; with expected type:
-;         Any
-\end{lstlisting}
-
-The type inferred for the local \lstinline|n| is \lstinline|(Array2 Nothing Number)|
-which tells the type system: it is never safe to write to this array, but
-it is safe to assume \lstinline|Number|s can be read from this array.
-
-To emphasise, Typed Clojure throws a static type error. Errors like this help Clojure programmers
-use foreign Java libraries more correctly.
-
-\begin{verbatim}
-Note that Java libraries are often large 
-and complex and programmers will probably
-enjoy the extra help from the type system.
-\end{verbatim}
-
 \subsection{Handling null}
@@
 We also want to represent \emph{null} in a way that is natural to Clojure programmers
-and approximates their mental model when reasoning about Clojure code.
+and models their reasoning of Clojure code.
 

### 2015-02-18 intro.tex
new structure, see small.tex

--- 2015-02-07 icfp15-typedclojure.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint,9pt]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{mmm}
-\usepackage{mathpartir}
-\usepackage{clj-grammar}
-\usepackage{xcolor}
-%\usepackage{url}
-%\usepackage{biblatex}
-\include{bibliography.bib}
-\usepackage{framed}
-
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
-
-\usepackage{amsmath}
-
-\newcommand\smallsection[1]{\bf \emph{#1}}
-
-\usepackage{amsthm}
-
-\newtheorem{assumption}{Assumption}
-\newtheorem{lemma}{Lemma}
-
-\theoremstyle{remark}
-\newtheorem*{case}{Case}
-
-\theoremstyle{remark}
-\newtheorem*{subcase}{Subcase}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-%\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Practical Optional Types for Clojure}
-
-\authorinfo{Ambrose Bonnaire-Sergeant}
-           {Indiana University}
-           {abonnair@indiana.edu}
-\authorinfo{Rowan Davies}
-           {University of Western Australia}
-           {rowan.davies@uwa.edu.au}
-\authorinfo{Sam Tobin-Hochstadt}
-           {Indiana University}
-           {samth@indiana.edu}
-
-\maketitle
-
-\begin{abstract}
-Clojure is a dynamically typed language hosted on the Java 
-Virtual Machine.
-Typed Racket is a valuable starting point for
-a gradual type system that targets Clojure.
-Building a similar type system for a new language gives the
-designer some flexibility to repurpose and extend features.
-This paper gives an overview of Typed Clojure, concentrating
-on the extensions and differences from Typed Racket. We also
-show where Typed Racket's features were particularly useful
-for type checking non-trivial Clojure idioms.
-\end{abstract}
-
-\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-\keywords
-gradual types, Clojure
-
 \section{Clojure with static typing}
@@
 
-$$
-\abs{\x{}}{\Union{\Nil}{\File}}
-{\letexp {\hinted{\File{}}{f}} {f}
-  {\ifexp {f} 
-    {\letexp {s} {\methodstaticexp {\File{}} {\String{}} {\String{}} {getParent} {f}}
-      {\doexp {\ifexp {s}
-                {\nil{}}
-                {\errorval{input}}}
-              {\newstaticexp {\String{}} {\File{}} {\File{}} {s}}}}
-    {\nil{}}
-}}
-              $$
+%$$
+%\abs{\x{}}{\Union{\Nil}{\File}}
+%{\letexp {\hinted{\File{}}{f}} {f}
+%  {\ifexp {f} 
+%    {\letexp {s} {\methodstaticexp {\File{}} {\String{}} {\String{}} {getParent} {f}}
+%      {\doexp {\ifexp {s}
+%                {\nil{}}
+%                {\errorval{input}}}
+%              {\newstaticexp {\String{}} {\File{}} {\File{}} {s}}}}
+%    {\nil{}}
+%}}
+%              $$
 
@@
 
+\subsection{Contributions}
+
+%The main idea is....... SPJ
+In this paper we present a practical type system for Clojure.
+
+\begin{itemize}
+
+  \item We formalise Typed Clojure and prove that typed code
+    \begin{itemize}
+      \item disallows null-pointer exceptions, (Section ?)
+      \item statically resolves Java interoperability (Section ?)
+    \end{itemize}
+  \item We reveal the surprising synthesis between Clojure multimethods
+        and occurrence typing.
+  \item We 
+\end{itemize}
+
 % what are the goals of Typed Clojure?
@@
 %   that knows about binding aliasing. Sounds straightforward?
-
-\input{contributions}
-
-\input{overview}
-
-\input{formal-model}
-
-\input{proof}
-
-\input{syntax-figure}
-
-\input{typing-rules}
-
-\input{tools-analyzer-figure}
-
-\input{subtyping}
-
-%$$
-%\begin{tdisplay}{Evaluation Contexts}
-%  \begin{altgrammar}
-%    \E{} &::=& [ ] % application rules
-%              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
-%              % map rules
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
-%              &\mbox{Evaluation Contexts}
-%  \end{altgrammar}
-%\end{tdisplay}
-%$$
-
-\input{convert-types-figure}
-\input{constant}
-
-\input{isa-figure}
-\input{get-method}
-
-\input{operational-semantics}
-\input{update}
-
-%\appendix
-%\section{Appendix Title}
-%
-%This is the text of the appendix, if you need one.
-
-%\acks
-%
-%Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{bibliography}
-
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2015-02-18 small.tex
new structure, see small.tex

--- 2015-02-04 icfp15-typedclojure.tex
@@
 
-\documentclass[preprint,9pt]{sigplanconf}
+\documentclass[preprint,9pt,twocolumn]{sigplanconf}
 
@@
 
-\usepackage{mmm}
-\usepackage{mathpartir}
-\usepackage{clj-grammar}
-\usepackage{xcolor}
-%\usepackage{url}
-%\usepackage{biblatex}
+\usepackage{common}
 \include{bibliography.bib}
-\usepackage{framed}
-
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
-
-\usepackage{amsmath}
-
-\newcommand\smallsection[1]{\bf \emph{#1}}
-
-\usepackage{amsthm}
-
-\newtheorem{assumption}{Assumption}
-\newtheorem{lemma}{Lemma}
-
-\theoremstyle{remark}
-\newtheorem*{case}{Case}
-
-\theoremstyle{remark}
-\newtheorem*{subcase}{Subcase}
 
@@
 
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
+\keywords
+optional types, Clojure
 
-\keywords
-gradual types, Clojure
-
-\section{Clojure with static typing}
-
-% current situation
-
-Programming is an error prone activity, especially in the presence of
-multiple languages working together. Over the past decade, it has become increasingly
-frequent to extend existing dynamically typed languages with optional type systems.
-The guarantees that a sound type system provides can be the difference between
-choosing a language that supports a type system or one that does not.
-
-% presenting clojure
-
-Clojure is a dynamically typed language built to run on the Java Virtual Machine (JVM)
-which has recently gained popularity as an alternative JVM language.
-It offers the flexibility of a Lisp dialect, including Common Lisp-style macros,
-emphasising a functional style via a standard library of immutable datastructures. 
-Clojure has good interoperability with existing Java code, allowing programmers to
-use existing Java libaries without leaving Clojure.
-
-%problem?
-
-Clojure
-
-Typed Clojure is a gradual type system for Clojure, based on the
-static portion of Typed Racket.
-
-\begin{lstlisting}
-(defn grandparent [^File f]
-  (when f
-    (let [f (. f (getParent))]
-      (when f
-        (. (new File f) (getParent))))))
-\end{lstlisting}
-%(some-> f (.getParent) (File.) (.getParent))
-
-\begin{lstlisting}
-(ann parent [(U nil File) -> (U nil String)])
-(defn parent [^File f]
-  (when f
-    (.getParent f)))
-\end{lstlisting}
-
-\begin{lstlisting}
-(defalias Expr
-  (U '{:op :if :test Expr :then Expr :else Expr}
-     '{:op :const :val Any} ...))
-(defmulti tree-str (fn [e :- Expr] (:op e)))
-(defmethod tree-str :if [{:keys [test then else]}] 
-  (str (tree-str test) (tree-str then) (tree-str else)))
-(defmethod tree-str :const [{:keys [val]}] 
-  (str val))
-\end{lstlisting}
-
-% what are the goals of Typed Clojure?
-% where do they differ from Typed Racket? why?
-
-% Things todo
-% - DONE small calculus based on previous TR formalisation
-%   that demos the ``flow'' environment & not/intersection type
-% - discussion and examples for flow filters
-% - discussion on the not/difference type, real code where it helps
-% - higher-rank polymorphism, show what it enables (eg. monads)
-%   - also show more complicated example, making an extensible
-%     type for clojure.core/conj, and why it's not straightforward
-% - discussion on limitations of local type inference
-% - demonstration of how we check multimethods
-% - dealing with array covariance
-% - note on how we avoid null-pointer exceptions
-% - discussion on why Typed Clojure only uses the ``static'' part
-%   of Typed Racket
-% - the equality filter problem (and draft solution?), a proposition
-%   that knows about binding aliasing. Sounds straightforward?
-
-\input{contributions}
+\input{intro}
 
@@
 
-\input{proof}
+\input{metatheory}
 
-\input{syntax-figure}
+\input{experience}
 
-\input{typing-rules}
+\input{related-work}
 
-\input{tools-analyzer-figure}
-
-\input{subtyping}
-
-%$$
-%\begin{tdisplay}{Evaluation Contexts}
-%  \begin{altgrammar}
-%    \E{} &::=& [ ] % application rules
-%              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
-%              % map rules
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
-%              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
-%              &\mbox{Evaluation Contexts}
-%  \end{altgrammar}
-%\end{tdisplay}
-%$$
-
-\input{convert-types-figure}
-\input{constant}
-
-\input{isa-figure}
-\input{get-method}
-
-\input{operational-semantics}
-\input{update}
-
-%\appendix
-%\section{Appendix Title}
-%
-%This is the text of the appendix, if you need one.
-
-%\acks
-%
-%Acknowledgments, if needed.
+\input{conclusion}
 
@@
 
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
 \end{document}

### 2015-02-19 formal-model.tex
smaller figures
isa figures
type hinting figure
add rewrite figure
type hint inference
tweak type hinting figure
java interop semantics
type hint prose

@@
 with differences highlighted in $\trdiff{\text{blue}}$.
-The main judgement
-
+The typing judgement 
 $
@@
 $
-
 says expression \e{} is of type \t{} in the 
 proposition environment $\propenv{}$, with 
-then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
+`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
 and object \object{}.
 
+The propositions inform conditionals on what they can assume down the then and else
+branches if the expression is used as a test.
+The object is a possibly-empty symbolic representation of the expression.
+
+\begin{figure*}
+  \footnotesize
+$$
+\begin{altgrammar}
+  \expd{}, \e{} &::=& \x{}
+                      \alt \v{} 
+                      \alt {\comb {\e{}} {\e{}}} 
+                      \alt {\abs {\x{}} {\t{}} {\e{}}}
+                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
+                      \alt {\trdiff{\doexp {\e{}} {\e{}}}}
+                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
+                &\mbox{Expressions} \\
+  \v{} &::=&          \true{} \alt \false{} \alt \trdiff{\nil{}}
+                      \alt \trdiff{\class{}}
+                      \alt \k{}
+                      \alt {\const{}}
+                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
+                &\mbox{Values} \\
+  {\const{}}           &::=& \classconst \alt \throwconst
+
+                &\mbox{Constants}\\
+  \s{}, \t{}    &::=& \Top 
+                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
+                      \alt
+                      {\ArrowOne {\x{}} {\t{}}
+                                   {\t{}}
+                                   {\filterset {\prop{}} {\prop{}}}
+                                   {\object{}}}
+                      \alt {\Value \singletonmeta{}} 
+                &\mbox{Types} \\
+  \singletonallsyntax{}
+                \\ \\
+  \occurrencetypingsyntax{}
+  %\\
+  %\classliteralallsyntax{}
+\end{altgrammar}
+$$
+\caption{Syntax of Terms, Types, Propositions and Objects}
+\end{figure*}
+
+\begin{figure*}
+  \footnotesize
+\begin{mathpar}
+  {\TLocal}
+
+  {\TTrue}
+
+{\TFalse}
+
+{\TNil}
+
+{\TDo}
+
+{\TIf}
+
+{\TLet}
+
+{\TApp}
+
+{\TAbs}
+
+         {\TSubsume}
+\end{mathpar}
+\caption{Standard typing rules}
+\end{figure*}
+
+\begin{figure*}
+  \footnotesize
+\begin{mathpar}
+  {\BLocal}
+
+  {\BDo}
+
+  {\BLet}
+
+  \BVal{}
+
+\BIfTrue{}
+
+\BIfFalse{}
+
+\BAbs{}
+
+\BBetaClosure{}
+
+\BDelta{}
+\end{mathpar}
+\caption{Semantics}
+\end{figure*}
+
+\begin{figure*}
+  \footnotesize
+\begin{mathpar}
+\standardsubtyping{}
+\end{mathpar}
+\caption{Subtyping rules}
+\end{figure*}
+
+\subsection{\isaliteral{}}
+
+One interesting Clojure function we modeled is \isaliteral.
+It takes two arguments, and returns \true{} if they are the same (equality mode)
+or if they are both classes and the left is a subtype of the right (subclassing mode),
+otherwise returns \false{}, as described in figure~\ref{figure:isasemantics}.
+
+For example, {\isaapp{\Keyword}{\Object}} = \true{} but 
+{\isaapp{\Object}{\Keyword}} = \false{}.
+This is useful in combination with \classconst{}
+with a fixed second argument, where {\isaapp{\appexp{\classconst}{\x{}}}{\Keyword}}
+is a test for \x{} being a \Keyword{}, and also
+as an equality test like
+{\isaapp{\x{}}{\makekw{a}}}.
+
+We encode this behaviour in figure~\ref{figure:isatyping}. 
+\isacompareliteral{} is the interesting part,
+which takes the object of the first argument 
+and the type of the second argument, and returns the then and else propositions
+of the expression.
+
+The first case handles subclassing mode
+in expressions like {\isaapp{\appexp{\classconst}{\x{}}}{\Object}},
+where the first argument has an object ending in \classpe{} and the second argument has a singleton class type.
+For example, in this case \isacomparetwo{\path{\classpe{}}{\x{}}}{\Value{\Object}}
+gives a then proposition {\isprop{\Object} {\x{}}}
+and an else proposition {\notprop{\Object}{\x{}}}.
+
+The reason to peel off one path element is subtle. For example,
+claiming {\isprop{\Value{\Object}} {\path{\classpe{}}{\x{}}}}
+might seem suitable, however it implies \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Object}
+which is wrong if \x{} = \makekw{a} where \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Keyword}.
+(In formulating \isacompareliteral we found a three year old bug in how core.typed handles this case).
+
+The second case handles the equality mode of \isaliteral{}
+with a non-class singleton type as the second argument, to be certain we are not in subclassing mode.
+Substitution is used to ensure well formed propositions.
+
+The final case gives up; we cannot distinguish between equality and subclassing mode.
+
+\isaliteral{} serves as the default dispatch mechanism for multimethods in Clojure.
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \isaopsemfigure{}
+
+    \BIsA{}
+  \end{mathpar}
+  \caption{\isaliteral{} semantics}
+  \label{figure:isasemantics}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \isapropsfigure{}
+
+    \TIsA{}
+  \end{mathpar}
+  \caption{\isaliteral{} typing rules}
+  \label{figure:isatyping}
+\end{figure}
+
 \subsection{Multimethods}
+
+\begin{figure}
+  \footnotesize
+$$
+\begin{altgrammar}
+  \e{} &::=& \ldots \alt \multimethodexpsyntax{} 
+                &\mbox{Expressions} \\
+  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
+                &\mbox{Values} \\
+
+ \disptablesyntax{} \\
+\end{altgrammar}
+$$
+\caption{Multimethod Syntax}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+\begin{mathpar}
+  \TDefMulti{}
+
+  \TDefMethod{}
+\end{mathpar}
+\caption{Multimethod typing rules}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+\begin{mathpar}
+  \Multisubtyping{}
+\end{mathpar}
+\caption{Multimethod subtyping rules}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \getmethodfigure{}
+\begin{mathpar}
+
+       \BDefMulti{}
+
+       \BDefMethod{}
+
+  \BBetaMulti{}
+\end{mathpar}
+\caption{Multimethod semantics}
+\end{figure}
 
@@
 
-\subsection{Extending occurrence typing}
+\subsection{Heterogenous maps}
+
+\begin{figure}
+  \footnotesize
+$$
+\begin{altgrammar}
+  \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
+                &\mbox{Expressions} \\
+  \v{} &::=& \ldots \alt {\emptymap{}}
+                &\mbox{Values} \\
+  \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
+                &\mbox{Types} \\
+  \auxhmapsyntax{}\\
+\end{altgrammar}
+$$
+\caption{HMap Terms and Types}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+\begin{mathpar}
+  \HMapsubtyping{}
+\end{mathpar}
+\caption{HMap subtyping rules}
+\end{figure}
 
 \subsection{Java interoperability}
+
+\begin{figure}
+  \footnotesize
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
+    &\mbox{Expressions} \\
+    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
+    &\mbox{Values} \\
+  \end{altgrammar}
+  $$
+  \caption{Java Interoperability Terms and Types}
+\end{figure}
 
@@
 
-Clojure comes with a small type system that helps improve
-performance in Clojure code: type hints. They are used to help
-the compiler resolve Java methods and prevent unboxing in
-certain situations. Tags are propagated bidirectionally
+\begin{figure}
+  \footnotesize
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
+    &\mbox{Expressions} \\
+  \end{altgrammar}
+  $$
+  \caption{Type Hinting and Non-Reflective Terms}
+\end{figure}
+
+\begin{figure*}
+  \footnotesize
+  \begin{mathpar}
+    \RAbs{}
+
+    \RNewElimRefl{}
+
+    \RMethodElimRefl{}
+
+    \RFieldElimRefl{}
+
+    \RLet{}
+
+    \RLetHint{}
+  \end{mathpar}
+\caption{Reflection Elimination}
+\label{figure:rewrite}
+\end{figure*}
+
+\begin{figure*}
+  \footnotesize
+  \begin{mathpar}
+    \TALocal{}
+
+    \TANil{}
+
+    \TANewStatic{}
+
+    \TALetHint{}
+
+    \TALet{}
+  \end{mathpar}
+\caption{Type Hint Inference}
+\label{figure:hintinfer}
+\end{figure*}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \BField{}
+
+    \BMethod{}
+
+    \BNew{}
+  \end{mathpar}
+  \caption{Java Interoperability Semantics}
+\end{figure}
+
+Clojure supports \emph{type hints}, an optional type system to improve performance
+and disambiguate Java calls.
+Programmers communicate to the Clojure compiler via type hints
+to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
 and the programmer is free to add them where they like.
+
+core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes solved
+reflection, however to tell the full story we include a rewrite relation 
+in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
+Figure~\ref{figure:hintinfer} models how type hints are propagated.
+
+We implicitly resolve reflection before type checking and remaining reflective calls are type errors.
 
@@
 \end{verbatim}
+
+\input{constant}

### 2015-02-19 formal-model.tex
example rewrite

@@
 {\TDo}
+
+{\TClass}
 
@@
   \begin{altgrammar}
-    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
+    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}\\
+            &\alt&  \localhintsyntax{} \alt \lethintsyntax{}
     &\mbox{Expressions} \\
@@
   \caption{Java Interoperability Terms and Types}
+  \label{figure:reflectivesyntax}
 \end{figure}
@@
   \begin{altgrammar}
-    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
-    &\mbox{Expressions} \\
+    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}\\
+    &&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
+    \ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
+
+    \mbox{Expressions}
   \end{altgrammar}
@@
   \caption{Type Hinting and Non-Reflective Terms}
+  \label{figure:nonreflectivesyntax}
 \end{figure}
@@
 
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    {\TNewStatic}
+
+    {\TFieldStatic}
+
+    {\TMethodStatic}
+  \end{mathpar}
+  \caption{Java Interoperability Typing Rules}
+  \label{figure:javatyping}
+\end{figure}
+
 Clojure supports \emph{type hints}, an optional type system to improve performance
@@
 
-core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes solved
+core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
 reflection, however to tell the full story we include a rewrite relation 
@@
 
-We implicitly resolve reflection before type checking and remaining reflective calls are type errors.
+We now have two languages: an internal language that may contain non-reflective Java calls (figure 
+  \ref{figure:nonreflectivesyntax}) 
+  and an external language that resembles Clojure, where we must rely on type-hints to resolve
+  reflective Java calls (figure~\ref{figure:reflectivesyntax}).
+
+For example, R-FieldElimRefl, which rewrites a reflective field call to be non-reflective,
+rewrites the target expression and uses the result to infer a class type hint as the field target.
+\fieldtypeliteral{} uses the inferred type hint to lookup if the inferred class has the appropriate
+field name and returns the actual target and result type hints which we store in the rewritten rule.
+
+We then extract these type hints in T-FieldStatic for typing checking 
+(figure~\ref{figure:javatyping}), which converts type hints
+with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
+dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).
+
+Method and constructor calls are similar, but T-MethodStatic and T-NewStatic disallow passing \nil{} as arguments, 
+we assume methods can return \nil{} and we assume constructors cannot.
+
+We rewrite to our internal language before type checking and remaining reflective calls are type errors.
 

### 2015-02-19 formal-model.tex
more figures

@@
                       \alt {\letexp {\x{}} {\e{}} {\e{}}}
+                      \alt {\errorvalv{}}
                 &\mbox{Expressions} \\
@@
                       \alt {\Value \singletonmeta{}} 
+                      \alt \trdiff{\class{}}
                 &\mbox{Types} \\
@@
 
+    {\TKw}
+
     {\TApp}
@@
     {\TAbs}
+
+    {\TError}
 
@@
 
-\begin{verbatim}
-TODO
-- example of normal MM usage
-- explain descrepency btwn TC syntax and formal
-  TR calculs syntax. eg. :object/:filter 
-    vs over arrow
-\end{verbatim}
-
 \subsection{Heterogenous maps}
@@
   \caption{HMap Terms and Types}
+  \label{figure:hmapsyntax}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    {\TGetHMap}
+
+    {\TGetAbsent}
+
+    {\TGetHMapPartialDefault}
+
+    {\TAssoc}
+  \end{mathpar}
+  \caption{HMap Typing Rules}
+  \label{figure:hmaptyping}
 \end{figure}
@@
 \end{figure}
+
+It is common in Clojure to use plain immutable hash maps to represent ad-hoc
+objects with a known set of keyword entries. We use \HMapliteral{}
+types to track known present, known absent and unknown keyword entries
+(figure~\ref{figure:hmapsyntax}).
+
+The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
+specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
+entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
+HMap can assume all entries not in {\mandatory{}} as absent.
+
+\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
+To simplify presentation, we only allow the empty map literal and
+restrict lookup and extension to keyword keys.
 
@@
   $$
-  \caption{Type Hinting and Non-Reflective Terms}
+  \caption{Non-Reflective Java Interoperability Syntax}
   \label{figure:nonreflectivesyntax}
@@
 
-\begin{verbatim}
-TODO 
-- write conversion rules for 
-  Java type -> Clojure type
-\end{verbatim}
-
 \input{constant}
+\input{update}
+\input{convert-types-figure}

### 2015-02-19 metatheory.tex
more figures

\section{Metatheory}

{\javaassumptionsall}

{\istruefalsedefinitions}

{\consistentwithdefinition}

\begin{lemma}
  {\soundnesslemmahypothesis}
\end{lemma}

### 2015-02-19 formal-model.tex
intro section 3

@@
 The object is a possibly-empty symbolic representation of the expression.
+
+Syntax is given in figure~\ref{figure:termsyntax}. Expressions include variables, values,
+application, abstractions, conditionals, do (sequencing) expressions and let expressions.
+We model throwable errors without a catching mechanism. All binding forms introduce fresh
+variables.
+
+Values include booleans, \nil{}, class literals, keywords, constants and closures.
+\false{} and \nil{} are the only logically false values.
+
+Types include the top type, untagged unions, functions, singleton types
+and class instances. We abbreviate \EmptyUnion{} as \Bot{}.
+Function types contain \emph{latent} propositions and object, which, along with the return type,
+may refer to the parameter identifier. They are latent because they are instantiated with the
+actual object of the argument in applications before they are used in the proposition environment.
+
+Propositions are identical to~\cite{TF10}; positive and negative type propositions, implications,
+conjunctions, disjuctions, trivial and impossible propositions.
+Type propositions contain a non-empty object which it claims type information about.
+
+Proposition environments are sequences of propositions in conjunctive normal form.
+
+Typing rules are in figure~\ref{figure:typingrules}.
+T-Error neither returns logically true or false as it does not return normally.
+T-Do uses the propositions learnt in the first expression to type check the second.
+Other than compensating for the extra logically false value \nil{}, the rest are standard.
+
+Figure~\ref{figure:standardopsem} contains the operational semantics, which are standard
+except for \nil{} being false.
+
+Subtyping (figure~\ref{figure:subtyping}) says singleton types are instances of their respective classes
+and \Object is transitively a supertype to everything except {\Nil{}},
+so \Top{} = {\Union{\Nil}{\Object}}.
 
@@
 \caption{Syntax of Terms, Types, Propositions and Objects}
+\label{figure:termsyntax}
 \end{figure*}
@@
   \end{mathpar}
-  \caption{Standard typing rules}
+  \caption{Typing rules}
+  \label{figure:typingrules}
 \end{figure*}
@@
   \end{mathpar}
-  \caption{Semantics}
+  \caption{Operational Semantics}
+  \label{figure:standardopsem}
 \end{figure*}
@@
 \caption{Subtyping rules}
+\label{figure:subtyping}
 \end{figure*}

### 2015-02-19 intro.tex
use minted for syntax highlighting

@@
 
-\begin{lstlisting}
+\begin{minted}[]{clojure}
 (ann parent [(U nil File) -> (U nil File)])
@@
       (new File s))))
-\end{lstlisting}
+\end{minted}
 
@@
 
-\begin{lstlisting}
+\begin{minted}[mathescape,
+  frame=lines,
+  framesep=2mm]{clojure}
 (defalias Expr
@@
   (str val))
-\end{lstlisting}
+\end{minted}
 

### 2015-02-19 small.tex
use minted for syntax highlighting
Abstract.

@@
 
-\documentclass[preprint,9pt,twocolumn]{sigplanconf}
+\documentclass[preprint,9pt,twocolumn,nocopyrightspace,authoryear]{sigplanconf}
 
@@
 \usepackage{common}
+\usepackage{minted}
 \include{bibliography.bib}
@@
 \begin{abstract}
-Clojure is a dynamically typed language hosted on the Java 
-Virtual Machine.
-Typed Racket is a valuable starting point for
-a gradual type system that targets Clojure.
-Building a similar type system for a new language gives the
-designer some flexibility to repurpose and extend features.
-This paper gives an overview of Typed Clojure, concentrating
-on the extensions and differences from Typed Racket. We also
-show where Typed Racket's features were particularly useful
-for type checking non-trivial Clojure idioms.
+Typed Clojure is an optional type system for Clojure, a dynamic
+language in the Lisp family that targets the JVM. Typed Clojure's type
+system build on the design of Typed Racket, repurposing in particular
+\emph{occurence typing}, its approach to statically reasoning about
+predicate tests. However, in adapting the type system to Clojure,
+changes and extensions are required to accomodate both additional
+language features and  idioms used by Clojure programmers.
+
+In this paper, we describe Typed Clojure and present these type system
+extensions, focusing on three features widely used in Clojure. First, Java
+interoperability is central to Clojure's mission but introduces
+challenges such as ubiquitous \texttt{null}; Typed Clojure handles
+Java interoperability while ensuring the absence of null-pointer
+exceptions in typed programs. Second, Clojure programmers
+idiomatically use immutable dictionaries for data structures; Typed
+Clojure handles this in the type system with multiple forms of
+heterogenous dictionary types. Third, multimethods provide extensible
+operations, and their Clojure semantics turns out to have a surprising
+synergy with the underlying occurence typing framework.
+
+We provide a formal model of the Typed Clojure type system
+incorporating these and other features, with a proof of
+soundness. Additionally, Typed Clojure is now in use by numerous
+corporations and developers working with Clojure, and we report on 
+experience with the system and its lessons for the future.
+
+%% Clojure is a dynamically typed language hosted on the Java 
+%% Virtual Machine.
+%% Typed Racket is a valuable starting point for
+%% a gradual type system that targets Clojure.
+%% Building a similar type system for a new language gives the
+%% designer some flexibility to repurpose and extend features.
+%% This paper gives an overview of Typed Clojure, concentrating
+%% on the extensions and differences from Typed Racket. We also
+%% show where Typed Racket's features were particularly useful
+%% for type checking non-trivial Clojure idioms.
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+%% \category{CR-number}{subcategory}{third-level}
 
-\keywords
-optional types, Clojure
+%% \keywords
+%% optional types, Clojure
 

### 2015-02-19 formal-model.tex
use minted for syntax highlighting
do example

@@
 \end{figure*}
+
+\subsection{\doliteral{}}
+
+Assertions in Clojure often expand to simple exceptional control flow reasoning.
+
+\begin{minted}{clojure}
+(fn [x :- (U nil Number)] {:pre (number? x)} (inc x))
+\end{minted}
+
+roughly expands to
+
+\begin{minted}{clojure}
+(fn [x :- (U nil Number)]
+  (do (if x (throw (new Exception)) nil)
+      (inc x)))
+\end{minted}
+
+With an elegant extension in T-Do (figure~\ref{figure:typingrules}) we can follow this reasoning
+to infer \mintinline{clojure}{(inc x)} is safe. The main insight is we can assume the
+disjunction of the propositions of the first \doliteral{} argument while checking the second.
+
+To sketch how this example is checked, first T-Abs type checks the body under the 
+environment \propenv{} = {\isprop{\Union{\nil}{\Number{}}}{\x{}}}.
+T-Do then extracts the propositions of its first argument:
+T-If tells us \thenprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} and 
+\elseprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
+using T-Error, T-Nil and T-Subsume.
+Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
+and we check \mintinline{clojure}{(inc x)} in the extended environment 
+\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}
+and by L-Update we know {\isprop{\Number}{\x{}}}
+so T-Local says \mintinline{clojure}{x} is a \Number.
 

### 2015-02-20 formal-model.tex
multimethod

@@
 Types include the top type, untagged unions, functions, singleton types
-and class instances. We abbreviate \EmptyUnion{} as \Bot{}.
+and class instances. We abbreviate \EmptyUnion{} as \Bot{}, \Value{\Nil{}} as \Nil{} and similarly
+for \True and \False.
 Function types contain \emph{latent} propositions and object, which, along with the return type,
@@
                 \\ \\
-  \occurrencetypingsyntax{}
+  \occurrencetypingsyntax{}\\
+  \openvsyntax{}
   %\\
@@
 To sketch how this example is checked, first T-Abs type checks the body under the 
-environment \propenv{} = {\isprop{\Union{\nil}{\Number{}}}{\x{}}}.
+environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
 T-Do then extracts the propositions of its first argument:
@@
 which is wrong if \x{} = \makekw{a} where \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Keyword}.
-(In formulating \isacompareliteral we found a three year old bug in how core.typed handles this case).
+(Formulating \isacompareliteral revealed a three year old bug in the implementation).
 
@@
 
-\begin{figure}
+\begin{figure*}
   \footnotesize
@@
                 &\mbox{Values} \\
+  \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
+                &\mbox{Types} \\
 
@@
 \caption{Multimethod Syntax}
-\end{figure}
+\label{figure:mmsyntax}
+\end{figure*}
 
@@
   \caption{Multimethod typing rules}
+  \label{figure:mmtyping}
 \end{figure}
@@
   \begin{mathpar}
-
     \BDefMulti{}
@@
   \caption{Multimethod semantics}
+  \label{figure:mmsemantics}
 \end{figure}
 
-Clojure provides multimethods which can dispatch on the result of an
-arbitrary function, the \emph{dispatch function}. \emph{Methods} are
-added to the multimethod
-which are associated with a \emph{dispatch value}, which is compared to the
-result of apply the dispatch function to the argument passed to the
-multimethod. If this comparison unambiguously dispatches to one method,
-that method is then called with the current arguments.
+A multimethod in Clojure is a function with a dispatch mechanism to
+of internal methods. Multimethods are created with {\mintinline{clojure}{defmulti}}
+and methods are installed via {\mintinline{clojure}{defmethod}}.
+They are called like regular functions.
 
-Occurrence typing, in particular when extended with paths,
-for checking method definitions for arbitrary dispatch functions.
+\begin{minted}{clojure}
+(ann foo [Any -> String])
+(defmulti foo class)
+(defmethod foo Number [x] (str (inc x)))
+(defmethod foo Keyword [x] (str (name x)))
+(foo :a) ;=> "a"
+(foo 1)  ;=> "2"
+\end{minted}
 
-\begin{lstlisting}
-(ns example.mm
-  (:require 
-    [clojure.core.typed :refer [ann]]))
+To ease presentation, we present \emph{immutable}
+multimethods (figure~\ref{figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
+without changing the original multimethod. The above is now written
 
-(ann my-mm [Any -> Number])
-(defmulti my-mm class)
+%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\mintinline{clojure}{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
+%                       {\classconst{}}}
+%                     {\\ bar}$
 
-(defmethod my-mm Float [f] (inc f))
-\end{lstlisting}
+\begin{minted}{clojure}
+(let [foo (defmulti [Any -> String] class)
+      foo (defmethod foo Number [x] (str (inc x)))
+      foo (defmethod foo Keyword [x] (str (name x)))]
+  (foo :a)) ;=> "a"
+\end{minted}
 
-\begin{lstlisting}
-(ann clojure.core/class
-  [Any -> (U nil Class)
-   :object {:id 0, :path [Class]}])
-\end{lstlisting}
+Multimethod semantics are in figure~\ref{figure:mmsemantics}.
+B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
+B-DefMethod returns a new multimethod with an extended dispatch table.
+B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
+which \getmethodliteral{} uses to choose the method to invoke.
+
+The typing rules (figure~\ref{figure:mmtyping}) utilse the {\MultiFntype{\s{}}{\t{}}}
+type, where \s{} is the interface type for beta reduction, and \t{} is the type for
+the dispatch function. T-DefMulti uses the provided interface type and dispatch function
+to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
+that must be true if this method is being invoked. The example can be type checked without
+changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
+method bodies.
 
@@
   \begin{altgrammar}
-    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}\\
-            &\alt&  \localhintsyntax{} \alt \lethintsyntax{}
+    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
+            \alt  \localhintsyntax{} \alt \lethintsyntax{}
     &\mbox{Expressions} \\
@@
     &\mbox{Values} \\
+    \tatypesyntax{}\\
+    \typehintenvsyntax{}\\
+    \classtableallsyntax{}
   \end{altgrammar}
@@
   \label{figure:reflectivesyntax}
-\end{figure}
+\end{figure*}
 
@@
   \begin{altgrammar}
-    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}\\
-    &&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
-    \ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
+    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
+    %&&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
+    %\ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
 
-    \mbox{Expressions}
+    &\mbox{Expressions}
   \end{altgrammar}
   $$
-  \caption{Non-Reflective Java Interoperability Syntax}
+  \caption{Internal Language Non-Reflective Java Interoperability Syntax}
   \label{figure:nonreflectivesyntax}
-\end{figure}
+\end{figure*}
 

### 2015-02-20 intro.tex
start on intro

@@
 % current situation
+
+The popularity of dynamically-typed languages languages in software
+development, combined with a recognition that types often improve
+programmer productivity, software reliability, and performance, has
+led to the recent development of a wide variety of optional and
+gradual type systems aimed at checking existing programs written in
+existing languages.  These include Microsoft's TypeScript for
+JavaScript, Facebook's Hack for PHP and Flow for JavaScript, and MyPy
+for Python among the optional systems, and Typed Racket, Reticulated
+Python, and GradualTalk among gradually-typed system.\footnote{We
+  reserve the term ``gradual typing'' for systems which soundly
+  interoperate between typed and untyped code; systems which do not
+  enforce type invariants we describe as ``optionally typed''.}
+%FIXME: add citations for all those systems.
+
+One key lesson of these systems, indeed a lesson known to early
+developers of optional type systems such as StrongTalk, is that type
+systems for existing languages must be designed to work with the
+features and idioms of the target language. Often this takes the form
+of a core language, be it of functions or classes and objects,
+together with extensions to handle distinctive language features.
 

### 2015-02-20 formal-model.tex
HMap prose

@@
   \caption{Multimethod subtyping rules}
+  \label{figure:mmsubtype}
 \end{figure}
@@
 
-The typing rules (figure~\ref{figure:mmtyping}) utilse the {\MultiFntype{\s{}}{\t{}}}
-type, where \s{} is the interface type for beta reduction, and \t{} is the type for
+The typing rules (figure~\ref{figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
+where \s{} is the interface type for beta reduction, and \t{} is the type for
 the dispatch function. T-DefMulti uses the provided interface type and dispatch function
@@
 changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
-method bodies.
+method bodies. T-App handles beta reduction because multimethods are subtypes of functions
+(figure~\ref{figure:mmsubtype}).
 
@@
   \begin{mathpar}
+    {\BAssoc}
+
+    {\BGet}
+
+    {\BGetMissing}
+  \end{mathpar}
+  \caption{HMap Semantics}
+  \label{figure:hmapsem}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
     {\TGetHMap}
@@
   \caption{HMap subtyping rules}
+  \label{figure:hmapsubtype}
 \end{figure}
@@
 entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
-HMap can assume all entries not in {\mandatory{}} as absent.
+HMap can assume all entries not in {\mandatory{}} as absent. We disallow keys to be both present and absent.
 
@@
 To simplify presentation, we only allow the empty map literal and
-restrict lookup and extension to keyword keys.
+restrict lookup and extension to keyword keys. The runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}
+is not part of the language syntax.
+
+Figure~\ref{figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
+entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
+map.
+
+Typing rules are given in figure~\ref{figure:hmaptyping}. T-GetHMap handles the simple
+case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
+a \Nil{} return type.
+T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
+T-Assoc extends the known keyword entries, overwriting previous entries with the same key.
+
+S-HMap (figure~\ref{figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
+subtypes and also cover the absent keys.
+Complete maps are subtypes of partial maps by S-HMapP.
 

### 2015-02-20 formal-model.tex
java interop

@@
 objects with a known set of keyword entries. We use \HMapliteral{}
-types to track known present, known absent and unknown keyword entries
-(figure~\ref{figure:hmapsyntax}).
+types (figure~\ref{figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.
 
@@
 
-\subsection{Java interoperability}
+\subsection{Java Interoperability and Type Hints}
 
@@
             \alt  \localhintsyntax{} \alt \lethintsyntax{}
-    &\mbox{Expressions} \\
-    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
-    &\mbox{Values} \\
-    \tatypesyntax{}\\
-    \typehintenvsyntax{}\\
-    \classtableallsyntax{}
+    &\mbox{Expressions}
   \end{altgrammar}
@@
 
-As Clojure supports Java interoperability, Typed Clojure has some
-integration with Java's type system. Calls to Java methods from
-Clojure can be type checked by Typed Clojure.
+Clojure supports interoperability with Java, including the ability to
+call constructors, methods and access fields. We model
+these operations in a restricted setting without class inheritance,
+overloading or Java Generics.
 
-We do not support Java Generics for several reasons. Firstly
-Clojure programmers already use \emph{type hints} to resolve
-Java methods which does not interact with generics. Secondly
-it would complicate the type system to need to consider advanced
-Generics features like existentials and f-bounded polymorphism
-which has complicated other type systems like Scala's.
-Finally Java interop is rare in Clojure code, so the extra
-work needed would not be worth the effort compared to the number
-of new bugs core.typed catches.
-
-Thanks to type erasure, only unparameterised classes, primitive
-types and array classes live at runtime. This simplifies 
-inference for \emph{tools.analyzer}, which uses type hints
-to resolve ambiguities in Java interop like overloaded methods.
-
-\subsection{Clojure type hints}
+We extend the external language with Java field lookups and calls to
+methods and constructors (figure~\ref{figure:nonreflectivesyntax}).
+There are no typing or evaluation rules for these syntaxes because
+Typed Clojure must be convinced that Java interactions are \emph{non-reflective},
+ie., are unambiguously resolved at compile-time. Programmers use
+type hinted expressions (figure~\ref{figure:nonreflectivesyntax}) 
+to help the compiler eliminate reflection.
 
@@
     \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
-    %&&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
-    %\ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
+    &\mbox{Expressions}\\
 
-    &\mbox{Expressions}
+    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
+    &\mbox{Values} \\
+
+    \tatypesyntax{}\\
+    \typehintenvsyntax{}\\
+    \classtableallsyntax{}
   \end{altgrammar}
   $$
-  \caption{Internal Language Non-Reflective Java Interoperability Syntax}
+  \caption{Internal Language Reflection Resolution Extensions}
   \label{figure:nonreflectivesyntax}
@@
 in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
-Figure~\ref{figure:hintinfer} models how type hints are propagated.
+Figure~\ref{figure:hintinfer} shows the interesting cases in type hint inference.
 
@@
 
-\subsection{Handling null}
-\label{sec:null}
-
-Probably the most common pain point in Java programming is dealing with \emph{null}.
-It is crucial that Typed Clojure deals with \emph{null} intelligently so that
-all potentially erroneous interactions with \emph{null} in typed code are caught at compile time.
-We also want to represent \emph{null} in a way that is natural to Clojure programmers
-and models their reasoning of Clojure code.
-
-Firstly, we separate the concepts of \emph{null} and reference types at the type level.
-This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
-We provide a singleton type \Nil{}, which contains just the value \lstinline|nil|,
-Clojure's equivalent of \emph{null}.
-
-Armed with general unions and the type \lstinline|nil|, 
-we make Typed Clojure's reference types \emph{non-nullable}.
-A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
-in Typed Clojure.
-
-Typed Racket already provides what is needed to write such types; indeed this is a repurposing
-of a common Typed Racket idiom of approximating an Option or Maybe type by
-making a union that includes \emph{\#f}, Racket's only false value.
-However, like handling covariant arrays, the subtlety lies in our interactions with foreign
-Java code.
-
-A set of conversion rules define how to convert a Java type signature into a null-safe
-Clojure version. These rules are designed to be conservative and can be explicitly overridden
-by programmers. Method parameters are non-nullable, while return types are nullable.
-Field types are nullable. Constructor parameters are non-nullable, and return types
-are also non-nullable: it is a Java invariant that all constructors return an instance
-of the class they are constructing.
-
 \input{constant}

### 2015-02-20 metatheory.tex
add theorem statements

@@
 \section{Metatheory}
+
+Our proof for type soundness is similar to~\cite{TF10}. We add
+errors and prove well-typed programs can never go wrong.
+
+We make our assumptions about Java explicit. We concede that
+method and constructor calls may diverge or error, but we assume they can
+never go wrong.
 
@@
 
+For readability we define logical truth in Clojure.
+
 {\istruefalsedefinitions}
+
+We use an extra lemma to support our main soundness lemma. Consistency
+ensures that occurrence typing does not claim the outer 
+\mintinline{clojure}{x} is a keyword.
+
+\begin{minted}{clojure}
+(let [x 1]
+  ((let [x :a]
+     (fn [] x))))
+\end{minted}
 
@@
 
-\begin{lemma}
+Our main lemma says if there is a defined reduction, then the propositions, object
+and type are correct.
+The metavariable \definedreduction{} ranges over \v{}, \errorvalv{} and \wrong{}.
+
+\begin{lemma}\label{lemma:soundness}
+
   {\soundnesslemmahypothesis}
+  \begin{proof}
+    By induction on the derivation of the typing judgement.
+  \end{proof}
 \end{lemma}
+
+% Uses above lemma:soundness label ^^^
+{\soundnesstheorem}
+
+{\wrongtheorem}

### 2015-02-21 intro.tex
formatting
tweak intro example. Ack.

@@
 Programming is an error prone activity, especially in the presence of
-multiple languages working together. Over the past decade, it has become increasingly
-frequent to extend existing dynamically typed languages with optional type systems.
-The guarantees that a sound type system provides can be the difference between
-choosing a language that supports a type system or one that does not.
+multiple languages working together. Over the past decade, it has
+become increasingly frequent to extend existing dynamically typed
+languages with optional type systems.  The guarantees that a sound
+type system provides can be the difference between choosing a language
+that supports a type system or one that does not.
 
@@
   (when f
-    (let [s (. f (getParent))]
+    (let [s (.getParent f)]
       (assert s "Expected parent string, got nil")
@@
 
-\begin{minted}[mathescape,
-  frame=lines,
-  framesep=2mm]{clojure}
+\begin{minted}[]{clojure}
 (defalias Expr

### 2015-02-21 overview.tex
add examples

@@
 
+\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg1.clj}
+\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg2.clj}
+\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg3.clj}
+\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg4.clj}
+\inputminted[firstline=4,lastline=15]{clojure}{code/demo/src/demo/eg5.clj}
+\inputminted[firstline=4,lastline=15]{clojure}{code/demo/src/demo/eg6.clj}
+\inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+
 % TODO references

### 2015-02-21 formal-model.tex
remove class table lookup figure from main paper

@@
 
-Syntax is given in figure~\ref{figure:termsyntax}. Expressions include variables, values,
+Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
 application, abstractions, conditionals, do (sequencing) expressions and let expressions.
@@
 
-Typing rules are in figure~\ref{figure:typingrules}.
+Typing rules are in figure~\ref{main:figure:typingrules}.
 T-Error neither returns logically true or false as it does not return normally.
@@
 
-Figure~\ref{figure:standardopsem} contains the operational semantics, which are standard
+Figure~\ref{main:figure:standardopsem} contains the operational semantics, which are standard
 except for \nil{} being false.
 
-Subtyping (figure~\ref{figure:subtyping}) says singleton types are instances of their respective classes
+Subtyping (figure~\ref{main:figure:subtyping}) says singleton types are instances of their respective classes
 and \Object is transitively a supertype to everything except {\Nil{}},
@@
 \caption{Syntax of Terms, Types, Propositions and Objects}
-\label{figure:termsyntax}
+\label{main:figure:termsyntax}
 \end{figure*}
@@
   \caption{Typing rules}
-  \label{figure:typingrules}
+  \label{main:figure:typingrules}
 \end{figure*}
@@
   \caption{Operational Semantics}
-  \label{figure:standardopsem}
+  \label{main:figure:standardopsem}
 \end{figure*}
@@
 \caption{Subtyping rules}
-\label{figure:subtyping}
+\label{main:figure:subtyping}
 \end{figure*}
@@
 
-With an elegant extension in T-Do (figure~\ref{figure:typingrules}) we can follow this reasoning
+With an elegant extension in T-Do (figure~\ref{main:figure:typingrules}) we can follow this reasoning
 to infer \mintinline{clojure}{(inc x)} is safe. The main insight is we can assume the
@@
 or if they are both classes and the left is a subtype of the right (subclassing mode),
-otherwise returns \false{}, as described in figure~\ref{figure:isasemantics}.
+otherwise returns \false{}, as described in figure~\ref{main:figure:isasemantics}.
 
@@
 
-We encode this behaviour in figure~\ref{figure:isatyping}. 
+We encode this behaviour in figure~\ref{main:figure:isatyping}. 
 \isacompareliteral{} is the interesting part,
@@
   \caption{\isaliteral{} semantics}
-  \label{figure:isasemantics}
+  \label{main:figure:isasemantics}
 \end{figure}
@@
   \caption{\isaliteral{} typing rules}
-  \label{figure:isatyping}
+  \label{main:figure:isatyping}
 \end{figure}
@@
 \caption{Multimethod Syntax}
-\label{figure:mmsyntax}
+\label{main:figure:mmsyntax}
 \end{figure*}
@@
   \caption{Multimethod typing rules}
-  \label{figure:mmtyping}
+  \label{main:figure:mmtyping}
 \end{figure}
@@
   \caption{Multimethod subtyping rules}
-  \label{figure:mmsubtype}
+  \label{main:figure:mmsubtype}
 \end{figure}
@@
   \caption{Multimethod semantics}
-  \label{figure:mmsemantics}
+  \label{main:figure:mmsemantics}
 \end{figure}
@@
 To ease presentation, we present \emph{immutable}
-multimethods (figure~\ref{figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
+multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
 without changing the original multimethod. The above is now written
@@
 
-Multimethod semantics are in figure~\ref{figure:mmsemantics}.
+Multimethod semantics are in figure~\ref{main:figure:mmsemantics}.
 B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
@@
 
-The typing rules (figure~\ref{figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
+The typing rules (figure~\ref{main:figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
 where \s{} is the interface type for beta reduction, and \t{} is the type for
@@
 method bodies. T-App handles beta reduction because multimethods are subtypes of functions
-(figure~\ref{figure:mmsubtype}).
+(figure~\ref{main:figure:mmsubtype}).
 
@@
   \caption{HMap Terms and Types}
-  \label{figure:hmapsyntax}
+  \label{main:figure:hmapsyntax}
 \end{figure}
@@
   \caption{HMap Semantics}
-  \label{figure:hmapsem}
+  \label{main:figure:hmapsem}
 \end{figure}
@@
   \caption{HMap Typing Rules}
-  \label{figure:hmaptyping}
+  \label{main:figure:hmaptyping}
 \end{figure}
@@
   \caption{HMap subtyping rules}
-  \label{figure:hmapsubtype}
+  \label{main:figure:hmapsubtype}
 \end{figure}
@@
 objects with a known set of keyword entries. We use \HMapliteral{}
-types (figure~\ref{figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.
+types (figure~\ref{main:figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.
 
@@
 
-Figure~\ref{figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
+Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
 entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
@@
 
-Typing rules are given in figure~\ref{figure:hmaptyping}. T-GetHMap handles the simple
+Typing rules are given in figure~\ref{main:figure:hmaptyping}. T-GetHMap handles the simple
 case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
@@
 
-S-HMap (figure~\ref{figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
+S-HMap (figure~\ref{main:figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
 subtypes and also cover the absent keys.
@@
   \caption{Java Interoperability Terms and Types}
-  \label{figure:reflectivesyntax}
+  \label{main:figure:reflectivesyntax}
 \end{figure*}
@@
 We extend the external language with Java field lookups and calls to
-methods and constructors (figure~\ref{figure:nonreflectivesyntax}).
+methods and constructors (figure~\ref{main:figure:nonreflectivesyntax}).
 There are no typing or evaluation rules for these syntaxes because
@@
 ie., are unambiguously resolved at compile-time. Programmers use
-type hinted expressions (figure~\ref{figure:nonreflectivesyntax}) 
+type hinted expressions (figure~\ref{main:figure:nonreflectivesyntax}) 
 to help the compiler eliminate reflection.
@@
   \caption{Internal Language Reflection Resolution Extensions}
-  \label{figure:nonreflectivesyntax}
+  \label{main:figure:nonreflectivesyntax}
 \end{figure*}
@@
   \end{mathpar}
-\caption{Reflection Elimination}
-\label{figure:rewrite}
+\caption{Reflection Elimination (select rules, figure~\ref{appendix:figure:rewrite} for full rules)}
+\label{main:figure:rewrite}
 \end{figure*}
@@
   \end{mathpar}
-\caption{Type Hint Inference}
-\label{figure:hintinfer}
+\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
+\label{main:figure:hintinfer}
 \end{figure*}
@@
   \caption{Java Interoperability Typing Rules}
-  \label{figure:javatyping}
+  \label{main:figure:javatyping}
 \end{figure}
@@
 reflection, however to tell the full story we include a rewrite relation 
-in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
-Figure~\ref{figure:hintinfer} shows the interesting cases in type hint inference.
+in figure~\ref{main:figure:rewrite} to eliminate reflection when sufficient type hints are provided.
+Figure~\ref{main:figure:hintinfer} shows the interesting cases in type hint inference.
 
 We now have two languages: an internal language that may contain non-reflective Java calls (figure 
-  \ref{figure:nonreflectivesyntax}) 
+  \ref{main:figure:nonreflectivesyntax}) 
   and an external language that resembles Clojure, where we must rely on type-hints to resolve
-  reflective Java calls (figure~\ref{figure:reflectivesyntax}).
+  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).
 
@@
 We then extract these type hints in T-FieldStatic for typing checking 
-(figure~\ref{figure:javatyping}), which converts type hints
+(figure~\ref{main:figure:javatyping}), which converts type hints
 with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
@@
 \input{update}
-\input{convert-types-figure}
+
+{\convertjavatypefigure{figure}}

### 2015-02-21 formal-model.tex
add map value metavariable. compact rules

@@
 Types include the top type, untagged unions, functions, singleton types
-and class instances. We abbreviate \EmptyUnion{} as \Bot{}, \Value{\Nil{}} as \Nil{} and similarly
+and class instances. We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{} and similarly
 for \True and \False.
@@
                       \alt {\letexp {\x{}} {\e{}} {\e{}}}
-                      \alt {\errorvalv{}}
+                      %\alt {\errorvalv{}}
                 &\mbox{Expressions} \\
-  \v{} &::=&          \true{} \alt \false{} \alt \trdiff{\nil{}}
-                      \alt \trdiff{\class{}}
-                      \alt \k{}
+  \v{} &::=&          \singletonmeta{}
                       \alt {\const{}}
@@
     {\TDo}
-
-    {\TClass}
-
+    %{\TClass}
     {\TIf}
-
     {\TLet}
 
-    {\TKw}
+    %{\TKw}
 
@@
 
-    {\TError}
+    %{\TError}
 
@@
     \BDefMulti{}
+    \BBetaMulti{}
 
     \BDefMethod{}
-
-    \BBetaMulti{}
   \end{mathpar}
@@
     {\BAssoc}
-
     {\BGet}
-
     {\BGetMissing}
@@
 To simplify presentation, we only allow the empty map literal and
-restrict lookup and extension to keyword keys. The runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}
-is not part of the language syntax.
+restrict lookup and extension to keyword keys. The metavariable \mapval{}
+ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.
 
@@
 
+    \BNew{}
+
     \BMethod{}
-
-    \BNew{}
   \end{mathpar}
@@
     {\TNewStatic}
-
     {\TFieldStatic}

### 2015-02-21 intro.tex

@@
 
-Programming is an error prone activity, especially in the presence of
-multiple languages working together. Over the past decade, it has
-become increasingly frequent to extend existing dynamically typed
-languages with optional type systems.  The guarantees that a sound
-type system provides can be the difference between choosing a language
-that supports a type system or one that does not.
+We synthesize these lessons to present \emph{Typed Clojure}, an
+optional type system for Clojure. Typed Clojure builds on the core
+type checking approach of Typed Racket, an existing gradual type
+system for Racket. However, Typed Clojure extends this basic framework
+in multiple ways to accommodate the unique idioms and features of
+Clojure, producing an expressive synthesis of ideas and demonstrating
+a surprising coincidence between the multiple dispatch approach of
+Clojure and the occurrence typing framework of Typed Racket.
 
-% presenting clojure
+The essence of Typed Clojure, of course, is Clojure, a dynamically
+typed language in the Lisp family built to run on the Java Virtual
+Machine (JVM) which has recently gained popularity as an alternative
+JVM language.  It offers the flexibility of a Lisp dialect, including
+macros, emphasizes a functional style via a
+standard library of immutable data structures, and provides
+interoperability with existing Java code, allowing programmers to use
+existing Java libraries without leaving Clojure.
+%
+Since its initial release in 2007, Clojure has been widely adopted for
+``backend'' development in places where its support for paralleism,
+functional programming, and Lisp-influenced abstraction is desired on
+the JVM. As a result, it now has an extensive base of existing untyped
+programs, many of whose developers can benefit from Typed Clojure. As
+a result, Typed Clojure is now used in industry, experience we discuss
+in this paper.
 
-Clojure is a dynamically typed language built to run on the Java Virtual Machine (JVM)
-which has recently gained popularity as an alternative JVM language.
-It offers the flexibility of a Lisp dialect, including Common Lisp-style macros,
-emphasising a functional style via a standard library of immutable datastructures. 
-Clojure has good interoperability with existing Java code, allowing programmers to
-use existing Java libaries without leaving Clojure.
-
-%problem?
-
-Clojure
-
-Typed Clojure is a gradual type system for Clojure, based on the
-static portion of Typed Racket.
-
+\begin{figure}
 \begin{minted}[]{clojure}
@@
 \end{minted}
+\caption{A simple Typed Clojure program}
+\label{fig:ex1}
+\end{figure}
+Figure~\ref{fig:ex1} presents a simple program demonstrating several
+aspects of our system, from simple type annotations to explicit
+handling of Java's \texttt{null} in interoperation, as well as an
+extended form of occurrence typing and Clojure's \emph{type hints},
+which are central to Typed Clojure's approach to interoperability.
+
+In the remainder of this paper, we describe Typed Clojure's central
+innovations, including Java interoperability, multimethods, and
+heterogenously-typed immutable maps, following an example-driven
+presentation in Section~\ref{sec:overview}. We then incrementatlly
+present a core calculus for Typed Clojure covering all of these
+features together in Section~\ref{sec:formal} and prove type
+soundness(section \ref{sec:sound}). We then discuss the full
+implementation of Typed Clojure, dubbed \texttt{core.typed}, which
+extends the formal model in many ways, and the experience gained from
+its use in Section~\ref{sec:exp}. Finally, we discuss related work and
+conclude.
 
@@
 
-\begin{minted}[]{clojure}
-(defalias Expr
-  (U '{:op :if :test Expr :then Expr :else Expr}
-     '{:op :const :val Any} ...))
-(ann tree-str [Expr -> String])
-(defmulti tree-str (fn [e :- Expr] (get :op e)))
-(defmethod tree-str :if [{:keys [test then else]}] 
-  (str (tree-str test) (tree-str then) (tree-str else)))
-(defmethod tree-str :const [{:keys [val]}] 
-  (str val))
-\end{minted}
+%% \begin{minted}[]{clojure}
+%% (defalias Expr
+%%   (U '{:op :if :test Expr :then Expr :else Expr}
+%%      '{:op :const :val Any} ...))
+%% (ann tree-str [Expr -> String])
+%% (defmulti tree-str (fn [e :- Expr] (get :op e)))
+%% (defmethod tree-str :if [{:keys [test then else]}] 
+%%   (str (tree-str test) (tree-str then) (tree-str else)))
+%% (defmethod tree-str :const [{:keys [val]}] 
+%%   (str val))
+%% \end{minted}
 
-\subsection{Contributions}
+%% \subsection{Contributions}
 
-%The main idea is....... SPJ
-In this paper we present a practical type system for Clojure.
+%% %The main idea is....... SPJ
+%% In this paper we present a practical type system for Clojure.
 
-\begin{itemize}
+%% \begin{itemize}
 
-  \item We formalise Typed Clojure and prove that typed code
-    \begin{itemize}
-      \item disallows null-pointer exceptions, (Section ?)
-      \item statically resolves Java interoperability (Section ?)
-    \end{itemize}
-  \item We reveal the surprising synthesis between Clojure multimethods
-        and occurrence typing.
-  \item We 
-\end{itemize}
+%%   \item We formalise Typed Clojure and prove that typed code
+%%     \begin{itemize}
+%%       \item disallows null-pointer exceptions, (Section ?)
+%%       \item statically resolves Java interoperability (Section ?)
+%%     \end{itemize}
+%%   \item We reveal the surprising synthesis between Clojure multimethods
+%%         and occurrence typing.
+%%   \item We 
+%% \end{itemize}
 

### 2015-02-21 intro.tex
add macro for core.typed

@@
 soundness(section \ref{sec:sound}). We then discuss the full
-implementation of Typed Clojure, dubbed \texttt{core.typed}, which
+implementation of Typed Clojure, dubbed \coretyped{}, which
 extends the formal model in many ways, and the experience gained from

### 2015-02-21 intro.tex
add opening example to tests

@@
 \begin{figure}
-\begin{minted}[]{clojure}
-(ann parent [(U nil File) -> (U nil File)])
-(defn parent [^File f]
-  (when f
-    (let [s (.getParent f)]
-      (assert s "Expected parent string, got nil")
-      (new File s))))
-\end{minted}
+\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/parent.clj}
 \caption{A simple Typed Clojure program}
@@
 \end{figure}
+
 Figure~\ref{fig:ex1} presents a simple program demonstrating several

### 2015-02-23 formal-model.tex
compact MM rules
let-aliasing rules
polish do/isa in section 3

@@
 application, abstractions, conditionals, do (sequencing) expressions and let expressions.
-We model throwable errors without a catching mechanism. All binding forms introduce fresh
-variables.
+All binding forms introduce fresh variables.
 
-Values include booleans, \nil{}, class literals, keywords, constants and closures.
-\false{} and \nil{} are the only logically false values.
+Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \mintinline{clojure}{:a}), 
+constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.
 
@@
 
-Typing rules are in figure~\ref{main:figure:typingrules}.
-T-Error neither returns logically true or false as it does not return normally.
-T-Do uses the propositions learnt in the first expression to type check the second.
-Other than compensating for the extra logically false value \nil{}, the rest are standard.
-
-Figure~\ref{main:figure:standardopsem} contains the operational semantics, which are standard
-except for \nil{} being false.
+Figures ~\ref{main:figure:typingrules} and~\ref{main:figure:standardopsem}
+contain the typing rules and operational semantics.
+They compensate for the extra logically false value \nil{}, otherwise they are standard.
 
@@
                       \alt {\ifexp {\e{}} {\e{}} {\e{}}}
-                      \alt {\trdiff{\doexp {\e{}} {\e{}}}}
+                      %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
                       \alt {\letexp {\x{}} {\e{}} {\e{}}}
@@
                 &\mbox{Values} \\
-  {\const{}}           &::=& \classconst \alt \throwconst
-
-                &\mbox{Constants}\\
+                \constantssyntax{}\\
   \s{}, \t{}    &::=& \Top 
@@
 
-    {\BDo}
+    %{\BDo}
 
@@
 
-\subsection{\doliteral{}}
+\subsection{Reasoning about Exceptional Control Flow}
 
-Assertions in Clojure often expand to simple exceptional control flow reasoning.
+Clojure assertions expand to simple exceptional control flow.
 
 \begin{minted}{clojure}
-(fn [x :- (U nil Number)] {:pre (number? x)} (inc x))
+(fn [x :- (U nil Number)] 
+  (assert (number? x))
+  (inc x))
 \end{minted}
@@
 (fn [x :- (U nil Number)]
-  (do (if x (throw (new Exception)) nil)
+  (do (if (number? x) nil (throw (new Exception)))
       (inc x)))
@@
 
-With an elegant extension in T-Do (figure~\ref{main:figure:typingrules}) we can follow this reasoning
-to infer \mintinline{clojure}{(inc x)} is safe. The main insight is we can assume the
-disjunction of the propositions of the first \doliteral{} argument while checking the second.
+We extend our model with sequencing expressions and errors, where {\errorvalv{}}
+is equivalent to a Clojure throw expression.
 
-To sketch how this example is checked, first T-Abs type checks the body under the 
+\smallskip
+$
+\begin{altgrammar}
+  \e{} &::=& \ldots \alt {\errorvalv{}} \alt {\doexp {\e{}} {\e{}}} &\mbox{Expressions} 
+\end{altgrammar}
+$
+
+\smallskip
+
+B-Do simply evaluates its arguments sequentially and returns the right argument.
+Since errors are not values, we define error propagation semantics
+like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).
+
+\begin{mathpar}
+    {\BDo}
+
+\infer [BE-Error]
+{}
+{ \opsem {\openv{}} 
+         {\errorvalv{}}
+         {\errorvalv{}}}
+
+\infer [BE-Do1]
+{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
+{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
+\end{mathpar}
+
+Our main insight is as follows: 
+if the left argument reduces to a value, then it is either logically true or false.
+Occurrence typing gives exactly the propositions for both cases, so their disjunction
+can be assumed for the right argument.
+T-Do formalises this intuition.
+
+\begin{mathpar}
+    {\TDo}
+
+    {\TError}
+\end{mathpar}
+
+The introduction of errors, specifically as an expression that is neither true nor false, 
+makes our insight interesting.
+For example, in checking \mintinline{clojure}{(do (throw (Exception.)) (inc :a))}
+by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \mintinline{clojure}{(inc :a)},
+which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
+which allows us to assume \mintinline{clojure}{(inc :a)} is unreachable (and it clearly is by BE-Do1).
+
+This works similarly with our motivating example. Intuitively, when evaluating
+\mintinline{clojure}{(inc x)} we know \mintinline{clojure}{(assert (number? x))} has evaluated to a value,
+so we assume \mintinline{clojure}{x} is a number.
+In detail, the example is checked first by T-Abs, which checks the body with the 
 environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
-T-Do then extracts the propositions of its first argument:
-T-If tells us \thenprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} and 
-\elseprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
-using T-Error, T-Nil and T-Subsume.
+Since the body is a \doliteral{},
+T-Do extracts the propositions of the left argument, an if:
+T-If says 
+\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
+by T-Nil and T-Subsume and
+\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
 Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
 and we check \mintinline{clojure}{(inc x)} in the extended environment 
-\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}
-and by L-Update we know {\isprop{\Number}{\x{}}}
-so T-Local says \mintinline{clojure}{x} is a \Number.
+\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
+By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
+so T-Local checks \mintinline{clojure}{x} as a \Number.
 
-\subsection{\isaliteral{}}
+\subsection{Multimethod dispatch mechanism: \isaliteral}
 
-One interesting Clojure function we modeled is \isaliteral.
-It takes two arguments, and returns \true{} if they are the same (equality mode)
-or if they are both classes and the left is a subtype of the right (subclassing mode),
-otherwise returns \false{}, as described in figure~\ref{main:figure:isasemantics}.
+Before tackling multimethods we model their core dispatch mechanism
+\isaliteral{}, a subclassing test for classes, otherwise an equality test.
+
+\begin{minted}{clojure}
+(isa? Keyword Object) ;=> true
+(isa? Object Keyword) ;=> false
+(isa? :a :a) ;=> true
+(isa? :a :b) ;=> false
+(isa? :a Keyword) ;=> false
+(isa? (class :a) Keyword) ;=> true
+\end{minted}
+
+B-IsA with the metafunction \isaopsemliteral{} give the operational semantics.
+
+\smallskip
+$
+\begin{array}{ll}
+  \vcenter{\hbox{\BIsA{}}}
+  &
+  \vcenter{\hbox{\isaopsemfigure{}}}
+\end{array}
+$
+\smallskip
+
+For the typing rules, we add rules for type checking keyword and class literals
+as singleton types, and define T-IsA, which infers the propositions of an \isaliteral{} 
+expression by passing
+ the \emph{object} of the left argument and the \emph{type}
+of the right argument to \isacompareliteral{}.
+
+\smallskip
+$
+\begin{array}{ll}
+\vcenter{\hbox{
+    \TIsA{}}} &
+
+  \vcenter{\hbox{$
+    \begin{array}{l}
+      {\TKw}\\\\
+      {\TClass}
+    \end{array}$}}
+\end{array}
+$
+\smallskip
+
+\begin{figure*}
+  \begin{mathpar}
+    \isapropsfigure{}
+  \end{mathpar}
+  \caption{\isaopsemliteral Proposition set Metafunction}
+\end{figure*}
+
+\begin{mathpar}
+    {\TKw}
+
+    {\TClass}
+
+    \isapropsfigure{}
+
+    \TIsA{}
+\end{mathpar}
 
@@
 \isaliteral{} serves as the default dispatch mechanism for multimethods in Clojure.
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \isaopsemfigure{}
-
-    \BIsA{}
-  \end{mathpar}
-  \caption{\isaliteral{} semantics}
-  \label{main:figure:isasemantics}
-\end{figure}
 
@@
   \begin{mathpar}
+    \BDefMethod{}
     \BDefMulti{}
     \BBetaMulti{}
-
-    \BDefMethod{}
   \end{mathpar}
-  \caption{Multimethod semantics}
+  \caption{Multimethod and method resolution semantics}
   \label{main:figure:mmsemantics}
@@
     {\TNewStatic}
+
     {\TFieldStatic}
-
     {\TMethodStatic}
@@
   \label{main:figure:javatyping}
-\end{figure}
+\end{figure*}
 
@@
 
+{\footnotesize
 \input{constant}
 \input{update}
+}
 
-{\convertjavatypefigure{figure}}
+{\convertjavatypefigure{figure}{\footnotesize}}

### 2015-02-23 intro.tex
use hmaps and let-aliasing in intro example

@@
 \begin{figure}
-\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/parent.clj}
+\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/parent2.clj}
 \caption{A simple Typed Clojure program}
@@
 
-Figure~\ref{fig:ex1} presents a simple program demonstrating several
+Figure~\ref{fig:ex1} presents a simple program demonstrating many
 aspects of our system, from simple type annotations to explicit
@@
 
-In the remainder of this paper, we describe Typed Clojure's central
-innovations, including Java interoperability, multimethods, and
-heterogenously-typed immutable maps, following an example-driven
-presentation in Section~\ref{sec:overview}. We then incrementatlly
+The \clj{parent} function has the type 
+$$
+\clj{['{:file (U nil File)} -> (U nil Str)]}
+$$
+which means that it takes a hash table whose \clj{:file} key maps to either
+\clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
+\clj{String}. The \clj{parent} function uses the \clj{:file} keyword
+as an accessor to get the file, checks that it isn't \clj{nil}, and
+then obtains the parent by making a Java method call.
+%
+The annotation \clj{^File} is a type hint, which instructs the Clojure
+compiler (running prior to Typed Clojure typechecking) to statically
+resolve method calls to the relevant object using the Java \clj{File}
+class, rather than using reflective method calls at runtime.
+
+In the remainder of this paper, we describe how Typed Clojure's
+central innovations, including Java interoperability, multimethods,
+and heterogenously-typed immutable maps, enable this example and many
+others. We begin with an example-driven presentation of the main type
+system features in Section~\ref{sec:overview}. We then incrementatlly
 present a core calculus for Typed Clojure covering all of these
@@
 soundness(section \ref{sec:sound}). We then discuss the full
-implementation of Typed Clojure, dubbed \coretyped{}, which
-extends the formal model in many ways, and the experience gained from
-its use in Section~\ref{sec:exp}. Finally, we discuss related work and
+implementation of Typed Clojure, dubbed \coretyped{}, which extends
+the formal model in many ways, and the experience gained from its use
+in Section~\ref{sec:exp}. Finally, we discuss related work and
 conclude.

### 2015-02-23 formal-model.tex
'update' covers all cases for ClassPE

@@
 
-Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \mintinline{clojure}{:a}), 
+Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \clj{:a}), 
 constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.
@@
 makes our insight interesting.
-For example, in checking \mintinline{clojure}{(do (throw (Exception.)) (inc :a))}
-by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \mintinline{clojure}{(inc :a)},
+For example, in checking \clj{(do (throw (Exception.)) (inc :a))}
+by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \clj{(inc :a)},
 which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
-which allows us to assume \mintinline{clojure}{(inc :a)} is unreachable (and it clearly is by BE-Do1).
+which allows us to assume \clj{(inc :a)} is unreachable (and it clearly is by BE-Do1).
 
 This works similarly with our motivating example. Intuitively, when evaluating
-\mintinline{clojure}{(inc x)} we know \mintinline{clojure}{(assert (number? x))} has evaluated to a value,
-so we assume \mintinline{clojure}{x} is a number.
+\clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
+so we assume \clj{x} is a number.
 In detail, the example is checked first by T-Abs, which checks the body with the 
@@
 Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
-and we check \mintinline{clojure}{(inc x)} in the extended environment 
+and we check \clj{(inc x)} in the extended environment 
 \ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
 By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
-so T-Local checks \mintinline{clojure}{x} as a \Number.
+so T-Local checks \clj{x} as a \Number.
 
@@
 
-B-IsA with the metafunction \isaopsemliteral{} give the operational semantics.
+B-IsA with the metafunction \isaopsemliteral{} reflect these semantics.
 
-\smallskip
-$
+$$
 \begin{array}{ll}
@@
 \end{array}
-$
-\smallskip
+$$
 
-For the typing rules, we add rules for type checking keyword and class literals
-as singleton types, and define T-IsA, which infers the propositions of an \isaliteral{} 
-expression by passing
- the \emph{object} of the left argument and the \emph{type}
-of the right argument to \isacompareliteral{}.
+For the typing rules, we add T-Kw and T-Class for type checking keyword and class literals
+as singleton types, and T-IsA for \isaliteral{} expressions, which 
+uses the \emph{object} of the left argument and the \emph{type} of the right argument
+to infer the final proposition set.
 
-\smallskip
-$
+$$
 \begin{array}{ll}
@@
 \end{array}
-$
-\smallskip
+$$
 
-\begin{figure*}
+The definition of \isacompareliteral{} (figure~\ref{main:figure:isaprops}) is deliberately conservative.
+The first line considers the case where the object of the left argument
+is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
+The simplest example
+
+\begin{minted}{clojure}
+  (isa? (class x) Keyword)
+\end{minted}
+
+has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
+that is if the expression is true then \clj{x} is a keyword, otherwise \clj{x} is not a keyword.
+
+We can also finally define \nil{} test
+
+\begin{minted}{clojure}
+(ann nil? (Pred nil))
+(defn nil? [x]
+  (isa? (class (class x)) Class))
+\end{minted}
+
+with the proposition set 
+\isacompare{\s{}}{\classpe{}({\path{\classpe{}}{\x{}}})}{\Value{\Class}}{\filterset{\isprop{\Class}{\path{\classpe{}}{\x{}}}}{\notprop{\Class}{\path{\classpe{}}{\x{}}}}},
+
+\begin{figure}
+  \footnotesize
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{\isaopsemliteral Proposition set Metafunction}
-\end{figure*}
 
-\begin{mathpar}
-    {\TKw}
-
-    {\TClass}
-
-    \isapropsfigure{}
-
-    \TIsA{}
-\end{mathpar}
+  \center{Takes \emph{object} of the left parameter and \emph{type} of the right parameter.}
+  \caption{Proposition Set for \isaliteral{} expressions}
+  \label{main:figure:isaprops}
+\end{figure}
 
@@
 \isaliteral{} serves as the default dispatch mechanism for multimethods in Clojure.
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \isapropsfigure{}
-
-    \TIsA{}
-  \end{mathpar}
-  \caption{\isaliteral{} typing rules}
-  \label{main:figure:isatyping}
-\end{figure}
 
@@
 A multimethod in Clojure is a function with a dispatch mechanism to
-of internal methods. Multimethods are created with {\mintinline{clojure}{defmulti}}
-and methods are installed via {\mintinline{clojure}{defmethod}}.
+of internal methods. Multimethods are created with {\clj{defmulti}}
+and methods are installed via {\clj{defmethod}}.
 They are called like regular functions.
@@
 
-%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\mintinline{clojure}{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
+%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\clj{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
 %                       {\classconst{}}}

### 2015-02-23 formal-model.tex
isa example

@@
 
-We can also finally define \nil{} test
+We can use \isaliteral{} as a crude \nil{} test
 
-\begin{minted}{clojure}
-(ann nil? (Pred nil))
-(defn nil? [x]
-  (isa? (class (class x)) Class))
-\end{minted}
+\inputminted[firstline=5]{clojure}{code/demo/src/demo/isa_nil.clj}
 
-with the proposition set 
-\isacompare{\s{}}{\classpe{}({\path{\classpe{}}{\x{}}})}{\Value{\Class}}{\filterset{\isprop{\Class}{\path{\classpe{}}{\x{}}}}{\notprop{\Class}{\path{\classpe{}}{\x{}}}}},
+because the then proposition is {\isprop{\Class}{\path{\classpe{}}{\x{}}}}, so
+we know {\isprop{\Object}{\x{}}} by L-Update, and combined with the fact {\isprop{\Union{\Nil{}}{\Number}}{\x{}}}
+we can infer {\isprop{\Number}{\x{}}} by L-Update.
 
@@
 
-For example, {\isaapp{\Keyword}{\Object}} = \true{} but 
-{\isaapp{\Object}{\Keyword}} = \false{}.
-This is useful in combination with \classconst{}
-with a fixed second argument, where {\isaapp{\appexp{\classconst}{\x{}}}{\Keyword}}
-is a test for \x{} being a \Keyword{}, and also
-as an equality test like
-{\isaapp{\x{}}{\makekw{a}}}.
+The second \isacompareliteral{} line handles the equality mode of \isaliteral{}
+with a non-class singleton type as the second argument.
 
-We encode this behaviour in figure~\ref{main:figure:isatyping}. 
-\isacompareliteral{} is the interesting part,
-which takes the object of the first argument 
-and the type of the second argument, and returns the then and else propositions
-of the expression.
-
-The first case handles subclassing mode
-in expressions like {\isaapp{\appexp{\classconst}{\x{}}}{\Object}},
-where the first argument has an object ending in \classpe{} and the second argument has a singleton class type.
-For example, in this case \isacomparetwo{\path{\classpe{}}{\x{}}}{\Value{\Object}}
-gives a then proposition {\isprop{\Object} {\x{}}}
-and an else proposition {\notprop{\Object}{\x{}}}.
-
-The reason to peel off one path element is subtle. For example,
-claiming {\isprop{\Value{\Object}} {\path{\classpe{}}{\x{}}}}
-might seem suitable, however it implies \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Object}
-which is wrong if \x{} = \makekw{a} where \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Keyword}.
-(Formulating \isacompareliteral revealed a three year old bug in the implementation).
-
-The second case handles the equality mode of \isaliteral{}
-with a non-class singleton type as the second argument, to be certain we are not in subclassing mode.
 Substitution is used to ensure well formed propositions.
@@
 The final case gives up; we cannot distinguish between equality and subclassing mode.
-
-\isaliteral{} serves as the default dispatch mechanism for multimethods in Clojure.
 

### 2015-02-23 formal-model.tex
formatting, remove/inline figures

@@
 
-\begin{figure*}
+\begin{figure}
   \footnotesize
 $$
-\begin{altgrammar}
+\begin{array}{lrlr}
   \expd{}, \e{} &::=& \x{}
@@
                       \alt {\comb {\e{}} {\e{}}} 
-                      \alt {\abs {\x{}} {\t{}} {\e{}}}
-                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
+                      \alt {\abs {\x{}} {\t{}} {\e{}}} &\mbox{Expressions} \\
+                      &\alt& {\ifexp {\e{}} {\e{}} {\e{}}}
                       %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
-                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
+                      \alt {\letexp {\x{}} {\e{}} {\e{}}}\\
                       %\alt {\errorvalv{}}
-                &\mbox{Expressions} \\
   \v{} &::=&          \singletonmeta{}
@@
                                    {\object{}}}
-                      \alt {\Value \singletonmeta{}} 
-                      \alt \trdiff{\class{}}
                 &\mbox{Types} \\
+                      &\alt& {\Value \singletonmeta{}} 
+                      \alt \trdiff{\class{}}\\
   \singletonallsyntax{}
@@
   %\classliteralallsyntax{}
-\end{altgrammar}
+\end{array}
 $$
@@
 \label{main:figure:termsyntax}
-\end{figure*}
+\end{figure}
 
@@
 
-    {\TDo}
+    %{\TDo}
     %{\TClass}
@@
 
-\begin{figure*}
+\begin{figure}
   \footnotesize
   \begin{mathpar}
-    {\BLocal}
-
-    %{\BDo}
-
-    {\BLet}
-
-    \BVal{}
-
+%    {\BLocal}
+%
+%    %{\BDo}
+%
+%    {\BLet}
+%
+%    \BVal{}
+%
     \BIfTrue{}
 
-    \BIfFalse{}
-
-    \BAbs{}
-
-    \BBetaClosure{}
-
-    \BDelta{}
+%    \BIfFalse{}
+%
+%    \BAbs{}
+%
+%    \BBetaClosure{}
+%
+%    \BDelta{}
   \end{mathpar}
@@
   \label{main:figure:standardopsem}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
+\begin{figure}
   \footnotesize
   \begin{mathpar}
-    \RAbs{}
-
-    \RNewElimRefl{}
-
-    \RMethodElimRefl{}
-
+    %\RAbs{}
+%
+%    \RNewElimRefl{}
+%
+%    \RMethodElimRefl{}
+%
     \RFieldElimRefl{}
-
-    \RLet{}
-
+%
+%    \RLet{}
+%
     \RLetHint{}
@@
 \label{main:figure:rewrite}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}
-  \footnotesize
-  \begin{mathpar}
-    \TALocal{}
-
-    \TANil{}
-
-    \TANewStatic{}
-
-    \TALetHint{}
-
-    \TALet{}
-  \end{mathpar}
-\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
-\label{main:figure:hintinfer}
-\end{figure*}
+%\begin{figure}
+%  \footnotesize
+%  \begin{mathpar}
+%%    \TALocal{}
+%%
+%%    \TANil{}
+%%
+%%    \TANewStatic{}
+%%
+%%    \TALetHint{}
+%
+%%    \TALet{}
+%  \end{mathpar}
+%\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
+%\label{main:figure:hintinfer}
+%\end{figure}
 
@@
 reflection, however to tell the full story we include a rewrite relation
-in figure~\ref{main:figure:rewrite} to eliminate reflection when sufficient type hints are provided.
-Figure~\ref{main:figure:hintinfer} shows the interesting cases in type hint inference.
+
+$$
+\rewrite {\taenv{}} {\e{}} {\ep{}}
+$$
+
+to eliminate reflection when sufficient type hints are provided.
+
+To help infer type hints
+
+$$
+\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
+$$
+
+gives the type hint \tatype{} of expression \e{} in environemtn \taenv{}.
+
+$$
+\TALetHint{}
+$$
 
@@
   reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).
+
+  $$
+    \RFieldElimRefl{}
+  $$
 
@@
 \input{constant}
-\input{update}
+
+\begin{figure*}
+  $$
+\begin{array}{lllll}
+\updatefigure
+\end{array}
+$$
+\caption{Type Update}
+\end{figure*}
 }
@@
 {\convertjavatypefigure{figure}{\footnotesize}}
+
+\begin{figure}
+  $$
+\begin{array}{lllll}
+  \restrictremovefigure{}
+\end{array}
+  $$
+  \caption{Restrict and Remove}
+\end{figure}

### 2015-02-23 overview.tex
unit tests
add appendix
formatting, remove/inline figures
section 2

@@
 
-Typed Clojure is a gradual type system for Clojure. It is designed
-to type check normal Clojure code by adding annotations. It is implemented
-as a library, and can be seamlessly included in any Clojure project; no
-separate compiler or language is needed.
+\subsection{Clojure}
 
-{\smallsection {Based on Typed Racket}}
-Initially, the similarities between untyped Racket and Clojure and Typed Racket's 
-ability to type check Racket code led us to investigate a similar type system for Clojure.
-After two years of development, the solid basis of Typed Racket 
-helps us type check many Clojure idioms without significant differences
-in implementation or theory. We found that extending Typed Clojure to check
-those idioms that have no obvious Racket equivalent did not significantly alter the structure
-of the type system.
+\subsection{Typed Racket and occurrence typing}
 
-{\smallsection {Occurrence typing}}
-\citet{TF08,TF10} developed \emph{occurrence typing}, which helps improve types at branches.
-Typed Clojure uses occurrence typing in a similar way to Typed Racket, with
-some extensions (discussed in in Section [?]). %FIXME
+Classic occurrence typing concentrates on conditional
+control flow like the following example.
 
-{\smallsection {Practical Variable-Arity Polymorphism}}
-Functions with non-trivial variable parameters are common in Racket.
-For example, Racket provides \emph{map} which takes a function and a
-variable number of collections and applies the function simultaneously
-to each element of the provided collections, returning a list of results.
-\citet*{STF09} developed a practical system that handles advanced variable parameters
-which can handle applications of functions like \emph{map}.
+\begin{listing}
+\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
+\caption{Occurrence typing}
+\end{listing}
 
-Clojure has a similar emphasis on variable-arity functions. In some ways,
-Clojure's core library encourages even more complicated variable-parameter schemes.
-The \emph{assoc} core function, for example, takes three parameters and
-then a quantity of variable parameters that is a multiple of two.
-This is beyond what Typed Racket (and Typed Clojure) can currently handle.
+Here we include the full namespace, which we omit in the following
+examples. Going down the then branch in this example means that we know
+\clj{x} is a number, which is automatically updated with occurrence typing.
 
-Functions like \emph{map} are common in Clojure, so we provide an implementation
-of variable-arity polymorphism which has similar capabilities as Typed Racket's
-implementation.
+In Clojure, exceptional control flow is almost as common, often in the form of
+runtime assertions.
 
-{\smallsection {Local Type Inference}}
-We use Pierce and Turner's Local Type Inference~\cite{PT00} to infer some polymorphic
-applications. Our implementation is based on Typed Racket's, which has extensions
-to support applications of polymorphic variable-arity functions like \emph{map}.
+\begin{listing}
+\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/do.clj}
+\caption{Exceptional control flow}
+\end{listing}
 
-{\smallsection {Unions and intersections}}
-Like Typed Racket, we include union and ordered intersection types. Unions define
-a least-upper-bound for a set of types. For example, we can express a type that is
-either \Number or \Symbol by including them in a union: {\Union {\Number} {\Symbol}}
+Clojure assertions expand to simple exceptional control flow,
+which we enhance our sequencing \doliteral{} expression to understand.
 
-Ordered intersections (described further by \citet{SA+12})
-are used for overloading function types. We can express a function that takes
-a \Number and returns a \Symbol, and vice-versa with an ordered intersection function type:
+\begin{minted}{clojure}
+(fn [x :- (U nil Number)]
+  (do (if (number? x) nil (throw (new Exception)))
+      (inc x)))
+\end{minted}
 
-\begin{lstlisting}[label=lst:ordered]
-(Fn [Number -> Symbol]
-    [Symbol -> Number])
-\end{lstlisting}
+Our main insight is that occurrence typing already provides
+valuable information about the left argument,
+namely if it returns a value, \clj{x} is a \clj{Number}.
+Since the right argument is evaluated after the left, we can assume
+\clj{x} is a \clj{Number} when evaluating \clj{(inc x)}, verifying
+the programmer's attempt to avoid null-pointer exceptions without further annotations.
 
-As our intersections are \emph{ordered}, we can express fine invariants in the
-case where arity parameter types overlap. Similar to a pattern match, earlier arities 
-are tried first, and the first arity to match ``wins''.
+\subsection{Heterogeneous hash-maps}
 
-For example, applying an \lstinline|Integer| argument to a function of type
+Hash-maps with keyword keys play a major role in Clojure programming.
 
-\begin{lstlisting}
-(Fn [Integer -> Integer]
-    [Number -> Number])
-\end{lstlisting}
+HMaps track the presense and absense of entries.
 
-returns an \lstinline|Integer|. Reversing the arities however gives
-type \lstinline|Number|, because the arity taking a \lstinline|Number|
-always matches first.
+$$
+\clj{(HMap :mandatory {:a Num})} 
+$$
 
-{\smallsection {Hosted on the Java Virtual Machine}}
-Clojure is built to run on the Java Virtual Machine (JVM),
-offering good interoperability with existing Java code.
-Typed Clojure helps programmers correctly call Java code
-by integrating with Java's type system.
+is the type of a map with at least an \clj{:a} entry mapped to a number,
+also written \clj{'{:a Num}}. If \clj{:b} is known to be an absent entry
+then
 
-We give Java arrays and Java's \emph{null} special treatment
-when involved with interoperability. Arrays are treated as \emph{read-only}
-when sourced from Java methods, discussed in Section \ref{sec:arrays}.
-We are explicit, and conservative by default, in the positions where
-Java's \emph{null} can be passed, discussed in Section \ref{sec:null}.
+% '#' not allowed in mintinline
+\smallskip
+\begin{minted}{clojure}
+  (HMap :mandatory {:a Num} :absent-keys #{:b})
+\end{minted}
 
-\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg1.clj}
+declares \clj{:b} absent. If the mandatory keys fully specify the
+entries, we write
+
+$$
+\clj{(HMap :mandatory {:a Num} :complete? true)}.
+$$
+
+\clj{:complete?} defaults to \clj{false}.
+The \clj{:optional} option
+
+$$
+\clj{(HMap :optional {:a Num})}
+$$
+
+abbreviates the union of present and absent entries
+
+\smallskip
+\begin{minted}{clojure}
+(U '{:a Num} (HMap :absent #{:a}))
+\end{minted}
+.
+
+We use \clj{defalias} to abbreviate HMap types.
+
+\begin{listing}
+\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
+\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
+\caption{A simple AST in terms of HMaps}
+\end{listing}
+
+HMaps are very malleable, including support for associate, merge and dissociate operations.
+
+\begin{listing}
+\inputminted[firstline=10,lastline=24]{clojure}{code/demo/src/demo/hmap.clj}
+\caption{HMap operations}
+\end{listing}
+
+Lookups on HMaps with multiple options combined with
+occurrence typing is an interesting problem.
+Testing a lookup on a mandatory entry is straightforward, since we
+know the lookup always succeeded.
+
+\begin{minted}{clojure}
+(fn [m :- '{:a (U nil Num)}] :- '{:a Num}
+  {:pre [(number? (:a m))]}
+  m)
+\end{minted}
+
+The result type is updated as expected to exclude \clj{nil}.
+Testing for \clj{nl} is more complicated, since it is the sentinel
+lookup failure indicator but may also appear as an entry value.
+If we test for \clj{nil} in an \clj{:optional} entry, the best we learn
+is that either the entry is missing or contains \clj{nil}.
+
+\begin{minted}{clojure}
+(fn [m :- (HMap :optional {:a (U nil Num)})] :- (HMap :optional {:a nil})
+  {:pre [(nil? (:a m))]}
+  m)
+\end{minted}
+
+\smallsection{Mandatory Entries}
+
+\smallsection{Known Absent Entries}
+
+\smallsection{Unknown Entries}
+
+\subsection{Java interoperability}
+
+Java
+
+\begin{listing}
+\inputminted[firstline=5,lastline=14]{clojure}{code/demo/src/demo/hmap.clj}
+\caption{Java Interoperability}
+\end{listing}
+
+\subsection{Multimethod}
+
+\subsection{Final example}
+
 \inputminted[firstline=4]{clojure}{code/demo/src/demo/eg2.clj}
@@
 \inputminted[firstline=4,lastline=15]{clojure}{code/demo/src/demo/eg6.clj}
-\inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+
+%% Old stuff vvvv
 
@@
 %  - implementation difference, AST walking vs syntax walking
+
+%Typed Clojure is a gradual type system for Clojure. It is designed
+%to type check normal Clojure code by adding annotations. It is implemented
+%as a library, and can be seamlessly included in any Clojure project; no
+%separate compiler or language is needed.
+%
+%{\smallsection {Based on Typed Racket}}
+%Initially, the similarities between untyped Racket and Clojure and Typed Racket's 
+%ability to type check Racket code led us to investigate a similar type system for Clojure.
+%After two years of development, the solid basis of Typed Racket 
+%helps us type check many Clojure idioms without significant differences
+%in implementation or theory. We found that extending Typed Clojure to check
+%those idioms that have no obvious Racket equivalent did not significantly alter the structure
+%of the type system.
+%
+%{\smallsection {Occurrence typing}}
+%\citet{TF08,TF10} developed \emph{occurrence typing}, which helps improve types at branches.
+%Typed Clojure uses occurrence typing in a similar way to Typed Racket, with
+%some extensions (discussed in in Section [?]). %FIXME
+%
+%{\smallsection {Practical Variable-Arity Polymorphism}}
+%Functions with non-trivial variable parameters are common in Racket.
+%For example, Racket provides \emph{map} which takes a function and a
+%variable number of collections and applies the function simultaneously
+%to each element of the provided collections, returning a list of results.
+%\citet*{STF09} developed a practical system that handles advanced variable parameters
+%which can handle applications of functions like \emph{map}.
+%
+%Clojure has a similar emphasis on variable-arity functions. In some ways,
+%Clojure's core library encourages even more complicated variable-parameter schemes.
+%The \emph{assoc} core function, for example, takes three parameters and
+%then a quantity of variable parameters that is a multiple of two.
+%This is beyond what Typed Racket (and Typed Clojure) can currently handle. 
+%
+%Functions like \emph{map} are common in Clojure, so we provide an implementation
+%of variable-arity polymorphism which has similar capabilities as Typed Racket's
+%implementation.
+%
+%{\smallsection {Local Type Inference}}
+%We use Pierce and Turner's Local Type Inference~\cite{PT00} to infer some polymorphic
+%applications. Our implementation is based on Typed Racket's, which has extensions
+%to support applications of polymorphic variable-arity functions like \emph{map}.
+%
+%{\smallsection {Unions and intersections}}
+%Like Typed Racket, we include union and ordered intersection types. Unions define
+%a least-upper-bound for a set of types. For example, we can express a type that is
+%either \Number or \Symbol by including them in a union: {\Union {\Number} {\Symbol}}
+%
+%Ordered intersections (described further by \citet{SA+12})
+%are used for overloading function types. We can express a function that takes
+%a \Number and returns a \Symbol, and vice-versa with an ordered intersection function type:
+%
+%\begin{lstlisting}[label=lst:ordered]
+%(Fn [Number -> Symbol]
+%    [Symbol -> Number])
+%\end{lstlisting}
+%
+%As our intersections are \emph{ordered}, we can express fine invariants in the
+%case where arity parameter types overlap. Similar to a pattern match, earlier arities 
+%are tried first, and the first arity to match ``wins''.
+%
+%For example, applying an \lstinline|Integer| argument to a function of type
+%
+%\begin{lstlisting}
+%(Fn [Integer -> Integer]
+%    [Number -> Number])
+%\end{lstlisting}
+%
+%returns an \lstinline|Integer|. Reversing the arities however gives
+%type \lstinline|Number|, because the arity taking a \lstinline|Number|
+%always matches first.
+%
+%{\smallsection {Hosted on the Java Virtual Machine}}
+%Clojure is built to run on the Java Virtual Machine (JVM),
+%offering good interoperability with existing Java code.
+%Typed Clojure helps programmers correctly call Java code
+%by integrating with Java's type system.
+%
+%We give Java arrays and Java's \emph{null} special treatment
+%when involved with interoperability. Arrays are treated as \emph{read-only}
+%when sourced from Java methods, discussed in Section \ref{sec:arrays}.
+%We are explicit, and conservative by default, in the positions where
+%Java's \emph{null} can be passed, discussed in Section \ref{sec:null}.

### 2015-02-23 formal-model.tex
java interop section 2

@@
 
-Clojure supports interoperability with Java, including the ability to
-call constructors, methods and access fields. We model
-these operations in a restricted setting without class inheritance,
+We model
+Java interoperability in a restricted setting without class inheritance,
 overloading or Java Generics.
@@
 
-Clojure supports \emph{type hints}, an optional type system to improve performance
-and disambiguate Java calls.
-Programmers communicate to the Clojure compiler via type hints
-to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
-and the programmer is free to add them where they like.
-
 core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved

### 2015-02-23 overview.tex
java interop section 2

@@
 
-Here we include the full namespace, which we omit in the following
+Here we include the full \clj{ns} form, which we omit in the following
 examples. Going down the then branch in this example means that we know
@@
 
-In Clojure, exceptional control flow is almost as common, often in the form of
+In Clojure, exceptional control flow is also common, often in the form of
 runtime assertions.
@@
 Our main insight is that occurrence typing already provides
-valuable information about the left argument,
+valuable information about the left argument of \doliteral{} expressions,
 namely if it returns a value, \clj{x} is a \clj{Number}.
@@
 is the type of a map with at least an \clj{:a} entry mapped to a number,
-also written \clj{'{:a Num}}. If \clj{:b} is known to be an absent entry
+also written \clj{'{:a Num}}. If \clj{:b} is also known to be an absent entry
 then
@@
 
-We use \clj{defalias} to abbreviate HMap types.
+\clj{defalias} defines a custom type abbreviation, often for unions of HMaps.
 
@@
 The result type is updated as expected to exclude \clj{nil}.
-Testing for \clj{nl} is more complicated, since it is the sentinel
+Testing for \clj{nil} is more complicated, since it is the sentinel
 lookup failure indicator but may also appear as an entry value.
@@
 \begin{minted}{clojure}
-(fn [m :- (HMap :optional {:a (U nil Num)})] :- (HMap :optional {:a nil})
+(fn [m :- (HMap :optional {:a (U nil Num)})] 
+  :- (HMap :optional {:a nil})
   {:pre [(nil? (:a m))]}
@@
 
-\smallsection{Mandatory Entries}
-
-\smallsection{Known Absent Entries}
-
-\smallsection{Unknown Entries}
-
 \subsection{Java interoperability}
 
-Java
+Clojure supports interoperability with Java, including the ability to
+call constructors, methods and access fields.
+
+Clojure supports \emph{type hints}, an optional type system to improve performance
+and disambiguate Java calls.
+Programmers communicate to the Clojure compiler via type hints
+to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
+and the programmer is free to add them where they like.
+
+Here we use \clj{ann} to annotate static types for the function, and the type hint
+\clj{^File} directs the Clojure compiler to assume \clj{f} is a \clj{File}
+when resolving Java calls at compile time.
 
 \begin{listing}
-\inputminted[firstline=5,lastline=14]{clojure}{code/demo/src/demo/hmap.clj}
+  \begin{minted}{clojure}
+(ann parent [(U nil File) -> (U nil Str)]
+(defn parent [^File f]
+  (when f
+    (.getParent f)))
+  \end{minted}
 \caption{Java Interoperability}
 \end{listing}
+
+While \clj{f} has the type hint \clj{File}, it also has the static type \clj{(U nil File)},
+and Typed Clojure disallows invoking \clj{nil} even if the compiler
+has statically resolved a method.
+Typed Clojure also conservatively assumes fields and method returns are nullable,
+and guarantees null-pointer exceptions cannot happen in typed code.
 

### 2015-02-23 formal-model.tex
final example

@@
 
-A multimethod in Clojure is a function with a dispatch mechanism to
-of internal methods. Multimethods are created with {\clj{defmulti}}
-and methods are installed via {\clj{defmethod}}.
-They are called like regular functions.
-
-\begin{minted}{clojure}
-(ann foo [Any -> String])
-(defmulti foo class)
-(defmethod foo Number [x] (str (inc x)))
-(defmethod foo Keyword [x] (str (name x)))
-(foo :a) ;=> "a"
-(foo 1)  ;=> "2"
-\end{minted}
-
 To ease presentation, we present \emph{immutable}

### 2015-02-23 overview.tex
final example

@@
 
+A multimethod in Clojure is a function with a dispatch mechanism to
+of internal methods. Multimethods are created with {\clj{defmulti}}
+and methods are installed via {\clj{defmethod}}.
+They are called like regular functions.
+
+\begin{minted}{clojure}
+(ann rep [Any -> String])
+(defmulti rep class)
+(defmethod rep Number [x] (str (inc x)))
+(defmethod rep Keyword [x] (str (name x)))
+
+(rep :a) ;=> "a"
+(rep 1)  ;=> "2"
+\end{minted}
+
+The \clj{class} function returns the class of an object, or \clj{nil}
+if given \clj{nil}. This multimethod's dispatch function returns the class of its
+argument. The underlying dispatch strategy features \clj{isa?}, a function
+that is a subclassing check when provided with classes
+
+\begin{minted}{clojure}
+  (isa? Keyword Object) ;=> true
+  (isa? Keyword Number) ;=> false
+\end{minted}
+
+otherwise an equality check.
+
+\begin{minted}{clojure}
+  (isa? :a :a) ;=> true
+  (isa? :a 1) ;=> false
+\end{minted}
+
+So in the previous example, \clj{(rep :a)} dispatched to the method
+where \clj{(isa? (class :a) v)} returned true, where \clj{v} is the dispatch value.
+Since \clj{(isa? (class :a) Keyword)} is true, the second method is chosen.
+
+The flexibility of \clj{isa?} is key to the generality of multimethods. 
+For instance, we can dispatch on the value of
+a map entry, even if it is not a class.
+
+\inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
+
+\clj{isa?} is further special-case with vectors: with two vectors of the
+same length, it returns the logical conjunction of the pairwise applications
+of \clj{isa?}.
+
+\begin{minted}{clojure}
+  (isa? [Keyword :a] [Object :a]) ;=> true
+  (isa? [Keyword Keyword] [Object Object]) ;=> true
+\end{minted}
+
+We can use this to dispatch on more than one thing. Here we perform
+double dispatch on the class of each argument.
+
+\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+
+No extra annotations are needed to follow the type-directed control flow.
+For instance, the middle method is has inferred the left argument is a number
+and thus safe to increment, and the right argument is a keyword and safe to
+extract its name.
+
+A dispatch value can also be the keyword \clj{:default} to specify a default
+method if no preferred method is found. The static type of both arguments
+remains unchanged.
+
 \subsection{Final example}
 
-\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg2.clj}
-\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg3.clj}
-\inputminted[firstline=4]{clojure}{code/demo/src/demo/eg4.clj}
-\inputminted[firstline=4,lastline=15]{clojure}{code/demo/src/demo/eg5.clj}
-\inputminted[firstline=4,lastline=15]{clojure}{code/demo/src/demo/eg6.clj}
-\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+The final example combines everything we will cover for the rest of the paper:
+multimethod dispatch, reflection resolution via type hints, Java method
+and constructor calls, conditional and exceptional flow reasoning
+and use of HMaps.
+
+\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
 

### 2015-02-24 intro.tex
Improvements to start of 2.

@@
 Python, and GradualTalk among gradually-typed system.\footnote{We
-  reserve the term ``gradual typing'' for systems which soundly
-  interoperate between typed and untyped code; systems which do not
+  reserve the term ``gradual typing'' for systems such as Typed Racket which soundly
+  interoperate between typed and untyped code; systems like Typed Clojure or TypeScript which do not
   enforce type invariants we describe as ``optionally typed''.}

### 2015-02-24 overview.tex
Improvements to start of 2.

@@
 
+\label{sec:overview}
+
+We now begin a tour of the central features of Typed Clojure,
+beginning with Clojure itself. In our presentation, we will make 
+use of the full Typed Clojure system to illustrate the key type system
+ideas, before studying the key features in detail in
+section~\ref{sec:formal}.
+
 \subsection{Clojure}
+
+Clojure~\cite{something-for-clojure} is a Lisp built to run on the
+Java Virtual Machine with exemplary support for concurrent programming
+and immutable data structures. It emphaisizes mostly-functional
+programming, restricting imperative updates to a limited set of
+structures which have specific thread synchronization behavior. By
+default, it provides fast implemenations of immutable lists, vectors,
+and hash tables, which are used for most data structures, although it
+also provides means for defining new records.
+
+One of Clojure's primary advantages is easy interoperation with
+existing Java libraries. It automatically generates appropriate JVM
+bytecode to make Java method and constructor calls, and treats Java
+values as any other Clojure value. However, this smooth
+interoperability comes at the cost of pervasive \clj{null}, which
+leads to the possibility of null pointer exceptions---a drawback we
+address in Typed Clojure.
 

### 2015-02-24 formal-model.tex
experience

@@
 
-Figures ~\ref{main:figure:typingrules} and~\ref{main:figure:standardopsem}
-contain the typing rules and operational semantics.
-They compensate for the extra logically false value \nil{}, otherwise they are standard.
+Figure ~\ref{main:figure:typingrules} contains the standard typing rules.
+The operational semantics are standard, except \nil{} is logically false:
+
+\begin{mathpar}
+    \BIfTrue{}
+
+    \BIfFalse{}
+\end{mathpar}
 
@@
 $$
-\begin{array}{lrlr}
+\begin{array}{lrll}
   \expd{}, \e{} &::=& \x{}
@@
   \occurrencetypingsyntax{}\\
+  \propenvsyntax{}\\
   \openvsyntax{}
@@
 
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
+%\begin{figure}
+%  \footnotesize
+%  \begin{mathpar}
 %    {\BLocal}
@@
 %
-    \BIfTrue{}
+    %\BIfTrue{}
 
@@
 %    \BDelta{}
-  \end{mathpar}
-  \caption{Operational Semantics}
-  \label{main:figure:standardopsem}
-\end{figure}
+%  \end{mathpar}
+%  \caption{Operational Semantics}
+%  \label{main:figure:standardopsem}
+%\end{figure}
 
@@
 We extend our model with sequencing expressions and errors, where {\errorvalv{}}
-is equivalent to a Clojure throw expression.
+is equivalent to Clojure's throw.
 
@@
 so T-Local checks \clj{x} as a \Number.
+
+\subsection{Heterogenous maps}
+
+\begin{figure}
+  \footnotesize
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
+    &\mbox{Expressions} \\
+    \v{} &::=& \ldots \alt {\emptymap{}}
+    &\mbox{Values} \\
+    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
+    &\mbox{Types} \\
+    \auxhmapsyntax{}\\
+    \pesyntax{}   &::=& \ldots \alt {\keype{\k{}}}
+                  &\mbox{Path Elements}
+  \end{altgrammar}
+  $$
+  \caption{HMap Terms and Types}
+  \label{main:figure:hmapsyntax}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    {\BAssoc}
+    {\BGet}
+    {\BGetMissing}
+  \end{mathpar}
+  \caption{HMap Semantics}
+  \label{main:figure:hmapsem}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    {\TGetHMap}
+
+    {\TGetAbsent}
+
+    {\TGetHMapPartialDefault}
+
+    {\TAssoc}
+  \end{mathpar}
+  \caption{HMap Typing Rules}
+  \label{main:figure:hmaptyping}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \HMapsubtyping{}
+  \end{mathpar}
+  \caption{HMap subtyping rules}
+  \label{main:figure:hmapsubtype}
+\end{figure}
+
+We now present \HMapliteral{} types in detail.
+Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
+\clj{(HMap :mandatory p :absent-keys a)} is written \HMapgeneric{\mandatory{}}{\absent{}}.
+The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
+specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
+entries and we abbreviate \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}}. Keys cannot be both present and absent.
+
+\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
+To simplify presentation, we only provide syntax for the empty map literal and
+restrict lookup and extension to keyword keys. The metavariable \mapval{}
+ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.
+
+We introduce our first path element {\keype{\k{}}}, the symbolic representation
+of a \getliteral{} expression on keyword \k{} where the sentinel value for
+a missing entry is \nil{}.
+
+Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps. B-Get returns the associated
+value if the entry is present otherwise  B-GetMissing returns \nil{}. B-Assoc returns a new extended
+map.
+
+Typing rules are given in figure~\ref{main:figure:hmaptyping}. T-GetHMap handles the simple
+case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
+a \Nil{} return type.
+T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
+T-Assoc extends the known keyword entries, overwriting previous entries with the same key.
+
+S-HMap (figure~\ref{main:figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
+subtypes and also cover the absent keys.
+Complete maps are subtypes of partial maps by S-HMapP.
+
+\subsection{Java Interoperability and Type Hints}
+
+We model
+Java interoperability in a restricted setting without class inheritance,
+overloading or Java Generics.
+
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
+            &\alt& \reflectiveexpsyntax{} 
+  \end{altgrammar}
+  $$
+
+We extend the external language with Java field lookups and calls to
+methods and constructors.
+These come without typing or evaluation rules:
+Typed Clojure requires the Clojure compiler to statically resolve all Java interactions
+in typed code.
+Type hinted expressions such as \localhintsyntax{} 
+to help the compiler eliminate reflection.
+
+\begin{figure*}
+  \footnotesize
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
+    &\mbox{Expressions}\\
+
+    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
+    &\mbox{Values} \\
+
+    \tatypesyntax{}\\
+    \typehintenvsyntax{}\\
+    \classtableallsyntax{}
+  \end{altgrammar}
+  $$
+  \caption{Internal Language Reflection Resolution Extensions}
+  \label{main:figure:nonreflectivesyntax}
+\end{figure*}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    %\RAbs{}
+%
+%    \RNewElimRefl{}
+%
+%    \RMethodElimRefl{}
+%
+    \RFieldElimRefl{}
+%
+%    \RLet{}
+%
+
+    \RLetHint{}
+  \end{mathpar}
+\caption{Reflection Elimination %(select rules, figure~\ref{appendix:figure:rewrite} for full rules)
+}
+\label{main:figure:rewrite}
+\end{figure}
+
+%\begin{figure}
+%  \footnotesize
+%  \begin{mathpar}
+%%    \TALocal{}
+%%
+%%    \TANil{}
+%%
+%%    \TANewStatic{}
+%%
+%%    \TALetHint{}
+%
+%%    \TALet{}
+%  \end{mathpar}
+%\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
+%\label{main:figure:hintinfer}
+%\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \BField{}
+
+    \BNew{}
+
+    \BMethod{}
+  \end{mathpar}
+  \caption{Java Interoperability Semantics}
+\end{figure}
+
+\begin{figure*}
+  \footnotesize
+  \begin{mathpar}
+    {\TNewStatic}
+
+    {\TFieldStatic}
+    {\TMethodStatic}
+  \end{mathpar}
+  \caption{Java Interoperability Typing Rules}
+  \label{main:figure:javatyping}
+\end{figure*}
+
+$$
+\rewrite {\taenv{}} {\e{}} {\ep{}}
+$$
+
+rewrites expression \e{} to a possibly-less reflective expression
+\ep{} with respect to type hint environment
+\taenv{} and Java class table \ct{}.
+
+$$
+\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
+$$
+
+infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
+
+%$$
+%\TALetHint{}
+%$$
+
+We now have two languages: an internal language that may contain non-reflective Java calls (figure 
+  \ref{main:figure:nonreflectivesyntax}) 
+  and an external language that resembles Clojure, where we must rely on type-hints to resolve
+  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).
+
+  $$
+    \RFieldElimRefl{}
+  $$
+
+For example, R-FieldElimRefl, which rewrites a reflective field call to be non-reflective,
+rewrites the target expression and uses the result to infer a class type hint as the field target.
+\fieldtypeliteral{} uses the inferred type hint to lookup if the inferred class has the appropriate
+field name and returns the actual target and result type hints which we store in the rewritten rule.
+
+We then extract these type hints in T-FieldStatic for typing checking 
+(figure~\ref{main:figure:javatyping}), which converts type hints
+with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
+dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).
+
+Method and constructor calls are similar, but T-MethodStatic and T-NewStatic disallow passing \nil{} as arguments, 
+we assume methods can return \nil{} and we assume constructors cannot.
+
+We rewrite to our internal language before type checking and remaining reflective calls are type errors.
+
+\subsection{\classconst{}}
+
+The \classconst{} function is modeled as a constant which returns
+an object's \Class{}, or \nil{} if provided \nil{}.
+
+\begin{mathpar}
+\constanttypefigure{}
+\end{mathpar}
+
+Our second path element \classpe{} is the symbolic representation
+of a call to \classconst{}.
+
+$$
+\begin{array}{lrlr}
+  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
+                &\mbox{Path Elements}
+\end{array}
+$$
 
@@
 
-\subsection{Heterogenous maps}
-
-\begin{figure}
-  \footnotesize
-  $$
-  \begin{altgrammar}
-    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
-    &\mbox{Expressions} \\
-    \v{} &::=& \ldots \alt {\emptymap{}}
-    &\mbox{Values} \\
-    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
-    &\mbox{Types} \\
-    \auxhmapsyntax{}\\
-  \end{altgrammar}
-  $$
-  \caption{HMap Terms and Types}
-  \label{main:figure:hmapsyntax}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    {\BAssoc}
-    {\BGet}
-    {\BGetMissing}
-  \end{mathpar}
-  \caption{HMap Semantics}
-  \label{main:figure:hmapsem}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    {\TGetHMap}
-
-    {\TGetAbsent}
-
-    {\TGetHMapPartialDefault}
-
-    {\TAssoc}
-  \end{mathpar}
-  \caption{HMap Typing Rules}
-  \label{main:figure:hmaptyping}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \HMapsubtyping{}
-  \end{mathpar}
-  \caption{HMap subtyping rules}
-  \label{main:figure:hmapsubtype}
-\end{figure}
-
-It is common in Clojure to use plain immutable hash maps to represent ad-hoc
-objects with a known set of keyword entries. We use \HMapliteral{}
-types (figure~\ref{main:figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.
-
-The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
-specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
-entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
-HMap can assume all entries not in {\mandatory{}} as absent. We disallow keys to be both present and absent.
-
-\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
-To simplify presentation, we only allow the empty map literal and
-restrict lookup and extension to keyword keys. The metavariable \mapval{}
-ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.
-
-Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
-entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
-map.
-
-Typing rules are given in figure~\ref{main:figure:hmaptyping}. T-GetHMap handles the simple
-case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
-a \Nil{} return type.
-T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
-T-Assoc extends the known keyword entries, overwriting previous entries with the same key.
-
-S-HMap (figure~\ref{main:figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
-subtypes and also cover the absent keys.
-Complete maps are subtypes of partial maps by S-HMapP.
-
-\subsection{Java Interoperability and Type Hints}
-
-\begin{figure*}
-  \footnotesize
-  $$
-  \begin{altgrammar}
-    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
-            \alt  \localhintsyntax{} \alt \lethintsyntax{}
-    &\mbox{Expressions}
-  \end{altgrammar}
-  $$
-  \caption{Java Interoperability Terms and Types}
-  \label{main:figure:reflectivesyntax}
-\end{figure*}
-
-We model
-Java interoperability in a restricted setting without class inheritance,
-overloading or Java Generics.
-
-We extend the external language with Java field lookups and calls to
-methods and constructors (figure~\ref{main:figure:nonreflectivesyntax}).
-There are no typing or evaluation rules for these syntaxes because
-Typed Clojure must be convinced that Java interactions are \emph{non-reflective},
-ie., are unambiguously resolved at compile-time. Programmers use
-type hinted expressions (figure~\ref{main:figure:nonreflectivesyntax}) 
-to help the compiler eliminate reflection.
-
-\begin{figure*}
-  \footnotesize
-  $$
-  \begin{altgrammar}
-    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
-    &\mbox{Expressions}\\
-
-    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
-    &\mbox{Values} \\
-
-    \tatypesyntax{}\\
-    \typehintenvsyntax{}\\
-    \classtableallsyntax{}
-  \end{altgrammar}
-  $$
-  \caption{Internal Language Reflection Resolution Extensions}
-  \label{main:figure:nonreflectivesyntax}
-\end{figure*}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    %\RAbs{}
-%
-%    \RNewElimRefl{}
-%
-%    \RMethodElimRefl{}
-%
-    \RFieldElimRefl{}
-%
-%    \RLet{}
-%
-    \RLetHint{}
-  \end{mathpar}
-\caption{Reflection Elimination (select rules, figure~\ref{appendix:figure:rewrite} for full rules)}
-\label{main:figure:rewrite}
-\end{figure}
-
-%\begin{figure}
-%  \footnotesize
-%  \begin{mathpar}
-%%    \TALocal{}
-%%
-%%    \TANil{}
-%%
-%%    \TANewStatic{}
-%%
-%%    \TALetHint{}
-%
-%%    \TALet{}
-%  \end{mathpar}
-%\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
-%\label{main:figure:hintinfer}
-%\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \BField{}
-
-    \BNew{}
-
-    \BMethod{}
-  \end{mathpar}
-  \caption{Java Interoperability Semantics}
-\end{figure}
-
-\begin{figure*}
-  \footnotesize
-  \begin{mathpar}
-    {\TNewStatic}
-
-    {\TFieldStatic}
-    {\TMethodStatic}
-  \end{mathpar}
-  \caption{Java Interoperability Typing Rules}
-  \label{main:figure:javatyping}
-\end{figure*}
-
-core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
-reflection, however to tell the full story we include a rewrite relation
-
-$$
-\rewrite {\taenv{}} {\e{}} {\ep{}}
-$$
-
-to eliminate reflection when sufficient type hints are provided.
-
-To help infer type hints
-
-$$
-\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
-$$
-
-gives the type hint \tatype{} of expression \e{} in environemtn \taenv{}.
-
-$$
-\TALetHint{}
-$$
-
-We now have two languages: an internal language that may contain non-reflective Java calls (figure 
-  \ref{main:figure:nonreflectivesyntax}) 
-  and an external language that resembles Clojure, where we must rely on type-hints to resolve
-  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).
-
-  $$
-    \RFieldElimRefl{}
-  $$
-
-For example, R-FieldElimRefl, which rewrites a reflective field call to be non-reflective,
-rewrites the target expression and uses the result to infer a class type hint as the field target.
-\fieldtypeliteral{} uses the inferred type hint to lookup if the inferred class has the appropriate
-field name and returns the actual target and result type hints which we store in the rewritten rule.
-
-We then extract these type hints in T-FieldStatic for typing checking 
-(figure~\ref{main:figure:javatyping}), which converts type hints
-with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
-dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).
-
-Method and constructor calls are similar, but T-MethodStatic and T-NewStatic disallow passing \nil{} as arguments, 
-we assume methods can return \nil{} and we assume constructors cannot.
-
-We rewrite to our internal language before type checking and remaining reflective calls are type errors.
-
 {\footnotesize
-\input{constant}
+\constantsemfigure{}
 

### 2015-02-24 formal-model.tex
experience intro

@@
 T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
+Even though their types vary significantly, we still use the same object.
 T-Assoc extends the known keyword entries, overwriting previous entries with the same key.
@@
 Complete maps are subtypes of partial maps by S-HMapP.
+
+\subsection{Proof system}
+
+The essential piece of occurrence typing's logical system is L-Update.
+(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
+
+\begin{mathpar}
+{\LUpdate}
+\end{mathpar}
+
+This is where we combine propositions with similar objects
+to learn new facts.
+The full definition of \updateliteral{} is given
+in
+
+\begin{figure*}
+  $$
+\begin{array}{lllll}
+\updatefigure
+\end{array}
+$$
+\caption{Type Update}
+\end{figure*}
 
@@
 
-\begin{figure*}
-  $$
-\begin{array}{lllll}
-\updatefigure
-\end{array}
-$$
-\caption{Type Update}
-\end{figure*}
-}
-
 {\convertjavatypefigure{figure}{\footnotesize}}

### 2015-02-24 formal-model.tex
don't footnote size the last 3 pages!

@@
 
-The essential piece of occurrence typing's logical system is L-Update.
+The essential piece of occurrence typing's logical system is L-Update,
+where we combine propositions that have similar objects to learn new facts.
 (The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
@@
 
-This is where we combine propositions with similar objects
-to learn new facts.
-The full definition of \updateliteral{} is given
-in
+For example, if \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
+and
+\inpropenv {\propenv{}} {\notprop {\Nil{}} {\x{}}},
+then
+$$
+\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype{\Nil{}}} {\emptypath{}}} {\x{}}},
+$$
+so we learn \inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).
+
+Now that we have our path elements \classpe{} and \keype{\k{}}
+and the typings rules that introduce them, we look at \updateliteral{} 
+(figure~\ref{main:figure:update}) and give one interesting example for each.
+
+\smallsection{\classpe{}}
+If \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
+and
+\inpropenv {\propenv{}} {\isprop {\Value{\Keyword}} {\path{\classpe{}}{\x{}}}}
+
+\smallsection{\keype{\k{}}}
+
+\smallsection{Combination}
 
@@
 \caption{Type Update}
+\label{main:figure:update}
 \end{figure*}
@@
 {\convertjavatypefigure{figure}{\footnotesize}}
+}
 
@@
   \caption{Restrict and Remove}
+  \label{main:figure:restrictremove}
 \end{figure}

### 2015-02-24 formal-model.tex
classpe example

@@
 \smallsection{\classpe{}}
-If \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
-and
-\inpropenv {\propenv{}} {\isprop {\Value{\Keyword}} {\path{\classpe{}}{\x{}}}}
+Consider the then branch of the following program:
+
+\begin{minted}{clojure}
+(fn [x :- (U nil Keyword)]
+  (when (class x)
+    ; here
+    ))
+\end{minted}
+
+Intuitively, since \clj{(class x)} is a true value it must be a class, therefore
+\clj{x} is an object (not \clj{nil}).
+
+From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
+The test has the object {\path{\classpe{}}{\x{}}} by T-App.
+By T-If we can assume
+{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
+Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
+{\path{\classpe{}}{\x{}}}, 
+by T-Update 
+$$
+\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
+$$
+
+This hits the sixth \classpe{} case of \updateliteral{}, which returns
+$$
+{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
+$$
+deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
+by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).
 

### 2015-02-24 formal-model.tex
section on update

@@
 
+\subsection{\classconst{}}
+
+The \classconst{} function is modeled as a constant which returns
+an object's \Class{}, or \nil{} if provided \nil{}.
+
+\begin{mathpar}
+\constanttypefigure{}
+\end{mathpar}
+
+Our second path element \classpe{} is the symbolic representation
+of a call to \classconst{}.
+
+$$
+\begin{array}{lrlr}
+  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
+                &\mbox{Path Elements}
+\end{array}
+$$
+
 \subsection{Proof system}
@@
 \smallsection{\keype{\k{}}}
+Consider the then branch of
 
-\smallsection{Combination}
+\begin{minted}{clojure}
+(fn [x :- '{:kw (U nil Keyword)}]
+  (when (get x :kw)
+    ; here
+    ))
+\end{minted}
+
+Similar to before,
+by T-Abs
+{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\x{}}},
+the test's object is {\path{\keype{\makekw{kw}}}{\x{}}} by T-App
+and by T-If we assume
+{\notprop {\falsy{}} {\path{\keype{\makekw{kw}}}{\x{}}}} in the then branch.
+
+T-Update combines these propositions
+$$
+\inpropenv {\propenv{}} {\isprop {\update {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\nottype {\falsy{}}} {\keype{\makekw{kw}}}} {\x{}}}.
+$$
+matching the \updateliteral{} case which pushes the \updateliteral{} inside the map
+$$
+{\HMapp{\mandatorysetsingleton{\makekw{kw}}{\update{\Union{\Nil}{\Keyword}}{\nottype {\falsy{}}}{\emptypath{}}}}{\emptyabsent}}
+$$
+concluding with
+$$
+\inpropenv {\propenv{}}
+{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Keyword}}{\emptyabsent}}   {\x{}}}.
+$$
 
@@
 We rewrite to our internal language before type checking and remaining reflective calls are type errors.
-
-\subsection{\classconst{}}
-
-The \classconst{} function is modeled as a constant which returns
-an object's \Class{}, or \nil{} if provided \nil{}.
-
-\begin{mathpar}
-\constanttypefigure{}
-\end{mathpar}
-
-Our second path element \classpe{} is the symbolic representation
-of a call to \classconst{}.
-
-$$
-\begin{array}{lrlr}
-  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
-                &\mbox{Path Elements}
-\end{array}
-$$
 

### 2015-02-24 overview.tex
clojure reference

@@
 
-Clojure~\cite{something-for-clojure} is a Lisp built to run on the
+Clojure~\cite{Hic08} is a Lisp built to run on the
 Java Virtual Machine with exemplary support for concurrent programming

### 2015-02-24 formal-model.tex
todo
clojure reference
model

@@
 \section{A Formal Model of Typed Clojure}
+
+\label{sec:formal}
 
@@
 
-The propositions inform conditionals on what they can assume down the then and else
-branches if the expression is used as a test.
-The object is a possibly-empty symbolic representation of the expression.
+The propositions capture what can be assumed if an expression tests as logically true or false
+respectively. The object is a possibly-empty symbolic representation of the expression.
 
 Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
-application, abstractions, conditionals, do (sequencing) expressions and let expressions.
+application, abstractions, conditionals and let expressions.
 All binding forms introduce fresh variables.
 
-Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \clj{:a}), 
+Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \makekw{a}), 
 constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.
@@
 
-Propositions are identical to~\cite{TF10}; positive and negative type propositions, implications,
+Propositions are identical to~\cite{TF10}---positive and negative type propositions, implications,
 conjunctions, disjuctions, trivial and impossible propositions.
@@
 $$
+
+% TODO semantics
 

### 2015-02-25 overview.tex
fixes from my office
hmap examples

@@
 Java Virtual Machine with exemplary support for concurrent programming
-and immutable data structures. It emphaisizes mostly-functional
+and immutable data structures. It emphasizes mostly-functional
 programming, restricting imperative updates to a limited set of
-structures which have specific thread synchronization behavior. By
-default, it provides fast implemenations of immutable lists, vectors,
+structures which have specific thread synchronization behaviour. By
+default, it provides fast implementations of immutable lists, vectors,
 and hash tables, which are used for most data structures, although it
@@
 
-Classic occurrence typing concentrates on conditional
-control flow like the following example.
+Tobin-Hochstadt and Felleisen~\cite{TF10}
+introduced Typed Racket with occurrence typing,
+an inference technique for conditional control flow.
+They introduce the concept of occurrence typing 
+with the following example.
 
-\begin{listing}
+\inputminted[firstline=1]{racket}{code/tr/example1.rkt}
+
+This function takes a value that is either \emph{\#f} % mintinline really hates #
+or a number, represented by an untagged \emph{union type}.
+The `then' branch has an implicit invariant
+that \rkt{x} is a number, which is automatically inferred with occurrence typing
+and type checked without further annotations.
+Here is same program in Typed Clojure.
+
+\begin{exmp}
 \inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
-\caption{Occurrence typing}
-\end{listing}
+\end{exmp}
 
-Here we include the full \clj{ns} form, which we omit in the following
-examples. Going down the then branch in this example means that we know
-\clj{x} is a number, which is automatically updated with occurrence typing.
+Clojure uses \emph{namespaces}, declared with the
+\clj{ns} form, to manage top-level \emph{vars}.
+This is a regular Clojure file compiled with
+the Clojure compiler except it declares a runtime dependency on
+\clj{clojure.core.typed}, Typed Clojure's core namespace.
+Programmers import typed versions of certain constructs as needed, like
+a \clj{fn} variant that support annotations,
+and use the provided \clj{check-ns} function to type check the current namespace
+when convenient.
 
-In Clojure, exceptional control flow is also common, often in the form of
-runtime assertions.
+Typed Clojure can check all the examples in the occurrence typing
+paper---the rest of this section describes the extensions necessary
+to check Clojure code.
 
-\begin{listing}
-\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/do.clj}
-\caption{Exceptional control flow}
-\end{listing}
+\subsection{Exceptional control flow}
 
-Clojure assertions expand to simple exceptional control flow,
-which we enhance our sequencing \doliteral{} expression to understand.
+Along with conditional control flow,
+Clojure programmers rely on \emph{exceptions}
+to assert type-related invariants.
 
-\begin{minted}{clojure}
-(fn [x :- (U nil Number)]
-  (do (if (number? x) nil (throw (new Exception)))
-      (inc x)))
-\end{minted}
+\begin{exmp}
+\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
+\end{exmp}
 
-Our main insight is that occurrence typing already provides
-valuable information about the left argument of \doliteral{} expressions,
-namely if it returns a value, \clj{x} is a \clj{Number}.
-Since the right argument is evaluated after the left, we can assume
-\clj{x} is a \clj{Number} when evaluating \clj{(inc x)}, verifying
-the programmer's attempt to avoid null-pointer exceptions without further annotations.
+We combine the sequencing form \clj{do} and a thrown exception with
+\clj{throw} to avoid a possible null-pointer exception by incrementing \clj{nil}.
+(We omit \clj{ns} forms for the rest of the examples).
+
+To check this example, we note that
+occurrence typing already provides
+valuable information about the conditional expression---namely that it cannot return logically true and if it returns logically false 
+then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
+We use this assumption to check \clj{(inc x)}, automatically
+ruling out the possibility of null-pointer exceptions
+(said formally in section~\ref{sec:doformal}).
 
@@
 Hash-maps with keyword keys play a major role in Clojure programming.
+HMap types model the most common usages of keyword maps.
 
-HMaps track the presense and absense of entries.
+\begin{exmp}
+\begin{listing}
+\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
+\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
+\end{listing}
+\end{exmp}
 
+Here we define a type abbreviation with \clj{defalias} called \clj{Expr}
+that describes the structure of a recursively-defined AST as a union of HMaps.
+\clj{an-exp} is a function verified to return an \clj{Expr}.
+
+HMaps support various options to track the presence and absence of entries.
+\clj{'{:a Num}} is a shorthand for the \clj{:mandatory} option---entries that
+must be present:
 $$
-\clj{(HMap :mandatory {:a Num})} 
+\clj{(HMap :mandatory {:a Num})}.
 $$
 
-is the type of a map with at least an \clj{:a} entry mapped to a number,
-also written \clj{'{:a Num}}. If \clj{:b} is also known to be an absent entry
-then
-
+\clj{:absent-keys} is a set of keywords that are known to not have entries.
+The value \clj{{:a 1 :b nil}} is not a member of the type
 % '#' not allowed in mintinline
@@
 \begin{minted}{clojure}
-  (HMap :mandatory {:a Num} :absent-keys #{:b})
+  (HMap :mandatory {:a Num} :absent-keys #{:b}).
 \end{minted}
 
-declares \clj{:b} absent. If the mandatory keys fully specify the
-entries, we write
-
-$$
-\clj{(HMap :mandatory {:a Num} :complete? true)}.
-$$
-
-\clj{:complete?} defaults to \clj{false}.
-The \clj{:optional} option
-
+\clj{:optional} is a map that declares entries either absent or
+present with a specific type. Concretely
 $$
@@
 $$
-
-abbreviates the union of present and absent entries
-
-\smallskip
+is the same as
 \begin{minted}{clojure}
-(U '{:a Num} (HMap :absent #{:a}))
-\end{minted}
-.
-
-\clj{defalias} defines a custom type abbreviation, often for unions of HMaps.
-
-\begin{listing}
-\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
-\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
-\caption{A simple AST in terms of HMaps}
-\end{listing}
-
-HMaps are very malleable, including support for associate, merge and dissociate operations.
-
-\begin{listing}
-\inputminted[firstline=10,lastline=24]{clojure}{code/demo/src/demo/hmap.clj}
-\caption{HMap operations}
-\end{listing}
-
-Lookups on HMaps with multiple options combined with
-occurrence typing is an interesting problem.
-Testing a lookup on a mandatory entry is straightforward, since we
-know the lookup always succeeded.
-
-\begin{minted}{clojure}
-(fn [m :- '{:a (U nil Num)}] :- '{:a Num}
-  {:pre [(number? (:a m))]}
-  m)
+      (U '{:a Num} (HMap :absent-keys #{:a})).
 \end{minted}
 
-The result type is updated as expected to exclude \clj{nil}.
-Testing for \clj{nil} is more complicated, since it is the sentinel
-lookup failure indicator but may also appear as an entry value.
-If we test for \clj{nil} in an \clj{:optional} entry, the best we learn
-is that either the entry is missing or contains \clj{nil}.
+\clj{:complete?} is \clj{true} if there are no further entries
+than those declared \clj{:mandatory} or \clj{:optional}.
 
-\begin{minted}{clojure}
-(fn [m :- (HMap :optional {:a (U nil Num)})] 
-  :- (HMap :optional {:a nil})
-  {:pre [(nil? (:a m))]}
-  m)
-\end{minted}
+\clj{:mandatory}, \clj{:optional} and \clj{:absent-keys} default to empty
+and \clj{:complete?} defaults to \clj{false}.
+
+HMaps interact with occurrence typing in interesting ways.
+What do we learn about the map \clj{m} in the following test?
+
+\begin{exmp}
+\inputminted[firstline=5,lastline=9]{racket}{code/demo/src/demo/hmap_path.clj}
+\end{exmp}
+
+In the then branch Typed Clojure infers \clj{'{:a Num}}, in the else branch
+\clj{(HMap :optional {:a nil})}. Failed lookups return \clj{nil},
+so we cannot distinguish between \clj{'{:a nil}} and \clj{(HMap :optional {:a nil})}
+by testing the result of looking up \clj{:a}.
+Section~\ref{sec:hmapformal} discusses this in greater detail.
 

### 2015-02-25 small.tex
tweak intro example. Ack.
add appendix
improve intro
fixes from my office
hmap examples

@@
 \include{bibliography.bib}
+\usepackage{chngcntr}
+
+\newcounter{ex}
+\newenvironment{exmp}{
+\refstepcounter{ex}%
+\par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
+\\[-2.5em]\nopagebreak}
+{\relax}
 
@@
 \doi{nnnnnnn.nnnnnnn}
+
+\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
+\newcommand{\rkt}[1]{\mintinline{racket}{#1}}
 
@@
 system build on the design of Typed Racket, repurposing in particular
-\emph{occurence typing}, its approach to statically reasoning about
+\emph{occurrence typing}, its approach to statically reasoning about
 predicate tests. However, in adapting the type system to Clojure,
-changes and extensions are required to accomodate both additional
+changes and extensions are required to accommodate both additional
 language features and  idioms used by Clojure programmers.
@@
 operations, and their Clojure semantics turns out to have a surprising
-synergy with the underlying occurence typing framework.
+synergy with the underlying occurrence typing framework.
 
@@
 
+\input{acknowledgements}
+
 % We recommend abbrvnat bibliography style.
@@
 \bibliography{bibliography}
+
+\clearpage
+
+\input{appendix}
 

### 2015-02-25 intro.tex
2.5

@@
 aspects of our system, from simple type annotations to explicit
-handling of Java's \texttt{null} in interoperation, as well as an
+handling of Java's \java{null} in interoperation, as well as an
 extended form of occurrence typing and Clojure's \emph{type hints},

### 2015-02-25 overview.tex
2.5

@@
 values as any other Clojure value. However, this smooth
-interoperability comes at the cost of pervasive \clj{null}, which
+interoperability comes at the cost of pervasive \java{null}, which
 leads to the possibility of null pointer exceptions---a drawback we
@@
 
-Tobin-Hochstadt and Felleisen~\cite{TF10}
+\citet{TF10}
 introduced Typed Racket with occurrence typing,
@@
 Programmers import typed versions of certain constructs as needed, like
-a \clj{fn} variant that support annotations,
+a \clj{fn} variant that supports annotations,
 and use the provided \clj{check-ns} function to type check the current namespace
@@
 We combine the sequencing form \clj{do} and a thrown exception with
-\clj{throw} to avoid a possible null-pointer exception by incrementing \clj{nil}.
-(We omit \clj{ns} forms for the rest of the examples).
+\clj{throw} to avoid a null-pointer exception from attempting to increment \clj{nil}.
+(We omit \clj{ns} forms for the rest of the examples)
+\footnote{See <this> github repo for full working examples}.
 
@@
 HMaps interact with occurrence typing in interesting ways.
-What do we learn about the map \clj{m} in the following test?
+The next example branches on the result of a lookup, and the types
+inferred for \clj{f} are subtle.
 
 \begin{exmp}
+\begin{listing}
 \inputminted[firstline=5,lastline=9]{racket}{code/demo/src/demo/hmap_path.clj}
+\end{listing}
 \end{exmp}
@@
 call constructors, methods and access fields.
+Method calls use prefix syntax, for example this call to 
+\clj{java.io.File}'s \clj{getParent} method.
 
-Clojure supports \emph{type hints}, an optional type system to improve performance
-and disambiguate Java calls.
-Programmers communicate to the Clojure compiler via type hints
-to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
-and the programmer is free to add them where they like.
+\begin{minted}{clj}
+  (fn [f] (.getParent f))
+\end{minted}
 
-Here we use \clj{ann} to annotate static types for the function, and the type hint
-\clj{^File} directs the Clojure compiler to assume \clj{f} is a \clj{File}
-when resolving Java calls at compile time.
+We have a problem: Clojure's compiler does not know which method to call.
+Instead, \clj{f} is inspected at runtime and an appropriate method is selected based
+and then invoked by \emph{Java reflection}.
+Unfortunately, reflection is slow and Clojure's algorithm for
+choosing methods at runtime is undefined in some cases.
 
+To resolve reflective calls, Clojure supports \emph{type-hints}.
+
+\begin{minted}{clj}
+  (fn [^File f] (.getParent f))
+\end{minted}
+
+The Clojure compiler uses the type hint on \clj{f}
+to statically resolve the method call to the \clj{java.io.File}
+method with the Java signature
+
+\begin{minted}{java}
+  public String getParent()
+\end{minted}
+
+Typed Clojure ignores type-hints, so to check this we must
+annotate the parameter's static type.
+
+\begin{minted}{clj}
+  (fn [f :- File] (.getParent f))
+\end{minted}
+
+However Typed Clojure disallows reflection in typed code, so we
+must add a type hint.
+
+\inputminted[firstline=10,lastline=10]{clojure}{code/demo/src/demo/parent3.clj}
+
+Now we have a well-typed expression free of reflection.
+
+Typed Clojure and Java treat \java{null} differently.
+In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
+called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
+This means the Java static type \java{String} is equivalent to
+\clj{(U nil String)} in Typed Clojure.
+
+To guarantee \java{null} is never accidentally leaked into a Typed Clojure program,
+we must assume methods are nullable.
+
+\begin{exmp}
+\inputminted[firstline=12,lastline=13]{clojure}{code/demo/src/demo/parent3.clj}
+\end{exmp}
+
+In contrast, JVM invariants guarantee that  constructor cannot return \java{null},
+so we are safe to assume constructors are non-nullable.
+
+\begin{exmp}
 \begin{listing}
-  \begin{minted}{clojure}
-(ann parent [(U nil File) -> (U nil Str)]
-(defn parent [^File f]
-  (when f
-    (.getParent f)))
-  \end{minted}
-\caption{Java Interoperability}
+\inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
 \end{listing}
+\end{exmp}
 
-While \clj{f} has the type hint \clj{File}, it also has the static type \clj{(U nil File)},
-and Typed Clojure disallows invoking \clj{nil} even if the compiler
-has statically resolved a method.
-Typed Clojure also conservatively assumes fields and method returns are nullable,
-and guarantees null-pointer exceptions cannot happen in typed code.
+Notice a type hint is used in an argument position to help choose the \java{File(String pathname)}
+constructor. By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
+but can be configured globally for particular positions if needed.
+
+Finally, Typed Clojure guarantees typed code cannot throw a null-pointer exception
+by attempting to dereference \clj{nil}. This can happen in several ways, including
+calling a method or a field on \clj{nil}.
+The next example calls a method on once-nullable local binding.
+(\clj{ann} annotates a var with an expected type).
+
+\begin{exmp}
+\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
+\end{exmp}
+
+The test on \clj{f} is essential to check this example---only then can Typed Clojure
+prove null-pointer exceptions are impossible.
 

### 2015-02-25 overview.tex
hmap syntax by example

@@
 
+\begin{Code}
 \begin{exmp}
-\begin{listing}
-\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
+\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
 \inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
-\end{listing}
 \end{exmp}
+\end{Code}
 
@@
 \clj{'{:a Num}} is a shorthand for the \clj{:mandatory} option---entries that
-must be present:
-$$
-\clj{(HMap :mandatory {:a Num})}.
-$$
+must be present.
+
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=58,lastline=64]{clojure}{code/demo/src/demo/hmap.clj}
+\end{exmp}
+\end{Code}
+
+Looking up \clj{:a} simply returns the associated type.
+Keys other than \clj{:a} are unaccounted for in \clj{Mand}, so 
+the best we can infer is \clj{Any}, the supertype of all types.
 
 \clj{:absent-keys} is a set of keywords that are known to not have entries.
-The value \clj{{:a 1 :b nil}} is not a member of the type
-% '#' not allowed in mintinline
-\smallskip
-\begin{minted}{clojure}
-  (HMap :mandatory {:a Num} :absent-keys #{:b}).
-\end{minted}
+We take advantage of this to infer a better type for lookups on known absent keys.
+
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=66,lastline=73]{clojure}{code/demo/src/demo/hmap.clj}
+\end{exmp}
+\end{Code}
 
 \clj{:optional} is a map that declares entries either absent or
-present with a specific type. Concretely
-$$
-\clj{(HMap :optional {:a Num})}
-$$
+present with a specific type.
+
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=5,lastline=12]{clojure}{code/demo/src/demo/hmap_path.clj}
+\end{exmp}
+\end{Code}
+
+Concretely \clj{(HMap :optional {:a Num})}
 is the same as
@@
 
-\clj{:mandatory}, \clj{:optional} and \clj{:absent-keys} default to empty
-and \clj{:complete?} defaults to \clj{false}.
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=17,lastline=23]{clojure}{code/demo/src/demo/hmap_path.clj}
+\end{exmp}
+\end{Code}
+
+Typed Clojure knows that this \clj{merge} will preserve the entries
+of the left hand map.
+
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=25,lastline=25]{clojure}{code/demo/src/demo/hmap_path.clj}
+\end{exmp}
+\end{Code}
 
@@
 
+\begin{Code}
 \begin{exmp}
-\begin{listing}
-\inputminted[firstline=5,lastline=9]{racket}{code/demo/src/demo/hmap_path.clj}
-\end{listing}
+\inputminted[firstline=14,lastline=15]{clojure}{code/demo/src/demo/hmap_path.clj}
 \end{exmp}
+\end{Code}
 
@@
 
+\begin{Code}
 \begin{exmp}
-\begin{listing}
 \inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
-\end{listing}
 \end{exmp}
+\end{Code}
 

### 2015-02-25 conclusion.tex

\section{Conclusion}
\label{sec:conclusion}

We have presented Typed Clojure, an optionally-typed version of
Clojure whose type system works with a wide variety of distinctive
Clojure idioms and features. Although based on the foundation of Typed
Racket's occurrence typing approach, Typed Clojure both extends the
fundamental control-flow based reasoning as well as applying it to
handle seemingly unrelated features such as multi-methods. In
addition, Typed Clojure supports crucial features such as heterogenous
maps and Java interoperability while integrating these features into
the core type system.

The result is a sound, expressive, and useful type system which, when
implemented in \coretyped with appropriate extensions, suitable for
typechecking significant amount of existing Clojure programs.
%
As a result, Typed Clojure is already successful: it is widely used in
the Clojure community among both enthusiasts and professional
programmers and recieves contributions from many developers.

However, there is much more that Typed Clojure can provide. Most
significantly, Typed Clojure currently does not provide \emph{gradual
  typing}---ineraction between typed and untyped code is unchecked and
thus unsound. We hope to explore the possibilites of using existing
mechanisms for contracts and proxies in Java and
Clojure~\cite{some-stuff} to enable sound gradual typing for Clojure.

Additionally, the Clojure compiler is unable to use Typed Clojure's
wealth of static information to optimze programs. Addressing this
requires not only first enabling sound gradual typing, but also
integrating Typed Clojure into the Clojure tool chain more deeply, so
that its information can be passed on to the compiler.

Finally, our case study and broader experience indicate that Clojure
programmers still find themselves unable to use Typed Clojure on some
of their programs for lack of expressiveness. This requires continued
effort to analyze and understand the relevant features and idioms and
develop new type checking approaches for them.

### 2015-02-25 overview.tex
add circleci example

@@
 
-Typed Clojure knows that this \clj{merge} will preserve the entries
-of the left hand map.
+The next example is a real Typed Clojure program written by CircleCI, who maintain a large production
+installation of Clojure (see section~\ref{sec:casestudy} for a case study).
 
@@
 \begin{exmp}
-\inputminted[firstline=25,lastline=25]{clojure}{code/demo/src/demo/hmap_path.clj}
+\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
 \end{exmp}
 \end{Code}
+
+\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
+removing the raw private key and attaching an encrypted private key.
+Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
+does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
+operation forgot to remove it.
 
@@
 
+\begin{Code}
+\begin{exmp}
 \begin{minted}{clojure}
@@
 \end{minted}
+\end{exmp}
+\end{Code}
 
@@
 
+\begin{Code}
+\begin{exmp}
 \inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
+\end{exmp}
+\end{Code}
 
@@
 
+\begin{Code}
+\begin{exmp}
 \inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+\end{exmp}
+\end{Code}
 
@@
 
+\begin{Code}
+\begin{exmp}
 \inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
+\end{exmp}
+\end{Code}
 

### 2015-02-25 overview.tex
HMap overview

@@
 
-\subsection{Multimethod}
+\subsection{Multimethods}
 
-A multimethod in Clojure is a function with a dispatch mechanism to
-of internal methods. Multimethods are created with {\clj{defmulti}}
-and methods are installed via {\clj{defmethod}}.
-They are called like regular functions.
+A multimethod in Clojure is a function that contains a dispatch
+function and methods. Multimethods are created with {\clj{defmulti}}.
+\begin{minted}{clojure}
+(ann rep [Any -> String])
+(defmulti rep class)
+\end{minted}
+\clj{rep} is a multimethod of type \clj{[Any -> String]} with an empty \emph{dispatch table}
+and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
+There are no methods to dispatch to so invocations of \clj{rep} will fail at runtime.
+\begin{minted}{clojure}
+  (rep :a) ;=> IllegalArgumentException ...
+\end{minted}
+
+Methods are installed via {\clj{defmethod}}.
+\begin{minted}{clojure}
+(defmethod rep Keyword [x] (str (name x)))
+\end{minted}
+We extend \clj{rep}'s dispatch table, mapping
+the \emph{dispatch value} \clj{Number} to the function
+\clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
+uses the value of \clj{(isa? (disp-fn arg) disp-val)}
+on each method entry to pick a winner \clj{mth}, and finally returns \clj{(mth arg)}.
+\clj{isa?} is a subclassing check when provided with classes
+\begin{minted}{clojure}
+  (isa? Keyword Object) ;=> true
+  (isa? Keyword Number) ;=> false
+\end{minted}
+otherwise an equality check.
+\begin{minted}{clojure}
+  (isa? :a :a) ;=> true
+  (isa? :a 1) ;=> false
+\end{minted}
+
+For example
+\clj{(rep :a)}
+picks the \clj{Keyword} method because
+\clj{(isa? (class :a) Keyword)} returns true,
+and finally returns \clj{"a"}, the value of
+\clj{((fn [x] (str (name x))) :a)}.
+We give the full definition of \clj{rep}.
 
@@
 (defmulti rep class)
+(defmethod rep Keyword [x] (str (name x)))
 (defmethod rep Number [x] (str (inc x)))
-(defmethod rep Keyword [x] (str (name x)))
 
@@
 \end{Code}
-
-The \clj{class} function returns the class of an object, or \clj{nil}
-if given \clj{nil}. This multimethod's dispatch function returns the class of its
-argument. The underlying dispatch strategy features \clj{isa?}, a function
-that is a subclassing check when provided with classes
-
-\begin{minted}{clojure}
-  (isa? Keyword Object) ;=> true
-  (isa? Keyword Number) ;=> false
-\end{minted}
-
-otherwise an equality check.
-
-\begin{minted}{clojure}
-  (isa? :a :a) ;=> true
-  (isa? :a 1) ;=> false
-\end{minted}
-
-So in the previous example, \clj{(rep :a)} dispatched to the method
-where \clj{(isa? (class :a) v)} returned true, where \clj{v} is the dispatch value.
-Since \clj{(isa? (class :a) Keyword)} is true, the second method is chosen.
+Typed Clojure does not statically ensure multimethod calls dispatch successfully---
+\clj{(rep "a")} type checks but throws a runtime error.
 
 The flexibility of \clj{isa?} is key to the generality of multimethods. 
-For instance, we can dispatch on the value of
-a map entry, even if it is not a class.
-
+We can dispatch on the \clj{:op} key 
+of our HMap AST \clj{Expr}.
+Keywords are functions that look themselves up in their argument, so \clj{:op}
+is our dispatch function.
 \begin{Code}
@@
 \end{Code}
+The destructuring syntax \clj{(fn [{:keys [val] :as m}] ...)} binds
+\clj{m} to the value of the first argument, and \clj{val} to \clj{(:val m)}.
 
-\clj{isa?} is further special-case with vectors: with two vectors of the
-same length, it returns the logical conjunction of the pairwise applications
-of \clj{isa?}.
-
+\clj{isa?} is special with vectors---vectors of the
+same length recursively call \clj{isa?} on the elements pairwise.
 \begin{minted}{clojure}
@@
   (isa? [Keyword Keyword] [Object Object]) ;=> true
+  (isa? [Keyword Object] [Object Object]) ;=> false
 \end{minted}
 
-We can use this to dispatch on more than one thing. Here we perform
-double dispatch on the class of each argument.
-
+Now multiple dispatch is possible---we dispatch on the class of both
+arguments simultaneously by defining a dispatch function that returns
+a vector containing the classes of the arguments.
 \begin{Code}
@@
 \end{Code}
+The dispatch values are also vectors---the first method is picked
+when the left argument is a \clj{Number} and the right is a \clj{Keyword},
+and is thus safe to increment and extract its name respectively.
+The dispatch value \clj{:default} specifies a default
+method if no preferred method is found.
 
-No extra annotations are needed to follow the type-directed control flow.
-For instance, the middle method is has inferred the left argument is a number
-and thus safe to increment, and the right argument is a keyword and safe to
-extract its name.
-
-A dispatch value can also be the keyword \clj{:default} to specify a default
-method if no preferred method is found. The static type of both arguments
-remains unchanged.
+No extra annotations are needed to follow type-directed control flow
+in multimethod dispatch.
 
@@
 and use of HMaps. 
-
 \begin{Code}
@@
 \end{Code}
+\clj{PayLoad} is either a HMap containing a file
+or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
+we know the \clj{:file} is a file.
+The body of the first method uses type-hints to resolve reflection
+and conditional control flow to prove null-pointer exceptions are impossible.
+The second method is similar except it uses exceptional control flow.
 

### 2015-02-25 overview.tex
finish overview

@@
 \citet{TF10}
-introduced Typed Racket with occurrence typing,
+presented Typed Racket with occurrence typing,
 an inference technique for conditional control flow.
@@
 then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
-We use this assumption to check \clj{(inc x)}, automatically
-ruling out the possibility of null-pointer exceptions
-(said formally in section~\ref{sec:doformal}).
+We use this assumption to check \clj{(inc x)} with an automatic
+guarantee that a null-pointer exception is impossible
+(modelled formally in section~\ref{sec:doformal} and proved
+in section~\ref{sec:metatheory}).
 
@@
 \inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
-\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
+\inputminted[firstline=27,lastline=29]{clojure}{code/demo/src/demo/hmap.clj}
 \end{exmp}
@@
 HMaps support various options to track the presence and absence of entries.
-\clj{'{:a Num}} is a shorthand for the \clj{:mandatory} option---entries that
-must be present.
+\clj{:mandatory} declares entries that must be present.
 
@@
 \end{Code}
-
-Looking up \clj{:a} simply returns the associated type.
+Looking up \clj{:a} simply returns the associated type \clj{Num}.
 Keys other than \clj{:a} are unaccounted for in \clj{Mand}, so 
 the best we can infer is \clj{Any}, the supertype of all types.
+\clj{'{:a Num}} is shorthand for \clj{(HMap :mandatory {:a Num})}.
 
 \clj{:absent-keys} is a set of keywords that are known to not have entries.
-We take advantage of this to infer a better type for lookups on known absent keys.
-
 \begin{Code}
@@
 \end{Code}
+We can infer a better type for \clj{(:b m)}---\clj{:b} is absent in \clj{Abs}
+so it looking up \clj{:b} returns \clj{nil}.
 
-\clj{:optional} is a map that declares entries either absent or
+\clj{:optional} is a map that declares entries either absent \emph{or}
 present with a specific type.
@@
 than those declared \clj{:mandatory} or \clj{:optional}.
-
 \begin{Code}
 \begin{exmp}
-\inputminted[firstline=17,lastline=23]{clojure}{code/demo/src/demo/hmap_path.clj}
+\inputminted[firstline=17,lastline=26]{clojure}{code/demo/src/demo/hmap_path.clj}
 \end{exmp}
 \end{Code}
+\clj{:a} is \clj{:mandatory} in \clj{Comp} so lookups return \clj{Num}.
+\clj{:b} is \clj{:optional} so lookups return \clj{(U nil Kw)}.
+\clj{:c} is not mentioned, but since \clj{:complete?} is \clj{true}
+it must be absent so a lookup returns \clj{nil}.
 
@@
 \clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
-removing the raw private key and attaching an encrypted private key.
+removing the raw \clj{:private-key} entry and associating an encrypted private key
+as \clj{:enc-private-key}.
 Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
@@
 
-In the then branch Typed Clojure infers \clj{'{:a Num}}, in the else branch
-\clj{(HMap :optional {:a nil})}. Failed lookups return \clj{nil},
-so we cannot distinguish between \clj{'{:a nil}} and \clj{(HMap :optional {:a nil})}
-by testing the result of looking up \clj{:a}.
+Typed Clojure infers \clj{m} as \clj{'{:a Num}} in the then branch
+and \clj{(HMap :optional {:a nil})}
+in the else branch.
+We cannot infer 
+\begin{minted}{clojure}
+ (HMap :absent-keys #{:a})
+\end{minted}
+in the else branch because \clj{m} could be the value \clj{{:a nil}}
+by the definiton of \clj{Opt}.
 Section~\ref{sec:hmapformal} discusses this in greater detail.

### 2015-02-25 overview.tex
formatting examples

@@
 \begin{exmp}
-\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
+\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
 \end{exmp}
@@
 \begin{exmp}
-\begin{minted}{clojure}
-(ann rep [Any -> String])
-(defmulti rep class)
-(defmethod rep Keyword [x] (str (name x)))
-(defmethod rep Number [x] (str (inc x)))
-
-(rep :a) ;=> "a"
-(rep 1)  ;=> "2"
-\end{minted}
+\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
 \end{exmp}
@@
 \begin{exmp}
-\inputminted[firstline=4,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
+\inputminted[firstline=5,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
 \end{exmp}

### 2015-02-25 formal-model.tex
hmap examples
review first half of formal model

@@
 
-We describe our type system in a similar way to Typed Racket (\citet{TF10}),
+We formalise our type system like \citet{TF10}
 with differences highlighted in $\trdiff{\text{blue}}$.
 The typing judgement 
-$
+$$
 {\judgement   {\propenv}
@@
   {\object{}}}
-$
+$$
 says expression \e{} is of type \t{} in the 
@@
 `then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}.
+and object \object{}. We sometimes write 
+{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.
 
@@
 All binding forms introduce fresh variables.
-
-Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \makekw{a}), 
-constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.
+Values include booleans, \nil{}, class literals, keywords, 
+constants and closures. \false{} and \nil{} are logically false, otherwise values are true.
 
@@
 
-Propositions are identical to~\cite{TF10}---positive and negative type propositions, implications,
-conjunctions, disjuctions, trivial and impossible propositions.
-Type propositions contain a non-empty object which it claims type information about.
+Propositions are standard---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+say object {\path {\pathelem{}} {\x{}}} is and is not of type \t{}
+respectively.
+We include implications,
+conjunctions, disjuctions, and the trivial and impossible propositions.
+Proposition environments are sequences of propositions.
 
-Proposition environments are sequences of propositions in conjunctive normal form.
-
-Figure ~\ref{main:figure:typingrules} contains the standard typing rules.
+Figure~\ref{main:figure:typingrules} contains the standard typing rules.
 The operational semantics are standard, except \nil{} is logically false:
@@
     {\TLet}
+    {\TConst}
 
@@
 \subsection{Reasoning about Exceptional Control Flow}
-
-Clojure assertions expand to simple exceptional control flow.
-
-\begin{minted}{clojure}
-(fn [x :- (U nil Number)] 
-  (assert (number? x))
-  (inc x))
-\end{minted}
-
-roughly expands to
-
-\begin{minted}{clojure}
-(fn [x :- (U nil Number)]
-  (do (if (number? x) nil (throw (new Exception)))
-      (inc x)))
-\end{minted}
+\label{sec:doformal}
 
@@
 \smallskip
-
-B-Do simply evaluates its arguments sequentially and returns the right argument.
-Since errors are not values, we define error propagation semantics
-like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).
-
-\begin{mathpar}
-    {\BDo}
-
-\infer [BE-Error]
-{}
-{ \opsem {\openv{}} 
-         {\errorvalv{}}
-         {\errorvalv{}}}
-
-\infer [BE-Do1]
-{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
-{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
-\end{mathpar}
+%
+%B-Do simply evaluates its arguments sequentially and returns the right argument.
+%Since errors are not values, we define error propagation semantics
+%like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).
+%
+%\begin{mathpar}
+%    {\BDo}
+%
+%\infer [BE-Error]
+%{}
+%{ \opsem {\openv{}} 
+%         {\errorvalv{}}
+%         {\errorvalv{}}}
+%
+%\infer [BE-Do1]
+%{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
+%{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
+%\end{mathpar}
 
@@
 which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
-which allows us to assume \clj{(inc :a)} is unreachable (and it clearly is by BE-Do1).
+which allows us to assume \clj{(inc :a)} is unreachable.
 
-This works similarly with our motivating example. Intuitively, when evaluating
+Now we can check example~\ref{example:doexception}.
+Intuitively, when evaluating
 \clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
@@
 \subsection{Heterogenous maps}
+\label{sec:hmapformal}
 
@@
   \begin{mathpar}
-    {\BAssoc}
-    {\BGet}
-    {\BGetMissing}
-  \end{mathpar}
-  \caption{HMap Semantics}
-  \label{main:figure:hmapsem}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
     {\TGetHMap}
@@
   \begin{mathpar}
+    {\BAssoc}
+    {\BGet}
+    {\BGetMissing}
+  \end{mathpar}
+  \caption{HMap Semantics}
+  \label{main:figure:hmapsem}
+\end{figure}
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
     \HMapsubtyping{}
@@
 Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
-\clj{(HMap :mandatory p :absent-keys a)} is written \HMapgeneric{\mandatory{}}{\absent{}}.
-The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
-specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
-entries and we abbreviate \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}}. Keys cannot be both present and absent.
+$$
+\clj{(HMap :mandatory p :absent-keys a :complete? b)}
+$$
+is written \HMapgeneric{\mandatory{}}{\absent{}},
+where the metavariable \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} otherwise {\partial{}}. 
+\mandatory{} specifies a set of known present keyword entries. \absent{} tracks known absent keys.
+\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. Keys cannot be both present and absent.
 
-\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
+\clj{(get m :a)} is identical to \clj{(:a m)}.
 To simplify presentation, we only provide syntax for the empty map literal and
@@
 We introduce our first path element {\keype{\k{}}}, the symbolic representation
-of a \getliteral{} expression on keyword \k{} where the sentinel value for
-a missing entry is \nil{}.
+of \clj{(get m k)} where \clj{m} represents the path after {\keype{\k{}}}---{\path{\keype{\makekw{b}}}{\path{\keype{\makekw{a}}}{\x{}}}}
+  represents \clj{(get (get x :a) :b)}.
 
-Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps. B-Get returns the associated
-value if the entry is present otherwise  B-GetMissing returns \nil{}. B-Assoc returns a new extended
-map.
-
-Typing rules are given in figure~\ref{main:figure:hmaptyping}. T-GetHMap handles the simple
-case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
-a \Nil{} return type.
-T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
-Even though their types vary significantly, we still use the same object.
-T-Assoc extends the known keyword entries, overwriting previous entries with the same key.
+Figure~\ref{main:figure:hmaptyping} contains the typing rules. T-GetHMap models a lookup
+that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
+and T-GetHMapPartialDefault a lookup with unknown results.
+T-Assoc extends a HMap with a mandatory entry.
+Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
+and \assocliteral{}.
 
@@
 The \classconst{} function is modeled as a constant which returns
-an object's \Class{}, or \nil{} if provided \nil{}.
+the argument's class or \nil{}.
+
+$$
+\begin{array}{lrlr}
+  \constantssyntax\\
+  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
+                &\mbox{Path Elements}
+\end{array}
+$$
 
@@
 Our second path element \classpe{} is the symbolic representation
-of a call to \classconst{}.
+of a call to \classconst{}---{\path{\classpe{}}{\path{\keype{\makekw{a}}}{\x{}}}}
+represents \clj{(class (get x :a))}.
+Figure~\ref{main:figure:primitivesem} contains the semantics for \classconst{}.
 
-$$
-\begin{array}{lrlr}
-  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
-                &\mbox{Path Elements}
-\end{array}
-$$
-
-% TODO semantics
+\constantsemfigure{main}
 
@@
 \end{mathpar}
-
 For example, if \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
@@
 $$
-so we learn \inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).
+so by the definition of \updateliteral{} (figure~\ref{main:figure:update})
+\inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).
 
 Now that we have our path elements \classpe{} and \keype{\k{}}
-and the typings rules that introduce them, we look at \updateliteral{} 
-(figure~\ref{main:figure:update}) and give one interesting example for each.
+and the typing rules that introduce them, we give a detailed example how
+each path element interacts with T-Update.
 
@@
 (fn [x :- (U nil Keyword)]
-  (when (class x)
+  (if (class x)
     ; here
-    ))
+    nil))
 \end{minted}
@@
 $$
-
 This hits the sixth \classpe{} case of \updateliteral{}, which returns
@@
 (fn [x :- '{:kw (U nil Keyword)}]
-  (when (get x :kw)
+  (if (get x :kw)
     ; here
-    ))
+    nil))
 \end{minted}
@@
 $$
-
 rewrites expression \e{} to a possibly-less reflective expression
 \ep{} with respect to type hint environment
-\taenv{} and Java class table \ct{}.
-
+\taenv{} and Java class table \ct{} and
 $$
@@
 $$
-
 infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
@@
 
-We can use \isaliteral{} as a crude \nil{} test
-
+We can rewrite example~\ref{example:conditionalflow} as
 \inputminted[firstline=5]{clojure}{code/demo/src/demo/isa_nil.clj}
@@
 
-Substitution is used to ensure well formed propositions.
+%Substitution is used to ensure well formed propositions.
 
-The final case gives up; we cannot distinguish between equality and subclassing mode.
+The final line gives up---we cannot distinguish between equality and subclassing mode.
 
@@
 multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
-without changing the original multimethod. The above is now written
-
-%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\clj{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
-%                       {\classconst{}}}
-%                     {\\ bar}$
-
+without changing the original multimethod. Example~\ref{example:rep} is now written
 \begin{minted}{clojure}
-(let [foo (defmulti [Any -> String] class)
-      foo (defmethod foo Number [x] (str (inc x)))
-      foo (defmethod foo Keyword [x] (str (name x)))]
-  (foo :a)) ;=> "a"
+(let [rep (defmulti [Any -> String] class)
+      rep (defmethod rep Keyword [x] (str (name x)))
+      rep (defmethod rep Number [x] (str (inc x)))]
+  (rep :a)) ;=> "a"
 \end{minted}
@@
 
-{\footnotesize
-\constantsemfigure{}
-
 {\convertjavatypefigure{figure}{\footnotesize}}
-}
 

### 2015-02-26 overview.tex
review first half of formal model
better examples. explain syntax

@@
 use of the full Typed Clojure system to illustrate the key type system
-ideas, before studying the key features in detail in
+ideas, before studying the core features in detail in
 section~\ref{sec:formal}.
@@
 presented Typed Racket with occurrence typing,
-an inference technique for conditional control flow.
+a technique for deriving type information from conditional control flow.
 They introduce the concept of occurrence typing 
@@
 \inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
+\label{example:conditionalflow}
 \end{exmp}
 
+Typed Clojure can check all the examples in~\citet{TF10}---the 
+rest of this section describes the extensions necessary
+to check Clojure code.
+
+\subsection{Clojure and Typed Clojure Primer}
+
+We introduce Clojure's syntax.
+Parenthesis indicate \emph{applications}, square brackets
+are syntax for literal \emph{vectors} and curly braces delimit literal
+\emph{hash-maps}.
+\clj{nil} is exactly Java's \java{null} and is one of two logically
+false values (the other is \clj{false}).
+\emph{Symbols} begin with an alphabetic character.
+A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
+The \clj{get} function takes a hash-map and a key returns the value
+associated with the key in the map, or returns \clj{nil} if there is none.
+Keywords are functions that take a map and applies \clj{get} to the map
+and itself.
+\clj{inc} and \clj{dec} increment and decrement numbers respectively.
+\clj{assoc} and \clj{dissoc} associate and dissociate entries
+from immutable maps.
+Everything is an instance of some class
+except for \clj{nil}. Predicates like \clj{number?} are equivalent to
+a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.
+
+Symbols support \emph{metadata}, an arbitrary map of
+data. The syntax \clj{^File x} is a single expression that is the symbol \clj{x} 
+with metadata \clj{{:tag File}}.
+In binding positions like \clj{(fn [^File x] ...)}
+the metadata is preserved in occurrences of \clj{x}.
+
+\emph{Commas} are always whitespace.
+\emph{Destructuring} is binding position syntax for pattern matching.
+A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
+to lookups on that keyword with the current argument, optionally terminated by
+\clj{:as} and a symbol, which binds the symbol to the current argument.
+$$
+\clj{(fn [{x :x, y :y, :as m}] ...)}
+$$ 
+binds \clj{m} to the first argument,
+\clj{x} to \clj{(get m :x)} and
+\clj{y} to \clj{(get m :y)} in the function body. Type hints are preserved as expected.
+
 Clojure uses \emph{namespaces}, declared with the
-\clj{ns} form, to manage top-level \emph{vars}.
-This is a regular Clojure file compiled with
+\clj{ns} form, to manage top-level \emph{vars} and
+class imports.
+Example~\ref{example:conditionalflow}
+is a regular Clojure file compiled with
 the Clojure compiler except it declares a runtime dependency on
@@
 when convenient.
+We omit \clj{ns} forms for the rest of the examples.
+\footnote{See <this> github repo for full working examples}
 
-Typed Clojure can check all the examples in the occurrence typing
-paper---the rest of this section describes the extensions necessary
-to check Clojure code.
+Java methods and fields are called in prefix notation
+like \clj{(.method target args*)} and \clj{(.field target)} respectively,
+with method and field names prefixed with a dot and methods taking some number of arguments.
+Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}.
+
+The \clj{if} special form is a standard two-armed conditional.
+The \clj{do} special form sequentially evaluates any number of expressions
+left-to-right, returning \nil{} on zero arguments.
+The \clj{when} macro is a one-armed \clj{if} that returns \clj{nil} if the test fails.
+The \clj{throw} special form is exactly Java's \java{throw}, which takes an
+instance of \clj{Throwable} and throws an exception with it.
+The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
+and any number of expressions in an implicit \clj{do} for the body of the function.
+The typed
+variant of \clj{fn} supports an optional keyword \clj{:-} and type following
+each parameter to declare their static types, and similarly after
+the parameter vector to declare the return type.
+
+In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
+Quoted maps \clj{'{:op ':if}} are HMap types---the keys are not quoted because only keyword keys
+are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
+any number of types.
 
@@
 \inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
+\label{example:doexception}
 \end{exmp}
-
-We combine the sequencing form \clj{do} and a thrown exception with
-\clj{throw} to avoid a null-pointer exception from attempting to increment \clj{nil}.
-(We omit \clj{ns} forms for the rest of the examples)
-\footnote{See <this> github repo for full working examples}.
 
@@
 \begin{exmp}
-\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
-\inputminted[firstline=27,lastline=29]{clojure}{code/demo/src/demo/hmap.clj}
+\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
 \end{exmp}
@@
 \clj{an-exp} is a function verified to return an \clj{Expr}.
-
-HMaps support various options to track the presence and absence of entries.
-\clj{:mandatory} declares entries that must be present.
-
-\begin{Code}
-\begin{exmp}
-\inputminted[firstline=58,lastline=64]{clojure}{code/demo/src/demo/hmap.clj}
-\end{exmp}
-\end{Code}
-Looking up \clj{:a} simply returns the associated type \clj{Num}.
-Keys other than \clj{:a} are unaccounted for in \clj{Mand}, so 
-the best we can infer is \clj{Any}, the supertype of all types.
-\clj{'{:a Num}} is shorthand for \clj{(HMap :mandatory {:a Num})}.
-
-\clj{:absent-keys} is a set of keywords that are known to not have entries.
-\begin{Code}
-\begin{exmp}
-\inputminted[firstline=66,lastline=73]{clojure}{code/demo/src/demo/hmap.clj}
-\end{exmp}
-\end{Code}
-We can infer a better type for \clj{(:b m)}---\clj{:b} is absent in \clj{Abs}
-so it looking up \clj{:b} returns \clj{nil}.
-
-\clj{:optional} is a map that declares entries either absent \emph{or}
-present with a specific type.
-
-\begin{Code}
-\begin{exmp}
-\inputminted[firstline=5,lastline=12]{clojure}{code/demo/src/demo/hmap_path.clj}
-\end{exmp}
-\end{Code}
-
-Concretely \clj{(HMap :optional {:a Num})}
-is the same as
-\begin{minted}{clojure}
-      (U '{:a Num} (HMap :absent-keys #{:a})).
-\end{minted}
-
-\clj{:complete?} is \clj{true} if there are no further entries
-than those declared \clj{:mandatory} or \clj{:optional}.
-\begin{Code}
-\begin{exmp}
-\inputminted[firstline=17,lastline=26]{clojure}{code/demo/src/demo/hmap_path.clj}
-\end{exmp}
-\end{Code}
-\clj{:a} is \clj{:mandatory} in \clj{Comp} so lookups return \clj{Num}.
-\clj{:b} is \clj{:optional} so lookups return \clj{(U nil Kw)}.
-\clj{:c} is not mentioned, but since \clj{:complete?} is \clj{true}
-it must be absent so a lookup returns \clj{nil}.
 
@@
 
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=10,lastline=23]{clojure}{code/demo/src/demo/key2.clj}
+\end{exmp}
+\end{Code}
+
 HMaps interact with occurrence typing in interesting ways.
-The next example branches on the result of a lookup, and the types
-inferred for \clj{f} are subtle.
 
@@
 \begin{exmp}
-\inputminted[firstline=14,lastline=15]{clojure}{code/demo/src/demo/hmap_path.clj}
+\inputminted[firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
 \end{exmp}
@@
 
-Typed Clojure infers \clj{m} as \clj{'{:a Num}} in the then branch
-and \clj{(HMap :optional {:a nil})}
-in the else branch.
-We cannot infer 
-\begin{minted}{clojure}
- (HMap :absent-keys #{:a})
-\end{minted}
-in the else branch because \clj{m} could be the value \clj{{:a nil}}
-by the definiton of \clj{Opt}.
 Section~\ref{sec:hmapformal} discusses this in greater detail.
@@
 
-To resolve reflective calls, Clojure supports \emph{type-hints}.
+To resolve reflective calls, Clojure supports \emph{type hints}.
 
@@
 
-Typed Clojure ignores type-hints, so to check this we must
+Typed Clojure ignores type hints, so to check this we must
 annotate the parameter's static type.
@@
 We extend \clj{rep}'s dispatch table, mapping
-the \emph{dispatch value} \clj{Number} to the function
+the \emph{dispatch value} \clj{Keyword} to the function
 \clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
@@
 \inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
+\label{example:rep}
 \end{exmp}
@@
 \begin{exmp}
-\inputminted[firstline=5,lastline=23]{clojure}{code/demo/src/demo/eg5.clj}
+\inputminted[firstline=5,lastline=25]{clojure}{code/demo/src/demo/eg5.clj}
 \end{exmp}
@@
 we know the \clj{:file} is a file.
-The body of the first method uses type-hints to resolve reflection
+The body of the first method uses type hints to resolve reflection
 and conditional control flow to prove null-pointer exceptions are impossible.

### 2015-02-26 formal-model.tex
section 4 better intro

@@
 
-We formalise our type system like \citet{TF10}
-with differences highlighted in $\trdiff{\text{blue}}$.
-The typing judgement 
+Now that we have demonstrated the core features Typed Clojure
+provides, we present a model that links them all together.
+Our presentation will start with a review of
+occurrence typing~\cite{TF10}.
+Then for the rest of the section we incrementally add each
+novel feature of Typed Clojure to the formalism,
+interleaving presentation of typing rules, operational semantics
+and
+subtyping.
+
+The first insight about occurrence typing is that
+propositions can be used to capture type information about our programs,
+in particular by referencing components or subcomponents of the runtime environment
+and relating them to types.
+\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
+``variable \clj{x} is not \clj{nil}''---
+in our logical system we write these as
+{\isprop{\Number}{\x{}}}
+and {\notprop{\ValueNil}{\x{}}}. 
+The other propositions are standard: the implication, conjunction,
+disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
+(figure~\ref{main:figure:termsyntax}).
+
+The particular part of the environment we reference in a positive or negative
+type proposition is called the \emph{object}.
+Every expression has an object that is either \emph{empty}, written \emptyobject{}, which says there is no
+known way to relate this expression to the current environment, or a 
+variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
+that exactly indicates how the value of this
+expression can be derived from the current environment.
+Type propositions can only reference non-empty objects.
+
+The second insight is that we can replace the traditional 
+representation of a
+type environment of a map from variables to types
+with a sequence of propositions, written \propenv{}. 
+Instead of mapping \clj{x} to
+the type \clj{Number}, we instead use the proposition {\isprop{\Number}{\x{}}}.
+
+The third insight is that we
+can use a logic system to learn new information about our programs.
+The key component of the logic system L-Update uses type propositions
+to derive new propositions.
 $$
-{\judgement   {\propenv}
-              {\hastype {\e{}} {\t{}}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-  {\object{}}}
+  {\LUpdate}
 $$
-says expression \e{} is of type \t{} in the 
-proposition environment $\propenv{}$, with 
-`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}. We sometimes write 
-{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.
+For example, with L-Update we can use the knowledge of
+\inpropenv{\propenv{}}{\isprop{\Union{\ValueNil{}}{\Number{}}}{\x{}}}
+and 
+\inpropenv{\propenv{}}{\notprop{\ValueNil{}}{\x{}}}
+to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
+(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
+We cover \updateliteral{} in section~\ref{sec:updateformalsection}.
 
-The propositions capture what can be assumed if an expression tests as logically true or false
-respectively. The object is a possibly-empty symbolic representation of the expression.
-
-Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
-application, abstractions, conditionals and let expressions.
-All binding forms introduce fresh variables.
-Values include booleans, \nil{}, class literals, keywords, 
-constants and closures. \false{} and \nil{} are logically false, otherwise values are true.
-
-Types include the top type, untagged unions, functions, singleton types
-and class instances. We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{} and similarly
-for \True and \False.
-Function types contain \emph{latent} propositions and object, which, along with the return type,
-may refer to the parameter identifier. They are latent because they are instantiated with the
-actual object of the argument in applications before they are used in the proposition environment.
-
-Propositions are standard---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-say object {\path {\pathelem{}} {\x{}}} is and is not of type \t{}
+The fourth insight is a consequence of
+programmers in dynamic languages
+using implicit type-based reasoning based on the result of
+conditional tests.
+For instance, example~\ref{example:conditionalflow}
+only utilises \clj{x} once
+the programmer is convinced it is safe to do so based whether
+\clj{(number? x)}
+is 
+a true or false value. The takeaway is that every expression has
+two relevant pieces of information to the logic system: a `then' proposition
+for when it reduces to a true value, and an `else' proposition
+when it reduces to a false value---for \clj{(number? x)}
+they are {\isprop{\Number}{\x{}}} and {\notprop{\Number}{\x{}}}
 respectively.
-We include implications,
-conjunctions, disjuctions, and the trivial and impossible propositions.
-Proposition environments are sequences of propositions.
-
-Figure~\ref{main:figure:typingrules} contains the standard typing rules.
-The operational semantics are standard, except \nil{} is logically false:
-
-\begin{mathpar}
-    \BIfTrue{}
-
-    \BIfFalse{}
-\end{mathpar}
-
-Subtyping (figure~\ref{main:figure:subtyping}) says singleton types are instances of their respective classes
-and \Object is transitively a supertype to everything except {\Nil{}},
-so \Top{} = {\Union{\Nil}{\Object}}.
+%\ref{main:figure:typingrules}
 
@@
 \end{figure}
+
+We formalise our type system following~\citet{TF10}
+(with differences highlighted in $\trdiff{\text{blue}}$).
+The typing judgement 
+$$
+{\judgement   {\propenv}
+              {\hastype {\e{}} {\t{}}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}}
+$$
+says expression \e{} is of type \t{} in the 
+proposition environment $\propenv{}$, with 
+`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+and object \object{}. We sometimes write 
+{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.
+
+Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
+application, abstractions, conditionals and let expressions.
+All binding forms introduce fresh variables.
+Values include booleans, \nil{}, class literals, keywords, 
+constants and closures. \false{} and \nil{} are logically false, otherwise values are true.
+
+Types include the top type, \emph{untagged} unions, functions, singleton types
+and class instances. We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
+{\ValueTrue} as \True and {\ValueFalse} as {\False}.
+Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
+may refer to the parameter identifier. They are latent because they are instantiated with the
+actual object of the argument in applications before they are used in the proposition environment.
+
+Figure~\ref{main:figure:typingrules} contains the standard typing rules.
+We can now demonstrate how to check example~\ref{example:conditionalflow}.
+$$
+\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{Boolean}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
+$$
+$$
+\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+$$
+$$
+\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
+$$
+$$
+\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{0}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+$$
+By T-Abs
+
+The operational semantics are standard, except \nil{} is logically false:
+
+\begin{mathpar}
+    \BIfTrue{}
+
+    \BIfFalse{}
+\end{mathpar}
+
+Subtyping (figure~\ref{main:figure:subtyping}) says singleton types are instances of their respective classes
+and \Object is transitively a supertype to everything except {\Nil{}},
+so \Top{} = {\Union{\Nil}{\Object}}.
 
@@
 We extend our model with sequencing expressions and errors, where {\errorvalv{}}
-is equivalent to Clojure's throw.
+models Clojure's throw special form.
 
@@
 if the left argument reduces to a value, then it is either logically true or false.
-Occurrence typing gives exactly the propositions for both cases, so their disjunction
-can be assumed for the right argument.
+If we learn the same information in both cases then we can use it as an assumption to check the right argument.
 T-Do formalises this intuition.
@@
     {\TDo}  
-
-    {\TError}
 \end{mathpar}
@@
 makes our insight interesting.
-For example, in checking \clj{(do (throw (Exception.)) (inc :a))}
-by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \clj{(inc :a)},
-which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
-which allows us to assume \clj{(inc :a)} is unreachable.
+
+\begin{mathpar}
+    {\TError}
+\end{mathpar}
 
 Now we can check example~\ref{example:doexception}.
-Intuitively, when evaluating
-\clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
-so we assume \clj{x} is a number.
-In detail, the example is checked first by T-Abs, which checks the body with the 
-environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
-Since the body is a \doliteral{},
-T-Do extracts the propositions of the left argument, an if:
-T-If says 
-\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
-by T-Nil and T-Subsume and
-\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
-Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
-and we check \clj{(inc x)} in the extended environment 
-\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
-By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
-so T-Local checks \clj{x} as a \Number.
+We check the test with T-App.
+$$
+\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{B}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
+$$
+By T-Nil and subsumption we derive the same information in both propositions.
+$$
+\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
+$$
+By T-Error we can assume anything about the return of an error.
+$$
+\judgement{\notprop{\clj{Num}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
+$$
+Using the above as premises to T-If we learn {\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}
+in both propositions.
+$$
+\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(if (number? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
+          {\filterset{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{Num}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
+$$
+By T-Local we know \clj{x} is a \clj{Num}.
+$$
+\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
+$$
+\clj{inc} requires a \clj{Num} argument which we just proved.
+$$
+\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+$$
 
@@
 
+\subsection{Paths}
+
 We introduce our first path element {\keype{\k{}}}, the symbolic representation
@@
 
-\subsection{\classconst{}}
-
-The \classconst{} function is modeled as a constant which returns
+The \classconst{} function is a constant which returns
 the argument's class or \nil{}.
@@
 \begin{array}{lrlr}
-  \constantssyntax\\
   \pesyntax{}   &::=& \ldots \alt {\classpe{}}
@@
 where we combine propositions that have similar objects to learn new facts.
-(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
 
@@
 and the typing rules that introduce them, we give a detailed example how
-each path element interacts with T-Update.
+each path element interacts with L-Update.
 
@@
 {\path{\classpe{}}{\x{}}}, 
-by T-Update 
+by L-Update 
 $$
@@
 
-T-Update combines these propositions
+L-Update combines these propositions
 $$

### 2015-02-26 overview.tex
section 4 better intro

@@
 from immutable maps.
+The \clj{=} function returns true if all its arguments are structurally equal---in this
+paper we treat it as pointer equality.
 Everything is an instance of some class
@@
 is a regular Clojure file compiled with
-the Clojure compiler except it declares a runtime dependency on
-\clj{clojure.core.typed}, Typed Clojure's core namespace.
+the Clojure compiler using \clj{:require} to declare a
+runtime dependency on Typed Clojure's core namespace
+\clj{clojure.core.typed}.
 Programmers import typed versions of certain constructs as needed, like
@@
 In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
-Quoted maps \clj{'{:op ':if}} are HMap types---the keys are not quoted because only keyword keys
+Quoted maps like \clj{'{:op ':const, :val Num}} are HMap types representing
+maps with known keyword entries with values of the associated type
+The keys are not quoted because only keyword keys
 are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
 any number of types.
+
+The \clj{defalias} macro defines a type abbreviation which can be recursively defined.
 

### 2015-02-26 formal-model.tex
section3 intro

@@
 in our logical system we write these as
-{\isprop{\Number}{\x{}}}
-and {\notprop{\ValueNil}{\x{}}}. 
+{\isprop{\clj{Number}}{\x{}}}
+and {\notprop{\clj{nil}}{\x{}}}. 
 The other propositions are standard: the implication, conjunction,
@@
 Instead of mapping \clj{x} to
-the type \clj{Number}, we instead use the proposition {\isprop{\Number}{\x{}}}.
+the type \clj{Number}, we instead use the proposition {\isprop{\clj{Number}}{\x{}}}.
 
@@
 For example, with L-Update we can use the knowledge of
-\inpropenv{\propenv{}}{\isprop{\Union{\ValueNil{}}{\Number{}}}{\x{}}}
+\inpropenv{\propenv{}}{\isprop{\clj{(U nil Number)}}{\x{}}}
 and 
-\inpropenv{\propenv{}}{\notprop{\ValueNil{}}{\x{}}}
-to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
+\inpropenv{\propenv{}}{\notprop{\clj{nil}}{\x{}}}
+to derive \inpropenv{\propenv{}}{\isprop{\clj{Number}}{\x{}}}.
 (The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
@@
 a true or false value. The takeaway is that every expression has
-two relevant pieces of information to the logic system: a `then' proposition
+two extra pieces of information relevant to the logic system: a `then' proposition
 for when it reduces to a true value, and an `else' proposition
 when it reduces to a false value---for \clj{(number? x)}
-they are {\isprop{\Number}{\x{}}} and {\notprop{\Number}{\x{}}}
-respectively.
+the then proposition is {\isprop{\clj{Number}}{\x{}}} and 
+the else proposition is {\notprop{\clj{Number}}{\x{}}}.
 %\ref{main:figure:typingrules}
@@
 We can now demonstrate how to check example~\ref{example:conditionalflow}.
+We abbreviate \clj{Num} as \clj{N},  \clj{Boolean} as \clj{B} and \clj{number?} as \clj{n?}.
+The test is checked with T-App, 
 $$
-\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{Boolean}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
+$$
+\clj{n?} has type
+$$
+{\ArrowOne{\x{}}{\clj{Any}}{\clj{B}}
+        {\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}}
+$$
+By T-Local.
+$$
+\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
+$$
+Now we check both branches.
+$$
+\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
 $$
-\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\clj{0}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
+We can now check the conditional with T-If.
 $$
-\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
+\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(if (n?
+x) (inc x) 0)}}{\clj{N}}}{\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}
 $$
+Finally the function can be checked with T-Abs.
 $$
-\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{0}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{}{\hastype{\clj{(fn ...)}}
+                                             {\ArrowOne{\x{}}{\clj{(U nil N)}}{\clj{N}}
+        {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
+  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
 $$

### 2015-02-26 formal-model.tex
explain subtyping

@@
 Types include the top type, \emph{untagged} unions, functions, singleton types
-and class instances. We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
+and class instances. 
+We write the class literal \clj{Boolean} as \Boolean{} and \clj{Keyword} as \Keyword{}.
+The type \Value{\Keyword} is inhabited by \clj{Keyword} and \clj{:a} is of type \Keyword{}.
+We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
 {\ValueTrue} as \True and {\ValueFalse} as {\False}.
@@
 Figure~\ref{main:figure:typingrules} contains the standard typing rules.
+The first interesting rule is T-If. The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
+are used as assumptions in the then and else branch respectively.
+If the result of the \ifliteral{} is true, then the value either
+came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
+which implies \thenprop{\prop{3}} is true. 
+The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} for
+similar reasons, but instead for when each branch is a false value.
+The second interesting rule is T-Local, where we actually utilise the logic system
+to extract a type for a local binding. This rule is why L-Update is crucial---T-Local
+looks for positive type propositions in \propenv{} with the same object as the binding.
+By subsumption we can choose the most specific positive proposition and thus
+T-Local can return the most specific type inferred by the logic system.
+
 We can now demonstrate how to check example~\ref{example:conditionalflow}.
@@
 $$
-\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
+\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
 $$
@@
 $$
+\clj{inc} has type
+$$
+{\ArrowOne{\x{}}{\clj{N}}{\clj{N}}
+        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
+$$
 We can now check the conditional with T-If.
@@
 $$
-Finally the function can be checked with T-Abs.
+Finally the function can be checked with T-Abs
 $$
@@
 $$
-By T-Abs
 
-The operational semantics are standard, except \nil{} is logically false:
+The operational semantics for our language are standard, except \nil{} is a false value:
 
@@
 
-Subtyping (figure~\ref{main:figure:subtyping}) says singleton types are instances of their respective classes
-and \Object is transitively a supertype to everything except {\Nil{}},
+Subtyping (figure~\ref{main:figure:subtyping}) 
+is a reflexive and transitive relation with top type \Top. All objects are subtypes
+of \emptyobject{}.
+Singleton types are instances of their respective classes---boolean singleton types
+are of type \Boolean{}, class literals are instances of \Class{} and keywords are
+instances of \Keyword{}.
+Instances of classes \class{} are subtypes of \Object{}. Since function types 
+are subtypes of \IFn{}, all types except for \Nil{} are subtypes of \Object{},
 so \Top{} = {\Union{\Nil}{\Object}}.
+Function subtyping is contravariant left of the arrow---latent propositions, object
+and the result type are covariant.
+Subtyping for untagged unions is standard.
 

### 2015-02-26 formal-model.tex
explain T-Do better

@@
 Values include booleans, \nil{}, class literals, keywords, 
-constants and closures. \false{} and \nil{} are logically false, otherwise values are true.
+constants and closures. 
+Value environments map local bindings to values.
+Closures contain the value environment \openv{} when the function was created,
+which is used during beta reduction to ensure the abstraction respects lexical scoping.
+\false{} and \nil{} are logically false, otherwise values are true.
 
@@
 \end{mathpar}
+
+For example,
+B-IfTrue evaluates \e{1} in the environment \openv{} with the result \v{1}.
+If \v{1} is not \nil{} or \false{} then the conditional returns the result of evaluating the then branch \e{2}
+in \openv{}.
 
@@
 We extend our model with sequencing expressions and errors, where {\errorvalv{}}
-models Clojure's throw special form.
+models the result of calling Clojure's \clj{throw} special form
+with some \clj{Throwable}.
 
@@
 $$
-\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{B}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\notprop{\clj{Num}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
+\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(if (number? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
-          {\filterset{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{Num}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
+\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(if (n? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
+          {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
 $$
-By T-Local we know \clj{x} is a \clj{Num}.
+By T-Local we know \clj{x} is a \clj{N}.
 $$
-\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
+\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
 $$
+By T-App,
 \clj{inc} requires a \clj{Num} argument which we just proved.
 $$
-\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+$$
+Finally we conclude with T-Do using the conditional and increment expressions as premises.
+$$
+\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(do (if (n? x) nil}\ {\errorvalv{}}{\clj{) (inc x))}}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
@@
   $$
-  \caption{HMap Terms and Types}
-  \label{main:figure:hmapsyntax}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{HMap Typing Rules}
-  \label{main:figure:hmaptyping}
+  \caption{HMap Terms, Types and Typing Rules}
+  \label{main:figure:hmapsyntax}
 \end{figure}
@@
 
-Figure~\ref{main:figure:hmaptyping} contains the typing rules. T-GetHMap models a lookup
+Figure~\ref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
 that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail

### 2015-02-26 conclusion.tex
tweak primer. more formal model

@@
 handle seemingly unrelated features such as multi-methods. In
-addition, Typed Clojure supports crucial features such as heterogenous
+addition, Typed Clojure supports crucial features such as heterogeneous
 maps and Java interoperability while integrating these features into

### 2015-02-26 formal-model.tex
tweak primer. more formal model

@@
 
-\subsection{Heterogenous maps}
+\subsection{Heterogeneous maps}
 \label{sec:hmapformal}
@@
   \end{mathpar}
-  \caption{HMap Semantics}
-  \label{main:figure:hmapsem}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{HMap subtyping rules}
-  \label{main:figure:hmapsubtype}
+  \caption{HMap Semantics and Subtyping rules}
+  \label{main:figure:hmapsem}
 \end{figure}
@@
 Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
+The syntax 
 $$
@@
 $$
-is written \HMapgeneric{\mandatory{}}{\absent{}},
-where the metavariable \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} otherwise {\partial{}}. 
-\mandatory{} specifies a set of known present keyword entries. \absent{} tracks known absent keys.
-\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. Keys cannot be both present and absent.
+we introduced in section~\ref{sec:overview}
+is now written \HMapgeneric{\mandatory{}}{\absent{}},
+where \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} and is {\partial{}} otherwise. 
+The \mandatory{} entry specifies a set of known present keyword entries and \absent{} tracks known absent keys.
+To ease presentation, 
+if a HMap is declared as \complete{} then \absent{} implicitly contains all other keywords not specified in \mandatory{}.
+%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
+Keys cannot be both present and absent.
 
-\clj{(get m :a)} is identical to \clj{(:a m)}.
+The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
+we only model the former to avoid the added complexity of keywords being functions.
 To simplify presentation, we only provide syntax for the empty map literal and
@@
 
-\subsection{Paths}
-
-We introduce our first path element {\keype{\k{}}}, the symbolic representation
-of \clj{(get m k)} where \clj{m} represents the path after {\keype{\k{}}}---{\path{\keype{\makekw{b}}}{\path{\keype{\makekw{a}}}{\x{}}}}
-  represents \clj{(get (get x :a) :b)}.
+Subtyping for HMaps (figure~\ref{main:figure:hmapsem})
+designate \MapLiteral{} as a common supertype for all HMaps.
+S-HMap says that a HMap is a subtype of another HMap if they agree
+on \completenessmeta{}, agree on mandatory entries with subtyping
+and at least cover the absent keys of the supertype.
+Complete maps are subtypes of partial maps
+as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).
 
@@
 and T-GetHMapPartialDefault a lookup with unknown results.
+We only handle unions of HMaps in T-GetHMap, where every element in the union
+contains the key we are looking up, in particular to support operations like
+looking up \clj{:op} on a map of type \clj{Expr} (example~\ref{example:decleaf}).
+The object in each T-Get rule is now more complicated than in T-Local---the 
+next section discusses this in detail.
 T-Assoc extends a HMap with a mandatory entry.
+
 Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
 and \assocliteral{}.
+The B-Get rule evaluates first evaluates the map and the key, asserts
+the key must be present in the map and returns the associated key. If
+the entry is missing, B-GetMissing says we return \nil{}.
 
-S-HMap (figure~\ref{main:figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
-subtypes and also cover the absent keys.
-Complete maps are subtypes of partial maps by S-HMapP.
+\subsection{Paths}
+
+Recall the first insight of occurrence typing---not
+only do we want to reason about components of the runtime environment, that
+is, variables, but \emph{subcomponents} of the environment.
+We have already mentioned paths \path{\pathelem{}}{\x{}}
+where \pathelem{} is a sequence of selectors on \x{} called \emph{path elements},
+but we have not yet defined what components of a path can be.
+\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
+to reason about selector operations on cons cells.
+We instead want to reason about HMap lookups.
+
+We introduce our first path element
+{\keype{\k{}}}, which represents the operation of looking up
+a key \k{}.
+We directly relate this to our typing rule T-GetHMap
+(figure~\ref{main:figure:hmapsyntax}) by
+demonstrating how the first conditional test is checked in example~\ref{example:decleaf}.
+First we unfold the destructuring syntax.
+\begin{minted}{clojure}
+  (fn [m :- Expr]
+    (let [op (:op m)]
+      (if (= :if (:op m))
+        {:op :if, ...}
+        (if ...))))
+\end{minted}
 
@@
 
-Our second path element \classpe{} is the symbolic representation
-of a call to \classconst{}---{\path{\classpe{}}{\path{\keype{\makekw{a}}}{\x{}}}}
-represents \clj{(class (get x :a))}.
-Figure~\ref{main:figure:primitivesem} contains the semantics for \classconst{}.
-
 \constantsemfigure{main}
-
-\subsection{Proof system}
-
-The essential piece of occurrence typing's logical system is L-Update,
-where we combine propositions that have similar objects to learn new facts.
-
-\begin{mathpar}
-  {\LUpdate}
-\end{mathpar}
-For example, if \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
-and
-\inpropenv {\propenv{}} {\notprop {\Nil{}} {\x{}}},
-then
-$$
-\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype{\Nil{}}} {\emptypath{}}} {\x{}}},
-$$
-so by the definition of \updateliteral{} (figure~\ref{main:figure:update})
-\inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).
-
-Now that we have our path elements \classpe{} and \keype{\k{}}
-and the typing rules that introduce them, we give a detailed example how
-each path element interacts with L-Update.
-
-\smallsection{\classpe{}}
-Consider the then branch of the following program:
-
-\begin{minted}{clojure}
-(fn [x :- (U nil Keyword)]
-  (if (class x)
-    ; here
-    nil))
-\end{minted}
-
-Intuitively, since \clj{(class x)} is a true value it must be a class, therefore
-\clj{x} is an object (not \clj{nil}).
-
-From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
-The test has the object {\path{\classpe{}}{\x{}}} by T-App.
-By T-If we can assume
-{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
-Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
-{\path{\classpe{}}{\x{}}}, 
-by L-Update 
-$$
-\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
-$$
-This hits the sixth \classpe{} case of \updateliteral{}, which returns
-$$
-{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
-$$
-deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
-by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).
-
-\smallsection{\keype{\k{}}}
-Consider the then branch of
-
-\begin{minted}{clojure}
-(fn [x :- '{:kw (U nil Keyword)}]
-  (if (get x :kw)
-    ; here
-    nil))
-\end{minted}
-
-Similar to before,
-by T-Abs
-{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\x{}}},
-the test's object is {\path{\keype{\makekw{kw}}}{\x{}}} by T-App
-and by T-If we assume
-{\notprop {\falsy{}} {\path{\keype{\makekw{kw}}}{\x{}}}} in the then branch.
-
-L-Update combines these propositions
-$$
-\inpropenv {\propenv{}} {\isprop {\update {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\nottype {\falsy{}}} {\keype{\makekw{kw}}}} {\x{}}}.
-$$
-matching the \updateliteral{} case which pushes the \updateliteral{} inside the map
-$$
-{\HMapp{\mandatorysetsingleton{\makekw{kw}}{\update{\Union{\Nil}{\Keyword}}{\nottype {\falsy{}}}{\emptypath{}}}}{\emptyabsent}}
-$$
-concluding with
-$$
-\inpropenv {\propenv{}}
-{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Keyword}}{\emptyabsent}}   {\x{}}}.
-$$
 

### 2015-02-26 intro.tex
tweak primer. more formal model

@@
 central innovations, including Java interoperability, multimethods,
-and heterogenously-typed immutable maps, enable this example and many
+and heterogeneously-typed immutable maps, enable this example and many
 others. We begin with an example-driven presentation of the main type

### 2015-02-26 overview.tex
tweak primer. more formal model

@@
 The \clj{=} function returns true if all its arguments are structurally equal---in this
-paper we treat it as pointer equality.
+paper we use it only for pointer equality.
 Everything is an instance of some class
@@
 \emph{Commas} are always whitespace.
-\emph{Destructuring} is binding position syntax for pattern matching.
-A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
-to lookups on that keyword with the current argument, optionally terminated by
-\clj{:as} and a symbol, which binds the symbol to the current argument.
-$$
-\clj{(fn [{x :x, y :y, :as m}] ...)}
-$$ 
-binds \clj{m} to the first argument,
-\clj{x} to \clj{(get m :x)} and
-\clj{y} to \clj{(get m :y)} in the function body. Type hints are preserved as expected.
 
@@
 with method and field names prefixed with a dot and methods taking some number of arguments.
-Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}.
+The canonical form for methods is \clj{(. target (method args*))} 
+and for fields \clj{(. target field)} using the dot special form, which we utilise occasionally
+for clarity.
+Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}
+or equivalently \clj{(new class args*)} with the \clj{new} special form.
 
@@
 The \clj{do} special form sequentially evaluates any number of expressions
-left-to-right, returning \nil{} on zero arguments.
-The \clj{when} macro is a one-armed \clj{if} that returns \clj{nil} if the test fails.
+left-to-right, returning the last expression or \nil{}.
+The \clj{when} macro is a one-armed \clj{if} with an implicit \clj{do}
+around arguments after the test, and returns \clj{nil} if the test fails.
 The \clj{throw} special form is exactly Java's \java{throw}, which takes an
-instance of \clj{Throwable} and throws an exception with it.
+instance of \clj{Throwable} and throws it as an exception.
 The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
 and any number of expressions in an implicit \clj{do} for the body of the function.
+The \clj{let} macro takes a vector of bindings, a flat vector of binding-value pairs,
+and scopes them in the body, an implicit \clj{do}.
+Scoping works like \clj{let*} in Scheme---that is values in the binding vector can refer
+to variables previously introduced in the same binder.
 The typed
@@
 the parameter vector to declare the return type.
+
+\emph{Destructuring} is binding position syntax for pattern matching.
+A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
+to lookups on that keyword with the current argument, optionally terminated by
+\clj{:as} and a symbol, which binds the symbol to the current argument. For example
+$$
+\clj{(fn [{^File x :x, ^String y :y, :as m}] ...)}
+$$ 
+expands to
+\begin{minted}{clojure}
+  (fn [m___0]
+    (let [m m___0,
+          ^File x (:x m),
+          ^String y (:y m)]
+      ...))
+\end{minted}
+Destructuring is supported similarly by \clj{let}.
 
@@
 
-The \clj{defalias} macro defines a type abbreviation which can be recursively defined.
+The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
 
@@
 \inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
+\label{example:decleaf}
 \end{exmp}

### 2015-02-26 small.tex
2.5
tweak primer. more formal model

@@
 \newcommand{\clj}[1]{\mintinline{clojure}{#1}}
+\newcommand{\java}[1]{\mintinline{java}{#1}}
 \newcommand{\rkt}[1]{\mintinline{racket}{#1}}
@@
 Clojure handles this in the type system with multiple forms of
-heterogenous dictionary types. Third, multimethods provide extensible
+heterogeneous dictionary types. Third, multimethods provide extensible
 operations, and their Clojure semantics turns out to have a surprising

### 2015-02-26 formal-model.tex
use formalism in judgements

@@
 in our logical system we write these as
-{\isprop{\clj{Number}}{\x{}}}
-and {\notprop{\clj{nil}}{\x{}}}. 
+{\isprop{\NumberFull}{\x{}}}
+and {\notprop{\Nil{}}{\x{}}}. 
 The other propositions are standard: the implication, conjunction,
@@
 Instead of mapping \clj{x} to
-the type \clj{Number}, we instead use the proposition {\isprop{\clj{Number}}{\x{}}}.
+the type \clj{Number}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
 
@@
 For example, with L-Update we can use the knowledge of
-\inpropenv{\propenv{}}{\isprop{\clj{(U nil Number)}}{\x{}}}
+\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
 and 
-\inpropenv{\propenv{}}{\notprop{\clj{nil}}{\x{}}}
-to derive \inpropenv{\propenv{}}{\isprop{\clj{Number}}{\x{}}}.
+\inpropenv{\propenv{}}{\notprop{\nil{}}{\x{}}}
+to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
 (The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
@@
 when it reduces to a false value---for \clj{(number? x)}
-the then proposition is {\isprop{\clj{Number}}{\x{}}} and 
-the else proposition is {\notprop{\clj{Number}}{\x{}}}.
+the then proposition is {\isprop{\NumberFull}{\x{}}} and 
+the else proposition is {\notprop{\NumberFull}{\x{}}}.
 %\ref{main:figure:typingrules}
@@
 and class instances. 
-We write the class literal \clj{Boolean} as \Boolean{} and \clj{Keyword} as \Keyword{}.
-The type \Value{\Keyword} is inhabited by \clj{Keyword} and \clj{:a} is of type \Keyword{}.
+We write the class literal \Booleanlong{} as \Boolean{} and \Keywordlong{} as \Keyword{}.
+The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
 We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
@@
 We can now demonstrate how to check example~\ref{example:conditionalflow}.
-We abbreviate \clj{Num} as \clj{N},  \clj{Boolean} as \clj{B} and \clj{number?} as \clj{n?}.
+We abbreviate \NumberFull{} as \Number{} and \Booleanlong{} as \Boolean{}.
 The test is checked with T-App, 
 $$
-\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\UnionNilNum{}}{\x{}}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
-\clj{n?} has type
+\numberhuh{} has type
 $$
-{\ArrowOne{\x{}}{\clj{Any}}{\clj{B}}
-        {\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}}
+{\ArrowOne{\x{}}{\Top}{\Boolean}
+        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
 $$
@@
 $$
-\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
+\judgement{\isprop{\Number}{\x{}}}{\hastype{\x{}}{\Number{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
 $$
@@
 $$
-\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
 $$
-\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\clj{0}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
-\clj{inc} has type
+\inc{} has type
 $$
-{\ArrowOne{\x{}}{\clj{N}}{\clj{N}}
+{\ArrowOne{\x{}}{\Number}{\Number}
         {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
@@
 $$
-\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(if (n?
-x) (inc x) 0)}}{\clj{N}}}{\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}
+\judgement{\isprop{\Number}{\x{}}}{\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\zeroliteral{}}}{\Number}}{\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{}{\hastype{\clj{(fn ...)}}
-                                             {\ArrowOne{\x{}}{\clj{(U nil N)}}{\clj{N}}
-        {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
+\judgement{}{\hastype{\abs{\x{}}{\UnionNilNum}{\ ...}}
+                                             {\ArrowOne{\x{}}{\UnionNilNum}{\Number}
+        {\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
   {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
@@
 
-The operational semantics for our language are standard, except \nil{} is a false value:
+The operational semantics for our language standard defined in big-step style, 
+The notable different is \nil{} is a false value, which require changes 
+to the semantics of \ifliteral{}:
 
@@
 Now we can check example~\ref{example:doexception}.
-We check the test with T-App.
+We check the test with T-App:
 $$
-\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
+\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
 $$
-Using the above as premises to T-If we learn {\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}
+Using the above as premises to T-If we learn {\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}
 in both propositions.
 $$
-\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(if (n? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
-          {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
+\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
+          {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
+          {\filterset{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\botprop{}}}}{\emptyobject{}}
 $$
-By T-Local we know \clj{x} is a \clj{N}.
+By T-Local we know \x{} is a \Number{}.
 $$
-\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
+\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\x{}}{\Number}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
 $$
 By T-App,
-\clj{inc} requires a \clj{Num} argument which we just proved.
+\inc{} requires a \Number{} argument which we just proved.
 $$
-\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(do (if (n? x) nil}\ {\errorvalv{}}{\clj{) (inc x))}}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
+          {\hastype{\doexp
+                     {\ifexp{\appexp{\numberhuh{}}{\x{}}}
+                        {\nil{}}
+                        {\errorvalv{}}}
+                     {\appexp{\inc{}}{\x{}}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
@@
   \end{mathpar}
-  \caption{HMap Terms, Types and Typing Rules}
-  \label{main:figure:hmapsyntax}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{HMap Semantics and Subtyping rules}
-  \label{main:figure:hmapsem}
+  \caption{HMap Terms, Types, Semantics, Subtyping and Typing Rules}
+  \label{main:figure:hmapsyntax}
 \end{figure}
@@
 restrict lookup and extension to keyword keys. The metavariable \mapval{}
-ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.
+ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
+usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
 
-Subtyping for HMaps (figure~\ref{main:figure:hmapsem})
+Subtyping for HMaps (figure~\ref{main:figure:hmapsyntax})
 designate \MapLiteral{} as a common supertype for all HMaps.
@@
 
-Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
+Figure~\ref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
 and \assocliteral{}.
@@
 is, variables, but \emph{subcomponents} of the environment.
-We have already mentioned paths \path{\pathelem{}}{\x{}}
-where \pathelem{} is a sequence of selectors on \x{} called \emph{path elements},
-but we have not yet defined what components of a path can be.
+We briefly mentioned paths, written \path{\pathelem{}}{\x{}}, but we concentrated on
+reasoning only about the variable \x{}---the notation \pathelem{} is a nested sequence of \emph{path elements}
+applied right-to-left to a variable \x{}.
 \citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
@@
 (figure~\ref{main:figure:hmapsyntax}) by
-demonstrating how the first conditional test is checked in example~\ref{example:decleaf}.
-First we unfold the destructuring syntax.
+checking the then branch of the first conditional test is checked in 
+an equivalent version of example~\ref{example:decleaf}.
 \begin{minted}{clojure}
   (fn [m :- Expr]
-    (let [op (:op m)]
-      (if (= :if (:op m))
+    (if (= (get m :op) :if)
       {:op :if, ...}
-        (if ...))))
+      (if ...)))
 \end{minted}
 
-The \classconst{} function is a constant which returns
-the argument's class or \nil{}.
+We do not specifically support \equivliteral{} in our calculus, 
+but on keyword arguments it works identically to \clj{isa?} which we model
+in section~\ref{sec:isaformal}.
+Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
+then \equivapp{\e{}}{\makekw{if}} has the proposition set 
+$$
+{\replacefor{\filterset{\isprop{\Value{\makekw{if}}}{\x{}}}{\notprop{\Value{\makekw{if}}}{\x{}}}}{\object{}}{\x{}}}
+$$
+where substitution reduces to \topprop{} if \object{} = \emptyobject{}.
 
+By T-Abs, checking the body of the function we have the proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
+By T-Local we have
+$$
+\judgement{\propenv{}}{\hastype{m}{\Expr{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{m}
+$$
+and since {\Expr{}} is a union of HMaps with the entry \makekw{op}, we use T-GetHMap to derive
+$$
+\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\object{1}}
+$$
+where \object{1} = {\path{\keype{\makekw{op}}}{m}}. Using our intuitive definition of \equivliteral{} above, we know
+$$
+\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\object{1}}}{\notprop{\Value{\makekw{if}}}{\object{1}}}}{\emptyobject{}}
+$$
+Now we check the then branch with our new assumption.
+$$
+\judgement{\propenvp{}}{\hastype{\curlymap{{\mapvalentrykwkeynoarrow{op}{\makekw{if}}}, ...}}{\Expr{}}}{...}{...}
+$$
+We are interested in what L-Update can derive in our new environment,
+\propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.
+
+With the premises
+\inpropenv{\propenvp{}}{\isprop{\Expr{}}{m}}
+and
+\inpropenv{\propenvp{}}{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}
+we derive
+$$
+\inpropenv{\propenvp{}}{\isprop{\update{\Expr{}}{\Value{\makekw{if}}}{\keype{\makekw{op}}}}{m}}
+$$
+which simplifies to 
+$$
+\inpropenv{\propenvp{}}{\isprop{\HMapp{\mandatoryset{{\mandatoryentrykwnoarrow{op}{\makekw{if}}}, {\mandatoryentrykwnoarrow{test}{\Expr{}}},
+                                       {\mandatoryentrykwnoarrow{then}{\Expr{}}},   {\mandatoryentrykwnoarrow{else}{\Expr{}}}}}
+                                   {\emptyabsent{}}}{m}}
+$$
+
+Our second path element \classpe{} is used in the latent
+object of the constant \classconst{} function. Like Clojure's \clj{class}
+function \classconst{} returns the argument's class or \nil{}
+if passed \nil{}.
 $$
@@
 $$
-
 \begin{mathpar}
@@
 \end{mathpar}
-
-\constantsemfigure{main}
-
-\begin{figure*}
-  $$
-\begin{array}{lllll}
-\updatefigure
-\end{array}
-$$
-\caption{Type Update}
-\label{main:figure:update}
-\end{figure*}
+The full semantics are given in figure~\ref{main:figure:primitivesem}.
+The definition of \updateliteral{} supports various idioms relating to \classpe{}
+which we introduce in section~\label{sec:isaformal}.
 
 \subsection{Java Interoperability and Type Hints}
+
+\begin{figure}[h]
+  \footnotesize
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
+    \\
+
+    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
+    &\mbox{Values} \\
+
+    \tatypesyntax{}\\
+    \typehintenvsyntax{}\\
+    \classtableallsyntax{}
+  \end{altgrammar}
+  $$
+  \begin{mathpar}
+    {\TNewStatic}
+    {\TFieldStatic}
+
+    {\TMethodStatic}
+  \end{mathpar}
+  \begin{mathpar}
+    \BField{}
+
+    \BNew{}
+
+    \BMethod{}
+  \end{mathpar}
+  \begin{mathpar}
+    %\RAbs{}
+%
+%    \RNewElimRefl{}
+%
+%    \RMethodElimRefl{}
+%
+    \RFieldElimRefl{}
+%
+%    \RLet{}
+%
+
+    \RLetHint{}
+  \end{mathpar}
+  \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
+  \label{main:figure:javatyping}
+\end{figure}
+
+In section~\ref{sec:overviewjavainterop} we discussed how Clojure programmers use
+type hints to help the compiler eliminate reflective calls.
+In this section we introduce our model of Java, including syntax
+for unresolved Java
 
@@
 
-\begin{figure*}
-  \footnotesize
-  $$
-  \begin{altgrammar}
-    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
-    &\mbox{Expressions}\\
-
-    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
-    &\mbox{Values} \\
-
-    \tatypesyntax{}\\
-    \typehintenvsyntax{}\\
-    \classtableallsyntax{}
-  \end{altgrammar}
-  $$
-  \caption{Internal Language Reflection Resolution Extensions}
-  \label{main:figure:nonreflectivesyntax}
-\end{figure*}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    %\RAbs{}
-%
-%    \RNewElimRefl{}
-%
-%    \RMethodElimRefl{}
-%
-    \RFieldElimRefl{}
-%
-%    \RLet{}
-%
-
-    \RLetHint{}
-  \end{mathpar}
-\caption{Reflection Elimination %(select rules, figure~\ref{appendix:figure:rewrite} for full rules)
-}
-\label{main:figure:rewrite}
-\end{figure}
-
 %\begin{figure}
@@
 
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \BField{}
-
-    \BNew{}
-
-    \BMethod{}
-  \end{mathpar}
-  \caption{Java Interoperability Semantics}
-\end{figure}
-
-\begin{figure*}
-  \footnotesize
-  \begin{mathpar}
-    {\TNewStatic}
-
-    {\TFieldStatic}
-    {\TMethodStatic}
-  \end{mathpar}
-  \caption{Java Interoperability Typing Rules}
-  \label{main:figure:javatyping}
-\end{figure*}
-
 $$
@@
 
-We now have two languages: an internal language that may contain non-reflective Java calls (figure 
-  \ref{main:figure:nonreflectivesyntax}) 
+We now have two languages: an internal language that may contain non-reflective Java calls (figure~\ref{main:figure:javatyping}) 
   and an external language that resembles Clojure, where we must rely on type-hints to resolve
@@
 \subsection{Multimethod dispatch mechanism: \isaliteral}
+
+\label{sec:isaformal}
 
@@
 
-The definition of \isacompareliteral{} (figure~\ref{main:figure:isaprops}) is deliberately conservative.
+The definition of \isacompareliteral{} (figure~\ref{main:figure:mmsyntax}) is deliberately conservative.
 The first line considers the case where the object of the left argument
@@
 
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \isapropsfigure{}
-  \end{mathpar}
-
-  \center{Takes \emph{object} of the left parameter and \emph{type} of the right parameter.}
-  \caption{Proposition Set for \isaliteral{} expressions}
-  \label{main:figure:isaprops}
-\end{figure}
-
 The second \isacompareliteral{} line handles the equality mode of \isaliteral{}
 with a non-class singleton type as the second argument.
+
+\constantsemfigure{main}
 
@@
 
-\begin{figure*}
+\begin{figure}
   \footnotesize
@@
 \begin{altgrammar}
-  \e{} &::=& \ldots \alt \multimethodexpsyntax{} 
-                &\mbox{Expressions} \\
+  \e{} &::=& \ldots \alt {\createmultiexp {\t{}} {\e{}}} &\mbox{Expressions} \\
+             &\alt& {\extendmultiexp {\e{}} {\e{}} {\e{}}}
+             \alt {\isaapp {\e{}} {\e{}}}\\
   \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
@@
 $$
-\caption{Multimethod Syntax}
-\label{main:figure:mmsyntax}
-\end{figure*}
-
-\begin{figure}
-  \footnotesize
+  \begin{mathpar}
+    \Multisubtyping{}
+  \end{mathpar}
+  \begin{mathpar}
+    \isapropsfigure{}
+  \end{mathpar}
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{Multimethod typing rules}
-  \label{main:figure:mmtyping}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \Multisubtyping{}
-  \end{mathpar}
-  \caption{Multimethod subtyping rules}
-  \label{main:figure:mmsubtype}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
   \getmethodfigure{}
@@
   \end{mathpar}
-  \caption{Multimethod and method resolution semantics}
-  \label{main:figure:mmsemantics}
+\caption{Multimethod Syntax, Subtyping, Typing Rules and Semantics}
+\label{main:figure:mmsyntax}
 \end{figure}
@@
 
-Multimethod semantics are in figure~\ref{main:figure:mmsemantics}.
+Multimethod semantics are in figure~\ref{main:figure:mmsyntax}.
 B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
@@
 
-The typing rules (figure~\ref{main:figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
+The typing rules (figure~\ref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
 where \s{} is the interface type for beta reduction, and \t{} is the type for
@@
 method bodies. T-App handles beta reduction because multimethods are subtypes of functions
-(figure~\ref{main:figure:mmsubtype}).
+(figure~\ref{main:figure:mmsyntax}).
+
+\begin{figure*}
+  $$
+\begin{array}{lllll}
+\updatefigure
+\end{array}
+$$
+\caption{Type Update}
+\label{main:figure:update}
+\end{figure*}
 

### 2015-02-26 overview.tex
use formalism in judgements

@@
 and type checked without further annotations.
-Here is same program in Typed Clojure.
+
+We chose to build on the ideas and implementation
+of Typed Racket to implement a type system targeting Clojure for several reasons.
+Initially, the similarities between Racket and Clojure drew us to
+investigate the effectiveness of repurposing occurrence typing
+for a Clojure type system---both languages share a Lisp heritage,
+similar standard functions 
+(for instance \clj{map}
+in both languages is variable-arity)
+and idioms.
+While Typed Racket is gradually typed and has sophisticated
+dynamic semantics for cross-language interaction, we 
+chose to first implement
+the static semantics
+with the hope to extend Typed Clojure to be gradually typed at a future date.
+Finally,
+Typed Racket's combination of bidirectional checking
+and occurrence typing presents a successful model for 
+type checking dynamically typed programs without compromising
+soundness, where other approaches like success typing~\cite{todo-dialyzer-paper}
+or soft typing~\cite{successtyping-todo}
+fail to prove strong properties about programs
+and have proved too complicated in practice respectively.
+
+Here is above program in Typed Clojure.
 
@@
 Symbols support \emph{metadata}, an arbitrary map of
-data. The syntax \clj{^File x} is a single expression that is the symbol \clj{x} 
+data. The syntax \clj{^File x}, shorthand for \clj{^{:tag File} x}, is a single expression that is the symbol \clj{x} 
 with metadata \clj{{:tag File}}.
 In binding positions like \clj{(fn [^File x] ...)}
-the metadata is preserved in occurrences of \clj{x}.
+the compile-time metadata is preserved in occurrences of \clj{x}
+and utilized by the compiler.
 
@@
 $$
-\clj{(fn [{^File x :x, ^String y :y, :as m}] ...)}
+\clj{(fn [{^File x :x, :as m}] ...)}
 $$ 
-expands to
-\begin{minted}{clojure}
-  (fn [m___0]
-    (let [m m___0,
-          ^File x (:x m),
-          ^String y (:y m)]
-      ...))
-\end{minted}
+expands to a \clj{let} that binds \clj{m} to the argument and 
+\clj{^File x} to \clj{(:x m)}.
 Destructuring is supported similarly by \clj{let}.
@@
 \subsection{Java interoperability}
+\label{sec:overviewjavainterop}
 

### 2015-02-26 formal-model.tex
describe nonreflective syntax

@@
 and 
-\inpropenv{\propenv{}}{\notprop{\nil{}}{\x{}}}
+\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
 to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
@@
 $$
-\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\UnionNilNum}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\Nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
+\judgement{\isprop{\UnionNilNum}{\x{}}}
           {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
@@
 $$
-\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
+\judgement{\isprop{\UnionNilNum}{\x{}}}
           {\hastype{\doexp
@@
 type hints to help the compiler eliminate reflective calls.
-In this section we introduce our model of Java, including syntax
-for unresolved Java
+In this section, first we introduce our model of Java and provide user-facing
+syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
+Then we model the Clojure compiler's compile-time reflection resolution
+algorithm.
+To achieve this we add notation for
+\emph{non-reflective} Java forms that carry sufficient tag information 
+to unambiguously identify a field, method or constructor.
+Then we define a rewrite relation that uses
+type hints to resolve reflection explicitly.
 
-We model
-Java interoperability in a restricted setting without class inheritance,
+We model Java interoperability in a restricted setting without class inheritance,
 overloading or Java Generics.
@@
 
-We extend the external language with Java field lookups and calls to
-methods and constructors.
-These come without typing or evaluation rules:
-Typed Clojure requires the Clojure compiler to statically resolve all Java interactions
-in typed code.
-Type hinted expressions such as \localhintsyntax{} 
-to help the compiler eliminate reflection.
+We extend the syntax with Java field lookups and calls to
+methods and constructors. We model the syntax after the `dot' special
+form 
+These come without typing or evaluation rules---Typed Clojure 
+requires the Clojure compiler to statically resolve all Java interactions
+in typed code, so they are implicitly type errors
+(as demonstrated in section~\ref{sec:overviewjavainterop}).
+We add type hinted syntax for local variables and let-bindings.
+
+Now we model the compiler's reflection-resolution algorithm.
+Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
+The method ${\methodstaticexp {\classhint{1}}
+                             {\overrightarrow {\classhint{i}}}
+                             {\classhint{2}}
+                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
+is a non-reflective call to the \mth{} method on class {\class{1}}, 
+with Java signature 
+$$
+{\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});
+$$
+The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
+calls the field on class {\classhint{1}} with signature
+$$
+{\classhint{2}}\ \fld{};
+$$
+The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
+                               {\class{}} {\overrightarrow{\e{i}}}}
+calls the constructor
+$$
+{\classhint{}}\ ({\overrightarrow{\classhint{i}}});
+$$
 
@@
 infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
-
 %$$
@@
 %$$
-
-We now have two languages: an internal language that may contain non-reflective Java calls (figure~\ref{main:figure:javatyping}) 
-  and an external language that resembles Clojure, where we must rely on type-hints to resolve
-  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).
 

### 2015-02-26 formal-model.tex
finish java formal

@@
 The definition of \updateliteral{} supports various idioms relating to \classpe{}
-which we introduce in section~\label{sec:isaformal}.
+which we introduce in section~\ref{sec:isaformal}.
 
@@
   $$
+ \classtablelookupsyntax{}
   \begin{mathpar}
@@
 %
-    \RFieldElimRefl{}
+%    \RFieldElimRefl{}
 %
@@
 
-    \RLetHint{}
+%    \RLetHint{}
   \end{mathpar}
+  \convertjavatypegrammar{}
   \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
@@
 
-In section~\ref{sec:overviewjavainterop} we discussed how Clojure programmers use
-type hints to help the compiler eliminate reflective calls.
-In this section, first we introduce our model of Java and provide user-facing
+In section~\ref{sec:overviewjavainterop} we discussed the role of type hints
+to help eliminate reflective calls.
+In this section, we introduce our model of Java and provide user-facing
 syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
@@
 algorithm.
-To achieve this we add notation for
-\emph{non-reflective} Java forms that carry sufficient tag information 
-to unambiguously identify a field, method or constructor.
+To achieve this, first we define notation for
+\emph{non-reflective} Java forms that 
+unambiguously call a field, method or constructor.
 Then we define a rewrite relation that uses
 type hints to resolve reflection explicitly.
+Finally we give typing rules that model how
+Typed Clojure interacts with non-reflective calls.
 
-We model Java interoperability in a restricted setting without class inheritance,
+We present Java interoperability in a restricted setting without class inheritance,
 overloading or Java Generics.
@@
 methods and constructors. We model the syntax after the `dot' special
-form 
+form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
+\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
+and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
 These come without typing or evaluation rules---Typed Clojure 
-requires the Clojure compiler to statically resolve all Java interactions
-in typed code, so they are implicitly type errors
+demands statically resolved Java interactions
+in typed code
 (as demonstrated in section~\ref{sec:overviewjavainterop}).
@@
 with Java signature 
-$$
-{\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});
-$$
+${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
 The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
-calls the field on class {\classhint{1}} with signature
-$$
-{\classhint{2}}\ \fld{};
-$$
+calls the field on class {\classhint{1}} with Java signature
+${\classhint{2}}\ \fld{};$.
 The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
@@
 calls the constructor
-$$
-{\classhint{}}\ ({\overrightarrow{\classhint{i}}});
-$$
+${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.
 
@@
 
+To obtain non-reflective calls we define our rewrite relation
 $$
@@
 $$
-rewrites expression \e{} to a possibly-less reflective expression
+which rewrites \e{} to a possibly-less reflective expression
 \ep{} with respect to type hint environment
-\taenv{} and Java class table \ct{} and
+\taenv{} and Java class table \ct{}.
+
+Type hint inference is modelled by
 $$
@@
 $$
-infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
+which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
 %$$
@@
 
+To demonstrate we consider the expression 
+$$
+\fieldexp{getParent}{\hinted{\File}{\x{}}}.
+$$
+where 
+$$
+\ct = \{ \File \mapsto \{ \ctfieldsliteral \mapsto \{\ctfldentry{getMethod}{\String}\} \} \}
+$$
+
+R-FieldElimRefl eliminates a reflective field.
   $$
@@
   $$
+First the target is rewritten \rewrite {\taenv{}} {\hinted{\File}{\x{}}}{\hinted{\File}{\x{}}}
+then we infer the type hint on the new expression \tajudgement {\taenv{}} {\hastype {\hinted{\File}{\x{}}} {\File}}
+We call \fieldtypeliteral{} (figure~\ref{main:figure:javatyping})
+to determine the exact signature of the field:
+\fieldtype{\ct{}}{\File}{getParent}{\File}{\String}
+and return the non-reflective field 
+$$
+\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}.
+$$
 
-For example, R-FieldElimRefl, which rewrites a reflective field call to be non-reflective,
-rewrites the target expression and uses the result to infer a class type hint as the field target.
-\fieldtypeliteral{} uses the inferred type hint to lookup if the inferred class has the appropriate
-field name and returns the actual target and result type hints which we store in the rewritten rule.
+Now we present the typing rules (figure~\ref{main:figure:javatyping}).
+To briefly finish our example, let us assume \propenv{} = {\isprop{\File}{\x{}}},
+then
+T-FieldStatic converts the target Java type to a non-nilable
+type \javatotc {\File} {\File}} and checks the target against it 
+\judgement{\propenv{}}{\hastype{\x{}}{\File}}, and finally converts the return Java type 
+to a nilable type and \javatotc {\String} {\Union{\Nil{}}{\String}}
+to finish with 
+$$
+\judgementtwo{\propenv{}}{\hastype{\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}}{\Union{\Nil{}}{\String}}}.
+$$
 
-We then extract these type hints in T-FieldStatic for typing checking 
-(figure~\ref{main:figure:javatyping}), which converts type hints
-with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
-dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).
+The other rules T-MethodStatic and T-NewStatic work similarly, and vary
+primarily in the choice of nilability in the conversion function---method returns
+are nilable and constructor returns are non-nilable.
 
-Method and constructor calls are similar, but T-MethodStatic and T-NewStatic disallow passing \nil{} as arguments, 
-we assume methods can return \nil{} and we assume constructors cannot.
-
-We rewrite to our internal language before type checking and remaining reflective calls are type errors.
+The operational semantics B-Field, B-New and B-Method simply evaluate their
+arguments and call the relevant JVM operation---section~\label{sec:metatheory}
+states our exact assumptions about each one.
 
@@
 
-{\convertjavatypefigure{figure}{\footnotesize}}
-
 \begin{figure}

### 2015-02-27 formal-model.tex
overview, Thursday evening changes

@@
-\section{A Formal Model of Typed Clojure}
+\section{A Formal Model of \lambdatc{}}
 
@@
 Now that we have demonstrated the core features Typed Clojure
-provides, we present a model that links them all together.
+provides, we link them together in a formal model called
+\lambdatc{}.
 Our presentation will start with a review of
@@
 interleaving presentation of typing rules, operational semantics
-and
-subtyping.
+and subtyping.
 
@@
 and relating them to types.
-\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
+\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
 ``variable \clj{x} is not \clj{nil}''---
@@
 type proposition is called the \emph{object}.
-Every expression has an object that is either \emph{empty}, written \emptyobject{}, which says there is no
-known way to relate this expression to the current environment, or a 
+Every expression has an object that is either \emph{empty}, written \emptyobject{}, 
+which says 
+this expression is not known to evaluate to a particular component
+  of the current environment, or a 
 variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
@@
 
-The third insight is that we
-can use a logic system to learn new information about our programs.
-The key component of the logic system L-Update uses type propositions
-to derive new propositions.
+Given a set of propositions, we can use logical reasoning to derive
+new information about our programs
+with the judgement \inpropenv{\propenv{}}{\prop{}}.
+In addition to the standard rules for the logical connectives, the key
+rule is L-Update, which combines multiple propositions about the same variable.
 $$
@@
 
-The fourth insight is a consequence of
-programmers in dynamic languages
+Finally, this approach allows the type system to track
+programming idioms from 
+dynamic languages
 using implicit type-based reasoning based on the result of
@@
 is 
-a true or false value. The takeaway is that every expression has
-two extra pieces of information relevant to the logic system: a `then' proposition
+true or false. 
+To express this in the type system, every expression 
+is described by two propositions: a `then' proposition
 for when it reduces to a true value, and an `else' proposition
@@
   \v{} &::=&          \singletonmeta{}
+                      \alt {\num{}}
                       \alt {\const{}}
@@
 `then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}. We sometimes write 
+and object \object{}. We write 
 {\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.
@@
 Values include booleans, \nil{}, class literals, keywords, 
+numbers,
 constants and closures. 
 Value environments map local bindings to values.
-Closures contain the value environment \openv{} when the function was created,
-which is used during beta reduction to ensure the abstraction respects lexical scoping.
-\false{} and \nil{} are logically false, otherwise values are true.
 
@@
 and class instances. 
-We write the class literal \Booleanlong{} as \Boolean{} and \Keywordlong{} as \Keyword{}.
+We abbreviate \Booleanlong{} as \Boolean{}, \Keywordlong{} as \Keyword{}
+and \NumberFull{} and \Number{}.
 The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
@@
 Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
-may refer to the parameter identifier. They are latent because they are instantiated with the
+may refer to the function argument.
+%Latent means they are relevant when the function is applied rather than evaluated.
+They are latent because they are instantiated with the
 actual object of the argument in applications before they are used in the proposition environment.
 
-Figure~\ref{main:figure:typingrules} contains the standard typing rules.
-The first interesting rule is T-If. The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
+Figure~\ref{main:figure:typingrules} contains the core typing rules.
+The key rule for reasoning about conditional control flow is
+T-If. The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
 are used as assumptions in the then and else branch respectively.
@@
 which implies \thenprop{\prop{3}} is true. 
-The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} for
-similar reasons, but instead for when each branch is a false value.
-The second interesting rule is T-Local, where we actually utilise the logic system
-to extract a type for a local binding. This rule is why L-Update is crucial---T-Local
-looks for positive type propositions in \propenv{} with the same object as the binding.
-By subsumption we can choose the most specific positive proposition and thus
-T-Local can return the most specific type inferred by the logic system.
+The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
+similarly.
+The T-Local rule connects the type system to the proof system over type propositions
+to derive a type for a variable.
+Using this rule, the type system can then appeal to L-Update to refine the type
+assigned to \x{}.
 
-We can now demonstrate how to check example~\ref{example:conditionalflow}.
-We abbreviate \NumberFull{} as \Number{} and \Booleanlong{} as \Boolean{}.
-The test is checked with T-App, 
+We are now equipped to type check 
+example~\ref{example:conditionalflow}.
+The test expression uses T-App, 
 $$
@@
 $$
-\numberhuh{} has type
-$$
+since \numberhuh{} has type
 {\ArrowOne{\x{}}{\Top}{\Boolean}
         {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
+      and \x{} has object \x{}.
+
+Now we check both branches using the extended proposition environment as specified by T-If.
 $$
-By T-Local.
-$$
-\judgement{\isprop{\Number}{\x{}}}{\hastype{\x{}}{\Number{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
-$$
-Now we check both branches.
-$$
-\judgement{\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{{\isprop{\UnionNilNum}{\x{}}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
 $$
-\judgement{\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{{\isprop{\UnionNilNum}{\x{}}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
 $$
-\inc{} has type
-$$
-{\ArrowOne{\x{}}{\Number}{\Number}
-        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
-$$
-We can now check the conditional with T-If.
-$$
-\judgement{\isprop{\Number}{\x{}}}{\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\zeroliteral{}}}{\Number}}{\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}
-$$
-Finally the function can be checked with T-Abs
-$$
-\judgement{}{\hastype{\abs{\x{}}{\UnionNilNum}{\ ...}}
-                                             {\ArrowOne{\x{}}{\UnionNilNum}{\Number}
-        {\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
-  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
-$$
+%\inc{} has type
+%$$
+%{\ArrowOne{\x{}}{\Number}{\Number}
+%        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
+%$$
+%We can now check the conditional with T-If.
+%$$
+%\judgement{\isprop{\Number}{\x{}}}{\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\zeroliteral{}}}{\Number}}{\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}
+%$$
+%Finally the function can be checked with T-Abs
+%$$
+%\judgement{}{\hastype{\abs{\x{}}{\UnionNilNum}{\ ...}}
+%                                             {\ArrowOne{\x{}}{\UnionNilNum}{\Number}
+%        {\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
+%  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+%$$
 
-The operational semantics for our language standard defined in big-step style, 
-The notable different is \nil{} is a false value, which require changes 
-to the semantics of \ifliteral{}:
+\paragraph{Operational semantics} We define the dynamic semantics for \lambdatc{}
+in a big-step style using an environment, following~\citet{TF10}.
+We include both errors and a \wrong{} value, which is provably ruled out by the
+type system.
+The main judgement is \opsem{\openv{}}{\e{}}{\definedreduction{}}
+which states that \e{} evaluates to answer \definedreduction{} in environment
+\openv{}.
+A notable difference is \nil{} is a false value, which affects the
+semantics of \ifliteral{}:
 
@@
 
-For example,
-B-IfTrue evaluates \e{1} in the environment \openv{} with the result \v{1}.
-If \v{1} is not \nil{} or \false{} then the conditional returns the result of evaluating the then branch \e{2}
-in \openv{}.
-
 Subtyping (figure~\ref{main:figure:subtyping}) 
-is a reflexive and transitive relation with top type \Top. All objects are subtypes
-of \emptyobject{}.
+is a reflexive and transitive relation with top type \Top. 
 Singleton types are instances of their respective classes---boolean singleton types
@@
 Our main insight is as follows: 
-if the left argument reduces to a value, then it is either logically true or false.
-If we learn the same information in both cases then we can use it as an assumption to check the right argument.
+if the first subexpression in a sequence reduces to a value, then it is either true or false.
+If we learn some proposition in both cases then we can use that proposition as an assumption to check the second subexpression.
 T-Do formalises this intuition.
@@
 
-The introduction of errors, specifically as an expression that is neither true nor false, 
+The introduction of errors, 
+which do not evaluate to either
+a true or false value,
 makes our insight interesting.
@@
 
-Now we can check example~\ref{example:doexception}.
-We check the test with T-App:
+Recall example~\ref{example:doexception}.
+\begin{minted}{clojure}
+      (do (if (number? x) nil (throw (Exception.)))
+          (inc x))
+\end{minted}
+
+As before, checking \appexp{\numberhuh{}}{\x{}} allows us to use the proposition \isprop{\Number}{\x{}}
+when checking the then branch.
+
+By T-Nil and subsumption we can propagate this  information to both propositions.
 $$
-\judgement{\isprop{\UnionNilNum}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\Nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
-By T-Nil and subsumption we derive the same information in both propositions.
+Furthermore, using T-Error
+and subsumption we can conclude anything in the else branch.
 $$
-\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\Nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
-By T-Error we can assume anything about the return of an error.
-$$
-\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
-$$
-Using the above as premises to T-If we learn {\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}
-in both propositions.
+Using the above as premises to T-If we conclude that if the first
+expression in the \doliteral{} evaluates successfully, \isprop{\Number}{\x{}} must be true.
 $$
@@
           {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
-          {\filterset{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\botprop{}}}}{\emptyobject{}}
+          {\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
-By T-Local we know \x{} is a \Number{}.
-$$
-\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\x{}}{\Number}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
-$$
-By T-App,
-\inc{} requires a \Number{} argument which we just proved.
-$$
-\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
-$$
-Finally we conclude with T-Do using the conditional and increment expressions as premises.
-$$
-\judgement{\isprop{\UnionNilNum}{\x{}}}
-          {\hastype{\doexp
-                     {\ifexp{\appexp{\numberhuh{}}{\x{}}}
-                        {\nil{}}
-                        {\errorvalv{}}}
-                     {\appexp{\inc{}}{\x{}}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
-$$
+We can now use \isprop{\Number}{\x{}} in the environment to check the second subexpression
+{\appexp{\inc{}}{\x{}}}, completing the example.
 
@@
  \classtablelookupsyntax{}
+  \convertjavatypegrammar{}
   \begin{mathpar}
@@
   \end{mathpar}
-  \convertjavatypegrammar{}
   \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}

### 2015-02-27 overview.tex
describe nonreflective syntax
overview, Thursday evening changes

@@
 address in Typed Clojure.
+
+\subsection{Clojure and Typed Clojure Primer}
+
+We introduce Clojure's syntax.
+Parenthesis indicate \emph{applications}, square brackets
+are syntax for literal \emph{vectors} and curly braces delimit literal
+\emph{hash-maps}.
+\clj{nil} is exactly Java's \java{null} and is one of two logically
+false values (the other is \clj{false}).
+\emph{Symbols} begin with an alphabetic character.
+A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
+The \clj{get} function takes a hash-map and a key returns the value
+associated with the key in the map, or returns \clj{nil} if there is none.
+Keywords are functions that take a map and applies \clj{get} to the map
+and itself.
+\clj{inc} and \clj{dec} increment and decrement numbers respectively.
+\clj{assoc} and \clj{dissoc} associate and dissociate entries
+from immutable maps.
+The \clj{=} function returns true if all its arguments are pointer equal.
+Everything is an instance of some class
+except for \clj{nil}. Predicates like \clj{number?} are equivalent to
+a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.
+
+Symbols support \emph{metadata}, an arbitrary map of
+data. The syntax \clj{^File x}, shorthand for \clj{^{:tag File} x}, is a single expression that is the symbol \clj{x} 
+with metadata \clj{{:tag File}}.
+In binding positions like \clj{(fn [^File x] ...)}
+the compile-time metadata is preserved in occurrences of \clj{x}
+and utilized by the compiler.
+
+\emph{Commas} are always whitespace.
+
+Clojure uses \emph{namespaces}, declared with the
+\clj{ns} form, to manage top-level \emph{vars} and
+class imports.
+Example~\ref{example:conditionalflow}
+is a regular Clojure file compiled with
+the Clojure compiler using \clj{:require} to declare a
+runtime dependency on Typed Clojure's core namespace
+\clj{clojure.core.typed}.
+Programmers import typed versions of certain constructs as needed, like
+a \clj{fn} variant that supports annotations,
+and use the provided \clj{check-ns} function to type check the current namespace
+when convenient.
+
+Java methods and fields are called in prefix notation
+like \clj{(.method target args*)} and \clj{(.field target)} respectively,
+with method and field names prefixed with a dot and methods taking some number of arguments.
+The canonical form for methods is \clj{(. target (method args*))} 
+and for fields \clj{(. target field)} using the dot special form, which we utilise occasionally
+for clarity.
+Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}
+or equivalently \clj{(new class args*)} with the \clj{new} special form.
+
+The \clj{if} special form is a standard two-armed conditional.
+The \clj{do} special form sequentially evaluates any number of expressions
+left-to-right, returning the last expression or \nil{}.
+The \clj{when} macro is a one-armed \clj{if} with an implicit \clj{do}
+around arguments after the test, and returns \clj{nil} if the test fails.
+The \clj{throw} special form is exactly Java's \java{throw}, which takes an
+instance of \clj{Throwable} and throws it as an exception.
+The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
+and any number of expressions in an implicit \clj{do} for the body of the function.
+The \clj{let} macro takes a vector of bindings, a flat vector of binding-value pairs,
+and scopes them in the body, an implicit \clj{do}.
+Scoping works like \clj{let*} in Scheme---that is values in the binding vector can refer
+to variables previously introduced in the same binder.
+The typed
+variant of \clj{fn} supports an optional keyword \clj{:-} and type following
+each parameter to declare their static types, and similarly after
+the parameter vector to declare the return type.
+
+\emph{Map destructuring} is binding position syntax for pattern matching.
+Parameters are replaced with a map containing entries of symbols to keywords that bind the symbols
+to lookups on that keyword with the current argument---the \clj{:as} option
+can follow which binds a symbol to the current argument. For example
+$$
+\clj{(fn [{^File x :x, :as m}] ...)}
+$$ 
+expands to a \clj{let} that binds \clj{m} to the argument and 
+\clj{^File x} to \clj{(:x m)}.
+
+In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
+Quoted maps like \clj{'{:op ':const, :val Num}} are HMap types representing
+maps with known keyword entries with values of the associated type
+The keys are not quoted because only keyword keys
+are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
+any number of types.
+
+The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
 
@@
 
-\subsection{Clojure and Typed Clojure Primer}
-
-We introduce Clojure's syntax.
-Parenthesis indicate \emph{applications}, square brackets
-are syntax for literal \emph{vectors} and curly braces delimit literal
-\emph{hash-maps}.
-\clj{nil} is exactly Java's \java{null} and is one of two logically
-false values (the other is \clj{false}).
-\emph{Symbols} begin with an alphabetic character.
-A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
-The \clj{get} function takes a hash-map and a key returns the value
-associated with the key in the map, or returns \clj{nil} if there is none.
-Keywords are functions that take a map and applies \clj{get} to the map
-and itself.
-\clj{inc} and \clj{dec} increment and decrement numbers respectively.
-\clj{assoc} and \clj{dissoc} associate and dissociate entries
-from immutable maps.
-The \clj{=} function returns true if all its arguments are structurally equal---in this
-paper we use it only for pointer equality.
-Everything is an instance of some class
-except for \clj{nil}. Predicates like \clj{number?} are equivalent to
-a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.
-
-Symbols support \emph{metadata}, an arbitrary map of
-data. The syntax \clj{^File x}, shorthand for \clj{^{:tag File} x}, is a single expression that is the symbol \clj{x} 
-with metadata \clj{{:tag File}}.
-In binding positions like \clj{(fn [^File x] ...)}
-the compile-time metadata is preserved in occurrences of \clj{x}
-and utilized by the compiler.
-
-\emph{Commas} are always whitespace.
-
-Clojure uses \emph{namespaces}, declared with the
-\clj{ns} form, to manage top-level \emph{vars} and
-class imports.
-Example~\ref{example:conditionalflow}
-is a regular Clojure file compiled with
-the Clojure compiler using \clj{:require} to declare a
-runtime dependency on Typed Clojure's core namespace
-\clj{clojure.core.typed}.
-Programmers import typed versions of certain constructs as needed, like
-a \clj{fn} variant that supports annotations,
-and use the provided \clj{check-ns} function to type check the current namespace
-when convenient.
-We omit \clj{ns} forms for the rest of the examples.
-\footnote{See <this> github repo for full working examples}
-
-Java methods and fields are called in prefix notation
-like \clj{(.method target args*)} and \clj{(.field target)} respectively,
-with method and field names prefixed with a dot and methods taking some number of arguments.
-The canonical form for methods is \clj{(. target (method args*))} 
-and for fields \clj{(. target field)} using the dot special form, which we utilise occasionally
-for clarity.
-Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}
-or equivalently \clj{(new class args*)} with the \clj{new} special form.
-
-The \clj{if} special form is a standard two-armed conditional.
-The \clj{do} special form sequentially evaluates any number of expressions
-left-to-right, returning the last expression or \nil{}.
-The \clj{when} macro is a one-armed \clj{if} with an implicit \clj{do}
-around arguments after the test, and returns \clj{nil} if the test fails.
-The \clj{throw} special form is exactly Java's \java{throw}, which takes an
-instance of \clj{Throwable} and throws it as an exception.
-The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
-and any number of expressions in an implicit \clj{do} for the body of the function.
-The \clj{let} macro takes a vector of bindings, a flat vector of binding-value pairs,
-and scopes them in the body, an implicit \clj{do}.
-Scoping works like \clj{let*} in Scheme---that is values in the binding vector can refer
-to variables previously introduced in the same binder.
-The typed
-variant of \clj{fn} supports an optional keyword \clj{:-} and type following
-each parameter to declare their static types, and similarly after
-the parameter vector to declare the return type.
-
-\emph{Destructuring} is binding position syntax for pattern matching.
-A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
-to lookups on that keyword with the current argument, optionally terminated by
-\clj{:as} and a symbol, which binds the symbol to the current argument. For example
-$$
-\clj{(fn [{^File x :x, :as m}] ...)}
-$$ 
-expands to a \clj{let} that binds \clj{m} to the argument and 
-\clj{^File x} to \clj{(:x m)}.
-Destructuring is supported similarly by \clj{let}.
-
-In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
-Quoted maps like \clj{'{:op ':const, :val Num}} are HMap types representing
-maps with known keyword entries with values of the associated type
-The keys are not quoted because only keyword keys
-are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
-any number of types.
-
-The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
-
 \subsection{Exceptional control flow}
@@
 Clojure programmers rely on \emph{exceptions}
-to assert type-related invariants.
+to assert type-related invariants, such as preventing
+null-pointer exceptions.
 
@@
 
-To check this example, we note that
-occurrence typing already provides
-valuable information about the conditional expression---namely that it cannot return logically true and if it returns logically false 
-then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
-We use this assumption to check \clj{(inc x)} with an automatic
-guarantee that a null-pointer exception is impossible
-(modelled formally in section~\ref{sec:doformal} and proved
-in section~\ref{sec:metatheory}).
+To check this example,
+occurrence typing 
+automatically
+assumes
+\clj{x} is a number when checking the second \clj{do} subexpression
+based on the first subexpression.
+\footnote{We omit \clj{ns} forms for the rest of the examples, see <this> github repo for full working examples}
+We model this formally (section~\ref{sec:doformal}) and prove
+null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).
 
@@
 
-\begin{Code}
 \begin{exmp}
@@
 \end{exmp}
-\end{Code}
 
-Here we define a type abbreviation with \clj{defalias} called \clj{Expr}
+Here \clj{defalias} defines a type abbreviation called \clj{Expr}
 that describes the structure of a recursively-defined AST as a union of HMaps.
@@
 
-The next example is a real Typed Clojure program written by CircleCI, who maintain a large production
-installation of Clojure (see section~\ref{sec:casestudy} for a case study).
+The next example is extracted from a Typed Clojure production system at CircleCI
+(section~\ref{sec:casestudy} presents a case study).
 
-\begin{Code}
 \begin{exmp}
@@
 \end{exmp}
-\end{Code}
 
 \clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
-removing the raw \clj{:private-key} entry and associating an encrypted private key
+dissociating the raw \clj{:private-key} entry and associating an encrypted private key
 as \clj{:enc-private-key}.
@@
 
-\begin{Code}
 \begin{exmp}
-\inputminted[firstline=10,lastline=23]{clojure}{code/demo/src/demo/key2.clj}
+\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
 \end{exmp}
-\end{Code}
 
+Next, 
 HMaps interact with occurrence typing in interesting ways.
 
-\begin{Code}
 \begin{exmp}
@@
 \end{exmp}
-\end{Code}
 
@@
 call constructors, methods and access fields.
-Method calls use prefix syntax, for example this call to 
-\clj{java.io.File}'s \clj{getParent} method.
 
@@
 
-We have a problem: Clojure's compiler does not know which method to call.
-Instead, \clj{f} is inspected at runtime and an appropriate method is selected based
-and then invoked by \emph{Java reflection}.
-Unfortunately, reflection is slow and Clojure's algorithm for
-choosing methods at runtime is undefined in some cases.
-
-To resolve reflective calls, Clojure supports \emph{type hints}.
+Unlike Java, Clojure is dynamically typed. We have no type information about \clj{f}
+but we still need to pick a method to call. The Clojure compiler delegates
+the choice to runtime using \emph{Java reflection}.
+Unfortunately reflection is slow and unpredictable, so Clojure supports \emph{type hints}
+to help eliminate it where possible,
 
@@
 
-The Clojure compiler uses the type hint on \clj{f}
-to statically resolve the method call to the \clj{java.io.File}
-method with the Java signature
+The Clojure compiler uses the type hint
+to statically resolve 
+the method call to the
+\mintinline{java}{public String getParent()}
+method of \clj{java.io.File}.
+The method is unambiguous but type checking fails---Typed Clojure
+considers \clj{f} to be of type \clj{Any}, which is unsafe 
+to use as the target of even a resolved method.
+We annotate the function parameter with type \clj{File}.
 
-\begin{minted}{java}
-  public String getParent()
+\begin{minted}{clj}
+  (fn [f :- File] (.getParent f)) ;; type error
 \end{minted}
 
-Typed Clojure ignores type hints, so to check this we must
-annotate the parameter's static type.
-
-\begin{minted}{clj}
-  (fn [f :- File] (.getParent f))
-\end{minted}
-
-However Typed Clojure disallows reflection in typed code, so we
-must add a type hint.
+Typed Clojure disallows reflection in typed code
+so we must add back the type hint to obtain a well-typed
+expression.
 
@@
 
-Now we have a well-typed expression free of reflection.
+The type hinting system and Typed Clojure's static type checking
+are separate, the latter predating the former by several years.
+The interaction between them is often not as obvious, for example type hints
+do not need to take \clj{nil} into account, while Typed Clojure guarantees
+null-pointer exceptions are impossible.
+
+\begin{exmp}
+\inputminted[firstline=18,lastline=19]{clojure}{code/demo/src/demo/parent3.clj}
+\end{exmp}
 
@@
 \begin{exmp}
-\inputminted[firstline=12,lastline=13]{clojure}{code/demo/src/demo/parent3.clj}
+\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
 \end{exmp}
@@
 \end{exmp}
-\end{Code}
 
-Notice a type hint is used in an argument position to help choose the \java{File(String pathname)}
-constructor. By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
+By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
 but can be configured globally for particular positions if needed.
-
-Finally, Typed Clojure guarantees typed code cannot throw a null-pointer exception
-by attempting to dereference \clj{nil}. This can happen in several ways, including
-calling a method or a field on \clj{nil}.
-The next example calls a method on once-nullable local binding.
-(\clj{ann} annotates a var with an expected type).
-
-\begin{exmp}
-\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
-\end{exmp}
-
-The test on \clj{f} is essential to check this example---only then can Typed Clojure
-prove null-pointer exceptions are impossible.
 
@@
 function and methods. Multimethods are created with {\clj{defmulti}}.
+\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
+The multimethod \clj{path} has type \clj{[Any -> String]}, an initially empty \emph{dispatch table}
+and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
+There are currently no methods so \clj{path} will fail at runtime.
 \begin{minted}{clojure}
-(ann rep [Any -> String])
-(defmulti rep class)
-\end{minted}
-\clj{rep} is a multimethod of type \clj{[Any -> String]} with an empty \emph{dispatch table}
-and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
-There are no methods to dispatch to so invocations of \clj{rep} will fail at runtime.
-\begin{minted}{clojure}
-  (rep :a) ;=> IllegalArgumentException ...
+  (path "dir/a") ;=> IllegalArgumentException ...
 \end{minted}
 
-Methods are installed via {\clj{defmethod}}.
-\begin{minted}{clojure}
-(defmethod rep Keyword [x] (str (name x)))
-\end{minted}
-We extend \clj{rep}'s dispatch table, mapping
-the \emph{dispatch value} \clj{Keyword} to the function
-\clj{(fn [x] (str (name x)))}. The call \clj{(rep arg)}
-uses the value of \clj{(isa? (disp-fn arg) disp-val)}
-on each method entry to pick a winner \clj{mth}, and finally returns \clj{(mth arg)}.
-\clj{isa?} is a subclassing check when provided with classes
+We can use {\clj{defmethod}} to add a method to \clj{path}.
+\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
+Now \clj{path}'s dispatch table contains a method mapping
+the \emph{dispatch value} \clj{String} to the function
+\clj{(fn [x] x)}. The call \clj{(path "dir/a")}
+now dispatches to the method because
+\clj{(isa? (class "dir/a") String)}
+is true, and finally returns \clj{((fn [x] x) "dir/a")}.
+\clj{isa?} is a subclassing check when provided with classes,
+otherwise it is an equality check.
 \begin{minted}{clojure}
   (isa? Keyword Object) ;=> true
-  (isa? Keyword Number) ;=> false
-\end{minted}
-otherwise an equality check.
-\begin{minted}{clojure}
   (isa? :a :a) ;=> true
-  (isa? :a 1) ;=> false
 \end{minted}
 
-For example
-\clj{(rep :a)}
-picks the \clj{Keyword} method because
-\clj{(isa? (class :a) Keyword)} returns true,
-and finally returns \clj{"a"}, the value of
-\clj{((fn [x] (str (name x))) :a)}.
-We give the full definition of \clj{rep}.
-
-\begin{Code}
 \begin{exmp}
@@
 \end{exmp}
-\end{Code}
-Typed Clojure does not statically ensure multimethod calls dispatch successfully---
-\clj{(rep "a")} type checks but throws a runtime error.
+
+Typed Clojure does not statically ensure multimethod calls dispatch successfully---\clj{(path :a)} 
+type checks but throws a runtime error.
 
@@
 is our dispatch function.
-\begin{Code}
+
 \begin{exmp}
@@
 \end{exmp}
-\end{Code}
-The destructuring syntax \clj{(fn [{:keys [val] :as m}] ...)} binds
-\clj{m} to the value of the first argument, and \clj{val} to \clj{(:val m)}.
 
@@
 \begin{minted}{clojure}
-  (isa? [Keyword :a] [Object :a]) ;=> true
   (isa? [Keyword Keyword] [Object Object]) ;=> true
-  (isa? [Keyword Object] [Object Object]) ;=> false
 \end{minted}
 
-Now multiple dispatch is possible---we dispatch on the class of both
-arguments simultaneously by defining a dispatch function that returns
-a vector containing the classes of the arguments.
-\begin{Code}
+Now we can simulate multiple dispatch by dispatching on
+a vector containing the class of both arguments.
+
 \begin{exmp}
-\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+\inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
 \end{exmp}
-\end{Code}
-The dispatch values are also vectors---the first method is picked
-when the left argument is a \clj{Number} and the right is a \clj{Keyword},
-and is thus safe to increment and extract its name respectively.
-The dispatch value \clj{:default} specifies a default
-method if no preferred method is found.
 
+We call three different \clj{File} constructors, each known at compile-time
+via type hints.
 No extra annotations are needed to follow type-directed control flow
@@
 and use of HMaps.
-\begin{Code}
+
 \begin{exmp}
@@
 \end{exmp}
-\end{Code}
+
 \clj{PayLoad} is either a HMap containing a file

### 2015-02-27 metatheory.tex
theorem
NPE theorem
fix appendix vs. main labelling
formatting, remove/inline figures
finish overview
formatting examples
overview

@@
 \section{Metatheory}
+\label{sec:metatheory}
 
-Our proof for type soundness is similar to~\cite{TF10}. We add
-errors and prove well-typed programs can never go wrong.
+Our proof for type soundness is similar to~\citet{TF10}. We add
+errors and \wrong{} value and prove well-typed programs
+rule out wrong programs.
 
@@
 
-{\javaassumptionsall}
+{\javanewassumption{main}}
 
-For readability we define logical truth in Clojure.
+%For readability we define logical truth in Clojure.
 
-{\istruefalsedefinitions}
+%{\istruefalsedefinitions{main}}
 
@@
 ensures that occurrence typing does not claim the outer 
-\mintinline{clojure}{x} is a keyword.
+\x{} is a keyword.
 
@@
 
-{\consistentwithdefinition}
+{\consistentwithonlydef{main}}
 
@@
 
-\begin{lemma}\label{lemma:soundness}
+\begin{lemma}\label{main:lemma:soundness}
 
@@
     By induction on the derivation of the typing judgement. 
+    (Full proof given as lemma~\ref{appendix:lemma:soundness}).
   \end{proof}
@@
 
-% Uses above lemma:soundness label ^^^
-{\soundnesstheorem}
+We can now state our soundness theorem.
 
-{\wrongtheorem}
+{\soundnesstheoremnoproof{main}}
+
+{\wrongtheoremnoproof{main}}
+
+Null-pointer exceptions are ruled-out in typed code.
+
+{\nilinvoketheoremnoproof{main}}

### 2015-02-27 overview.tex
overview

@@
 
-We introduce Clojure's syntax.
-Parenthesis indicate \emph{applications}, square brackets
-are syntax for literal \emph{vectors} and curly braces delimit literal
-\emph{hash-maps}.
-\clj{nil} is exactly Java's \java{null} and is one of two logically
-false values (the other is \clj{false}).
+We now review the basics of the Clojure subset we use in the paper.
+
+\paragraph{Syntax} Parentheses indicate \emph{applications}, brackets
+delimit
+\emph{vectors}, braces
+delimit
+\emph{hash-maps}
+and double quotes delimit Java \emph{strings}.
+\clj{nil} is exactly Java's \java{null}.
 \emph{Symbols} begin with an alphabetic character.
-A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
-The \clj{get} function takes a hash-map and a key returns the value
-associated with the key in the map, or returns \clj{nil} if there is none.
-Keywords are functions that take a map and applies \clj{get} to the map
+A colon prefixed symbol like \clj{:a} is a \emph{keyword}.
+\emph{Commas} are always whitespace.
+
+\paragraph{Core forms} \clj{if} is a standard two-armed conditional.
+\clj{do} sequences two expressions returning the latter.
+\clj{throw} corresponds to Java's \java{throw}.
+\clj{fn} creates a function from a parameter vector and a body expression.
+\clj{let} takes a flat binding vector (which can refer to previous bindings like Scheme's \mintinline{scheme}{let*})
+and a body expression.
+\clj{def} binds a top-level \emph{var}.
+\clj{defn} combines \clj{def} and \clj{fn}.
+
+\paragraph{Semantics} \clj{get} applied to a hash-map and a key returns the value
+associated with the key in the map, or \clj{nil} if there is none.
+Keywords are functions that apply \clj{get} to their argument
 and itself.
-\clj{inc} and \clj{dec} increment and decrement numbers respectively.
+\clj{inc} and \clj{dec} increment and decrement numbers;
 \clj{assoc} and \clj{dissoc} associate and dissociate entries
 from immutable maps.
-The \clj{=} function returns true if all its arguments are pointer equal.
-Everything is an instance of some class
-except for \clj{nil}. Predicates like \clj{number?} are equivalent to
-a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.
+The \clj{=} function returns true if both arguments are equal.
+\clj{number?} is
+a Java \java{instanceof} test of \clj{java.lang.Number}.
+\clj{nil} and \clj{false} are the false values.
 
-Symbols support \emph{metadata}, an arbitrary map of
-data. The syntax \clj{^File x}, shorthand for \clj{^{:tag File} x}, is a single expression that is the symbol \clj{x} 
-with metadata \clj{{:tag File}}.
-In binding positions like \clj{(fn [^File x] ...)}
-the compile-time metadata is preserved in occurrences of \clj{x}
-and utilized by the compiler.
+\paragraph{Namespaces} \clj{ns}
+declares a
+\emph{namespace}
+which manage var and class imports. 
+\clj{:require} takes a number of namespace dependencies and \clj{:refer}
+brings vars into scope by name.
+\clj{:refer-clojure} \clj{:exclude} unmaps core vars from the current namespace.
 
-\emph{Commas} are always whitespace.
-
-Clojure uses \emph{namespaces}, declared with the
-\clj{ns} form, to manage top-level \emph{vars} and
-class imports.
-Example~\ref{example:conditionalflow}
-is a regular Clojure file compiled with
-the Clojure compiler using \clj{:require} to declare a
-runtime dependency on Typed Clojure's core namespace
-\clj{clojure.core.typed}.
-Programmers import typed versions of certain constructs as needed, like
-a \clj{fn} variant that supports annotations,
-and use the provided \clj{check-ns} function to type check the current namespace
-when convenient.
-
-Java methods and fields are called in prefix notation
+\paragraph{Java Interoperability} Java methods and fields have prefix notation
 like \clj{(.method target args*)} and \clj{(.field target)} respectively,
 with method and field names prefixed with a dot and methods taking some number of arguments.
-The canonical form for methods is \clj{(. target (method args*))} 
-and for fields \clj{(. target field)} using the dot special form, which we utilise occasionally
-for clarity.
-Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}
-or equivalently \clj{(new class args*)} with the \clj{new} special form.
+Constructors are called with the class name with a dot suffix and arguments like \clj{(class. args*)}.
 
-The \clj{if} special form is a standard two-armed conditional.
-The \clj{do} special form sequentially evaluates any number of expressions
-left-to-right, returning the last expression or \nil{}.
-The \clj{when} macro is a one-armed \clj{if} with an implicit \clj{do}
-around arguments after the test, and returns \clj{nil} if the test fails.
-The \clj{throw} special form is exactly Java's \java{throw}, which takes an
-instance of \clj{Throwable} and throws it as an exception.
-The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
-and any number of expressions in an implicit \clj{do} for the body of the function.
-The \clj{let} macro takes a vector of bindings, a flat vector of binding-value pairs,
-and scopes them in the body, an implicit \clj{do}.
-Scoping works like \clj{let*} in Scheme---that is values in the binding vector can refer
-to variables previously introduced in the same binder.
-The typed
-variant of \clj{fn} supports an optional keyword \clj{:-} and type following
-each parameter to declare their static types, and similarly after
-the parameter vector to declare the return type.
+\paragraph{Metadata} Symbols support \emph{metadata}---the syntax \clj{^File x}
+is a single expression of a symbol \clj{x} with metadata \clj{{:tag File}}.
+In binding positions like \clj{(fn [^File x] ...)}
+syntactic occurrences preserve metadata.
 
-\emph{Map destructuring} is binding position syntax for pattern matching.
-Parameters are replaced with a map containing entries of symbols to keywords that bind the symbols
-to lookups on that keyword with the current argument---the \clj{:as} option
-can follow which binds a symbol to the current argument. For example
-$$
+\paragraph{Destructuring} \emph{Map destructuring} is binding position syntax for pattern matching.
+Parameters are replaced with maps of symbols to keywords that bind the symbols
+to lookups on that keyword, optionally terminated by \clj{:as}
+which aliases the parameter. For example
 \clj{(fn [{^File x :x, :as m}] ...)}
-$$ 
-expands to a \clj{let} that binds \clj{m} to the argument and 
+expands to a \clj{let} binding \clj{m} to the first argument and 
 \clj{^File x} to \clj{(:x m)}.
 
-In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
-Quoted maps like \clj{'{:op ':const, :val Num}} are HMap types representing
-maps with known keyword entries with values of the associated type
-The keys are not quoted because only keyword keys
-are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
+\paragraph{Static type syntax} Single quoted keywords are singleton types containing just the keyword.
+Single quoted maps mapping keywords to types 
+are HMap types representing
+maps with known keyword entries with values of the associated type.
+\clj{U} creates an \emph{untagged union} type, which can take
 any number of types.
 
+\paragraph{Typed Clojure Macros}
 The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
+The typed \clj{fn} has optional annotations by following a parameter
+or binding vector by \clj{:-} and a type.
 
@@
 This function takes a value that is either \emph{\#f} % mintinline really hates #
-or a number, represented by an untagged \emph{union type}.
+or a number, represented by an \emph{untagged} union type.
 The `then' branch has an implicit invariant
@@
 \end{exmp}
+
+This is a regular Clojure file compiled with
+the Clojure compiler. It has a
+runtime dependency 
+on
+\clj{clojure.core.typed},
+Typed Clojure's core namespace.
+Programmers import typed versions of `typed' macros and qualified types to ease annotation.
+The \clj{check-ns} function type checks the current namespace.
 
@@
 
+\begin{exmp}
 \inputminted[firstline=10,lastline=10]{clojure}{code/demo/src/demo/parent3.clj}
+\end{exmp}
 
@@
 \begin{minted}{clojure}
-  (path "dir/a") ;=> IllegalArgumentException ...
+  (path (File. "dir/a"));=> IllegalArgumentException
 \end{minted}
@@
 \inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
-Now \clj{path}'s dispatch table contains a method mapping
-the \emph{dispatch value} \clj{String} to the function
-\clj{(fn [x] x)}. The call \clj{(path "dir/a")}
-now dispatches to the method because
+Now the dispatch table maps
+the \emph{dispatch value} \clj{File} to the function
+\clj{(fn [^File x] (.getPath x))}. 
+The call 
+$$
+\clj{(path (File. "dir/a"))}
+$$
+now dispatches to the \clj{File} method because
+$$
 \clj{(isa? (class "dir/a") String)}
-is true, and finally returns \clj{((fn [x] x) "dir/a")}.
-\clj{isa?} is a subclassing check when provided with classes,
-otherwise it is an equality check.
+$$
+is true, and finally returns 
+$$
+\clj{((fn [^File x] (.getPath x)) "dir/a")}.
+$$
+
+The \clj{isa?} function first tries an equality check
+on its arguments, then if that fails
+and both arguments are classes a subclassing
+check is returned.
 \begin{minted}{clojure}
+  (isa? :a :a) ;=> true
   (isa? Keyword Object) ;=> true
-  (isa? :a :a) ;=> true
 \end{minted}
 
+In the full definition for \clj{path} we add a method for \clj{nil}
+which uses the equality check provied by \clj{isa?}.
+
+\begin{Code}
 \begin{exmp}
@@
 \end{exmp}
+\end{Code}
 
-Typed Clojure does not statically ensure multimethod calls dispatch successfully---\clj{(path :a)} 
-type checks but throws a runtime error.
+Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
+type checks because \clj{:a} agrees with the parameter type \clj{Any},
+but throws an error at runtime.
 
@@
 of our HMap AST \clj{Expr}.
-Keywords are functions that look themselves up in their argument, so \clj{:op}
-is our dispatch function.
+Since keywords are functions that look themselves up in their argument, we simply
+use \clj{:op} as the dispatch function.
 
 \begin{exmp}
-\inputminted[firstline=5,lastline=25]{clojure}{code/demo/src/demo/eg5.clj}
+\inputminted[firstline=5,lastline=31]{clojure}{code/demo/src/demo/eg5.clj}
 \end{exmp}
@@
 multimethod dispatch, reflection resolution via type hints, Java method
-and constructor calls, conditional and exceptional flow reasoning
-and use of HMaps.
+and constructor calls, conditional and exceptional flow reasoning,
+and HMaps.
 
@@
 
-\clj{PayLoad} is either a HMap containing a file
-or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
+We dispatch on \clj{:p} to distinguish the two cases of \clj{FSM}---for example on \clj{:F}
 we know the \clj{:file} is a file.

### 2015-02-27 formal-model.tex
overview
isa? formal

@@
 
-The particular part of the environment we reference in a positive or negative
+The particular part of the environment we reference in a
 type proposition is called the \emph{object}.
@@
     {\TLet}
-    {\TConst}
-
-    %{\TKw}
+    \begin{array}{c}
+      {\TConst}\\\\
+      {\TKw}\\\\
+      {\TClass}
+    \end{array}
 
@@
 \begin{minted}{clojure}
-      (do (if (number? x) nil (throw (Exception.)))
-          (inc x))
+...  (do (if (number? x) nil (throw (Exception.)))
+         (inc x)) ...
 \end{minted}
@@
 
-Before tackling multimethods we model their core dispatch mechanism
-\isaliteral{}, a subclassing test for classes, otherwise an equality test.
+We now consider the core dispatch mechanism for multimethods. 
+Recalling the examples in section~\ref{sec:multioverview},
+\isaliteral{} is
+a subclassing test for classes, otherwise an equality test---we do not model
+the semantics for vectors.
 
-\begin{minted}{clojure}
-(isa? Keyword Object) ;=> true
-(isa? Object Keyword) ;=> false
-(isa? :a :a) ;=> true
-(isa? :a :b) ;=> false
-(isa? :a Keyword) ;=> false
-(isa? (class :a) Keyword) ;=> true
-\end{minted}
+The main interest for the typing rule is \isacompareliteral{} 
+(figure~\ref{main:figure:mmsyntax}).
 
-B-IsA with the metafunction \isaopsemliteral{} reflect these semantics.
+\begin{mathpar}
+  \TIsA{}
+\end{mathpar}
+
+To demonstrate,
+\mintinline{clojure}{(isa? (class x) Keyword)}
+has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.
+
+B-IsA models the semantics. \isaopsemliteral{} explicitly handles classes in the second case.
 
@@
 
-For the typing rules, we add T-Kw and T-Class for type checking keyword and class literals
-as singleton types, and T-IsA for \isaliteral{} expressions, which 
-uses the \emph{object} of the left argument and the \emph{type} of the right argument
-to infer the final proposition set.
-
-$$
-\begin{array}{ll}
-\vcenter{\hbox{
-    \TIsA{}}} &
-
-  \vcenter{\hbox{$
-    \begin{array}{l}
-      {\TKw}\\\\
-      {\TClass}
-    \end{array}$}}
-\end{array}
-$$
-
-The definition of \isacompareliteral{} (figure~\ref{main:figure:mmsyntax}) is deliberately conservative.
-The first line considers the case where the object of the left argument
-is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
-The simplest example
-
-\begin{minted}{clojure}
-  (isa? (class x) Keyword)
-\end{minted}
-
-has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
-that is if the expression is true then \clj{x} is a keyword, otherwise \clj{x} is not a keyword.
-
-We can rewrite example~\ref{example:conditionalflow} as
-\inputminted[firstline=5]{clojure}{code/demo/src/demo/isa_nil.clj}
-
-because the then proposition is {\isprop{\Class}{\path{\classpe{}}{\x{}}}}, so
-we know {\isprop{\Object}{\x{}}} by L-Update, and combined with the fact {\isprop{\Union{\Nil{}}{\Number}}{\x{}}}
-we can infer {\isprop{\Number}{\x{}}} by L-Update.
-
-The second \isacompareliteral{} line handles the equality mode of \isaliteral{}
-with a non-class singleton type as the second argument.
+%The definition of \isacompareliteral{} (figure~\ref{main:figure:mmsyntax}) is deliberately conservative.
+%The first line considers the case where the object of the left argument
+%is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
 
 \constantsemfigure{main}
-
-%Substitution is used to ensure well formed propositions.
-
-The final line gives up---we cannot distinguish between equality and subclassing mode.
 

### 2015-02-27 overview.tex
isa? formal

@@
 from immutable maps.
+\clj{class} returns the argument's class, or \clj{nil} if passed \clj{nil}.
 The \clj{=} function returns true if both arguments are equal.
@@
 The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
-The typed \clj{fn} has optional annotations by following a parameter
-or binding vector by \clj{:-} and a type.
+The typed \clj{fn} supports optional annotations by 
+adding
+\clj{:-} and a type after a parameter
+position
+or binding vector 
+to annotate parameter types
+and return types respectively.
 
@@
 a technique for deriving type information from conditional control flow.
-They introduce the concept of occurrence typing 
+They introduced the concept of occurrence typing 
 with the following example.
@@
 type checking dynamically typed programs without compromising
-soundness, where other approaches like success typing~\cite{todo-dialyzer-paper}
-or soft typing~\cite{successtyping-todo}
-fail to prove strong properties about programs
-and have proved too complicated in practice respectively.
+soundness, which is appealing over success typing~\cite{todo-dialyzer-paper}
+which cannot prove strong properties about programs
+and soft typing~\cite{successtyping-todo}
+which has proved too complicated in practice.
 
@@
 Clojure programmers rely on \emph{exceptions}
-to assert type-related invariants, such as preventing
-null-pointer exceptions.
+to assert type-related invariants.
 
@@
 \end{exmp}
+
+In this example a \clj{throw} expression guards \clj{(inc x)}
+from being evaluated if \clj{x} is \clj{nil}, preventing
+a possible null-pointer exception.
 
@@
 based on the first subexpression.
-\footnote{We omit \clj{ns} forms for the rest of the examples, see <this> github repo for full working examples}
+\footnote{See \url{https://github.com/typedclojure/examples}
+  for full examples. From here we omit \clj{ns} forms.}
 We model this formally (section~\ref{sec:doformal}) and prove
@@
 
-The next example is extracted from a Typed Clojure production system at CircleCI
+The next example is extracted from a production system at CircleCI
 (section~\ref{sec:casestudy} presents a case study).
@@
 operation forgot to remove it.
+
+The next example is the same except we use the \clj{:absent-keys}
+HMap option
+and
+utilize destructuring.
 
@@
 
-A multimethod in Clojure is a function that contains a dispatch
-function and methods. Multimethods are created with {\clj{defmulti}}.
+\label{sec:multioverview}
+
+A multimethod in Clojure is a function with a \emph{dispatch
+function} and a \emph{dispath table} of methods. Multimethods are created with {\clj{defmulti}}.
 \inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
-The multimethod \clj{path} has type \clj{[Any -> String]}, an initially empty \emph{dispatch table}
-and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
+The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
+and \emph{dispatch function} \clj{class}.
 There are currently no methods so \clj{path} will fail at runtime.

### 2015-02-27 overview.tex
overview

@@
 
-The next example is extracted from a production system at CircleCI
+\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI
 (section~\ref{sec:casestudy} presents a case study).
@@
 
-Next, 
-HMaps interact with occurrence typing in interesting ways.
+We would still catch the bug if \clj{dissoc} forgot to remove \clj{:private-key},
+but now stashing the raw private key as \clj{:secret-key} is allowed.
+
+\paragraph{Branching on HMaps} Finally, testing on HMap properties
+allows us to refine its type down branches. \clj{dec-map} takes an
+\clj{Expr}, traverses to its nodes and decrements their values, then
+builds the \clj{Expr} back up with the decremented nodes.
 
 \begin{exmp}
-\inputminted[firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
+\inputminted[linenos,firstnumber=1,firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
+\label{example:decmap}
 \end{exmp}
 
-Section~\ref{sec:hmapformal} discusses this in greater detail.
+If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
+we remove
+the \clj{:do} and \clj{:const}
+Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
+and we are left with an \clj{:if} Expr.
+On line 8,
+we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
+Line 9 we know \clj{m}
+we remove \clj{:const} because it contradicts \clj{(= (:op m) :do)} being true,
+and we know \clj{m} is a \clj{:do} Expr.
+Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
+so we are left with \clj{m} being a \clj{:const} Expr.
+
+Section~\ref{sec:formalpaths} discusses how this automatic reasoning is achieved.
 
@@
 
-The flexibility of \clj{isa?} is key to the generality of multimethods. 
+\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
 We can dispatch on the \clj{:op} key 
@@
 \begin{exmp}
-\inputminted[firstline=5,lastline=31]{clojure}{code/demo/src/demo/eg5.clj}
+\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
+\label{example:incmap}
 \end{exmp}
 
-\clj{isa?} is special with vectors---vectors of the
+\clj{inc-map} is like example~\ref{example:decmap} except the nodes are incremented.
+The reasoning is similar, except we only use positive
+information since we only consider the \emph{dispatch value} and how it relates
+to the \emph{dispatch function}.
+In the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
+narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
+and \clj{:const} methods.
+
+\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
 same length recursively call \clj{isa?} on the elements pairwise.
@@
 Now we can simulate multiple dispatch by dispatching on
-a vector containing the class of both arguments.
+a vector containing the class of both arguments. \clj{open}
+takes two arguments which can be strings or files and returns
+a new file that concatenates their paths.
 
@@
 via type hints.
-No extra annotations are needed to follow type-directed control flow
-in multimethod dispatch.
+Multiple dispatch follows the same kind of reasoning as example~\ref{example:incmap},
+except we update multiple bindings simultaneously.
 

### 2015-02-27 overview.tex
overview

@@
 \clj{inc-map} is like example~\ref{example:decmap} except the nodes are incremented.
-The reasoning is similar, except we only use positive
-information since we only consider the \emph{dispatch value} and how it relates
+The reasoning is similar, except we only consider one branch (the current method) by
+locally considering the current \emph{dispatch value} and reasoning about how it relates
 to the \emph{dispatch function}.
-In the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
+For example, 
+in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
 narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}

### 2015-02-27 overview.tex
formatting

@@
 \paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
-We can dispatch on the \clj{:op} key 
+In example~\ref{example:incmap} we
+dispatch on the \clj{:op} key 
 of our HMap AST \clj{Expr}.
@@
 
+\begin{figure}
 \begin{exmp}
@@
 \end{exmp}
+\end{figure}
 
@@
 
-Now we can simulate multiple dispatch by dispatching on
+Example~\ref{example:multidispatch}
+simulates multiple dispatch by dispatching on
 a vector containing the class of both arguments. \clj{open}
@@
 
+\begin{figure}
 \begin{exmp}
 \inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
+\label{example:multidispatch}
 \end{exmp}
+\end{figure}
 
@@
 
-The final example combines everything we will cover for the rest of the paper:
+Example~\ref{example:final}
+combines everything we will cover for the rest of the paper:
 multimethod dispatch, reflection resolution via type hints, Java method
@@
 
+\begin{figure}
 \begin{exmp}
 \inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
+\label{example:final}
 \end{exmp}
+\end{figure}
 

### 2015-02-27 formal-model.tex
overview
tweak figure
some bib

@@
 novel feature of Typed Clojure to the formalism,
-interleaving presentation of typing rules, operational semantics
+interleaving presentation of syntax, typing rules, operational semantics
 and subtyping.
 
+%TODO EXPLAIN what is a proposition?
 The first insight about occurrence typing is that
-propositions can be used to capture type information about our programs,
-in particular by referencing components or subcomponents of the runtime environment
-and relating them to types.
+logical formulas
+can be used to represent type information about our programs
+by relating parts of the runtime environment to types.
+Occurrence typing uses propositional logic 
 \emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
@@
 and {\notprop{\Nil{}}{\x{}}}. 
-The other propositions are standard: the implication, conjunction,
-disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
+The other propositions are standard logical connectives: implications, conjunctions,
+disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
 (figure~\ref{main:figure:termsyntax}).
@@
 \subsection{Paths}
+\label{sec:formalpaths}
 
@@
  \classtablelookupsyntax{}
-  \convertjavatypegrammar{}
+ \begin{mathpar}
+  \begin{altgrammar}
+    \convertjavatypenil{}
+  \end{altgrammar}
+  \begin{altgrammar}
+    \convertjavatypenonnil{}
+  \end{altgrammar}
+\end{mathpar}
   \begin{mathpar}
@@
     {\TMethodStatic}
+
+    %{\TInstance}
   \end{mathpar}

### 2015-02-27 intro.tex
overview, Thursday evening changes
some bib

@@
 \begin{figure}
-\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/parent2.clj}
+\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/parent2.clj}
 \caption{A simple Typed Clojure program}
@@
 aspects of our system, from simple type annotations to explicit
-handling of Java's \java{null} in interoperation, as well as an
+handling of Java's \java{null} (written \clj{nil}) in interoperation, as well as an
 extended form of occurrence typing and Clojure's \emph{type hints},

### 2015-02-27 related-work.tex
some bib

\section{Related Work}

\paragraph{Multimethods} 
Millstein and Chambers~\cite{MS02}
describe Dubious, a simple statically typed core language including multimethods that
dispatch on the type of its arguments. They tackle a key challenge for statically typing
multimethods: ``it is possible for two modules containing arbitrary multimethods to typecheck
successfully in isolation but generate type errors when linked together.''~\cite{MS02}

\paragraph{Occurrence Typing} 
Occurrence typing~\cite{TF08,TF10} extends the type 
system with a \emph{proposition environment} that represents 
the information on the types of bindings down conditional branches.
These propositions are then used to update the types associated
with bindings in the \emph{type environment} down branches
so binding occurrences are given different types 
depending on the branches they appear in, and the conditionals
that lead to that branch.

\paragraph{Record Types} 
O'Caml-style extensible record types have been the subject of extensive research 
(eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91})
most inference

Dependent JavaScript~\cite{Chugh:2012:DTJ} can track similar invariants as HMaps with types
for JS objects. They must deal with mutable objects, they feature refinement types and strong updates to the heap
to track changes to objects.

Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track entries in a mutable Lua table.
Typed Lua changes the dynamic semantics of Lua to accommodate mutability: Typed Lua raises a runtime error
for lookups on missing keys---HMaps consider lookups on missing keys normal.

\paragraph{Optional type systems}
\begin{itemize}
  \item Reticulated Python~\cite{Vitousek14}
  \item GradualTalk
  \item Flow
\end{itemize}

Typed Racket is a statically typed sister language of Racket. It
attempts to preserve existing Racket idioms and aims to type check
existing Racket code by simply adding top level type annotations~\cite{Tob10}.

Typed Racket fully expands all macro calls before type checking~\cite{Tob10},
avoiding the complex semantics of type checking macro definitions, an ongoing research area summarised
by~\citet{Her10}. Typed Clojure also expands macros before type checking.

\paragraph{Variable-arity Polymorphism}

~\citet{STF09} invented a type system supporting variable-arity polymorphism  % doesn't fit on a line
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

\paragraph{Java Interoperability in Statically Typed Languages}

Scala~\cite{OCD+} has nullable references for compatibility with Java.
Programmers must manually check for
\java{null} like in Java to avoid null-pointer exceptions.

% def checkNullOrEmpty(v:Seq[Any]):Boolean = {
%   return v.getClass
% }
% 
% checkNullOrEmpty(null) ;=> NPE
%
% null.getClass

### 2015-02-27 formal-model.tex
move T-If to prose

@@
 can be used to represent type information about our programs
-by relating parts of the runtime environment to types.
-Occurrence typing uses propositional logic 
+by relating parts of the runtime environment to types
+via propositional logic.
 \emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
-``variable \clj{x} is not \clj{nil}''---
-in our logical system we write these as
+``variable \clj{x} is not \clj{nil}''---in our logical system we write these as
 {\isprop{\NumberFull}{\x{}}}
@@
 type environment of a map from variables to types
-with a sequence of propositions, written \propenv{}. 
+with a set of propositions, written \propenv{}. 
 Instead of mapping \clj{x} to
@@
 In addition to the standard rules for the logical connectives, the key
-rule is L-Update, which combines multiple propositions about the same variable.
+rule is L-Update, which combines multiple propositions about the same variable,
+allowing us to refine its type.
 $$
@@
 The key rule for reasoning about conditional control flow is
-T-If. The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
+T-If.
+
+\begin{mathpar}
+\TIf{}
+\end{mathpar}
+
+The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
 are used as assumptions in the then and else branch respectively.
@@
   \begin{mathpar}
-    {\TLocal}
-
-    {\TTrue}
-
-    {\TFalse}
-
-    {\TNil}
-
     %{\TDo}
     %{\TClass}
-    {\TIf}
-    {\TLet}
+    %{\TIf}
+    {\TAbs}
+    {\TSubsume}
     \begin{array}{c}
@@
     \end{array}
+    \begin{array}{c}
+      {\TTrue}\\\\
+      {\TFalse}\\\\
+      {\TNil}
+    \end{array}
+
+    {\TLet}
+    {\TLocal}
 
     {\TApp}
-
-    {\TAbs}
-
     %{\TError}
 
-    {\TSubsume}
   \end{mathpar}

### 2015-02-27 intro.tex
move T-If to prose

@@
 %
-The annotation \clj{^File} is a type hint, which instructs the Clojure
+The annotation \clj{^File f} is a type hint on \clj{f}, which instructs the Clojure
 compiler (running prior to Typed Clojure typechecking) to statically
-resolve method calls to the relevant object using the Java \clj{File}
-class, rather than using reflective method calls at runtime.
+resolve the \clj{getParent} 
+call to \clj{File}'s \clj{getParent} method with signature \java{String getParent();}, rather than using reflective method calls at runtime.
 

### 2015-02-27 overview.tex
move T-If to prose

@@
 that describes the structure of a recursively-defined AST as a union of HMaps.
+A quoted keyword in a type such as \clj{':if} is a singleton type that contains just the keyword.
 \clj{an-exp} is a function verified to return an \clj{Expr}.
@@
 \inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
+\label{example:circleci}
 \end{exmp}
@@
 we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
-Line 9 we know \clj{m}
-we remove \clj{:const} because it contradicts \clj{(= (:op m) :do)} being true,
+Line 9 know we can
+remove the \clj{:const} Expr from the type of \clj{m} because it contradicts \clj{(= (:op m) :do)} being true,
 and we know \clj{m} is a \clj{:do} Expr.
@@
 method of \clj{java.io.File}.
-The method is unambiguous but type checking fails---Typed Clojure
+The call to \clj{getParent} is unambiguous at runtime but type 
+checking fails---Typed Clojure
 considers \clj{f} to be of type \clj{Any}, which is unsafe 
 to use as the target of even a resolved method.
-We annotate the function parameter with type \clj{File}.
+If instead we annotate the function parameter with type \clj{File},
+we get a static type error.
 
@@
 
+% type hints give no guarantees
+% TODO
+
 The type hinting system and Typed Clojure's static type checking
 are separate, the latter predating the former by several years.
-The interaction between them is often not as obvious, for example type hints
-do not need to take \clj{nil} into account, while Typed Clojure guarantees
+The interaction between them is often not as obvious, for example Typed Clojure
+has an explicit type for \clj{null}
 null-pointer exceptions are impossible.
+
+do not need to take \clj{nil} into account,
 
@@
 
-To guarantee \java{null} is never accidentally leaked into a Typed Clojure program,
-we must assume methods are nullable.
+Reference types in Java are nullable, so to guarantee a method call does not
+leak \java{null} into a Typed Clojure program we
+must assume methods can return \clj{nil}.
 
@@
 
-In contrast, JVM invariants guarantee that  constructor cannot return \java{null},
+In contrast, JVM invariants guarantee that  constructors cannot return \java{null},
 so we are safe to assume constructors are non-nullable.
@@
 A multimethod in Clojure is a function with a \emph{dispatch
-function} and a \emph{dispath table} of methods. Multimethods are created with {\clj{defmulti}}.
+function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
 \inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
@@
 
-\begin{figure}
 \begin{exmp}
@@
 \end{exmp}
-\end{figure}
 

### 2015-02-27 formal-model.tex
3.4

@@
 (The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
-We cover \updateliteral{} in section~\ref{sec:updateformalsection}.
+We cover L-Update in more detail in section~\ref{sec:formalpaths}.
 
@@
 \begin{mathpar}
-\TIf{}
+  {\TIf}
 \end{mathpar}
@@
 
-We introduce our first path element
+\paragraph{Key path element} We introduce our first path element
 {\keype{\k{}}}, which represents the operation of looking up
@@
 
-By T-Abs, checking the body of the function we have the proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
-By T-Local we have
+We start with proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
+Since {\Expr{}} is a union of HMaps, each with the entry \makekw{op}, we can use T-GetHMap.
 $$
-\judgement{\propenv{}}{\hastype{m}{\Expr{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{m}
+\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\path{\keype{\makekw{op}}}{m}}
 $$
-and since {\Expr{}} is a union of HMaps with the entry \makekw{op}, we use T-GetHMap to derive
+Using our intuitive definition of \equivliteral{} above, we know
 $$
-\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\object{1}}
+\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}{\notprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}}{\emptyobject{}}
 $$
-where \object{1} = {\path{\keype{\makekw{op}}}{m}}. Using our intuitive definition of \equivliteral{} above, we know
-$$
-\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\object{1}}}{\notprop{\Value{\makekw{if}}}{\object{1}}}}{\emptyobject{}}
-$$
-Now we check the then branch with our new assumption.
-$$
-\judgement{\propenvp{}}{\hastype{\curlymap{{\mapvalentrykwkeynoarrow{op}{\makekw{if}}}, ...}}{\Expr{}}}{...}{...}
-$$
-We are interested in what L-Update can derive in our new environment,
+Going down the then branch gives us the extended environment
 \propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.
-
-With the premises
-\inpropenv{\propenvp{}}{\isprop{\Expr{}}{m}}
-and
-\inpropenv{\propenvp{}}{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}
-we derive
-$$
-\inpropenv{\propenvp{}}{\isprop{\update{\Expr{}}{\Value{\makekw{if}}}{\keype{\makekw{op}}}}{m}}
-$$
-which simplifies to 
+Using L-Update we can combine what we know about object $m$ and object
+{\path{\keype{\makekw{op}}}{m}}
+to derive
 $$
@@
 
-Our second path element \classpe{} is used in the latent
+\paragraph{Class path element} Our second path element \classpe{} is used in the latent
 object of the constant \classconst{} function. Like Clojure's \clj{class}
@@
     {\TNewStatic}
-    {\TFieldStatic}
 
@@
 
-We extend the syntax with Java field lookups and calls to
+We extend the syntax with type hinted expressions, Java field lookups and calls to
 methods and constructors. We model the syntax after the `dot' special
@@
 and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
-These come without typing or evaluation rules---Typed Clojure 
-demands statically resolved Java interactions
-in typed code
+The Java expressions come without typing rules because Typed Clojure
+only reasons about resolved reflection
 (as demonstrated in section~\ref{sec:overviewjavainterop}).
-We add type hinted syntax for local variables and let-bindings.
 
@@
                                {\class{}} {\overrightarrow{\e{i}}}}
-calls the constructor
+calls the constructor with Java signature
 ${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.
@@
 To obtain non-reflective calls we define our rewrite relation
-$$
 \rewrite {\taenv{}} {\e{}} {\ep{}}
-$$
 which rewrites \e{} to a possibly-less reflective expression
@@
 
+For example,
+R-FieldElimRefl emits a non-reflective field if it can
+find a field matching the type hint inferred on \ep{}.
+  $$
+    \RFieldElimRefl{}
+  $$
+
 Type hint inference is modelled by
-$$
 \tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
-$$
 which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
@@
 
-To demonstrate we consider the expression 
+For demonstration purposes let us rewrite
+a simple field expression, with the assumptions that
+that we have a class \Point{} with a field \getx{} of Java type \Number
+and \intaenv {\taenv{}} {\x{}} {\Point{}}.
+In rewriting
+the expression \fieldexp{\getx}{\x{}},
+{\Point{}} is inferred as the type hint of \x{}
+and \Number{} is the field type
+by \fieldtypeliteral{} (figure~\ref{main:figure:javatyping}).
+Now we just plug in the new information into our new expression
 $$
-\fieldexp{getParent}{\hinted{\File}{\x{}}}.
-$$
-where 
-$$
-\ct = \{ \File \mapsto \{ \ctfieldsliteral \mapsto \{\ctfldentry{getMethod}{\String}\} \} \}
+\fieldstaticexp{\Point}{\Number}{\getx}{\x{}}.
 $$
 
-R-FieldElimRefl eliminates a reflective field.
-  $$
-    \RFieldElimRefl{}
-  $$
-First the target is rewritten \rewrite {\taenv{}} {\hinted{\File}{\x{}}}{\hinted{\File}{\x{}}}
-then we infer the type hint on the new expression \tajudgement {\taenv{}} {\hastype {\hinted{\File}{\x{}}} {\File}}
-We call \fieldtypeliteral{} (figure~\ref{main:figure:javatyping})
-to determine the exact signature of the field:
-\fieldtype{\ct{}}{\File}{getParent}{\File}{\String}
-and return the non-reflective field 
-$$
-\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}.
-$$
+Now we present the typing rules for resolved Java interoperability.
+T-FieldStatic checks a resolved field expression by ensuring the target has
+the correct static type, then returns a nilable type corresponding the Java type.
 
-Now we present the typing rules (figure~\ref{main:figure:javatyping}).
-To briefly finish our example, let us assume \propenv{} = {\isprop{\File}{\x{}}},
-then
-T-FieldStatic converts the target Java type to a non-nilable
-type \javatotc {\File} {\File}} and checks the target against it 
-\judgement{\propenv{}}{\hastype{\x{}}{\File}}, and finally converts the return Java type 
-to a nilable type and \javatotc {\String} {\Union{\Nil{}}{\String}}
-to finish with 
-$$
-\judgementtwo{\propenv{}}{\hastype{\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}}{\Union{\Nil{}}{\String}}}.
-$$
+\begin{mathpar}
+    {\TFieldStatic}
+\end{mathpar}
 
-The other rules T-MethodStatic and T-NewStatic work similarly, and vary
-primarily in the choice of nilability in the conversion function---method returns
+To continue our example, let us assume \propenv{} = {\isprop{\Point}{\x{}}}.
+T-FieldStatic checks \x{} under the non-nilable type {\Point}
+and returns the nilable type {\Union{\Nil}{\Number}}.
+
+The other rules T-MethodStatic and T-NewStatic work similarly (figure~\ref{main:figure:javatyping}), varying
+in the choice of nilability in the conversion function---method returns
 are nilable and constructor returns are non-nilable.
 
-The operational semantics B-Field, B-New and B-Method simply evaluate their
+The operational semantics B-Field, B-New and B-Method (figure~\ref{main:figure:javatyping}) simply evaluate their
 arguments and call the relevant JVM operation---section~\label{sec:metatheory}
-states our exact assumptions about each one.
+states our exact assumptions about each.
 

### 2015-02-27 overview.tex
3.4

@@
 function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
-\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
+\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
 The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
 and \emph{dispatch function} \clj{class}.
-There are currently no methods so \clj{path} will fail at runtime.
-\begin{minted}{clojure}
-  (path (File. "dir/a"));=> IllegalArgumentException
-\end{minted}
 
-We can use {\clj{defmethod}} to add a method to \clj{path}.
+We can use {\clj{defmethod}} to install a method to \clj{path}.
+\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
+Now the dispatch table maps
+the \emph{dispatch value} \clj{String} to the function
+\clj{(fn [x] x)}. 
+We add another method
+which maps
+\clj{File} to the function
+\clj{(fn [^File x] (.getPath x))}
+in the dispatch table.
 \inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
-Now the dispatch table maps
-the \emph{dispatch value} \clj{File} to the function
-\clj{(fn [^File x] (.getPath x))}. 
-The call 
+
+After installing both methods, the call 
 $$
@@
 $$
-now dispatches to the \clj{File} method because
+dispatches to the second method we installed because
 $$
@@
 
-In the full definition for \clj{path} we add a method for \clj{nil}
-which uses the equality check provied by \clj{isa?}.
+We include the above sequence of definitions as example~\ref{example:rep}.
 
@@
 \begin{exmp}
-\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
+\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
 \label{example:rep}

### 2015-02-27 intro.tex

--- 2015-02-27 intro.tex
@@
 for Python among the optional systems, and Typed Racket, Reticulated
-Python, and GradualTalk among gradually-typed system.\footnote{We
+Python, and GradualTalk among gradually-typed systems.\footnote{We
   reserve the term ``gradual typing'' for systems such as Typed Racket which soundly
@@
 Clojure, producing an expressive synthesis of ideas and demonstrating
-a surprising coincidence between the multiple dispatch approach of
-Clojure and the occurrence typing framework of Typed Racket.
+a surprising coincidence between multiple dispatch in
+Clojure and Typed Racket's  occurrence typing framework.
 
@@
 the JVM. As a result, it now has an extensive base of existing untyped
-programs, many of whose developers can benefit from Typed Clojure. As
-a result, Typed Clojure is now used in industry, experience we discuss
+programs, whose developers can now benefit from Typed Clojure. As
+a result, Typed Clojure is used in industry, experience we discuss
 in this paper.
@@
 %
-The annotation \clj{^File} is a type hint, which instructs the Clojure
+The annotation \clj{^File f} is a type hint on \clj{f}, which instructs the Clojure
 compiler (running prior to Typed Clojure typechecking) to statically
-resolve method calls to the relevant object using the Java \clj{File}
-class, rather than using reflective method calls at runtime.
+resolve the \clj{getParent} 
+call to \clj{File}'s \clj{getParent} method with signature \java{String getParent();}, rather than using reflection at runtime.
 

### 2015-02-27 small.tex
overview, Thursday evening changes
isa? formal

@@
 \usepackage{chngcntr}
+\usepackage{url}
 
@@
 \par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
-\\[-2.5em]\nopagebreak}
+\\[-2.4em]\nopagebreak}
 {\relax}
@@
 system build on the design of Typed Racket, repurposing in particular
-\emph{occurrence typing}, its approach to statically reasoning about
+\emph{occurrence typing}, an approach to statically reasoning about
 predicate tests. However, in adapting the type system to Clojure,
-changes and extensions are required to accommodate both additional
+changes and extensions are required to accommodate  additional
 language features and  idioms used by Clojure programmers.

### 2015-02-27 related-work.tex
update figure

@@
 
-\paragraph{Occurrence Typing} 
-Occurrence typing~\cite{TF08,TF10} extends the type 
-system with a \emph{proposition environment} that represents 
-the information on the types of bindings down conditional branches.
-These propositions are then used to update the types associated
-with bindings in the \emph{type environment} down branches
-so binding occurrences are given different types 
-depending on the branches they appear in, and the conditionals
-that lead to that branch.
+% one sentence
+% TC based on TR, already covered
+
+%\paragraph{Occurrence Typing} 
+%Occurrence typing~\cite{TF08,TF10} extends the type 
+%system with a \emph{proposition environment} that represents 
+%the information on the types of bindings down conditional branches.
+%These propositions are then used to update the types associated
+%with bindings in the \emph{type environment} down branches
+%so binding occurrences are given different types 
+%depending on the branches they appear in, and the conditionals
+%that lead to that branch.
+
+% What's diff about TC from the related work
+% small summary for deisel....
+% - diesel supports x
+%- - calculus supports some subset of x
+% we support y, which covers most of x but also foo
+
+% eg. multiple dispatch
+%     nominal vs structural
+
+% eg. run abritrary metaprogramming over dispatch in CLOS
+%  more expressive
+
+% type systems for mm or rows
+% rows vs HMap
+% - no poly in HMap
+% - based on subtyping
+% - rows based on polymorphism
 

### 2015-02-27 formal-model.tex
update figure
move syntax descriptions to overview

@@
 disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
-(figure~\ref{main:figure:termsyntax}).
+(\figref{main:figure:termsyntax}).
 
@@
 (The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
-We cover L-Update in more detail in section~\ref{sec:formalpaths}.
+We cover L-Update in more detail in \secref{sec:formalpaths}.
 
@@
 conditional tests.
-For instance, example~\ref{example:conditionalflow}
+For instance, \egref{example:conditionalflow}
 only utilises \clj{x} once
@@
 
-Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
+Syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
 application, abstractions, conditionals and let expressions.
@@
 
-Figure~\ref{main:figure:typingrules} contains the core typing rules.
+\figref{main:figure:typingrules} contains the core typing rules.
 The key rule for reasoning about conditional control flow is
@@
 The T-Local rule connects the type system to the proof system over type propositions
+via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
 to derive a type for a variable.
@@
 We are now equipped to type check 
-example~\ref{example:conditionalflow}.
+\egref{example:conditionalflow}:
+$$
+\clj{(if (number? x) (inc x) 0)}
+$$
+
 The test expression uses T-App, 
@@
 
-Subtyping (figure~\ref{main:figure:subtyping}) 
+Subtyping (\figref{main:figure:subtyping}) 
 is a reflexive and transitive relation with top type \Top. 
@@
 
-Recall example~\ref{example:doexception}.
+Recall \egref{example:doexception}.
 \begin{minted}{clojure}
@@
 We now present \HMapliteral{} types in detail.
-Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
+\figref{main:figure:hmapsyntax} extends the syntax for terms and types.
 The syntax 
@@
 $$
-we introduced in section~\ref{sec:overview}
+we introduced in \secref{sec:overview}
 is now written \HMapgeneric{\mandatory{}}{\absent{}},
@@
 
-Subtyping for HMaps (figure~\ref{main:figure:hmapsyntax})
+Subtyping for HMaps (\figref{main:figure:hmapsyntax})
 designate \MapLiteral{} as a common supertype for all HMaps.
@@
 
-Figure~\ref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
+\figref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
 that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
@@
 contains the key we are looking up, in particular to support operations like
-looking up \clj{:op} on a map of type \clj{Expr} (example~\ref{example:decleaf}).
+looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf}).
 The object in each T-Get rule is now more complicated than in T-Local---the 
@@
 
-Figure~\ref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
+\figref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
 and \assocliteral{}.
@@
 We directly relate this to our typing rule T-GetHMap
-(figure~\ref{main:figure:hmapsyntax}) by
+(\figref{main:figure:hmapsyntax}) by
 checking the then branch of the first conditional test is checked in 
-an equivalent version of example~\ref{example:decleaf}.
+an equivalent version of \egref{example:decleaf}.
 \begin{minted}{clojure}
@@
 but on keyword arguments it works identically to \clj{isa?} which we model
-in section~\ref{sec:isaformal}.
+in \secref{sec:isaformal}.
 Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
@@
 \end{mathpar}
-The full semantics are given in figure~\ref{main:figure:primitivesem}.
+The full semantics are given in \figref{main:figure:primitivesem}.
 The definition of \updateliteral{} supports various idioms relating to \classpe{}
-which we introduce in section~\ref{sec:isaformal}.
+which we introduce in \secref{sec:isaformal}.
 
@@
 
-In section~\ref{sec:overviewjavainterop} we discussed the role of type hints
+In \secref{sec:overviewjavainterop} we discussed the role of type hints
 to help eliminate reflective calls.
@@
 only reasons about resolved reflection
-(as demonstrated in section~\ref{sec:overviewjavainterop}).
+(as demonstrated in \secref{sec:overviewjavainterop}).
 
 Now we model the compiler's reflection-resolution algorithm.
-Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
+\figref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
 The method ${\methodstaticexp {\classhint{1}}
@@
 and \Number{} is the field type
-by \fieldtypeliteral{} (figure~\ref{main:figure:javatyping}).
+by \fieldtypeliteral{} (\figref{main:figure:javatyping}).
 Now we just plug in the new information into our new expression
@@
 
-The other rules T-MethodStatic and T-NewStatic work similarly (figure~\ref{main:figure:javatyping}), varying
+The other rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
 in the choice of nilability in the conversion function---method returns
@@
 
-The operational semantics B-Field, B-New and B-Method (figure~\ref{main:figure:javatyping}) simply evaluate their
-arguments and call the relevant JVM operation---section~\label{sec:metatheory}
+The operational semantics B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
+arguments and call the relevant JVM operation---\secref{sec:metatheory}
 states our exact assumptions about each.
@@
 We now consider the core dispatch mechanism for multimethods. 
-Recalling the examples in section~\ref{sec:multioverview},
+Recalling the examples in \secref{sec:multioverview},
 \isaliteral{} is
@@
 The main interest for the typing rule is \isacompareliteral{} 
-(figure~\ref{main:figure:mmsyntax}).
+(\figref{main:figure:mmsyntax}).
 
@@
 To demonstrate,
-\mintinline{clojure}{(isa? (class x) Keyword)}
+\isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
 has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.
@@
 To ease presentation, we present \emph{immutable}
-multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
-without changing the original multimethod. Example~\ref{example:rep} is now written
+multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
+without changing the original multimethod. \egref{example:rep} is now written
 \begin{minted}{clojure}
@@
 
-Multimethod semantics are in figure~\ref{main:figure:mmsyntax}.
+Multimethod semantics are in \figref{main:figure:mmsyntax}.
 B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
@@
 
-The typing rules (figure~\ref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
+The typing rules (\figref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
 where \s{} is the interface type for beta reduction, and \t{} is the type for
@@
 method bodies. T-App handles beta reduction because multimethods are subtypes of functions
-(figure~\ref{main:figure:mmsyntax}).
+(\figref{main:figure:mmsyntax}).
 

### 2015-02-27 intro.tex
move syntax descriptions to overview

@@
 
-The popularity of dynamically-typed languages languages in software
+The popularity of dynamically-typed languages in software
 development, combined with a recognition that types often improve
@@
 Since its initial release in 2007, Clojure has been widely adopted for
-``backend'' development in places where its support for paralleism,
+``backend'' development in places where its support for parallelism,
 functional programming, and Lisp-influenced abstraction is desired on
@@
 others. We begin with an example-driven presentation of the main type
-system features in Section~\ref{sec:overview}. We then incrementatlly
+system features in \secref{sec:overview}. We then incrementally
 present a core calculus for Typed Clojure covering all of these
-features together in Section~\ref{sec:formal} and prove type
-soundness(section \ref{sec:sound}). We then discuss the full
+features together in \secref{sec:formal} and prove type
+soundness (\secref{sec:metatheory}). We then discuss the full
 implementation of Typed Clojure, dubbed \coretyped{}, which extends
 the formal model in many ways, and the experience gained from its use
-in Section~\ref{sec:exp}. Finally, we discuss related work and
+in \secref{sec:experience}. Finally, we discuss related work and
 conclude.

### 2015-02-27 overview.tex
move syntax descriptions to overview

@@
 
-\subsection{Clojure and Typed Clojure Primer}
+\subsection{Clojure Syntax}
 
-We now review the basics of the Clojure subset we use in the paper.
+We describe new syntax as they appear in each example, but
+we also include the essential basics of Clojure syntax.
 
-\paragraph{Syntax} Parentheses indicate \emph{applications}, brackets
+\clj{nil} is exactly Java's \java{null}.
+Parentheses indicate \emph{applications}, brackets
 delimit
@@
 \emph{hash-maps}
-and double quotes delimit Java \emph{strings}.
-\clj{nil} is exactly Java's \java{null}.
-\emph{Symbols} begin with an alphabetic character.
-A colon prefixed symbol like \clj{:a} is a \emph{keyword}.
-\emph{Commas} are always whitespace.
+and double quotes delimit \emph{Java strings}.
+\emph{Symbols} begin with an alphabetic character,
+and a colon prefixed symbol like \clj{:a} is a \emph{keyword}.
 
-\paragraph{Core forms} \clj{if} is a standard two-armed conditional.
-\clj{do} sequences two expressions returning the latter.
-\clj{throw} corresponds to Java's \java{throw}.
-\clj{fn} creates a function from a parameter vector and a body expression.
-\clj{let} takes a flat binding vector (which can refer to previous bindings like Scheme's \mintinline{scheme}{let*})
-and a body expression.
-\clj{def} binds a top-level \emph{var}.
-\clj{defn} combines \clj{def} and \clj{fn}.
-
-\paragraph{Semantics} \clj{get} applied to a hash-map and a key returns the value
-associated with the key in the map, or \clj{nil} if there is none.
-Keywords are functions that apply \clj{get} to their argument
-and itself.
-\clj{inc} and \clj{dec} increment and decrement numbers;
-\clj{assoc} and \clj{dissoc} associate and dissociate entries
-from immutable maps.
-\clj{class} returns the argument's class, or \clj{nil} if passed \clj{nil}.
-The \clj{=} function returns true if both arguments are equal.
-\clj{number?} is
-a Java \java{instanceof} test of \clj{java.lang.Number}.
-\clj{nil} and \clj{false} are the false values.
-
-\paragraph{Namespaces} \clj{ns}
-declares a
-\emph{namespace}
-which manage var and class imports. 
-\clj{:require} takes a number of namespace dependencies and \clj{:refer}
-brings vars into scope by name.
-\clj{:refer-clojure} \clj{:exclude} unmaps core vars from the current namespace.
-
-\paragraph{Java Interoperability} Java methods and fields have prefix notation
-like \clj{(.method target args*)} and \clj{(.field target)} respectively,
-with method and field names prefixed with a dot and methods taking some number of arguments.
-Constructors are called with the class name with a dot suffix and arguments like \clj{(class. args*)}.
-
-\paragraph{Metadata} Symbols support \emph{metadata}---the syntax \clj{^File x}
-is a single expression of a symbol \clj{x} with metadata \clj{{:tag File}}.
-In binding positions like \clj{(fn [^File x] ...)}
-syntactic occurrences preserve metadata.
-
-\paragraph{Destructuring} \emph{Map destructuring} is binding position syntax for pattern matching.
-Parameters are replaced with maps of symbols to keywords that bind the symbols
-to lookups on that keyword, optionally terminated by \clj{:as}
-which aliases the parameter. For example
-\clj{(fn [{^File x :x, :as m}] ...)}
-expands to a \clj{let} binding \clj{m} to the first argument and 
-\clj{^File x} to \clj{(:x m)}.
-
-\paragraph{Static type syntax} Single quoted keywords are singleton types containing just the keyword.
-Single quoted maps mapping keywords to types 
-are HMap types representing
-maps with known keyword entries with values of the associated type.
-\clj{U} creates an \emph{untagged union} type, which can take
-any number of types.
-
-\paragraph{Typed Clojure Macros}
-The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
-The typed \clj{fn} supports optional annotations by 
-adding
-\clj{:-} and a type after a parameter
-position
-or binding vector 
-to annotate parameter types
-and return types respectively.
+\emph{Commas} are always \emph{whitespace}.
 
@@
 
-Here is above program in Typed Clojure.
+Here is the above program in Typed Clojure.
 
@@
 This is a regular Clojure file compiled with
-the Clojure compiler. It has a
-runtime dependency 
+the Clojure compiler, with the \clj{ns} form
+declaring a
+\emph{namespace}
+for managing var and class imports.
+Here \clj{:require} declares a runtime dependency 
 on
 \clj{clojure.core.typed},
-Typed Clojure's core namespace.
-Programmers import typed versions of `typed' macros and qualified types to ease annotation.
-The \clj{check-ns} function type checks the current namespace.
+Typed Clojure's core namespace, and
+\clj{:refer}
+brings a collection of vars into scope by name.
+The \clj{:refer-clojure} \clj{:exclude} option unmaps core vars from the current 
+namespace---here we unmap \clj{clojure.core/fn},
+which creates a function from a parameter vector and a body expression,
+and import a typed variant of \clj{fn}
+that supports type annotations.
 
-Typed Clojure can check all the examples in~\citet{TF10}---the 
+The typed \clj{fn} supports optional annotations by 
+adding
+\clj{:-} and a type after a parameter
+position
+or binding vector 
+to annotate parameter types
+and return types respectively.
+Typed Clojure provides a \clj{check-ns} function to type checks the current namespace.
+\clj{number?} is
+a Java \java{instanceof} test of \clj{java.lang.Number}.
+As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
+any number of types.
+
+Typed Clojure can already check all of the examples in~\citet{TF10}---the 
 rest of this section describes the extensions necessary
@@
 
-In this example a \clj{throw} expression guards \clj{(inc x)}
+The \clj{do} form sequences two expressions returning the latter,
+\clj{throw} corresponds to Java's \java{throw}
+and \clj{(class. args*)} is the syntax for Java constructors---that 
+is a class name with a dot suffix as the operator followed the arguments to the constructor.
+
+In this example a \clj{throw} expression guards \clj{(inc x)}, the increment function for numbers,
 from being evaluated if \clj{x} is \clj{nil}, preventing
@@
 
-Here \clj{defalias} defines a type abbreviation called \clj{Expr}
+The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
+The function \clj{an-exp} is verified to return an \clj{Expr}.
+
+The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
+Here \clj{defalias} defines \clj{Expr}
 that describes the structure of a recursively-defined AST as a union of HMaps.
-A quoted keyword in a type such as \clj{':if} is a singleton type that contains just the keyword.
-\clj{an-exp} is a function verified to return an \clj{Expr}.
+A quoted keyword in a type, such as \clj{':if}, is a singleton type that contains just the keyword.
+A type that is a quoted map like \clj{'{:op ':if}} is a
+HMap type with a fixed number of keyword entries of the specified types
+known to be \emph{present},
+zero entries known to absolutely be \emph{absent},
+and an infinite number of \emph{unknown} entries entries.
+Since only keyword keys are allowed, they do not require quoting.
 
@@
 \clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
-dissociating the raw \clj{:private-key} entry and associating an encrypted private key
-as \clj{:enc-private-key}.
+non-destructively dissociating the raw \clj{:private-key} entry with \clj{dissoc}
+and associating an encrypted private key
+as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
+The expression \clj{(:private-key kp)} shows that keywords are also 
+functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
 Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
@@
 
-The next example is the same except we use the \clj{:absent-keys}
-HMap option
-and
-utilize destructuring.
+The next example is the same except we use the \clj{:absent-keys} HMap option.
+We also
+utilize \emph{map destructuring}, which is binding position syntax for pattern matching.
+Parameters are replaced with maps of symbols to keywords that bind the symbols
+to lookups on that keyword, optionally terminated by \clj{:as}
+which aliases the parameter. For example
+\clj{(fn [{^File x :x, :as m}] ...)}
+expands to a \clj{let} binding \clj{m} to the first argument and 
+\clj{^File x} to \clj{(:x m)}.
+Clojure's \clj{let} takes a flat binding vector which can refer to previous bindings like Scheme's \mintinline{scheme}{let*},
+and a body expression.
 
@@
 
-We would still catch the bug if \clj{dissoc} forgot to remove \clj{:private-key},
-but now stashing the raw private key as \clj{:secret-key} is allowed.
+Since this example enforces that \clj{:private-key} must not appear
+in a \clj{EncKeyPair}
+Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
+from the return value.
+Now, however we could stash the raw private key in another entry
+like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
+without Typed Clojure noticing.
 
@@
 allows us to refine its type down branches. \clj{dec-map} takes an
-\clj{Expr}, traverses to its nodes and decrements their values, then
+\clj{Expr}, traverses to its nodes and decrements their values by \clj{dec}, then
 builds the \clj{Expr} back up with the decremented nodes.
@@
 
+Calls to Java methods and fields have prefix notation
+like \clj{(.method target args*)} and \clj{(.field target)} respectively,
+with method and field names prefixed with a dot and methods taking some number of arguments.
+
 Unlike Java, Clojure is dynamically typed. We have no type information about \clj{f}
@@
 \end{minted}
+
+Symbols support \emph{metadata}---the syntax \clj{^File f}
+is a single expression that is a symbol \clj{f} with metadata \clj{{:tag File}}.
+In binding positions like \clj{(fn [^File f] ...)}
+syntactic occurrences preserve metadata.
 
@@
 The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
-and \emph{dispatch function} \clj{class}.
+and \emph{dispatch function} \clj{class}, a function that
+returns the class of its argument or \clj{nil} if passed \clj{nil}.
 

### 2015-02-28 formal-model.tex
start of 3

@@
 
-%TODO EXPLAIN what is a proposition?
 The first insight about occurrence typing is that
@@
 via propositional logic.
-\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
-``variable \clj{x} is not \clj{nil}''---in our logical system we write these as
+\emph{Type Propositions} \prop{} make assertions like ``variable \x{} is of type \NumberFull{}'' or
+``variable \x{} is not \nil{}''---in our logical system we write these as
 {\isprop{\NumberFull}{\x{}}}
@@
 
-The particular part of the environment we reference in a
+The particular part of the runtime environment we reference in a
 type proposition is called the \emph{object}.
-Every expression has an object that is either \emph{empty}, written \emptyobject{}, 
+The typing judgement relates an object to every expression in the language.
+An object is either \emph{empty}, written \emptyobject{}, 
 which says 
-this expression is not known to evaluate to a particular component
-  of the current environment, or a 
+this expression is not known to evaluate to a particular part
+  of the current runtime environment, or a 
 variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
 that exactly indicates how the value of this
-expression can be derived from the current environment.
+expression can be derived from the current runtime environment.
 Type propositions can only reference non-empty objects.
@@
 representation of a
-type environment of a map from variables to types
+type environment (eg., a map from variables to types)
 with a set of propositions, written \propenv{}. 
-Instead of mapping \clj{x} to
-the type \clj{Number}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
+Instead of mapping \x{} to
+the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
 
@@
 {\ValueTrue} as \True and {\ValueFalse} as {\False}.
-Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
+Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
 may refer to the function argument.
@@
 
-The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
-are used as assumptions in the then and else branch respectively.
-If the result of the \ifliteral{} is true, then the value either
+The propositions of the test expression \e{1}, \thenprop{\prop{1}} and \elseprop{\prop{1}}, are 
+used as assumptions in the then and else branch respectively.
+If the result of the \ifliteral{} is a true value, then it either
 came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
@@
 We are now equipped to type check
-\egref{example:conditionalflow}:
+\egref{example:conditionalflow}, starting at body:
 $$
-\clj{(if (number? x) (inc x) 0)}
+\clj{... (if (number? x) (inc x) 0) ...}
 $$
 
+We know {\propenv{}} = {\isprop{\UnionNilNum{}}{\x{}}}.
 The test expression uses T-App, 
 $$
-\judgement{\isprop{\UnionNilNum{}}{\x{}}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\propenv{}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 
-Now we check both branches using the extended proposition environment as specified by T-If.
+Finally we check both branches using the extended proposition environment as specified by T-If.
+Going down the then branch, our new assumption {\isprop{\Number}{\x{}}} is crucial to check
 $$
-\judgement{{\isprop{\UnionNilNum}{\x{}}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\hastype{\x{}}{\Number{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
 $$
+because we can now satisify the premise of T-Local:
 $$
-\judgement{{\isprop{\UnionNilNum}{\x{}}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+\inpropenv{{\propenv{}},\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}.
 $$
+%\judgement{{\propenv{}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+%$$
+%$$
+%\judgement{{\propenv{}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+%$$
+
 %\inc{} has type

### 2015-02-28 formal-model.tex
paths

@@
 which states that \e{} evaluates to answer \definedreduction{} in environment
-\openv{}.
-A notable difference is \nil{} is a false value, which affects the
+\openv{}. We chose to omit the core rules (see \figref{appendix:figure:opsem})
+however a notable difference is \nil{} is a false value, which affects the
 semantics of \ifliteral{}:
@@
 \figref{main:figure:hmapsyntax} extends the syntax for terms and types.
-The syntax 
-$$
-\clj{(HMap :mandatory p :absent-keys a :complete? b)}
-$$
-we introduced in \secref{sec:overview}
-is now written \HMapgeneric{\mandatory{}}{\absent{}},
-where \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} and is {\partial{}} otherwise. 
-The \mandatory{} entry specifies a set of known present keyword entries and \absent{} tracks known absent keys.
+
+We write \HMapgeneric{\mandatory{}}{\absent{}}
+where {\mandatory{}} is a map of \emph{present} entries (mapping keywords to types),
+\absent{} is a set of keyword keys that are known to be \emph{absent}
+and
+\completenessmeta{} is {\complete{}} (``complete'') if the map is fully specified by \mandatory{},
+and {\partial{}} (``partial'') if there are \emph{unknown} entries.
 To ease presentation, 
-if a HMap is declared as \complete{} then \absent{} implicitly contains all other keywords not specified in \mandatory{}.
+if a HMap is declared as \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
 %\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
@@
 and T-GetHMapPartialDefault a lookup with unknown results.
-We only handle unions of HMaps in T-GetHMap, where every element in the union
-contains the key we are looking up, in particular to support operations like
-looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf}).
-The object in each T-Get rule is now more complicated than in T-Local---the 
+Lookups on unions of HMaps are only supported in T-GetHMap, 
+in particular to support
+looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf})
+where every element in the union
+contains the key we are looking up.
+The objects in the T-Get rules are more complicated than those in T-Local---the 
 next section discusses this in detail.
-T-Assoc extends a HMap with a mandatory entry.
+Finally T-AssocHMap extends a HMap with a mandatory entry while preserving completeness
+and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
+formed types.
 
@@
 
-Recall the first insight of occurrence typing---not
-only do we want to reason about components of the runtime environment, that
-is, variables, but \emph{subcomponents} of the environment.
-We briefly mentioned paths, written \path{\pathelem{}}{\x{}}, but we concentrated on
-reasoning only about the variable \x{}---the notation \pathelem{} is a nested sequence of \emph{path elements}
-applied right-to-left to a variable \x{}.
+Recall the first insight of occurrence typing---we can reason
+about specific \emph{parts} of the runtime environment
+using propositions.
+The way to refer to \emph{parts} of the runtime environment with
+occurrencing is via \emph{path elements}.
+A \emph{path} consists of a series of path elements
+applied right-to-left to a variable
+written
+\path{\pathelem{}}{\x{}}.
 \citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
 to reason about selector operations on cons cells.
-We instead want to reason about HMap lookups.
+We instead want to reason about HMap lookups and calls to \classconst{}.
 

### 2015-02-28 overview.tex
experience

@@
 
-\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI
+\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI,
+a company with a large production Typed Clojure system
 (section~\ref{sec:casestudy} presents a case study).

### 2015-02-28 formal-model.tex
experience
done mm formal

@@
     {\TAbs}
-    {\TSubsume}
+    \begin{array}{c}
+      {\TSubsume}\\\\
+      {\TNum}
+    \end{array}
     \begin{array}{c}
@@
 
-The main interest for the typing rule is \isacompareliteral{} 
-(\figref{main:figure:mmsyntax}).
+The main interest for the T-IsA typing rule is the \isacompareliteral{} 
+metafunction
+(\figref{main:figure:mmsyntax}), used to calculate the propositions for
+\isaliteral{} expressions.
 
@@
 \begin{minted}{clojure}
-(let [rep (defmulti [Any -> String] class)
-      rep (defmethod rep Keyword [x] (str (name x)))
-      rep (defmethod rep Number [x] (str (inc x)))]
-  (rep :a)) ;=> "a"
+(let [path (defmulti [Any -> (U nil String)] class)]
+  (let [path (defmethod path String [x] x)]
+    (let [path (defmethod path File [^File x] 
+                 (.getPath x))]
+      (path "dir/a")))) ;=> "a"
 \end{minted}
+where subsequent bindings in the binder is sugar for nested let expressions.
+
+The syntax for multimethods is given in \figref{main:figure:mmsyntax}.
+{\MultiFntype {\s{}} {\sp{}}} is the type of multimethods with \emph{interface type}
+{\s{}} and \emph{dispatch function type} {\sp{}}.
+The expression {\createmultiexp {\s{}} {\e{}}} expression creates a multimethod
+with interface type \s{} and dispatch function \e{}.
+The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
+takes multimethod \e{m} and returns a new multimethod with an extended dispatch
+table from dispatch value {\e{v}} to {\e{f}}.
+The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
+with dispatch function {\v{}} and dispatch table {\disptable{}}.
+
+The typing rules are in \figref{main:figure:mmsyntax}.
+The T-DefMulti rule checks to ensure the type of dispatch function
+is a function that has at least as permissive a parameter type
+as the interface type.
+
+For example, let us check the \defmultiliteral{} from our translation of \egref{example:rep}
+using T-DefMulti.
+
+$$
+\judgement{\propenv{}}
+{\hastype
+{\createmultiexp 
+      {\s{}}
+      {\classconst{}}}
+  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+$$
+where \s{}  = {\ArrowOne {\x{}} {\Top{}} {\t{}} {\filterset {\topprop{}} {\topprop{}}} {\emptyobject{}}}
+  and \sp{} = {\ArrowOne {\x{}} {\Top{}} {\Union{\Nil}{\Class}} {\filterset {\topprop{}} {\topprop{}}} {\path{\classpe{}}{\x{}}}}.
+  Since the parameter types agree, this is well-typed.
+
+The T-DefMethod rule is carefully constructed to ensure we have a syntactic
+lambda expression as the right-most subexpression.
+This way we can manually check the body of the lambda under an extended
+environment like we sketched in \egref{example:incmap}.
+We use \isacompareliteral{} to infer the proposition for this method.
+
+Let us continue with the next line of the translation of \egref{example:rep}.
+From our previous line we have \propenvp{} = \propenv{},{\isprop{\MultiFntype {\s{}}{\sp{}}}{path}}.
+$$
+\judgement{\propenvp{}}
+{\hastype
+  {\extendmultiexp {prop} {\String}
+                   {\abs {\x{}} {\Top{}} {\x{}}}}
+  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+$$
+We know \emph{prop} is a multimethod by \propenvp{}, so now we check the body.
+$$
+\judgement{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
+{\hastype
+  {\x{}}
+  {\String}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+$$
+This is checked by T-Local since {\inpropenv{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
+The new proposition {\isprop{\String}{\x{}}} was inferred by 
+$$
+  \isacompare{\Top{}}{\path{\classpe{}}{\x{}}}{\Value{\File{}}}
+             {\filterset{\isprop{\String}{\x{}}}
+                        {\notprop{\String}{\x{}}}}
+$$
+
+The next line is similarly. The final line is checked by T-App because
+subtyping for multimethods (\figref{main:figure:mmsyntax}), specifically S-PMultiFn
+says that {\MultiFntype {\s{}}{\sp{}}} can be upcast to its interface type {\s{}}.
+Since {\s{}} has parameter type \Top{}, this holds trivially.
 
@@
 B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
-which \getmethodliteral{} uses to choose the method to invoke.
-
-The typing rules (\figref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
-where \s{} is the interface type for beta reduction, and \t{} is the type for
-the dispatch function. T-DefMulti uses the provided interface type and dispatch function
-to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
-that must be true if this method is being invoked. The example can be type checked without
-changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
-method bodies. T-App handles beta reduction because multimethods are subtypes of functions
-(\figref{main:figure:mmsyntax}).
+and uses \getmethodliteral{} (which models Clojure's \clj{get-method}) 
+to obtain a method. The call to \getmethodliteral{} only returns a value
+if there is exactly one method such that the corresponding dispatch value
+is compatible with the current result of the dispatch function by \isaopsemliteral{}.
+Finally we return the result of applying the obtained method and the original evaluated argument.
 

### 2015-02-28 formal-model.tex
3.4

@@
   \begin{mathpar}
+    \HMapsubtyping{}
+  \end{mathpar}
+  \begin{mathpar}
     {\BAssoc}
@@
   \end{mathpar}
-  \begin{mathpar}
-    \HMapsubtyping{}
-  \end{mathpar}
-  \caption{HMap Terms, Types, Semantics, Subtyping and Typing Rules}
+  \caption{HMap Syntax, Typing and Operational Semantics}
   \label{main:figure:hmapsyntax}
@@
   \footnotesize
+
   $$
   \begin{altgrammar}
-    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
+  \end{altgrammar}
+  $$
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
+            &\alt& \reflectiveexpsyntax{} \\
+     &\alt& \mininonreflectiveexpsyntax{}
     \\
@@
   $$
+  \begin{mathpar}
+    {\TNewStatic}
+
+    {\TMethodStatic}
+
+    %{\TInstance}
+  \end{mathpar}
  \classtablelookupsyntax{}
@@
   \end{altgrammar}
-\end{mathpar}
-  \begin{mathpar}
-    {\TNewStatic}
-
-    {\TMethodStatic}
-
-    %{\TInstance}
 \end{mathpar}
@@
   \end{mathpar}
-  \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
+  \caption{Java Interoperability Syntax, Typing and Operational Semantics}
   \label{main:figure:javatyping}
@@
 
-  $$
-  \begin{altgrammar}
-    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
-            &\alt& \reflectiveexpsyntax{} 
-  \end{altgrammar}
-  $$
-
-We extend the syntax with type hinted expressions, Java field lookups and calls to
-methods and constructors. We model the syntax after the `dot' special
+We extend the syntax in \figref{main:figure:javatyping} with
+type hinted expressions, reflective and non-reflective Java field lookups and calls to
+methods and constructors. 
+We model the syntax after the `dot' special
 form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
@@
 and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
-The Java expressions come without typing rules because Typed Clojure
+The reflective expressions come without typing rules because Typed Clojure
 only reasons about resolved reflection
 (as demonstrated in \secref{sec:overviewjavainterop}).
-
-Now we model the compiler's reflection-resolution algorithm.
-\figref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
-The method ${\methodstaticexp {\classhint{1}}
+%
+The method call ${\methodstaticexp {\classhint{1}}
                              {\overrightarrow {\classhint{i}}}
@@
 ${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
-The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
+The field access {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
 calls the field on class {\classhint{1}} with Java signature
 ${\classhint{2}}\ \fld{};$.
-The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
+The constructor invocation {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                                {\class{}} {\overrightarrow{\e{i}}}}
@@
 
-To obtain non-reflective calls we define our rewrite relation
+We model Clojure's reflection resolution algorithm as a 
+rewrite relation
 \rewrite {\taenv{}} {\e{}} {\ep{}}
@@
 \taenv{} and Java class table \ct{}.
-
+%
 For example,
@@
 
-Type hint inference is modelled by
+Type hint inference is given by the judgement
 \tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
-which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
+which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in type hint environment \taenv{}.
+We defer the remainder of the rules for rewriting and the definition of type hint inference
+to the supplemental material.
 %$$
@@
 
-For demonstration purposes let us rewrite
-a simple field expression, with the assumptions that
-that we have a class \Point{} with a field \getx{} of Java type \Number
-and \intaenv {\taenv{}} {\x{}} {\Point{}}.
+As an example, we rewrite
+a simple field reference, with the assumptions that
+that the class \Point{} has a field $x$ of Java type \Number
+and \intaenv {\taenv{}} {p} {\Point{}}.
 In rewriting
-the expression \fieldexp{\getx}{\x{}},
-{\Point{}} is inferred as the type hint of \x{}
+the expression \fieldexp{x}{p},
+{\Point{}} is inferred as the type hint of $p$
 and \Number{} is the field type
@@
 $$
-\fieldstaticexp{\Point}{\Number}{\getx}{\x{}}.
+\fieldstaticexp{\Point}{\Number}{x}{p}.
 $$
@@
 
-To continue our example, let us assume \propenv{} = {\isprop{\Point}{\x{}}}.
-T-FieldStatic checks \x{} under the non-nilable type {\Point}
-and returns the nilable type {\Union{\Nil}{\Number}}.
+To continue our example, assume \propenv{} = {\isprop{\Point}{p}}.
+T-FieldStatic therefore produces the type
+{\Union{\Nil}{\Number}}
+for the entire expression.
 
-The other rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
-in the choice of nilability in the conversion function---method returns
-are nilable and constructor returns are non-nilable.
+The rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
+in the choice of nilability in the conversion function---methods
+can return nil but constructors cannot.
 
-The operational semantics B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
-arguments and call the relevant JVM operation---\secref{sec:metatheory}
+The evaluation rules B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
+arguments and call the relevant JVM operation, which do not model---\secref{sec:metatheory}
 states our exact assumptions about each.
@@
   \begin{mathpar}
-    \Multisubtyping{}
+    \TDefMulti{}
+
+    \TDefMethod{}
   \end{mathpar}
@@
   \begin{mathpar}
-    \TDefMulti{}
-
-    \TDefMethod{}
+    \Multisubtyping{}
   \end{mathpar}
-  \getmethodfigure{}
   \begin{mathpar}
@@
   \end{mathpar}
-\caption{Multimethod Syntax, Subtyping, Typing Rules and Semantics}
+  \getmethodfigure{}
+\caption{Multimethod Syntax, Typing and Operational Semantics}
 \label{main:figure:mmsyntax}
@@
 \end{minted}
-where subsequent bindings in the binder is sugar for nested let expressions.
 

### 2015-02-28 metatheory.tex
3.4

@@
 Our proof for type soundness is similar to~\citet{TF10}. We add
-errors and \wrong{} value and prove well-typed programs
-rule out wrong programs.
+errors and \wrong{} value and prove
+well-typed programs do not go wrong.
 
@@
 We use an extra lemma to support our main soundness lemma. Consistency
-ensures that occurrence typing does not claim the outer 
-\x{} is a keyword.
-
-\begin{minted}{clojure}
-(let [x 1]
-  ((let [x :a]
-     (fn [] x))))
-\end{minted}
+ensures that occurrence typing does not refer to variables
+hidden inside a closure.
 

### 2015-02-28 formal-model.tex
figure 3.2

@@
   \end{mathpar}
-  \caption{Subtyping rules}
+  \caption{Core Subtyping rules}
   \label{main:figure:subtyping}
@@
 
-\subsection{Heterogeneous maps}
+\subsection{Precise Types for Heterogeneous maps}
 \label{sec:hmapformal}
@@
 
-We now present \HMapliteral{} types in detail.
-\figref{main:figure:hmapsyntax} extends the syntax for terms and types.
-
-We write \HMapgeneric{\mandatory{}}{\absent{}}
-where {\mandatory{}} is a map of \emph{present} entries (mapping keywords to types),
-\absent{} is a set of keyword keys that are known to be \emph{absent}
+\figref{main:figure:hmapsyntax} presents syntax, typing rules
+and dynamic semantics in detail.
+%
+The type \HMapgeneric{\mandatory{}}{\absent{}}
+includes {\mandatory{}}, a map of \emph{present} entries (mapping keywords to types),
+\absent{}, a set of keyword keys that are known to be \emph{absent}
 and
-\completenessmeta{} is {\complete{}} (``complete'') if the map is fully specified by \mandatory{},
+tag \completenessmeta{} which is either {\complete{}} (``complete'') if the map is fully specified by \mandatory{},
 and {\partial{}} (``partial'') if there are \emph{unknown} entries.
 To ease presentation, 
-if a HMap is declared as \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
+if an HMap has completeness tag \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
 %\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
@@
 
-Subtyping for HMaps (\figref{main:figure:hmapsyntax})
+Subtyping for HMaps
 designate \MapLiteral{} as a common supertype for all HMaps.
-S-HMap says that a HMap is a subtype of another HMap if they agree
+S-HMap says that an HMap is a subtype of another HMap if they agree
 on \completenessmeta{}, agree on mandatory entries with subtyping
@@
 
-\figref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
+The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
 that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
@@
 next section discusses this in detail.
-Finally T-AssocHMap extends a HMap with a mandatory entry while preserving completeness
+Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
 and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
@@
 
-\figref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
-and \assocliteral{}.
-The B-Get rule evaluates first evaluates the map and the key, asserts
-the key must be present in the map and returns the associated key. If
-the entry is missing, B-GetMissing says we return \nil{}.
+The semantics for \getliteral{} and \assocliteral{} are straightforward.
+If the entry is missing, B-GetMissing produces \nil{}.
 
@@
 Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
-then \equivapp{\e{}}{\makekw{if}} has the proposition set 
+then \equivapp{\e{}}{\makekw{if}} has the true and false propositions
 $$
@@
 
-\subsection{Multimethod dispatch mechanism: \isaliteral}
+\subsection{Multimethod preliminaries: \isaliteral}
 
@@
 
-We now consider the core dispatch mechanism for multimethods. 
+We now consider the \isaliteral{} operation, a core part of the dispatch mechanism for multimethods. 
 Recalling the examples in \secref{sec:multioverview},
@@
 
-The main interest for the T-IsA typing rule is the \isacompareliteral{} 
+The key component of the T-IsA rule is the \isacompareliteral{} 
 metafunction
 (\figref{main:figure:mmsyntax}), used to calculate the propositions for
-\isaliteral{} expressions.
+\isaliteral{} tests.
 
@@
 
-To demonstrate,
+As an example,
 \isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
-has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.
+has the true and false propositions 
+\isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
+meaning that if this expression produces true, \x{} is a keyword, otherwise it is not.
 
-B-IsA models the semantics. \isaopsemliteral{} explicitly handles classes in the second case.
-
-$$
-\begin{array}{ll}
-  \vcenter{\hbox{\BIsA{}}}
-  &
-  \vcenter{\hbox{\isaopsemfigure{}}}
-\end{array}
-$$
+The operational behavior of \isaliteral{} is given by B-IsA (\figref{main:figure:mmsyntax}). \isaopsemliteral{} explicitly handles classes in the second case.
 
@@
   \getmethodfigure{}
+$$
+\begin{array}{ll}
+  \vcenter{\hbox{\BIsA{}}}
+  &
+  \vcenter{\hbox{\isaopsemfigure{}}}
+\end{array}
+$$
 \caption{Multimethod Syntax, Typing and Operational Semantics}

### 2015-02-28 formal-model.tex
formatting

@@
 
-\begin{figure*}
+\begin{figure}
   \footnotesize
   \begin{mathpar}
-    \standardsubtyping{}
+    \SRefl{}
+
+    \STop{}
+
+\SUnionSuper{}
+
+\SUnionSub{}
+
+\SFunMono{}
+
+\SObject{}
+
+\SClass{}
+
+\SSBool{}
+
+\SFun{}
+
+\SSKw{}
+
   \end{mathpar}
@@
   \label{main:figure:subtyping}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \footnotesize
-
-  $$
-  \begin{altgrammar}
-  \end{altgrammar}
-  $$
   $$
@@
 arguments and call the relevant JVM operation, which do not model---\secref{sec:metatheory}
-states our exact assumptions about each.
+states our exact assumptions.
 
@@
 \isaliteral{} tests.
-
 \begin{mathpar}
@@
 
-\begin{figure}
+\begin{figure}[t!]
   \footnotesize
@@
 To ease presentation, we present \emph{immutable}
-multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
+multimethods, with syntax and semantics given in \figref{main:figure:mmsyntax}. 
+\defmethodliteral{} returns a new extended multimethod
 without changing the original multimethod. \egref{example:rep} is now written
@@
 
-The syntax for multimethods is given in \figref{main:figure:mmsyntax}.
-{\MultiFntype {\s{}} {\sp{}}} is the type of multimethods with \emph{interface type}
+The type {\MultiFntype {\s{}} {\sp{}}} characterizes multimethods with \emph{interface type}
 {\s{}} and \emph{dispatch function type} {\sp{}}.
-The expression {\createmultiexp {\s{}} {\e{}}} expression creates a multimethod
+The expression {\createmultiexp {\s{}} {\e{}}} defines a multimethod
 with interface type \s{} and dispatch function \e{}.
 The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
-takes multimethod \e{m} and returns a new multimethod with an extended dispatch
-table from dispatch value {\e{v}} to {\e{f}}.
+extends multimethod \e{m} and to map
+dispatch value {\e{v}} to {\e{f}} in an extended dispatch table.
 The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
@@
 
-The typing rules are in \figref{main:figure:mmsyntax}.
-The T-DefMulti rule checks to ensure the type of dispatch function
-is a function that has at least as permissive a parameter type
+The T-DefMulti rule ensures that the type of the dispatch function
+has at least as permissive a parameter type
 as the interface type.
-
-For example, let us check the \defmultiliteral{} from our translation of \egref{example:rep}
+%
+For example, we can check the definition from our translation above of \egref{example:rep}
 using T-DefMulti.
-
 $$
-\judgement{\propenv{}}
+\judgement{}%{\propenv{}}
 {\hastype
@@
 This way we can manually check the body of the lambda under an extended
-environment like we sketched in \egref{example:incmap}.
-We use \isacompareliteral{} to infer the proposition for this method.
+environment as sketched in \egref{example:incmap}.
+We use \isacompareliteral{} to compute the proposition for this method,
+since \isaliteral{} is used at runtime in multimethod dispatch.
 
-Let us continue with the next line of the translation of \egref{example:rep}.
-From our previous line we have \propenvp{} = \propenv{},{\isprop{\MultiFntype {\s{}}{\sp{}}}{path}}.
+We continue with the next line of the translation of \egref{example:rep}.
+From the previous line we have \propenv{} = {\isprop{\MultiFntype {\s{}}{\sp{}}}{path}},
+so
 $$
-\judgement{\propenvp{}}
+\judgement{\propenv{}}
 {\hastype
@@
 $$
-We know \emph{prop} is a multimethod by \propenvp{}, so now we check the body.
+We know \emph{prop} is a multimethod by \propenv{}, so now we check the body
+of this method.
 $$
-\judgement{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
+\judgement{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
 {\hastype
@@
 $$
-This is checked by T-Local since {\inpropenv{\propenvp{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
-The new proposition {\isprop{\String}{\x{}}} was inferred by 
+%This is checked by T-Local since {\inpropenv{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
+The new proposition {\isprop{\String}{\x{}}} is derived by 
 $$
@@
              {\filterset{\isprop{\String}{\x{}}}
-                        {\notprop{\String}{\x{}}}}
+                        {\notprop{\String}{\x{}}}}.
 $$
+%
+The body of the \clj{let} is checked by T-App because
+{\MultiFntype {\s{}}{\sp{}}} is a subtype of its interface type {\s{}}.
 
-The next line is similarly. The final line is checked by T-App because
-subtyping for multimethods (\figref{main:figure:mmsyntax}), specifically S-PMultiFn
-says that {\MultiFntype {\s{}}{\sp{}}} can be upcast to its interface type {\s{}}.
-Since {\s{}} has parameter type \Top{}, this holds trivially.
-
-Multimethod semantics are in \figref{main:figure:mmsyntax}.
-B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
-B-DefMethod returns a new multimethod with an extended dispatch table.
-B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
+Multimethod definition semantics are straightforward.
+B-DefMulti creates a multimethod with the given dispatch function and an empty dispatch table.
+B-DefMethod produces a new multimethod with an extended dispatch table.
+B-BetaMulti invokes the dispatch function with the evaluated argument to obtain the dispatch value,
 and uses \getmethodliteral{} (which models Clojure's \clj{get-method}) 
-to obtain a method. The call to \getmethodliteral{} only returns a value
-if there is exactly one method such that the corresponding dispatch value
-is compatible with the current result of the dispatch function by \isaopsemliteral{}.
-Finally we return the result of applying the obtained method and the original evaluated argument.
+to extract the appropriate method. The call to \getmethodliteral{} only returns a value
+if there is \emph{exactly one} method such that the corresponding dispatch value
+is compatible, using \isaopsemliteral{}, with the result of the dispatch function.
+Finally we return the result of applying the extracted method and the original argument.
 

### 2015-02-28 formal-model.tex
meta

@@
 
+The full definition of \updateliteral{} is given in \figref{main:figure:update}
+which considers both keys a path elements as well as the \classconst{}
+path element described below.
+In the absence of paths, update simply performs set-theoretic operations
+on types; see \figref{main:figure:restrictremove} for details.
+
 \paragraph{Class path element} Our second path element \classpe{} is used in the latent
@@
 \end{mathpar}
-The full semantics are given in \figref{main:figure:primitivesem}.
+The dynamic semantics are given in \figref{main:figure:primitivesem}.
 The definition of \updateliteral{} supports various idioms relating to \classpe{}

### 2015-02-28 metatheory.tex
formatting
meta

@@
 
-Our proof for type soundness is similar to~\citet{TF10}. We add
-errors and \wrong{} value and prove
+We prove type soundness follow using the same technique as~\citet{TF10}. 
+We also include errors and a \wrong{} value and prove
 well-typed programs do not go wrong.
 
-We make our assumptions about Java explicit. We concede that
+Rather than modelling Java's dynamic semantics, we instead
+make our assumptions about Java explicit. We concede that
 method and constructor calls may diverge or error, but we assume they can
-never go wrong.
+never go wrong. (Assumptions for other operations are given in the supplemental
+material).
 
@@
 
-We use an extra lemma to support our main soundness lemma. Consistency
+For the purposes of our soundness proof, we require that all values
+are \emph{consistent}.
+Consistency
 ensures that occurrence typing does not refer to variables
@@
 {\wrongtheoremnoproof{main}}
-
-Null-pointer exceptions are ruled-out in typed code.
-
-{\nilinvoketheoremnoproof{main}}
+%
+%{\nilinvoketheoremnoproof{main}}

### 2015-02-28 related-work.tex
ocaml

@@
 \paragraph{Multimethods} 
-Millstein and Chambers~\cite{MS02}
-describe Dubious, a simple statically typed core language including multimethods that
-dispatch on the type of its arguments. They tackle a key challenge for statically typing
-multimethods: ``it is possible for two modules containing arbitrary multimethods to typecheck
-successfully in isolation but generate type errors when linked together.''~\cite{MS02}
+\citet{MS02} describe Dubious, a simple statically typed core language including multimethods that
+dispatch on the type of its arguments. 
+In Dubious methods declare the types of arguments that they expect which corresponds to
+exclusively using \clj{class} as the dispatch function in Typed Clojure.
 
@@
 \paragraph{Record Types} 
-O'Caml-style extensible record types have been the subject of extensive research 
-(eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91})
-most inference
+OCaml-style extensible record types (eg. \citet{Wan89}, \citet{CM91}, \citet{HP91})
+are defined polymorphically in terms of row variables.
+HMaps in Typed Clojure are instead designed to be used with subtyping.
 
@@
 \end{itemize}
-
-Typed Racket is a statically typed sister language of Racket. It
-attempts to preserve existing Racket idioms and aims to type check
-existing Racket code by simply adding top level type annotations~\cite{Tob10}.
-
-Typed Racket fully expands all macro calls before type checking~\cite{Tob10},
-avoiding the complex semantics of type checking macro definitions, an ongoing research area summarised
-by~\citet{Her10}. Typed Clojure also expands macros before type checking.
-
-\paragraph{Variable-arity Polymorphism}
-
-~\citet{STF09} invented a type system supporting variable-arity polymorphism  % doesn't fit on a line
-a version of which is included in the current implementation of Typed Racket.
-Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
-of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
-which are used to check definitions and usages of functions with non-trivial rest parameters
-
-\paragraph{Java Interoperability in Statically Typed Languages}
-
-Scala~\cite{OCD+} has nullable references for compatibility with Java.
-Programmers must manually check for
-\java{null} like in Java to avoid null-pointer exceptions.
-
-% def checkNullOrEmpty(v:Seq[Any]):Boolean = {
-%   return v.getClass
-% }
-% 
-% checkNullOrEmpty(null) ;=> NPE
-%
-% null.getClass

### 2015-02-28 related-work.tex
retic

@@
 \paragraph{Optional type systems}
-\begin{itemize}
-  \item Reticulated Python~\cite{Vitousek14}
+Reticulated Python~\cite{Vitousek14} is a gradually typed
+system for Python, implemented as a source-to-source
+translation that inserts dynamic checks at language boundaries
+and supporting Python's first-class object system.
+Typed Clojure is implemented as a syntactic check on Clojure
+code with no option to insert dynamic checks automatically.
+Typed Clojure does not support a first-class object system
+because Java has nominal classes, however HMaps offer a subset 
+of the structural features offered by Reticulated.
+
   \item GradualTalk

### 2015-02-28 related-work.tex

@@
 
+% Cite a few of the early papers here.
+%http://www.cs.washington.edu/research/projects/cecil/www/pubs/
 \paragraph{Multimethods} 
-\citet{MS02} describe Dubious, a simple statically typed core language including multimethods that
-dispatch on the type of its arguments. 
-In Dubious methods declare the types of arguments that they expect which corresponds to
-exclusively using \clj{class} as the dispatch function in Typed Clojure.
+\citeauthor{MS02} and collaborators present a sequence of
+systems~\cite{MS02, cecil, more} with statically-typed multimethods
+and modular type checking.  In contrast to Typed Clojure, in these
+system methods declare the types of arguments that they expect which
+corresponds to exclusively using \clj{class} as the dispatch function
+in Typed Clojure. However, Typed Clojure does not attempt to rule out
+failed dispatches at runtime.
 
@@
 
-\paragraph{Record Types} 
-OCaml-style extensible record types (eg. \citet{Wan89}, \citet{CM91}, \citet{HP91})
-are defined polymorphically in terms of row variables.
-HMaps in Typed Clojure are instead designed to be used with subtyping.
+\paragraph{Record Types} Row polymorphism~\cite{Wan89,CM91,HP91}, used
+in systems such as the OCaml object system, provides many of the
+features of HMap types, but defined using universally-quantified row
+variables. HMaps in Typed Clojure are instead designed to be used with
+subtyping, but nonetheless provide similar expressiveness, including
+the ability to require presence and absence of certain keys.
 
-Dependent JavaScript~\cite{Chugh:2012:DTJ} can track similar invariants as HMaps with types
-for JS objects. They must deal with mutable objects, they feature refinement types and strong updates to the heap
-to track changes to objects.
+Dependent JavaScript~\cite{Chugh:2012:DTJ} can track similar
+invariants as HMaps with types for JS objects. They must deal with
+mutable objects, they feature refinement types and strong updates to
+the heap to track changes to objects.
 
-Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track entries in a mutable Lua table.
-Typed Lua changes the dynamic semantics of Lua to accommodate mutability: Typed Lua raises a runtime error
-for lookups on missing keys---HMaps consider lookups on missing keys normal.
+Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track
+entries in a mutable Lua table.  Typed Lua changes the dynamic
+semantics of Lua to accommodate mutability: Typed Lua raises a runtime
+error for lookups on missing keys---HMaps consider lookups on missing
+keys normal.
+
+The integration of completeness information, crucial for many examples
+in Typed Clojure, is not provided by any of these systems.
+
+\paragraph{Java Interoperability in Statically Typed Languages}
+Scala~\cite{OCD+} has nullable references for compatibility with Java.
+Programmers must manually check for
+\java{null} as in Java to avoid null-pointer exceptions.
 

### 2015-02-28 related-work.tex
fix
records

--- 2015-02-28 related-work.tex
@@
 
-\paragraph{Record Types} 
-OCaml-style extensible record types (eg. \citet{Wan89}, \citet{CM91}, \citet{HP91})
-are defined polymorphically in terms of row variables.
-HMaps in Typed Clojure are instead designed to be used with subtyping.
+\paragraph{Record Types} Row polymorphism~\cite{Wan89,CM91,HP91}, used
+in systems such as the OCaml object system, provides many of the
+features of HMap types, but defined using universally-quantified row
+variables. HMaps in Typed Clojure are instead designed to be used with
+subtyping, but nonetheless provide similar expressiveness, including
+the ability to require presence and absence of certain keys.
 
-Dependent JavaScript~\cite{Chugh:2012:DTJ} can track similar invariants as HMaps with types
-for JS objects. They must deal with mutable objects, they feature refinement types and strong updates to the heap
-to track changes to objects.
+Dependent JavaScript~\cite{Chugh:2012:DTJ} can track similar
+invariants as HMaps with types for JS objects. They must deal with
+mutable objects, they feature refinement types and strong updates to
+the heap to track changes to objects.
 
-Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track entries in a mutable Lua table.
-Typed Lua changes the dynamic semantics of Lua to accommodate mutability: Typed Lua raises a runtime error
-for lookups on missing keys---HMaps consider lookups on missing keys normal.
+Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track
+entries in a mutable Lua table.  Typed Lua changes the dynamic
+semantics of Lua to accommodate mutability: Typed Lua raises a runtime
+error for lookups on missing keys---HMaps consider lookups on missing
+keys normal.
+
+The integration of completeness information, crucial for many examples
+in Typed Clojure, is not provided by any of these systems.
 

### 2015-02-28 related-work.tex
Related work.

--- 2015-02-28 related-work.tex
@@
 \paragraph{Optional type systems}
-Reticulated Python~\cite{Vitousek14} is a gradually typed
-system for Python, implemented as a source-to-source
-translation that inserts dynamic checks at language boundaries
-and supporting Python's first-class object system.
-Typed Clojure is implemented as a syntactic check on Clojure
-code with no option to insert dynamic checks automatically.
-Typed Clojure does not support a first-class object system
-because Java has nominal classes, however HMaps offer a subset 
-of the structural features offered by Reticulated.
-
+\begin{itemize}
+  \item Reticulated Python~\cite{Vitousek14}
   \item GradualTalk

### 2015-02-28 related-work.tex
related

--- 2015-02-28 related-work.tex
@@
 \citeauthor{MS02} and collaborators present a sequence of
-systems~\cite{MS02, cecil, more} with statically-typed multimethods
+systems~\cite{Chambers:1992:OMC, Chambers:1994:TMM, MS02} with statically-typed multimethods
 and modular type checking.  In contrast to Typed Clojure, in these
@@
 
-\paragraph{Record Types} Row polymorphism~\cite{Wan89,CM91,HP91}, used
+\paragraph{Record Types} Row polymorphism~\cite{Wand89typeinference,CM91,HP91}, used
 in systems such as the OCaml object system, provides many of the
@@
 
-  \item GradualTalk
-  \item Flow
-\end{itemize}
+%  \item GradualTalk
+%  \item Flow
+%\end{itemize}

### 2015-02-28 related-work.tex

@@
 
-\paragraph{Optional type systems}
-Reticulated Python~\cite{Vitousek14} is a gradually typed
-system for Python, implemented as a source-to-source
-translation that inserts dynamic checks at language boundaries
-and supporting Python's first-class object system.
-Typed Clojure is implemented as a syntactic check on Clojure
-code with no option to insert dynamic checks automatically.
-Typed Clojure does not support a first-class object system
-because Java has nominal classes, however HMaps offer a subset 
-of the structural features offered by Reticulated.
+\paragraph{Other optional and gradual type systems}
+In addition to Typed Racket, several other gradual type
+systems have been developed recently, targeting existing
+dynamically-typed languages.  Reticulated Python~\cite{Vitousek14} is
+an experimental gradually typed system for Python, implemented as a
+source-to-source translation that inserts dynamic checks at language
+boundaries and supporting Python's first-class object system.  Typed
+Clojure does not support a first-class object system because Java (and
+Clojure) have nominal classes, however HMaps offer an alternative to
+the structural objects offered by Reticulated. Similarly,
+GradualTalk~\cite{gradualtalk} offers gradual typing for SmallTalk,
+with nominal classes.
+
+Optional types, requiring less implementation effort and avoiding any
+runtime cost, have been widely adopted in industry, including Hack, an
+extension to PHP~\cite{hack}, and Flow~\cite{flow} and
+TypeScript~\cite{typescript}, two extensions of JavaScript. These
+systems all support some form of occurrence typing, but not in the
+generality presented here, nor do they include the other features we
+have presented.
 

### 2015-02-28 overview.tex
3.4
cites

@@
 type checking dynamically typed programs without compromising
-soundness, which is appealing over success typing~\cite{todo-dialyzer-paper}
+soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
 which cannot prove strong properties about programs
-and soft typing~\cite{successtyping-todo}
+and soft typing~\cite{CF91}
 which has proved too complicated in practice.
@@
 
-\begin{figure}
-\begin{exmp}
-\inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
-\label{example:multidispatch}
-\end{exmp}
-\end{figure}
-
 We call three different \clj{File} constructors, each known at compile-time
@@
 \begin{exmp}
+\inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
+\label{example:multidispatch}
+\end{exmp}
+\begin{exmp}
 \inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
@@
 \end{exmp}
+\caption{Multimethod Examples}
 \end{figure}

### 2015-02-28 related-work.tex
optional and gradual

@@
 \citeauthor{MS02} and collaborators present a sequence of
-systems~\cite{Chambers:1992:OMC, Chambers:1994:TMM, MS02} with statically-typed multimethods
+systems~\cite{MS02, cecil, more} with statically-typed multimethods
 and modular type checking.  In contrast to Typed Clojure, in these
@@
 
-\paragraph{Record Types} Row polymorphism~\cite{Wand89typeinference,CM91,HP91}, used
+\paragraph{Record Types} Row polymorphism~\cite{Wan89,CM91,HP91}, used
 in systems such as the OCaml object system, provides many of the

### 2015-02-28 conclusion.tex
Related work.
missing cites.

@@
 
+% Delete the following paragraphs if space is needed.
+
 However, there is much more that Typed Clojure can provide. Most
@@
 mechanisms for contracts and proxies in Java and
-Clojure~\cite{some-stuff} to enable sound gradual typing for Clojure.
+Clojure to enable sound gradual typing for Clojure.
 
@@
 effort to analyze and understand the relevant features and idioms and
-develop new type checking approaches for them.
+develop new type checking approaches.

### 2015-02-28 overview.tex
sub
egref

@@
 
+\newpage
 Here is the above program in Typed Clojure.
-
 \begin{exmp}
@@
 expression.
-
+\newpage
 \begin{exmp}
@@
 
-We include the above sequence of definitions as example~\ref{example:rep}.
+We include the above sequence of definitions as \egref{example:rep}.
 
@@
 \paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
-In example~\ref{example:incmap} we
+In \egref{example:incmap} we
 dispatch on the \clj{:op} key 
@@
 
-\clj{inc-map} is like example~\ref{example:decmap} except the nodes are incremented.
+\clj{inc-map} is like \egref{example:decmap} except the nodes are incremented.
 The reasoning is similar, except we only consider one branch (the current method) by
@@
 
-Example~\ref{example:multidispatch}
+\egref{example:multidispatch}
 simulates multiple dispatch by dispatching on
@@
 via type hints.
-Multiple dispatch follows the same kind of reasoning as example~\ref{example:incmap},
+Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
 except we update multiple bindings simultaneously.
@@
 
-Example~\ref{example:final}
+\egref{example:final}
 combines everything we will cover for the rest of the paper:

### 2015-03-01 conclusion.tex
corrections

@@
 the Clojure community among both enthusiasts and professional
-programmers and recieves contributions from many developers.
+programmers and receives contributions from many developers.
 
@@
 significantly, Typed Clojure currently does not provide \emph{gradual
-  typing}---ineraction between typed and untyped code is unchecked and
-thus unsound. We hope to explore the possibilites of using existing
+  typing}---interaction between typed and untyped code is unchecked and
+thus unsound. We hope to explore the possibilities of using existing
 mechanisms for contracts and proxies in Java and
@@
 Additionally, the Clojure compiler is unable to use Typed Clojure's
-wealth of static information to optimze programs. Addressing this
+wealth of static information to optimize programs. Addressing this
 requires not only first enabling sound gradual typing, but also

### 2015-03-01 formal-model.tex
formal model
corrections

@@
 type proposition is called the \emph{object}.
-The typing judgement relates an object to every expression in the language.
+The typing judgment relates an object to every expression in the language.
 An object is either \emph{empty}, written \emptyobject{}, 
@@
 new information about our programs
-with the judgement \inpropenv{\propenv{}}{\prop{}}.
+with the judgment \inpropenv{\propenv{}}{\prop{}}.
 In addition to the standard rules for the logical connectives, the key
@@
 For instance, \egref{example:conditionalflow}
-only utilises \clj{x} once
+only utilizes \clj{x} once
 the programmer is convinced it is safe to do so based whether
@@
 
-We formalise our type system following~\citet{TF10}
+We formalize our type system following~\citet{TF10}
 (with differences highlighted in $\trdiff{\text{blue}}$).
-The typing judgement 
+The typing judgment 
 $$
@@
 $$
-because we can now satisify the premise of T-Local:
+because we can now satisfy the premise of T-Local:
 $$
@@
 type system.
-The main judgement is \opsem{\openv{}}{\e{}}{\definedreduction{}}
+The main judgment is \opsem{\openv{}}{\e{}}{\definedreduction{}}
 which states that \e{} evaluates to answer \definedreduction{} in environment
@@
 
-\SSBool{}
+\begin{array}{l}
+\SSBool{}\\\\
+\SSKw{}
+\end{array}
 
 \SFun{}
-
-\SSKw{}
 
@@
 If we learn some proposition in both cases then we can use that proposition as an assumption to check the second subexpression.
-T-Do formalises this intuition.
+T-Do formalizes this intuition.
 
@@
 using propositions.
-The way to refer to \emph{parts} of the runtime environment with
-occurrencing is via \emph{path elements}.
-A \emph{path} consists of a series of path elements
-applied right-to-left to a variable
-written
-\path{\pathelem{}}{\x{}}.
+We refer to parts of the runtime environment via 
+a \emph{path} that consists of a series of
+\emph{path elements} applied right-to-left to a variable
+written \path{\pathelem{}}{\x{}}.
 \citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
@@
 
-Type hint inference is given by the judgement
+Type hint inference is given by the judgment
 \tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}

### 2015-03-01 metatheory.tex
corrections

@@
 
-Rather than modelling Java's dynamic semantics, we instead
+Rather than modeling Java's dynamic semantics, we instead
 make our assumptions about Java explicit. We concede that
@@
   \begin{proof}
-    By induction on the derivation of the typing judgement. 
+    By induction on the derivation of the typing judgment. 
     (Full proof given as lemma~\ref{appendix:lemma:soundness}).

### 2015-03-01 overview.tex
corrections

@@
 \clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
-non-destructively dissociating the raw \clj{:private-key} entry with \clj{dissoc}
+dissociating the raw \clj{:private-key} entry with \clj{dissoc}
 and associating an encrypted private key
@@
 
-The next example is the same except we use the \clj{:absent-keys} HMap option.
-We also
-utilize \emph{map destructuring}, which is binding position syntax for pattern matching.
-Parameters are replaced with maps of symbols to keywords that bind the symbols
-to lookups on that keyword, optionally terminated by \clj{:as}
-which aliases the parameter. For example
-\clj{(fn [{^File x :x, :as m}] ...)}
-expands to a \clj{let} binding \clj{m} to the first argument and 
-\clj{^File x} to \clj{(:x m)}.
-Clojure's \clj{let} takes a flat binding vector which can refer to previous bindings like Scheme's \mintinline{scheme}{let*},
-and a body expression.
+\egref{example:absentkeys}
+is like \egref{example:circleci}
+except the \clj{:absent-keys} HMap option is used
+instead of \clj{:complete?},
+which takes a \emph{set literal} of keywords that do not appear in the map, written 
+with \emph{\#}-prefixed braces.
+The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
+aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
+in the function body.
 
@@
 \inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
+\label{example:absentkeys}
 \end{exmp}
@@
 \end{exmp}
-
-% type hints give no guarantees
-% TODO
-
-The type hinting system and Typed Clojure's static type checking
-are separate, the latter predating the former by several years.
-The interaction between them is often not as obvious, for example Typed Clojure
-has an explicit type for \clj{null}
-null-pointer exceptions are impossible.
-
-do not need to take \clj{nil} into account,
+A type hint does not correspond to a static property so they are ignored by Typed Clojure.
+Type hints are still required, however, as Typed Clojure has no mechanism to provide
+optimisations (discussed in \secref{sec:experience}).
 
@@
 The second method is similar except it uses exceptional control flow.
-
-%% Old stuff vvvv
-
-% TODO references
-% why is TR such a good base?
-% - immutability
-% - common lisp ancestry
-% differences?
-% - Clojure is built on JVM
-% - interop with JVM
-% - Clojure's idiomatic primitives are different
-% - multimethods + protocols
-% - less sophisticated macro system
-%  - not an issue
-%  - implementation difference, AST walking vs syntax walking
-
-%Typed Clojure is a gradual type system for Clojure. It is designed
-%to type check normal Clojure code by adding annotations. It is implemented
-%as a library, and can be seamlessly included in any Clojure project; no
-%separate compiler or language is needed.
-%
-%{\smallsection {Based on Typed Racket}}
-%Initially, the similarities between untyped Racket and Clojure and Typed Racket's 
-%ability to type check Racket code led us to investigate a similar type system for Clojure.
-%After two years of development, the solid basis of Typed Racket 
-%helps us type check many Clojure idioms without significant differences
-%in implementation or theory. We found that extending Typed Clojure to check
-%those idioms that have no obvious Racket equivalent did not significantly alter the structure
-%of the type system.
-%
-%{\smallsection {Occurrence typing}}
-%\citet{TF08,TF10} developed \emph{occurrence typing}, which helps improve types at branches.
-%Typed Clojure uses occurrence typing in a similar way to Typed Racket, with
-%some extensions (discussed in in Section [?]). %FIXME
-%
-%{\smallsection {Practical Variable-Arity Polymorphism}}
-%Functions with non-trivial variable parameters are common in Racket.
-%For example, Racket provides \emph{map} which takes a function and a
-%variable number of collections and applies the function simultaneously
-%to each element of the provided collections, returning a list of results.
-%\citet*{STF09} developed a practical system that handles advanced variable parameters
-%which can handle applications of functions like \emph{map}.
-%
-%Clojure has a similar emphasis on variable-arity functions. In some ways,
-%Clojure's core library encourages even more complicated variable-parameter schemes.
-%The \emph{assoc} core function, for example, takes three parameters and
-%then a quantity of variable parameters that is a multiple of two.
-%This is beyond what Typed Racket (and Typed Clojure) can currently handle. 
-%
-%Functions like \emph{map} are common in Clojure, so we provide an implementation
-%of variable-arity polymorphism which has similar capabilities as Typed Racket's
-%implementation.
-%
-%{\smallsection {Local Type Inference}}
-%We use Pierce and Turner's Local Type Inference~\cite{PT00} to infer some polymorphic
-%applications. Our implementation is based on Typed Racket's, which has extensions
-%to support applications of polymorphic variable-arity functions like \emph{map}.
-%
-%{\smallsection {Unions and intersections}}
-%Like Typed Racket, we include union and ordered intersection types. Unions define
-%a least-upper-bound for a set of types. For example, we can express a type that is
-%either \Number or \Symbol by including them in a union: {\Union {\Number} {\Symbol}}
-%
-%Ordered intersections (described further by \citet{SA+12})
-%are used for overloading function types. We can express a function that takes
-%a \Number and returns a \Symbol, and vice-versa with an ordered intersection function type:
-%
-%\begin{lstlisting}[label=lst:ordered]
-%(Fn [Number -> Symbol]
-%    [Symbol -> Number])
-%\end{lstlisting}
-%
-%As our intersections are \emph{ordered}, we can express fine invariants in the
-%case where arity parameter types overlap. Similar to a pattern match, earlier arities 
-%are tried first, and the first arity to match ``wins''.
-%
-%For example, applying an \lstinline|Integer| argument to a function of type
-%
-%\begin{lstlisting}
-%(Fn [Integer -> Integer]
-%    [Number -> Number])
-%\end{lstlisting}
-%
-%returns an \lstinline|Integer|. Reversing the arities however gives
-%type \lstinline|Number|, because the arity taking a \lstinline|Number|
-%always matches first.
-%
-%{\smallsection {Hosted on the Java Virtual Machine}}
-%Clojure is built to run on the Java Virtual Machine (JVM),
-%offering good interoperability with existing Java code.
-%Typed Clojure helps programmers correctly call Java code
-%by integrating with Java's type system.
-%
-%We give Java arrays and Java's \emph{null} special treatment
-%when involved with interoperability. Arrays are treated as \emph{read-only}
-%when sourced from Java methods, discussed in Section \ref{sec:arrays}.
-%We are explicit, and conservative by default, in the positions where
-%Java's \emph{null} can be passed, discussed in Section \ref{sec:null}.

### 2015-04-22 formal-model.tex
add reviews

@@
 
-Syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
+The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
 application, abstractions, conditionals and let expressions.
@@
 As an example, we rewrite
-a simple field reference, with the assumptions that
+a simple field reference, with the assumptions
 that the class \Point{} has a field $x$ of Java type \Number
@@
 The evaluation rules B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
-arguments and call the relevant JVM operation, which do not model---\secref{sec:metatheory}
+arguments and call the relevant JVM operation, which we do not model---\secref{sec:metatheory}
 states our exact assumptions.
@@
                  (.getPath x))]
-      (path "dir/a")))) ;=> "a"
+      (path (File. "dir/a"))))) ;=> "dir/a"
 \end{minted}

### 2015-04-22 overview.tex
add reviews

@@
 and return types respectively.
-Typed Clojure provides a \clj{check-ns} function to type checks the current namespace.
+Typed Clojure provides a \clj{check-ns} function to type check the current namespace.
 \clj{number?} is
@@
 
-\clj{inc-map} is like \egref{example:decmap} except the nodes are incremented.
+\clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
 The reasoning is similar, except we only consider one branch (the current method) by

### 2015-05-15 intro.tex
corrections
delete ns form. delete some type hints. use `new` over dot.

@@
 \begin{figure}
-\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/parent2.clj}
+\inputminted[firstline=5]{clojure}{code/demo/src/demo/parent2.clj}
 \caption{A simple Typed Clojure program}
@@
 handling of Java's \java{null} (written \clj{nil}) in interoperation, as well as an
-extended form of occurrence typing and Clojure's \emph{type hints},
-which are central to Typed Clojure's approach to interoperability.
+extended form of occurrence typing and method resolution of
+Java interoperability based on static type information.
 
@@
 then obtains the parent by making a Java method call.
-%
-The annotation \clj{^File f} is a type hint on \clj{f}, which instructs the Clojure
-compiler (running prior to Typed Clojure typechecking) to statically
-resolve the \clj{getParent} 
-call to \clj{File}'s \clj{getParent} method with signature \java{String getParent();}, rather than using reflection at runtime.
 
@@
 
-%$$
-%\abs{\x{}}{\Union{\Nil}{\File}}
-%{\letexp {\hinted{\File{}}{f}} {f}
-%  {\ifexp {f} 
-%    {\letexp {s} {\methodstaticexp {\File{}} {\String{}} {\String{}} {getParent} {f}}
-%      {\doexp {\ifexp {s}
-%                {\nil{}}
-%                {\errorval{input}}}
-%              {\newstaticexp {\String{}} {\File{}} {\File{}} {s}}}}
-%    {\nil{}}
-%}}
-%              $$
+\subsection{Contributions}
 
-%(some-> f (.getParent) (File.) (.getParent))
+These are our main contributions.
 
-%\begin{lstlisting}
-%(ann parent [(U nil File) -> (U nil String)])
-%(defn parent [^File f]
-%  (when f
-%    (.getParent f)))
-%\end{lstlisting}
-
-%% \begin{minted}[]{clojure}
-%% (defalias Expr
-%%   (U '{:op :if :test Expr :then Expr :else Expr}
-%%      '{:op :const :val Any} ...))
-%% (ann tree-str [Expr -> String])
-%% (defmulti tree-str (fn [e :- Expr] (get :op e)))
-%% (defmethod tree-str :if [{:keys [test then else]}] 
-%%   (str (tree-str test) (tree-str then) (tree-str else)))
-%% (defmethod tree-str :const [{:keys [val]}] 
-%%   (str val))
-%% \end{minted}
-
-%% \subsection{Contributions}
-
-%% %The main idea is....... SPJ
-%% In this paper we present a practical type system for Clojure.
-
-%% \begin{itemize}
-
-%%   \item We formalise Typed Clojure and prove that typed code
-%%     \begin{itemize}
-%%       \item disallows null-pointer exceptions, (Section ?)
-%%       \item statically resolves Java interoperability (Section ?)
-%%     \end{itemize}
-%%   \item We reveal the surprising synthesis between Clojure multimethods
-%%         and occurrence typing.
-%%   \item We 
-%% \end{itemize}
-
-% what are the goals of Typed Clojure?
-% where do they differ from Typed Racket? why?
-
-% Things todo
-% - DONE small calculus based on previous TR formalisation
-%   that demos the ``flow'' environment & not/intersection type
-% - discussion and examples for flow filters
-% - discussion on the not/difference type, real code where it helps
-% - higher-rank polymorphism, show what it enables (eg. monads)
-%   - also show more complicated example, making an extensible
-%     type for clojure.core/conj, and why it's not straightforward
-% - discussion on limitations of local type inference
-% - demonstration of how we check multimethods
-% - dealing with array covariance
-% - note on how we avoid null-pointer exceptions
-% - discussion on why Typed Clojure only uses the ``static'' part
-%   of Typed Racket
-% - the equality filter problem (and draft solution?), a proposition
-%   that knows about binding aliasing. Sounds straightforward?
+\begin{itemize}
+  \item We motivate, describe and evaluate Typed Clojure, an optional type system for Clojure
+        that understands existing Clojure idioms.
+  \item We present the theory and implementation of a type system
+        based on \emph{occurrence typing} that statically prevents null-pointer 
+        exceptions in typed code in the presence of Java interoperability, 
+        applicable to existing JVM languages.
+  \item We prove type soundness for Typed Clojure.
+\end{itemize}

### 2015-05-15 overview.tex
delete ns form. delete some type hints. use `new` over dot.

@@
 
-\inputminted[firstline=1]{racket}{code/tr/example1.rkt}
+\inputminted[firstline=3]{racket}{code/tr/example1.rkt}
 
@@
 
-\newpage
 Here is the above program in Typed Clojure.
 \begin{exmp}
-\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
+\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
 \label{example:conditionalflow}
@@
 
-This is a regular Clojure file compiled with
-the Clojure compiler, with the \clj{ns} form
-declaring a
-\emph{namespace}
-for managing var and class imports.
-Here \clj{:require} declares a runtime dependency 
-on
-\clj{clojure.core.typed},
-Typed Clojure's core namespace, and
-\clj{:refer}
-brings a collection of vars into scope by name.
-The \clj{:refer-clojure} \clj{:exclude} option unmaps core vars from the current 
-namespace---here we unmap \clj{clojure.core/fn},
-which creates a function from a parameter vector and a body expression,
-and import a typed variant of \clj{fn}
-that supports type annotations.
-
-The typed \clj{fn} supports optional annotations by 
+The \clj{fn} macro (provided by core.typed) supports optional annotations by 
 adding
@@
 and return types respectively.
-Typed Clojure provides a \clj{check-ns} function to type check the current namespace.
 \clj{number?} is
@@
 
-The \clj{do} form sequences two expressions returning the latter,
-\clj{throw} corresponds to Java's \java{throw}
-and \clj{(class. args*)} is the syntax for Java constructors---that 
-is a class name with a dot suffix as the operator followed the arguments to the constructor.
-
-In this example a \clj{throw} expression guards \clj{(inc x)}, the increment function for numbers,
-from being evaluated if \clj{x} is \clj{nil}, preventing
+The fully expanded increment function in~\egref{example:doexception}
+guards its final call with a number check, preventing
 a possible null-pointer exception.
+Without this check, the type system would reject the program.
 
@@
 \footnote{See \url{https://github.com/typedclojure/examples}
-  for full examples. From here we omit \clj{ns} forms.}
+  for full examples.}
 We model this formally (section~\ref{sec:doformal}) and prove

### 2015-05-16 formal-model.tex
JVM ops are dynamically typed

@@
 \begin{minted}{clojure}
-...  (do (if (number? x) nil (throw (Exception.)))
-         (inc x)) ...
+...
+  (do (if (number? x) nil (throw (new Exception)))
+      (inc x)) 
+...
 \end{minted}
@@
 
-\subsection{Java Interoperability and Type Hints}
+\subsection{Java Interoperability}
 
@@
   \begin{altgrammar}
-    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
-            &\alt& \reflectiveexpsyntax{} \\
-     &\alt& \mininonreflectiveexpsyntax{}
+    \e{} &::=& \ldots \alt \reflectiveexpsyntax{} &\mbox{Expressions}\\
     \\
@@
 
-    \tatypesyntax{}\\
-    \typehintenvsyntax{}\\
     \classtableallsyntax{}
@@
   \begin{mathpar}
-    {\TNewStatic}
+    {\TNew}
 
-    {\TMethodStatic}
+    {\TMethod}
 
@@
   \end{mathpar}
-  \begin{mathpar}
-    %\RAbs{}
-%
-%    \RNewElimRefl{}
-%
-%    \RMethodElimRefl{}
-%
-%    \RFieldElimRefl{}
-%
-%    \RLet{}
-%
-
-%    \RLetHint{}
-  \end{mathpar}
   \caption{Java Interoperability Syntax, Typing and Operational Semantics}
@@
 
-In \secref{sec:overviewjavainterop} we discussed the role of type hints
-to help eliminate reflective calls.
-In this section, we introduce our model of Java and provide user-facing
-syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
-Then we model the Clojure compiler's compile-time reflection resolution
-algorithm.
-To achieve this, first we define notation for
-\emph{non-reflective} Java forms that 
-unambiguously call a field, method or constructor.
-Then we define a rewrite relation that uses
-type hints to resolve reflection explicitly.
-Finally we give typing rules that model how
-Typed Clojure interacts with non-reflective calls.
-
 We present Java interoperability in a restricted setting without class inheritance,
@@
 
-We extend the syntax in \figref{main:figure:javatyping} with
-type hinted expressions, reflective and non-reflective Java field lookups and calls to
+We extend the syntax in \figref{main:figure:javatyping} with Java field lookups and calls to
 methods and constructors. 
@@
 \clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
-and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
-The reflective expressions come without typing rules because Typed Clojure
-only reasons about resolved reflection
-(as demonstrated in \secref{sec:overviewjavainterop}).
-%
-The method call ${\methodstaticexp {\classhint{1}}
-                             {\overrightarrow {\classhint{i}}}
-                             {\classhint{2}}
-                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
-is a non-reflective call to the \mth{} method on class {\class{1}}, 
-with Java signature 
-${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
-The field access {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
-calls the field on class {\classhint{1}} with Java signature
-${\classhint{2}}\ \fld{};$.
-The constructor invocation {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
-                               {\class{}} {\overrightarrow{\e{i}}}}
-calls the constructor with Java signature
-${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.
+and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
 
-%\begin{figure}
-%  \footnotesize
-%  \begin{mathpar}
-%%    \TALocal{}
-%%
-%%    \TANil{}
-%%
-%%    \TANewStatic{}
-%%
-%%    \TALetHint{}
-%
-%%    \TALet{}
-%  \end{mathpar}
-%\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
-%\label{main:figure:hintinfer}
-%\end{figure}
-
-We model Clojure's reflection resolution algorithm as a 
-rewrite relation
-\rewrite {\taenv{}} {\e{}} {\ep{}}
-which rewrites \e{} to a possibly-less reflective expression
-\ep{} with respect to type hint environment
-\taenv{} and Java class table \ct{}.
-%
-For example,
-R-FieldElimRefl emits a non-reflective field if it can
-find a field matching the type hint inferred on \ep{}.
-  $$
-    \RFieldElimRefl{}
-  $$
-
-Type hint inference is given by the judgment
-\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
-which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in type hint environment \taenv{}.
-We defer the remainder of the rules for rewriting and the definition of type hint inference
-to the supplemental material.
-%$$
-%\TALetHint{}
-%$$
-
-As an example, we rewrite
-a simple field reference, with the assumptions
-that the class \Point{} has a field $x$ of Java type \Number
-and \intaenv {\taenv{}} {p} {\Point{}}.
-In rewriting
-the expression \fieldexp{x}{p},
-{\Point{}} is inferred as the type hint of $p$
-and \Number{} is the field type
-by \fieldtypeliteral{} (\figref{main:figure:javatyping}).
-Now we just plug in the new information into our new expression
-$$
-\fieldstaticexp{\Point}{\Number}{x}{p}.
-$$
-
-Now we present the typing rules for resolved Java interoperability.
-T-FieldStatic checks a resolved field expression by ensuring the target has
+T-Field checks a resolved field expression by ensuring the target has
 the correct static type, then returns a nilable type corresponding the Java type.
@@
 \begin{mathpar}
-    {\TFieldStatic}
+    {\TField}
 \end{mathpar}
 
-To continue our example, assume \propenv{} = {\isprop{\Point}{p}}.
-T-FieldStatic therefore produces the type
-{\Union{\Nil}{\Number}}
-for the entire expression.
-
-The rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
+The rules T-Method and T-New work similarly (\figref{main:figure:javatyping}), varying
 in the choice of nilability in the conversion function---methods
-can return nil but constructors cannot.
+can return nil but constructors cannot. Void also does not have a constructor.
 

### 2015-05-16 overview.tex
JVM ops are dynamically typed

@@
 
-Unlike Java, Clojure is dynamically typed. We have no type information about \clj{f}
-but we still need to pick a method to call. The Clojure compiler delegates
-the choice to runtime using \emph{Java reflection}.
-Unfortunately reflection is slow and unpredictable, so Clojure supports \emph{type hints}
-to help eliminate it where possible,
-
-\begin{minted}{clj}
-  (fn [^File f] (.getParent f))
-\end{minted}
-
-Symbols support \emph{metadata}---the syntax \clj{^File f}
-is a single expression that is a symbol \clj{f} with metadata \clj{{:tag File}}.
-In binding positions like \clj{(fn [^File f] ...)}
-syntactic occurrences preserve metadata.
-
-The Clojure compiler uses the type hint
-to statically resolve 
-the method call to the
-\mintinline{java}{public String getParent()}
-method of \clj{java.io.File}.
-The call to \clj{getParent} is unambiguous at runtime but type 
-checking fails---Typed Clojure
-considers \clj{f} to be of type \clj{Any}, which is unsafe 
-to use as the target of even a resolved method.
-If instead we annotate the function parameter with type \clj{File},
-we get a static type error.
-
-\begin{minted}{clj}
-  (fn [f :- File] (.getParent f)) ;; type error
-\end{minted}
-
-Typed Clojure disallows reflection in typed code
-so we must add back the type hint to obtain a well-typed
-expression.
-\newpage
-\begin{exmp}
-\inputminted[firstline=10,lastline=10]{clojure}{code/demo/src/demo/parent3.clj}
-\end{exmp}
-A type hint does not correspond to a static property so they are ignored by Typed Clojure.
-Type hints are still required, however, as Typed Clojure has no mechanism to provide
-optimisations (discussed in \secref{sec:experience}).
-
 \begin{exmp}
@@
 \clj{File} to the function
-\clj{(fn [^File x] (.getPath x))}
+\clj{(fn [x] (.getPath x))}
 in the dispatch table.
@@
 $$
-\clj{(path (File. "dir/a"))}
+\clj{(path (new File "dir/a"))}
 $$
@@
 $$
-\clj{((fn [^File x] (.getPath x)) "dir/a")}.
+\clj{((fn [x] (.getPath x)) "dir/a")}.
 $$
@@
 \begin{exmp}
-\inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
+\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
 \label{example:multidispatch}

### 2015-05-20 formal-model.tex
flatten \hastype in \judgement

@@
 {\judgement   {\propenv}
-              {\hastype {\e{}} {\t{}}}
+              {\e{}} {\t{}}
   {\filterset {\thenprop {\prop{}}}
@@
 and object \object{}. We write 
-{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.
+{\judgementtwo{\propenv}{\e{}} {\t{}} if we are only interested in the type.
 
@@
 $$
-\judgement{\propenv{}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\propenv{}}{\appexp{\numberhuh{}}{\x{}}}{\Boolean}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\hastype{\x{}}{\Number{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
+\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\x{}}{\Number{}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\Nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\isprop{\Number}{\x{}}}{\nil{}}{\Nil{}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 $$
-\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\notprop{\Number}{\x{}}}{\errorvalv{}}{\Bot}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
 $$
@@
 \judgement{\isprop{\UnionNilNum}{\x{}}}
-          {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
+          {\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}
           {\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
@@
 in \secref{sec:isaformal}.
-Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
+Intuitively, if {\judgement{\propenv{}}{\e{}}{\t{}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
 then \equivapp{\e{}}{\makekw{if}} has the true and false propositions
@@
 $$
-\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\path{\keype{\makekw{op}}}{m}}
+\judgement{\propenv{}}{\getexp{m}{\makekw{op}}}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\path{\keype{\makekw{op}}}{m}}
 $$
@@
 $$
-\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}{\notprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}}{\emptyobject{}}
+\judgement{\propenv{}}{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}{\filterset{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}{\notprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}}{\emptyobject{}}
 $$
@@
 \judgement{}%{\propenv{}}
-{\hastype
 {\createmultiexp 
@@
       {\classconst{}}}
-  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+  {\MultiFntype {\s{}}{\sp{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
 $$
@@
 \judgement{\propenv{}}
-{\hastype
   {\extendmultiexp {prop} {\String}
                    {\abs {\x{}} {\Top{}} {\x{}}}}
-  {\MultiFntype {\s{}}{\sp{}}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+  {\MultiFntype {\s{}}{\sp{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
 $$
@@
 \judgement{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
-{\hastype
   {\x{}}
-  {\String}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+  {\String}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
 $$

### 2015-05-20 formal-model.tex
add trivial structural rewrite rules for everything but java interop
revert to statically typed Java interop

@@
   \begin{altgrammar}
-    \e{} &::=& \ldots \alt \reflectiveexpsyntax{} &\mbox{Expressions}\\
+    \e{} &::=& \ldots   
+            \alt \reflectiveexpsyntax{} &\mbox{Expressions}\\
+     &\alt& \mininonreflectiveexpsyntax{}
     \\
@@
   \begin{mathpar}
-    {\TNew}
+    {\TNewStatic}
 
-    {\TMethod}
+    {\TMethodStatic}
 
@@
 
-T-Field checks a resolved field expression by ensuring the target has
+T-FieldStatic checks a resolved field expression by ensuring the target has
 the correct static type, then returns a nilable type corresponding the Java type.
@@
 \begin{mathpar}
-    {\TField}
+    {\TFieldStatic}
 \end{mathpar}
 
-The rules T-Method and T-New work similarly (\figref{main:figure:javatyping}), varying
+The rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
 in the choice of nilability in the conversion function---methods

### 2015-06-15 formal-model.tex
add T-New, T-Method
remove resolve functions.
Use Mth/Ctor/Fld in T-{Method,New,Field}.
TODO TJ metafunction
updating proof
finish BetaMulti proof
update metatheory

@@
 $$
+{\judgementrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}
+  {\ep{}}}
+$$
+says expression \e{} is of type \t{} in the 
+proposition environment $\propenv{}$, with 
+`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+and object \object{}, and rewrites to \ep{}. We write 
+{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
+  or
 {\judgement   {\propenv}
@@
   {\object{}}}
-$$
-says expression \e{} is of type \t{} in the 
-proposition environment $\propenv{}$, with 
-`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}. We write 
-{\judgementtwo{\propenv}{\e{}} {\t{}} if we are only interested in the type.
+for expressions that rewrite to itself.
 
@@
   \footnotesize
+%   \judgbox{
+%{\judgementrewrite   {\propenv}
+%              {\e{}} {\t{}}
+%  {\filterset {\thenprop {\prop{}}}
+%              {\elseprop {\prop{}}}}
+%  {\object{}}{\ep{}}}}
+%           {Under proposition environment $\propenv{}$, 
+%             expression \e{} is of type \t{}
+%             with  \\
+%
+%`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+%and object \object{} and rewrites to \ep{}.}
   \begin{mathpar}
@@
   \end{mathpar}
- \classtablelookupsyntax{}
+ %\classtablelookupsyntax{}
  \begin{mathpar}
@@
     \convertjavatypenonnil{}
+  \end{altgrammar}
+  \\
+  \begin{altgrammar}
+    \converttctype{}
   \end{altgrammar}
@@
 \begin{mathpar}
-    {\TFieldStatic}
+    {\TField}
 \end{mathpar}

### 2015-06-28 intro.tex
Rewrote intro.

@@
 gradual type systems aimed at checking existing programs written in
-existing languages.  These include Microsoft's TypeScript for
-JavaScript, Facebook's Hack for PHP and Flow for JavaScript, and MyPy
+existing languages.  These include  TypeScript and Flow for
+JavaScript, Hack for PHP, and MyPy
 for Python among the optional systems, and Typed Racket, Reticulated
 Python, and GradualTalk among gradually-typed systems.\footnote{We
-  reserve the term ``gradual typing'' for systems such as Typed Racket which soundly
-  interoperate between typed and untyped code; systems like Typed Clojure or TypeScript which do not
+  use ``gradual typing'' for systems like Typed Racket with sound
+  interoperation between typed and untyped code; Typed Clojure or
+ TypeScript which don't
   enforce type invariants we describe as ``optionally typed''.}
@@
 We synthesize these lessons to present \emph{Typed Clojure}, an
-optional type system for Clojure. Typed Clojure builds on the core
-type checking approach of Typed Racket, an existing gradual type
-system for Racket. However, Typed Clojure extends this basic framework
-in multiple ways to accommodate the unique idioms and features of
-Clojure, producing an expressive synthesis of ideas and demonstrating
-a surprising coincidence between multiple dispatch in
-Clojure and Typed Racket's  occurrence typing framework.
-
-The essence of Typed Clojure, of course, is Clojure, a dynamically
+optional type system for Clojure. 
+%
+Clojure is a dynamically
 typed language in the Lisp family built to run on the Java Virtual
@@
 
-\begin{figure}
-\inputminted[firstline=5]{clojure}{code/demo/src/demo/parent2.clj}
+Since Clojure is a language in the
+Lisp family, we apply the lessons of Typed Racket, an existing gradual type
+system for Racket, to the core of Typed Clojure, consisting of an extended
+$\lambda$-calculus over a variety of base types shared between all Lisp systems.
+%
+Furthermore, Typed Racket's \emph{occurrence typing} has proved
+necessary for type checking realistic Clojure programs.
+
+\begin{figure}[t!]
+\inputminted[firstline=5,lastline=12]{clojure}{code/demo/src/demo/ex-intro.clj}
 \caption{A simple Typed Clojure program}
@@
 
-Figure~\ref{fig:ex1} presents a simple program demonstrating many
-aspects of our system, from simple type annotations to explicit
-handling of Java's \java{null} (written \clj{nil}) in interoperation, as well as an
-extended form of occurrence typing and method resolution of
-Java interoperability based on static type information.
+However, Clojure goes beyond Racket in many ways, requiring several
+new type system features which we detail in this paper.
+%
+Most significantly, Clojure supports, and Clojure developers use,
+\textbf{multimethods} to structure their code in extensible
+fashion. Furthermore, since Clojure is an untyped language, dispatch
+within multimethods is determined by application of dynamic predicates
+to argument values. 
+%
+Fortunately, the dynamic dispatch used by multimethods has surprising
+symmetry with the conditional dispatch handled by occurrence
+typing. Typed Clojure is therefore able to effectively handle complex
+and highly dynamic dispatch as present in existing Clojure programs.
 
-The \clj{parent} function has the type 
-$$
-\clj{['{:file (U nil File)} -> (U nil Str)]}
-$$
-which means that it takes a hash table whose \clj{:file} key maps to either
-\clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
-\clj{String}. The \clj{parent} function uses the \clj{:file} keyword
-as an accessor to get the file, checks that it isn't \clj{nil}, and
-then obtains the parent by making a Java method call.
+But multimethods are not the only Clojure feature crucial to type
+checking existing programs. As a language built on the Java Virtual
+Machine, Clojure provides flexible and transparent access to existing
+Java libraries, and \textbf{Clojure/Java interoperation} is found in almost
+every significant Clojure code base. Typed Clojure therefore builds in
+an understanding of the Java type system and handles interoperation
+appropriately. Notably, Typed Clojure avoids conflating Java's
+\texttt{null} with all other types, leading to automatic type-enforced
+absence of null-pointer exceptions.
 
-In the remainder of this paper, we describe how Typed Clojure's
-central innovations, including Java interoperability, multimethods,
-and heterogeneously-typed immutable maps, enable this example and many
-others. We begin with an example-driven presentation of the main type
-system features in \secref{sec:overview}. We then incrementally
-present a core calculus for Typed Clojure covering all of these
-features together in \secref{sec:formal} and prove type
-soundness (\secref{sec:metatheory}). We then discuss the full
-implementation of Typed Clojure, dubbed \coretyped{}, which extends
-the formal model in many ways, and the experience gained from its use
-in \secref{sec:experience}. Finally, we discuss related work and
-conclude.
+A simple example of these features in combination is given in
+figure~\ref{fig:ex1}. Here, the \texttt{tostr} multimethod dispatches
+on the \texttt{class} of the argument---when it is a \texttt{String},
+the first method implementation is called, for a Java \texttt{File}, the
+second. The method for \texttt{File} uses Java interoperation to call
+the \texttt{.getPath} method, and Typed Clojure uses the type
+information to help determine which JVM method call to construct.
 
-\subsection{Contributions}
+Finally, when not working with existing Java classes and objects, the
+most common Clojure data structure is the dictionary, a
+high-performance immutable table which Clojure programmers repurpose
+for all manner of data types. Simply treating them as uniformly-typed
+key-value mappings would be entirely insufficient for existing
+programs and programming styles. Instead, Typed Clojure provides a
+flexible \textbf{heterogenous maps}, in which mandatory, optional, and
+absent keys can be specified.
 
-These are our main contributions.
+While these features may seem disparate, they are unified in important
+ways. First, they all leverage the type system mechanisms that Typed
+Clojure inherits from Typed Racket---multimethods when using 
+dispatch via predicates, Java interoperation for handling
+\texttt{null} tests, and heterogenous maps using union types and
+reasoning about subcomponents of data. Second, and more significantly,
+they are the crucial features for handling Clojure code in
+practice. Typed Clojure's use in real Clojure deployments would not be
+possible without effective handling of these three Clojure features.
+
+%\subsection{Contributions}
+
+Our main contributions are as follows:
 
 \begin{itemize}
-  \item We motivate, describe and evaluate Typed Clojure, an optional type system for Clojure
-        that understands existing Clojure idioms.
-  \item We present the theory and implementation of a type system
-        based on \emph{occurrence typing} that statically prevents null-pointer 
-        exceptions in typed code in the presence of Java interoperability, 
-        applicable to existing JVM languages.
-  \item We prove type soundness for Typed Clojure.
+  \item We motivate and describe  Typed Clojure, an optional
+    type system for Clojure that understands existing Clojure idioms.
+  \item We present a sound formal model for the three crucial type
+    system features that Typed Clojure relies on: multi-methods, Java
+    interoperability, and heterogenous maps.
+  \item We evaluate the use of Typed Clojure features on existing
+    Typed Clojure code, including both open source and in-house systems.
 \end{itemize}
+
+%% \begin{figure}
+%% \inputminted[firstline=5]{clojure}{code/demo/src/demo/parent2.clj}
+%% \caption{A simple Typed Clojure program}
+%% \label{fig:ex1}
+%% \end{figure}
+
+%% Figure~\ref{fig:ex1} presents a simple program demonstrating many
+%% aspects of our system, from simple type annotations to explicit
+%% handling of Java's \java{null} (written \clj{nil}) in interoperation, as well as an
+%% extended form of occurrence typing and method resolution of
+%% Java interoperability based on static type information.
+
+%% The \clj{parent} function has the type 
+%% $$
+%% \clj{['{:file (U nil File)} -> (U nil Str)]}
+%% $$
+%% which means that it takes a hash table whose \clj{:file} key maps to either
+%% \clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
+%% \clj{String}. The \clj{parent} function uses the \clj{:file} keyword
+%% as an accessor to get the file, checks that it isn't \clj{nil}, and
+%% then obtains the parent by making a Java method call.
+
+ The remainder of this papers begins with an example-driven
+ presentation of the main type system features in
+ \secref{sec:overview}. We then incrementally present a core calculus
+ for Typed Clojure covering all of these features together in
+ \secref{sec:formal} and prove type soundness
+ (\secref{sec:metatheory}). We then discuss the full implementation of
+ Typed Clojure, dubbed \coretyped{}, which extends the formal model in
+ many ways, and empirical analysis of significant code bases written
+ in \coretyped{} in \secref{sec:experience}. Finally, we discuss
+ related work and conclude.

### 2015-06-28 small.tex
anon authors
Rewrote intro.

@@
+% rubber: set arguments --shell-escape
+% arara: pdflatex: {shell: true}
 %-----------------------------------------------------------------------------
@@
 
-\authorinfo{Ambrose Bonnaire-Sergeant}
-           {Indiana University}
-           {abonnair@indiana.edu}
-\authorinfo{Rowan Davies}
-           {University of Western Australia}
-           {rowan.davies@uwa.edu.au}
-\authorinfo{Sam Tobin-Hochstadt}
-           {Indiana University}
-           {samth@indiana.edu}
+\authorinfo{}{}{\vspace*{-2mm}}
+%\authorinfo{Ambrose Bonnaire-Sergeant}
+%           {Indiana University}
+%           {abonnair@indiana.edu}
+%\authorinfo{Rowan Davies}
+%           {University of Western Australia}
+%           {rowan.davies@uwa.edu.au}
+%\authorinfo{Sam Tobin-Hochstadt}
+%           {Indiana University}
+%           {samth@indiana.edu}
 
@@
 Typed Clojure is an optional type system for Clojure, a dynamic
-language in the Lisp family that targets the JVM. Typed Clojure's type
-system build on the design of Typed Racket, repurposing in particular
+language in the Lisp family that targets the JVM. Typed Clojure
+enables Clojure programmers to gain greater confidence in the
+correctness of their code via static type checking while remaining in
+the Clojure world, and has acquired significant adoption in the
+Clojure community. Typed Clojure's type
+system repurposes Typed Racket's
 \emph{occurrence typing}, an approach to statically reasoning about
-predicate tests. However, in adapting the type system to Clojure,
-changes and extensions are required to accommodate  additional
-language features and  idioms used by Clojure programmers.
+predicate tests, and also includes several new type system features to
+handle existing Clojure idioms.
 
 In this paper, we describe Typed Clojure and present these type system
-extensions, focusing on three features widely used in Clojure. First, Java
+extensions, focusing on three features widely used in Clojure. 
+%
+ First, multimethods provide extensible
+operations, and their Clojure semantics turns out to have a surprising
+synergy with the underlying occurrence typing framework.
+%
+Second, Java
 interoperability is central to Clojure's mission but introduces
@@
 Java interoperability while ensuring the absence of null-pointer
-exceptions in typed programs. Second, Clojure programmers
+exceptions in typed programs. 
+%
+Third, Clojure programmers
 idiomatically use immutable dictionaries for data structures; Typed
-Clojure handles this in the type system with multiple forms of
-heterogeneous dictionary types. Third, multimethods provide extensible
-operations, and their Clojure semantics turns out to have a surprising
-synergy with the underlying occurrence typing framework.
+Clojure handles this with multiple forms of
+heterogeneous dictionary types.
+%
 
@@
 corporations and developers working with Clojure, and we report on 
-experience with the system and its lessons for the future.
+a quantiative analysis of the use of the features described in this
+paper in two substantial code bases.
 

### 2015-06-29 small.tex
typo

@@
 corporations and developers working with Clojure, and we report on 
-a quantiative analysis of the use of the features described in this
+a quantitative analysis of the use of the features described in this
 paper in two substantial code bases.

### 2015-07-03 formal-model.tex
Add section 2 intro. Move exceptional flow to section 3 unedited.

@@
 \label{sec:doformal}
+
+Along with conditional control flow,
+Clojure programmers rely on \emph{exceptions}
+to assert type-related invariants.
+
+\begin{exmp}
+\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
+\label{example:doexception}
+\end{exmp}
+
+The fully expanded increment function in~\egref{example:doexception}
+guards its final call with a number check, preventing
+a possible null-pointer exception.
+Without this check, the type system would reject the program.
+
+To check this example,
+occurrence typing 
+automatically
+assumes
+\clj{x} is a number when checking the second \clj{do} subexpression
+based on the first subexpression.
+\footnote{See \url{https://github.com/typedclojure/examples}
+  for full examples.}
+We model this formally %(section~\ref{sec:doformal}) 
+and prove
+null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).
 

### 2015-07-03 overview.tex
More editing.
Add section 2 intro. Move exceptional flow to section 3 unedited.

@@
 
-\subsection{Clojure Syntax}
+%\paragraph{Clojure Syntax}
+%
+%We describe new syntax as they appear in each example, but
+%begin with include the essential basics of Clojure syntax.
+%
+%\clj{nil} is exactly Java's \java{null}.
+%Parentheses indicate \emph{applications}, brackets
+%delimit
+%\emph{vectors}, braces
+%delimit
+%\emph{hash-maps}
+%and double quotes delimit \emph{Java strings}.
+%\emph{Symbols} begin with an alphabetic character,
+%and a colon prefixed symbol like \clj{:a} is a \emph{keyword}.
+%
+%\emph{Commas} are always \emph{whitespace}.
 
-We describe new syntax as they appear in each example, but
-we also include the essential basics of Clojure syntax.
+\subsection{Typed Clojure Introduction}
 
-\clj{nil} is exactly Java's \java{null}.
-Parentheses indicate \emph{applications}, brackets
-delimit
-\emph{vectors}, braces
-delimit
-\emph{hash-maps}
-and double quotes delimit \emph{Java strings}.
-\emph{Symbols} begin with an alphabetic character,
-and a colon prefixed symbol like \clj{:a} is a \emph{keyword}.
+Here is a simple program in Typed Clojure.
 
-\emph{Commas} are always \emph{whitespace}.
+We define a global variable \clj{greet} as a one-argument function
+taking and returning a string.
 
-\subsection{Typed Racket and occurrence typing}
+\begin{minted}{clj}
+(ann  greet [Str -> Str])
+(defn greet [n]
+  (str "Hello, " n "!"))
+(greet "Grace") ; :- Str
+;=> "Hello, Grace!"
+(greet nil)    ; Type Error:
+               ;   Expected Str, given nil
+\end{minted}
 
-\citet{TF10}
-presented Typed Racket with occurrence typing,
-a technique for deriving type information from conditional control flow.
-They introduced the concept of occurrence typing 
-with the following example.
+The function may take strings, but providing \clj{nil} (Clojure's name for Java's \emph{null})
+is a static type error --- \clj{nil} is not a string in Typed Clojure.
 
-\inputminted[firstline=3]{racket}{code/tr/example1.rkt}
+\paragraph{Unions} We can make the annotation more permissive with \emph{ad-hoc unions}
+to allow \clj{nil}.
 
-This function takes a value that is either \emph{\#f} % mintinline really hates #
-or a number, represented by an \emph{untagged} union type.
-The `then' branch has an implicit invariant
-that \rkt{x} is a number, which is automatically inferred with occurrence typing
-and type checked without further annotations.
+\begin{minted}{clj}
+(ann  greet-nil [(U nil Str) -> Str])
+(defn greet-nil [n]
+  (str "Hello" (when n (str ", " n)) "!"))
+(greet-nil "Donald") ; :- Str
+;=> "Hello, Donald!"
+(greet-nil nil)    ; :- Str
+;=> "Hello!"
+\end{minted}
 
-We chose to build on the ideas and implementation
-of Typed Racket to implement a type system targeting Clojure for several reasons.
-Initially, the similarities between Racket and Clojure drew us to
-investigate the effectiveness of repurposing occurrence typing
-for a Clojure type system---both languages share a Lisp heritage,
-similar standard functions 
-(for instance \clj{map}
-in both languages is variable-arity)
-and idioms.
-While Typed Racket is gradually typed and has sophisticated
-dynamic semantics for cross-language interaction, we 
-chose to first implement
-the static semantics
-with the hope to extend Typed Clojure to be gradually typed at a future date.
-Finally,
-Typed Racket's combination of bidirectional checking
-and occurrence typing presents a successful model for 
-type checking dynamically typed programs without compromising
-soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
-which cannot prove strong properties about programs
-and soft typing~\cite{CF91}
-which has proved too complicated in practice.
+All Clojure values are true, except \clj{nil} and \clj{false}, so the
+comma is only added when the argument is non-nil.
 
-Here is the above program in Typed Clojure.
-\begin{exmp}
-\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
-\label{example:conditionalflow}
-\end{exmp}
+\paragraph{Flow analysis} Typed Clojure uses occurrence typing to
+model type-based control flow.
 
-The \clj{fn} macro (provided by core.typed) supports optional annotations by 
-adding
-\clj{:-} and a type after a parameter
-position
-or binding vector 
-to annotate parameter types
-and return types respectively.
-\clj{number?} is
-a Java \java{instanceof} test of \clj{java.lang.Number}.
-As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
-any number of types.
+\begin{minted}{clj}
+(ann  greetings [Str (U nil Int) -> Str])
+(defn greetings [n i]
+  (str "Hello, "
+       (when i
+         (apply str (repeat i (str "hello, "))))
+       n "!"))
+(greetings "Donald" 2)  ; :- Str
+;=> "Hello, hello, hello, Donald!"
+(greetings "Grace" nil) ; :- Str
+;=> "Hello, Grace!"
+\end{minted}
 
-Typed Clojure can already check all of the examples in~\citet{TF10}---the 
-rest of this section describes the extensions necessary
-to check Clojure code.
+Providing \clj{repeat} with a nilable first argument is disallowed,
+but the \clj{when} expression ensures it can only be passed an integer.
 
-\subsection{Exceptional control flow}
-
-Along with conditional control flow,
-Clojure programmers rely on \emph{exceptions}
-to assert type-related invariants.
-
-\begin{exmp}
-\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
-\label{example:doexception}
-\end{exmp}
-
-The fully expanded increment function in~\egref{example:doexception}
-guards its final call with a number check, preventing
-a possible null-pointer exception.
-Without this check, the type system would reject the program.
-
-To check this example,
-occurrence typing 
-automatically
-assumes
-\clj{x} is a number when checking the second \clj{do} subexpression
-based on the first subexpression.
-\footnote{See \url{https://github.com/typedclojure/examples}
-  for full examples.}
-We model this formally (section~\ref{sec:doformal}) and prove
-null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).
+%\subsection{Type System Basics}
+%
+%\citet{TF10}
+%presented Typed Racket with occurrence typing,
+%a technique for deriving type information from conditional control flow.
+%They introduced the concept of occurrence typing 
+%with the following example.
+%
+%\inputminted[firstline=3]{racket}{code/tr/example1.rkt}
+%
+%This function takes a value that is either \emph{\#f} % mintinline really hates #
+%or a number, represented by an \emph{untagged} union type.
+%The `then' branch has an implicit invariant
+%that \rkt{x} is a number, which is automatically inferred with occurrence typing
+%and type checked without further annotations.
+%
+%We chose to build on the ideas and implementation
+%of Typed Racket to implement a type system targeting Clojure for several reasons.
+%Initially, the similarities between Racket and Clojure drew us to
+%investigate the effectiveness of repurposing occurrence typing
+%for a Clojure type system---both languages share a Lisp heritage,
+%similar standard functions 
+%(for instance \clj{map}
+%in both languages is variable-arity)
+%and idioms.
+%While Typed Racket is gradually typed and has sophisticated
+%dynamic semantics for cross-language interaction, we 
+%chose to first implement
+%the static semantics
+%with the hope to extend Typed Clojure to be gradually typed at a future date.
+%Finally,
+%Typed Racket's combination of bidirectional checking
+%and occurrence typing presents a successful model for 
+%type checking dynamically typed programs without compromising
+%soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
+%which cannot prove strong properties about programs
+%and soft typing~\cite{CF91}
+%which has proved too complicated in practice.
+%
+%Here is the above program in Typed Clojure.
+%\begin{exmp}
+%\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
+%\label{example:conditionalflow}
+%\end{exmp}
+%
+%The \clj{fn} macro (provided by core.typed) supports optional annotations by 
+%adding
+%\clj{:-} and a type after a parameter
+%position
+%or binding vector 
+%to annotate parameter types
+%and return types respectively.
+%\clj{number?} is
+%a Java \java{instanceof} test of \clj{java.lang.Number}.
+%As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
+%any number of types.
+%
+%Typed Clojure can already check all of the examples in~\citet{TF10}---the 
+%rest of this section describes the extensions necessary
+%to check Clojure code.
 
@@
 
-The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
-The function \clj{an-exp} is verified to return an \clj{Expr}.
-
-The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
-Here \clj{defalias} defines \clj{Expr}
+%The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
+%The function \clj{an-exp} is verified to return an \clj{Expr}.
+%
+Here \clj{defalias} defines \clj{Expr}, a type abbreviation
 that describes the structure of a recursively-defined AST as a union of HMaps.
-A quoted keyword in a type, such as \clj{':if}, is a singleton type that contains just the keyword.
+A quoted keyword in a type --- \clj{':if} --- is a singleton type that contains just the keyword.
 A type that is a quoted map like \clj{'{:op ':if}} is a

### 2015-07-05 intro.tex
More editing.
New page for bib. Changes to sec 1 & 2

@@
 \begin{figure}[t!]
-\inputminted[firstline=5,lastline=12]{clojure}{code/demo/src/demo/ex-intro.clj}
+\inputminted[firstline=5]{clojure}{code/demo/src/demo/ex-intro.clj}
 \caption{A simple Typed Clojure program}
@@
 A simple example of these features in combination is given in
-figure~\ref{fig:ex1}. Here, the \texttt{tostr} multimethod dispatches
+figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
 on the \texttt{class} of the argument---when it is a \texttt{String},
 the first method implementation is called, for a Java \texttt{File}, the
-second. The method for \texttt{File} uses Java interoperation to call
-the \texttt{.getPath} method, and Typed Clojure uses the type
-information to help determine which JVM method call to construct.
+second. The method for \texttt{String} uses Java interoperation
+to call a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
+method for \texttt{File} invokes
+the \texttt{.getName} method with a non-nil target, returning a nilable
+type. Typed Clojure uses type information to resolve JVM method or constructor overloads,
+avoiding expensive runtime reflective calls.
 
@@
 programs and programming styles. Instead, Typed Clojure provides a
-flexible \textbf{heterogenous maps}, in which mandatory, optional, and
+flexible \textbf{heterogenous map} type, in which mandatory, optional, and
 absent keys can be specified.
@@
 
-\begin{itemize}
+\begin{enumerate}
   \item We motivate and describe  Typed Clojure, an optional
@@
     Typed Clojure code, including both open source and in-house systems.
-\end{itemize}
+\end{enumerate}
 
@@
 
- The remainder of this papers begins with an example-driven
+\noindent
+ The remainder of this paper begins with an example-driven
  presentation of the main type system features in

### 2015-07-05 overview.tex
New page for bib. Changes to sec 1 & 2

@@
 
-\subsection{Typed Clojure Introduction}
+\subsection{Typed Clojure}
 
@@
 ;=> "Hello, Grace!"
+\end{minted}
+
+The function may take strings, but providing \clj{nil} (Clojure's name for Java's \java{null})
+is a static type error --- \clj{nil} is not a string in Typed Clojure.
+
+\begin{minted}{clj}
 (greet nil)    ; Type Error:
                ;   Expected Str, given nil
+               ; in: nil
 \end{minted}
-
-The function may take strings, but providing \clj{nil} (Clojure's name for Java's \emph{null})
-is a static type error --- \clj{nil} is not a string in Typed Clojure.
 
@@
 
-All Clojure values are true, except \clj{nil} and \clj{false}, so the
+All Clojure values are true except \clj{nil} and \clj{false}, so the
 comma is only added when the argument is non-nil.
@@
 model type-based control flow.
+In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
+is never passed a \clj{nil} argument.
 
@@
 
-Providing \clj{repeat} with a nilable first argument is disallowed,
-but the \clj{when} expression ensures it can only be passed an integer.
+Removing the \clj{when} expression is a static type error
+--- \clj{repeat} cannot be passed \clj{nil}.
+
+\begin{minted}{clj}
+(ann  greetings-bad [Str (U nil Int) -> Str])
+(defn greetings-bad [n i]
+  (str "Hello, "
+       (apply str (repeat i (str "hello, ")))
+       ; Type Error:
+       ;   Expected Num, given (U nil Num).
+       ; in: i
+       n "!"))
+\end{minted}
 
@@
 %to check Clojure code.
+
+\subsection{Java interoperability}
+\label{sec:overviewjavainterop}
+
+Clojure supports interoperability with Java, including the ability to
+call constructors, invoke methods, and access fields.
+
+\begin{minted}{clj}
+  (.getParent (new File "a/b"))  ; :- (U nil Str)
+  ;=> "a"
+\end{minted}
+
+If a specific constructor, method, or field cannot be found based on the
+static types of its arguments, a type error is thrown.
+
+\begin{minted}{clj}
+  (fn [f] 
+    (.getParent f)
+    ; Type error:
+    ;   Unresolved host interop: getParent
+    )
+\end{minted}
+
+Function arguments default to \clj{Any}, the most permissive type. Ascribing
+a parameter type helps Typed Clojure find a specific method.
+
+%Calls to Java methods and fields have prefix notation
+%like \clj{(.method target args*)} and \clj{(.field target)} respectively,
+%with method and field names prefixed with a dot and methods taking some number of arguments.
+
+\begin{exmp}
+\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
+\end{exmp}
+
+%\begin{exmp}
+%\inputminted[firstline=18,lastline=19]{clojure}{code/demo/src/demo/parent3.clj}
+%\end{exmp}
+
+The conditional guards from dereferencing \clj{nil}, and as before
+removing it is a static type error.
+
+\begin{minted}{clj}
+(defn parent-bad-in [f :- (U nil File)]
+  (.getParent f)
+  ; Type error:
+  ;   Cannot call instance method 
+  ;   java.io.File/getParent on type 
+  ;   (U nil File).
+  )
+\end{minted}
+
+Any Java method returning a reference can also return \java{null} ---
+Typed Clojure rejects programs that assume otherwise.
+
+\begin{minted}{clj}
+(defn parent-bad-out [f :- File] :- Str
+  (.getParent f)
+  ; Type error:
+  ;   Expected Str, given (U nil Str).
+  ; in: (.getParent f)
+  )
+\end{minted}
+
+%Typed Clojure and Java treat \java{null} differently.
+%In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
+%called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
+%This means the Java static type \java{String} is equivalent to
+%\clj{(U nil String)} in Typed Clojure.
+%
+%Reference types in Java are nullable, so to guarantee a method call does not
+%leak \java{null} into a Typed Clojure program we
+%must assume methods can return \clj{nil}.
+
+In contrast, JVM invariants guarantee constructors return a non-null reference.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
+
+\begin{exmp}
+\inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
+\end{exmp}
+
+By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
+but can be configured globally for particular non-target positions if needed.
+
+\subsection{Multimethod}
+
+A \emph{multimethod} is a kind of extensible function. It consists of
+a \emph{dispatching function}, given with \clj{defmulti}, and a collection of \emph{methods} associated with
+\emph{dispatch values}, each installed via \clj{defmethod}.
+
+Recall the multimethod in figure~\ref{fig:ex1}.
+
+\begin{minted}{clj}
+(ann pname [(U File String) -> (U nil String)])
+(defmulti pname class)
+(defmethod pname String [s] (pname (new File s)))
+(defmethod pname File [f] (.getName f))
+\end{minted}
+
+Its dispatching function is
+\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
+respectively.
+
+When called, the dispatching function is applied to the provided arguments to
+yield a \emph{dispatching value}.
+The \clj{isa?} function is then used to compare the current dispatching value to each method's dispatch value.
+An unambiguous match leads to the corresponding method being applied to the arguments,
+giving the final result.
+
+The following Typed Clojure program is semantically identical to figure~\ref{fig:ex1}.
+
+\begin{minted}{clj}
+(ann pname [(U Str File) -> (U nil Str)])
+(defn pname [x]
+  (let [dfn  class ; dispatching function
+        dval (dfn x)] ; dispatching value
+    (cond
+      (and (isa? dval String)       ; if
+           (not (isa? dval File))) 
+      (pname (new File x))          ; then
+
+      (and (not (isa? dval String)) ; else if
+           (isa? dval File))       
+      (.getName x)                  ; then
+      :else (throw (Exception. "No match")))))
+\end{minted}
+
+The \clj{isa?} predicate 
+is a subclassing check for classes, otherwise
+an equality check.
+
+\begin{minted}{clojure}
+  (isa? :a :a) ;=> true
+  (isa? :a :b) ;=> false
+  (isa? String Object) ;=> true
+  (isa? Object String) ;=> false
+\end{minted}
+
+\subsection{Multimethods}
+
+\label{sec:multioverview}
+
+%A multimethod in Clojure is a function with a \emph{dispatch
+%function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
+%\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
+%The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
+%and \emph{dispatch function} \clj{class}, a function that
+%returns the class of its argument or \clj{nil} if passed \clj{nil}.
+%
+%We can use {\clj{defmethod}} to install a method to \clj{path}.
+%\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
+%Now the dispatch table maps
+%the \emph{dispatch value} \clj{String} to the function
+%\clj{(fn [x] x)}. 
+%We add another method
+%which maps
+%\clj{File} to the function
+%\clj{(fn [x] (.getPath x))}
+%in the dispatch table.
+%\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
+%
+%After installing both methods, the call 
+%$$
+%\clj{(path (new File "dir/a"))}
+%$$
+%dispatches to the second method we installed because
+%$$
+%\clj{(isa? (class "dir/a") String)}
+%$$
+%is true, and finally returns 
+%$$
+%\clj{((fn [x] (.getPath x)) "dir/a")}.
+%$$
+
+We include the above sequence of definitions as \egref{example:rep}.
+
+\begin{Code}
+\begin{exmp}
+\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
+\label{example:rep}
+\end{exmp}
+\end{Code}
+
+Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
+type checks because \clj{:a} agrees with the parameter type \clj{Any},
+but throws an error at runtime.
+
+\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
+In \egref{example:incmap} we
+dispatch on the \clj{:op} key 
+of our HMap AST \clj{Expr}.
+Since keywords are functions that look themselves up in their argument, we simply
+use \clj{:op} as the dispatch function.
+
+\begin{exmp}
+\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
+\label{example:incmap}
+\end{exmp}
+
+The function \clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
+The reasoning is similar, except we only consider one branch (the current method) by
+locally considering the current \emph{dispatch value} and reasoning about how it relates
+to the \emph{dispatch function}.
+For example, 
+in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
+narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
+and \clj{:const} methods.
+
+\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
+same length recursively call \clj{isa?} on the elements pairwise.
+\begin{minted}{clojure}
+  (isa? [Keyword Keyword] [Object Object]) ;=> true
+\end{minted}
+
+\egref{example:multidispatch}
+simulates multiple dispatch by dispatching on
+a vector containing the class of both arguments. \clj{open}
+takes two arguments which can be strings or files and returns
+a new file that concatenates their paths.
+
+We call three different \clj{File} constructors, each known at compile-time
+via type hints.
+Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
+except we update multiple bindings simultaneously.
 
@@
 
-\subsection{Java interoperability}
-\label{sec:overviewjavainterop}
-
-Clojure supports interoperability with Java, including the ability to
-call constructors, methods and access fields.
-
-\begin{minted}{clj}
-  (fn [f] (.getParent f))
-\end{minted}
-
-Calls to Java methods and fields have prefix notation
-like \clj{(.method target args*)} and \clj{(.field target)} respectively,
-with method and field names prefixed with a dot and methods taking some number of arguments.
-
-\begin{exmp}
-\inputminted[firstline=18,lastline=19]{clojure}{code/demo/src/demo/parent3.clj}
-\end{exmp}
-
-Typed Clojure and Java treat \java{null} differently.
-In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
-called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
-This means the Java static type \java{String} is equivalent to
-\clj{(U nil String)} in Typed Clojure.
-
-Reference types in Java are nullable, so to guarantee a method call does not
-leak \java{null} into a Typed Clojure program we
-must assume methods can return \clj{nil}.
-
-\begin{exmp}
-\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
-\end{exmp}
-
-In contrast, JVM invariants guarantee that  constructors cannot return \java{null},
-so we are safe to assume constructors are non-nullable.
-
-\begin{exmp}
-\inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
-\end{exmp}
-
-By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
-but can be configured globally for particular positions if needed.
-
-\subsection{Multimethods}
-
-\label{sec:multioverview}
-
-A multimethod in Clojure is a function with a \emph{dispatch
-function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
-\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
-The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
-and \emph{dispatch function} \clj{class}, a function that
-returns the class of its argument or \clj{nil} if passed \clj{nil}.
-
-We can use {\clj{defmethod}} to install a method to \clj{path}.
-\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
-Now the dispatch table maps
-the \emph{dispatch value} \clj{String} to the function
-\clj{(fn [x] x)}. 
-We add another method
-which maps
-\clj{File} to the function
-\clj{(fn [x] (.getPath x))}
-in the dispatch table.
-\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
-
-After installing both methods, the call 
-$$
-\clj{(path (new File "dir/a"))}
-$$
-dispatches to the second method we installed because
-$$
-\clj{(isa? (class "dir/a") String)}
-$$
-is true, and finally returns 
-$$
-\clj{((fn [x] (.getPath x)) "dir/a")}.
-$$
-
-The \clj{isa?} function first tries an equality check
-on its arguments, then if that fails
-and both arguments are classes a subclassing
-check is returned.
-\begin{minted}{clojure}
-  (isa? :a :a) ;=> true
-  (isa? Keyword Object) ;=> true
-\end{minted}
-
-We include the above sequence of definitions as \egref{example:rep}.
-
-\begin{Code}
-\begin{exmp}
-\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
-\label{example:rep}
-\end{exmp}
-\end{Code}
-
-Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
-type checks because \clj{:a} agrees with the parameter type \clj{Any},
-but throws an error at runtime.
-
-\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
-In \egref{example:incmap} we
-dispatch on the \clj{:op} key 
-of our HMap AST \clj{Expr}.
-Since keywords are functions that look themselves up in their argument, we simply
-use \clj{:op} as the dispatch function.
-
-\begin{exmp}
-\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
-\label{example:incmap}
-\end{exmp}
-
-\clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
-The reasoning is similar, except we only consider one branch (the current method) by
-locally considering the current \emph{dispatch value} and reasoning about how it relates
-to the \emph{dispatch function}.
-For example, 
-in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
-narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
-and \clj{:const} methods.
-
-\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
-same length recursively call \clj{isa?} on the elements pairwise.
-\begin{minted}{clojure}
-  (isa? [Keyword Keyword] [Object Object]) ;=> true
-\end{minted}
-
-\egref{example:multidispatch}
-simulates multiple dispatch by dispatching on
-a vector containing the class of both arguments. \clj{open}
-takes two arguments which can be strings or files and returns
-a new file that concatenates their paths.
-
-We call three different \clj{File} constructors, each known at compile-time
-via type hints.
-Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
-except we update multiple bindings simultaneously.
-
 \subsection{Final example}

### 2015-07-05 conclusion.tex
Edit conclusion.

@@
 
-We have presented Typed Clojure, an optionally-typed version of
-Clojure whose type system works with a wide variety of distinctive
+Optional type systems must be designed with close attention to the
+languague that they are intendend to work for.
+We have therefore designed Typed Clojure, an optionally-typed version of
+Clojure, with a type system that works with a wide variety of distinctive
 Clojure idioms and features. Although based on the foundation of Typed
@@
 handle seemingly unrelated features such as multi-methods. In
-addition, Typed Clojure supports crucial features such as heterogeneous
-maps and Java interoperability while integrating these features into
-the core type system.
+addition, Typed Clojure supports crucial features such as
+heterogeneous maps and Java interoperability while integrating these
+features into the core type system. Not only are each of these
+features important in isolation to Clojure and Typed Clojure
+programmers, but they must fit together smoothly to ensure that
+existing Clojure programs are easy to convert to Typed Clojure.
 
@@
 programmers and receives contributions from many developers.
+
+Our empirical analysis of existing Typed Clojure programs bears out
+our design choices. Multimethods, Java interoperation, and
+heterogenous maps are indeed common in both Clojure and Typed Clojure,
+meaning that our type system must accomodate them. Furthermore, they
+are commonly used together, and the features of each are mutually
+reinforcing. Additionally, the choice to make Java's \clj{null}
+explicit in the type system is validated by the many Typed Clojure
+programs that use the ability to specify non-nullable types.
 
@@
 
-Finally, our case study and broader experience indicate that Clojure
+Finally, our case study, evaluation, and broader experience indicate that Clojure
 programmers still find themselves unable to use Typed Clojure on some

### 2015-07-05 formal-model.tex
Rip out most of section 2

@@
   or
-{\judgement   {\propenv}
+{\judgementselfrewrite   {\propenv}
               {\e{}} {\t{}}
@@
   {\object{}}}
-for expressions that rewrite to itself.
+for expressions that rewrite to themselves.
 

### 2015-07-05 overview.tex
Rip out most of section 2

@@
 \begin{minted}{clj}
-(greet nil)    ; Type Error:
+(greet 
+  nil ; Type Error:
       ;   Expected Str, given nil
-               ; in: nil
+  )
 \end{minted}
@@
   (str "Hello, "
-       (when i
-         (apply str (repeat i (str "hello, "))))
+       (when i  ; when i is a non-nil integer
+         (apply str (repeat i "hello, ")))
        n "!"))
@@
   (str "Hello, "
-       (apply str (repeat i (str "hello, ")))
-       ; Type Error:
-       ;   Expected Num, given (U nil Num).
-       ; in: i
+       (apply str 
+         (repeat 
+           i ; Type Error:
+             ;   Expected Int, given (U nil Int).
+           "hello, "))
        n "!"))
@@
   (fn [f] 
-    (.getParent f)
-    ; Type error:
-    ;   Unresolved host interop: getParent
+    (.getParent f) ; Type Error:
+                   ;   Unresolved interop: getParent
     )
@@
 (defn parent-bad-in [f :- (U nil File)]
-  (.getParent f)
-  ; Type error:
+  (.getParent f) ; Type Error:
                  ;   Cannot call instance method 
@@
 (defn parent-bad-out [f :- File] :- Str
-  (.getParent f)
-  ; Type error:
+  (.getParent f) ; Type Error:
                  ;   Expected Str, given (U nil Str).
-  ; in: (.getParent f)
   )
@@
 
-\subsection{Multimethod}
+\subsection{Multimethods}
 
-A \emph{multimethod} is a kind of extensible function. It consists of
-a \emph{dispatching function}, given with \clj{defmulti}, and a collection of \emph{methods} associated with
-\emph{dispatch values}, each installed via \clj{defmethod}.
+\label{sec:multioverview}
 
-Recall the multimethod in figure~\ref{fig:ex1}.
+A \emph{multimethod} is a kind of extensible function. It combines
+a \emph{dispatch function} with one or more \emph{methods}.
+When invoked, the arguments are first supplied to the dispatch function, yielding
+a \emph{dispatch value}. A method is then chosen
+based on the dispatch value---this method is applied to the arguments
+to finally return a value for the entire expression.
+
+Recall the multimethod in figure~\ref{fig:ex1}, duplicated below.
 
@@
 
-Its dispatching function is
-\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
-respectively.
+%Its dispatching function is
+%\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
+%respectively.
 
-When called, the dispatching function is applied to the provided arguments to
-yield a \emph{dispatching value}.
-The \clj{isa?} function is then used to compare the current dispatching value to each method's dispatch value.
-An unambiguous match leads to the corresponding method being applied to the arguments,
-giving the final result.
+The dispatch function \clj{class}---associated at multimethod creation with \clj{defmulti}---dictates 
+whether the \clj{String} or \clj{File} method is chosen---both installed via \clj{defmethod}.
 
-The following Typed Clojure program is semantically identical to figure~\ref{fig:ex1}.
+The multimethod dispatch rules involve
+\clj{isa?}, a hybrid predicate which includes a subclassing check for classes and
+an equality check for most other values.
 
-\begin{minted}{clj}
-(ann pname [(U Str File) -> (U nil Str)])
-(defn pname [x]
-  (let [dfn  class ; dispatching function
-        dval (dfn x)] ; dispatching value
-    (cond
-      (and (isa? dval String)       ; if
-           (not (isa? dval File))) 
-      (pname (new File x))          ; then
-
-      (and (not (isa? dval String)) ; else if
-           (isa? dval File))       
-      (.getName x)                  ; then
-      :else (throw (Exception. "No match")))))
+\begin{minted}{clojure}
+  (isa? String Object) ;=> true
+  (isa? Object String) ;=> false
+  ; keywords are interned symbols with a : prefix
+  (isa? :a :a) ;=> true
+  (isa? :a :b) ;=> false
 \end{minted}
 
-The \clj{isa?} predicate 
-is a subclassing check for classes, otherwise
-an equality check.
+The current dispatch value and---in turn---each method's associated dispatch value
+is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.
+
+To demonstrate,
+  \clj{(pname "STAINS/JELLY")}
+chooses the \clj{String} method because
+\clj{(isa? (class "STAINS/JELLY") String)}
+is true and
+\clj{(isa? (class "STAINS/JELLY") File)}
+is false. Then the \clj{String} method body
+\clj{(pname (new File "STAINS/JELLY"))}
+chooses the \clj{File} method for opposite reasons,
+resulting in 
+\begin{minted}{clojure}
+(.getName (new File "STAINS/JELLY"))  ; :- (U nil Str)
+;=> "JELLY"
+\end{minted}
+
+%The following Typed Clojure program is semantically identical to figure~\ref{fig:ex1}.
+%
+%\begin{minted}{clj}
+%(ann pname [(U Str File) -> (U nil Str)])
+%(defn pname [x]
+%  ; dispatch value calculated by applying dispatch
+%  ; function `class` to argument `x`.
+%  (cond
+%    ; if (class x) subclasses String, but not File
+%    (and (isa? (class x) String)
+%         (not (isa? (class x) File)))
+%    ; then choose the String method
+%    (pname (new File x))
+%
+%    ; else if (class x) subclasses File, but not String
+%    (and (isa? (class x) File)         
+%         (not (isa? (class x) String)))
+%    ; then choose the File method
+%    (.getName x)
+%    :else (throw (Exception. "No match"))))
+%\end{minted}
+%
+%An unambiguous match leads to the corresponding method being applied to the arguments,
+%giving the final result.
+
+Dispatching on a non-class value is common, as well as specifying
+a \clj{:default} method for when no methods are matched.
 
 \begin{minted}{clojure}
-  (isa? :a :a) ;=> true
-  (isa? :a :b) ;=> false
-  (isa? String Object) ;=> true
-  (isa? Object String) ;=> false
+(ann hi [Kw -> Str])
+(defmulti hi identity)
+(defmethod hi :en [_] "hello")
+(defmethod hi :fr [_] "bonjour")
+(defmethod hi :default [_] "um...")
+
+(str (hi :en) " " (hi :fr)) ;=> "hello bonjour"
+(hi :bocce)   ;=> "um..."
 \end{minted}
 
-\subsection{Multimethods}
+The \clj{hi} multimethod dispatches on its arguments via the identity function.
 
-\label{sec:multioverview}
-
+%\subsection{Multimethods}
+%
 %A multimethod in Clojure is a function with a \emph{dispatch
@@
 
-We include the above sequence of definitions as \egref{example:rep}.
+%We include the above sequence of definitions as \egref{example:rep}.
+%
+%\begin{Code}
+%\begin{exmp}
+%\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
+%\label{example:rep}
+%\end{exmp}
+%\end{Code}
+%
+%Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
+%type checks because \clj{:a} agrees with the parameter type \clj{Any},
+%but throws an error at runtime.
 
-\begin{Code}
-\begin{exmp}
-\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
-\label{example:rep}
-\end{exmp}
-\end{Code}
+%\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
+%same length recursively call \clj{isa?} on the elements pairwise.
+%\begin{minted}{clojure}
+%  (isa? [Keyword Keyword] [Object Object]) ;=> true
+%\end{minted}
+%
+%\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+%
+%\egref{example:multidispatch}
+%simulates multiple dispatch by dispatching on
+%a vector containing the class of both arguments. \clj{open}
+%takes two arguments which can be strings or files and returns
+%a new file that concatenates their paths.
+%
+%We call three different \clj{File} constructors, each known at compile-time
+%via type hints.
+%Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
+%except we update multiple bindings simultaneously.
 
-Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
-type checks because \clj{:a} agrees with the parameter type \clj{Any},
-but throws an error at runtime.
+\subsection{Heterogeneous hash-maps}
+
+Immutable hash-maps with keyword keys play a major role in Clojure programming.
+
+\begin{minted}{clojure}
+(def breakfast {:en "waffles" :fr "croissants"})
+
+breakfast ; :- (HMap :mandatory {:en Str :fr Str}
+          ;          :complete? true)
+
+(get breakfast :en)    ; :- Str
+;=> "waffles"
+; keywords are functions that look themselves up
+(:fr breakfast)        ; :- Str
+;=> "croissants"
+(:bocce breakfast) ; :- nil
+;=> nil
+\end{minted}
+
+HMap types model the most common usages of keyword maps.
+The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
+and \clj{:fr} have string values (in \clj{:mandatory}) and that 
+every other key is absent (in \clj{:complete?}).
+
+HMap's default to being partially specified. The HMap shorthand annotating
+\clj{lunch}---below---forgets 
+information about absent keys, only providing \clj{:mandatory}
+information.
+
+\begin{minted}{clojure}
+(ann lunch '{:en Str :fr Str})
+(def lunch {:en "muffin" :fr "baguette"})
+
+(:en lunch)    ; :- Str
+;=> "muffin"
+(:fr lunch)    ; :- Str
+;=> "baguette"
+; Unknown lookups are now less accurate
+(:bocce lunch) ; :- Any
+;=> nil
+\end{minted}
+
+\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI,
+a company with a large production Typed Clojure system
+(section~\ref{sec:casestudy} presents a case study).
+
+\begin{minted}{clojure}
+(defalias RawKeyPair
+  "Unencrypted keypair -- extra keys disallowed"
+  (HMap :mandatory {:public-key RawKey,
+                    :private-key RawKey},
+        :complete? true))
+
+(defalias EncKeyPair
+  "Encrypted keypair -- extra keys disallowed"
+  (HMap :mandatory {:public-key RawKey,
+                    :enc-private-key EncKey},
+        :complete? true))
+
+(ann enc-keypair [RawKeyPair -> EncKeyPair])
+(defn enc-keypair "Encrypt an unencrypted keypair"
+  [{pk :private-key :as kp}] ; original map is kp
+  (assoc 
+    ; remove unencrypted private key
+    (dissoc kp :private-key)
+    ; add encrypted private key
+    :enc-private-key (encrypt pk)))
+\end{minted}
+%\begin{exmp}
+%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
+%\label{example:circleci}
+%\end{exmp}
+Fully specified HMap's are used to
+statically enforce the private key is not accidentally left in a supposedly
+encrypted keypair,
+
+\begin{minted}{clojure}
+(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
+(defn enc-keypair-bad
+  [{pk :private-key :as kp}]
+  (assoc kp :enc-private-key (encrypt pk)))
+\end{minted}
+
+%\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
+%dissociating the raw \clj{:private-key} entry with \clj{dissoc}
+%and associating an encrypted private key
+%as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
+%The expression \clj{(:private-key kp)} shows that keywords are also 
+%functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
+%Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
+%does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
+%operation forgot to remove it.
+
+%\egref{example:absentkeys}
+%is like \egref{example:circleci}
+%except the \clj{:absent-keys} HMap option is used
+%instead of \clj{:complete?},
+%which takes a \emph{set literal} of keywords that do not appear in the map, written 
+%with \emph{\#}-prefixed braces.
+%The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
+%aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
+%in the function body.
+%
+%\begin{exmp}
+%\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
+%\label{example:absentkeys}
+%\end{exmp}
+%
+%Since this example enforces that \clj{:private-key} must not appear
+%in a \clj{EncKeyPair}
+%Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
+%from the return value.
+%Now, however we could stash the raw private key in another entry
+%like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
+%without Typed Clojure noticing.
+
+%\paragraph{Branching on HMaps} Finally, testing on HMap properties
+%allows us to refine its type down branches. \clj{dec-map} takes an
+%\clj{Expr}, traverses to its nodes and decrements their values by \clj{dec}, then
+%builds the \clj{Expr} back up with the decremented nodes.
+%
+%\begin{exmp}
+%\inputminted[linenos,firstnumber=1,firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
+%\label{example:decmap}
+%\end{exmp}
+%
+%If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
+%we remove
+%the \clj{:do} and \clj{:const}
+%Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
+%and we are left with an \clj{:if} Expr.
+%On line 8,
+%we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
+%Line 9 know we can
+%remove the \clj{:const} Expr from the type of \clj{m} because it contradicts \clj{(= (:op m) :do)} being true,
+%and we know \clj{m} is a \clj{:do} Expr.
+%Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
+%so we are left with \clj{m} being a \clj{:const} Expr.
+%
+%Section~\ref{sec:formalpaths} discusses how this automatic reasoning is achieved.
+
+Partially specified HMap's with a common dispatch key
+combine naturally with ad-hoc unions.
+
+\begin{minted}{clojure}
+(defalias Order   ; define type abbreviation
+  "A meal order, tracking dessert quantities."
+  (U '{:Meal ':lunch ; keyword singleton type
+       :desserts Num}
+     '{:Meal ':dinner :desserts Num}
+     '{:Meal ':combo :meal1 Order :meal2 Order}))
+
+(ann desserts [Order -> Int])
+(defmulti desserts "Number of desserts per order."
+          :Meal)  ; dispatch on :Meal entry
+; map destructuring backwards, d is :desserts entry
+(defmethod desserts :lunch [{d :desserts}] d)
+(defmethod desserts :dinner [{d :desserts}] d)
+(defmethod desserts :combo [{m1 :meal1 m2 :meal2]
+  (+ (desserts m1) (desserts m2)))
+
+(cutlery {:Meal :combo 
+          :meal1 {:Meal :lunch :desserts 1}
+          :meal2 {:Meal :dinner :desserts 2}})
+;=> 3
+\end{minted}
+
+%\begin{exmp}
+%\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
+%\label{example:decleaf}
+%\end{exmp}
+%
+%The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
+%The function \clj{an-exp} is verified to return an \clj{Expr}.
+%
+%Here \clj{defalias} defines \clj{Expr}, a type abbreviation
+%that describes the structure of a recursively-defined AST as a union of HMaps.
+%Keyword singleton types are quoted---\clj{':lunch}.
+%A type that is a quoted map like \clj{'{:op ':if}} is a
+%HMap type with a fixed number of keyword entries of the specified types
+%known to be \emph{present},
+%zero entries known to absolutely be \emph{absent},
+%and an infinite number of \emph{unknown} entries entries.
+%Since only keyword keys are allowed, they do not require quoting.
 
@@
 and \clj{:const} methods.
-
-\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
-same length recursively call \clj{isa?} on the elements pairwise.
-\begin{minted}{clojure}
-  (isa? [Keyword Keyword] [Object Object]) ;=> true
-\end{minted}
-
-\egref{example:multidispatch}
-simulates multiple dispatch by dispatching on
-a vector containing the class of both arguments. \clj{open}
-takes two arguments which can be strings or files and returns
-a new file that concatenates their paths.
-
-We call three different \clj{File} constructors, each known at compile-time
-via type hints.
-Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
-except we update multiple bindings simultaneously.
-
-\subsection{Heterogeneous hash-maps}
-
-Hash-maps with keyword keys play a major role in Clojure programming.
-HMap types model the most common usages of keyword maps.
-
-\begin{exmp}
-\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
-\label{example:decleaf}
-\end{exmp}
-
-%The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
-%The function \clj{an-exp} is verified to return an \clj{Expr}.
-%
-Here \clj{defalias} defines \clj{Expr}, a type abbreviation
-that describes the structure of a recursively-defined AST as a union of HMaps.
-A quoted keyword in a type --- \clj{':if} --- is a singleton type that contains just the keyword.
-A type that is a quoted map like \clj{'{:op ':if}} is a
-HMap type with a fixed number of keyword entries of the specified types
-known to be \emph{present},
-zero entries known to absolutely be \emph{absent},
-and an infinite number of \emph{unknown} entries entries.
-Since only keyword keys are allowed, they do not require quoting.
-
-\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI,
-a company with a large production Typed Clojure system
-(section~\ref{sec:casestudy} presents a case study).
-
-\begin{exmp}
-\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
-\label{example:circleci}
-\end{exmp}
-
-\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
-dissociating the raw \clj{:private-key} entry with \clj{dissoc}
-and associating an encrypted private key
-as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
-The expression \clj{(:private-key kp)} shows that keywords are also 
-functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
-Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
-does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
-operation forgot to remove it.
-
-\egref{example:absentkeys}
-is like \egref{example:circleci}
-except the \clj{:absent-keys} HMap option is used
-instead of \clj{:complete?},
-which takes a \emph{set literal} of keywords that do not appear in the map, written 
-with \emph{\#}-prefixed braces.
-The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
-aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
-in the function body.
-
-\begin{exmp}
-\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
-\label{example:absentkeys}
-\end{exmp}
-
-Since this example enforces that \clj{:private-key} must not appear
-in a \clj{EncKeyPair}
-Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
-from the return value.
-Now, however we could stash the raw private key in another entry
-like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
-without Typed Clojure noticing.
-
-\paragraph{Branching on HMaps} Finally, testing on HMap properties
-allows us to refine its type down branches. \clj{dec-map} takes an
-\clj{Expr}, traverses to its nodes and decrements their values by \clj{dec}, then
-builds the \clj{Expr} back up with the decremented nodes.
-
-\begin{exmp}
-\inputminted[linenos,firstnumber=1,firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
-\label{example:decmap}
-\end{exmp}
-
-If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
-we remove
-the \clj{:do} and \clj{:const}
-Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
-and we are left with an \clj{:if} Expr.
-On line 8,
-we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
-Line 9 know we can
-remove the \clj{:const} Expr from the type of \clj{m} because it contradicts \clj{(= (:op m) :do)} being true,
-and we know \clj{m} is a \clj{:do} Expr.
-Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
-so we are left with \clj{m} being a \clj{:const} Expr.
-
-Section~\ref{sec:formalpaths} discusses how this automatic reasoning is achieved.
 

### 2015-07-06 overview.tex
Change end of section 2

@@
 
-The \clj{hi} multimethod dispatches on its arguments via the identity function.
+The \clj{hi} multimethod dispatches on its argument via the identity function.
 
@@
 The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
-and \clj{:fr} have string values (in \clj{:mandatory}) and that 
-every other key is absent (in \clj{:complete?}).
+and \clj{:fr} have string values (expressed as a \clj{:mandatory} entry) and that 
+every other key is absent (by \clj{:complete?} being \clj{true}).
 
-HMap's default to being partially specified. The HMap shorthand annotating
-\clj{lunch}---below---forgets 
+HMap's default to being partially specified---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
+\clj{'{:en Str :fr Str}}
+forgets 
 information about absent keys, only providing \clj{:mandatory}
@@
 (def lunch {:en "muffin" :fr "baguette"})
-
 (:en lunch)    ; :- Str
@@
 
-\paragraph{HMaps in Practice} The next example is extracted from a production system at CircleCI,
+\paragraph{HMaps in practice} The next example is extracted from a production system at CircleCI,
 a company with a large production Typed Clojure system
@@
 %\end{exmp}
-Fully specified HMap's are used to
+Fully specified HMap's
 statically enforce the private key is not accidentally left in a supposedly
-encrypted keypair,
+encrypted keypair.
 
@@
   [{pk :private-key :as kp}]
-  (assoc kp :enc-private-key (encrypt pk)))
+  (assoc kp :enc-private-key (encrypt pk))
+  ; Type Error:
+  ;   Expected EncKeyPair, given 
+  ;   (HMap :mandatory {:enc-private-key EncKey
+  ;                     :private-key RawKey
+  ;                     :public-key RawKey}
+  ;         :complete? true)
+  )
 \end{minted}
+
+The extra \clj{:private-key} entry disagrees with \clj{EncKeyPair}, so a type error
+is given.
 
@@
 
-Partially specified HMap's with a common dispatch key
+\subsection{HMaps and multimethods, joined at the hip}
+
+Since HMaps are the primary way of specifying the structure of data in Clojure,
+and multimethods are the primary tool for dispatching on data, they are inevitably
+linked.
+There are infinite ways of both structuring and dispatching on data, so there is no
+hope predicting them all, or even manually covering useful a subset.
+
+Thankfully, occurrence typing is both extensible \emph{and} compositional.
+By extending occurrence typing with
+a handful of rules based on HMaps and other functions, 
+we can automatically cover almost all the common cases---as well others
+that compose simple rules in arbitrary ways.
+
+Futhermore, no special attention is needed for multimethod dispatch---the primitive branching
+mechanism is still the humble \clj{if} conditional. Only a small number of rules are needed
+to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
+In practice, this means all conditional-based control flow understood by occurrence typing
+also extends to the context of multimethod dispatch, and vice-versa.
+
+We first demonstrate a very common, simple dispatch style,
+then move on to deeper structural dispatching where occurrence typing's
+compositionality shines.
+
+\paragraph{HMaps and unions} Partially specified HMap's with a common dispatch key
 combine naturally with ad-hoc unions.
+An \clj{Order} is one of three kinds of HMaps.
 
@@
   (U '{:Meal ':lunch ; keyword singleton type
-       :desserts Num}
-     '{:Meal ':dinner :desserts Num}
+       :desserts Int}
+     '{:Meal ':dinner :desserts Int}
      '{:Meal ':combo :meal1 Order :meal2 Order}))
+\end{minted}
 
+The \clj{:Meal} entry is common to each HMap, always mapped to a known keyword singleton
+type.
+It's natural to dispatch on the \clj{class} of an instance---it's similarly
+natural to dispatch on a known entry like \clj{:Meal}.
+
+\begin{minted}{clojure}
 (ann desserts [Order -> Int])
@@
 (defmethod desserts :dinner [{d :desserts}] d)
-(defmethod desserts :combo [{m1 :meal1 m2 :meal2]
+(defmethod desserts :combo [{m1 :meal1 m2 :meal2}]
   (+ (desserts m1) (desserts m2)))
 
-(cutlery {:Meal :combo 
+(desserts {:Meal :combo 
            :meal1 {:Meal :lunch :desserts 1}
@@
 \end{minted}
+
+The \clj{:combo} method is verified to only structurally recur
+on \clj{Order}s. This is achieved because we learn the argument
+must be at least of type
+\clj{'{:Meal :combo}}
+since
+\clj{(isa? (:Meal o) :combo)}
+must be true. This eliminates possibility of \clj{:lunch} and \clj{:dinner}
+orders, deducing
+\begin{minted}{clojure}
+'{:Meal ':combo :meal1 Order :meal2 Order}
+\end{minted}
+as the argument type, which can safely structurally recur.
+
+\paragraph{Arbitrary dispatch}
+An equally valid dispatch mechanism for \clj{desserts}
+would be on the \clj{class} of the \clj{:desserts} key.
+We have already seen dispatch on \clj{class} and on keywords
+in isolation---occurrence typing automatically understands
+control flow that combines its simple building blocks.
+
+\begin{minted}{clojure}
+(ann desserts' [Order -> Int])
+(defmulti desserts' 
+          (fn [o :- Order] (class (:desserts o))))
+(defmethod desserts' Long [{d :desserts}] d)
+(defmethod desserts' nil [{m1 :meal1 m2 :meal2}]
+  (+ (desserts' m1) (desserts' m2)))
+
+(desserts' {:Meal :combo 
+            :meal1 {:Meal :lunch :desserts 1}
+            :meal2 {:Meal :dinner :desserts 2}})
+;=> 3
+\end{minted}
+
+In the \clj{Long} method, Typed Clojure learns that
+its argument is at least of type \clj{'{:desserts Long}}---since
+\begin{minted}{clojure}
+(isa? (class (:desserts o)) Long)
+\end{minted}
+must be true.
+
+Combined with the knowledge that it is also an
+\clj{Order}, we can eliminate the possibility of
+a \clj{:combo} meal, and infer the type
+\begin{minted}{clojure}
+(U '{:Meal ':lunch :desserts Int}
+   '{:Meal ':dinner :desserts Int})
+\end{minted}
+Looking up \clj{:desserts} on this type returns an \clj{Int}.
+
+In the \clj{nil} method, we know
+\begin{minted}{clojure}
+(isa? (class (:desserts o)) nil)
+\end{minted}
+is true---which implies \clj{(class (:desserts o))} is \clj{nil}.
+
+Since lookups on missing keys return \clj{nil}, there are
+now two possibilities:
+\begin{itemize}
+  \item either \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
+  \item \clj{o} is missing a \clj{:desserts} entry.
+\end{itemize}
+In a type, we learn \clj{o} is at least
+\begin{minted}{clojure}
+(U '{:desserts nil}
+   ; :absent-keys, a set of known absent entries
+   (HMap :absent-keys #{:desserts}))
+\end{minted}
+
+This is enough to infer \clj{o} is a \clj{:combo} meal, thus verifying both
+structural recursions.
 
@@
 
-\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
-In \egref{example:incmap} we
-dispatch on the \clj{:op} key 
-of our HMap AST \clj{Expr}.
-Since keywords are functions that look themselves up in their argument, we simply
-use \clj{:op} as the dispatch function.
-
-\begin{exmp}
-\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
-\label{example:incmap}
-\end{exmp}
-
-The function \clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
-The reasoning is similar, except we only consider one branch (the current method) by
-locally considering the current \emph{dispatch value} and reasoning about how it relates
-to the \emph{dispatch function}.
-For example, 
-in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
-narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
-and \clj{:const} methods.
-
-\subsection{Final example}
-
-\egref{example:final}
-combines everything we will cover for the rest of the paper:
-multimethod dispatch, reflection resolution via type hints, Java method
-and constructor calls, conditional and exceptional flow reasoning,
-and HMaps.
-
-\begin{figure}
-\begin{exmp}
-\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
-\label{example:multidispatch}
-\end{exmp}
-\begin{exmp}
-\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
-\label{example:final}
-\end{exmp}
-\caption{Multimethod Examples}
-\end{figure}
-
-We dispatch on \clj{:p} to distinguish the two cases of \clj{FSM}---for example on \clj{:F}
-we know the \clj{:file} is a file.
-The body of the first method uses type hints to resolve reflection
-and conditional control flow to prove null-pointer exceptions are impossible.
-The second method is similar except it uses exceptional control flow.
+%\paragraph{HMap dispatch} The flexibility of \clj{isa?} is key to the generality of multimethods. 
+%In \egref{example:incmap} we
+%dispatch on the \clj{:op} key 
+%of our HMap AST \clj{Expr}.
+%Since keywords are functions that look themselves up in their argument, we simply
+%use \clj{:op} as the dispatch function.
+%
+%\begin{exmp}
+%\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
+%\label{example:incmap}
+%\end{exmp}
+%
+%The function \clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
+%The reasoning is similar, except we only consider one branch (the current method) by
+%locally considering the current \emph{dispatch value} and reasoning about how it relates
+%to the \emph{dispatch function}.
+%For example, 
+%in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
+%narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
+%and \clj{:const} methods.
+%
+%
+%\subsection{Final example}
+%
+%\egref{example:final}
+%combines everything we will cover for the rest of the paper:
+%multimethod dispatch, reflection resolution via type hints, Java method
+%and constructor calls, conditional and exceptional flow reasoning,
+%and HMaps. 
+%
+%
+%\begin{figure}
+%\begin{exmp}
+%\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
+%\label{example:multidispatch}
+%\end{exmp}
+%\begin{exmp}
+%\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
+%\label{example:final}
+%\end{exmp}
+%\caption{Multimethod Examples}
+%\end{figure}
+%
+%We dispatch on \clj{:p} to distinguish the two cases of \clj{FSM}---for example on \clj{:F}
+%we know the \clj{:file} is a file.
+%The body of the first method uses type hints to resolve reflection
+%and conditional control flow to prove null-pointer exceptions are impossible.
+%The second method is similar except it uses exceptional control flow.

### 2015-07-06 overview.tex
wrap up section 2

@@
 comma is only added when the argument is non-nil.
+
+Typed Clojure guarantees that well-typed code cannot
+dereference the null-pointer.
+This is especially important for Clojure programs---\clj{nil}
+is treated like any other distinct datum in Clojure,  and its status as
+one of only two false values means it is a common choice to indicate
+``nothing'' or ``failure''.
 
@@
 
+The following Typed Clojure program constructs a new \clj{File}
+instance and calls the \clj{getParent} method on the result.
+
 \begin{minted}{clj}
@@
 
-If a specific constructor, method, or field cannot be found based on the
+Typed Clojure helps the Clojure compiler avoid expensive reflective 
+calls,
+however if a specific constructor, method, or field cannot be found based on the
 static types of its arguments, a type error is thrown.
@@
 
-The conditional guards from dereferencing \clj{nil}, and as before
-removing it is a static type error.
+The conditional guards from dereferencing \clj{nil}, and---as before---removing 
+it is a static type error, as typed code could possibly dereference \clj{nil}.
 
@@
 a \emph{dispatch function} with one or more \emph{methods}.
-When invoked, the arguments are first supplied to the dispatch function, yielding
+
+\paragraph{Value-based dispatch}
+Here is a simple multimethod.
+
+\begin{minted}{clojure}
+(ann hi [Kw -> Str]) ; interface type for `hi`
+(defmulti hi identity) ; dispatch function `identity`
+(defmethod hi :en [_] "hello") ; method for `:en`
+(defmethod hi :fr [_] "bonjour") ; method for `:fr`
+(defmethod hi :default [_] "um...") ; default method
+\end{minted}
+
+When invoked, the arguments are first supplied to the dispatch function---\clj{identity}---yielding
 a \emph{dispatch value}. A method is then chosen
@@
 to finally return a value for the entire expression.
+
+\begin{minted}{clojure}
+(map hi [:en :fr :bocce]) ; map over keyword vector 
+;=> ("hello" "bonjour" "um...")
+\end{minted}
+
+For example, 
+\clj{(hi :en)} evaluates to \clj{"hello"}---it executes
+the \clj{:en} method
+because \clj{(= (identity :en) :en)} is true
+and \clj{(= (identity :en) :fr)} is false.
+
+This is only part of the story for multimethod dispatch.
+
+\paragraph{Class-based dispatch}
+For class values, multimethods choose methods based on subclassing
+relationships.
 
@@
 
-The dispatch function \clj{class}---associated at multimethod creation with \clj{defmulti}---dictates 
-whether the \clj{String} or \clj{File} method is chosen---both installed via \clj{defmethod}.
+The dispatch function \clj{class}
+%---associated at multimethod creation with \clj{defmulti}---
+dictates 
+whether the \clj{String} or \clj{File} method is chosen.
+%---both installed via \clj{defmethod}
 
@@
 \begin{minted}{clojure}
-  (isa? String Object) ;=> true
-  (isa? Object String) ;=> false
-  ; keywords are interned symbols with a : prefix
-  (isa? :a :a) ;=> true
-  (isa? :a :b) ;=> false
+(isa? (identity :en) :en) ;=> true
+(isa? (identity :en) :fr) ;=> false
+(isa? (class "STAINS/JELLY") String) ;=> true
+(isa? (class "STAINS/JELLY") Object) ;=> true
+(isa? (class (new File "JELLY")) String) ;=> false
 \end{minted}
@@
 %giving the final result.
-
-Dispatching on a non-class value is common, as well as specifying
-a \clj{:default} method for when no methods are matched.
-
-\begin{minted}{clojure}
-(ann hi [Kw -> Str])
-(defmulti hi identity)
-(defmethod hi :en [_] "hello")
-(defmethod hi :fr [_] "bonjour")
-(defmethod hi :default [_] "um...")
-
-(str (hi :en) " " (hi :fr)) ;=> "hello bonjour"
-(hi :bocce)   ;=> "um..."
-\end{minted}
-
-The \clj{hi} multimethod dispatches on its argument via the identity function.
 
@@
 
-Immutable hash-maps with keyword keys play a major role in Clojure programming.
+Immutable hash-maps with keyword keys play a major role in Clojure programming
+as the primary way to structure data.
 
@@
 (def breakfast {:en "waffles" :fr "croissants"})
-
-breakfast ; :- (HMap :mandatory {:en Str :fr Str}
-          ;          :complete? true)
 
@@
 HMap types model the most common usages of keyword maps.
+
+\begin{minted}{clojure}
+breakfast ; :- (HMap :mandatory {:en Str :fr Str}
+          ;          :complete? true)
+\end{minted}
+
 The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
@@
 to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
-In practice, this means all conditional-based control flow understood by occurrence typing
+In practice, this means all conditional-based control flow analysis
 also extends to the context of multimethod dispatch, and vice-versa.
@@
 (ann desserts [Order -> Int])
-(defmulti desserts "Number of desserts per order."
+(defmulti desserts "Total desserts per order."
           :Meal)  ; dispatch on :Meal entry
@@
 The \clj{:combo} method is verified to only structurally recur
-on \clj{Order}s. This is achieved because we learn the argument
-must be at least of type
+on \clj{Order}s. This is achieved because we learn the argument--- \clj{o} ---must % em dash needs space here, looks weird
+be of type
 \clj{'{:Meal :combo}}
@@
 \clj{(isa? (:Meal o) :combo)}
-must be true. This eliminates possibility of \clj{:lunch} and \clj{:dinner}
-orders, deducing
+must be true. Combining \clj{'{:Meal :combo}}
+with the fact that \clj{o} is an \clj{Order}
+eliminates possibility of \clj{:lunch} and \clj{:dinner}
+orders, simplifying \clj{o} to
 \begin{minted}{clojure}
@@
 \end{minted}
-as the argument type, which can safely structurally recur.
+which can safely structurally recur.
 
-\paragraph{Arbitrary dispatch}
+\paragraph{Nested dispatch}
 An equally valid dispatch mechanism for \clj{desserts}
@@
           (fn [o :- Order] (class (:desserts o))))
-(defmethod desserts' Long [{d :desserts}] d)
-(defmethod desserts' nil [{m1 :meal1 m2 :meal2}]
+(defmethod desserts' Long [{d :desserts :as o}] 
+  ; o :- (U '{:Meal ':lunch :desserts Int}
+  ;         '{:Meal ':dinner :desserts Int})
+  d)
+(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
+  ; o :- '{:Meal ':combo :meal1 Order :meal2 Order}
   (+ (desserts' m1) (desserts' m2)))
-
-(desserts' {:Meal :combo 
-            :meal1 {:Meal :lunch :desserts 1}
-            :meal2 {:Meal :dinner :desserts 2}})
-;=> 3
 \end{minted}
+%
+%(desserts' {:Meal :combo 
+%            :meal1 {:Meal :lunch :desserts 1}
+%            :meal2 {:Meal :dinner :desserts 2}})
+%;=> 3
 
@@
 
-Combined with the knowledge that it is also an
+Knowing \clj{o} is also an
 \clj{Order}, we can eliminate the possibility of
@@
 
-Since lookups on missing keys return \clj{nil}, there are
-now two possibilities:
+Since lookups on missing keys return \clj{nil}, either
 \begin{itemize}
-  \item either \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
+  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
   \item \clj{o} is missing a \clj{:desserts} entry.
 \end{itemize}
-In a type, we learn \clj{o} is at least
+Equivalently, we learn \clj{o} is at least of type
 \begin{minted}{clojure}
@@
 \end{minted}
+This intersected with \clj{o} being an \clj{Order} simplifies to a \clj{:combo} meal, 
+\begin{minted}{clojure}
+'{:Meal ':combo :meal1 Order :meal2 Order}
+\end{minted}
+thus verifying both structural recursions.
 
-This is enough to infer \clj{o} is a \clj{:combo} meal, thus verifying both
-structural recursions.
+\paragraph{From multiple to arbitrary dispatch}
+Multimethod dispatch goes even further, supporting
+multiple dispatch via vectors.
+
+Fully specifying multimethod dispatch is beyond the scope of this paper,
+but the same intuition applies---adding support for multiple dispatch
+automatically allows arbitrary combinations and nestings
+of it and previous simple dispatch rules.
 

### 2015-07-06 formal-model.tex
Rewrite some of start of section 3

--- 2015-06-15 formal-model.tex
@@
 \lambdatc{}.
+
+Building on occurrence typing,
+we incrementally add each
+novel feature of Typed Clojure to the formalism,
+interleaving presentation of syntax, typing rules, operational semantics,
+and subtyping.
+
+\subsection{Occurrence typing}
+
 Our presentation will start with a review of
-occurrence typing~\cite{TF10}.
-Then for the rest of the section we incrementally add each
-novel feature of Typed Clojure to the formalism,
-interleaving presentation of syntax, typing rules, operational semantics
-and subtyping.
+occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
+
+\paragraph{Logical formulas over type environments}
+Standard type systems use type environments to track lexical type information.
+Introducing a lexically scoped variable $x$ is a matter of associating 
+associating $x$ with its type $\Number$ (a number) in the type environment.
+Correspondingly, checking the type of a variable is a simple map lookup.
+$$
+\begin{array}{lr}
+\judgementtwo{\x{} \mapsto \Number}{\appexp{\inc{}}{\x{}}}{\Number}
+&
+\infer [LC-Local]
+{ {\propenv{}}(\x{}) = {\t{}}
+}
+{ \judgementtwo {\propenv{}} 
+            {\x{}} {\t{}}
+           }
+\end{array}
+$$
+
+Occurrence typing takes a different approach.
+The same information is conveyed by \emph{logical formulas}.
+The logical statement {\isprop{\Number{}}{\x{}}} says the value
+obtained via $x$ is of type \Number{}. We now must \emph{prove}
+a variable's type via a \emph{logical system}.
+$$
+\begin{array}{lr}
+\judgementtwo{\isprop{\Number{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\Number}
+&
+\infer [T-LocalProve]
+{ \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
+{ \judgementtwo {\propenv{}} 
+            {\x{}} {\t{}}
+           }
+\end{array}
+$$
+The proof system judgment { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
+says, under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
+
+\paragraph{Symbolic representations}
+As we saw in \secref{sec:overview},
+occurrence typing is capable of reasoning
+about deeply nested expressions.
+For each expression, the type system assigns a symbolic representation
+with respect to the current lexical environment.
+
+$x$ is represented as $x$; $\appexpone{\ccclass{\appexp{\makekw{lunch}}{m}}}$ as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{m}}}$;
+$42$ has no representation, called \emptyobject{}.
+
+This symbolic representation called the \emph{object}---\emptyobject{} is the
+\emph{empty object}.
+
+%The particular part of the runtime environment we reference in a
+%type proposition is called the \emph{object}.
+%The typing judgment relates an object to every expression in the language.
+%An object is either \emph{empty}, written \emptyobject{}, 
+%which says 
+%this expression is not known to evaluate to a particular part
+%  of the current runtime environment, or a 
+%variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
+%that exactly indicates how the value of this
+%expression can be derived from the current runtime environment.
+%Type propositions can only reference non-empty objects.
+
+\paragraph{Logical statements about objects}
+So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}},
+saying variable \x{} is of type \t{}.
+We extend logical statements to reason about \emph{arbitrary} non-empty 
+objects.
+
+To illustrate, recall \egref{example:desserts-on-meal}.
+We know the parameter $o$ is of type $\Order$
+by the annotation on $desserts$---written
+\begin{equation}
+\qquad {\isprop{\Order}{o}}
+\end{equation}
+as a logical statement.
+
+In checking the ${\makekw{combo}}$ method, we also assume
+${\appexp{\makekw{Meal}}{o}}$ is ${\makekw{combo}}$,
+based on multimethod dispatch rules. This is written
+\begin{equation}
+  \qquad {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{o}}}
+\end{equation}
+and pronounced ``the ${\makekw{Meal}}$ path of variable $o$ is of type
+{\Value{\makekw{combo}}}'', a keyword singleton type.
+
+\paragraph{Logical system in action} Using our two facts, T-LocalProve demands
+we must solve for \t{} in
+$$
+{ \inpropenv 
+  {{\isprop{\Order}{o}},
+    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{o}}}}
+  {\isprop {\t{}} {o}}}
+$$
+Now the purpose of a logical representation can be conveyed---we rely on the logical
+system to combine pieces of type information to deduce more accurate types for lexical
+bindings.
+
+% TODO introduction rules for logical statements
+\paragraph{}
+
+\paragraph{Extended to propositional logic}
 
@@
 (\figref{main:figure:termsyntax}).
-
-The particular part of the runtime environment we reference in a
-type proposition is called the \emph{object}.
-The typing judgment relates an object to every expression in the language.
-An object is either \emph{empty}, written \emptyobject{}, 
-which says 
-this expression is not known to evaluate to a particular part
-  of the current runtime environment, or a 
-variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
-that exactly indicates how the value of this
-expression can be derived from the current runtime environment.
-Type propositions can only reference non-empty objects.
 
@@
 conditional tests.
-For instance, \egref{example:conditionalflow}
-only utilizes \clj{x} once
+For instance,
+\egref{example:parent-if}
+only utilizes \clj{f} once
 the programmer is convinced it is safe to do so based whether
-\clj{(number? x)}
+\clj{f}
 is
@@
 for when it reduces to a true value, and an `else' proposition
-when it reduces to a false value---for \clj{(number? x)}
-the then proposition is {\isprop{\NumberFull}{\x{}}} and 
-the else proposition is {\notprop{\NumberFull}{\x{}}}.
+when it reduces to a false value---for \clj{f}
+the then proposition is {\notprop{\falsy}{f}} and 
+the else proposition is {\isprop{\falsy}{f}}.
 %\ref{main:figure:typingrules}
@@
 We are now equipped to type check
-\egref{example:conditionalflow}, starting at body:
+\egref{example:parent-if}:
 $$
-\clj{... (if (number? x) (inc x) 0) ...}
+\clj{(if f (.getParent f) nil)}
 $$
 
-We know {\propenv{}} = {\isprop{\UnionNilNum{}}{\x{}}}.
-The test expression uses T-App, 
+With {\propenv{}} = {\isprop{\UnionNilFile{}}{f}},
 $$
-\judgement{\propenv{}}{\appexp{\numberhuh{}}{\x{}}}{\Boolean}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
+\judgement{\propenv{}}{f}{\UnionNilFile{}}{\localfilterset{f}}{f}
 $$
-since \numberhuh{} has type
-{\ArrowOne{\x{}}{\Top}{\Boolean}
-        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
-      and \x{} has object \x{}.
+via T-Local.
 
-Finally we check both branches using the extended proposition environment as specified by T-If.
-Going down the then branch, our new assumption {\isprop{\Number}{\x{}}} is crucial to check
+Checking the then branch involves extending
+the proposition environment with {\notprop{\falsy}{f}}
 $$
@@
 
-\subsection{Reasoning about Exceptional Control Flow}
-\label{sec:doformal}
-
-We extend our model with sequencing expressions and errors, where {\errorvalv{}}
-models the result of calling Clojure's \clj{throw} special form
-with some \clj{Throwable}.
-
-\smallskip
-$
-\begin{altgrammar}
-  \e{} &::=& \ldots \alt {\errorvalv{}} \alt {\doexp {\e{}} {\e{}}} &\mbox{Expressions} 
-\end{altgrammar}
-$
-
-\smallskip
+%\subsection{Reasoning about Exceptional Control Flow}
+%\label{sec:doformal}
 %
-%B-Do simply evaluates its arguments sequentially and returns the right argument.
-%Since errors are not values, we define error propagation semantics
-%like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).
+%Along with conditional control flow,
+%Clojure programmers rely on \emph{exceptions}
+%to assert type-related invariants.
+%
+%\begin{exmp}
+%\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
+%\label{example:doexception}
+%\end{exmp}
+%
+%The fully expanded increment function in~\egref{example:doexception}
+%guards its final call with a number check, preventing
+%a possible null-pointer exception.
+%Without this check, the type system would reject the program.
+%
+%To check this example,
+%occurrence typing 
+%automatically
+%assumes
+%\clj{x} is a number when checking the second \clj{do} subexpression
+%based on the first subexpression.
+%\footnote{See \url{https://github.com/typedclojure/examples}
+%  for full examples.}
+%We model this formally %(section~\ref{sec:doformal}) 
+%and prove
+%null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).
+%
+%
+%We extend our model with sequencing expressions and errors, where {\errorvalv{}}
+%models the result of calling Clojure's \clj{throw} special form
+%with some \clj{Throwable}.
+%
+%\smallskip
+%$
+%\begin{altgrammar}
+%  \e{} &::=& \ldots \alt {\errorvalv{}} \alt {\doexp {\e{}} {\e{}}} &\mbox{Expressions} 
+%\end{altgrammar}
+%$
+%
+%\smallskip
+%%
+%%B-Do simply evaluates its arguments sequentially and returns the right argument.
+%%Since errors are not values, we define error propagation semantics
+%%like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).
+%%
+%%\begin{mathpar}
+%%    {\BDo}
+%%
+%%\infer [BE-Error]
+%%{}
+%%{ \opsem {\openv{}} 
+%%         {\errorvalv{}}
+%%         {\errorvalv{}}}
+%%
+%%\infer [BE-Do1]
+%%{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
+%%{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
+%%\end{mathpar}
+%
+%Our main insight is as follows: 
+%if the first subexpression in a sequence reduces to a value, then it is either true or false.
+%If we learn some proposition in both cases then we can use that proposition as an assumption to check the second subexpression.
+%T-Do formalizes this intuition.
 %
 %\begin{mathpar}
-%    {\BDo}
+%    {\TDo}  
+%\end{mathpar}
 %
-%\infer [BE-Error]
-%{}
-%{ \opsem {\openv{}} 
-%         {\errorvalv{}}
-%         {\errorvalv{}}}
+%The introduction of errors, 
+%which do not evaluate to either
+%a true or false value,
+%makes our insight interesting.
 %
-%\infer [BE-Do1]
-%{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
-%{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
+%\begin{mathpar}
+%    {\TError}
 %\end{mathpar}
-
-Our main insight is as follows: 
-if the first subexpression in a sequence reduces to a value, then it is either true or false.
-If we learn some proposition in both cases then we can use that proposition as an assumption to check the second subexpression.
-T-Do formalizes this intuition.
-
-\begin{mathpar}
-    {\TDo}  
-\end{mathpar}
-
-The introduction of errors, 
-which do not evaluate to either
-a true or false value,
-makes our insight interesting.
-
-\begin{mathpar}
-    {\TError}
-\end{mathpar}
-
-Recall \egref{example:doexception}.
-\begin{minted}{clojure}
-...
-  (do (if (number? x) nil (throw (new Exception)))
-      (inc x)) 
-...
-\end{minted}
-
-As before, checking \appexp{\numberhuh{}}{\x{}} allows us to use the proposition \isprop{\Number}{\x{}}
-when checking the then branch.
-
-By T-Nil and subsumption we can propagate this  information to both propositions.
-$$
-\judgement{\isprop{\Number}{\x{}}}{\nil{}}{\Nil{}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
-$$
-Furthermore, using T-Error
-and subsumption we can conclude anything in the else branch.
-$$
-\judgement{\notprop{\Number}{\x{}}}{\errorvalv{}}{\Bot}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
-$$
-Using the above as premises to T-If we conclude that if the first
-expression in the \doliteral{} evaluates successfully, \isprop{\Number}{\x{}} must be true.
-$$
-\judgement{\isprop{\UnionNilNum}{\x{}}}
-          {\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}
-          {\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
-$$
-We can now use \isprop{\Number}{\x{}} in the environment to check the second subexpression
-{\appexp{\inc{}}{\x{}}}, completing the example.
+%
+%Recall \egref{example:doexception}.
+%\begin{minted}{clojure}
+%...
+%  (do (if (number? x) nil (throw (new Exception)))
+%      (inc x)) 
+%...
+%\end{minted}
+%
+%As before, checking \appexp{\numberhuh{}}{\x{}} allows us to use the proposition \isprop{\Number}{\x{}}
+%when checking the then branch.
+%
+%By T-Nil and subsumption we can propagate this  information to both propositions.
+%$$
+%\judgement{\isprop{\Number}{\x{}}}{\nil{}}{\Nil{}}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
+%$$
+%Furthermore, using T-Error
+%and subsumption we can conclude anything in the else branch.
+%$$
+%\judgement{\notprop{\Number}{\x{}}}{\errorvalv{}}{\Bot}{\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
+%$$
+%Using the above as premises to T-If we conclude that if the first
+%expression in the \doliteral{} evaluates successfully, \isprop{\Number}{\x{}} must be true.
+%$$
+%\judgement{\isprop{\UnionNilNum}{\x{}}}
+%          {\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}
+%          {\filterset{\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}}{\emptyobject{}}
+%$$
+%We can now use \isprop{\Number}{\x{}} in the environment to check the second subexpression
+%{\appexp{\inc{}}{\x{}}}, completing the example.
 
@@
     \e{} &::=& \ldots   
-            \alt \reflectiveexpsyntax{} &\mbox{Expressions}\\
+                      {\fieldexp {\fld{}} {\e{}}}
+                      \alt {\methodexp {\mth{}} {\e{}} {\overrightarrow{\e{}}}}&\mbox{Expressions}\\
+                      &\alt& {\newexp {\class{}} {\overrightarrow{\e{}}}}\\
      &\alt& \mininonreflectiveexpsyntax{}

### 2015-07-06 overview.tex
Rewrite some of start of section 3

@@
 \begin{exmp}
-\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/parent3.clj}
+\begin{minted}{clojure}
+(ann parent [(U nil File) -> (U nil Str)])
+(defn parent [f]
+  (if f (.getParent f) nil))
+\end{minted}
+\label{example:parent-if}
 \end{exmp}
@@
 HMap types model the most common usages of keyword maps.
-
 \begin{minted}{clojure}
@@
 \end{minted}
-
 The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
@@
 
+\begin{exmp}
 \begin{minted}{clojure}
@@
 \end{minted}
+\label{example:desserts-on-meal}
+\end{exmp}
 

### 2015-07-06 formal-model.tex
Reorder section 3.

@@
 occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
+We build up the occurrence typing calculus for illustrative purposes, 
+and present the full syntax at the end of the section.
 
@@
 The proof system judgment { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
-says, under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
+says logical assumptions {\propenv{}} prove variable \x{} is of type \t{}.
 
@@
 For each expression, the type system assigns a symbolic representation
-with respect to the current lexical environment.
-
-$x$ is represented as $x$; $\appexpone{\ccclass{\appexp{\makekw{lunch}}{m}}}$ as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{m}}}$;
-$42$ has no representation, called \emptyobject{}.
-
-This symbolic representation called the \emph{object}---\emptyobject{} is the
-\emph{empty object}.
+with respect to the current lexical environment. For example,
+
+\begin{itemize}
+  \item $x$ is represented as $x$; 
+  \item $\appexpone{\ccclass{\appexp{\makekw{lunch}}{m}}}$ is represented as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{m}}}$;
+  \item $42$ has no representation, called \emptyobject{}.
+\end{itemize}
+
+This symbolic representation is called the expression's \emph{object}. 
+\figref{main:figure:termsyntax} gives the syntax for objects.
+\emptyobject{} is the
+\emph{empty object}. Non-empty objects \path{\pathelem{}}{\x{}} consist of a symbol \x{} and a \emph{path} \pathelem{},
+which is a possibly-empty sequence of \emph{path elements} applied right-to-left.
+Path elements are {\classpe{}} and {\keype{k}}, representing a call to \classconst{} and
+a keyword lookup on $k$ respectively.
 
@@
 We extend logical statements to reason about \emph{arbitrary} non-empty 
-objects.
+objects {\isprop{\t{}}{\path{\pathelem{}}{\x{}}}}.
 
@@
 as a logical statement.
-
+%
 In checking the ${\makekw{combo}}$ method, we also assume
@@
 
-\paragraph{Logical system in action} Using our two facts, T-LocalProve demands
-we must solve for \t{} in
+\paragraph{Logical system in action} At some point in checking the {\makekw{combo}} method,
+we will need the type of $o$.
+Our proposition environment consists of both above facts
+$$
+\propenv{} = {{\isprop{\Order}{o}},
+    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{o}}}}
+$$
+and T-LocalProve says $o$ is of type \t{} if \propenv{} \emph{proves} $o$ is of type \t{}
 $$
 { \inpropenv 
-  {{\isprop{\Order}{o}},
-    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{o}}}}
+  {\propenv{}}
   {\isprop {\t{}} {o}}}
 $$
+
 Now the purpose of a logical representation can be conveyed---we rely on the logical
@@
 % TODO introduction rules for logical statements
-\paragraph{}
-
-\paragraph{Extended to propositional logic}
-
-The first insight about occurrence typing is that
-logical formulas
-can be used to represent type information about our programs
-by relating parts of the runtime environment to types
-via propositional logic.
-\emph{Type Propositions} \prop{} make assertions like ``variable \x{} is of type \NumberFull{}'' or
-``variable \x{} is not \nil{}''---in our logical system we write these as
-{\isprop{\NumberFull}{\x{}}}
-and {\notprop{\Nil{}}{\x{}}}. 
+\paragraph{Extension to propositional logic}
+The kinds of logical statements are extended to \emph{propositional logic}.
+\figref{main:figure:termsyntax} describes the syntax we will use for the rest of the paper.
+The proposition metavariable \prop{} 
+consists of positive and negative \emph{type propositions} 
+about non-empty objects---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+respectively---the former we have seen before, 
+the latter pronounced ``the object {\path {\pathelem{}} {\x{}}} is \emph{not} of type \t{}''.
 The other propositions are standard logical connectives: implications, conjunctions,
-disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
-(\figref{main:figure:termsyntax}).
-
-The second insight is that we can replace the traditional 
-representation of a
-type environment (eg., a map from variables to types)
-with a set of propositions, written \propenv{}. 
-Instead of mapping \x{} to
-the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
-
-Given a set of propositions, we can use logical reasoning to derive
-new information about our programs
-with the judgment \inpropenv{\propenv{}}{\prop{}}.
-In addition to the standard rules for the logical connectives, the key
-rule is L-Update, which combines multiple propositions about the same variable,
-allowing us to refine its type.
-$$
-  {\LUpdate}
-$$
-For example, with L-Update we can use the knowledge of
-\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
-and 
-\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
-to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
-(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
-We cover L-Update in more detail in \secref{sec:formalpaths}.
-
-Finally, this approach allows the type system to track
-programming idioms from 
-dynamic languages
-using implicit type-based reasoning based on the result of
-conditional tests.
-For instance,
-\egref{example:parent-if}
-only utilizes \clj{f} once
-the programmer is convinced it is safe to do so based whether
-\clj{f}
-is
-true or false. 
-To express this in the type system, every expression 
-is described by two propositions: a `then' proposition
-for when it reduces to a true value, and an `else' proposition
-when it reduces to a false value---for \clj{f}
-the then proposition is {\notprop{\falsy}{f}} and 
-the else proposition is {\isprop{\falsy}{f}}.
+disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions.
+
+%The first insight about occurrence typing is that
+%logical formulas
+%can be used to represent type information about our programs
+%by relating parts of the runtime environment to types
+%via propositional logic.
+%\emph{Type propositions}  make assertions like ``variable \x{} is of type \NumberFull{}'' or
+%``variable \x{} is not \nil{}''---in our logical system we write these as
+%{\isprop{\NumberFull}{\x{}}}
+%and {\notprop{\Nil{}}{\x{}}}. 
+%
+%The second insight is that we can replace the traditional 
+%representation of a
+%type environment (eg., a map from variables to types)
+%with a set of propositions, written \propenv{}. 
+%Instead of mapping \x{} to
+%the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
+
+\paragraph{`Then' and `else' propositions} Along with an object, each expression carries two
+propositions---a `then' proposition that can be assumed if the expression evaluates to a
+true value, and an `else' proposition for evaluating to a false value.
+
+For example, the `then' proposition for the expression $o$ is $\notprop{\falsy{}}{o}$.
+This says if the expression $o$ evaluates to a true value, then $o$ is neither
+of type \False or \Nil. Naturally, the `else' proposition $\isprop{\falsy{}}{o}$
+says $o$ is either \False or \Nil.
+
+Together, both propositions are called a \emph{proposition set}.
+
+\paragraph{The full judgment}
+
+We formalize our type system following~\citet{TF10}.
+%(with differences highlighted in $\trdiff{\text{blue}}$)
+The typing judgment 
+$$
+{\judgementrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}
+  {\ep{}}}
+$$
+says expression \e{} rewrites to \ep{}, which
+is of type \t{} in the 
+proposition environment $\propenv{}$, with 
+`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+and object \object{}. We write 
+{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
+  or
+{\judgementselfrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}}
+for expressions that rewrite to themselves.
+
+The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
+application, abstractions, conditionals, and let expressions.
+All binding forms introduce fresh variables.
+Values include booleans, \nil{}, class literals, keywords, 
+numbers,
+constants and closures. 
+Value environments map local bindings to values.
+
+Types include the top type, \emph{untagged} unions, functions, singleton types
+and class instances. 
+We abbreviate \Booleanlong{} as \Boolean{}, \Keywordlong{} as \Keyword{}
+and \NumberFull{} and \Number{}.
+The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
+We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
+{\ValueTrue} as \True and {\ValueFalse} as {\False}.
+Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
+which may refer to the function argument.
+%Latent means they are relevant when the function is applied rather than evaluated.
+They are latent because they are instantiated with the
+actual object of the argument in applications before they are used in the proposition environment.
+
+\paragraph{Elimination rule for propositions}
+We can now review the main elimination rule for propositions, T-Local.
+$$
+    {\TLocal}
+$$
+It says, if proposition environment \propenv{} proves variable \x{} is 
+of type \t{}, then under proposition environment \propenv{}
+variable \x{} is of type \t{}. Also, if \x{} is a true value, then
+we know \x{} is neither \False or \Nil, and if \x{} is a false value,
+then \x{} is either \False or \Nil. Finally, the symbolic representation
+for \x{}---its object---is \x{}.
+
+\paragraph{Introduction rules for propositions}
+Every program branch is an opportunity to extend the proposition environment
+since we learn whether a test expression evaluates true or false. The T-If
+rule assumes the test's `then' proposition \thenprop{\prop{1}}
+in the then branch, and the test's `else' proposition \elseprop{\prop{1}}
+in the else branch.
+\begin{mathpar}
+  {\TIf}
+\end{mathpar}
+The `then' proposition is {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}}
+because if the entire expression reduces to a true value,
+then either the `then' branch is true---implying {\thenprop {\prop{2}}}---or
+the `else' branch is true---implying {\thenprop {\prop{3}}}.
+
+%Given a set of propositions, we can use logical reasoning to derive
+%new information about our programs
+%with the judgment \inpropenv{\propenv{}}{\prop{}}.
+%In addition to the standard rules for the logical connectives, the key
+%rule is L-Update, which combines multiple propositions about the same variable,
+%allowing us to refine its type.
+%$$
+%  {\LUpdate}
+%$$
+%For example, with L-Update we can use the knowledge of
+%\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
+%and 
+%\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
+%to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
+%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
+%We cover L-Update in more detail in \secref{sec:formalpaths}.
+%
+%Finally, this approach allows the type system to track
+%programming idioms from 
+%dynamic languages
+%using implicit type-based reasoning based on the result of
+%conditional tests.
+%For instance,
+%\egref{example:parent-if}
+%only utilizes \clj{f} once
+%the programmer is convinced it is safe to do so based whether
+%\clj{f}
+%is
+%true or false. 
+%To express this in the type system, every expression 
+%is described by two propositions: a `then' proposition
+%for when it reduces to a true value, and an `else' proposition
+%when it reduces to a false value---for \clj{f}
+%the then proposition is {\notprop{\falsy}{f}} and 
+%the else proposition is {\isprop{\falsy}{f}}.
 %\ref{main:figure:typingrules}
@@
 
-We formalize our type system following~\citet{TF10}
-(with differences highlighted in $\trdiff{\text{blue}}$).
-The typing judgment 
-$$
-{\judgementrewrite   {\propenv}
-              {\e{}} {\t{}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
+%\figref{main:figure:typingrules} contains the core typing rules.
+%The key rule for reasoning about conditional control flow is
+%T-If. 
+%
+%\begin{mathpar}
+%  {\TIf}
+%\end{mathpar}
+
+%The propositions of the test expression \e{1}, \thenprop{\prop{1}} and \elseprop{\prop{1}}, are 
+%used as assumptions in the then and else branch respectively.
+
+The introduction rule for lambda abstractions also introduces 
+propositions to bring its parameter into scope when checking the body.
+The proposition set says lambda expressions cannot be false.
+$$
+{\TAbs}
+$$
+The object \x{} is in scope in the latent propositions \thenprop{\prop{}} and
+\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. These are substituted with the
+actual argument object with the application rule for functions.
+$$
+    {\TApp}
+$$
+
+The rule for let expressions links inferred information about
+\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.
+$$
+%\footnotesize
+\infer [T-Let]
+{ \judgementrewrite {\propenv{}} {\e{1}} {\s{}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
+             {\object{1}}
+             {\ep{1}}
+             \\\\
+             \propp{} = {\impprop {\notprop {\falsydiff{}} {\x{}}} {\thenprop {\prop{1}}}}
+             \\\\
+             \proppp{} = {\impprop {\isprop {\falsydiff{}} {\x{}}} {\elseprop {\prop{1}}}}
+  \\\\
+  \judgementrewrite
+       {\propenv{}, {\isprop {\s{}} {\x{}}},
+         {\propp{}},
+         {\proppp{}}}
+             {\e{}} {\t{}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
              {\object{}} 
-  {\ep{}}}
-$$
-says expression \e{} is of type \t{} in the 
-proposition environment $\propenv{}$, with 
-`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}, and rewrites to \ep{}. We write 
-{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
-  or
-{\judgementselfrewrite   {\propenv}
-              {\e{}} {\t{}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-  {\object{}}}
-for expressions that rewrite to themselves.
-
-The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
-application, abstractions, conditionals and let expressions.
-All binding forms introduce fresh variables.
-Values include booleans, \nil{}, class literals, keywords, 
-numbers,
-constants and closures. 
-Value environments map local bindings to values.
-
-Types include the top type, \emph{untagged} unions, functions, singleton types
-and class instances. 
-We abbreviate \Booleanlong{} as \Boolean{}, \Keywordlong{} as \Keyword{}
-and \NumberFull{} and \Number{}.
-The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
-We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
-{\ValueTrue} as \True and {\ValueFalse} as {\False}.
-Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
-may refer to the function argument.
-%Latent means they are relevant when the function is applied rather than evaluated.
-They are latent because they are instantiated with the
-actual object of the argument in applications before they are used in the proposition environment.
-
-\figref{main:figure:typingrules} contains the core typing rules.
-The key rule for reasoning about conditional control flow is
-T-If.
-
-\begin{mathpar}
-  {\TIf}
-\end{mathpar}
-
-The propositions of the test expression \e{1}, \thenprop{\prop{1}} and \elseprop{\prop{1}}, are 
-used as assumptions in the then and else branch respectively.
-If the result of the \ifliteral{} is a true value, then it either
-came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
-which implies \thenprop{\prop{3}} is true. 
-The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
-similarly.
-The T-Local rule connects the type system to the proof system over type propositions
-via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
-to derive a type for a variable.
-Using this rule, the type system can then appeal to L-Update to refine the type
-assigned to \x{}.
-
-We are now equipped to type check
-\egref{example:parent-if}:
-$$
-\clj{(if f (.getParent f) nil)}
-$$
-
-With {\propenv{}} = {\isprop{\UnionNilFile{}}{f}},
-$$
-\judgement{\propenv{}}{f}{\UnionNilFile{}}{\localfilterset{f}}{f}
-$$
-via T-Local.
-
-Checking the then branch involves extending
-the proposition environment with {\notprop{\falsy}{f}}
-$$
-\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\x{}}{\Number{}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
-$$
-because we can now satisfy the premise of T-Local:
-$$
-\inpropenv{{\propenv{}},\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}.
-$$
+             {\ep{}}
+             }
+{ \judgementrewrite {\propenv{}} {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}
+             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
+                          {\object{1}}
+                          {\x{}}}
+             {\replacefor {\object{}} 
+                          {\object{1}}
+                          {\x{}}}
+             {\letexp {\x{}} {\ep{1}} {\ep{}}}
+             }
+$$
+
+\paragraph{Other typing rules} The remaining core typing rules
+include a subsumption rule, and rules for other values that reflect
+\nil{} and \false{} as being the only false values.
+$$
+\footnotesize
+    %{\TDo}
+    %{\TClass}
+    %{\TIf}
+    %{\TAbs}
+    %\begin{array}{c}
+    %  {\TSubsume}\\\\
+    %  {\TNum}
+    %\end{array}
+    \begin{array}{c}
+      {\TNum}\\\\
+      {\TConst}\\\\
+      {\TKw}\\\\
+      {\TClass}\\\\
+      {\TTrue}\\\\
+    \end{array}
+    \begin{array}{c}
+      {\TSubsume}\\\\
+      {\TNil}\\\\
+      {\TFalse}
+    \end{array}
+$$
+
+%The T-Local rule connects the type system to the proof system over type propositions
+%via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
+%to derive a type for a variable.
+%Using this rule, the type system can then appeal to L-Update to refine the type
+%assigned to \x{}.
+%
+%We are now equipped to type check
+%\egref{example:parent-if}:
+%$$
+%\clj{(if f (.getParent f) nil)}
+%$$
+%
+%With {\propenv{}} = {\isprop{\UnionNilFile{}}{f}},
+%$$
+%\judgement{\propenv{}}{f}{\UnionNilFile{}}{\localfilterset{f}}{f}
+%$$
+%via T-Local.
+%
+%Checking the then branch involves extending
+%the proposition environment with {\notprop{\falsy}{f}}
+%$$
+%\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\x{}}{\Number{}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
+%$$
+%because we can now satisfy the premise of T-Local:
+%$$
+%\inpropenv{{\propenv{}},\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}.
+%$$
 %\judgement{{\propenv{}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
@@
 
-Subtyping (\figref{main:figure:subtyping}) 
-is a reflexive and transitive relation with top type \Top. 
+\paragraph{Subtyping}
+\figref{main:figure:subtyping} presents subtyping
+as a reflexive and transitive relation with top type \Top. 
 Singleton types are instances of their respective classes---boolean singleton types
@@
 
-\begin{figure*}
-  \footnotesize
-%   \judgbox{
-%{\judgementrewrite   {\propenv}
-%              {\e{}} {\t{}}
-%  {\filterset {\thenprop {\prop{}}}
-%              {\elseprop {\prop{}}}}
-%  {\object{}}{\ep{}}}}
-%           {Under proposition environment $\propenv{}$, 
-%             expression \e{} is of type \t{}
-%             with  \\
-%
-%`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-%and object \object{} and rewrites to \ep{}.}
-  \begin{mathpar}
-    %{\TDo}
-    %{\TClass}
-    %{\TIf}
-    {\TAbs}
-    \begin{array}{c}
-      {\TSubsume}\\\\
-      {\TNum}
-    \end{array}
-    \begin{array}{c}
-      {\TConst}\\\\
-      {\TKw}\\\\
-      {\TClass}
-    \end{array}
-    \begin{array}{c}
-      {\TTrue}\\\\
-      {\TFalse}\\\\
-      {\TNil}
-    \end{array}
-
-    {\TLet}
-    {\TLocal}
-
-    {\TApp}
-    %{\TError}
-
-  \end{mathpar}
-  \caption{Typing rules}
-  \label{main:figure:typingrules}
-\end{figure*}
+\paragraph{class} The \classconst{} function is implemented as a constant
+function (via T-Const), which returns a latent object which is the argument applied to the
+\classpe{} path element
+$$
+\constanttypefigure{}
+$$
+The dynamic semantics are given in \figref{main:figure:primitivesem}.
+%The definition of \updateliteral{} supports various idioms relating to \classpe{}
+%which we introduce in \secref{sec:isaformal}.
+
+%\begin{figure*}
+%  \footnotesize
+%%%   \judgbox{
+%%%{\judgementrewrite   {\propenv}
+%%%              {\e{}} {\t{}}
+%%%  {\filterset {\thenprop {\prop{}}}
+%%%              {\elseprop {\prop{}}}}
+%%%  {\object{}}{\ep{}}}}
+%%%           {Under proposition environment $\propenv{}$, 
+%%%             expression \e{} is of type \t{}
+%%%             with  \\
+%%%
+%%%`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+%%%and object \object{} and rewrites to \ep{}.}
+%  \begin{mathpar}
+%    %{\TDo}
+%    %{\TClass}
+%    %{\TIf}
+%    %{\TAbs}
+%    %\begin{array}{c}
+%    %  {\TSubsume}\\\\
+%    %  {\TNum}
+%    %\end{array}
+%    \begin{array}{c}
+%      {\TNum}\\\\
+%      {\TConst}\\\\
+%      {\TKw}\\\\
+%      {\TClass}\\\\
+%      {\TTrue}\\\\
+%    \end{array}
+%    \begin{array}{c}
+%      {\TSubsume}\\\\
+%      {\TNil}\\\\
+%      {\TFalse}
+%    \end{array}
+%
+%    %{\TLet}
+%    %{\TLocal}
+%
+%    %{\TApp}
+%    %{\TError}
+%
+%  \end{mathpar}
+%  \caption{Typing rules}
+%  \label{main:figure:typingrules}
+%\end{figure*}
 
@@
 
-\subsection{Precise Types for Heterogeneous maps}
-\label{sec:hmapformal}
-
-\begin{figure}
-  \footnotesize
-  $$
-  \begin{altgrammar}
-    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
-    &\mbox{Expressions} \\
-    \v{} &::=& \ldots \alt {\emptymap{}}
-    &\mbox{Values} \\
-    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
-    &\mbox{Types} \\
-    \auxhmapsyntax{}\\
-    \pesyntax{}   &::=& \ldots \alt {\keype{\k{}}}
-                  &\mbox{Path Elements}
-  \end{altgrammar}
-  $$
-  \begin{mathpar}
-    {\TGetHMap}
-
-    {\TGetAbsent}
-
-    {\TGetHMapPartialDefault}
-
-    {\TAssoc}
-  \end{mathpar}
-  \begin{mathpar}
-    \HMapsubtyping{}
-  \end{mathpar}
-  \begin{mathpar}
-    {\BAssoc}
-    {\BGet}
-    {\BGetMissing}
-  \end{mathpar}
-  \caption{HMap Syntax, Typing and Operational Semantics}
-  \label{main:figure:hmapsyntax}
-\end{figure}
-
-\figref{main:figure:hmapsyntax} presents syntax, typing rules
-and dynamic semantics in detail.
-%
-The type \HMapgeneric{\mandatory{}}{\absent{}}
-includes {\mandatory{}}, a map of \emph{present} entries (mapping keywords to types),
-\absent{}, a set of keyword keys that are known to be \emph{absent}
-and
-tag \completenessmeta{} which is either {\complete{}} (``complete'') if the map is fully specified by \mandatory{},
-and {\partial{}} (``partial'') if there are \emph{unknown} entries.
-To ease presentation, 
-if an HMap has completeness tag \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
-%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
-Keys cannot be both present and absent.
-
-The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
-we only model the former to avoid the added complexity of keywords being functions.
-To simplify presentation, we only provide syntax for the empty map literal and
-restrict lookup and extension to keyword keys. The metavariable \mapval{}
-ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
-usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
-
-Subtyping for HMaps
-designate \MapLiteral{} as a common supertype for all HMaps.
-S-HMap says that an HMap is a subtype of another HMap if they agree
-on \completenessmeta{}, agree on mandatory entries with subtyping
-and at least cover the absent keys of the supertype.
-Complete maps are subtypes of partial maps
-as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).
-
-The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
-that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
-and T-GetHMapPartialDefault a lookup with unknown results.
-Lookups on unions of HMaps are only supported in T-GetHMap, 
-in particular to support
-looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf})
-where every element in the union
-contains the key we are looking up.
-The objects in the T-Get rules are more complicated than those in T-Local---the 
-next section discusses this in detail.
-Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
-and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
-formed types.
-
-The semantics for \getliteral{} and \assocliteral{} are straightforward.
-If the entry is missing, B-GetMissing produces \nil{}.
-
-\subsection{Paths}
-\label{sec:formalpaths}
-
-Recall the first insight of occurrence typing---we can reason
-about specific \emph{parts} of the runtime environment
-using propositions.
-We refer to parts of the runtime environment via 
-a \emph{path} that consists of a series of
-\emph{path elements} applied right-to-left to a variable
-written \path{\pathelem{}}{\x{}}.
-\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
-to reason about selector operations on cons cells.
-We instead want to reason about HMap lookups and calls to \classconst{}.
-
-\paragraph{Key path element} We introduce our first path element
-{\keype{\k{}}}, which represents the operation of looking up
-a key \k{}.
-We directly relate this to our typing rule T-GetHMap
-(\figref{main:figure:hmapsyntax}) by
-checking the then branch of the first conditional test is checked in 
-an equivalent version of \egref{example:decleaf}.
-\begin{minted}{clojure}
-  (fn [m :- Expr]
-    (if (= (get m :op) :if)
-      {:op :if, ...}
-      (if ...)))
-\end{minted}
-
-We do not specifically support \equivliteral{} in our calculus, 
-but on keyword arguments it works identically to \clj{isa?} which we model
-in \secref{sec:isaformal}.
-Intuitively, if {\judgement{\propenv{}}{\e{}}{\t{}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
-then \equivapp{\e{}}{\makekw{if}} has the true and false propositions
-$$
-{\replacefor{\filterset{\isprop{\Value{\makekw{if}}}{\x{}}}{\notprop{\Value{\makekw{if}}}{\x{}}}}{\object{}}{\x{}}}
-$$
-where substitution reduces to \topprop{} if \object{} = \emptyobject{}.
-
-We start with proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
-Since {\Expr{}} is a union of HMaps, each with the entry \makekw{op}, we can use T-GetHMap.
-$$
-\judgement{\propenv{}}{\getexp{m}{\makekw{op}}}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\path{\keype{\makekw{op}}}{m}}
-$$
-Using our intuitive definition of \equivliteral{} above, we know
-$$
-\judgement{\propenv{}}{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}{\filterset{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}{\notprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}}{\emptyobject{}}
-$$
-Going down the then branch gives us the extended environment
-\propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.
-Using L-Update we can combine what we know about object $m$ and object
-{\path{\keype{\makekw{op}}}{m}}
-to derive
-$$
-\inpropenv{\propenvp{}}{\isprop{\HMapp{\mandatoryset{{\mandatoryentrykwnoarrow{op}{\makekw{if}}}, {\mandatoryentrykwnoarrow{test}{\Expr{}}},
-                                       {\mandatoryentrykwnoarrow{then}{\Expr{}}},   {\mandatoryentrykwnoarrow{else}{\Expr{}}}}}
-                                   {\emptyabsent{}}}{m}}
-$$
-
-The full definition of \updateliteral{} is given in \figref{main:figure:update}
-which considers both keys a path elements as well as the \classconst{}
-path element described below.
-In the absence of paths, update simply performs set-theoretic operations
-on types; see \figref{main:figure:restrictremove} for details.
-
-\paragraph{Class path element} Our second path element \classpe{} is used in the latent
-object of the constant \classconst{} function. Like Clojure's \clj{class}
-function \classconst{} returns the argument's class or \nil{}
-if passed \nil{}.
-$$
-\begin{array}{lrlr}
-  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
-                &\mbox{Path Elements}
-\end{array}
-$$
-\begin{mathpar}
-\constanttypefigure{}
-\end{mathpar}
-The dynamic semantics are given in \figref{main:figure:primitivesem}.
-The definition of \updateliteral{} supports various idioms relating to \classpe{}
-which we introduce in \secref{sec:isaformal}.
-
 \subsection{Java Interoperability}
@@
 We model the syntax after the `dot' special
-form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
-\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
-and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
-
-T-FieldStatic checks a resolved field expression by ensuring the target has
-the correct static type, then returns a nilable type corresponding the Java type.
-
-\begin{mathpar}
+form to prevent ambiguity---field accesses are written \fieldexp{\fld{}}{\e{}},
+and method calls are written $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$.
+%and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
+
+The rule for unresolved fields ensures the target has
+the correct static type, then returns a nilable type corresponding the Java type
+after rewriting to a non-reflective field access.
+$$
     {\TField}
-\end{mathpar}
-
-The rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
+$$
+
+The rules T-Method and T-New work similarly (\figref{main:figure:javatyping}), varying
 in the choice of nilability in the conversion function---methods
@@
 states our exact assumptions.
+There are no evaluation rules for reflective Java interoperability, since there are no typing
+rules that rewrite to reflective calls.
+
+%\subsection{Paths}
+%\label{sec:formalpaths}
+%
+%Recall the first insight of occurrence typing---we can reason
+%about specific \emph{parts} of the runtime environment
+%using propositions.
+%We refer to parts of the runtime environment via 
+%a \emph{path} that consists of a series of
+%\emph{path elements} applied right-to-left to a variable
+%written \path{\pathelem{}}{\x{}}.
+%\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
+%to reason about selector operations on cons cells.
+%We instead want to reason about HMap lookups and calls to \classconst{}.
+%
+%\paragraph{Key path element} We introduce our first path element
+%{\keype{\k{}}}, which represents the operation of looking up
+%a key \k{}.
+%We directly relate this to our typing rule T-GetHMap
+%(\figref{main:figure:hmapsyntax}) by
+%checking the then branch of the first conditional test is checked in 
+%an equivalent version of \egref{example:decleaf}.
+%\begin{minted}{clojure}
+%  (fn [m :- Expr]
+%    (if (= (get m :op) :if)
+%      {:op :if, ...}
+%      (if ...)))
+%\end{minted}
+%
+%We do not specifically support \equivliteral{} in our calculus, 
+%but on keyword arguments it works identically to \clj{isa?} which we model
+%in \secref{sec:isaformal}.
+%Intuitively, if {\judgement{\propenv{}}{\e{}}{\t{}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
+%then \equivapp{\e{}}{\makekw{if}} has the true and false propositions
+%$$
+%{\replacefor{\filterset{\isprop{\Value{\makekw{if}}}{\x{}}}{\notprop{\Value{\makekw{if}}}{\x{}}}}{\object{}}{\x{}}}
+%$$
+%where substitution reduces to \topprop{} if \object{} = \emptyobject{}.
+%
+%We start with proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
+%Since {\Expr{}} is a union of HMaps, each with the entry \makekw{op}, we can use T-GetHMap.
+%$$
+%\judgement{\propenv{}}{\getexp{m}{\makekw{op}}}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\path{\keype{\makekw{op}}}{m}}
+%$$
+%Using our intuitive definition of \equivliteral{} above, we know
+%$$
+%\judgement{\propenv{}}{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}{\filterset{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}{\notprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}}{\emptyobject{}}
+%$$
+%Going down the then branch gives us the extended environment
+%\propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.
+%Using L-Update we can combine what we know about object $m$ and object
+%{\path{\keype{\makekw{op}}}{m}}
+%to derive
+%$$
+%\inpropenv{\propenvp{}}{\isprop{\HMapp{\mandatoryset{{\mandatoryentrykwnoarrow{op}{\makekw{if}}}, {\mandatoryentrykwnoarrow{test}{\Expr{}}},
+%                                       {\mandatoryentrykwnoarrow{then}{\Expr{}}},   {\mandatoryentrykwnoarrow{else}{\Expr{}}}}}
+%                                   {\emptyabsent{}}}{m}}
+%$$
+%
+%The full definition of \updateliteral{} is given in \figref{main:figure:update}
+%which considers both keys a path elements as well as the \classconst{}
+%path element described below.
+%In the absence of paths, update simply performs set-theoretic operations
+%on types; see \figref{main:figure:restrictremove} for details.
+%
+%\paragraph{Class path element} Our second path element \classpe{} is used in the latent
+%object of the constant \classconst{} function. Like Clojure's \clj{class}
+%function \classconst{} returns the argument's class or \nil{}
+%if passed \nil{}.
+%$$
+%\begin{array}{lrlr}
+%  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
+%                &\mbox{Path Elements}
+%\end{array}
+%$$
+%\begin{mathpar}
+%\constanttypefigure{}
+%\end{mathpar}
+%The dynamic semantics are given in \figref{main:figure:primitivesem}.
+%The definition of \updateliteral{} supports various idioms relating to \classpe{}
+%which we introduce in \secref{sec:isaformal}.
 
@@
                 &\mbox{Values} \\
+ \disptablesyntax{} \\\\
   \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
-                &\mbox{Types} \\
-
- \disptablesyntax{} \\
+                &\mbox{Types}
 \end{altgrammar}
@@
 
+\subsection{Precise Types for Heterogeneous maps}
+\label{sec:hmapformal}
+
+\begin{figure}
+  \footnotesize
+  $$
+  \begin{altgrammar}
+    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
+    &\mbox{Expressions} \\
+    \v{} &::=& \ldots \alt {\emptymap{}}
+    &\mbox{Values} \\
+    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
+    &\mbox{Types} \\
+    \auxhmapsyntax{}\\
+%    \pesyntax{}   &::=& \ldots \alt {\keype{\k{}}}
+%                  &\mbox{Path Elements}
+  \end{altgrammar}
+  $$
+  \begin{mathpar}
+    {\TGetHMap}
+
+    {\TGetAbsent}
+
+    {\TGetHMapPartialDefault}
+
+    {\TAssoc}
+  \end{mathpar}
+  \begin{mathpar}
+    \HMapsubtyping{}
+  \end{mathpar}
+  \begin{mathpar}
+    {\BAssoc}
+    {\BGet}
+    {\BGetMissing}
+  \end{mathpar}
+  \caption{HMap Syntax, Typing and Operational Semantics}
+  \label{main:figure:hmapsyntax}
+\end{figure}
+
+\figref{main:figure:hmapsyntax} presents syntax, typing rules
+and dynamic semantics in detail.
+%
+The type \HMapgeneric{\mandatory{}}{\absent{}}
+includes {\mandatory{}}, a map of \emph{present} entries (mapping keywords to types),
+\absent{}, a set of keyword keys that are known to be \emph{absent}
+and
+tag \completenessmeta{} which is either {\complete{}} (``complete'') if the map is fully specified by \mandatory{},
+and {\partial{}} (``partial'') if there are \emph{unknown} entries.
+To ease presentation, 
+if an HMap has completeness tag \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
+%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
+Keys cannot be both present and absent.
+
+The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
+we only model the former to avoid the added complexity of keywords being functions.
+To simplify presentation, we only provide syntax for the empty map literal and
+restrict lookup and extension to keyword keys. The metavariable \mapval{}
+ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
+usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
+
+Subtyping for HMaps
+designate \MapLiteral{} as a common supertype for all HMaps.
+S-HMap says that an HMap is a subtype of another HMap if they agree
+on \completenessmeta{}, agree on mandatory entries with subtyping
+and at least cover the absent keys of the supertype.
+Complete maps are subtypes of partial maps
+as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).
+
+The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
+that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
+and T-GetHMapPartialDefault a lookup with unknown results.
+Lookups on unions of HMaps are only supported in T-GetHMap, 
+for cases like \egref{example:desserts-on-meal}
+where every element in the union
+contains the key we are looking up.
+The objects in the T-Get rules are more complicated than those in T-Local---the 
+next section discusses this in detail.
+Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
+and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
+formed types.
+
+The semantics for \getliteral{} and \assocliteral{} are straightforward.
+If the entry is missing, B-GetMissing produces \nil{}.
+
 \begin{figure*}

### 2015-07-07 formal-model.tex
rewrite middle of section 3

@@
 $$
+The logical system deduces \t{} as being a {\makekw{combo}} meal.
 
@@
 
-% TODO introduction rules for logical statements
 \paragraph{Extension to propositional logic}
@@
 numbers,
-constants and closures. 
-Value environments map local bindings to values.
+constants, and strings. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
+close over functions with value environments \openv{}, which map bindings to values.
 
@@
 and class instances.
-We abbreviate \Booleanlong{} as \Boolean{}, \Keywordlong{} as \Keyword{}
-and \NumberFull{} and \Number{}.
-The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
-We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
-{\ValueTrue} as \True and {\ValueFalse} as {\False}.
+We abbreviate 
+\Booleanlong{} to \Boolean{}, 
+\Keywordlong{} to \Keyword{},
+\NumberFull{}  to \Number{}, and 
+\StringFull{}  to \String{}.
+The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \makekw{a} is of type \Keyword{}.
+We abbreviate 
+\EmptyUnion{}     to \Bot{}, 
+{\ValueNil}       to \Nil{}, 
+{\ValueTrue}      to \True, and
+{\ValueFalse} to {\False}.
 Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
@@
                       \alt {\const{}}
+                      \alt {\str{}}
                       \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
@@
 \elseprop{\prop{}}, latent object \object{}, and latent type \t{}. These are substituted with the
-actual argument object with the application rule for functions.
+actual argument object---\objectp{}---with the application rule for functions.
 $$
@@
       {\TSubsume}\\\\
+      {\TStr}\\\\
       {\TNil}\\\\
@@
 semantics of \ifliteral{}:
-
 \begin{mathpar}
@@
   \begin{mathpar}
-    {\TNew}
-
-    {\TMethod}
+    {\TField}
+
+    %{\TMethod}
 
@@
 methods and constructors. 
-We model the syntax after the `dot' special
-form to prevent ambiguity---field accesses are written \fieldexp{\fld{}}{\e{}},
+To prevent ambiguity between zero-argument methods and fields, field accesses are written \fieldexp{\fld{}}{\e{}},
 and method calls are written $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$.
@@
 
-The rule for unresolved fields ensures the target has
-the correct static type, then returns a nilable type corresponding the Java type
-after rewriting to a non-reflective field access.
-$$
-    {\TField}
-$$
-
-The rules T-Method and T-New work similarly (\figref{main:figure:javatyping}), varying
-in the choice of nilability in the conversion function---methods
-can return nil but constructors cannot. Void also does not have a constructor.
+\egref{example:getparent-direct-constructor}, copied here,
+$$
+\clj{(.getParent (new File "a/b"))}
+$$
+is written
+\begin{equation}  \label{eq:unresolved}
+  \qquad {\methodexp {\getparent{}} {\newexp {\File{}} {\makestr{a/b}}} {}}
+\end{equation}
+
+But both the constructor and method are unresolved.
+We introduce \emph{non-reflective} expressions for specifying exact Java overloads.
+\begin{equation} \label{eq:resolved}
+\qquad {\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
+\end{equation}
+From the left, the one-argument constructor for \File takes a \String, and the 
+\getparent{} method of
+\File{} takes zero arguments
+and
+returns a \String.
+
+We now walk through the conversion from unresolved expression~\ref{eq:unresolved} to 
+resolved expression~\ref{eq:resolved}.
+
+The T-New typing rule checks and rewrites constructors.
+$$
+    {\TNew}
+$$
+To check
+{\newexp {\File{}} {\makestr{a/b}}}
+we first resolve the constructor overload in the class table---there is at most one
+to simplify presentation.
+With \classhint{1} = \String,
+we convert to a nilable type the argument with \t{1} = \Union{\Nil}{\String}
+and type check {\makestr{a/b}} against \t{1}.
+Typed Clojure defaults to allowing non-nilable arguments, but this
+can be overridden, so we model the more general case.
+% which erases nil
+The return Java type \File is converted to a non-nil
+Typed Clojure type \t{} = \File for the return type,
+and the proposition set says constructors can never be false---constructors
+can never produce the internal boolean value that Clojure uses for \false{}, or \nil{}.
+Finally, the constructor rewrites to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}.
+
+The T-Method rule checks unresolved methods.
+$$
+    {\TMethod}
+$$
+To check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}},
+we verify the target type \s{} = \File is non-nil before erasing \Nil
+by converting to a Java type \classhint{1} = \File.
+The specific overload is chosen from the class table based on \classhint{1}---there is at most one.
+Then {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
+is checked against a nilable conversion of \classhint{1}, \t{1} = \Union{\Nil}{\File},
+which succeeds by the previous rule.
+The nilable return type \t{} = \Union{\Nil}{\String} is given, and---finally---the 
+entire expression rewrites to expression \ref{eq:resolved}.
+%
+%We allow arguments to constructors and methods to be nilable, but not method
+%and field targets.
+
+The T-Field rule is included in \figref{main:figure:javatyping}, and is like T-Method, but without arguments.
 
@@
 \isaliteral{} is
-a subclassing test for classes, otherwise an equality test---we do not model
-the semantics for vectors.
+a subclassing test for classes, otherwise an equality test.
+%---we do not model the semantics for vectors
 
@@
   \begin{mathpar}
-    \TDefMulti{}
-
-    \TDefMethod{}
+    %\TDefMulti{}
+
+    %\TDefMethod{}
   \end{mathpar}
@@
 
-To ease presentation, we present \emph{immutable}
-multimethods, with syntax and semantics given in \figref{main:figure:mmsyntax}. 
-\defmethodliteral{} returns a new extended multimethod
-without changing the original multimethod. \egref{example:rep} is now written
+We use \emph{immutable}
+multimethods without default methods to ease presentation.
+%Syntax and semantics are given in \figref{main:figure:mmsyntax}. 
+%Multimethods can error if no matching method is chosen (rules in the supplemental material).
+Below is the translation of \egref{example:hi-multimethod} to \lambdatc{}.
 \begin{minted}{clojure}
-(let [path (defmulti [Any -> (U nil String)] class)]
-  (let [path (defmethod path String [x] x)]
-    (let [path (defmethod path File [^File x] 
-                 (.getPath x))]
-      (path (File. "dir/a"))))) ;=> "dir/a"
+(ann hi [Kw -> Str])
+(defmulti hi identity)
+(defmethod hi :en [_] "hello")
+(defmethod hi :fr [_] "bonjour")
+(hi :en) ;=> "hello"
 \end{minted}
-
-The type {\MultiFntype {\s{}} {\sp{}}} characterizes multimethods with \emph{interface type}
-{\s{}} and \emph{dispatch function type} {\sp{}}.
-The expression {\createmultiexp {\s{}} {\e{}}} defines a multimethod
-with interface type \s{} and dispatch function \e{}.
-The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
-extends multimethod \e{m} and to map
-dispatch value {\e{v}} to {\e{f}} in an extended dispatch table.
-The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
-with dispatch function {\v{}} and dispatch table {\disptable{}}.
-
-The T-DefMulti rule ensures that the type of the dispatch function
-has at least as permissive a parameter type
-as the interface type.
-%
-For example, we can check the definition from our translation above of \egref{example:rep}
-using T-DefMulti.
-$$
-\judgement{}%{\propenv{}}
-{\createmultiexp 
-      {\s{}}
-      {\classconst{}}}
-  {\MultiFntype {\s{}}{\sp{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
-$$
-where \s{}  = {\ArrowOne {\x{}} {\Top{}} {\t{}} {\filterset {\topprop{}} {\topprop{}}} {\emptyobject{}}}
-  and \sp{} = {\ArrowOne {\x{}} {\Top{}} {\Union{\Nil}{\Class}} {\filterset {\topprop{}} {\topprop{}}} {\path{\classpe{}}{\x{}}}}.
-  Since the parameter types agree, this is well-typed.
-
-The T-DefMethod rule is carefully constructed to ensure we have a syntactic
-lambda expression as the right-most subexpression.
-This way we can manually check the body of the lambda under an extended
-environment as sketched in \egref{example:incmap}.
-We use \isacompareliteral{} to compute the proposition for this method,
-since \isaliteral{} is used at runtime in multimethod dispatch.
-
-We continue with the next line of the translation of \egref{example:rep}.
-From the previous line we have \propenv{} = {\isprop{\MultiFntype {\s{}}{\sp{}}}{path}},
-so
-$$
-\judgement{\propenv{}}
-  {\extendmultiexp {prop} {\String}
-                   {\abs {\x{}} {\Top{}} {\x{}}}}
-  {\MultiFntype {\s{}}{\sp{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
-$$
-We know \emph{prop} is a multimethod by \propenv{}, so now we check the body
-of this method.
-$$
-\judgement{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
-  {\x{}}
-  {\String}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
-$$
-%This is checked by T-Local since {\inpropenv{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
-The new proposition {\isprop{\String}{\x{}}} is derived by 
-$$
-  \isacompare{\Top{}}{\path{\classpe{}}{\x{}}}{\Value{\File{}}}
-             {\filterset{\isprop{\String}{\x{}}}
-                        {\notprop{\String}{\x{}}}}.
-$$
-%
-The body of the \clj{let} is checked by T-App because
-{\MultiFntype {\s{}}{\sp{}}} is a subtype of its interface type {\s{}}.
-
-Multimethod definition semantics are straightforward.
+\begin{listing}
+$\letexp{hi_0} {\createmultiexp {\ArrowOne{\x{}}{\Keyword}{\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}} {\abs{\x{}}{\Keyword}{\x{}}}}
+  {\\\text{\quad}
+    \letexp{hi_1} {\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
+      {\\\text{\quad\quad}
+        \letexp{hi_2} {\extendmultiexp {hi_1} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}
+        {\\\text{\quad\quad\quad
+          \appexp{hi_2}{\makekw{en}}}}}}
+$
+\end{listing}
+
+For convenience, examples in this section are flattened when they are really nested
+let bindings. We also elide trivial latent propositions and objects.
+The following is an abbreviation of the previous expression.
+\begin{lstlisting}
+${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
+${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
+${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
+$\appexp{hi}{\makekw{en}}$
+\end{lstlisting}
+
+%\defmethodliteral{} returns a new extended multimethod
+%without changing the original multimethod. 
+%
+%\begin{minted}{clojure}
+%(let [hi (defmulti [Kw -> Str] identity)]
+%  (let [hi (defmethod hi :en [_] "hello")]
+%    (let [hi (defmethod hi :fr [_] "bonjour")]
+%      (hi :en))) ;=> "hello"
+%\end{minted}
+
+\paragraph{How to check}
+We now walk through type checking the expression
+\begin{equation}  \label{eq:hi-defmulti}
+{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}
+\end{equation}
+
+The expression {\createmultiexp {\s{}} {\e{}}} creates a multimethod with \emph{interface type} \s{}, and dispatch function \e{}
+of type \sp{},
+producing a value of type
+{\MultiFntype {\s{}} {\sp{}}}. % with interface type {\s{}} and dispatch function type {\sp{}}.
+The typing rule for \defmultiliteral{} simply checks the dispatch function, and
+verifies both the interface and dispatch type's domain agree.
+$$
+    \TDefMulti{}
+$$
+Expression~\ref{eq:hi-defmulti} checks with \t{} = \Keyword, interface type \s{} = {\ArrowTwo{\x{}}{\Keyword}{\String}},
+dispatch function type \sp{} = {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}, and overall type
+$$
+{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
+              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
+$$
+
+Next, we walk through checking
+\begin{equation}  \label{eq:hi-defmethoden}
+{\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
+\end{equation}
+
+The expression 
+{\extendmultiexp {\e{m}} {\e{v}} {\e{f}}} creates a new multimethod
+that extends multimethod \e{m}'s dispatch table, mapping dispatch value
+\e{v} to method \e{f}. The typing rule for \defmethodliteral{}
+checks \e{m} is a multimethod with dispatch function type \t{d},
+then calculates the extra information we know based on the current
+dispatch value {\thenprop{\proppp{}}}, which is assumed when checking the method
+body.
+$$
+    \TDefMethod{}
+$$
+Expression~\ref{eq:hi-defmethoden} checks with \e{m} being of type
+$$
+{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
+              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
+$$
+with \objectp{} = {\x{}} and \t{v} = \Value{\makekw{en}}. Then {\thenprop{\proppp{}}} = 
+{\isprop {\Value{\makekw{en}}}{\x{}}} by
+$$
+\isacompare{}{\x{}}{\Value{\makekw{en}}}
+{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
+$$
+Since \t{} = \Keyword{}, we check the method body like so
+$$
+\judgement{{\isprop{\Keyword}{\x{}}},{\isprop {\Value{\makekw{en}}}{\x{}}}}
+  {\makestr{hello}}
+  {\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
+$$
+Finally from the interface type \t{m}, we know \thenprop{\prop{}} = \elseprop{\prop{}} = \topprop{},
+and \object{} = \emptyobject{}, which also agrees with the method body, above.
+
+%The expression {\createmultiexp {\s{}} {\e{}}} 
+%defines a multimethod
+%with interface type \s{} and dispatch function \e{}.
+%The expression {\extendmultiexp {\e{m}} {\e{v}}{\e{f}}}
+%extends multimethod \e{m} and to map
+%dispatch value {\e{v}} to {\e{f}} in an extended dispatch table.
+%The value {\multi {\v{}} {\disptable{}}} is the runtime value of a multimethod
+%with dispatch function {\v{}} and dispatch table {\disptable{}}.
+%
+%The T-DefMulti rule ensures that the type of the dispatch function
+%has at least as permissive a parameter type
+%as the interface type.
+%
+%For example, we can check the definition from our translation above of \egref{example:rep}
+%using T-DefMulti.
+%$$
+%\judgement{}%{\propenv{}}
+%{\createmultiexp 
+%      {\s{}}
+%      {\classconst{}}}
+%  {\MultiFntype {\s{}}{\sp{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+%$$
+%where \s{}  = {\ArrowOne {\x{}} {\Top{}} {\t{}} {\filterset {\topprop{}} {\topprop{}}} {\emptyobject{}}}
+%  and \sp{} = {\ArrowOne {\x{}} {\Top{}} {\Union{\Nil}{\Class}} {\filterset {\topprop{}} {\topprop{}}} {\path{\classpe{}}{\x{}}}}.
+%  Since the parameter types agree, this is well-typed.
+%
+%The T-DefMethod rule requires a syntactic lambda expression as the method.
+%This way we can manually check the body of the lambda under an extended
+%environment as sketched in \egref{example:incmap}.
+%We use \isacompareliteral{} to compute the proposition for this method,
+%since \isaliteral{} is used at runtime in multimethod dispatch.
+%
+%We continue with the next line of the translation of \egref{example:rep}.
+%From the previous line we have \propenv{} = {\isprop{\MultiFntype {\s{}}{\sp{}}}{path}},
+%so
+%$$
+%\judgement{\propenv{}}
+%  {\extendmultiexp {prop} {\String}
+%                   {\abs {\x{}} {\Top{}} {\x{}}}}
+%  {\MultiFntype {\s{}}{\sp{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+%$$
+%We know \emph{prop} is a multimethod by \propenv{}, so now we check the body
+%of this method.
+%$$
+%\judgement{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}
+%  {\x{}}
+%  {\String}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
+%$$
+%%This is checked by T-Local since {\inpropenv{\propenv{},{\isprop{\Top}{\x{}}},{\isprop{\String}{\x{}}}}{\isprop{\String}{\x{}}}}.
+%The new proposition {\isprop{\String}{\x{}}} is derived by 
+%$$
+%  \isacompare{\Top{}}{\path{\classpe{}}{\x{}}}{\Value{\File{}}}
+%             {\filterset{\isprop{\String}{\x{}}}
+%                        {\notprop{\String}{\x{}}}}.
+%$$
+%%
+%The body of the \clj{let} is checked by T-App because
+%{\MultiFntype {\s{}}{\sp{}}} is a subtype of its interface type {\s{}}.
+
+\paragraph{Semantics}
+Multimethod definition semantics are straightforward, and are given 
+in \figref{main:figure:mmsyntax}. 
 B-DefMulti creates a multimethod with the given dispatch function and an empty dispatch table.
@@
 contains the key we are looking up.
-The objects in the T-Get rules are more complicated than those in T-Local---the 
-next section discusses this in detail.
+The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
+is a symbolic representation for a keyword lookup of $k$ in \object{}.
+The substitution for {\x{}} handles the case where \object{} is empty.
+$$
+\begin{array}{rcl}
+{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
+{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
+\end{array}
+$$
+
+%The objects in the T-Get rules are more complicated than those in T-Local---the 
+%next section discusses this in detail.
 Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
@@
 
+\subsection{Multimethods and Java}
+\subsection{Multimethods and HMaps}
+
 \begin{figure*}

### 2015-07-07 formal-model.tex
Rearrange some things in section 3

@@
 which is a possibly-empty sequence of \emph{path elements} applied right-to-left.
-Path elements are {\classpe{}} and {\keype{k}}, representing a call to \classconst{} and
+This paper uses two path elements---{\classpe{}} and {\keype{k}}---representing a call to \classconst{} and
 a keyword lookup on $k$ respectively.
@@
 \paragraph{`Then' and `else' propositions} Along with an object, each expression carries two
-propositions---a `then' proposition that can be assumed if the expression evaluates to a
-true value, and an `else' proposition for evaluating to a false value.
+propositions---a `then' and an `else' proposition,
+that can be assumed if the expression evaluates to a
+true value and a false value respectively.
 
@@
 This says if the expression $o$ evaluates to a true value, then $o$ is neither
-of type \False or \Nil. Naturally, the `else' proposition $\isprop{\falsy{}}{o}$
+of type \False or \Nil. Naturally, the `else' proposition $\isprop{\falsy{}}{o}$---which
+is true when $o$ is a false value---
 says $o$ is either \False or \Nil.
 
-Together, both propositions are called a \emph{proposition set}.
+The pair of propositions are called a \emph{proposition set}, written {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}.
 
@@
 
-The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
+\paragraph{Expressions} The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
 application, abstractions, conditionals, and let expressions.
@@
 
-Types include the top type, \emph{untagged} unions, functions, singleton types
+\paragraph{Types} Types include the top type, \emph{untagged} unions, functions, singleton types
 and class instances.
-We abbreviate 
+We abbreviate the classes
 \Booleanlong{} to \Boolean{}, 
 \Keywordlong{} to \Keyword{},
-\NumberFull{}  to \Number{}, and 
-\StringFull{}  to \String{}.
-The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \makekw{a} is of type \Keyword{}.
-We abbreviate 
+\NumberFull{}  to \Number{},
+\StringFull{}  to \String{}, and 
+\FileFull{}  to \File{}.
+We also abbreviate the types
 \EmptyUnion{}     to \Bot{}, 
@@
 {\ValueFalse} to {\False}.
+
+The difference between the types
+\Value{\Keyword} and \Keyword{} is subtle.
+The former is inhabited by the class literal \Keyword{} and the result of 
+\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
+
 Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
@@
 
+\paragraph{Constructors} First we check and convert {\newexp {\File{}} {\makestr{a/b}}} to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
 The T-New typing rule checks and rewrites constructors.
@@
 
+\paragraph{Methods} Next we convert {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
+to
+{\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
 The T-Method rule checks unresolved methods.
@@
     \BDefMulti{}
-    \BBetaMulti{}
+    %\BBetaMulti{}
   \end{mathpar}
@@
 
+\paragraph{Subtyping}
+Multimethods are also functions, which is encoded via the following subtyping rule.
+$$
+\SPMultiFn{}
+$$
+This rule says a multimethod can be upcast to its interface type. This means
+multimethod call sites can be handled by T-App via T-Subsume. Other rules are given in
+in \figref{main:figure:mmsyntax}.
+
 \paragraph{Semantics}
-Multimethod definition semantics are straightforward, and are given 
+Multimethod definition semantics are also given 
 in \figref{main:figure:mmsyntax}. 
@@
 B-DefMethod produces a new multimethod with an extended dispatch table.
-B-BetaMulti invokes the dispatch function with the evaluated argument to obtain the dispatch value,
-and uses \getmethodliteral{} (which models Clojure's \clj{get-method}) 
-to extract the appropriate method. The call to \getmethodliteral{} only returns a value
-if there is \emph{exactly one} method such that the corresponding dispatch value
-is compatible, using \isaopsemliteral{}, with the result of the dispatch function.
-Finally we return the result of applying the extracted method and the original argument.
+
+The overall dispatch mechanism is summarised by B-BetaMulti.
+$$
+    \BBetaMulti{}
+$$
+First the dispatch function \v{d} is applied to the argument \vp{} to obtain
+the dispatch value \v{e}.
+Based on \v{e},
+the \getmethodliteral{} metafunction (\figref{main:figure:mmsyntax})
+extracts a method \v{f} from the method table {\disptable{}}
+and applies it to the original argument for the final result.
 

### 2015-07-08 formal-model.tex
Add feeds2imap experiment

@@
 
-For example, the `then' proposition for the expression $o$ is $\notprop{\falsy{}}{o}$.
+The `then' proposition {\thenprop{\prop{}}} for the variable $o$ is $\notprop{\falsy{}}{o}$, for example.
 This says if the expression $o$ evaluates to a true value, then $o$ is neither
-of type \False or \Nil. Naturally, the `else' proposition $\isprop{\falsy{}}{o}$---which
-is true when $o$ is a false value---
-says $o$ is either \False or \Nil.
-
-The pair of propositions are called a \emph{proposition set}, written {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}.
+of type \False or \Nil. Naturally, the `else' proposition {\elseprop{\prop{}}} is $\isprop{\falsy{}}{o}$---which
+is true when $o$ is a false value---says $o$ is either \False or \Nil.
+
+The pair of propositions {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} is called a \emph{proposition set}.
 
@@
 methods and constructors. 
-To prevent ambiguity between zero-argument methods and fields, field accesses are written \fieldexp{\fld{}}{\e{}},
-and method calls are written $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$.
+To prevent ambiguity, % between zero-argument methods and fields, 
+field accesses are written \fieldexp{\fld{}}{\e{}}
+and method calls $\methodexp{\mth{}}{\e{}}{\overrightarrow{e}}$.
 %and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
 
-\egref{example:getparent-direct-constructor}, copied here,
+\egref{example:getparent-direct-constructor} translates like so.
 $$
@@
 $$
-is written
 \begin{equation}  \label{eq:unresolved}

### 2015-07-08 formal-model.tex
edits to sections 3 and 5

@@
 
-\egref{example:getparent-direct-constructor} translates like so.
+\egref{example:getparent-direct-constructor} translates to expression~\ref{eq:unresolved}.
 $$
@@
 $$
-Since \t{} = \Keyword{}, we check the method body like so
+Since \t{} = \Keyword{}, we check the method body with
 $$
@@
 
-\subsection{Multimethods and Java}
-\subsection{Multimethods and HMaps}
+\subsection{Proof system}
+
+% checks obvious prop logic, but update is most interesting rule.
 

### 2015-07-09 overview.tex
rewrite middle of section 3
Add feeds2imap experiment
Edits to 2.

@@
 Here is a simple program in Typed Clojure.
-
-We define a global variable \clj{greet} as a one-argument function
+%
+We define \clj{greet} as a one-argument function
 taking and returning a string.
@@
 \begin{minted}{clj}
-(greet 
-  nil ; Type Error:
+(greet nil) ; Type Error:
             ;   Expected Str, given nil
-  )
 \end{minted}
@@
 
-\paragraph{Flow analysis} Typed Clojure uses occurrence typing to
+\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{thf10} to
 model type-based control flow.
@@
 
+\begin{exmp}
 \begin{minted}{clj}
@@
 \end{minted}
+\label{example:getparent-direct-constructor}
+\end{exmp}
 
@@
 \begin{minted}{clj}
-  (fn [f] 
-    (.getParent f) ; Type Error:
-                   ;   Unresolved interop: getParent
-    )
+  (fn [f]           ; Type Error:                    
+    (.getParent f)) ;   Unresolved interop: getParent
+
 \end{minted}
@@
 Function arguments default to \clj{Any}, the most permissive type. Ascribing
-a parameter type helps Typed Clojure find a specific method.
+a parameter type allows Typed Clojure to find a specific method.
 
@@
 it is a static type error, as typed code could possibly dereference \clj{nil}.
-
 \begin{minted}{clj}
 (defn parent-bad-in [f :- (U nil File)]
-  (.getParent f) ; Type Error:
+  (.getParent f)) ; Type Error:
                   ;   Cannot call instance method 
@@
                   ;   (U nil File).
-  )
 \end{minted}
+Since Java-level types do not provide information about nullability,
+by default Typed Clojure conservatively assumes that Java method and
+constructor arguments, such as to @clj{.getParent}
+ must be \emph{non-nullable}, but this can be
+configured for  particular calls as needed.
 
@@
 (defn parent-bad-out [f :- File] :- Str
-  (.getParent f) ; Type Error:
+  (.getParent f)) ; Type Error:
                   ;   Expected Str, given (U nil Str).
-  )
 \end{minted}
@@
 
-By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
-but can be configured globally for particular non-target positions if needed.
-
 \subsection{Multimethods}
@@
 
-A \emph{multimethod} is a kind of extensible function. It combines
-a \emph{dispatch function} with one or more \emph{methods}.
+\emph{Multimethods} are a kind of extensible function, and they are
+widely used to define Clojure operations.  It combines a
+\emph{dispatch function} with one or more \emph{methods}.
 
 \paragraph{Value-based dispatch}
-Here is a simple multimethod.
+This simple multimethod says hello in different languages, as
+specified by a keyword argument.
 
+\begin{exmp}
 \begin{minted}{clojure}
@@
 \end{minted}
+\label{example:hi-multimethod}
+\end{exmp}
 
@@
 
-This is only part of the story for multimethod dispatch.
+Dispatching based on literal values enables certain forms of method
+definition, but this is only part of the story for multimethod dispatch.
 
 \paragraph{Class-based dispatch}
-For class values, multimethods choose methods based on subclassing
+For class values, multimethods can choose methods based on subclassing
 relationships.
 
-Recall the multimethod in figure~\ref{fig:ex1}, duplicated below.
-
+Recall the multimethod from figure~\ref{fig:ex1}, reproduced here.
 \begin{minted}{clj}
@@
 %respectively.
-
+\noindent
 The dispatch function \clj{class}
@@
 %---both installed via \clj{defmethod}
-
-The multimethod dispatch rules involve
-\clj{isa?}, a hybrid predicate which includes a subclassing check for classes and
-an equality check for most other values.
+%
+The multimethod dispatch rules use
+\clj{isa?}, a hybrid predicate which is a subclassing check for classes and
+an equality check for other values.
 
@@
 
-To demonstrate,
+In our example,
   \clj{(pname "STAINS/JELLY")}
-chooses the \clj{String} method because
+chooses the  method for \clj{String} because
 \clj{(isa? (class "STAINS/JELLY") String)}
@@
 \clj{(isa? (class "STAINS/JELLY") File)}
-is false. Then the \clj{String} method body
+is false. The \clj{String} method body
 \clj{(pname (new File "STAINS/JELLY"))}
@@
 
-Immutable hash-maps with keyword keys play a major role in Clojure programming
-as the primary way to structure data.
+Beyond primitives and Java objects, the most common Clojure data
+structure is the 
+immutable hash-map, typicially with keyword keys. This structure is
+the primary way to represent compound data in Clojure programs.
 
+Hash-maps are accessed with the \clj{get} function:
 \begin{minted}{clojure}
@@
 ;=> "waffles"
-; keywords are functions that look themselves up
+\end{minted}
+Additionally, keywords are functions that look themselves up in a
+map. Missing keys produce \clj{nil}.
+\begin{minted}{clojure}
 (:fr breakfast)        ; :- Str
@@
 
-HMap types model the most common usages of keyword maps.
+In Typed Clojure, \emph{HMap types} describe the most common usages of
+keyword-keyed maps.
 \begin{minted}{clojure}
@@
 \end{minted}
-The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
-and \clj{:fr} have string values (expressed as a \clj{:mandatory} entry) and that 
-every other key is absent (by \clj{:complete?} being \clj{true}).
+The inferred type for \clj{breakfast} holds two kinds of information---the known entries--\clj{:en}
+and \clj{:fr}---and their types, which are  \clj{:mandatory}, and that 
+no other key is present, since \clj{:complete?} is \clj{true}.
 
-HMap's default to being partially specified---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
+HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
 \clj{'{:en Str :fr Str}}
-forgets 
-information about absent keys, only providing \clj{:mandatory}
-information.
+omits
+information about absent keys, only providing information on
+\clj{:mandatory} keys.
 
@@
 a company with a large production Typed Clojure system
-(section~\ref{sec:casestudy} presents a case study).
+(section~\ref{sec:casestudy} presents a case study and empirical
+result from this code base).
 
+\begin{exmp}
 \begin{minted}{clojure}
@@
 \end{minted}
-%\begin{exmp}
+\label{example:circleci}
+\end{exmp}
+
 %\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
-%\label{example:circleci}
-%\end{exmp}
-Fully specified HMap's
+Using ully specified HMaps
 statically enforce the private key is not accidentally left in a supposedly
@@
   [{pk :private-key :as kp}]
-  (assoc kp :enc-private-key (encrypt pk))
+  (assoc kp :enc-private-key (encrypt pk)))
   ; Type Error:
@@
   ;         :complete? true)
-  )
 \end{minted}
-
-The extra \clj{:private-key} entry disagrees with \clj{EncKeyPair}, so a type error
-is given.
+The extra \clj{:private-key} entry does not match \clj{EncKeyPair}, so a type error
+is raised.
 
@@
 linked.
-There are infinite ways of both structuring and dispatching on data, so there is no
-hope predicting them all, or even manually covering useful a subset.
+There are infinite ways of both structuring and dispatching on data,
+so we cannot hope to merely add a set of special case rules for
+handling these features. Instead, as type system designers, we must
+search for a compositional approach.
 
-Thankfully, occurrence typing is both extensible \emph{and} compositional.
-By extending occurrence typing with
+Thankfully, occurrence typing, originally designed for reasoning about
+\clj{if} tests, provide the compositional approach we need.
+By extending the system with
 a handful of rules based on HMaps and other functions, 
-we can automatically cover almost all the common cases---as well others
+we can automatically cover both easy cases and those
 that compose simple rules in arbitrary ways.
 
-Futhermore, no special attention is needed for multimethod dispatch---the primitive branching
-mechanism is still the humble \clj{if} conditional. Only a small number of rules are needed
-to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
-In practice, this means all conditional-based control flow analysis
-also extends to the context of multimethod dispatch, and vice-versa.
+Futhermore, this approach extends to multimethod dispatch---the primitive branching
+mechanism works like the humble \clj{if} conditional. 
+Only a small number of rules are needed
+to encode the \clj{isa?}-based dispatch, themselves made of simple pieces.
+In practice, this means that conditional-based control flow typing
+ extends to  multimethod dispatch, and vice-versa.
 
@@
 The \clj{:combo} method is verified to only structurally recur
-on \clj{Order}s. This is achieved because we learn the argument--- \clj{o} ---must % em dash needs space here, looks weird
+on \clj{Order}s. This is achieved because we learn the argument---\clj{o}---must % em dash needs space here, looks weird
 be of type
@@
 \end{minted}
-which can safely structurally recur.
+which is an appropriate argument for the recursive call.
 
@@
 must be true.
-
+%
 Knowing \clj{o} is also an
 \clj{Order}, we can eliminate the possibility of
-a \clj{:combo} meal, and infer the type
+a \clj{:combo} meal, and deduce the type:
 \begin{minted}{clojure}
@@
 
-In the \clj{nil} method, we know
+In the \clj{nil} case, we know
 \begin{minted}{clojure}
@@
 \end{minted}
-This intersected with \clj{o} being an \clj{Order} simplifies to a \clj{:combo} meal, 
+Since \clj{o} is originally an \clj{Order}, this simplifies to a \clj{:combo} meal, 
 \begin{minted}{clojure}
@@
 \end{minted}
-thus verifying both structural recursions.
+thus allowing both recursive calls to type check.
 
-\paragraph{From multiple to arbitrary dispatch}
-Multimethod dispatch goes even further, supporting
-multiple dispatch via vectors.
-
-Fully specifying multimethod dispatch is beyond the scope of this paper,
+\paragraph{From multimethod to multiple dispatch}
+Clojure multimethod dispatch, and Typed Clojure's handling of it, goes
+even further, supporting dispatch on multiple arguments via vectors.
+%
+Dispatch on multiple arguments is beyond the scope of this paper,
 but the same intuition applies---adding support for multiple dispatch

### 2015-07-10 metatheory.tex
modified some of proof
update metatheory
edits to 4

@@
 
-We prove type soundness follow using the same technique as~\citet{TF10}. 
-We also include errors and a \wrong{} value and prove
-well-typed programs do not go wrong.
+We prove type soundness following \citet{TF10}.  Our model is extended
+to include errors and a \wrong{} value, and we prove well-typed
+programs do not go wrong; this is therefore a stronger theorem than
+proved by \citet{TF10}.
 
-Rather than modeling Java's dynamic semantics, we instead
-make our assumptions about Java explicit. We concede that
-method and constructor calls may diverge or error, but we assume they can
-never go wrong. (Assumptions for other operations are given in the supplemental
-material).
+Rather than modeling Java's dynamic semantics, a task of daunting
+complexity, we instead make our assumptions about Java explicit. We
+concede that method and constructor calls may diverge or error, but
+assume they can never go wrong. (Assumptions for other operations are
+given in the supplemental material).
 
@@
 For the purposes of our soundness proof, we require that all values
-are \emph{consistent}.
-Consistency
-ensures that occurrence typing does not refer to variables
-hidden inside a closure.
+are \emph{consistent}.  Consistency (defined in the supplemental
+material) states that the types of closures are well-scoped---they do
+not claim propositions about variables hidden in their closures.
 
-{\consistentwithonlydef{main}}
+%{\consistentwithonlydef{main}}
 
@@
   \begin{proof}
-    By induction on the derivation of the typing judgment. 
+    By induction on the derivation of \opsem {\openv{}} {\e{}} {\a{}}.
     (Full proof given as lemma~\ref{appendix:lemma:soundness}).
@@
 
-We can now state our soundness theorem.
+We can now state our soundness theorems.
 

### 2015-07-10 overview.tex
Fixes to section 2

@@
 by default Typed Clojure conservatively assumes that Java method and
-constructor arguments, such as to @clj{.getParent}
+constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
+\egref{example:getparent-direct-constructor},
  must be \emph{non-nullable}, but this can be
 configured for  particular calls as needed.
+The target of method invocations is always non-nullable.
 
@@
 %\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
-Using ully specified HMaps
+Using fully specified HMaps
 statically enforce the private key is not accidentally left in a supposedly
@@
 Thankfully, occurrence typing, originally designed for reasoning about
-\clj{if} tests, provide the compositional approach we need.
+\clj{if} tests, provides the compositional approach we need.
 By extending the system with
@@
 \end{minted}
-which is an appropriate argument for the recursive call.
+which contains appropriate arguments for both recursive calls.
 

### 2015-07-10 formal-model.tex
move typing rules to figures

@@
 Standard type systems use type environments to track lexical type information.
-Introducing a lexically scoped variable $x$ is a matter of associating 
+Introducing a lexically scoped variable $x$ is a matter of
 associating $x$ with its type $\Number$ (a number) in the type environment.
@@
 The logical statement {\isprop{\Number{}}{\x{}}} says the value
-obtained via $x$ is of type \Number{}. We now must \emph{prove}
-a variable's type via a \emph{logical system}.
+obtained via $x$ is of type \Number{}. A \emph{logical system}
+must now \emph{prove}
+a variable's type.
 $$
@@
 The proof system judgment { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
-says logical assumptions {\propenv{}} prove variable \x{} is of type \t{}.
-
-\paragraph{Symbolic representations}
+says under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
+%We later define the more general T-Local.
+
+\paragraph{Symbolic representation}
 As we saw in \secref{sec:overview},
@@
 with respect to the current lexical environment. For example,
-
 \begin{itemize}
-  \item $x$ is represented as $x$; 
-  \item $\appexpone{\ccclass{\appexp{\makekw{lunch}}{m}}}$ is represented as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{m}}}$;
+  \item variable \makelocal{m} is represented as \makelocal{m}; 
+  \item $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ is represented as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$;
   \item $42$ has no representation, called \emptyobject{}.
 \end{itemize}
-
 This symbolic representation is called the expression's \emph{object}. 
@@
 \emptyobject{} is the
-\emph{empty object}. Non-empty objects \path{\pathelem{}}{\x{}} consist of a symbol \x{} and a \emph{path} \pathelem{},
-which is a possibly-empty sequence of \emph{path elements} applied right-to-left.
-This paper uses two path elements---{\classpe{}} and {\keype{k}}---representing a call to \classconst{} and
-a keyword lookup on $k$ respectively.
+\emph{empty object}. Non-empty objects \path{\pathelem{}}{\x{}} consist of a root variable \x{} and a \emph{path} \pathelem{},
+a possibly-empty sequence of \emph{path elements} applied right-to-left starting from the root variable.
+We use two path elements---{\classpe{}} and {\keype{k}}---representing the result
+of a calling \classconst{} and the result of looking up a keyword $k$ respectively.
 
@@
 \paragraph{Logical statements about objects}
-So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}},
+So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}}---
 saying variable \x{} is of type \t{}.
@@
 To illustrate, recall \egref{example:desserts-on-meal}.
-We know the parameter $o$ is of type $\Order$
+We know the parameter \makelocal{o} is of type $\Order$
 by the annotation on $desserts$---written
 \begin{equation}
-\qquad {\isprop{\Order}{o}}
+\qquad {\isprop{\Order}{\makelocal{o}}}
 \end{equation}
@@
 In checking the ${\makekw{combo}}$ method, we also assume
-${\appexp{\makekw{Meal}}{o}}$ is ${\makekw{combo}}$,
+${\appexp{\makekw{Meal}}{\makelocal{o}}}$ is ${\makekw{combo}}$,
 based on multimethod dispatch rules. This is written
 \begin{equation}
-  \qquad {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{o}}}
+  \qquad {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}
 \end{equation}
-and pronounced ``the ${\makekw{Meal}}$ path of variable $o$ is of type
+and pronounced ``the ${\makekw{Meal}}$ path of variable \makelocal{o} is of type
 {\Value{\makekw{combo}}}'', a keyword singleton type.
 
-\paragraph{Logical system in action} At some point in checking the {\makekw{combo}} method,
-we will need the type of $o$.
-Our proposition environment consists of both above facts
-$$
-\propenv{} = {{\isprop{\Order}{o}},
-    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{o}}}}
-$$
-and T-LocalProve says $o$ is of type \t{} if \propenv{} \emph{proves} $o$ is of type \t{}
-$$
+\paragraph{Logical system in action} 
+To attain the type of \makelocal{o}, 
+T-LocalProve says we must solve for \t{} in
+$
 { \inpropenv 
   {\propenv{}}
-  {\isprop {\t{}} {o}}}
-$$
-The logical system deduces \t{} as being a {\makekw{combo}} meal.
-
+  {\isprop {\t{}} {\makelocal{o}}}}
+$,
+under proposition environment 
+$$
+\propenv{} = {{\isprop{\Order}{\makelocal{o}}},
+    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
+$$
+
+The main 
+$$
+\LUpdatePos
+$$
 Now the purpose of a logical representation can be conveyed---we rely on the logical
@@
 
-\paragraph{Extension to propositional logic}
-The kinds of logical statements are extended to \emph{propositional logic}.
+\paragraph{Extended to propositional logic}
+We extend our logical statements to \emph{propositional logic}.
 \figref{main:figure:termsyntax} describes the syntax we will use for the rest of the paper.
@@
 
-The pair of propositions {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} is called a \emph{proposition set}.
+%The pair of propositions {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} is called a \emph{proposition set}.
 
@@
 
-\paragraph{Elimination rule for propositions}
-We can now review the main elimination rule for propositions, T-Local.
-$$
-    {\TLocal}
-$$
-It says, if proposition environment \propenv{} proves variable \x{} is 
-of type \t{}, then under proposition environment \propenv{}
-variable \x{} is of type \t{}. Also, if \x{} is a true value, then
-we know \x{} is neither \False or \Nil, and if \x{} is a false value,
-then \x{} is either \False or \Nil. Finally, the symbolic representation
-for \x{}---its object---is \x{}.
-
-\paragraph{Introduction rules for propositions}
-Every program branch is an opportunity to extend the proposition environment
-since we learn whether a test expression evaluates true or false. The T-If
-rule assumes the test's `then' proposition \thenprop{\prop{1}}
-in the then branch, and the test's `else' proposition \elseprop{\prop{1}}
-in the else branch.
-\begin{mathpar}
-  {\TIf}
-\end{mathpar}
-The `then' proposition is {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}}
-because if the entire expression reduces to a true value,
-then either the `then' branch is true---implying {\thenprop {\prop{2}}}---or
-the `else' branch is true---implying {\thenprop {\prop{3}}}.
+\paragraph{Typing rules}
+The core type rules are
+given as \figref{main:figure:othertypingrules}. We introduce
+the interesting rules while demonstrating how to check
+the complement number predicate
+\begin{equation}
+\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+\end{equation}
+%, including
+%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
+%as impossible (\botprop{}) `then' propositions.
+
+The lambda rule T-Abs introduces \isprop{\Top}{\makelocal{d}}
+to check the body.
+With \propenv{} = \isprop{\Top}{\makelocal{d}},
+T-If then checks the test {\appexp{\numberhuh{}}{\makelocal{d}}}
+with T-App with three main steps.
+
+First, the operator \numberhuh{} is checked with T-Const, which
+uses 
+the \constanttypeliteral{} relation to give types to constants.
+\numberhuh{} is a predicate over numbers, and
+\classconst{} returns its argument's class.
+$$
+\constanttypefigure{}
+$$
+The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.
+
+Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
+the operand \makelocal{d} is checked with
+T-Local as
+$$
+\judgementselfrewrite{\propenv{}}
+                     {\makelocal{d}}
+                     {\UnionNilNum}
+                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
+                                {\isprop{\falsy}{\makelocal{d}}}}
+                     {\makelocal{d}}
+$$
+which encodes the type, proposition, and object information
+previously discussed about variables.
+
+Finally, T-App substitutes the operand's object for the parameter
+---here \makelocal{d} for \numberhuh{}'s \x{}---yielding
+$$
+\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
+  {\Boolean{}}
+  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+             {\notprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+$$
+
+To check {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
+T-If then
+introduces \isprop{\Number{}}{\makelocal{d}}
+in the `then' branch and \notprop{\Number{}}{\makelocal{d}}
+in the `else' branch.
+%The object \x{} over latent propositions \thenprop{\prop{}} and
+%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
+%actual argument object---\objectp{}---is substituted in at application by T-App.
+
+The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
+The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
+\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
+  {\Boolean{}}
+  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+             {\notprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+
+The T-If refines each branch based on the test---a
+{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
+for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
+for checking the `else' branch.
+Information from each branch is combined via subsumption for the overall type, propositions, and object.
+
+For example
+$
+\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
+  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+$
+type checks because T-Subsume allows us to check both branches as
+$
+\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
+  {\false{}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+$
+and
+$
+\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
+  {\true{}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+$
+respectively.
+%
+%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
+%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
+%we use T-Subsume on both branches
+%to introduce a logical disjunction
+%since at least one must be true if the entire expression is true.
 
@@
 
-%\figref{main:figure:typingrules} contains the core typing rules.
-%The key rule for reasoning about conditional control flow is
-%T-If. 
-%
-%\begin{mathpar}
-%  {\TIf}
-%\end{mathpar}
-
-%The propositions of the test expression \e{1}, \thenprop{\prop{1}} and \elseprop{\prop{1}}, are 
-%used as assumptions in the then and else branch respectively.
-
-The introduction rule for lambda abstractions also introduces 
-propositions to bring its parameter into scope when checking the body.
-The proposition set says lambda expressions cannot be false.
-$$
-{\TAbs}
-$$
-The object \x{} is in scope in the latent propositions \thenprop{\prop{}} and
-\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. These are substituted with the
-actual argument object---\objectp{}---with the application rule for functions.
-$$
-    {\TApp}
-$$
-
-The rule for let expressions links inferred information about
-\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.
-$$
-%\footnotesize
-\infer [T-Let]
-{ \judgementrewrite {\propenv{}} {\e{1}} {\s{}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
-             {\object{1}}
-             {\ep{1}}
-             \\\\
-             \propp{} = {\impprop {\notprop {\falsydiff{}} {\x{}}} {\thenprop {\prop{1}}}}
-             \\\\
-             \proppp{} = {\impprop {\isprop {\falsydiff{}} {\x{}}} {\elseprop {\prop{1}}}}
-  \\\\
-  \judgementrewrite
-       {\propenv{}, {\isprop {\s{}} {\x{}}},
-         {\propp{}},
-         {\proppp{}}}
-             {\e{}} {\t{}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-             {\object{}} 
-             {\ep{}}
-             }
-{ \judgementrewrite {\propenv{}} {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}
-             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
-                          {\object{1}}
-                          {\x{}}}
-             {\replacefor {\object{}} 
-                          {\object{1}}
-                          {\x{}}}
-             {\letexp {\x{}} {\ep{1}} {\ep{}}}
-             }
-$$
-
-\paragraph{Other typing rules} The remaining core typing rules
-include a subsumption rule, and rules for other values that reflect
-\nil{} and \false{} as being the only false values.
+\begin{figure*}
 $$
@@
   \begin{array}{c}
-      {\TNum}\\\\
-      {\TConst}\\\\
-      {\TKw}\\\\
-      {\TClass}\\\\
-      {\TTrue}\\\\
-    \end{array}
+    {\TLocal} \ \ \  \ \ \ 
+    {\TIf}\ \ \  \ \ \ \ \ 
+    {\TAbs} \ \ \ \ 
     \begin{array}{c}
-      {\TSubsume}\\\\
-      {\TStr}\\\\
       {\TNil}\\\\
@@
     \end{array}
-$$
+    \\\\
+    {\TLet} \ \ \  \ \ \ \ \ \ 
+    {\TApp}
+    \\\\
+    \begin{array}{c}
+      {\TNum}\\\\
+      {\TConst}
+    \end{array} \ \ \ \
+    \begin{array}{c}
+      {\TKw}\\\\
+      {\TNum}\\\\
+      {\TClass}
+    \end{array} \ \ \ \
+    \begin{array}{c}
+      {\TSubsume}
+    \end{array} \ \ \ \ 
+    \begin{array}{c}
+      {\TStr}\\\\
+      {\TTrue}
+    \end{array}
+  \end{array}
+    %\begin{array}{c}
+    %  {\TSubsume}\\\\
+    %  {\TStr}\\\\
+    %  {\TNil}\\\\
+    %  {\TFalse}
+    %\end{array}
+$$
+  \caption{Core typing rules}
+  \label{main:figure:othertypingrules}
+\end{figure*}
+
+%\figref{main:figure:typingrules} contains the core typing rules.
+%The key rule for reasoning about conditional control flow is
+%T-If. 
+%
+%\begin{mathpar}
+%  {\TIf}
+%\end{mathpar}
+
+%The propositions of the test expression \e{1}, \thenprop{\prop{1}} and \elseprop{\prop{1}}, are 
+%used as assumptions in the then and else branch respectively.
+
+The let rule T-Let links inferred information about
+\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.
 
@@
 Subtyping for untagged unions is standard.
-
-\paragraph{class} The \classconst{} function is implemented as a constant
-function (via T-Const), which returns a latent object which is the argument applied to the
-\classpe{} path element
-$$
-\constanttypefigure{}
-$$
-The dynamic semantics are given in \figref{main:figure:primitivesem}.
 %The definition of \updateliteral{} supports various idioms relating to \classpe{}
@@
 Typed Clojure type \t{} = \File for the return type,
-and the proposition set says constructors can never be false---constructors
+and the propositions say constructors can never be false---constructors
 can never produce the internal boolean value that Clojure uses for \false{}, or \nil{}.
@@
 
-\begin{figure}
-  $$
-\begin{array}{lllll}
-  \restrictremovefigure{}
-\end{array}
-  $$
-  \caption{Restrict and Remove}
-  \label{main:figure:restrictremove}
-\end{figure}
+%\begin{figure}
+%  $$
+%\begin{array}{lllll}
+%  \restrictremovefigure{}
+%\end{array}
+%  $$
+%  \caption{Restrict and Remove}
+%  \label{main:figure:restrictremove}
+%\end{figure}

### 2015-07-10 formal-model.tex
rewrite 3.1

@@
 \paragraph{Typing rules}
-The core type rules are
+The core typing rules are
 given as \figref{main:figure:othertypingrules}. We introduce
-the interesting rules while demonstrating how to check
-the complement number predicate
+the interesting rules with the complement number predicate
+as a running example.
 \begin{equation}
@@
 
-The lambda rule T-Abs introduces \isprop{\Top}{\makelocal{d}}
+The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
 to check the body.
 With \propenv{} = \isprop{\Top}{\makelocal{d}},
-T-If then checks the test {\appexp{\numberhuh{}}{\makelocal{d}}}
-with T-App with three main steps.
-
-First, the operator \numberhuh{} is checked with T-Const, which
+the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
+via the T-App rule, with three main steps.
+
+First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
 uses 
@@
 Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
-the operand \makelocal{d} is checked with
+in T-App the operand \ep{} = \makelocal{d} is checked with
 T-Local as
-$$
+\begin{equation}
 \judgementselfrewrite{\propenv{}}
                      {\makelocal{d}}
-                     {\UnionNilNum}
+                     {\Top}
                      {\filterset{\notprop{\falsy}{\makelocal{d}}}
@@
                      {\makelocal{d}}
-$$
+\end{equation}
 which encodes the type, proposition, and object information
-previously discussed about variables.
-
-Finally, T-App substitutes the operand's object for the parameter
----here \makelocal{d} for \numberhuh{}'s \x{}---yielding
-$$
-\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
+about variables
+as previously discussed.
+
+Finally, the T-App rule substitutes the operand's object \objectp{}
+for the parameter \x{} in the latent type, propositions, and object.
+\begin{equation}
+\judgementselfrewrite{\propenv{}}
+  {\appexp{\numberhuh{}}{\makelocal{d}}}
   {\Boolean{}}
@@
   {\emptyobject{}}
-$$
-
-To check {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
-T-If then
-introduces \isprop{\Number{}}{\makelocal{d}}
-in the `then' branch and \notprop{\Number{}}{\makelocal{d}}
-in the `else' branch.
-%The object \x{} over latent propositions \thenprop{\prop{}} and
-%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
-%actual argument object---\objectp{}---is substituted in at application by T-App.
-
-The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
-The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
-\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
-  {\Boolean{}}
-  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
-             {\notprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-
-The T-If refines each branch based on the test---a
-{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
-for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
-for checking the `else' branch.
-Information from each branch is combined via subsumption for the overall type, propositions, and object.
-
-For example
-$
-\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
-  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
-  {\Boolean{}}
-  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-             {\isprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-$
-type checks because T-Subsume allows us to check both branches as
-$
-\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
+\end{equation}
+To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
+the latent `then' proposition of \constanttype{\numberhuh{}} with 
+\makelocal{d}, giving
+{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
+{\isprop{\Number{}}{\makelocal{d}}}.
+
+To check the branches of {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
+T-If
+introduces \thenprop{\prop{1}} = \isprop{\Number{}}{\makelocal{d}}
+in the `then' branch 
+\e{2} = {\false{}},
+and \elseprop{\prop{1}} = \notprop{\Number{}}{\makelocal{d}}
+in the `else' branch
+\e{3} = \true{}.
+
+The branches are checked with T-False and T-True respectively,
+%$$
+%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
+%  {\false{}}
+%  {\False{}}
+%  {\filterset{\botprop{}}
+%             {\topprop{}}}
+%  {\emptyobject{}}
+%$$
+then T-Subsume.
+$$
+\begin{array}{c}
+\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
   {\false{}}
@@
   {\emptyobject{}}
-$
-and
-$
-\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
+  \\
+\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
   {\true{}}
@@
   {\emptyobject{}}
-$
-respectively.
+\end{array}
+$$
+%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
+%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
+%and \object{} = {\emptyobject{}}.
+%
+%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
+%\issubtypein{}{\t{}}{\tp{}}.
+%and 
+%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+%from 
+%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
+The T-Subsume conditions
+\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
+and
+\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+allow us to pick compatible propositions.
+
+Finally T-Abs concludes, using the T-If outputs, with
+$$
+{\judgementselfrewrite{}{\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}}
+                    {\ArrowOne {\makelocal{d}} {\Top{}}
+                                      {\Boolean{}}
+                                      {\filterset {\notprop{\Number{}}{\makelocal{d}}}
+                                                  {\isprop{\Number{}}{\makelocal{d}}}}
+                                      {\emptyobject{}}}
+                    {\filterset {\topprop{}}
+                                {\botprop{}}}
+                    {\emptyobject{}}}
+$$
+
+%The object \x{} over latent propositions \thenprop{\prop{}} and
+%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
+%actual argument object---\objectp{}---is substituted in at application by T-App.
+
+%The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
+%The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
+%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
+%  {\Boolean{}}
+%  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+%             {\notprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+
+%The T-If refines each branch based on the test---a
+%{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
+%for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
+%for checking the `else' branch.
+%%Information from each branch is combined via subsumption for the overall type, propositions, and object.
+%The T-Subsume rule
+
+%For example
+%$
+%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
+%  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%type checks because T-Subsume allows us to check both branches as
+%$
+%\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
+%  {\false{}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%and
+%$
+%\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
+%  {\true{}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%respectively.
 %
@@
 
-The let rule T-Let links inferred information about
-\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.
-
+%The let rule T-Let links inferred information about
+%\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.
+%
 %The T-Local rule connects the type system to the proof system over type propositions
@@
 
-\paragraph{Constructors} First we check and convert {\newexp {\File{}} {\makestr{a/b}}} to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
+\paragraph{Constructors} First we check and convert {\newexp {\File{}} {\makestr{a/b}}} to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}.
 The T-New typing rule checks and rewrites constructors.
@@
 
-For convenience, examples in this section are flattened when they are really nested
-let bindings. We also elide trivial latent propositions and objects.
-The following is an abbreviation of the previous expression.
-\begin{lstlisting}
-${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
-${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
-${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
-$\appexp{hi}{\makekw{en}}$
-\end{lstlisting}
+%For convenience, examples in this section are flattened when they are really nested
+%let bindings. We also elide trivial latent propositions and objects.
+%The following is an abbreviation of the previous expression.
+%\begin{lstlisting}
+%${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
+%${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
+%${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
+%$\appexp{hi}{\makekw{en}}$
+%\end{lstlisting}
 

### 2015-07-10 formal-model.tex
tweak 3.1

@@
 \lambdatc{}.
-
+%
 Building on occurrence typing,
@@
 occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
-We build up the occurrence typing calculus for illustrative purposes, 
-and present the full syntax at the end of the section.
+%We build up the occurrence typing calculus for illustrative purposes, 
+%and present the full syntax at the end of the section.
 
 \paragraph{Logical formulas over type environments}
-Standard type systems use type environments to track lexical type information.
-Introducing a lexically scoped variable $x$ is a matter of
-associating $x$ with its type $\Number$ (a number) in the type environment.
-Correspondingly, checking the type of a variable is a simple map lookup.
+In standard type systems, the 
+%use type environments to track lexical type information.
+rules for variables use \emph{type environments}.
+%associating $x$ with its type $\Number$ (a number) in the type environment,
+%%Correspondingly, 
+%and checking the type of a variable is a simple map lookup.
 $$
 \begin{array}{lr}
-\judgementtwo{\x{} \mapsto \Number}{\appexp{\inc{}}{\x{}}}{\Number}
+\infer [LC-Let]
+{ \judgementtwo {{\propenv{}}}
+                {\e{1}} {\s{}}
+  \\
+  \judgementtwo {{\propenv{}},\x{} \mapsto {\s{}}}
+                {\e{2}} {\t{}}
+}
+{ 
+  \judgementtwo {\propenv{}} 
+            {\letexp{\x{}}{\e{1}}{\e{2}}} {\t{}}
+           }
 &
@@
 
-Occurrence typing takes a different approach.
-The same information is conveyed by \emph{logical formulas}.
-The logical statement {\isprop{\Number{}}{\x{}}} says the value
-obtained via $x$ is of type \Number{}. A \emph{logical system}
-must now \emph{prove}
-a variable's type.
+Occurrence typing instead uses \emph{logical formulas}
+and a \emph{proof system}.
+The logical statement {\isprop{\s{}}{\x{}}} says
+variable $x$ is of type \s{}. 
+%A \emph{logical system}
+%must now \emph{prove}
+%a variable's type.
 $$
 \begin{array}{lr}
-\judgementtwo{\isprop{\Number{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\Number}
+\infer [T0-Let]
+{ \judgementtwo {{\propenv{}}}
+                {\e{1}} {\s{}}
+  \\
+  \judgementtwo {{\propenv{}},\isprop{\s{}}{\x{}}}
+                {\e{2}} {\t{}}
+}
+{ 
+  \judgementtwo {\propenv{}} 
+            {\letexp{\x{}}{\e{1}}{\e{2}}}
+            {\replacefor{\t{}}{\emptyobject{}}{\x{}}}
+           }
+%\judgementtwo{\isprop{\Number{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\Number}
 &
-\infer [T-LocalProve]
+\infer [T0-Local]
 { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
@@
 says under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
+Types can now refer to variables, so we erase \x{} from the final type
+in T0-Let.
 %We later define the more general T-Local.
@@
 \paragraph{Symbolic representation}
-As we saw in \secref{sec:overview},
-occurrence typing is capable of reasoning
-about deeply nested expressions.
-For each expression, the type system assigns a symbolic representation
-with respect to the current lexical environment. For example,
-\begin{itemize}
-  \item variable \makelocal{m} is represented as \makelocal{m}; 
-  \item $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ is represented as ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$;
-  \item $42$ has no representation, called \emptyobject{}.
-\end{itemize}
-This symbolic representation is called the expression's \emph{object}. 
-\figref{main:figure:termsyntax} gives the syntax for objects.
-\emptyobject{} is the
-\emph{empty object}. Non-empty objects \path{\pathelem{}}{\x{}} consist of a root variable \x{} and a \emph{path} \pathelem{},
+%As we saw in \secref{sec:overview},
+%occurrence typing is capable of reasoning
+%about deeply nested expressions.
+To reason about nested expressions,
+each expression is associated with 
+a symbolic representation
+called an \emph{object}.
+%with respect to the current lexical environment. 
+For example,
+  variable \makelocal{m} has object \makelocal{m};
+  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
+  $42$ has the \emph{empty} object \emptyobject{}.
+
+\figref{main:figure:termsyntax} gives the syntax for objects---non-empty objects 
+\path{\pathelem{}}{\x{}} consist of a root variable \x{} and a \emph{path} \pathelem{},
+consisting of
 a possibly-empty sequence of \emph{path elements} applied right-to-left starting from the root variable.
@@
 \paragraph{Logical statements about objects}
-So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}}---
-saying variable \x{} is of type \t{}.
+%So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}}---
+%saying variable \x{} is of type \t{}.
 We extend logical statements to reason about \emph{arbitrary} non-empty 
@@
 To attain the type of \makelocal{o}, 
-T-LocalProve says we must solve for \t{} in
+T0-Local says we must solve for \t{} in
 $
@@
 $$
-
-The main 
-$$
-\LUpdatePos
-$$
-Now the purpose of a logical representation can be conveyed---we rely on the logical
-system to combine pieces of type information to deduce more accurate types for lexical
-bindings.
+which deduces \t{} to be a {\makekw{combo}} meal.
+The logical
+system \emph{combines} pieces of type information to deduce more accurate types for lexical
+bindings---this is explained in \secref{formalmodel:proofsystem}.
 
@@
 
-\paragraph{`Then' and `else' propositions} Along with an object, each expression carries two
-propositions---a `then' and an `else' proposition,
-that can be assumed if the expression evaluates to a
-true value and a false value respectively.
-
-The `then' proposition {\thenprop{\prop{}}} for the variable $o$ is $\notprop{\falsy{}}{o}$, for example.
-This says if the expression $o$ evaluates to a true value, then $o$ is neither
-of type \False or \Nil. Naturally, the `else' proposition {\elseprop{\prop{}}} is $\isprop{\falsy{}}{o}$---which
-is true when $o$ is a false value---says $o$ is either \False or \Nil.
+\paragraph{Expressions} The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
+application, abstractions, conditionals, and let expressions.
+All binding forms introduce fresh variables.
+Values include booleans, \nil{}, class literals, keywords, 
+numbers,
+constants, and strings. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
+close value environments \openv{} over functions, which map bindings to values.
+
+\paragraph{Types} Types include the top type, \emph{untagged} unions, functions, singleton types
+and class instances.
+We abbreviate the classes
+\Booleanlong{} to \Boolean{}, 
+\Keywordlong{} to \Keyword{},
+\NumberFull{}  to \Number{},
+\StringFull{}  to \String{}, and 
+\FileFull{}  to \File{}.
+We also abbreviate the types
+\EmptyUnion{}     to \Bot{}, 
+{\ValueNil}       to \Nil{}, 
+{\ValueTrue}      to \True, and
+{\ValueFalse} to {\False}.
+
+The difference between the types
+\Value{\Keyword} and \Keyword{} is subtle.
+The former is inhabited by the class literal \Keyword{} and the result of 
+\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
+
+Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
+which may refer to the function argument.
+%Latent means they are relevant when the function is applied rather than evaluated.
+They are latent because they are instantiated with the
+actual object of the argument in applications before they are used in the proposition environment.
+
+\paragraph{Branching propositions} 
+%Expressions in test position communicate propositions to each branch.
+Each test expression \e{1} in
+\ifexp{\e{1}}{\e{2}}{\e{3}} has a `then' and an `else' proposition to check
+\e{2} and \e{3} respectively.
+For example, in \ifexp{\makelocal{o}}{\e{2}}{\e{3}}
+we know {\makelocal{o}} is true in \e{2} and false in \e{3}.
+%that can be assumed if the expression evaluates to a
+%true value and a false value respectively.
+%
+The `then' and `else' propositions 
+for \makelocal{o}
+are $\notprop{\falsy{}}{\makelocal{o}}$ and $\isprop{\falsy{}}{\makelocal{o}}$ respectively.
 
@@
 for expressions that rewrite to themselves.
-
-\paragraph{Expressions} The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
-application, abstractions, conditionals, and let expressions.
-All binding forms introduce fresh variables.
-Values include booleans, \nil{}, class literals, keywords, 
-numbers,
-constants, and strings. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
-close over functions with value environments \openv{}, which map bindings to values.
-
-\paragraph{Types} Types include the top type, \emph{untagged} unions, functions, singleton types
-and class instances.
-We abbreviate the classes
-\Booleanlong{} to \Boolean{}, 
-\Keywordlong{} to \Keyword{},
-\NumberFull{}  to \Number{},
-\StringFull{}  to \String{}, and 
-\FileFull{}  to \File{}.
-We also abbreviate the types
-\EmptyUnion{}     to \Bot{}, 
-{\ValueNil}       to \Nil{}, 
-{\ValueTrue}      to \True, and
-{\ValueFalse} to {\False}.
-
-The difference between the types
-\Value{\Keyword} and \Keyword{} is subtle.
-The former is inhabited by the class literal \Keyword{} and the result of 
-\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
-
-Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
-which may refer to the function argument.
-%Latent means they are relevant when the function is applied rather than evaluated.
-They are latent because they are instantiated with the
-actual object of the argument in applications before they are used in the proposition environment.
 
@@
 %{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
-The T-Subsume conditions
+The T-Subsume premises
 \inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
@@
 \inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-allow us to pick compatible propositions.
+allow us to pick compatible propositions for both branches.
 
@@
 
+\label{formalmodel:proofsystem}
 \subsection{Proof system}

### 2015-07-10 formal-model.tex
Rearrange 3.1

@@
 
-\subsection{Occurrence typing}
+\subsection{Core type system}
 
@@
 
-\paragraph{Logical formulas over type environments}
-In standard type systems, the 
-%use type environments to track lexical type information.
-rules for variables use \emph{type environments}.
-%associating $x$ with its type $\Number$ (a number) in the type environment,
-%%Correspondingly, 
-%and checking the type of a variable is a simple map lookup.
+\paragraph{Expressions} Syntax is given in \figref{main:figure:termsyntax}. Expressions \e{} 
+include variables \x{}, values \v{},
+applications, abstractions, conditionals, and let expressions.
+All binding forms introduce fresh variables.
+Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
+numbers {\num{}},
+constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
+close value environments \openv{} over functions, which map bindings to values.
+
+\paragraph{Types} Types \s{} or \t{} include the top type, \emph{untagged} unions, functions, singleton types
+and class instances.
+We abbreviate the classes
+\Booleanlong{} to \Boolean{}, 
+\Keywordlong{} to \Keyword{},
+\NumberFull{}  to \Number{},
+\StringFull{}  to \String{}, and 
+\FileFull{}  to \File{}.
+We also abbreviate the types
+\EmptyUnion{}     to \Bot{}, 
+{\ValueNil}       to \Nil{}, 
+{\ValueTrue}      to \True, and
+{\ValueFalse} to {\False}.
+
+The difference between the types
+\Value{\Keyword} and \Keyword{} is subtle.
+The former is inhabited by the class literal \Keyword{} and the result of 
+\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
+
+Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
+which may refer to the function argument.
+%Latent means they are relevant when the function is applied rather than evaluated.
+They are latent because they are instantiated with the
+actual object of the argument in applications before they are used in the proposition environment.
+
+\paragraph{Objects}
+%As we saw in \secref{sec:overview},
+%occurrence typing is capable of reasoning
+%about deeply nested expressions.
+To reason about nested expressions,
+each expression is associated with 
+a symbolic representation
+called an \emph{object}.
+%with respect to the current lexical environment. 
+For example,
+  variable \makelocal{m} has object \makelocal{m};
+  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
+  $42$ has the \emph{empty} object \emptyobject{}.
+
+\figref{main:figure:termsyntax} gives the syntax for objects \object{}---non-empty objects 
+\path{\pathelem{}}{\x{}} combine of a root variable \x{} and a \emph{path} \pathelem{},
+which consists of
+a possibly-empty sequence of \emph{path elements} (\pesyntax{}) applied right-to-left from the root variable.
+We use two path elements---{\classpe{}} and {\keype{k}}---representing the results
+of calling \classconst{} and of looking up a keyword $k$ respectively.
+
+\paragraph{Propositions}
+In standard type systems, \emph{type environments}
+track variables.
 $$
@@
 $$
-The proof system judgment { \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
-says under logical assumptions {\propenv{}}, variable \x{} is of type \t{}.
 Types can now refer to variables, so we erase \x{} from the final type
 in T0-Let.
-%We later define the more general T-Local.
-
-\paragraph{Symbolic representation}
-%As we saw in \secref{sec:overview},
-%occurrence typing is capable of reasoning
-%about deeply nested expressions.
-To reason about nested expressions,
-each expression is associated with 
-a symbolic representation
-called an \emph{object}.
-%with respect to the current lexical environment. 
-For example,
-  variable \makelocal{m} has object \makelocal{m};
-  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
-  $42$ has the \emph{empty} object \emptyobject{}.
-
-\figref{main:figure:termsyntax} gives the syntax for objects---non-empty objects 
-\path{\pathelem{}}{\x{}} consist of a root variable \x{} and a \emph{path} \pathelem{},
-consisting of
-a possibly-empty sequence of \emph{path elements} applied right-to-left starting from the root variable.
-We use two path elements---{\classpe{}} and {\keype{k}}---representing the result
-of a calling \classconst{} and the result of looking up a keyword $k$ respectively.
-
-%The particular part of the runtime environment we reference in a
-%type proposition is called the \emph{object}.
-%The typing judgment relates an object to every expression in the language.
-%An object is either \emph{empty}, written \emptyobject{}, 
-%which says 
-%this expression is not known to evaluate to a particular part
-%  of the current runtime environment, or a 
-%variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
-%that exactly indicates how the value of this
-%expression can be derived from the current runtime environment.
-%Type propositions can only reference non-empty objects.
-
-\paragraph{Logical statements about objects}
-%So far we have seen logical statements about the simplest object, a variable---{\isprop{\t{}}{\x{}}}---
-%saying variable \x{} is of type \t{}.
 We extend logical statements to reason about \emph{arbitrary} non-empty 
 objects {\isprop{\t{}}{\path{\pathelem{}}{\x{}}}}.
-
-To illustrate, recall \egref{example:desserts-on-meal}.
-We know the parameter \makelocal{o} is of type $\Order$
-by the annotation on $desserts$---written
-\begin{equation}
-\qquad {\isprop{\Order}{\makelocal{o}}}
-\end{equation}
-as a logical statement.
-%
-In checking the ${\makekw{combo}}$ method, we also assume
-${\appexp{\makekw{Meal}}{\makelocal{o}}}$ is ${\makekw{combo}}$,
-based on multimethod dispatch rules. This is written
-\begin{equation}
-  \qquad {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}
-\end{equation}
-and pronounced ``the ${\makekw{Meal}}$ path of variable \makelocal{o} is of type
-{\Value{\makekw{combo}}}'', a keyword singleton type.
-
-\paragraph{Logical system in action} 
-To attain the type of \makelocal{o}, 
-T0-Local says we must solve for \t{} in
-$
-{ \inpropenv 
-  {\propenv{}}
-  {\isprop {\t{}} {\makelocal{o}}}}
-$,
-under proposition environment 
-$$
-\propenv{} = {{\isprop{\Order}{\makelocal{o}}},
-    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
-$$
-which deduces \t{} to be a {\makekw{combo}} meal.
-The logical
-system \emph{combines} pieces of type information to deduce more accurate types for lexical
-bindings---this is explained in \secref{formalmodel:proofsystem}.
-
-\paragraph{Extended to propositional logic}
-We extend our logical statements to \emph{propositional logic}.
-\figref{main:figure:termsyntax} describes the syntax we will use for the rest of the paper.
-The proposition metavariable \prop{} 
-consists of positive and negative \emph{type propositions} 
-about non-empty objects---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
-respectively---the former we have seen before, 
-the latter pronounced ``the object {\path {\pathelem{}} {\x{}}} is \emph{not} of type \t{}''.
-The other propositions are standard logical connectives: implications, conjunctions,
-disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions.
-
-%The first insight about occurrence typing is that
-%logical formulas
-%can be used to represent type information about our programs
-%by relating parts of the runtime environment to types
-%via propositional logic.
-%\emph{Type propositions}  make assertions like ``variable \x{} is of type \NumberFull{}'' or
-%``variable \x{} is not \nil{}''---in our logical system we write these as
-%{\isprop{\NumberFull}{\x{}}}
-%and {\notprop{\Nil{}}{\x{}}}. 
-%
-%The second insight is that we can replace the traditional 
-%representation of a
-%type environment (eg., a map from variables to types)
-%with a set of propositions, written \propenv{}. 
-%Instead of mapping \x{} to
-%the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
-
-\paragraph{Expressions} The syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
-application, abstractions, conditionals, and let expressions.
-All binding forms introduce fresh variables.
-Values include booleans, \nil{}, class literals, keywords, 
-numbers,
-constants, and strings. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
-close value environments \openv{} over functions, which map bindings to values.
-
-\paragraph{Types} Types include the top type, \emph{untagged} unions, functions, singleton types
-and class instances.
-We abbreviate the classes
-\Booleanlong{} to \Boolean{}, 
-\Keywordlong{} to \Keyword{},
-\NumberFull{}  to \Number{},
-\StringFull{}  to \String{}, and 
-\FileFull{}  to \File{}.
-We also abbreviate the types
-\EmptyUnion{}     to \Bot{}, 
-{\ValueNil}       to \Nil{}, 
-{\ValueTrue}      to \True, and
-{\ValueFalse} to {\False}.
-
-The difference between the types
-\Value{\Keyword} and \Keyword{} is subtle.
-The former is inhabited by the class literal \Keyword{} and the result of 
-\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
-
-Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
-which may refer to the function argument.
-%Latent means they are relevant when the function is applied rather than evaluated.
-They are latent because they are instantiated with the
-actual object of the argument in applications before they are used in the proposition environment.
-
-\paragraph{Branching propositions} 
-%Expressions in test position communicate propositions to each branch.
-Each test expression \e{1} in
-\ifexp{\e{1}}{\e{2}}{\e{3}} has a `then' and an `else' proposition to check
-\e{2} and \e{3} respectively.
-For example, in \ifexp{\makelocal{o}}{\e{2}}{\e{3}}
-we know {\makelocal{o}} is true in \e{2} and false in \e{3}.
-%that can be assumed if the expression evaluates to a
-%true value and a false value respectively.
-%
-The `then' and `else' propositions 
-for \makelocal{o}
-are $\notprop{\falsy{}}{\makelocal{o}}$ and $\isprop{\falsy{}}{\makelocal{o}}$ respectively.
-
-%The pair of propositions {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} is called a \emph{proposition set}.
-
-\paragraph{The full judgment}
-
-We formalize our type system following~\citet{TF10}.
-%(with differences highlighted in $\trdiff{\text{blue}}$)
-The typing judgment 
-$$
-{\judgementrewrite   {\propenv}
-              {\e{}} {\t{}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-  {\object{}}
-  {\ep{}}}
-$$
-says expression \e{} rewrites to \ep{}, which
-is of type \t{} in the 
-proposition environment $\propenv{}$, with 
-`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}. We write 
-{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
-  or
-{\judgementselfrewrite   {\propenv}
-              {\e{}} {\t{}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-  {\object{}}}
-for expressions that rewrite to themselves.
-
-\paragraph{Typing rules}
-The core typing rules are
-given as \figref{main:figure:othertypingrules}. We introduce
-the interesting rules with the complement number predicate
-as a running example.
-\begin{equation}
-\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
-\end{equation}
-%, including
-%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
-%as impossible (\botprop{}) `then' propositions.
-
-The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
-to check the body.
-With \propenv{} = \isprop{\Top}{\makelocal{d}},
-the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
-via the T-App rule, with three main steps.
-
-First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
-uses 
-the \constanttypeliteral{} relation to give types to constants.
-\numberhuh{} is a predicate over numbers, and
-\classconst{} returns its argument's class.
-$$
-\constanttypefigure{}
-$$
-The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.
-
-Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
-in T-App the operand \ep{} = \makelocal{d} is checked with
-T-Local as
-\begin{equation}
-\judgementselfrewrite{\propenv{}}
-                     {\makelocal{d}}
-                     {\Top}
-                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
-                                {\isprop{\falsy}{\makelocal{d}}}}
-                     {\makelocal{d}}
-\end{equation}
-which encodes the type, proposition, and object information
-about variables
-as previously discussed.
-
-Finally, the T-App rule substitutes the operand's object \objectp{}
-for the parameter \x{} in the latent type, propositions, and object.
-\begin{equation}
-\judgementselfrewrite{\propenv{}}
-  {\appexp{\numberhuh{}}{\makelocal{d}}}
-  {\Boolean{}}
-  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
-             {\notprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-\end{equation}
-To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
-the latent `then' proposition of \constanttype{\numberhuh{}} with 
-\makelocal{d}, giving
-{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
-{\isprop{\Number{}}{\makelocal{d}}}.
-
-To check the branches of {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
-T-If
-introduces \thenprop{\prop{1}} = \isprop{\Number{}}{\makelocal{d}}
-in the `then' branch 
-\e{2} = {\false{}},
-and \elseprop{\prop{1}} = \notprop{\Number{}}{\makelocal{d}}
-in the `else' branch
-\e{3} = \true{}.
-
-The branches are checked with T-False and T-True respectively,
-%$$
-%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
-%  {\false{}}
-%  {\False{}}
-%  {\filterset{\botprop{}}
-%             {\topprop{}}}
-%  {\emptyobject{}}
-%$$
-then T-Subsume.
-$$
-\begin{array}{c}
-\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
-  {\false{}}
-  {\Boolean{}}
-  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-             {\isprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-  \\
-\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
-  {\true{}}
-  {\Boolean{}}
-  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-             {\isprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-\end{array}
-$$
-%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
-%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
-%and \object{} = {\emptyobject{}}.
-%
-%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
-%\issubtypein{}{\t{}}{\tp{}}.
-%and 
-%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-%from 
-%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
-The T-Subsume premises
-\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
-and
-\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-allow us to pick compatible propositions for both branches.
-
-Finally T-Abs concludes, using the T-If outputs, with
-$$
-{\judgementselfrewrite{}{\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}}
-                    {\ArrowOne {\makelocal{d}} {\Top{}}
-                                      {\Boolean{}}
-                                      {\filterset {\notprop{\Number{}}{\makelocal{d}}}
-                                                  {\isprop{\Number{}}{\makelocal{d}}}}
-                                      {\emptyobject{}}}
-                    {\filterset {\topprop{}}
-                                {\botprop{}}}
-                    {\emptyobject{}}}
-$$
-
-%The object \x{} over latent propositions \thenprop{\prop{}} and
-%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
-%actual argument object---\objectp{}---is substituted in at application by T-App.
-
-%The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
-%The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
-%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
-%  {\Boolean{}}
-%  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
-%             {\notprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-
-%The T-If refines each branch based on the test---a
-%{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
-%for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
-%for checking the `else' branch.
-%%Information from each branch is combined via subsumption for the overall type, propositions, and object.
-%The T-Subsume rule
-
-%For example
-%$
-%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
-%  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
-%  {\Boolean{}}
-%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-%             {\isprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-%$
-%type checks because T-Subsume allows us to check both branches as
-%$
-%\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
-%  {\false{}}
-%  {\Boolean{}}
-%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-%             {\isprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-%$
-%and
-%$
-%\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
-%  {\true{}}
-%  {\Boolean{}}
-%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-%             {\isprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-%$
-%respectively.
-%
-%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
-%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
-%we use T-Subsume on both branches
-%to introduce a logical disjunction
-%since at least one must be true if the entire expression is true.
-
-%Given a set of propositions, we can use logical reasoning to derive
-%new information about our programs
-%with the judgment \inpropenv{\propenv{}}{\prop{}}.
-%In addition to the standard rules for the logical connectives, the key
-%rule is L-Update, which combines multiple propositions about the same variable,
-%allowing us to refine its type.
-%$$
-%  {\LUpdate}
-%$$
-%For example, with L-Update we can use the knowledge of
-%\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
-%and 
-%\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
-%to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
-%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
-%We cover L-Update in more detail in \secref{sec:formalpaths}.
-%
-%Finally, this approach allows the type system to track
-%programming idioms from 
-%dynamic languages
-%using implicit type-based reasoning based on the result of
-%conditional tests.
-%For instance,
-%\egref{example:parent-if}
-%only utilizes \clj{f} once
-%the programmer is convinced it is safe to do so based whether
-%\clj{f}
-%is
-%true or false. 
-%To express this in the type system, every expression 
-%is described by two propositions: a `then' proposition
-%for when it reduces to a true value, and an `else' proposition
-%when it reduces to a false value---for \clj{f}
-%the then proposition is {\notprop{\falsy}{f}} and 
-%the else proposition is {\isprop{\falsy}{f}}.
-%\ref{main:figure:typingrules}
+The proof system judgment in T0-Local
+$$
+{ \inpropenv {\propenv{}} {\isprop {\t{}} {\path{\pathelem{}}{\x{}}}} }
+$$
+says under logical assumptions {\propenv{}}, object {\path{\pathelem{}}{\x{}}} is of type \t{}.
+%We later define the more general T-Local.
 
@@
 \begin{array}{lrll}
-  \expd{}, \e{} &::=& \x{}
+  \e{} &::=& \x{}
                       \alt \v{} 
@@
 \end{figure}
+
+We further extend logical statements to \emph{propositional logic}.
+\figref{main:figure:termsyntax} describes the syntax
+for propositions \prop{},
+consisting of positive and negative \emph{type propositions} 
+about non-empty objects---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
+respectively---the latter pronounced ``the object {\path {\pathelem{}} {\x{}}} is \emph{not} of type \t{}''.
+The other propositions are standard logical connectives: implications, conjunctions,
+disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions.
+
+Each expression is associated with two propositions---when expression
+\e{1} is in test position like
+\ifexp{\e{1}}{\e{2}}{\e{3}},
+the type system extracts \e{1}'s `then' and `else' proposition to check
+\e{2} and \e{3} respectively.
+For example, in \ifexp{\makelocal{o}}{\e{2}}{\e{3}}
+we learn {\makelocal{o}} is true in \e{2} via {\makelocal{o}}'s `then' proposition $\notprop{\falsy{}}{\makelocal{o}}$, and 
+that {\makelocal{o}} is false in \e{3} via {\makelocal{o}}'s `else' proposition $\isprop{\falsy{}}{\makelocal{o}}$.
+
+To illustrate, recall \egref{example:desserts-on-meal}.
+We know the parameter \makelocal{o} is of type $\Order$,
+%by the annotation on $desserts$
+written
+{\isprop{\Order}{\makelocal{o}}}
+as a proposition.
+%
+In checking the ${\makekw{combo}}$ method, we also assume
+${\appexp{\makekw{Meal}}{\makelocal{o}}}$ is ${\makekw{combo}}$,
+based on multimethod dispatch rules. This is written
+  {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}},
+pronounced ``the ${\makekw{Meal}}$ path of variable \makelocal{o} is of type
+{\Value{\makekw{combo}}}''.
+
+%\paragraph{Logical system in action} 
+To attain the type of \makelocal{o}, 
+we must solve for \t{} in
+$
+{ \inpropenv 
+  {\propenv{}}
+  {\isprop {\t{}} {\makelocal{o}}}}
+$,
+under proposition environment
+$
+\propenv{} = {{\isprop{\Order}{\makelocal{o}}},
+    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
+$
+which deduces \t{} to be a {\makekw{combo}} meal.
+The logical
+system \emph{combines} pieces of type information to deduce more accurate types for lexical
+bindings---this is explained in \secref{formalmodel:proofsystem}.
+
+%The first insight about occurrence typing is that
+%logical formulas
+%can be used to represent type information about our programs
+%by relating parts of the runtime environment to types
+%via propositional logic.
+%\emph{Type propositions}  make assertions like ``variable \x{} is of type \NumberFull{}'' or
+%``variable \x{} is not \nil{}''---in our logical system we write these as
+%{\isprop{\NumberFull}{\x{}}}
+%and {\notprop{\Nil{}}{\x{}}}. 
+%
+%The second insight is that we can replace the traditional 
+%representation of a
+%type environment (eg., a map from variables to types)
+%with a set of propositions, written \propenv{}. 
+%Instead of mapping \x{} to
+%the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \SRefl{}
+
+    \STop{}
+
+\SUnionSuper{}
+
+\SUnionSub{}
+
+\SFunMono{}
+
+\SObject{}
+
+\SClass{}
+
+\begin{array}{l}
+\SSBool{}\\\\
+\SSKw{}
+\end{array}
+
+\SFun{}
+
+  \end{mathpar}
+  \caption{Core Subtyping rules}
+  \label{main:figure:subtyping}
+\end{figure}
+
+\paragraph{Subtyping}
+\figref{main:figure:subtyping} presents subtyping
+as a reflexive and transitive relation with top type \Top. 
+Singleton types are instances of their respective classes---boolean singleton types
+are of type \Boolean{}, class literals are instances of \Class{} and keywords are
+instances of \Keyword{}.
+Instances of classes \class{} are subtypes of \Object{}. Since function types 
+are subtypes of \IFn{}, all types except for \Nil{} are subtypes of \Object{},
+so \Top{} = {\Union{\Nil}{\Object}}.
+Function subtyping is contravariant left of the arrow---latent propositions, object
+and the result type are covariant.
+Subtyping for untagged unions is standard.
+
+\paragraph{The full judgment}
+
+We formalize our type system following~\citet{TF10}.
+%(with differences highlighted in $\trdiff{\text{blue}}$)
+The typing judgment 
+$$
+{\judgementrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}
+  {\ep{}}}
+$$
+says expression \e{} rewrites to \ep{}, which
+is of type \t{} in the 
+proposition environment $\propenv{}$, with 
+`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+and object \object{}. We write 
+{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
+  or
+{\judgementselfrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}}
+for expressions that rewrite to themselves.
+
+\paragraph{Typing rules}
+The core typing rules are
+given as \figref{main:figure:othertypingrules}. We introduce
+the interesting rules with the complement number predicate
+as a running example.
+\begin{equation}
+\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+\end{equation}
+%, including
+%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
+%as impossible (\botprop{}) `then' propositions.
+
+The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
+to check the body.
+With \propenv{} = \isprop{\Top}{\makelocal{d}},
+the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
+via the T-App rule, with three main steps.
+
+First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
+uses 
+\constanttypeliteral{} to type constants.
+\numberhuh{} is a predicate over numbers, and
+\classconst{} returns its argument's class.
+$$
+\constanttypefigure{}
+$$
+The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.
+
+Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
+in T-App the operand \ep{} = \makelocal{d} is checked with
+T-Local as
+\begin{equation}
+\judgementselfrewrite{\propenv{}}
+                     {\makelocal{d}}
+                     {\Top}
+                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
+                                {\isprop{\falsy}{\makelocal{d}}}}
+                     {\makelocal{d}}
+\end{equation}
+which encodes the type, proposition, and object information
+about variables
+as previously discussed.
+
+Finally, the T-App rule substitutes the operand's object \objectp{}
+for the parameter \x{} in the latent type, propositions, and object.
+\begin{equation}
+\judgementselfrewrite{\propenv{}}
+  {\appexp{\numberhuh{}}{\makelocal{d}}}
+  {\Boolean{}}
+  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+             {\notprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+\end{equation}
+To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
+the latent `then' proposition of \constanttype{\numberhuh{}} with 
+\makelocal{d}, giving
+{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
+{\isprop{\Number{}}{\makelocal{d}}}.
+
+To check the branches of {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
+T-If
+introduces \thenprop{\prop{1}} = \isprop{\Number{}}{\makelocal{d}}
+to check \e{2} = {\false{}},
+and \elseprop{\prop{1}} = \notprop{\Number{}}{\makelocal{d}}
+to check 
+\e{3} = \true{}.
+%
+The branches are first checked with T-False and T-True respectively,
+then T-Subsume.
+%$$
+%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
+%  {\false{}}
+%  {\False{}}
+%  {\filterset{\botprop{}}
+%             {\topprop{}}}
+%  {\emptyobject{}}
+%$$
+$$
+\begin{array}{c}
+\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
+  {\false{}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+  \\
+\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
+  {\true{}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+\end{array}
+$$
+%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
+%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
+%and \object{} = {\emptyobject{}}.
+%
+%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
+%\issubtypein{}{\t{}}{\tp{}}.
+%and 
+%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+%from 
+%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
+The T-Subsume premises
+\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
+and
+\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+allow us to pick compatible propositions for both branches.
+
+Finally T-Abs concludes, using the T-If outputs, with
+$$
+{\judgementselfrewrite{}{\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}}
+                    {\ArrowOne {\makelocal{d}} {\Top{}}
+                                      {\Boolean{}}
+                                      {\filterset {\notprop{\Number{}}{\makelocal{d}}}
+                                                  {\isprop{\Number{}}{\makelocal{d}}}}
+                                      {\emptyobject{}}}
+                    {\filterset {\topprop{}}
+                                {\botprop{}}}
+                    {\emptyobject{}}}
+$$
+
+%The object \x{} over latent propositions \thenprop{\prop{}} and
+%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
+%actual argument object---\objectp{}---is substituted in at application by T-App.
+
+%The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
+%The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
+%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
+%  {\Boolean{}}
+%  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+%             {\notprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+
+%The T-If refines each branch based on the test---a
+%{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
+%for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
+%for checking the `else' branch.
+%%Information from each branch is combined via subsumption for the overall type, propositions, and object.
+%The T-Subsume rule
+
+%For example
+%$
+%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
+%  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%type checks because T-Subsume allows us to check both branches as
+%$
+%\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
+%  {\false{}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%and
+%$
+%\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
+%  {\true{}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%respectively.
+%
+%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
+%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
+%we use T-Subsume on both branches
+%to introduce a logical disjunction
+%since at least one must be true if the entire expression is true.
+
+%Given a set of propositions, we can use logical reasoning to derive
+%new information about our programs
+%with the judgment \inpropenv{\propenv{}}{\prop{}}.
+%In addition to the standard rules for the logical connectives, the key
+%rule is L-Update, which combines multiple propositions about the same variable,
+%allowing us to refine its type.
+%$$
+%  {\LUpdate}
+%$$
+%For example, with L-Update we can use the knowledge of
+%\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
+%and 
+%\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
+%to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
+%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
+%We cover L-Update in more detail in \secref{sec:formalpaths}.
+%
+%Finally, this approach allows the type system to track
+%programming idioms from 
+%dynamic languages
+%using implicit type-based reasoning based on the result of
+%conditional tests.
+%For instance,
+%\egref{example:parent-if}
+%only utilizes \clj{f} once
+%the programmer is convinced it is safe to do so based whether
+%\clj{f}
+%is
+%true or false. 
+%To express this in the type system, every expression 
+%is described by two propositions: a `then' proposition
+%for when it reduces to a true value, and an `else' proposition
+%when it reduces to a false value---for \clj{f}
+%the then proposition is {\notprop{\falsy}{f}} and 
+%the else proposition is {\isprop{\falsy}{f}}.
+%\ref{main:figure:typingrules}
 
@@
 
-\paragraph{Subtyping}
-\figref{main:figure:subtyping} presents subtyping
-as a reflexive and transitive relation with top type \Top. 
-Singleton types are instances of their respective classes---boolean singleton types
-are of type \Boolean{}, class literals are instances of \Class{} and keywords are
-instances of \Keyword{}.
-Instances of classes \class{} are subtypes of \Object{}. Since function types 
-are subtypes of \IFn{}, all types except for \Nil{} are subtypes of \Object{},
-so \Top{} = {\Union{\Nil}{\Object}}.
-Function subtyping is contravariant left of the arrow---latent propositions, object
-and the result type are covariant.
-Subtyping for untagged unions is standard.
 %The definition of \updateliteral{} supports various idioms relating to \classpe{}
@@
 
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \SRefl{}
-
-    \STop{}
-
-\SUnionSuper{}
-
-\SUnionSub{}
-
-\SFunMono{}
-
-\SObject{}
-
-\SClass{}
-
-\begin{array}{l}
-\SSBool{}\\\\
-\SSKw{}
-\end{array}
-
-\SFun{}
-
-  \end{mathpar}
-  \caption{Core Subtyping rules}
-  \label{main:figure:subtyping}
-\end{figure}
-
 %\subsection{Reasoning about Exceptional Control Flow}

### 2015-07-10 formal-model.tex
add LOC. add HMap defalias counts. explain zero MM in f2imap
initial POPL submission

@@
 \subsection{Core type system}
+\label{sec:coretypesystem}
 
@@
 \begin{equation}  \label{eq:hi-defmethoden}
-{\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
+{\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
 \end{equation}
@@
 restrict lookup and extension to keyword keys. The metavariable \mapval{}
-ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
-usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
+ranges over the runtime value of maps %{\curlymapvaloverright{\k{}}{\v{}}},
+%usually 
+written 
+{\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
 
@@
 
-The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
+The typing rules for \getliteral{} consider three cases. T-GetHMap models a lookup
 that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
@@
 
-\label{formalmodel:proofsystem}
-\subsection{Proof system}
-
-% checks obvious prop logic, but update is most interesting rule.
-
 \begin{figure*}
@@
 \end{figure*}
+
+\label{formalmodel:proofsystem}
+\subsection{Proof system}
+
+% checks obvious prop logic, but update is most interesting rule.
+Now that we have covered \lambdatc{}, we now review
+the proof system, in particular the T-Update rule.
+The rest of the proof system, which uses standard proposition logic, 
+we defer to the supplemental material---\figref{appendix:figure:proofsystem} contains the full rules.
+$$
+\LUpdate{}
+$$
+The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).
+The rule says if we know positive information \t{} about object \path{\pathelemp{}}{\x{}}, and some
+either positive or negative information \propisnotmeta{} about some extension of that path,
+\path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}, 
+the \updateliteral{} metafunction can derive some new positive information about \path{\pathelemp{}}{\x{}}.
+
+The example in 
+\secref{sec:coretypesystem}
+that combines information
+$$
+{ \inpropenv 
+  {{\isprop{\Order}{\makelocal{o}}},
+    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
+  {\isprop {\t{}} {\makelocal{o}}}}
+$$
+to deduce \t{},
+uses T-Update where \path{\pathelemp{}}{\x{}} = {\path{\keype{\makekw{Meal}}}{\makelocal{o}}}
+and
+$$
+\update{\Order}{\makekw{combo}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}
+$$
+Since {\Order} is just a union of HMaps, we update each member of the union
+via the last HMap case of
+\figref{main:figure:update}.
+Each recursive call hits the first HMap case, since {\makekw{combo}}
+is always present, finally calling \restrictliteral{} in the penultimate case.
+
+The non-combo HMaps change to \Bot, because \restrict{\makekw{lunch}}{\makekw{combo}} = \Bot,
+leaving just the \makekw{combo} HMap.
 

### 2015-07-10 formal-model.tex
Revert "initial POPL submission"
This reverts commit 8df022f4b37844e348bca957d47c99921b08163c.

@@
 restrict lookup and extension to keyword keys. The metavariable \mapval{}
-ranges over the runtime value of maps %{\curlymapvaloverright{\k{}}{\v{}}},
-%usually 
-written 
-{\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
+ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
+usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
 
@@
 
-The typing rules for \getliteral{} consider three cases. T-GetHMap models a lookup
+The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
 that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail

### 2015-07-10 conclusion.tex
Fits.

@@
 Optional type systems must be designed with close attention to the
-languague that they are intendend to work for.
+language that they are intended to work for.
 We have therefore designed Typed Clojure, an optionally-typed version of
@@
 
-The result is a sound, expressive, and useful type system which, when
+The result is a sound, expressive, and useful type system which, as
 implemented in \coretyped with appropriate extensions, suitable for
@@
 %
-As a result, Typed Clojure is already successful: it is widely used in
+As a result, Typed Clojure is already successful: it is used in
 the Clojure community among both enthusiasts and professional
@@
 our design choices. Multimethods, Java interoperation, and
-heterogenous maps are indeed common in both Clojure and Typed Clojure,
-meaning that our type system must accomodate them. Furthermore, they
+heterogeneous maps are indeed common in both Clojure and Typed Clojure,
+meaning that our type system must accommodate them. Furthermore, they
 are commonly used together, and the features of each are mutually
@@
 explicit in the type system is validated by the many Typed Clojure
-programs that use the ability to specify non-nullable types.
+programs that  specify non-nullable types.
 
@@
 wealth of static information to optimize programs. Addressing this
-requires not only first enabling sound gradual typing, but also
-integrating Typed Clojure into the Clojure tool chain more deeply, so
-that its information can be passed on to the compiler.
+requires not only  enabling sound gradual typing, but also
+integrating Typed Clojure into the Clojure tool so
+that its information can be communicated to the compiler.
 
@@
 of their programs for lack of expressiveness. This requires continued
-effort to analyze and understand the relevant features and idioms and
+effort to analyze and understand the features and idioms and
 develop new type checking approaches.

### 2015-07-10 related-work.tex
Fits.

--- 2015-02-28 related-work.tex
@@
 
-Optional types, requiring less implementation effort and avoiding any
-runtime cost, have been widely adopted in industry, including Hack, an
-extension to PHP~\cite{hack}, and Flow~\cite{flow} and
+Optional types, requiring less implementation effort and avoiding
+runtime cost, have been  adopted in industry, including Hack for PHP~\cite{hack}, and Flow~\cite{flow} and
 TypeScript~\cite{typescript}, two extensions of JavaScript. These
-systems all support some form of occurrence typing, but not in the
+systems  support  forms of occurrence typing, but not in the
 generality presented here, nor do they include the other features we
-have presented.
+present.
 

### 2015-07-10 overview.tex
Fix egregious formatting problems.

@@
 \begin{minted}{clj}
-  (.getParent (new File "a/b"))  ; :- (U nil Str)
+  ; :- (U nil Str)
+  (.getParent (new File "a/b"))  
   ;=> "a"
@@
 \begin{minted}{clojure}
-(ann hi [Kw -> Str]) ; interface type for `hi`
+(ann hi [Kw -> Str]) ; multimethod type
 (defmulti hi identity) ; dispatch function `identity`

### 2015-07-11 formal-model.tex
add GT citations. better update figure and section 3.6

--- 2015-07-10 formal-model.tex
@@
 Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
-numbers {\num{}},
+integer numbers {\nat{}},
 constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
@@
   \v{} &::=&          \singletonmeta{}
-                      \alt {\num{}}
+                      \alt {\nat{}}
                       \alt {\const{}}
@@
   \end{mathpar}
-  \caption{Core Subtyping rules}
+  \caption{Core subtyping rules}
   \label{main:figure:subtyping}
@@
 given as \figref{main:figure:othertypingrules}. We introduce
-the interesting rules with the complement number predicate
+the interesting rules with the complement natural number predicate
 as a running example.
@@
 uses 
-\constanttypeliteral{} to type constants.
-\numberhuh{} is a predicate over numbers, and
+\constanttypeliteral{} (\figref{main:figure:constanttyping})
+to type constants.
+\numberhuh{} is a predicate over the natural numbers, and
 \classconst{} returns its argument's class.
+
+\begin{figure}
   $$
@@
   $$
+  \caption{Constant typing}
+  \label{main:figure:constanttyping}
+\end{figure}
+
 The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.
@@
 
-\label{formalmodel:proofsystem}
-\subsection{Proof system}
-
-% checks obvious prop logic, but update is most interesting rule.
-
 \begin{figure*}
   $$
-\begin{array}{lllll}
-\updatefigure
+\begin{array}{llll}
+\updatefigure{}
 \end{array}
 $$
-\caption{Type Update}
+$$
+\begin{array}{lr}
+  \begin{array}{llll}
+    \restrictfigure{}
+  \end{array}
+  \begin{array}{llll}
+    \removefigure{}
+  \end{array}
+\end{array}
+$$
+\caption{Type update (the metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables), 
+  \notsubtypein{}{\Nil{}}{\nottype{\t{}}} when \issubtypein{}{\Nil{}}{\t{}})}
 \label{main:figure:update}
@@
 %  $$
-%\begin{array}{lllll}
+%\begin{array}{llll}
 %  \restrictremovefigure{}
@@
 %\end{figure}
+
+\subsection{Proof system}
+\label{formalmodel:proofsystem}
+
+\begin{figure}
+  $$
+\LUpdate{}
+$$
+%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
+  \caption{Proof system update rule (full rules \figref{appendix:figure:proofsystem})}
+  \label{main:figure:lupdate}
+\end{figure}
+
+The proof system for occurrence typing uses standard propositional logic,
+except for where nested information is combined. This is
+handled by the L-Update rule (\figref{main:figure:lupdate})---it says
+under \propenv{}, if object \path{\pathelemp{}}{\x{}} is of type \t{}, and 
+some extension
+\path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}
+is of possibly-negative type \propisnotmeta{}, then
+{\update{\t{}}{\propisnotmeta{}}{\pathelem{}}}
+is the new type of \path{\pathelemp{}}{\x{}} under \propenv{}.
+
+Recall \egref{example:desserts-on-meal}.
+%, resuming from
+%\secref{sec:coretypesystem}. 
+Solving
+$
+{ \inpropenv 
+  {{\isprop{\Order}{\makelocal{o}}},
+    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
+  {\isprop {\t{}} {\makelocal{o}}}}
+$
+uses L-Update, where \pathelem{} = {\emptypath{}} and \pathelemp{} = [{\keype{\makekw{Meal}}}].
+$$
+\inpropenv{\propenv{}}{\isprop{\update{\Order}{\Value{\makekw{combo}}}{[{\keype{\makekw{Meal}}}]}}{\makelocal{o}}}
+$$
+Since {\Order} is a union of HMaps, we structurally recur on the first case of \updateliteral{}
+(\figref{main:figure:update}),
+which preserves \pathelem{}.
+Each initial recursion hits the first HMap case, since there is some \t{} such that
+{\inmandatory{\k{}}{\t{}}{\mandatory{}}} and 
+\completenessmeta{} accepts partial maps \partial{}.
+
+To demonstrate,
+\makekw{lunch} meals hit the first HMap case and
+update to {\HMapp {\extendmandatoryset {\mandatory{}}{\Valkw{Meal}}{\sp{}}} {\emptyabsent{}}}
+where \sp{} = {\update{\Valkw{lunch}}{\Valkw{combo}}{\emptypath{}}}
+and \mandatory{} = \mandatoryset{\mandatoryentry{\Valkw{Meal}}{\Valkw{lunch}},{\mandatoryentry{\Valkw{desserts}}{\Number{}}}}.
+\sp{} updates to \Bot via the penultimate \updateliteral{} case,
+because \restrict{\Value{\makekw{lunch}}}{\Value{\makekw{combo}}} = \Bot
+by the first \restrictliteral{} case.
+The same happens to \makekw{dinner} meals,
+leaving just the \makekw{combo} HMap.
+
+In \egref{example:desserts-on-class},
+$
+\inpropenv{\propenv{}}{\isprop{\update{\Order}{\Long}{[{\classpe{}}, {\keype{\makekw{desserts}}}]}}{\makelocal{o}}}
+$
+updates the argument in the {\Long} method.
+This recurs and extra time for each meal to handle the {\classpe{}}
+path element.
+%$
+%{\update{\Number}{\Long}{[{\classpe{}}]}}}
+%$
+%
+%$
+%{\update{\Int}{\Long}{[{\classpe{}}]}}}
+%$
+%
+%

### 2015-07-11 overview.tex
add GT citations. better update figure and section 3.6

@@
 The following Typed Clojure program constructs a new \clj{File}
-instance and calls the \clj{getParent} method on the result.
+instance and calls the \clj{getParent} method on the result,
+returning a string \clj{"a"}, inferred as nullable.
 
@@
 \begin{minted}{clj}
+(.getParent (new File "a/b"))  
 ; :- (U nil Str)
-  (.getParent (new File "a/b"))  
 ;=> "a"
@@
 
+In the first method, the dispatch value is the concrete class \clj{Long}, a subtype
+of \clj{Int}, and the second \clj{nil}, the sentinel value for a failed map lookup.
+In practice, \clj{:lunch} and \clj{:dinner} meals will dispatch to the \clj{Long}
+method, but Typed Clojure infers a slightly more general type due to the definition
+of \clj{:combo} meals.
+
+\begin{exmp}
 \begin{minted}{clojure}
@@
   ; o :- (U '{:Meal ':lunch :desserts Int}
-  ;         '{:Meal ':dinner :desserts Int})
+  ;         '{:Meal ':dinner :desserts Int}
+  ;         '{:Meal ':combo :desserts Int
+  ;           :meal1 Order :meal2 Order})
   d)
@@
 \end{minted}
+\label{example:desserts-on-class}
+\end{exmp}
 %
@@
 %
-Knowing \clj{o} is also an
-\clj{Order}, we can eliminate the possibility of
-a \clj{:combo} meal, and deduce the type:
-\begin{minted}{clojure}
-(U '{:Meal ':lunch :desserts Int}
-   '{:Meal ':dinner :desserts Int})
-\end{minted}
-Looking up \clj{:desserts} on this type returns an \clj{Int}.
+%Knowing \clj{o} is also an
+%\clj{Order},
+In this method, we deduce
+the entry \clj{:desserts}
+\emph{must} be a present and mapped to an \clj{Int},
+even in a \clj{:combo} meal
+which does not state \clj{:desserts}
+as present or absent.
 
-In the \clj{nil} case, we know
-\begin{minted}{clojure}
-(isa? (class (:desserts o)) nil)
-\end{minted}
-is true---which implies \clj{(class (:desserts o))} is \clj{nil}.
-
+In the \clj{nil} method,
+\clj{(isa? (class (:desserts o)) nil)}
+must be true---which implies \clj{(class (:desserts o))} is \clj{nil}.
+%
 Since lookups on missing keys return \clj{nil}, either
 \begin{itemize}
-  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
-  \item \clj{o} is missing a \clj{:desserts} entry.
+  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
+  \item \clj{o} is missing a \clj{:desserts} entry, like \clj{{}}.
 \end{itemize}
@@
 \end{minted}
-Since \clj{o} is originally an \clj{Order}, this simplifies to a \clj{:combo} meal, 
-\begin{minted}{clojure}
-'{:Meal ':combo :meal1 Order :meal2 Order}
-\end{minted}
-thus allowing both recursive calls to type check.
+This eliminates non-\clj{:combo} meals
+since their \clj{'{:desserts Int}} type does not agree
+with this new information (because \clj{:desserts}
+is neither mapped to \clj{nil} or absent).
+
+%simplifies to a \clj{:combo} meal, 
+%\begin{minted}{clojure}
+%'{:Meal ':combo :meal1 Order :meal2 Order}
+%\end{minted}
+%thus allowing both recursive calls to type check.
 

### 2015-07-11 formal-model.tex
submit

@@
 Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
-integer numbers {\nat{}},
+integers {\nat{}},
 constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
@@
 
-\paragraph{Types} Types \s{} or \t{} include the top type, \emph{untagged} unions, functions, singleton types
-and class instances.
+\paragraph{Types} Types \s{} or \t{} 
+include the top type \Top,
+\emph{untagged} unions {\Unionsplice {\overrightarrow{\t{}}}}, 
+singletons ${\Value \singletonmeta{}}$,
+and class instances \class{}.
 We abbreviate the classes
@@
 {\ValueFalse} to {\False}.
-
+%
 The difference between the types
@@
 \appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
-
-Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions, object, and return type,
-which may refer to the function argument.
+%
+Function types 
+$
+{\ArrowOne {\x{}} {\s{}}
+             {\t{}}
+             {\filterset {\prop{}} {\prop{}}}
+             {\object{}}}
+$
+contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions \prop{}, object \object{}, and return type
+\t{},
+which may refer to the function argument \x{}.
 %Latent means they are relevant when the function is applied rather than evaluated.
-They are latent because they are instantiated with the
-actual object of the argument in applications before they are used in the proposition environment.
+They are instantiated with the
+actual object of the argument in applications. % before they are used in the proposition environment.
 
@@
   $42$ has the \emph{empty} object \emptyobject{}.
-
+%
 \figref{main:figure:termsyntax} gives the syntax for objects \object{}---non-empty objects 
@@
 
-\paragraph{Propositions}
+\paragraph{Propositions with a logical system}
 In standard type systems, \emph{type environments}
@@
 
-Occurrence typing instead uses \emph{logical formulas}
-and a \emph{proof system}.
+Occurrence typing instead pairs \emph{logical formulas},
+that can reason about arbitrary non-empty objects,
+with a \emph{proof system}.
 The logical statement {\isprop{\s{}}{\x{}}} says
@@
             {\letexp{\x{}}{\e{1}}{\e{2}}}
-            {\replacefor{\t{}}{\emptyobject{}}{\x{}}}
+            {\t{}}
            }
@@
 $$
-Types can now refer to variables, so we erase \x{} from the final type
-in T0-Let.
-We extend logical statements to reason about \emph{arbitrary} non-empty 
-objects {\isprop{\t{}}{\path{\pathelem{}}{\x{}}}}.
-The proof system judgment in T0-Local
-$$
+In T0-Local, 
+$
 { \inpropenv {\propenv{}} {\isprop {\t{}} {\path{\pathelem{}}{\x{}}}} }
-$$
-says under logical assumptions {\propenv{}}, object {\path{\pathelem{}}{\x{}}} is of type \t{}.
+$
+appeals to the proof system to solve \t{}.
+%says under logical assumptions {\propenv{}}, object {\path{\pathelem{}}{\x{}}} is of type \t{}.
 %We later define the more general T-Local.
@@
 disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions.
+%
+The full proof system judgement
+$$
+{ \inpropenv {\propenv{}} {\prop{}} }
+$$
+says \emph{proposition environment} {\propenv{}} proves proposition \prop{}.
 
@@
 For example, in \ifexp{\makelocal{o}}{\e{2}}{\e{3}}
-we learn {\makelocal{o}} is true in \e{2} via {\makelocal{o}}'s `then' proposition $\notprop{\falsy{}}{\makelocal{o}}$, and 
+we learn variable {\makelocal{o}} is true in \e{2} via {\makelocal{o}}'s `then' proposition $\notprop{\falsy{}}{\makelocal{o}}$, and 
 that {\makelocal{o}} is false in \e{3} via {\makelocal{o}}'s `else' proposition $\isprop{\falsy{}}{\makelocal{o}}$.
@@
 
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \SRefl{}
-
-    \STop{}
-
-\SUnionSuper{}
-
-\SUnionSub{}
-
-\SFunMono{}
-
-\SObject{}
-
-\SClass{}
-
-\begin{array}{l}
-\SSBool{}\\\\
-\SSKw{}
-\end{array}
-
-\SFun{}
-
-  \end{mathpar}
-  \caption{Core subtyping rules}
-  \label{main:figure:subtyping}
-\end{figure}
-
-\paragraph{Subtyping}
-\figref{main:figure:subtyping} presents subtyping
-as a reflexive and transitive relation with top type \Top. 
-Singleton types are instances of their respective classes---boolean singleton types
-are of type \Boolean{}, class literals are instances of \Class{} and keywords are
-instances of \Keyword{}.
-Instances of classes \class{} are subtypes of \Object{}. Since function types 
-are subtypes of \IFn{}, all types except for \Nil{} are subtypes of \Object{},
-so \Top{} = {\Union{\Nil}{\Object}}.
-Function subtyping is contravariant left of the arrow---latent propositions, object
-and the result type are covariant.
-Subtyping for untagged unions is standard.
-
 \paragraph{The full judgment}
-
-We formalize our type system following~\citet{TF10}.
-%(with differences highlighted in $\trdiff{\text{blue}}$)
-The typing judgment 
-$$
-{\judgementrewrite   {\propenv}
-              {\e{}} {\t{}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-  {\object{}}
-  {\ep{}}}
-$$
-says expression \e{} rewrites to \ep{}, which
-is of type \t{} in the 
-proposition environment $\propenv{}$, with 
-`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
-and object \object{}. We write 
-{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
-  or
-{\judgementselfrewrite   {\propenv}
-              {\e{}} {\t{}}
-  {\filterset {\thenprop {\prop{}}}
-              {\elseprop {\prop{}}}}
-  {\object{}}}
-for expressions that rewrite to themselves.
-
-\paragraph{Typing rules}
-The core typing rules are
-given as \figref{main:figure:othertypingrules}. We introduce
-the interesting rules with the complement natural number predicate
-as a running example.
-\begin{equation}
-\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
-\end{equation}
-%, including
-%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
-%as impossible (\botprop{}) `then' propositions.
-
-The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
-to check the body.
-With \propenv{} = \isprop{\Top}{\makelocal{d}},
-the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
-via the T-App rule, with three main steps.
-
-First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
-uses 
-\constanttypeliteral{} (\figref{main:figure:constanttyping})
-to type constants.
-\numberhuh{} is a predicate over the natural numbers, and
-\classconst{} returns its argument's class.
-
-\begin{figure}
-  $$
-\constanttypefigure{}
-  $$
-  \caption{Constant typing}
-  \label{main:figure:constanttyping}
-\end{figure}
-
-The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.
-
-Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
-in T-App the operand \ep{} = \makelocal{d} is checked with
-T-Local as
-\begin{equation}
-\judgementselfrewrite{\propenv{}}
-                     {\makelocal{d}}
-                     {\Top}
-                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
-                                {\isprop{\falsy}{\makelocal{d}}}}
-                     {\makelocal{d}}
-\end{equation}
-which encodes the type, proposition, and object information
-about variables
-as previously discussed.
-
-Finally, the T-App rule substitutes the operand's object \objectp{}
-for the parameter \x{} in the latent type, propositions, and object.
-\begin{equation}
-\judgementselfrewrite{\propenv{}}
-  {\appexp{\numberhuh{}}{\makelocal{d}}}
-  {\Boolean{}}
-  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
-             {\notprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-\end{equation}
-To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
-the latent `then' proposition of \constanttype{\numberhuh{}} with 
-\makelocal{d}, giving
-{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
-{\isprop{\Number{}}{\makelocal{d}}}.
-
-To check the branches of {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
-T-If
-introduces \thenprop{\prop{1}} = \isprop{\Number{}}{\makelocal{d}}
-to check \e{2} = {\false{}},
-and \elseprop{\prop{1}} = \notprop{\Number{}}{\makelocal{d}}
-to check 
-\e{3} = \true{}.
-%
-The branches are first checked with T-False and T-True respectively,
-then T-Subsume.
-%$$
-%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
-%  {\false{}}
-%  {\False{}}
-%  {\filterset{\botprop{}}
-%             {\topprop{}}}
-%  {\emptyobject{}}
-%$$
-$$
-\begin{array}{c}
-\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
-  {\false{}}
-  {\Boolean{}}
-  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-             {\isprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-  \\
-\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
-  {\true{}}
-  {\Boolean{}}
-  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-             {\isprop{\Number{}}{\makelocal{d}}}}
-  {\emptyobject{}}
-\end{array}
-$$
-%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
-%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
-%and \object{} = {\emptyobject{}}.
-%
-%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
-%\issubtypein{}{\t{}}{\tp{}}.
-%and 
-%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-%from 
-%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
-The T-Subsume premises
-\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
-and
-\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-allow us to pick compatible propositions for both branches.
-
-Finally T-Abs concludes, using the T-If outputs, with
-$$
-{\judgementselfrewrite{}{\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}}
-                    {\ArrowOne {\makelocal{d}} {\Top{}}
-                                      {\Boolean{}}
-                                      {\filterset {\notprop{\Number{}}{\makelocal{d}}}
-                                                  {\isprop{\Number{}}{\makelocal{d}}}}
-                                      {\emptyobject{}}}
-                    {\filterset {\topprop{}}
-                                {\botprop{}}}
-                    {\emptyobject{}}}
-$$
-
-%The object \x{} over latent propositions \thenprop{\prop{}} and
-%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
-%actual argument object---\objectp{}---is substituted in at application by T-App.
-
-%The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
-%The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
-%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
-%  {\Boolean{}}
-%  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
-%             {\notprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-
-%The T-If refines each branch based on the test---a
-%{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
-%for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
-%for checking the `else' branch.
-%%Information from each branch is combined via subsumption for the overall type, propositions, and object.
-%The T-Subsume rule
-
-%For example
-%$
-%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
-%  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
-%  {\Boolean{}}
-%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-%             {\isprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-%$
-%type checks because T-Subsume allows us to check both branches as
-%$
-%\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
-%  {\false{}}
-%  {\Boolean{}}
-%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-%             {\isprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-%$
-%and
-%$
-%\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
-%  {\true{}}
-%  {\Boolean{}}
-%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
-%             {\isprop{\Number{}}{\makelocal{d}}}}
-%  {\emptyobject{}}
-%$
-%respectively.
-%
-%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
-%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
-%we use T-Subsume on both branches
-%to introduce a logical disjunction
-%since at least one must be true if the entire expression is true.
-
-%Given a set of propositions, we can use logical reasoning to derive
-%new information about our programs
-%with the judgment \inpropenv{\propenv{}}{\prop{}}.
-%In addition to the standard rules for the logical connectives, the key
-%rule is L-Update, which combines multiple propositions about the same variable,
-%allowing us to refine its type.
-%$$
-%  {\LUpdate}
-%$$
-%For example, with L-Update we can use the knowledge of
-%\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
-%and 
-%\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
-%to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
-%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
-%We cover L-Update in more detail in \secref{sec:formalpaths}.
-%
-%Finally, this approach allows the type system to track
-%programming idioms from 
-%dynamic languages
-%using implicit type-based reasoning based on the result of
-%conditional tests.
-%For instance,
-%\egref{example:parent-if}
-%only utilizes \clj{f} once
-%the programmer is convinced it is safe to do so based whether
-%\clj{f}
-%is
-%true or false. 
-%To express this in the type system, every expression 
-%is described by two propositions: a `then' proposition
-%for when it reduces to a true value, and an `else' proposition
-%when it reduces to a false value---for \clj{f}
-%the then proposition is {\notprop{\falsy}{f}} and 
-%the else proposition is {\isprop{\falsy}{f}}.
-%\ref{main:figure:typingrules}
 
@@
 
+We formalize our type system following~\citet{TF10}.
+%(with differences highlighted in $\trdiff{\text{blue}}$)
+The typing judgment 
+$$
+{\judgementrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}
+  {\ep{}}}
+$$
+says expression \e{} rewrites to \ep{}, which
+is of type \t{} in the 
+proposition environment $\propenv{}$, with 
+`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
+and object \object{}.
+
+We write 
+{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} 
+to mean 
+{\judgementrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\propp{}}}
+              {\elseprop {\propp{}}}}
+  {\objectp{}}
+  {\ep{}}}
+for some {\thenprop {\propp{}}}, {\elseprop {\propp{}}}
+and {\objectp{}},
+and
+  abbreviate self rewriting judgements
+{\judgementrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}
+  {\e{}}}
+to
+{\judgementselfrewrite   {\propenv}
+              {\e{}} {\t{}}
+  {\filterset {\thenprop {\prop{}}}
+              {\elseprop {\prop{}}}}
+  {\object{}}}.
+
+\paragraph{Typing rules}
+
+The core typing rules are
+given as \figref{main:figure:othertypingrules}. We introduce
+the interesting rules with the complement number predicate
+as a running example.
+\begin{equation}
+\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+\end{equation}
+%, including
+%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
+%as impossible (\botprop{}) `then' propositions.
+
+The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
+to check the body.
+With \propenv{} = \isprop{\Top}{\makelocal{d}},
+the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
+via the T-App rule, with three main steps.
+
+First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
+uses 
+\constanttypeliteral{} (\figref{main:figure:constanttyping}, dynamic semantics in the supplemental material)
+to type constants.
+\numberhuh{} is a predicate over numbers, and
+\classconst{} returns its argument's class.
+
+\begin{figure}
+  $$
+\constanttypefigure{}
+  $$
+  \caption{Constant typing}
+  \label{main:figure:constanttyping}
+\end{figure}
+
+Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
+in T-App the operand \ep{} = \makelocal{d} is checked with
+T-Local as
+\begin{equation}
+\judgementselfrewrite{\propenv{}}
+                     {\makelocal{d}}
+                     {\Top}
+                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
+                                {\isprop{\falsy}{\makelocal{d}}}}
+                     {\makelocal{d}}
+\end{equation}
+which encodes the type, proposition, and object information
+about variables
+as previously discussed.
+
+Finally, the T-App rule substitutes the operand's object \objectp{}
+for the parameter \x{} in the latent type, propositions, and object.
+\begin{equation}
+\judgementselfrewrite{\propenv{}}
+  {\appexp{\numberhuh{}}{\makelocal{d}}}
+  {\Boolean{}}
+  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+             {\notprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+\end{equation}
+To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
+the latent `then' proposition of \constanttype{\numberhuh{}} with 
+\makelocal{d}, giving
+{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
+{\isprop{\Number{}}{\makelocal{d}}}.
+
+To check the branches of {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
+T-If
+introduces \thenprop{\prop{1}} = \isprop{\Number{}}{\makelocal{d}}
+to check \e{2} = {\false{}},
+and \elseprop{\prop{1}} = \notprop{\Number{}}{\makelocal{d}}
+to check 
+\e{3} = \true{}.
+%
+The branches are first checked with T-False and T-True respectively,
+then T-Subsume.
+%$$
+%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
+%  {\false{}}
+%  {\False{}}
+%  {\filterset{\botprop{}}
+%             {\topprop{}}}
+%  {\emptyobject{}}
+%$$
+$$
+\begin{array}{c}
+\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
+  {\false{}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+  \\
+\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
+  {\true{}}
+  {\Boolean{}}
+  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+             {\isprop{\Number{}}{\makelocal{d}}}}
+  {\emptyobject{}}
+\end{array}
+$$
+%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
+%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
+%and \object{} = {\emptyobject{}}.
+%
+%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
+%\issubtypein{}{\t{}}{\tp{}}.
+%and 
+%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+%from 
+%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
+The T-Subsume premises
+\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
+and
+\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+allow us to pick compatible propositions for both branches.
+
+Finally T-Abs concludes, using the T-If outputs, with
+$$
+{\judgementselfrewrite{}{\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}}
+                    {\ArrowOne {\makelocal{d}} {\Top{}}
+                                      {\Boolean{}}
+                                      {\filterset {\notprop{\Number{}}{\makelocal{d}}}
+                                                  {\isprop{\Number{}}{\makelocal{d}}}}
+                                      {\emptyobject{}}}
+                    {\filterset {\topprop{}}
+                                {\botprop{}}}
+                    {\emptyobject{}}}
+$$
+
+%The object \x{} over latent propositions \thenprop{\prop{}} and
+%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
+%actual argument object---\objectp{}---is substituted in at application by T-App.
+
+%The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
+%The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
+%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
+%  {\Boolean{}}
+%  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
+%             {\notprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+
+%The T-If refines each branch based on the test---a
+%{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
+%for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
+%for checking the `else' branch.
+%%Information from each branch is combined via subsumption for the overall type, propositions, and object.
+%The T-Subsume rule
+
+%For example
+%$
+%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
+%  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%type checks because T-Subsume allows us to check both branches as
+%$
+%\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
+%  {\false{}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%and
+%$
+%\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
+%  {\true{}}
+%  {\Boolean{}}
+%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
+%             {\isprop{\Number{}}{\makelocal{d}}}}
+%  {\emptyobject{}}
+%$
+%respectively.
+%
+%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
+%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
+%we use T-Subsume on both branches
+%to introduce a logical disjunction
+%since at least one must be true if the entire expression is true.
+
+%Given a set of propositions, we can use logical reasoning to derive
+%new information about our programs
+%with the judgment \inpropenv{\propenv{}}{\prop{}}.
+%In addition to the standard rules for the logical connectives, the key
+%rule is L-Update, which combines multiple propositions about the same variable,
+%allowing us to refine its type.
+%$$
+%  {\LUpdate}
+%$$
+%For example, with L-Update we can use the knowledge of
+%\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
+%and 
+%\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
+%to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
+%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
+%We cover L-Update in more detail in \secref{sec:formalpaths}.
+%
+%Finally, this approach allows the type system to track
+%programming idioms from 
+%dynamic languages
+%using implicit type-based reasoning based on the result of
+%conditional tests.
+%For instance,
+%\egref{example:parent-if}
+%only utilizes \clj{f} once
+%the programmer is convinced it is safe to do so based whether
+%\clj{f}
+%is
+%true or false. 
+%To express this in the type system, every expression 
+%is described by two propositions: a `then' proposition
+%for when it reduces to a true value, and an `else' proposition
+%when it reduces to a false value---for \clj{f}
+%the then proposition is {\notprop{\falsy}{f}} and 
+%the else proposition is {\isprop{\falsy}{f}}.
+%\ref{main:figure:typingrules}
+
 %\figref{main:figure:typingrules} contains the core typing rules.
@@
 %$$
+
+\begin{figure}
+  \footnotesize
+  \begin{mathpar}
+    \SRefl{}
+
+    \STop{}
+
+\SUnionSuper{}
+
+\SUnionSub{}
+
+\SFunMono{}
+
+\SObject{}
+
+\SClass{}
+
+\begin{array}{l}
+\SSBool{}\\\\
+\SSKw{}
+\end{array}
+
+\SFun{}
+
+  \end{mathpar}
+  \caption{Core subtyping rules}
+  \label{main:figure:subtyping}
+\end{figure}
+
+\paragraph{Subtyping}
+\figref{main:figure:subtyping} presents subtyping
+as a reflexive and transitive relation with top type \Top. 
+Singleton types are instances of their respective classes---boolean singleton types
+are of type \Boolean{}, class literals are instances of \Class{} and keywords are
+instances of \Keyword{}.
+Instances of classes \class{} are subtypes of \Object{}. Function types 
+are subtypes of \IFn{}. All types except for \Nil{} are subtypes of \Object{},
+so \Top{} is similar to {\Union{\Nil}{\Object}}.
+Function subtyping is contravariant left of the arrow---latent propositions, object
+and result type are covariant.
+Subtyping for untagged unions is standard.
 
@@
   \begin{mathpar}
+    {\TNew}
+
+    {\TMethod}
+
     {\TField}
-
-    %{\TMethod}
 
@@
 
-\egref{example:getparent-direct-constructor} translates to expression~\ref{eq:unresolved}.
-$$
-\clj{(.getParent (new File "a/b"))}
-$$
+\egref{example:getparent-direct-constructor} \clj{(.getParent (new File "a/b"))}
+translates to
 \begin{equation}  \label{eq:unresolved}
@@
 The T-New typing rule checks and rewrites constructors.
-$$
-    {\TNew}
-$$
+%$$
+%    {\TNew}
+%$$
 To check
@@
 {\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
-The T-Method rule checks unresolved methods.
-$$
-    {\TMethod}
-$$
-To check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}},
-we verify the target type \s{} = \File is non-nil before erasing \Nil
+%The T-Method rule checks unresolved methods.
+%$$
+%    {\TMethod}
+%$$
+We use T-Method to check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}, 
+which checks unresolved methods.
+We verify the target type \s{} = \File is non-nil before erasing \Nil
 by converting to a Java type \classhint{1} = \File.
@@
 We now consider the \isaliteral{} operation, a core part of the dispatch mechanism for multimethods. 
-Recalling the examples in \secref{sec:multioverview},
+Recalling the \secref{sec:multioverview} examples,
 \isaliteral{} is
@@
 %---we do not model the semantics for vectors
-
-The key component of the T-IsA rule is the \isacompareliteral{} 
-metafunction
-(\figref{main:figure:mmsyntax}), used to calculate the propositions for
+%
+The key component of the T-IsA rule (\figref{main:figure:mmsyntax})
+is the \isacompareliteral{} 
+metafunction, used to calculate the propositions for
 \isaliteral{} tests.
-\begin{mathpar}
-  \TIsA{}
-\end{mathpar}
+%\begin{mathpar}
+%  \TIsA{}
+%\end{mathpar}
 
@@
 
-\constantsemfigure{main}
+%\constantsemfigure{main}
 
@@
   \begin{mathpar}
-    %\TDefMulti{}
-
-    %\TDefMethod{}
+    \TDefMulti{}
+
+    \TDefMethod{}
+
+    \TIsA{}
   \end{mathpar}
@@
 
-We use \emph{immutable}
+\figref{main:figure:mmsyntax}
+presents \emph{immutable}
 multimethods without default methods to ease presentation.
@@
 Below is the translation of \egref{example:hi-multimethod} to \lambdatc{}.
-\begin{minted}{clojure}
-(ann hi [Kw -> Str])
-(defmulti hi identity)
-(defmethod hi :en [_] "hello")
-(defmethod hi :fr [_] "bonjour")
-(hi :en) ;=> "hello"
-\end{minted}
+%\begin{minted}{clojure}
+%(ann hi [Kw -> Str])
+%(defmulti hi identity)
+%(defmethod hi :en [_] "hello")
+%(defmethod hi :fr [_] "bonjour")
+%(hi :en) ;=> "hello"
+%\end{minted}
+
 \begin{listing}
@@
 
-\paragraph{How to check}
-We now walk through type checking the expression
-\begin{equation}  \label{eq:hi-defmulti}
-{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}
-\end{equation}
-
+%\paragraph{How to check}
+We now show to check
+{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}.
+%
 The expression {\createmultiexp {\s{}} {\e{}}} creates a multimethod with \emph{interface type} \s{}, and dispatch function \e{}
@@
 {\MultiFntype {\s{}} {\sp{}}}. % with interface type {\s{}} and dispatch function type {\sp{}}.
-The typing rule for \defmultiliteral{} simply checks the dispatch function, and
+The T-DefMulti typing rule simply checks the dispatch function, and
 verifies both the interface and dispatch type's domain agree.
-$$
-    \TDefMulti{}
-$$
-Expression~\ref{eq:hi-defmulti} checks with \t{} = \Keyword, interface type \s{} = {\ArrowTwo{\x{}}{\Keyword}{\String}},
+%$$
+%    \TDefMulti{}
+%$$
+Our example checks with \t{} = \Keyword, interface type \s{} = {\ArrowTwo{\x{}}{\Keyword}{\String}},
 dispatch function type \sp{} = {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}, and overall type
-$$
+$
 {\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
               {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
-$$
-
-Next, we walk through checking
-\begin{equation}  \label{eq:hi-defmethoden}
+$.
+
+Next, show how to check
+$
 {\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
-\end{equation}
-
+$.
+%
 The expression 
@@
 that extends multimethod \e{m}'s dispatch table, mapping dispatch value
-\e{v} to method \e{f}. The typing rule for \defmethodliteral{}
+\e{v} to method \e{f}. The T-DefMulti typing rule
 checks \e{m} is a multimethod with dispatch function type \t{d},
@@
 body.
-$$
-    \TDefMethod{}
-$$
-Expression~\ref{eq:hi-defmethoden} checks with \e{m} being of type
-$$
+%$$
+%    \TDefMethod{}
+%$$
+Our example checks with \e{m} being of type
+$
 {\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
               {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
-$$
+$
 with \objectp{} = {\x{}} and \t{v} = \Value{\makekw{en}}. Then {\thenprop{\proppp{}}} = 
 {\isprop {\Value{\makekw{en}}}{\x{}}} by
-$$
+$
 \isacompare{}{\x{}}{\Value{\makekw{en}}}
 {\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
-$$
+$.
 Since \t{} = \Keyword{}, we check the method body with
-$$
+$
 \judgement{{\isprop{\Keyword}{\x{}}},{\isprop {\Value{\makekw{en}}}{\x{}}}}
@@
   {\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
-$$
+$.
 Finally from the interface type \t{m}, we know \thenprop{\prop{}} = \elseprop{\prop{}} = \topprop{},
@@
 \paragraph{Subtyping}
-Multimethods are also functions, which is encoded via the following subtyping rule.
-$$
-\SPMultiFn{}
-$$
+Multimethods are also functions, which is encoded via S-PMultiFn.
+%$$
+%\SPMultiFn{}
+%$$
 This rule says a multimethod can be upcast to its interface type. This means
@@
 
-\figref{main:figure:hmapsyntax} presents syntax, typing rules
-and dynamic semantics in detail.
-%
+\figref{main:figure:hmapsyntax}
+presents
+heterogeneous map types.
+The partially specified map of
+\clj{lunch} in \egref{example:lunchpartial}
+is written
+$$
+\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}
+$$
+We abbreviate this type as \Lunch in this section.
 The type \HMapgeneric{\mandatory{}}{\absent{}}
-includes {\mandatory{}}, a map of \emph{present} entries (mapping keywords to types),
+contains {\mandatory{}}, a map of \emph{present} entries (mapping keywords to types),
 \absent{}, a set of keyword keys that are known to be \emph{absent}
@@
 and {\partial{}} (``partial'') if there are \emph{unknown} entries.
+
+The type of the fully specified map
+\clj{breakfast} in \egref{example:breakfastcomplete} elides the absent entries,
+written
+$$
+\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}
+$$
+We abbreviate this type as \Breakfast in this section.
 To ease presentation, 
@@
 
-The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
-we only model the former to avoid the added complexity of keywords being functions.
-To simplify presentation, we only provide syntax for the empty map literal and
-restrict lookup and extension to keyword keys. The metavariable \mapval{}
+The metavariable \mapval{}
 ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
 usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.
-
+We %do not model keywords as functions,
+only provide syntax for the empty map literal,
+however when convenient we abbreviate non-empty map literals
+to be a series of \assocliteral{} operations on the empty map.
+We restrict lookup and extension to keyword keys.
+
+\paragraph{How to check}
+A mandatory lookup is checked by T-GetHMap.
+$$
+\abs{\makelocal{b}}{\Breakfast}{\getexp{\makelocal{b}}{\makekw{en}}}
+$$
+The result type is \String, and the return object is \path{\keype{\makekw{en}}}{\makelocal{b}}.
+The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
+is a symbolic representation for a keyword lookup of $k$ in \object{}.
+The substitution for {\x{}} handles the case where \object{} is empty.
+$$
+\begin{array}{rcl}
+{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
+{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
+\end{array}
+$$
+
+An absent lookup is checked by T-GetHMapAbsent.
+$$
+\abs{\makelocal{b}}{\Breakfast}{\getexp{\makelocal{b}}{\makekw{bocce}}}
+$$
+The result type is \Nil, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{b}}.
+
+A lookup of a key that is neither present or absent is checked by
+T-GetHMapPartialDefault.
+$$
+\abs{\makelocal{u}}{\Lunch}{\getexp{\makelocal{u}}{\makekw{bocce}}}
+$$
+The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
+
+For presentational reasons, lookups on unions of HMaps are only supported in T-GetHMap. 
+Furthermore, each element of the union must contain the key we are looking up.
+$$
+\abs{\makelocal{u}}{\Unionsplice{\Breakfast \Lunch}}{\getexp{\makelocal{u}}{\makekw{en}}}
+$$
+The result type is \String, and the return object is \path{\keype{\makekw{en}}}{\makelocal{u}}.
+However, lookups on \makekw{bocce} on {\Unionsplice{\Breakfast \Lunch}} maps are unsupported.
+This restriction still allows us to check many of the examples in \secref{sec:overview}---in
+particular we can check 
+\egref{example:desserts-on-meal}, as \makekw{Meal} is in common with both HMaps,
+but cannot check \egref{example:desserts-on-class}
+because a \makekw{combo} meal lacks a \makekw{desserts} entry.
+
+Extending a map with T-Assoc preserves its completeness.
+$$
+\abs{\makelocal{b}}{\Breakfast}{\assocexp{\makelocal{b}}{\makekw{au}}{\makestr{beans}}}
+$$
+The result type is
+$
+\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}
+        ,{\mandatoryentrynoarrow{\Valkw{au}}{\String}}}}
+$.
+
+%for cases like \egref{example:desserts-on-meal}
+%where every element in the union
+%contains the key we are looking up.
+
+\paragraph{Subtyping}
 Subtyping for HMaps
@@
 
-The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
-that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
-and T-GetHMapPartialDefault a lookup with unknown results.
-Lookups on unions of HMaps are only supported in T-GetHMap, 
-for cases like \egref{example:desserts-on-meal}
-where every element in the union
-contains the key we are looking up.
-The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
-is a symbolic representation for a keyword lookup of $k$ in \object{}.
-The substitution for {\x{}} handles the case where \object{} is empty.
-$$
-\begin{array}{rcl}
-{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
-{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
-\end{array}
-$$
+%The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
+%that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
+%and T-GetHMapPartialDefault a lookup with unknown results.
 

### 2015-07-11 overview.tex
submit

@@
 \end{minted}
-
-The function may take strings, but providing \clj{nil} (Clojure's name for Java's \java{null})
-is a static type error --- \clj{nil} is not a string in Typed Clojure.
-
+%
+Strings are accepted, but providing \clj{nil} (Clojure's name for Java's \java{null})
+is a static type error---\clj{nil} is not a string in Typed Clojure.
+%
 \begin{minted}{clj}
@@
 
-\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{thf10} to
+\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{TF10} to
 model type-based control flow.
@@
 is never passed a \clj{nil} argument.
-
+%
 \begin{minted}{clj}
@@
 \end{minted}
-
-Removing the \clj{when} expression is a static type error
---- \clj{repeat} cannot be passed \clj{nil}.
-
+%
+Removing the \clj{when} expression is a static type error---\clj{repeat} 
+cannot be passed \clj{nil}.
+%
 \begin{minted}{clj}
@@
 Typed Clojure rejects programs that assume otherwise.
-
+%
 \begin{minted}{clj}
@@
 In contrast, JVM invariants guarantee constructors return a non-null reference.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
-
+%
 \begin{exmp}
@@
 a \emph{dispatch value}. A method is then chosen
-based on the dispatch value---this method is applied to the arguments
+based on the dispatch value---the arguments are then passed to the method
 to finally return a value for the entire expression.
-
+%
 \begin{minted}{clojure}
@@
 \end{minted}
-
+%
 For example, 
@@
 relationships.
-
+%
 Recall the multimethod from figure~\ref{fig:ex1}, reproduced here.
@@
 \end{minted}
-
+%
 The current dispatch value and---in turn---each method's associated dispatch value
@@
 Hash-maps are accessed with the \clj{get} function:
+\begin{exmp}
 \begin{minted}{clojure}
-(def breakfast {:en "waffles" :fr "croissants"})
+(def breakfast 
+  {:en "waffles" :fr "croissants"})
 
@@
 \end{minted}
+\label{example:breakfastcomplete}
+\end{exmp}
 Additionally, keywords are functions that look themselves up in a
@@
 \clj{:mandatory} keys.
-
+%
+\begin{exmp}
 \begin{minted}{clojure}
@@
 \end{minted}
+\label{example:lunchpartial}
+\end{exmp}
 
@@
 %\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
-Using fully specified HMaps
-statically enforce the private key is not accidentally left in a supposedly
-encrypted keypair.
-
+If we forget to remove the unencrypted private key,
+a type error is given, because \clj{EncKeyPair} is fully specified.
+%
 \begin{minted}{clojure}
@@
           :Meal)  ; dispatch on :Meal entry
-; map destructuring backwards, d is :desserts entry
+; destructuring reads backwards, d is :desserts entry
 (defmethod desserts :lunch [{d :desserts}] d)
@@
 
-In the first method, the dispatch value is the concrete class \clj{Long}, a subtype
-of \clj{Int}, and the second \clj{nil}, the sentinel value for a failed map lookup.
+In the first method, the dispatch value is the class \clj{Long}, a subtype
+of \clj{Int}, and the second method has dispatch value \clj{nil}, the sentinel value for a failed map lookup.
 In practice, \clj{:lunch} and \clj{:dinner} meals will dispatch to the \clj{Long}
@@
 the entry \clj{:desserts}
-\emph{must} be a present and mapped to an \clj{Int},
+\emph{must} be a present and mapped to a \clj{Long},
 even in a \clj{:combo} meal

### 2015-07-11 formal-model.tex
final POPL submission

@@
 The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
+Notice the propositions attached to each \getliteral{} rule are trivial---propositions
+are erased once they enter a HMap type.
 
@@
 
-Extending a map with T-Assoc preserves its completeness.
+Extending a map with T-AssocHMap preserves its completeness.
 $$
@@
         ,{\mandatoryentrynoarrow{\Valkw{au}}{\String}}}}
-$.
+$,
+which is complete, like its input.
+T-AssocHMap also enforces ${\k{}} \not\in {\absent{}}$ to prevent badly formed types.
 
@@
 %next section discusses this in detail.
-Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
-and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
-formed types.
+%Finally T-AssocHMap extends an HMap with a mandatory entry while preserving completeness
+%and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
+%formed types.
 
@@
 under \propenv{}, if object \path{\pathelemp{}}{\x{}} is of type \t{}, and 
-some extension
+an extension
 \path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}
@@
 {\update{\t{}}{\propisnotmeta{}}{\pathelem{}}}
-is the new type of \path{\pathelemp{}}{\x{}} under \propenv{}.
+is \path{\pathelemp{}}{\x{}}'s type under \propenv{}.
 
@@
 updates the argument in the {\Long} method.
-This recurs and extra time for each meal to handle the {\classpe{}}
+This recurs twice for each meal to handle the {\classpe{}}
 path element.
+
+We describe the other \updateliteral{} cases.
+The first \classpe{} case updates
+to \class{} if \classconst{} returns \Value{\class{}}.
+The second \keype{\k{}} case detects contradictions in absent
+keys. % not overlapping with \Nil{}.
+The third \keype{\k{}} case updates unknown entries to be mapped to \Nil{} or absent.
+The fourth \keype{\k{}} case updates unknown entries to be \emph{present}
+when they do no overlap with \Nil{}.
+
 %$

### 2015-07-11 formal-model.tex
some typos, submitted

@@
 keys. % not overlapping with \Nil{}.
-The third \keype{\k{}} case updates unknown entries to be mapped to \Nil{} or absent.
+The third \keype{\k{}} case updates unknown entries to be mapped to \t{} or absent.
 The fourth \keype{\k{}} case updates unknown entries to be \emph{present}
-when they do no overlap with \Nil{}.
+when they do not overlap with \Nil{}.
 

### 2015-10-08 formal-model.tex
typo
convert to ESOP format

@@
 
-We formalize our type system following~\citet{TF10}.
+We formalize our type system following~\cite{TF10}.
 %(with differences highlighted in $\trdiff{\text{blue}}$)
@@
 \paragraph{Operational semantics} We define the dynamic semantics for \lambdatc{}
-in a big-step style using an environment, following~\citet{TF10}.
+in a big-step style using an environment, following~\cite{TF10}.
 We include both errors and a \wrong{} value, which is provably ruled out by the
@@
 
-\begin{figure}[t]
+\begin{figure}%[t]
   \footnotesize
@@
 %written \path{\pathelem{}}{\x{}}.
-%\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
+%\cite{TF10} introduce the path elements \carpe{} and \cdrpe{}
 %to reason about selector operations on cons cells.
@@
 
-\begin{figure}[t!]
+\begin{figure}%[t!]
   \footnotesize
@@
 The result type is \String, and the return object is \path{\keype{\makekw{en}}}{\makelocal{u}}.
-However, lookups on \makekw{bocce} on {\Unionsplice{\Breakfast \Lunch}} maps are unsupported.
+However, lookups of \makekw{bocce} on {\Unionsplice{\Breakfast \Lunch}} maps are unsupported.
 This restriction still allows us to check many of the examples in \secref{sec:overview}---in

### 2015-10-08 metatheory.tex
convert to ESOP format

@@
 
-We prove type soundness following \citet{TF10}.  Our model is extended
+We prove type soundness following \cite{TF10}.  Our model is extended
 to include errors and a \wrong{} value, and we prove well-typed
 programs do not go wrong; this is therefore a stronger theorem than
-proved by \citet{TF10}.
+proved by \cite{TF10}.
 

### 2015-10-08 related-work.tex
convert to ESOP format

@@
 \paragraph{Multimethods} 
-\citeauthor{MS02} and collaborators present a sequence of
+\cite{MS02} and collaborators present a sequence of
 systems~\cite{Chambers:1992:OMC, Chambers:1994:TMM, MS02} with statically-typed multimethods

### 2015-10-13 formal-model.tex
fix esop formatting

@@
                       \alt {\comb {\e{}} {\e{}}} 
-                      \alt {\abs {\x{}} {\t{}} {\e{}}} &\mbox{Expressions} \\
-                      &\alt& {\ifexp {\e{}} {\e{}} {\e{}}}
+                      \alt {\abs {\x{}} {\t{}} {\e{}}} 
+                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                       %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
-                      \alt {\letexp {\x{}} {\e{}} {\e{}}}\\
+                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                       %\alt {\errorvalv{}}
+                      &\mbox{Expressions} \\
   \v{} &::=&          \singletonmeta{}
@@
                                    {\object{}}}
+                      \alt {\Value \singletonmeta{}} 
+                      \alt \trdiff{\class{}}
                 &\mbox{Types} \\
-                      &\alt& {\Value \singletonmeta{}} 
-                      \alt \trdiff{\class{}}\\
   \singletonallsyntax{}
@@
 \begin{figure*}
-$$
 \footnotesize
@@
     %\end{array}
-  \begin{array}{c}
-    {\TLocal} \ \ \  \ \ \ 
-    {\TIf}\ \ \  \ \ \ \ \ 
-    {\TAbs} \ \ \ \ 
-    \begin{array}{c}
-      {\TNil}\\\\
+  \begin{mathpar}
+        {\TLocal}
+
+        {\TIf}
+
+        {\TAbs}\\
+
+        {\TNil}
+
     {\TFalse}
-    \end{array}
-    \\\\
-    {\TLet} \ \ \  \ \ \ \ \ \ 
+
+    {\TLet}\\
+
     {\TApp}
-    \\\\
-    \begin{array}{c}
+
+    {\TNum}\\
+
+    {\TConst}
+
+    {\TKw}
+
     {\TNum}\\\\
-      {\TConst}
-    \end{array} \ \ \ \
-    \begin{array}{c}
-      {\TKw}\\\\
-      {\TNum}\\\\
+
     {\TClass}
-    \end{array} \ \ \ \
-    \begin{array}{c}
+
     {\TSubsume}
-    \end{array} \ \ \ \ 
-    \begin{array}{c}
-      {\TStr}\\\\
+
+    {\TStr}\\
+
     {\TTrue}
-    \end{array}
-  \end{array}
+  \end{mathpar}
     %\begin{array}{c}
@@
     %\end{array}
-$$
   \caption{Core typing rules}
@@
 \begin{altgrammar}
-  \e{} &::=& \ldots \alt {\createmultiexp {\t{}} {\e{}}} &\mbox{Expressions} \\
-             &\alt& {\extendmultiexp {\e{}} {\e{}} {\e{}}}
-             \alt {\isaapp {\e{}} {\e{}}}\\
+  \e{} &::=& \ldots \alt {\createmultiexp {\t{}} {\e{}}} \alt
+             {\extendmultiexp {\e{}} {\e{}} {\e{}}}
+             \alt {\isaapp {\e{}} {\e{}}} &\mbox{Expressions} \\
   \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
@@
 $$
-    \BBetaMulti{}
+\footnotesize \BBetaMulti{}
 $$
@@
 is written
-$$
-\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}
-$$
+\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}.
 We abbreviate this type as \Lunch in this section.
@@
 written
-$$
-\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}
-$$
+\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}.
 We abbreviate this type as \Breakfast in this section.
@@
 
-\begin{figure}
-  $$
-\LUpdate{}
-$$
-%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
-  \caption{Proof system update rule (full rules \figref{appendix:figure:proofsystem})}
-  \label{main:figure:lupdate}
-\end{figure}
-
 The proof system for occurrence typing uses standard propositional logic,
 except for where nested information is combined. This is
-handled by the L-Update rule (\figref{main:figure:lupdate})---it says
+handled by the L-Update rule.
+  $$
+\LUpdate{}
+$$
+It says
 under \propenv{}, if object \path{\pathelemp{}}{\x{}} is of type \t{}, and 

### 2015-10-13 overview.tex
convert to ESOP format
fix esop formatting

@@
 (ann  greet [Str -> Str])
-(defn greet [n]
-  (str "Hello, " n "!"))
+(defn greet [n] (str "Hello, " n "!"))
 (greet "Grace") ; :- Str
@@
 \begin{minted}{clj}
-(greet nil) ; Type Error:
-            ;   Expected Str, given nil
+(greet nil) ; Type Error: Expected Str, given nil
 \end{minted}
@@
 (ann  greet-nil [(U nil Str) -> Str])
-(defn greet-nil [n]
-  (str "Hello" (when n (str ", " n)) "!"))
+(defn greet-nil [n] (str "Hello" (when n (str ", " n)) "!"))
 (greet-nil "Donald") ; :- Str
@@
 (defn greetings [n i]
-  (str "Hello, "
-       (when i  ; when i is a non-nil integer
+  (str "Hello, " (when i  ; when i is a non-nil integer
                    (apply str (repeat i "hello, ")))
@@
 (defn greetings-bad [n i]
-  (str "Hello, "
-       (apply str 
-         (repeat 
-           i ; Type Error:
-             ;   Expected Int, given (U nil Int).
+  (str "Hello, " (apply str 
+                   (repeat i ; Type Error: Expected Int, given (U nil Int).
                            "hello, "))
@@
 %
-%\citet{TF10}
+%\cite{TF10}
 %presented Typed Racket with occurrence typing,
@@
 %
-%Typed Clojure can already check all of the examples in~\citet{TF10}---the 
+%Typed Clojure can already check all of the examples in~\cite{TF10}---the 
 %rest of this section describes the extensions necessary
@@
 \begin{minted}{clj}
-(.getParent (new File "a/b"))  
-; :- (U nil Str)
+(.getParent (new File "a/b"))  ; :- (U nil Str)
 ;=> "a"
@@
 \begin{minted}{clj}
-(fn [f]           ; Type Error:                    
-  (.getParent f)) ;   Unresolved interop: getParent
+(fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
 
@@
 (ann parent [(U nil File) -> (U nil Str)])
-(defn parent [f]
-  (if f (.getParent f) nil))
+(defn parent [f] (if f (.getParent f) nil))
 \end{minted}
@@
 (defn parent-bad-in [f :- (U nil File)]
-  (.getParent f)) ; Type Error:
-                  ;   Cannot call instance method 
-                  ;   java.io.File/getParent on type 
-                  ;   (U nil File).
+  (.getParent f)) ; Type Error: Cannot call instance method 
+                  ; java.io.File/getParent on type (U nil File).
 \end{minted}
@@
 (defn parent-bad-out [f :- File] :- Str
-  (.getParent f)) ; Type Error:
-                  ;   Expected Str, given (U nil Str).
+  (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
 \end{minted}
@@
 \begin{minted}{clojure}
-(def breakfast 
-  {:en "waffles" :fr "croissants"})
-
+(def breakfast {:en "waffles" :fr "croissants"})
 (get breakfast :en)    ; :- Str
@@
 \begin{minted}{clojure}
-breakfast ; :- (HMap :mandatory {:en Str :fr Str}
-          ;          :complete? true)
+breakfast ; :- (HMap :mandatory {:en Str, :fr Str}, :complete? true)
 \end{minted}
@@
 ;=> "baguette"
-; Unknown lookups are now less accurate
-(:bocce lunch) ; :- Any
+(:bocce lunch) ; :- Any     ; Unknown lookups are now less accurate
 ;=> nil
@@
 \begin{minted}{clojure}
-(defalias RawKeyPair
-  "Unencrypted keypair -- extra keys disallowed"
-  (HMap :mandatory {:public-key RawKey,
-                    :private-key RawKey},
-        :complete? true))
+(defalias RawKeyPair ; extra keys disallowed
+  (HMap :mandatory {:pub RawKey, :priv RawKey}, :complete? true))
 
-(defalias EncKeyPair
-  "Encrypted keypair -- extra keys disallowed"
-  (HMap :mandatory {:public-key RawKey,
-                    :enc-private-key EncKey},
-        :complete? true))
+(defalias EncKeyPair ; extra keys disallowed
+  (HMap :mandatory {:pub RawKey, :enc-priv EncKey}, :complete? true))
 
 (ann enc-keypair [RawKeyPair -> EncKeyPair])
-(defn enc-keypair "Encrypt an unencrypted keypair"
-  [{pk :private-key :as kp}] ; original map is kp
-  (assoc 
-    ; remove unencrypted private key
-    (dissoc kp :private-key)
-    ; add encrypted private key
-    :enc-private-key (encrypt pk)))
+(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
+  (assoc (dissoc kp :priv)       ; remove unencrypted private key
+         :enc-priv (encrypt pk))) ; add encrypted private key
 \end{minted}
@@
 (ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
-(defn enc-keypair-bad
-  [{pk :private-key :as kp}]
-  (assoc kp :enc-private-key (encrypt pk)))
-  ; Type Error:
-  ;   Expected EncKeyPair, given 
-  ;   (HMap :mandatory {:enc-private-key EncKey
-  ;                     :private-key RawKey
-  ;                     :public-key RawKey}
+(defn enc-keypair-bad [{pk :priv :as kp}]
+  (assoc kp :enc-priv (encrypt pk)))
+  ; Type Error: Expected EncKeyPair, given 
+  ;   (HMap :mandatory {:enc-priv EncKey, :priv RawKey, :pub RawKey}
   ;         :complete? true)
 \end{minted}
-The extra \clj{:private-key} entry does not match \clj{EncKeyPair}, so a type error
+The extra \clj{:priv} entry does not match \clj{EncKeyPair}, so a type error
 is raised.
@@
 %\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
-%dissociating the raw \clj{:private-key} entry with \clj{dissoc}
+%dissociating the raw \clj{:priv} entry with \clj{dissoc}
 %and associating an encrypted private key
-%as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
-%The expression \clj{(:private-key kp)} shows that keywords are also 
+%as \clj{:enc-priv} on an immutable map with \clj{assoc}.
+%The expression \clj{(:priv kp)} shows that keywords are also 
 %functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
 %Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
-%does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
+%does not contain an entry \clj{:priv}, and would complain if the \clj{dissoc}
 %operation forgot to remove it.
@@
 %with \emph{\#}-prefixed braces.
-%The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
-%aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
+%The syntax \clj{(fn [{pkey :priv, :as kp}] ...)}
+%aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:priv m)}
 %in the function body.
@@
 %
-%Since this example enforces that \clj{:private-key} must not appear
+%Since this example enforces that \clj{:priv} must not appear
 %in a \clj{EncKeyPair}
-%Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
+%Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:priv}
 %from the return value.
@@
 
+%FIXME define defalias above and keyword singletons
+
 \begin{minted}{clojure}
-(defalias Order   ; define type abbreviation
-  "A meal order, tracking dessert quantities."
-  (U '{:Meal ':lunch ; keyword singleton type
-       :desserts Int}
-     '{:Meal ':dinner :desserts Int}
+(defalias Order "A meal order, tracking dessert quantities."
+  (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner :desserts Int}
      '{:Meal ':combo :meal1 Order :meal2 Order}))
@@
 (ann desserts [Order -> Int])
-(defmulti desserts "Total desserts per order."
-          :Meal)  ; dispatch on :Meal entry
-; destructuring reads backwards, d is :desserts entry
-(defmethod desserts :lunch [{d :desserts}] d)
+(defmulti desserts :Meal)  ; dispatch on :Meal entry
+(defmethod desserts :lunch [{d :desserts}] d) ; d is :desserts entry
 (defmethod desserts :dinner [{d :desserts}] d)
@@
 
-(desserts {:Meal :combo 
-           :meal1 {:Meal :lunch :desserts 1}
+(desserts {:Meal :combo, :meal1 {:Meal :lunch :desserts 1}, 
            :meal2 {:Meal :dinner :desserts 2}})
@@
 (ann desserts' [Order -> Int])
-(defmulti desserts' 
-          (fn [o :- Order] (class (:desserts o))))
+(defmulti desserts' (fn [o :- Order] (class (:desserts o))))
 (defmethod desserts' Long [{d :desserts :as o}] 
-  ; o :- (U '{:Meal ':lunch :desserts Int}
-  ;         '{:Meal ':dinner :desserts Int}
-  ;         '{:Meal ':combo :desserts Int
-  ;           :meal1 Order :meal2 Order})
+  ; o :- (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner, :desserts Int}
+  ;         '{:Meal ':combo :desserts Int, :meal1 Order, :meal2 Order})
   d)
 (defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
-  ; o :- '{:Meal ':combo :meal1 Order :meal2 Order}
+  ; o :- '{:Meal ':combo, :meal1 Order, :meal2 Order}
   (+ (desserts' m1) (desserts' m2)))
@@
 \end{itemize}
-Equivalently, we learn \clj{o} is at least of type
+Equivalently, we learn \clj{o} is of type
+% Note: mintedinline doesn't work with hash characters #
 \begin{minted}{clojure}
-(U '{:desserts nil}
 ; :absent-keys, a set of known absent entries
-   (HMap :absent-keys #{:desserts}))
+(U '{:desserts nil} (HMap :absent-keys #{:desserts}))
 \end{minted}

### 2015-10-13 related-work.tex
fix esop formatting

@@
 \cite{MS02} and collaborators present a sequence of
-systems~\cite{Chambers:1992:OMC, Chambers:1994:TMM, MS02} with statically-typed multimethods
+systems~\cite{Chambers:1992:OMC,Chambers:1994:TMM,MS02} with statically-typed multimethods
 and modular type checking.  In contrast to Typed Clojure, in these

### 2015-10-14 conclusion.tex
need to cut 1.5 pages

@@
 % Delete the following paragraphs if space is needed.
-
-However, there is much more that Typed Clojure can provide. Most
-significantly, Typed Clojure currently does not provide \emph{gradual
-  typing}---interaction between typed and untyped code is unchecked and
-thus unsound. We hope to explore the possibilities of using existing
-mechanisms for contracts and proxies in Java and
-Clojure to enable sound gradual typing for Clojure.
-
-Additionally, the Clojure compiler is unable to use Typed Clojure's
-wealth of static information to optimize programs. Addressing this
-requires not only  enabling sound gradual typing, but also
-integrating Typed Clojure into the Clojure tool so
-that its information can be communicated to the compiler.
-
-Finally, our case study, evaluation, and broader experience indicate that Clojure
-programmers still find themselves unable to use Typed Clojure on some
-of their programs for lack of expressiveness. This requires continued
-effort to analyze and understand the features and idioms and
-develop new type checking approaches.
+%
+%However, there is much more that Typed Clojure can provide. Most
+%significantly, Typed Clojure currently does not provide \emph{gradual
+%  typing}---interaction between typed and untyped code is unchecked and
+%thus unsound. We hope to explore the possibilities of using existing
+%mechanisms for contracts and proxies in Java and
+%Clojure to enable sound gradual typing for Clojure.
+%
+%Additionally, the Clojure compiler is unable to use Typed Clojure's
+%wealth of static information to optimize programs. Addressing this
+%requires not only  enabling sound gradual typing, but also
+%integrating Typed Clojure into the Clojure tool so
+%that its information can be communicated to the compiler. 
+%
+%Finally, our case study, evaluation, and broader experience indicate that Clojure
+%programmers still find themselves unable to use Typed Clojure on some
+%of their programs for lack of expressiveness. This requires continued
+%effort to analyze and understand the features and idioms and
+%develop new type checking approaches.

### 2015-10-14 formal-model.tex
need to cut 1.5 pages

@@
 track variables.
-$$
-\begin{array}{lr}
+\begin{mathpar}
 \infer [LC-Let]
@@
            }
-&
+
 \infer [LC-Local]
@@
            }
-\end{array}
-$$
+\end{mathpar}
 
@@
 %a variable's type.
-$$
-\begin{array}{lr}
+\begin{mathpar}
 \infer [T0-Let]
@@
 %\judgementtwo{\isprop{\Number{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\Number}
-&
+
 \infer [T0-Local]
@@
            }
-\end{array}
-$$
+\end{mathpar}
 In T0-Local, 
@@
 The full proof system judgement
-$$
+$
 { \inpropenv {\propenv{}} {\prop{}} }
-$$
+$
 says \emph{proposition environment} {\propenv{}} proves proposition \prop{}.
@@
 
-\paragraph{The full judgment}
+\paragraph{Typing judgment}
 
@@
         {\TLocal}
-
+        {\TAbs}
         {\TIf}
-
-        {\TAbs}\\
-
-        {\TNil}
-
-    {\TFalse}
-
-    {\TLet}\\
-
-    {\TApp}
-
+    \\
+    \begin{array}{c}
+    {\TKw}\\
       {\TNum}\\
-
+    \end{array}
+    \begin{array}{c}
+      {\TNil}\\
+      {\TFalse}\\
     {\TConst}
-
-    {\TKw}
-
-    {\TNum}\\\\
-
-    {\TClass}
-
+    \end{array}
+    \begin{array}{c}
+    {\TStr}\\
+    {\TClass}\\
+    {\TTrue}
+  \end{array}
+        \\
+
+    {\TLet}
+    \\
+
+    {\TApp}\ \ 
     {\TSubsume}
-
-    {\TStr}\\
-
-    {\TTrue}
+    \\
   \end{mathpar}
@@
 
-We formalize our type system following~\cite{TF10}.
+\begin{figure}%[t!]
+  \footnotesize
+  \begin{mathpar}
+
+\SUnionSuper{}\ \ \ 
+\SUnionSub{}\ \ \ 
+\SFunMono{}\ \ \ 
+\begin{array}{l}
+\SObject{}\\
+\SClass{}\\
+\SSBool{}
+\end{array}
+
+\SFun{}
+\begin{array}{l}
+    \SRefl{}\ \ \ 
+    \STop{}\\
+\SSKw{}
+\end{array}
+
+  \end{mathpar}
+  \caption{Core subtyping rules}
+  \label{main:figure:subtyping}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+    \BIfTrue{}
+
+    \BIfFalse{}
+\end{mathpar}
+  \caption{Select core semantics}
+\label{main:figure:coresemantics}
+\end{figure}
+
+We formalize our system following Tobin-Hochstadt and Felleisen \cite{TF10}.
 %(with differences highlighted in $\trdiff{\text{blue}}$)
@@
 \classconst{} returns its argument's class.
-
-\begin{figure}
-  $$
-\constanttypefigure{}
-  $$
-  \caption{Constant typing}
-  \label{main:figure:constanttyping}
-\end{figure}
 
@@
 
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \SRefl{}
-
-    \STop{}
-
-\SUnionSuper{}
-
-\SUnionSub{}
-
-\SFunMono{}
-
-\SObject{}
-
-\SClass{}
-
-\begin{array}{l}
-\SSBool{}\\\\
-\SSKw{}
-\end{array}
-
-\SFun{}
-
-  \end{mathpar}
-  \caption{Core subtyping rules}
-  \label{main:figure:subtyping}
-\end{figure}
-
 \paragraph{Subtyping}
@@
 however a notable difference is \nil{} is a false value, which affects the
-semantics of \ifliteral{}:
-\begin{mathpar}
-    \BIfTrue{}
-
-    \BIfFalse{}
-\end{mathpar}
+semantics of \ifliteral{} (\figref{main:figure:coresemantics}).
 
@@
 
-\begin{figure}%[t]
+\begin{figure}
   \footnotesize
@@
   \begin{altgrammar}
-    \e{} &::=& \ldots   
-                      {\fieldexp {\fld{}} {\e{}}}
-                      \alt {\methodexp {\mth{}} {\e{}} {\overrightarrow{\e{}}}}&\mbox{Expressions}\\
-                      &\alt& {\newexp {\class{}} {\overrightarrow{\e{}}}}\\
-     &\alt& \mininonreflectiveexpsyntax{}
-    \\
+    \e{} &::=& \ldots   {\fieldexp {\fld{}} {\e{}}} \alt {\methodexp {\mth{}} {\e{}} {\overrightarrow{\e{}}}}
+                      \alt {\newexp {\class{}} {\overrightarrow{\e{}}}}
+                      &\mbox{Expressions}\\
+     &\alt& \nonreflectiveexpsyntax{} &\mbox{Non-reflective Expressions}\\
 
@@
   \end{altgrammar}
-  \\
   \begin{altgrammar}
@@
   \begin{mathpar}
-    \BField{}
-
+    \BField{}\ \ \ 
+%
     \BNew{}
@@
   \label{main:figure:javatyping}
+\end{figure}
+
+\begin{figure}
+  $$
+\constanttypefigure{}
+  $$
+  \caption{Constant typing}
+  \label{main:figure:constanttyping}
 \end{figure}
@@
 
-\begin{figure}%[t!]
+\begin{figure}
   \footnotesize
@@
                 &\mbox{Values} \\
- \disptablesyntax{} \\\\
+ \disptablesyntax{} \\
   \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
@@
     \Multisubtyping{}
+
+    \BDefMulti{}
   \end{mathpar}
@@
     \BDefMethod{}
-    \BDefMulti{}
     %\BBetaMulti{}
@@
   \getmethodfigure{}
-$$
-\begin{array}{ll}
-  \vcenter{\hbox{\BIsA{}}}
-  &
-  \vcenter{\hbox{\isaopsemfigure{}}}
-\end{array}
-$$
+\begin{mathpar}
+  {\BIsA{}}
+  {\isaopsemfigure{}}
+  \\
+\BBetaMulti{}
+\end{mathpar}
 \caption{Multimethod Syntax, Typing and Operational Semantics}
@@
 The overall dispatch mechanism is summarised by B-BetaMulti.
-$$
-\footnotesize \BBetaMulti{}
-$$
 First the dispatch function \v{d} is applied to the argument \vp{} to obtain
@@
   \begin{mathpar}
+    {\TAssoc}
+
     {\TGetHMap}
@@
     {\TGetHMapPartialDefault}
-
-    {\TAssoc}
+    \ \ \ 
+  {\SHMapMono}
   \end{mathpar}
   \begin{mathpar}
-    \HMapsubtyping{}
+  {\SHMapP}\ \ 
+  {\SHMap}
+
   \end{mathpar}
   \begin{mathpar}
-    {\BAssoc}
-    {\BGet}
+    {\BAssoc}\ \ 
+    {\BGet}\ \ 
     {\BGetMissing}
@@
   \label{main:figure:hmapsyntax}
+\end{figure}
+
+\begin{figure}
+  $$
+\LUpdate{}
+$$
+\caption{Update rule}
+\label{main:figure:updaterule}
+\end{figure}
+
+\begin{figure}
+$$
+\begin{array}{lr}
+  \begin{array}{llll}
+    \restrictfigure{}
+  \end{array}
+  \ \ 
+  \begin{array}{llll}
+    \removefigure{}
+  \end{array}
+\end{array}
+$$
+\caption{Restrict and remove}
+\label{main:figure:restrictremove}
 \end{figure}
@@
 The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
-Notice the propositions attached to each \getliteral{} rule are trivial---propositions
-are erased once they enter a HMap type.
-
-For presentational reasons, lookups on unions of HMaps are only supported in T-GetHMap. 
-Furthermore, each element of the union must contain the key we are looking up.
+Notice propositions are erased once they enter a HMap type.
+
+For presentational reasons, lookups on unions of HMaps are only supported in T-GetHMap
+and each element of the union must contain the relevant key.
 $$
@@
 The semantics for \getliteral{} and \assocliteral{} are straightforward.
-If the entry is missing, B-GetMissing produces \nil{}.
-
-\begin{figure*}
+%If the entry is missing, B-GetMissing produces \nil{}.
+
+\begin{figure}
   $$
@@
 $$
-$$
-\begin{array}{lr}
-  \begin{array}{llll}
-    \restrictfigure{}
-  \end{array}
-  \begin{array}{llll}
-    \removefigure{}
-  \end{array}
-\end{array}
-$$
 \caption{Type update (the metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables), 
-  \notsubtypein{}{\Nil{}}{\nottype{\t{}}} when \issubtypein{}{\Nil{}}{\t{}})}
+  \notsubtypein{}{\Nil{}}{\nottype{\t{}}} when \issubtypein{}{\Nil{}}{\t{}}, see
+\figref{main:figure:restrictremove} for \restrictliteral{} and \removeliteral{}.
+  )}
 \label{main:figure:update}
-\end{figure*}
+\end{figure}
 
@@
 
-The proof system for occurrence typing uses standard propositional logic,
+The occurrence typing proof system uses standard propositional logic,
 except for where nested information is combined. This is
-handled by the L-Update rule.
-  $$
-\LUpdate{}
-$$
+handled by L-Update (\figref{main:figure:updaterule}).
+
 It says

### 2015-10-14 metatheory.tex
fix esop formatting
need to cut 1.5 pages

@@
 concede that method and constructor calls may diverge or error, but
-assume they can never go wrong. (Assumptions for other operations are
+assume they can never go wrong
+(other assumptions
 given in the supplemental material).
@@
 
-We can now state our soundness theorems.
+We can now state our soundness theorem.
 
@@
 
-{\wrongtheoremnoproof{main}}
+%{\wrongtheoremnoproof{main}}
 %

### 2015-10-14 overview.tex
need to cut 1.5 pages

@@
 ideas, before studying the core features in detail in
-section~\ref{sec:formal}.
+\secref{sec:formal}.
 
@@
 to allow \clj{nil}.
+All Clojure values are true except \clj{nil} and \clj{false}, so a \clj{nil} test
+amounts to a test on the value.
 
@@
 
-All Clojure values are true except \clj{nil} and \clj{false}, so the
-comma is only added when the argument is non-nil.
-
 Typed Clojure guarantees that well-typed code cannot
@@
 model type-based control flow.
-In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
-is never passed a \clj{nil} argument.
+In \clj{greetings}, a branch ensures \clj{repeat}
+is never passed \clj{nil}.
 %
@@
 (defn greetings [n i]
-  (str "Hello, " (when i  ; when i is a non-nil integer
-                   (apply str (repeat i "hello, ")))
-       n "!"))
+  (str "Hello, " (when i (apply str (repeat i "hello, "))) n "!"))
 (greetings "Donald" 2)  ; :- Str
@@
 %
-Removing the \clj{when} expression is a static type error---\clj{repeat} 
-cannot be passed \clj{nil}.
+Removing the branch is a static type error---\clj{repeat} cannot be passed \clj{nil}.
 %
@@
 (ann  greetings-bad [Str (U nil Int) -> Str])
-(defn greetings-bad [n i]
-  (str "Hello, " (apply str 
-                   (repeat i ; Type Error: Expected Int, given (U nil Int).
-                           "hello, "))
-       n "!"))
+(defn greetings-bad [n i]           ; Expected Int, given (U nil Int)
+  (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
 \end{minted}
@@
 \end{minted}
-
+%
 %Typed Clojure and Java treat \java{null} differently.
@@
 %must assume methods can return \clj{nil}.
-
-In contrast, JVM invariants guarantee constructors return a non-null reference.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
+%
+In contrast, JVM invariants guarantee constructors return non-null.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
 %
 \begin{exmp}
-\inputminted[firstline=15,lastline=16]{clojure}{code/demo/src/demo/parent3.clj}
+\inputminted[firstline=15,lastline=15]{clojure}{code/demo/src/demo/parent3.clj}
 \end{exmp}
@@
 \begin{minted}{clojure}
-(map hi [:en :fr :bocce]) ; map over keyword vector 
-;=> ("hello" "bonjour" "um...")
+(map hi [:en :fr :bocce]) ;=> ("hello" "bonjour" "um...")
 \end{minted}
@@
 \begin{minted}{clojure}
-(.getName (new File "STAINS/JELLY")) ; :- (U nil Str)
-;=> "JELLY"
+(.getName (new File "STAINS/JELLY")) ;=> "JELLY" :- (U nil Str)
 \end{minted}
@@
 
-Hash-maps are accessed with the \clj{get} function:
+Keywords look themselves up in a map when called, or return \clj{nil}.
+%
 \begin{exmp}
@@
 (def breakfast {:en "waffles" :fr "croissants"})
-(get breakfast :en)    ; :- Str
-;=> "waffles"
+(:en breakfast)  ;=> "waffles" :- Str
+(:bocce breakfast) ; nil :- nil
 \end{minted}
@@
 \end{exmp}
-Additionally, keywords are functions that look themselves up in a
-map. Missing keys produce \clj{nil}.
-\begin{minted}{clojure}
-(:fr breakfast)        ; :- Str
-;=> "croissants"
-(:bocce breakfast) ; :- nil
-;=> nil
-\end{minted}
 
-In Typed Clojure, \emph{HMap types} describe the most common usages of
+\emph{HMap types} describe the most common usages of
 keyword-keyed maps.
@@
 \end{minted}
-The inferred type for \clj{breakfast} holds two kinds of information---the known entries--\clj{:en}
+The inferred type for \clj{breakfast} holds two pieces of information---the known entries--\clj{:en}
 and \clj{:fr}---and their types, which are  \clj{:mandatory}, and that 
@@
 (def lunch {:en "muffin" :fr "baguette"})
-(:en lunch)    ; :- Str
-;=> "muffin"
-(:fr lunch)    ; :- Str
-;=> "baguette"
-(:bocce lunch) ; :- Any     ; Unknown lookups are now less accurate
-;=> nil
+(:bocce lunch) ; nil :- Any ; less accurate type
 \end{minted}
@@
 \end{exmp}
+%(:en lunch)    ; :- Str
+%;=> "muffin"
+%(:fr lunch)    ; :- Str
+%;=> "baguette"
 
@@
 a company with a large production Typed Clojure system
-(section~\ref{sec:casestudy} presents a case study and empirical
+(\secref{sec:casestudy} presents a case study and empirical
 result from this code base).
@@
   (HMap :mandatory {:pub RawKey, :priv RawKey}, :complete? true))
-
 (defalias EncKeyPair ; extra keys disallowed
@@
 (defn enc-keypair [{pk :priv :as kp}] ; original map is kp
-  (assoc (dissoc kp :priv)       ; remove unencrypted private key
-         :enc-priv (encrypt pk))) ; add encrypted private key
+  (assoc (dissoc kp :priv) :enc-priv (encrypt pk)))
 \end{minted}
@@
 \end{exmp}
+
+%(ann enc-keypair [RawKeyPair -> EncKeyPair])
+%(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
+%  (assoc (dissoc kp :priv)       ; remove unencrypted private key
+%         :enc-priv (encrypt pk))) ; add encrypted private key
 
@@
 (defn enc-keypair-bad [{pk :priv :as kp}]
-  (assoc kp :enc-priv (encrypt pk)))
-  ; Type Error: Expected EncKeyPair, given 
-  ;   (HMap :mandatory {:enc-priv EncKey, :priv RawKey, :pub RawKey}
-  ;         :complete? true)
+  (assoc kp :enc-priv (encrypt pk))) ; Type error: :pub disallowed
 \end{minted}
@@
 linked.
-There are infinite ways of both structuring and dispatching on data,
-so we cannot hope to merely add a set of special case rules for
+There are infinite ways of structuring and dispatching on data,
+so we cannot merely add a set of special case rules for
 handling these features. Instead, as type system designers, we must
@@
 we can automatically cover both easy cases and those
-that compose simple rules in arbitrary ways.
+that compose rules in arbitrary ways.
 
@@
 
-In the first method, the dispatch value is the class \clj{Long}, a subtype
-of \clj{Int}, and the second method has dispatch value \clj{nil}, the sentinel value for a failed map lookup.
+The first method has dispatch value \clj{Long}, a subtype
+of \clj{Int}, and the second method has \clj{nil}, the sentinel value for a failed map lookup.
 In practice, \clj{:lunch} and \clj{:dinner} meals will dispatch to the \clj{Long}
@@
   \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
-  \item \clj{o} is missing a \clj{:desserts} entry, like \clj{{}}.
+  \item \clj{o} is missing a \clj{:desserts} entry.%, like \clj{{}}.
 \end{itemize}
-Equivalently, we learn \clj{o} is of type
+We can express this type with the \clj{:absent-keys} HMap option
+%Equivalently, \clj{o} is of type
 % Note: mintedinline doesn't work with hash characters #
 \begin{minted}{clojure}
-; :absent-keys, a set of known absent entries
 (U '{:desserts nil} (HMap :absent-keys #{:desserts}))
@@
 with this new information (because \clj{:desserts}
-is neither mapped to \clj{nil} or absent).
+is neither \clj{nil} or absent).
 

### 2015-10-16 formal-model.tex
shaved 1/4 page

@@
 
-Now that we have demonstrated the core features Typed Clojure
+After demonstrating the core features Typed Clojure
 provides, we link them together in a formal model called
@@
 
-Our presentation will start with a review of
+We start with a review of
 occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
@@
 To illustrate, recall \egref{example:desserts-on-meal}.
-We know the parameter \makelocal{o} is of type $\Order$,
+The parameter \makelocal{o} is of type $\Order$,
 %by the annotation on $desserts$
@@
 %
-In checking the ${\makekw{combo}}$ method, we also assume
+In the ${\makekw{combo}}$ method, we know
 ${\appexp{\makekw{Meal}}{\makelocal{o}}}$ is ${\makekw{combo}}$,
@@
 %the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.
-
-\paragraph{Typing judgment}
 
@@
 
+\paragraph{Typing judgment}
+
 We formalize our system following Tobin-Hochstadt and Felleisen \cite{TF10}.
@@
 The typing judgment 
-$$
+$
 {\judgementrewrite   {\propenv}
@@
   {\ep{}}}
-$$
+$
 says expression \e{} rewrites to \ep{}, which
@@
 With \propenv{} = \isprop{\Top}{\makelocal{d}},
-the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
-via the T-App rule, with three main steps.
+T-If first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
+via the T-App rule, with three steps.
 
@@
 which encodes the type, proposition, and object information
-about variables
-as previously discussed.
+about variables.
 
@@
 
-We now walk through the conversion from unresolved expression~\ref{eq:unresolved} to 
-resolved expression~\ref{eq:resolved}.
+We now walk through this conversion.% from unresolved expression~\ref{eq:unresolved} to 
+%resolved expression~\ref{eq:resolved}.
 
@@
 \begin{figure}
+  \footnotesize
   $$
@@
 \begin{figure}
+  \footnotesize
 $$
@@
 
-A lookup of a key that is neither present or absent is checked by
+A lookup that is not present or absent is checked by
 T-GetHMapPartialDefault.
@@
 $,
-which is complete, like its input.
+a complete map.
 T-AssocHMap also enforces ${\k{}} \not\in {\absent{}}$ to prevent badly formed types.
@@
 designate \MapLiteral{} as a common supertype for all HMaps.
-S-HMap says that an HMap is a subtype of another HMap if they agree
+S-HMap says that HMaps are subtypes if they agree
 on \completenessmeta{}, agree on mandatory entries with subtyping

### 2015-10-16 overview.tex
shaved 1/4 page

@@
 We now begin a tour of the central features of Typed Clojure,
-beginning with Clojure itself. In our presentation, we will make 
-use of the full Typed Clojure system to illustrate the key type system
+beginning with Clojure itself. Our presentation
+uses the full Typed Clojure system to illustrate key type system
 ideas, before studying the core features in detail in
@@
 
-Clojure~\cite{Hic08} is a Lisp built to run on the
-Java Virtual Machine with exemplary support for concurrent programming
-and immutable data structures. It emphasizes mostly-functional
-programming, restricting imperative updates to a limited set of
-structures which have specific thread synchronization behaviour. By
-default, it provides fast implementations of immutable lists, vectors,
-and hash tables, which are used for most data structures, although it
-also provides means for defining new records.
-
-One of Clojure's primary advantages is easy interoperation with
-existing Java libraries. It automatically generates appropriate JVM
-bytecode to make Java method and constructor calls, and treats Java
-values as any other Clojure value. However, this smooth
+Clojure~\cite{Hic08} is a Lisp that runs on the
+Java Virtual Machine with support for concurrent programming
+and immutable data structures in a mostly-functional
+style.
+%, restricting imperative updates to a limited set of
+%structures each with specific thread synchronization behaviour. 
+%Fast implementations of immutable vectors, and hash tables are featured,
+%and a means for defining new records.
+%
+Clojure provides easy interoperation with
+existing Java libraries,
+with
+Java
+values being as any other Clojure value. However, this smooth
 interoperability comes at the cost of pervasive \java{null}, which
@@
 instance and calls the \clj{getParent} method on the result,
-returning a string \clj{"a"}, inferred as nullable.
+returning a nullable string \clj{"a"}.
 
@@
 (defn parent-bad-in [f :- (U nil File)]
-  (.getParent f)) ; Type Error: Cannot call instance method 
-                  ; java.io.File/getParent on type (U nil File).
+  (.getParent f)) ; Type Error: Cannot call instance method on nil.
 \end{minted}
@@
 but the same intuition applies---adding support for multiple dispatch
-automatically allows arbitrary combinations and nestings
-of it and previous simple dispatch rules.
+admits arbitrary combinations and nestings
+of it and previous dispatch rules.
 

### 2015-10-16 related-work.tex
shaved 1/4 page

@@
 in Typed Clojure. However, Typed Clojure does not attempt to rule out
-failed dispatches at runtime.
+failed dispatches.
 
@@
 \paragraph{Other optional and gradual type systems}
-In addition to Typed Racket, several other gradual type
-systems have been developed recently, targeting existing
+%In addition to Typed Racket, 
+Several other gradual type
+systems have been developed for existing
 dynamically-typed languages.  Reticulated Python~\cite{Vitousek14} is
@@
 source-to-source translation that inserts dynamic checks at language
-boundaries and supporting Python's first-class object system.  Typed
-Clojure does not support a first-class object system because Java (and
-Clojure) have nominal classes, however HMaps offer an alternative to
+boundaries and supporting Python's first-class object system. 
+Clojure's nominal classes avoids the need to support
+first-class object system in Typed Clojure, however HMaps offer an alternative to
 the structural objects offered by Reticulated. Similarly,
@@
 
-Optional types, requiring less implementation effort and avoiding
-runtime cost, have been  adopted in industry, including Hack for PHP~\cite{hack}, and Flow~\cite{flow} and
+Optional types
+%, requiring less implementation effort and avoiding
+%runtime cost, 
+have been  adopted in industry, including Hack for PHP~\cite{hack}, and Flow~\cite{flow} and
 TypeScript~\cite{typescript}, two extensions of JavaScript. These
-systems  support  forms of occurrence typing, but not in the
-generality presented here, nor do they include the other features we
+systems  support  limited forms of occurrence typing,
+and do not include the other features we
 present.

### 2015-10-16 intro.tex
need to cut 1.5 pages
1/2 page to go

@@
 Clojure is a dynamically
-typed language in the Lisp family built to run on the Java Virtual
-Machine (JVM) which has recently gained popularity as an alternative
+typed language in the Lisp family---built on the Java Virtual
+Machine (JVM)---which has recently gained popularity as an alternative
 JVM language.  It offers the flexibility of a Lisp dialect, including
-macros, emphasizes a functional style via a
-standard library of immutable data structures, and provides
+macros, emphasizes a functional style via
+immutable data structures, and provides
 interoperability with existing Java code, allowing programmers to use
@@
 functional programming, and Lisp-influenced abstraction is desired on
-the JVM. As a result, it now has an extensive base of existing untyped
-programs, whose developers can now benefit from Typed Clojure. As
+the JVM. As a result, it has an extensive base of existing untyped
+programs, whose developers can benefit from Typed Clojure. As
 a result, Typed Clojure is used in industry, experience we discuss
@@
 \begin{figure}[t!]
-\inputminted[firstline=5]{clojure}{code/demo/src/demo/ex-intro.clj}
+\inputminted[firstline=15]{clojure}{code/demo/src/demo/ex-intro.clj}
 \caption{A simple Typed Clojure program}
@@
 an understanding of the Java type system and handles interoperation
-appropriately. Notably, Typed Clojure avoids conflating Java's
-\texttt{null} with all other types, leading to automatic type-enforced
+appropriately. Notably, \texttt{null} is a distinct type in Typed Clojure,
+leading to automatic type-enforced
 absence of null-pointer exceptions.
 
-A simple example of these features in combination is given in
+An example of these features is given in
 figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
-on the \texttt{class} of the argument---when it is a \texttt{String},
-the first method implementation is called, for a Java \texttt{File}, the
-second. The method for \texttt{String} uses Java interoperation
-to call a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
-method for \texttt{File} invokes
-the \texttt{.getName} method with a non-nil target, returning a nilable
-type. Typed Clojure uses type information to resolve JVM method or constructor overloads,
-avoiding expensive runtime reflective calls.
+on the \texttt{class} of the argument---for \texttt{String}s,
+the first method implementation is called, for \texttt{File}s, the
+second. The \texttt{String} method calls
+a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
+\texttt{.getName} method 
+on \texttt{File} requires a non-nil target, returning a nilable
+type.  Typed Clojure offers an opt-in mode that
+resolves JVM overloading, avoiding expensive runtime reflective calls.
 
-Finally, when not working with existing Java classes and objects, the
-most common Clojure data structure is the dictionary, a
-high-performance immutable table which Clojure programmers repurpose
-for all manner of data types. Simply treating them as uniformly-typed
-key-value mappings would be entirely insufficient for existing
+Finally, high-performance immutable dictionaries
+are the most common Clojure data structure,
+which programmers repurpose liberally.
+Simply treating them as uniformly-typed
+key-value mappings would be insufficient for existing
 programs and programming styles. Instead, Typed Clojure provides a
-flexible \textbf{heterogenous map} type, in which mandatory, optional, and
-absent keys can be specified.
+flexible \textbf{heterogenous map} type, in which specific entries can be specified.
 
 While these features may seem disparate, they are unified in important
-ways. First, they all leverage the type system mechanisms that Typed
-Clojure inherits from Typed Racket---multimethods when using 
+ways. First, they leverage the type system mechanisms
+inherited from Typed Racket---multimethods when using 
 dispatch via predicates, Java interoperation for handling
 \texttt{null} tests, and heterogenous maps using union types and
-reasoning about subcomponents of data. Second, and more significantly,
-they are the crucial features for handling Clojure code in
+reasoning about subcomponents of data. Second,
+they are crucial features for handling Clojure code in
 practice. Typed Clojure's use in real Clojure deployments would not be
@@
     type system for Clojure that understands existing Clojure idioms.
-  \item We present a sound formal model for the three crucial type
-    system features that Typed Clojure relies on: multi-methods, Java
+  \item We present a sound formal model for three crucial type
+    system features: multi-methods, Java
     interoperability, and heterogenous maps.
@@
  \secref{sec:formal} and prove type soundness
- (\secref{sec:metatheory}). We then discuss the full implementation of
- Typed Clojure, dubbed \coretyped{}, which extends the formal model in
- many ways, and empirical analysis of significant code bases written
- in \coretyped{} in \secref{sec:experience}. Finally, we discuss
- related work and conclude.
+ (\secref{sec:metatheory}). We then 
+ %discuss the full implementation of
+ %Typed Clojure, \coretyped{}, which extends the formal model in many ways, 
+ present an empirical analysis of significant code bases written
+ in the full implementation of Typed Clojure---\coretyped{}---in \secref{sec:experience}. 
+ Finally, we discuss related work and conclude.

### 2015-10-16 overview.tex
1/2 page to go

@@
 
-Here is a simple program in Typed Clojure.
+Here is a simple one-argument function
 %
-We define \clj{greet} as a one-argument function
-taking and returning a string.
+\clj{greet} that takes and returns strings.
 
@@
 %
-Strings are accepted, but providing \clj{nil} (Clojure's name for Java's \java{null})
-is a static type error---\clj{nil} is not a string in Typed Clojure.
+Providing \clj{nil} (exactly Java's \java{null})
+is a static type error---\clj{nil} is not a string.
 %
@@
 
-\paragraph{Unions} We can make the annotation more permissive with \emph{ad-hoc unions}
-to allow \clj{nil}.
-All Clojure values are true except \clj{nil} and \clj{false}, so a \clj{nil} test
-amounts to a test on the value.
+\paragraph{Unions} To allow \clj{nil}, we use \emph{ad-hoc unions} (\clj{nil} and \clj{false}
+are logically false).
 
@@
 (defn greet-nil [n] (str "Hello" (when n (str ", " n)) "!"))
-(greet-nil "Donald") ; :- Str
-;=> "Hello, Donald!"
-(greet-nil nil)      ; :- Str
-;=> "Hello!"
+(greet-nil "Donald") ;=> "Hello, Donald!" :- Str 
+(greet-nil nil)      ;=> "Hello!"         :- Str
 \end{minted}
 
-Typed Clojure guarantees that well-typed code cannot
-dereference the null-pointer.
-This is especially important for Clojure programs---\clj{nil}
-is treated like any other distinct datum in Clojure,  and its status as
-one of only two false values means it is a common choice to indicate
-``nothing'' or ``failure''.
+Typed Clojure prevents well-typed code from dereferencing \clj{nil}.
+%This is important for Clojure programs---\clj{nil}
+%is treated like any other distinct datum in Clojure.
 
-\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{TF10} to
-model type-based control flow.
+\paragraph{Flow analysis} Occurrence typing~\cite{TF10}
+models type-based control flow.
 In \clj{greetings}, a branch ensures \clj{repeat}
@@
   (str "Hello, " (when i (apply str (repeat i "hello, "))) n "!"))
-(greetings "Donald" 2)  ; :- Str
-;=> "Hello, hello, hello, Donald!"
-(greetings "Grace" nil) ; :- Str
-;=> "Hello, Grace!"
+(greetings "Donald" 2)  ;=> "Hello, hello, hello, Donald!" :- Str
+(greetings "Grace" nil) ;=> "Hello, Grace!"                :- Str
 \end{minted}
@@
 
-Clojure supports interoperability with Java, including the ability to
-call constructors, invoke methods, and access fields.
-
-The following Typed Clojure program constructs a new \clj{File}
-instance and calls the \clj{getParent} method on the result,
-returning a nullable string \clj{"a"}.
+Clojure can interact with Java constructors, methods, and fields.
+This program calls the \clj{getParent} on a constructed
+\clj{File}
+instance, inferred as a nullable string.
 
@@
 \begin{minted}{clj}
-(.getParent (new File "a/b"))  ; :- (U nil Str)
-;=> "a"
+(.getParent (new File "a/b"))  ;=> "a" :- (U nil Str)
 \end{minted}
@@
 \end{exmp}
-
-Typed Clojure helps the Clojure compiler avoid expensive reflective 
-calls,
-however if a specific constructor, method, or field cannot be found based on the
+%
+Typed Clojure can optionally integrate with the Clojure compiler to avoid expensive reflective 
+calls, however if a specific constructor, method, or field cannot be found based on the
 static types of its arguments, a type error is thrown.
-
+%
 \begin{minted}{clj}
 (fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
-
 \end{minted}
-
+%
 Function arguments default to \clj{Any}, the most permissive type. Ascribing
@@
 \end{minted}
-Since Java-level types do not provide information about nullability,
-by default Typed Clojure conservatively assumes that Java method and
-constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
-\egref{example:getparent-direct-constructor},
- must be \emph{non-nullable}, but this can be
-configured for  particular calls as needed.
-The target of method invocations is always non-nullable.
+Since Java-level types lack nullability information,
+Typed Clojure conservatively asserts Java methods and
+constructor arguments cannot be passed \clj{nil}---this restriction can be
+adjusted per method. Method targets can never be \clj{nil}.
 
-Any Java method returning a reference can also return \java{null} ---
-Typed Clojure rejects programs that assume otherwise.
+Typed Clojure rejects programs that assume methods cannot return \clj{nil}.
 %
@@
 %
-Recall the multimethod from figure~\ref{fig:ex1}, reproduced here.
-\begin{minted}{clj}
-(ann pname [(U File String) -> (U nil String)])
-(defmulti pname class)
-(defmethod pname String [s] (pname (new File s)))
-(defmethod pname File [f] (.getName f))
-\end{minted}
-
+Recall the multimethod from \figref{fig:ex1}. %, reproduced here.
+%\begin{minted}{clj}
+%(ann pname [(U File String) -> (U nil String)])
+%(defmulti pname class)
+%(defmethod pname String [s] (pname (new File s)))
+%(defmethod pname File [f] (.getName f))
+%\end{minted}
+%
 %Its dispatching function is
@@
 %respectively.
-\noindent
+%\noindent
 The dispatch function \clj{class}
@@
 
+%(isa? (class "STAINS/JELLY") Object) ;=> true
+%(isa? (identity :en) :fr) ;=> false
+%(isa? (class (new File "JELLY")) String) ;=> false
 \begin{minted}{clojure}
-(isa? (identity :en) :en) ;=> true
-(isa? (identity :en) :fr) ;=> false
-(isa? (class "STAINS/JELLY") String) ;=> true
-(isa? (class "STAINS/JELLY") Object) ;=> true
-(isa? (class (new File "JELLY")) String) ;=> false
+(isa? :en :en)       ;=> true
+(isa? String Object) ;=> true
 \end{minted}
@@
 is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.
-
-In our example,
+%
+For example,
+the call
   \clj{(pname "STAINS/JELLY")}
-chooses the  method for \clj{String} because
-\clj{(isa? (class "STAINS/JELLY") String)}
-is true and
-\clj{(isa? (class "STAINS/JELLY") File)}
-is false. The \clj{String} method body
-\clj{(pname (new File "STAINS/JELLY"))}
-chooses the \clj{File} method for opposite reasons,
-resulting in 
-\begin{minted}{clojure}
-(.getName (new File "STAINS/JELLY")) ;=> "JELLY" :- (U nil Str)
-\end{minted}
+picks the \clj{String} method because \clj{(isa? String String)}
+is true, and
+\clj{(isa? String File)}
+is not.
+%---\clj{(class "STAINS/JELLY")}
+%is \clj{String}.
+%
+%The \clj{String} method body
+%\clj{(pname (new File "STAINS/JELLY"))}
+%chooses the \clj{File} method for opposite reasons.
 
@@
 (:en breakfast)    ;=> "waffles" :- Str
-(:bocce breakfast) ; nil :- nil
+(:bocce breakfast) ;=> nil       :- nil
 \end{minted}
@@
 (def lunch {:en "muffin" :fr "baguette"})
-(:bocce lunch) ; nil :- Any ; less accurate type
+(:bocce lunch) ;=> nil :- Any ; less accurate type
 \end{minted}
@@
 %\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
-If we forget to remove the unencrypted private key,
-a type error is given, because \clj{EncKeyPair} is fully specified.
+As \clj{EncKeyPair} is fully specified, we must remove extra keys like \clj{:priv}.
 %
 \begin{minted}{clojure}
-(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
 (defn enc-keypair-bad [{pk :priv :as kp}]
-  (assoc kp :enc-priv (encrypt pk))) ; Type error: :pub disallowed
+  (assoc kp :enc-priv (encrypt pk))) ; Type error: :priv disallowed
 \end{minted}
-The extra \clj{:priv} entry does not match \clj{EncKeyPair}, so a type error
-is raised.
 

### 2015-10-16 small.tex
New page for bib. Changes to sec 1 & 2
CircleCI data
add GT citations. better update figure and section 3.6
convert to ESOP format
fix esop formatting
need to cut 1.5 pages
1/2 page to go

@@
 
-\documentclass[preprint,9pt,twocolumn,nocopyrightspace,authoryear]{sigplanconf}
+\documentclass{llncs}
 
@@
 
+\bibliographystyle{splncs03}
+
 \usepackage{common}
@@
 \usepackage{url}
-
 \newcounter{ex}
@@
 \par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
-\\[-2.4em]\nopagebreak}
+\\[-2.5em]\nopagebreak}
 {\relax}
@@
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
+%
+%\setminted{%fontsize=\footnotesize
+%}
 
@@
 
-\authorinfo{}{}{\vspace*{-2mm}}
-%\authorinfo{Ambrose Bonnaire-Sergeant}
+%\authorinfo{}{}{\vspace*{-2mm}}
+\author{\ma{\text{Ambrose Bonnaire-Sergeant}^\dagger\text{, Rowan Davies*}, \text{Sam Tobin-Hochstadt}^\dagger}}
+\institute{Indiana University$\dagger$; Omnia Team, Commonwealth Bank of Australia*}
 %       {Indiana University}
 %       {abonnair@indiana.edu}
-%\authorinfo{Rowan Davies}
 %       {University of Western Australia}
-%           {rowan.davies@uwa.edu.au}
-%\authorinfo{Sam Tobin-Hochstadt}
+%       {Rowan.Davies@cba.com.au}
 %       {Indiana University}
@@
 the Clojure world, and has acquired significant adoption in the
-Clojure community. Typed Clojure's type
-system repurposes Typed Racket's
+Clojure community. Typed Clojure repurposes Typed Racket's
 \emph{occurrence typing}, an approach to statically reasoning about
@@
 
-\input{acknowledgements}
+%\input{acknowledgements}
 
@@
 
-\bibliographystyle{abbrvnat}
-
 % The bibliography should be embedded for final submission.
+\newpage
 

### 2015-10-16 metatheory.tex
It fits.

@@
 
-Our main lemma says if there is a defined reduction, then the propositions, object
-and type are correct.
-The metavariable \definedreduction{} ranges over \v{}, \errorvalv{} and \wrong{}.
+We can now state our main lemma and soundness theorem.  The
+metavariable \definedreduction{} ranges over \v{}, \errorvalv{} and
+\wrong{}. Proofs are deferred to \ref{appendix:lemma:soundness}.
 
@@
   {\soundnesslemmahypothesis}
-  \begin{proof}
-    By induction on the derivation of \opsem {\openv{}} {\e{}} {\a{}}.
-    (Full proof given as lemma~\ref{appendix:lemma:soundness}).
-  \end{proof}
 \end{lemma}
-
-We can now state our soundness theorem.
 

### 2015-10-17 formal-model.tex
It fits.
first submission
review first half

@@
 
-After demonstrating the core features Typed Clojure
-provides, we link them together in a formal model called
+After demonstrating the core features of Typed Clojure, 
+we link them together in a formal model called
 \lambdatc{}.
@@
 applications, abstractions, conditionals, and let expressions.
-All binding forms introduce fresh variables.
+All binding forms introduce fresh variables---a subtle but important point since our type environments
+are not simply dictionaries.
 Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
@@
 constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
-close value environments \openv{} over functions, which map bindings to values.
+close value environments \openv{}---which map bindings to values---over functions.
 
@@
 The difference between the types
-\Value{\Keyword} and \Keyword{} is subtle.
-The former is inhabited by the class literal \Keyword{} and the result of 
-\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
+\Value{\class{}} and \class{} is subtle.
+The former is inhabited by class literals like \Keyword{} and the result of 
+\appexp{\classconst{}}{\makekw{a}}---the latter by \emph{instances} of classes,
+like a keyword literal \makekw{a}, an instance of the type \Keyword{}.
 %
@@
 %about deeply nested expressions.
-To reason about nested expressions,
-each expression is associated with 
+Each expression is associated with 
 a symbolic representation
@@
 We use two path elements---{\classpe{}} and {\keype{k}}---representing the results
-of calling \classconst{} and of looking up a keyword $k$ respectively.
+of calling \classconst{} and looking up a keyword $k$, respectively.
 
 \paragraph{Propositions with a logical system}
-In standard type systems, \emph{type environments}
-track variables.
+In standard type systems, association lists often
+track the types of variables, like in LC-Let and LC-Local.
 \begin{mathpar}
@@
 $
-{ \inpropenv {\propenv{}} {\isprop {\t{}} {\path{\pathelem{}}{\x{}}}} }
-$
-appeals to the proof system to solve \t{}.
+{ \inpropenv {\propenv{}} {\isprop {\t{}}{\x{}}}}
+$
+appeals to the proof system to solve for \t{}.
 %says under logical assumptions {\propenv{}}, object {\path{\pathelem{}}{\x{}}} is of type \t{}.
@@
 \paragraph{Methods} Next we convert {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
-to
+to the non-reflective expression
 {\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
@@
 
-As an example,
+As an example, the expression
 \isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
@@
 \end{listing}
-
+%
 %For convenience, examples in this section are flattened when they are really nested
@@
 %\end{lstlisting}
-
+%
 %\defmethodliteral{} returns a new extended multimethod
@@
 %\end{minted}
-
+%
 %\paragraph{How to check}
-We now show to check
-{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}.
-%
-The expression {\createmultiexp {\s{}} {\e{}}} creates a multimethod with \emph{interface type} \s{}, and dispatch function \e{}
+To check 
+{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}},
+%
+we note
+{\createmultiexp {\s{}} {\e{}}} creates a multimethod with \emph{interface type} \s{}, and dispatch function \e{}
 of type \sp{},
@@
 {\MultiFntype {\s{}} {\sp{}}}. % with interface type {\s{}} and dispatch function type {\sp{}}.
-The T-DefMulti typing rule simply checks the dispatch function, and
+The T-DefMulti typing rule checks the dispatch function, and
 verifies both the interface and dispatch type's domain agree.
@@
   \label{main:figure:hmapsyntax}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  $$
-\LUpdate{}
-$$
-\caption{Update rule}
-\label{main:figure:updaterule}
 \end{figure}
@@
 The substitution for {\x{}} handles the case where \object{} is empty.
-$$
+\begin{mathpar}
 \begin{array}{rcl}
 {\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
+\end{array}
+\ \ \ \ \ \ \ 
+\begin{array}{rcl}
 {\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
 \end{array}
-$$
+\end{mathpar}
 
@@
 except for where nested information is combined. This is
-handled by L-Update (\figref{main:figure:updaterule}).
+handled by L-Update:
+{  \footnotesize
+  $$
+\LUpdate{}
+$$
+}
 
@@
 To demonstrate,
-\makekw{lunch} meals hit the first HMap case and
+\makekw{lunch} meals are handled by the first HMap case and
 update to {\HMapp {\extendmandatoryset {\mandatory{}}{\Valkw{Meal}}{\sp{}}} {\emptyabsent{}}}

### 2015-10-17 intro.tex
review first half

@@
 gradual type systems aimed at checking existing programs written in
-existing languages.  These include  TypeScript and Flow for
-JavaScript, Hack for PHP, and MyPy
-for Python among the optional systems, and Typed Racket, Reticulated
-Python, and GradualTalk among gradually-typed systems.\footnote{We
+existing languages.  These include  TypeScript~\cite{typescript} and Flow~\cite{flow} for
+JavaScript, Hack~\cite{hack} for PHP, and mypy~\cite{mypy}
+for Python among the optional systems, and Typed Racket~\cite{TF08}, Reticulated
+Python~\cite{Vitousek14}, and GradualTalk~\cite{gradualtalk} among gradually-typed systems.\footnote{We
   use ``gradual typing'' for systems like Typed Racket with sound
@@
 functional programming, and Lisp-influenced abstraction is desired on
-the JVM. As a result, it has an extensive base of existing untyped
-programs, whose developers can benefit from Typed Clojure. As
-a result, Typed Clojure is used in industry, experience we discuss
-in this paper.
+the JVM. As a result, there is an extensive base of existing untyped
+programs whose developers can benefit from Typed Clojure,
+an experience we discuss in this paper.
 
@@
 appropriately. Notably, \texttt{null} is a distinct type in Typed Clojure,
-leading to automatic type-enforced
-absence of null-pointer exceptions.
+designed to automatically rule out null-pointer exceptions.
 
 An example of these features is given in
-figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
+\figref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
 on the \texttt{class} of the argument---for \texttt{String}s,
@@
 on \texttt{File} requires a non-nil target, returning a nilable
-type.  Typed Clojure offers an opt-in mode that
-resolves JVM overloading, avoiding expensive runtime reflective calls.
+type.  
+%Typed Clojure offers an opt-in mode that
+%resolves JVM overloading, avoiding expensive runtime reflective calls.
 
-Finally, high-performance immutable dictionaries
-are the most common Clojure data structure,
-which programmers repurpose liberally.
+Finally, flexible, high-performance immutable dictionaries
+are the most common Clojure data structure.
 Simply treating them as uniformly-typed
@@
  present an empirical analysis of significant code bases written
- in the full implementation of Typed Clojure---\coretyped{}---in \secref{sec:experience}. 
+ in \coretyped{}---the full implementation of Typed Clojure---in \secref{sec:experience}. 
  Finally, we discuss related work and conclude.

### 2015-10-17 overview.tex
review first half

@@
 %
-Clojure provides easy interoperation with
-existing Java libraries,
-with
-Java
-values being as any other Clojure value. However, this smooth
-interoperability comes at the cost of pervasive \java{null}, which
-leads to the possibility of null pointer exceptions---a drawback we
-address in Typed Clojure.
+Clojure provides easy interoperation with existing Java libraries, with Java values being like any other Clojure value. 
+However, this smooth interoperability comes at the cost of pervasive \java{null}, which leads to the possibility of null pointer exceptions---a drawback we address in Typed Clojure.
 
@@
 (defn greet [n] (str "Hello, " n "!"))
-(greet "Grace") ; :- Str
-;=> "Hello, Grace!"
+(greet "Grace") ;=> "Hello, Grace!" :- Str
 \end{minted}
@@
 are logically false).
-
+%
 \begin{minted}{clj}
@@
 \end{minted}
-
+%
+%
 Typed Clojure prevents well-typed code from dereferencing \clj{nil}.
@@
 \clj{File}
-instance, inferred as a nullable string.
+instance, returning a nullable string.
 
@@
 %
-Typed Clojure can optionally integrate with the Clojure compiler to avoid expensive reflective 
-calls, however if a specific constructor, method, or field cannot be found based on the
-static types of its arguments, a type error is thrown.
+Typed Clojure can integrate with the Clojure compiler to avoid expensive reflective 
+calls like \clj{getParent}, however if a specific overload cannot be found based on the
+surrounding static context, a type error is thrown.
 %
@@
 \end{minted}
-Since Java-level types lack nullability information,
-Typed Clojure conservatively asserts Java methods and
-constructor arguments cannot be passed \clj{nil}---this restriction can be
-adjusted per method. Method targets can never be \clj{nil}.
 
@@
 \end{minted}
+Since Java-level types lack nullability information,
+Typed Clojure conservatively assumes Java methods and
+constructor arguments cannot be passed \clj{nil}---this restriction can be
+adjusted per method, but method targets can never be \clj{nil}.
+
 %
@@
 
-\emph{Multimethods} are a kind of extensible function, and they are
-widely used to define Clojure operations.  It combines a
-\emph{dispatch function} with one or more \emph{methods}.
+\emph{Multimethods} are a kind of extensible function---combining a \emph{dispatch function} with 
+one or more \emph{methods}---widely used to define Clojure operations.
 
 \paragraph{Value-based dispatch}
-This simple multimethod says hello in different languages, as
-specified by a keyword argument.
+This simple multimethod says hello in different languages.%, as
+%specified by a keyword argument.
 
@@
 a \emph{dispatch value}. A method is then chosen
-based on the dispatch value---the arguments are then passed to the method
-to finally return a value for the entire expression.
+based on the dispatch value, to which the arguments are then passed to return a value.
 %
@@
 The multimethod dispatch rules use
-\clj{isa?}, a hybrid predicate which is a subclassing check for classes and
+\clj{isa?}, a hybrid predicate which is both a subclassing check for classes and
 an equality check for other values.
@@
 
-Beyond primitives and Java objects, the most common Clojure data
-structure is the 
-immutable hash-map, typicially with keyword keys. This structure is
-the primary way to represent compound data in Clojure programs.
-
-Keywords look themselves up in a map when called, or return \clj{nil}.
+%Beyond primitives and Java objects, 
+The most common way to represent compound data in Clojure 
+are immutable hash-maps, typicially with keyword keys.
+%
+Keywords double as functions that
+look themselves up in a map, or return \clj{nil} if absent.
 %
@@
 \end{minted}
-The inferred type for \clj{breakfast} holds two pieces of information---the known entries--\clj{:en}
-and \clj{:fr}---and their types, which are  \clj{:mandatory}, and that 
-no other key is present, since \clj{:complete?} is \clj{true}.
+This says
+\clj{:en} and \clj{:fr} are known entries mapped to strings,
+and the map is fully specified---that is, no other entries exist---by \clj{:complete?} being \clj{true}.
 
-HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
-\clj{'{:en Str :fr Str}}
-omits
-information about absent keys, only providing information on
-\clj{:mandatory} keys.
+HMap types default to partial specification, with
+\clj{'{:en Str :fr Str}} abbreviating \clj{(HMap :mandatory {:en Str, :fr Str})}.
 %
@@
 (defalias RawKeyPair ; extra keys disallowed
-  (HMap :mandatory {:pub RawKey, :priv RawKey}, :complete? true))
+  (HMap :mandatory {:pub RawKey, :priv RawKey}, 
+        :complete? true))
 (defalias EncKeyPair ; extra keys disallowed
@@
 (ann enc-keypair [RawKeyPair -> EncKeyPair])
-(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
-  (assoc (dissoc kp :priv) :enc-priv (encrypt pk)))
+(defn enc-keypair [kp]
+  (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
 \end{minted}
@@
 \begin{minted}{clojure}
-(defn enc-keypair-bad [{pk :priv :as kp}]
-  (assoc kp :enc-priv (encrypt pk))) ; Type error: :priv disallowed
+(defn enc-keypair-bad [kp] ; Type error: :priv disallowed
+  (assoc kp :enc-priv (encrypt (:priv kp))))
 \end{minted}
@@
 
-Since HMaps are the primary way of specifying the structure of data in Clojure,
-and multimethods are the primary tool for dispatching on data, they are inevitably
-linked.
-There are infinite ways of structuring and dispatching on data,
-so we cannot merely add a set of special case rules for
-handling these features. Instead, as type system designers, we must
-search for a compositional approach.
+HMaps and multimethods are the primary ways for representing
+and dispatching on data respectively, and so are intrinsically linked.
+As type system designers, we must
+search for a compositional approach that can anticipate
+any combination of these features.
 
@@
 
-Futhermore, this approach extends to multimethod dispatch---the primitive branching
-mechanism works like the humble \clj{if} conditional. 
-Only a small number of rules are needed
-to encode the \clj{isa?}-based dispatch, themselves made of simple pieces.
-In practice, this means that conditional-based control flow typing
+Futhermore, this approach extends to multimethod dispatch by reusing
+occurrence typing's approach to conditionals
+%, whose branching
+%mechanism may appear complex, but
+%can be understood in terms of the humble \clj{if} conditional. 
+and
+encoding a small number of rules to handle
+the \clj{isa?}-based dispatch.
+In practice, conditional-based control flow typing
 extends to multimethod dispatch, and vice-versa.
@@
 natural to dispatch on a known entry like \clj{:Meal}.
-
+%
 \begin{exmp}
@@
 (defmulti desserts :Meal)  ; dispatch on :Meal entry
-(defmethod desserts :lunch [{d :desserts}] d) ; d is :desserts entry
-(defmethod desserts :dinner [{d :desserts}] d)
-(defmethod desserts :combo [{m1 :meal1 m2 :meal2}] 
-  (+ (desserts m1) (desserts m2)))
+(defmethod desserts :lunch [o] (:desserts d))
+(defmethod desserts :dinner [o] (:desserts d))
+(defmethod desserts :combo [o] 
+  (+ (desserts (:meal1 o)) (desserts (:meal2 o))))
 
 (desserts {:Meal :combo, :meal1 {:Meal :lunch :desserts 1}, 
-           :meal2 {:Meal :dinner :desserts 2}})
-;=> 3
+           :meal2 {:Meal :dinner :desserts 2}}) ;=> 3
 \end{minted}
@@
 \end{exmp}
-
+%
 The \clj{:combo} method is verified to only structurally recur
-on \clj{Order}s. This is achieved because we learn the argument---\clj{o}---must % em dash needs space here, looks weird
+on \clj{Order}s. This is achieved because we learn the argument \clj{o} must
 be of type
@@
 \clj{(isa? (:Meal o) :combo)}
-must be true. Combining \clj{'{:Meal :combo}}
+is true. Combining this
 with the fact that \clj{o} is an \clj{Order}
@@
 orders, simplifying \clj{o} to
-\begin{minted}{clojure}
-'{:Meal ':combo :meal1 Order :meal2 Order}
-\end{minted}
+\clj{'{:Meal ':combo :meal1 Order :meal2 Order}}
 which contains appropriate arguments for both recursive calls.
@@
 \paragraph{Nested dispatch}
-An equally valid dispatch mechanism for \clj{desserts}
-would be on the \clj{class} of the \clj{:desserts} key.
+A more exotic dispatch mechanism for \clj{desserts}
+might be on the \clj{class} of the \clj{:desserts} key.
+If the result is a number, then we know the \clj{:desserts}
+key is a number, otherwise the input is a \clj{:combo} meal.
 We have already seen dispatch on \clj{class} and on keywords
@@
 (ann desserts' [Order -> Int])
-(defmulti desserts' (fn [o :- Order] (class (:desserts o))))
-(defmethod desserts' Long [{d :desserts :as o}] 
-  ; o :- (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner, :desserts Int}
-  ;         '{:Meal ':combo :desserts Int, :meal1 Order, :meal2 Order})
-  d)
-(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
+(defmulti desserts' 
+  (fn [o :- Order] (class (:desserts o))))
+(defmethod desserts' Long [o] 
+; o :- (U '{:Meal (U ':dinner ':lunch), :desserts Int}
+;       '{:Meal ':combo, :desserts Int, :meal1 Order, :meal2 Order})
+  (:desserts o))
+(defmethod desserts' nil [o]
   ; o :- '{:Meal ':combo, :meal1 Order, :meal2 Order}
-  (+ (desserts' m1) (desserts' m2)))
+  (+ (desserts' (:meal1 o)) (desserts' (:meal2 o))))
 \end{minted}
@@
 its argument is at least of type \clj{'{:desserts Long}}---since
-\begin{minted}{clojure}
-(isa? (class (:desserts o)) Long)
-\end{minted}
+\clj{(isa? (class (:desserts o)) Long)}
 must be true.
@@
 %\clj{Order},
-In this method, we deduce
-the entry \clj{:desserts}
-\emph{must} be a present and mapped to a \clj{Long},
-even in a \clj{:combo} meal
-which does not state \clj{:desserts}
+Here the
+\clj{:desserts} entry
+\emph{must} be present and mapped to a \clj{Long}---even in a \clj{:combo} meal,
+which does not specify \clj{:desserts}
 as present or absent.
@@
 
-\paragraph{From multimethod to multiple dispatch}
+\paragraph{From multiple to arbitrary dispatch}
 Clojure multimethod dispatch, and Typed Clojure's handling of it, goes

### 2015-10-17 small.tex
review first half

@@
 \usepackage{url}
+% seems like a nightmare to use, but you get clickable links
+%\usepackage{hyperref}
+
 \newcounter{ex}
@@
 \par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
-\\[-2.5em]\nopagebreak}
+\\[-3.23em]\nopagebreak}
 {\relax}
@@
 soundness. Additionally, Typed Clojure is now in use by numerous
-corporations and developers working with Clojure, and we report on 
-a quantitative analysis of the use of the features described in this
-paper in two substantial code bases.
+corporations and developers working with Clojure, and we present
+a quantitative analysis on the use of type system features
+in two substantial code bases.
 

### 2015-10-17 formal-model.tex
add tejas, no table

@@
 The branches are first checked with T-False and T-True respectively,
-then T-Subsume.
+the T-Subsume premises
+\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
+and
+\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
+allow us to pick compatible propositions for both branches.
 %$$
@@
 %{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
-The T-Subsume premises
-\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
-and
-\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
-allow us to pick compatible propositions for both branches.
-
-Finally T-Abs concludes, using the T-If outputs, with
+
+Finally T-Abs assigns a type to the overall function:
 $$
@@
 
-\egref{example:getparent-direct-constructor} \clj{(.getParent (new File "a/b"))}
+In \egref{example:getparent-direct-constructor}, \clj{(.getParent (new File "a/b"))}
 translates to
@@
 %Multimethods can error if no matching method is chosen (rules in the supplemental material).
-Below is the translation of \egref{example:hi-multimethod} to \lambdatc{}.
+\figref{main:figure:mmexample}
+is the translation of \egref{example:hi-multimethod} to \lambdatc{}.
 %\begin{minted}{clojure}
@@
 
-\begin{listing}
+\begin{figure}
 $\letexp{hi_0} {\createmultiexp {\ArrowOne{\x{}}{\Keyword}{\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}} {\abs{\x{}}{\Keyword}{\x{}}}}
@@
 $
-\end{listing}
+\caption{Multimethod example}
+\label{main:figure:mmexample}
+\end{figure}
 %
@@
 \paragraph{Subtyping}
-Multimethods are also functions, which is encoded via S-PMultiFn.
+Multimethods are functions, via S-PMultiFn,
 %$$
@@
 %$$
-This rule says a multimethod can be upcast to its interface type. This means
-multimethod call sites can be handled by T-App via T-Subsume. Other rules are given in
+which says a multimethod can be upcast to its interface type. 
+Multimethod call sites are then handled by T-App via T-Subsume. Other rules are given
 in \figref{main:figure:mmsyntax}.
@@
 heterogeneous map types.
-The partially specified map of
-\clj{lunch} in \egref{example:lunchpartial}
-is written
-\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}.
-We abbreviate this type as \Lunch in this section.
 The type \HMapgeneric{\mandatory{}}{\absent{}}
@@
 and {\partial{}} (``partial'') if there are \emph{unknown} entries.
-
+%
+The partially specified map of
+\clj{lunch} in \egref{example:lunchpartial}
+is written
+\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}.
+We abbreviate this type as \Lunch in this section.
+%
 The type of the fully specified map

### 2015-10-17 related-work.tex
It fits.
add tejas, no table

@@
 
+TeJaS~\cite{TeJaS}, another type system for JavaScript,
+also supports similar HMaps, with the ability to
+record the presence and absence of entries, but lacks a compositional
+flow-checking approach like occurrence typing.
+
 Typed Lua~\cite{Maidl:2014:TLO} has \emph{table types} which track
@@
 keys normal.
-
-The integration of completeness information, crucial for many examples
-in Typed Clojure, is not provided by any of these systems.
 
@@
 %runtime cost, 
-have been  adopted in industry, including Hack for PHP~\cite{hack}, and Flow~\cite{flow} and
+have been  adopted in industry, including Hack~\cite{hack}, and Flow~\cite{flow} and
 TypeScript~\cite{typescript}, two extensions of JavaScript. These

### 2015-12-03 esop-rebuttal.txt
esop draft rebuttal

Major issues:

1) Section 5.1 should use tables instead of prose.

We will restate the data in tables.

==========================================================================

Q: What is the delta between occurrence typing in ICFP'10 and the as presented here?

A:

Q: Is TC decidable?

Q: Why can't the type system check Example 9? What restrictions are imposed on HMaps?

A: For presentational reasons, we only support a specific pattern of looking up
a known key in a union of HMaps...

Q: What specific patterns could not be supported by TC that led to CircleCI
   abandoning it?

Q: How is TC compiled/used in practice? Is it sealed from untyped namespaces?

A: core.typed

Q: "Java cannot be passed nil"?

A: This is a conservative assumption, as the paper states in the same sentence.
   In practice this is a reasonable default, as discussed in the CircleCI evaluation (5.1).

Q: Does (dissoc k) just remove its argument-entry from a map?

A: (dissoc m k) returns a new map that is like m, but without a k entry.

Q: How symbolic representations of objects in OT relate to Scala's
   path-dependent types?

A: ??

Q: Formula (3) -- do the positive/negative statements mean that d
   is of type Nat, so N_d should be read as a proposition?

A: Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.

Q: Are Java's generic classes (in particular collections) supported
   in TC type checking algorithm?

A: No. We chose to avoid the added complexity of Java generics ..

Q: How do Clojure IDE's "incrementalize" type checking for code completion?

A: Cursive Clojure prioritizes type hints assigned to variables in the local scope
   during method name autocompletion.

Q: "Then (new[S] F "a/b") is checked against a nilable conversion..."?

A: FIXME I have no idea, looks like a bug?

Q: What is the type of (defmethod e_m e_v e_f)?

A: As shown in T-DefMethod, it returns the same type as e_m.

Q: How overlapping multi-method clauses are handled?

A: In the type system, each method is checked in isolation with respect to 
   just the current dispatch value and the dispatch function type.
   In the operational semantics, B-BetaMulti can only fire if there is an unambiguous
   match, encoded in the GM metafunction.

Q: Citation for Footnote 1?

A: FIXME

Q: What category do you consider e.g. C# 4.0, or Scala's Dynamic, or
   Dart? (WRT gradual typing)

Q: Is the additional of an Unknown or Dynamic type in a static
    type system enough to make it gradual?

Q: Is "optional" here a synonym for Bracha's "pluggable" type system?

A: Yes, because the type system does not affect the final program.
  ... But also no...

Q: One way of thinking about those
   annotations is that in fact Typed Clojure actually is a gradual type
   system as well as an optional type system, with the annotations
   effectively acting as type assumptions which will be checked
   dynamically to interact with other untyped code.

A: Typed Clojure does not check these assumptions at runtime, thus is not
   gradually typed in any sense, just optionally typed like TypeScript.

Q: What is "ann"?

A: (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.

Q: What is the "Any" type?

A: "Any" is effectively the same as (U nil Object), and has no special runtime semantics.
   This also holds in the implementation, as JVM primitives pun their boxed equivalents
   as static types, reflecting how Clojure deals with primitives.

Q: Why does 42 have the object "0"?

A: There is no known way to derive 42 from the current environment, so the object is "empty".
   More recent forumlations of occurrence typing assign the object "42" to aid symbolic
   computations.

Q: Is the complexity of multimethods worth it, given the case studies?

A:

### 2015-12-04 esop-rebuttal.txt
esop rebuttal

@@
-Major issues:
+We thank the reviewers for their detailed feedback.  We are thrilled the
+reviewers found the paper "a joy to read", "an excellent demonstration 
+of occurrence typing's power", and providing a "substantial practical 
+contribution".
 
-1) Section 5.1 should use tables instead of prose.
+The reviewers' primary concerns are with two issues:
 
-We will restate the data in tables.
+1. What are the contributions beyond occurrence typing as presented
+   in ICFP'10?
+
+There are two main extensions in relation to occurrence typing.
+
+The first is the presentation of HMap's, with support for flow checking
+on sub-parts of maps. This required several extra typing rules (Fig. 10)
+and subtle, but clean, extensions to the `update` rule (Fig. 12, clauses
+4-7).
+
+The second is multimethods, which is mainly contained in the T-DefMethod
+typing rule. It is in the same spirit as T-If, which was present in ICFP'10,
+but demonstrates occurrence typing extends gracefully to support complex 
+control flow patterns.
+
+The main takeaway is that occurrence typing as formulated in ICFP'10
+serves as a valuable base to support for the nuances of a Clojure type system,
+without any particular foresight from the original authors.
+
+2. Are multimethods worth the extra complexity in the type system, given
+   they are used so little in practice?
+
+Yes.
+
+The numbers presented in the evaluation section are possibly misleading.
+Even though a low number of multimethods are defined in practice, extending
+them via `defmethod` accounts for a large amount of code.
+
+A comparable analogy to a Java code base might be assuming interfaces are
+a minor feature if a small number of interfaces are defined, yet a large number
+of classes extend them.
+
+As noted in Section 5.1, these multimethods are single-dispatch on a
+fixed map entry, which further gives them similar roles as Java interfaces,
+which perform single dispatch on the first argument of methods.
+
+Following are responses to specific statements or queries.
 
@@
 
-Q: What is the delta between occurrence typing in ICFP'10 and the as presented here?
+Q. Section 5.1 should use tables instead of prose.
 
-A:
+A. We will restate the data in tables.
 
-Q: Is TC decidable?
+Q: Is TC decidable? `restrict` seems problematic.
+
+A: The calculus leaves the details of `restrict` up to implementations. core.typed
+   features an ad-hoc intersection type, so `restrict` simply returns the intersection 
+   of both types, thus making the occurrence typing implementation decidable.
 
@@
 
-A: For presentational reasons, we only support a specific pattern of looking up
-a known key in a union of HMaps...
+A: For presentational reasons we only support a specific pattern of looking up
+   a common, known key in a union of HMaps (notice T-GetHMap (Fig. 10) is the only rule
+   supporting lookups on unions of HMaps). Example 9 is checkable in core.typed, 
+   but not the calculus, and serves as a demonstration of the kinds of patterns
+   that might occur in Clojure code.
 
@@
 
+A: We refer to the detailed blog post they published:
+
+     http://blog.circleci.com/why-were-no-longer-using-core-typed/
+
+Q: Are the case studies representative programs that would be good candidates for optiona
+   typing?
+
+A: We chose the CircleCI code base as it was a substantial size. The other case
+   study was a randomly chosen open source project. A goal of Typed Clojure is to support
+   as many Clojure idioms as possible, and we learnt valuable lessons from both projects
+   in this regard.
+
 Q: How is TC compiled/used in practice? Is it sealed from untyped namespaces?
 
-A: core.typed
+A: TC is simply compiled to be normal Clojure code. As alluded to in Footnote 1,
+   Typed Clojure is not gradually typed.
 
@@
 
-Q: How symbolic representations of objects in OT relate to Scala's
+Q: How do symbolic representations of objects in OT relate to Scala's
    path-dependent types?
 
-A: ??
+A: They are unrelated.
 
@@
 
-A: No. We chose to avoid the added complexity of Java generics ..
+A: No. TC however supports parametric polymorphism for Clojure collections and a manual subset
+   of Java collections. In practice, this is sufficient for many interop situations.
 
-Q: How do Clojure IDE's "incrementalize" type checking for code completion?
+Q: The simplified syntax chosen for method calls works for compilers, but
+   not IDE's. How do Clojure IDE's "incrementalize" type checking for code completion?
 
-A: Cursive Clojure prioritizes type hints assigned to variables in the local scope
+A: This syntax is not arbitrary; (.method o) is sugar for (. o (method)) in Clojure.
+   Cursive Clojure prioritizes type hints assigned to variables in the local scope
    during method name autocompletion.
@@
 
-A: FIXME I have no idea, looks like a bug?
+A: The prose should be fixed to instead read "non-nilable".
 
@@
 
-A: FIXME
+A: Siek et al SNAPL'15 is a good candidate for describing gradual typing, however optional
+   typing is not explicitly defined.
 
@@
 
+A: C# 4.0 is gradually typed. Scala's Dynamic is an unrelated metaprogramming construct, 
+   comparable to a statically resolved "method_missing" in Ruby.  Dart is 
+   optionally typed.
+
 Q: Is the additional of an Unknown or Dynamic type in a static
@@
 
+A: A gradual type system is defined by its operational semantics, in particular
+   how the runtime prevents untyped code from breaking static invariants.
+
 Q: Is "optional" here a synonym for Bracha's "pluggable" type system?
 
-A: Yes, because the type system does not affect the final program.
-  ... But also no...
+A: They are in the same spirit.
 
-Q: One way of thinking about those
-   annotations is that in fact Typed Clojure actually is a gradual type
-   system as well as an optional type system, with the annotations
-   effectively acting as type assumptions which will be checked
+Q: One way of thinking about those annotations is that in fact Typed Clojure 
+   actually is a gradual type system as well as an optional type system, with 
+   the annotations effectively acting as type assumptions which will be checked
    dynamically to interact with other untyped code.
@@
 A: (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.
+   We will explain this explicitly.
 
-Q: What is the "Any" type?
+Q: What is the "Any" type? "Kw"?
 
 A: "Any" is effectively the same as (U nil Object), and has no special runtime semantics.
-   This also holds in the implementation, as JVM primitives pun their boxed equivalents
-   as static types, reflecting how Clojure deals with primitives.
+   It is *not* a Dynamic type.
+   "Kw" is the type for keywords.
 
@@
 
-A: There is no known way to derive 42 from the current environment, so the object is "empty".
-   More recent forumlations of occurrence typing assign the object "42" to aid symbolic
-   computations.
-
-Q: Is the complexity of multimethods worth it, given the case studies?
-
-A:
+A: There is no known way to derive 42 from the current environment, so we say the 
+   object is "empty".

### 2015-12-04 esop-rebuttal.txt
move gradual typing to major section, respond to specific reviews

@@
 
-The reviewers' primary concerns are with two issues:
+The reviewers' primary concerns are with three issues:
 
@@
 
-There are two main extensions in relation to occurrence typing.
+As described in Section 1, our overall contributions are
+
+  - a type system for Clojure that support Clojure idioms,
+  - a sound formal model for three crucial type system features, and
+  - an evaluation of the type system in practice.
+
+There are two main extensions specifically in relation to occurrence typing.
 
@@
 
-Following are responses to specific statements or queries.
+3. What is the distinction between optional and gradual typing?
+
+Optional typing refers to type systems that do not enforce type invariants
+at runtime, and are comparable to Bracha's "pluggable" type systems (as
+suggested by Review 2).
+
+Typed Clojure does not check assumptions at runtime, thus is not
+gradually typed in any sense, just optionally typed like TypeScript.
+
+Review 2 requests to categorize the following:
+ - C# 4.0 is gradually typed. 
+ - Scala's Dynamic is an unrelated metaprogramming construct, 
+   comparable to a statically resolved "method_missing" in Ruby. 
+ - Dart is optionally typed.
+
+A gradual type system is defined by its operational semantics, in particular
+how the runtime prevents untyped code from breaking static invariants. An
+addition of an "Unknown or Dynamic type" (Review 2) is not enough to make
+a gradual type system.
+
+We distinguish optional and gradual typing in footnote 1. A good reference for 
+this distinction is Siek et al SNAPL'15 (as requested by Review 2).
+
+"Any" is effectively the same as `(U nil Object)`, and has no special runtime semantics.
+It is *not* a Dynamic type (Review 2).
 
@@
 
-Q. Section 5.1 should use tables instead of prose.
+Following are responses to specific statements or queries.
 
-A. We will restate the data in tables.
+REVIEW 1: Section 5.1 should use tables instead of prose.
 
-Q: Is TC decidable? `restrict` seems problematic.
+  We will restate the data in tables.
 
-A: The calculus leaves the details of `restrict` up to implementations. core.typed
+REVIEW 1: Is TC decidable? `restrict` seems problematic.
+
+  The calculus leaves the details of `restrict` up to implementations. core.typed
   features an ad-hoc intersection type, so `restrict` simply returns the intersection 
@@
 
-Q: Why can't the type system check Example 9? What restrictions are imposed on HMaps?
+REVIEW 1: Why can't the type system check Example 9? 
+          What restrictions are imposed on HMaps?
 
-A: For presentational reasons we only support a specific pattern of looking up
+  For presentational reasons we only support a specific pattern of looking up
   a common, known key in a union of HMaps (notice T-GetHMap (Fig. 10) is the only rule
@@
 
-Q: What specific patterns could not be supported by TC that led to CircleCI
+REVIEW 1: What specific patterns could not be supported by TC that led to CircleCI
           abandoning it?
 
-A: We refer to the detailed blog post they published:
+  We refer to the detailed blog post they published:
 
@@
 
-Q: Are the case studies representative programs that would be good candidates for optiona
-   typing?
+REVIEW 3: Are the case studies representative programs that would be good candidates 
+          for optional typing?
 
-A: We chose the CircleCI code base as it was a substantial size. The other case
+  We chose the CircleCI code base as it was a substantial size. The other case
   study was a randomly chosen open source project. A goal of Typed Clojure is to support
@@
 
-Q: How is TC compiled/used in practice? Is it sealed from untyped namespaces?
+REVIEW 1: How is TC compiled/used in practice? Is it sealed from untyped namespaces?
 
-A: TC is simply compiled to be normal Clojure code. As alluded to in Footnote 1,
+  TC is compiled to as normal Clojure code. As alluded to in Footnote 1,
   Typed Clojure is not gradually typed.
 
-Q: "Java cannot be passed nil"?
+REVIEW 1: "Java cannot be passed nil"?
 
-A: This is a conservative assumption, as the paper states in the same sentence.
+  This is a conservative assumption, as the paper states in the same sentence.
   In practice this is a reasonable default, as discussed in the CircleCI evaluation (5.1).
 
-Q: Does (dissoc k) just remove its argument-entry from a map?
+REVIEW 1: Does (dissoc k) just remove its argument-entry from a map?
 
-A: (dissoc m k) returns a new map that is like m, but without a k entry.
+  (dissoc m k) returns a new map that is like m, but without a k entry.
 
-Q: How do symbolic representations of objects in OT relate to Scala's
+REVIEW 1: How do symbolic representations of objects in OT relate to Scala's
           path-dependent types?
 
-A: They are unrelated.
+  They are unrelated.
 
-Q: Formula (3) -- do the positive/negative statements mean that d
+REVIEW 1: Formula (3) -- do the positive/negative statements mean that d
           is of type Nat, so N_d should be read as a proposition?
 
-A: Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.
+  Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.
 
-Q: Are Java's generic classes (in particular collections) supported
+REVIEW 1: Are Java's generic classes (in particular collections) supported
           in TC type checking algorithm?
 
-A: No. TC however supports parametric polymorphism for Clojure collections and a manual subset
+  No. TC however supports parametric polymorphism for Clojure collections and a manual subset
   of Java collections. In practice, this is sufficient for many interop situations.
 
-Q: The simplified syntax chosen for method calls works for compilers, but
+REVIEW 1: The simplified syntax chosen for method calls works for compilers, but
           not IDE's. How do Clojure IDE's "incrementalize" type checking for code completion?
 
-A: This syntax is not arbitrary; (.method o) is sugar for (. o (method)) in Clojure.
+  This syntax is not arbitrary; (.method o) is sugar for (. o (method)) in Clojure.
   Cursive Clojure prioritizes type hints assigned to variables in the local scope
@@
 
-Q: "Then (new[S] F "a/b") is checked against a nilable conversion..."?
+REVIEW 1: "Then (new[S] F "a/b") is checked against a nilable conversion..."?
 
-A: The prose should be fixed to instead read "non-nilable".
+  The prose should be fixed to instead read "non-nilable".
 
-Q: What is the type of (defmethod e_m e_v e_f)?
+REVIEW 1: What is the type of (defmethod e_m e_v e_f)?
 
-A: As shown in T-DefMethod, it returns the same type as e_m.
+  As shown in T-DefMethod, it returns the same type as e_m.
 
-Q: How overlapping multi-method clauses are handled?
+REVIEW 1: How overlapping multi-method clauses are handled?
 
-A: In the type system, each method is checked in isolation with respect to 
+  In the type system, each method is checked in isolation with respect to 
   just the current dispatch value and the dispatch function type.
@@
 
-Q: Citation for Footnote 1?
+REVIEW 2: What is "ann"?
 
-A: Siek et al SNAPL'15 is a good candidate for describing gradual typing, however optional
-   typing is not explicitly defined.
-
-Q: What category do you consider e.g. C# 4.0, or Scala's Dynamic, or
-   Dart? (WRT gradual typing)
-
-A: C# 4.0 is gradually typed. Scala's Dynamic is an unrelated metaprogramming construct, 
-   comparable to a statically resolved "method_missing" in Ruby.  Dart is 
-   optionally typed.
-
-Q: Is the additional of an Unknown or Dynamic type in a static
-   type system enough to make it gradual?
-
-A: A gradual type system is defined by its operational semantics, in particular
-   how the runtime prevents untyped code from breaking static invariants.
-
-Q: Is "optional" here a synonym for Bracha's "pluggable" type system?
-
-A: They are in the same spirit.
-
-Q: One way of thinking about those annotations is that in fact Typed Clojure 
-   actually is a gradual type system as well as an optional type system, with 
-   the annotations effectively acting as type assumptions which will be checked
-   dynamically to interact with other untyped code.
-
-A: Typed Clojure does not check these assumptions at runtime, thus is not
-   gradually typed in any sense, just optionally typed like TypeScript.
-
-Q: What is "ann"?
-
-A: (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.
+  (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.
   We will explain this explicitly.
 
-Q: What is the "Any" type? "Kw"?
+REVIEW 2: What is the "Kw" type?
 
-A: "Any" is effectively the same as (U nil Object), and has no special runtime semantics.
-   It is *not* a Dynamic type.
    "Kw" is the type for keywords.
 
-Q: Why does 42 have the object "0"?
+REVIEW 2: Why does 42 have the object "0"?
 
-A: There is no known way to derive 42 from the current environment, so we say the 
+  There is no known way to derive 42 from the current environment, so we say the 
   object is "empty".

### 2015-12-04 esop-rebuttal.txt
practical contribution

@@
   - a type system for Clojure that support Clojure idioms,
-  - a sound formal model for three crucial type system features, and
+  - a sound formal model for three crucial type system features,
+  - an implementation successfully used in practice and industry, and
   - an evaluation of the type system in practice.

### 2015-12-04 esop-rebuttal.txt
will revise paper according to reviews

@@
 
-Following are responses to specific statements or queries.
+We now address a few specific questions from the reviewers. The
+reviewers also make several suggestions for presentation revisions and
+improvements to the evaluation, including comparisons to other systems---we 
+will make these changes in a revised version.
 

### 2015-12-05 esop-rebuttal.txt
Edits.

@@
-We thank the reviewers for their detailed feedback.  We are thrilled the
+We thank the reviewers for their detailed feedback.  We're happy the
 reviewers found the paper "a joy to read", "an excellent demonstration 
-of occurrence typing's power", and providing a "substantial practical 
+of occurrence typing's power", and provided "substantial practical 
 contribution".
 
-The reviewers' primary concerns are with three issues:
+We begin with three high-level issues:
 
@@
 
-  - a type system for Clojure that support Clojure idioms,
+  - a type system for Clojure supporting Clojure idioms,
   - a sound formal model for three crucial type system features,
@@
 
-There are two main extensions specifically in relation to occurrence typing.
+There are two main extensions specifically related to occurrence typing.
 
-The first is the presentation of HMap's, with support for flow checking
-on sub-parts of maps. This required several extra typing rules (Fig. 10)
-and subtle, but clean, extensions to the `update` rule (Fig. 12, clauses
-4-7).
+The first is the presentation of HMap's, with support for flow
+checking on sub-parts of maps. This required several extra typing
+rules (Fig. 10) and subtle extensions to the `update` rule (Fig. 12,
+clauses 4-7).
 
-The second is multimethods, which is mainly contained in the T-DefMethod
-typing rule. It is in the same spirit as T-If, which was present in ICFP'10,
-but demonstrates occurrence typing extends gracefully to support complex 
-control flow patterns.
+The second is multimethods, which is mainly contained in the
+T-DefMethod typing rule. It reuses the machinery T-If, which was
+present in ICFP'10, but demonstrates occurrence typing is able to
+support new and previously-unhandled control flow.
 
-The main takeaway is that occurrence typing as formulated in ICFP'10
-serves as a valuable base to support for the nuances of a Clojure type system,
-without any particular foresight from the original authors.
-
-2. Are multimethods worth the extra complexity in the type system, given
-   they are used so little in practice?
+2. Are multimethods worth the extra complexity in the type system, if
+   they are used rarely?
 
@@
 
-The numbers presented in the evaluation section are possibly misleading.
-Even though a low number of multimethods are defined in practice, extending
-them via `defmethod` accounts for a large amount of code.
+The numbers presented in the evaluation section do not indicate that
+multimethods are are.  Even though a low number of distinct
+multimethods are defined in practice, extending them via `defmethod`
+accounts for a large amount of code.
 
-A comparable analogy to a Java code base might be assuming interfaces are
-a minor feature if a small number of interfaces are defined, yet a large number
-of classes extend them.
+Analagously, though Java code may define few interfaces, they will be
+used in more class definitions, and then in many method calls.
 
-As noted in Section 5.1, these multimethods are single-dispatch on a
-fixed map entry, which further gives them similar roles as Java interfaces,
-which perform single dispatch on the first argument of methods.
+As noted in Section 5.1, these multimethods are often single-dispatch
+on a fixed map entry, making them even more similar to Java
+interfaces, which perform single dispatch on the first argument of
+methods.
 
@@
 Typed Clojure does not check assumptions at runtime, thus is not
-gradually typed in any sense, just optionally typed like TypeScript.
+gradually typed, just optionally typed like TypeScript.
 
 Review 2 requests to categorize the following:
- - C# 4.0 is gradually typed. 
- - Scala's Dynamic is an unrelated metaprogramming construct, 
+ - C# 4.0: gradually typed. 
+ - Scala's Dynamic: a mostly-unrelated metaprogramming construct, 
    comparable to a statically resolved "method_missing" in Ruby. 
- - Dart is optionally typed.
+ - Dart: optionally typed.
 
 A gradual type system is defined by its operational semantics, in particular
-how the runtime prevents untyped code from breaking static invariants. An
+that the runtime prevents untyped code from breaking static invariants. An
 addition of an "Unknown or Dynamic type" (Review 2) is not enough to make
-a gradual type system.
+a gradual type system without this enforcement.
 
-We distinguish optional and gradual typing in footnote 1. A good reference for 
-this distinction is Siek et al SNAPL'15 (as requested by Review 2).
+We distinguish optional and gradual typing in footnote 1; see [Siek et
+al, SNAPL'15] for more on the distinction.
 
-"Any" is effectively the same as `(U nil Object)`, and has no special runtime semantics.
-It is *not* a Dynamic type (Review 2).
+"Any" is the top type, effectively the same as `(U nil Object)`, and
+has no special runtime semantics.  It is *not* a Dynamic type (Review
+2).
 
@@
 
-REVIEW 1: Section 5.1 should use tables instead of prose.
-
-  We will restate the data in tables.
-
 REVIEW 1: Is TC decidable? `restrict` seems problematic.
 
-  The calculus leaves the details of `restrict` up to implementations. core.typed
-  features an ad-hoc intersection type, so `restrict` simply returns the intersection 
-  of both types, thus making the occurrence typing implementation decidable.
+  The calculus leaves the details of `restrict` up to
+  implementations. core.typed features an ad-hoc intersection type, so
+  `restrict` simply returns the intersection of both types, thus
+  making this function simple.
 
@@
 
-  For presentational reasons we only support a specific pattern of looking up
-  a common, known key in a union of HMaps (notice T-GetHMap (Fig. 10) is the only rule
-  supporting lookups on unions of HMaps). Example 9 is checkable in core.typed, 
-  but not the calculus, and serves as a demonstration of the kinds of patterns
+  For presentational reasons we only support a specific pattern of
+  looking up a common, known key in a union of HMaps (notice T-GetHMap
+  (Fig. 10) is the only rule supporting lookups on unions of
+  HMaps). Example 9 is checkable in our implementation, but not the
+  calculus, and serves as a demonstration of the kinds of patterns
   that might occur in Clojure code.
 
-REVIEW 1: What specific patterns could not be supported by TC that led to CircleCI
-          abandoning it?
+REVIEW 1: What specific patterns could not be supported by TC that led
+          to CircleCI abandoning it?
 
@@
 
-REVIEW 3: Are the case studies representative programs that would be good candidates 
-          for optional typing?
+REVIEW 3: Are the case studies representative programs that would be
+          good candidates for optional typing?
 
-  We chose the CircleCI code base as it was a substantial size. The other case
-  study was a randomly chosen open source project. A goal of Typed Clojure is to support
-  as many Clojure idioms as possible, and we learnt valuable lessons from both projects
-  in this regard.
+  We chose the CircleCI code base as it was a substantial size. The
+  other case study was a randomly chosen open source project. A goal
+  of Typed Clojure is to support as many Clojure idioms as possible,
+  and we learnt valuable lessons from both projects in this regard.
 
-REVIEW 1: How is TC compiled/used in practice? Is it sealed from untyped namespaces?
+REVIEW 1: How is TC compiled/used in practice? Is it sealed from
+untyped namespaces?
 
-  TC is compiled to as normal Clojure code. As alluded to in Footnote 1,
-  Typed Clojure is not gradually typed.
+  TC is compiled to as normal Clojure code. As discussed above, Typed
+  Clojure is not gradually typed.
 
@@
 
-  This is a conservative assumption, as the paper states in the same sentence.
-  In practice this is a reasonable default, as discussed in the CircleCI evaluation (5.1).
+  This is a conservative assumption, as we mention.  In practice this
+  is a reasonable default, as discussed in the CircleCI evaluation
+  (5.1).
 
@@
 
-  No. TC however supports parametric polymorphism for Clojure collections and a manual subset
-  of Java collections. In practice, this is sufficient for many interop situations.
+  No. TC however supports parametric polymorphism for Clojure
+  collections and a manual subset of Java collections. In practice,
+  this is sufficient for most interop situations.
 
-REVIEW 1: The simplified syntax chosen for method calls works for compilers, but
-          not IDE's. How do Clojure IDE's "incrementalize" type checking for code completion?
+REVIEW 1: The simplified syntax chosen for method calls works for
+          compilers, but not IDE's. How do Clojure IDE's
+          "incrementalize" type checking for code completion?
 
-  This syntax is not arbitrary; (.method o) is sugar for (. o (method)) in Clojure.
-  Cursive Clojure prioritizes type hints assigned to variables in the local scope
-  during method name autocompletion.
+  This is not our choice of syntax; (.method o) is sugar for (. o
+  (method)) in Clojure.  The Cursive Clojure IDE prioritizes type
+  hints assigned to variables in the local scope during method name
+  autocompletion.
 
@@
 
-REVIEW 1: How overlapping multi-method clauses are handled?
+REVIEW 1: How are overlapping multi-method clauses handled?
 
-  In the type system, each method is checked in isolation with respect to 
-  just the current dispatch value and the dispatch function type.
-  In the operational semantics, B-BetaMulti can only fire if there is an unambiguous
-  match, encoded in the GM metafunction.
+  In the type system, each method is checked in isolation with respect
+  to the current dispatch value and the dispatch function type.  In
+  the operational semantics, B-BetaMulti can only fire if there is an
+  unambiguous match, encoded in the GM metafunction.
 
@@
 
-  (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.
-  We will explain this explicitly.
+  (ann v t) 'ann'otates variable v as type t. We will explain this
+  explicitly.
 
@@
 
-  There is no known way to derive 42 from the current environment, so we say the 
-  object is "empty".
+  Objects only describe references to the runtime environment, so we
+  say the object of 42 is "empty".

### 2015-12-26 formal-model.tex
fix italic gamma

@@
 overloading or Java Generics.
-
+%
 We extend the syntax in \figref{main:figure:javatyping} with Java field lookups and calls to
 methods and constructors. 
-To prevent ambiguity, % between zero-argument methods and fields, 
+To prevent ambiguity between zero-argument methods and fields,
+we use Clojure's primitive ``dot'' syntax:
 field accesses are written \fieldexp{\fld{}}{\e{}}
@@
 %$$
-We use T-Method to check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}, 
-which checks unresolved methods.
-We verify the target type \s{} = \File is non-nil before erasing \Nil
-by converting to a Java type \classhint{1} = \File.
-The specific overload is chosen from the class table based on \classhint{1}---there is at most one.
-Then {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
-is checked against a nilable conversion of \classhint{1}, \t{1} = \Union{\Nil}{\File},
-which succeeds by the previous rule.
-The nilable return type \t{} = \Union{\Nil}{\String} is given, and---finally---the 
+The T-Method rule for unresolved methods
+checks {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
+We verify the target type \s{} = \File is non-nil by T-New.
+The overload is chosen from the class table based on \classhint{1} = \File---there is at most one.
+The nilable return type \t{} = \Union{\Nil}{\String} is given, and the 
 entire expression rewrites to expression \ref{eq:resolved}.
@@
 
-The T-Field rule is included in \figref{main:figure:javatyping}, and is like T-Method, but without arguments.
+The T-Field rule (\figref{main:figure:javatyping}) is like T-Method, but without arguments.
 
@@
 
-\figref{main:figure:mmsyntax}
-presents \emph{immutable}
-multimethods without default methods to ease presentation.
+\figref{main:figure:mmsyntax} presents \emph{immutable} multimethods without default methods to ease presentation.
 %Syntax and semantics are given in \figref{main:figure:mmsyntax}. 
 %Multimethods can error if no matching method is chosen (rules in the supplemental material).
-\figref{main:figure:mmexample}
-is the translation of \egref{example:hi-multimethod} to \lambdatc{}.
+\figref{main:figure:mmexample} translates the mutable \egref{example:hi-multimethod} to \lambdatc{}.
 %\begin{minted}{clojure}

### 2015-12-30 formal-model.tex
more TODO's

@@
 which encodes the type, proposition, and object information
-about variables.
+about variables. The proposition {\notprop{\falsy}{\makelocal{d}}}
+says ``it is not the case that variable {\makelocal{d}} is of type {\falsy}'';
+{\isprop{\falsy}{\makelocal{d}}} says ``{\makelocal{d}} is of type {\falsy}''.
 
 Finally, the T-App rule substitutes the operand's object \objectp{}
-for the parameter \x{} in the latent type, propositions, and object.
+for the parameter \x{} in the latent type, propositions, and object. The proposition
+{\isprop{\Number{}}{\makelocal{d}}} says ``{\makelocal{d}} is of type {\Number{}}'';
+{\notprop{\Number{}}{\makelocal{d}}} says ``it is not the case that {\makelocal{d}}
+is of type {\Number{}}''. The object {\makelocal{d}} is the symbolic representation
+of what the expression {\makelocal{d}} evaluates to.
 \begin{equation}
@@
 
-Next, show how to check
+Next, we show how to check
 $
@@
 and \object{} = \emptyobject{}, which also agrees with the method body, above.
+Notice the overall type of a \defmethodliteral{} is the same as its first subexpression \e{m}.
 

### 2015-12-30 overview.tex
fix italic gamma
more TODO's

@@
 \end{exmp}
-
+%
 %(ann enc-keypair [RawKeyPair -> EncKeyPair])
@@
 %         :enc-priv (encrypt pk))) ; add encrypted private key
-
+%
 %\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
-As \clj{EncKeyPair} is fully specified, we must remove extra keys like \clj{:priv}.
+As \clj{EncKeyPair} is fully specified, we remove extra keys like \clj{:priv}
+via \clj{dissoc}, which returns a new map that is the first argument without the
+entry named by the second argument. Notice removing \clj{dissoc} causes a type error.
 %

### 2015-12-31 metatheory.tex
error values

@@
 
-We prove type soundness following \cite{TF10}.  Our model is extended
-to include errors and a \wrong{} value, and we prove well-typed
+We prove type soundness following Tobin-Hochstadt and Felleisen~\cite{TF10}.  Our model is extended
+to include errors \errorvalv{} and a \wrong{} value, and we prove well-typed
 programs do not go wrong; this is therefore a stronger theorem than
-proved by \cite{TF10}.
+proved by Tobin-Hochstadt and Felleisen~\cite{TF10}. 
+Errors behave like Java exceptions---they can be thrown and propagate ``upwards'' in the evaluation rules
+(\errorvalv{} rules are deferred to the appendix).
 
@@
 assume they can never go wrong
-(other assumptions
-given in the supplemental material).
+(other assumptions given in the supplemental material).
 

### 2015-12-31 formal-model.tex
overlapping
clarify where object came from

@@
 $
-with \objectp{} = {\x{}} and \t{v} = \Value{\makekw{en}}. Then {\thenprop{\proppp{}}} = 
+with \objectp{} = {\x{}} (from below the arrow on the right argument of the previous type) and \t{v} = \Value{\makekw{en}}. 
+Then {\thenprop{\proppp{}}} = 
 {\isprop {\Value{\makekw{en}}}{\x{}}} by
@@
 Notice the overall type of a \defmethodliteral{} is the same as its first subexpression \e{m}.
+
+It is worth noting the lack of special typing rules for overlapping methods---each
+method is checked independently based on local type information.
 

### 2015-12-31 formal-model.tex
clarify isaprops

@@
 
-We now consider the \isaliteral{} operation, a core part of the dispatch mechanism for multimethods. 
-Recalling the \secref{sec:multioverview} examples,
+We now consider the \isaliteral{} operation, a core part of the multimethod dispatch mechanism. 
+Recalling the examples in \secref{sec:multioverview},
 \isaliteral{} is
-a subclassing test for classes, otherwise an equality test.
+a subclassing test for classes, but otherwise is an equality test.
 %---we do not model the semantics for vectors
 %
-The key component of the T-IsA rule (\figref{main:figure:mmsyntax})
-is the \isacompareliteral{} 
-metafunction, used to calculate the propositions for
-\isaliteral{} tests.
+The T-IsA rule uses \isacompareliteral{}
+(\figref{main:figure:mmsyntax}), a metafunction which produces the propositions for
+\isaliteral{} expressions.
 %\begin{mathpar}
@@
 
-As an example, the expression
+To demonstrate the first \isacompareliteral{} case,
+the expression
 \isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
-has the true and false propositions 
+is true if \x{} is a keyword, otherwise false.
+When checked with T-IsA,
+the object of the left subexpression \object{} = {\path{\classpe{}}{\x{}}}
+(which starts with the {\classpe{}} path element)
+and the type of the right subexpression \t{} = {\Value{\Keyword}} (a singleton class type)
+together trigger the first \isacompareliteral{} case
 \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
-meaning that if this expression produces true, \x{} is a keyword, otherwise it is not.
+giving propositions that correspond to our informal description {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} = {\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.
+
+The second \isacompareliteral{} case captures the simple equality mode for non-class singleton types.
+For example,
+the expression
+\isaapp{\x{}}{\makekw{en}} produces true 
+when \x{} evaluates to {\makekw{en}}, otherwise it produces false.
+Using T-IsA,
+it has the propositions {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} = 
+\isacompare{}{\x{}}{\Value{\makekw{en}}}{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
+since \object{} = {\x{}} and \t{} = {\Value{\makekw{en}}}.
+%
+The side condition on the second \isacompareliteral{} case ensures we are in equality mode---if \x{} can possibly be a class in 
+\isaapp{\x{}}{\Object{}}, \isacompareliteral{} uses its conservative default case,
+since if \x{} is a class literal, subclassing mode could be triggered.
+%
+Capture-avoiding substitution of objects {\replacefor {} {\object{}} {\x{}}} used in this case erases propositions
+that would otherwise have \emptyobject{} substituted in for their objects---it
+is defined in the appendix.
 
@@
 Then {\thenprop{\proppp{}}} = 
-{\isprop {\Value{\makekw{en}}}{\x{}}} by
+{\isprop {\Value{\makekw{en}}}{\x{}}}
+from
 $
@@
 {\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
-$.
+$
+(see \secref{sec:isaformal}).
 Since \t{} = \Keyword{}, we check the method body with

### 2016-01-07 intro.tex
add tejas, no table
explain ann, Any, Kw

@@
   enforce type invariants we describe as ``optionally typed''.}
-%FIXME: add citations for all those systems.
 
@@
 An example of these features is given in
-\figref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
-on the \texttt{class} of the argument---for \texttt{String}s,
-the first method implementation is called, for \texttt{File}s, the
-second. The \texttt{String} method calls
-a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
-\texttt{.getName} method 
-on \texttt{File} requires a non-nil target, returning a nilable
+\figref{fig:ex1}. Here, the \clj{pname} multimethod dispatches
+on the \clj{class} of the argument---for \clj{String}s,
+the first method implementation is called, for \clj{File}s, the
+second. The \clj{String} method calls
+a \clj{File} constructor, returning a non-nil \clj{File} instance---the 
+\clj{.getName} method 
+on \clj{File} requires a non-nil target, returning a nilable
 type.  

### 2016-01-07 overview.tex
explain ann, Any, Kw

@@
 
-Here is a simple one-argument function
-%
-\clj{greet} that takes and returns strings.
+A simple one-argument function \clj{greet} is annotated with \clj{ann} to take and return strings.
 
@@
 %
-Function arguments default to \clj{Any}, the most permissive type. Ascribing
+Function arguments default to \clj{Any}, which is similar to a union of all types. Ascribing
 a parameter type allows Typed Clojure to find a specific method.
@@
 \paragraph{Value-based dispatch}
-This simple multimethod says hello in different languages.%, as
+This simple multimethod takes a keyword (\clj{Kw}) and says hello in different languages.%, as
 %specified by a keyword argument.

### 2016-01-07 overview.tex
nil restrictions

@@
 \end{minted}
-Since Java-level types lack nullability information,
-Typed Clojure conservatively assumes Java methods and
-constructor arguments cannot be passed \clj{nil}---this restriction can be
-adjusted per method, but method targets can never be \clj{nil}.
+Method targets can never be \clj{nil}.
+Typed Clojure also prevents passing \clj{nil} as Java method or
+constructor arguments by default---this restriction can be
+adjusted per method.
 

### 2016-01-07 formal-model.tex
explain hmap lookups

@@
 $$
-The result type is \Nil, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{b}}.
+The result type is \Nil---since \Breakfast is fully specified---with return object \path{\keype{\makekw{bocce}}}{\makelocal{b}}.
 
@@
 $$
-The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
+The result type is \Top---since {\Lunch} has an unknown \makekw{bocce} entry---with return object \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
 Notice propositions are erased once they enter a HMap type.

### 2016-01-07 formal-model.tex
explain absent entries

@@
 is written
-\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}.
-We abbreviate this type as \Lunch in this section.
+\HMapp{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}{\emptyabsent{}}
+(abbreviated \Lunch).
 %
@@
 written
-\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}.
-We abbreviate this type as \Breakfast in this section.
+\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}}}
+(abbreviated \Breakfast).
 To ease presentation, 
-if an HMap has completeness tag \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
+if an HMap has completeness tag \complete{} then \absent{} is elided and implicitly contains all keywords not in the domain of 
+\mandatory{}---dissociating keys is not modelled, so the set of absent entries otherwise
+never grows.
+Keys cannot be both present and absent.
 %\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}.
-Keys cannot be both present and absent.
 

### 2016-01-07 formal-model.tex
42 is unimportant, thus empty object

@@
   variable \makelocal{m} has object \makelocal{m};
-  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
-  $42$ has the \emph{empty} object \emptyobject{}.
+  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and $42$ has the \emph{empty} object \emptyobject{} since it is unimportant in our system.
 %

### 2016-01-07 conclusion.tex
experience section

@@
 The result is a sound, expressive, and useful type system which, as
-implemented in \coretyped with appropriate extensions, suitable for
-typechecking significant amount of existing Clojure programs.
+implemented in \coretyped with appropriate extensions, is suitable for
+typechecking a significant amount of existing Clojure programs.
 %
@@
 the Clojure community among both enthusiasts and professional
-programmers and receives contributions from many developers.
+programmers.% and receives contributions from many developers.
 
-Our empirical analysis of existing Typed Clojure programs bears out
-our design choices. Multimethods, Java interoperation, and
-heterogeneous maps are indeed common in both Clojure and Typed Clojure,
-meaning that our type system must accommodate them. Furthermore, they
-are commonly used together, and the features of each are mutually
-reinforcing. Additionally, the choice to make Java's \clj{null}
-explicit in the type system is validated by the many Typed Clojure
-programs that  specify non-nullable types.
+%Our empirical analysis of existing Typed Clojure programs bears out
+%our design choices. Multimethods, Java interoperation, and
+%heterogeneous maps are indeed common in both Clojure and Typed Clojure,
+%meaning that our type system must accommodate them. Furthermore, they
+%are commonly used together, and the features of each are mutually
+%reinforcing. Additionally, the choice to make Java's \clj{null}
+%explicit in the type system is validated by the many Typed Clojure
+%programs that  specify non-nullable types.
 
 % Delete the following paragraphs if space is needed.
-%
+
 %However, there is much more that Typed Clojure can provide. Most
@@
 %that its information can be communicated to the compiler.
-%
+
 %Finally, our case study, evaluation, and broader experience indicate that Clojure

### 2016-01-07 conclusion.tex
page break, cite core.typed

@@
 programmers, but they must fit together smoothly to ensure that
-existing Clojure programs are easy to convert to Typed Clojure.
+existing untyped programs are easy to convert to Typed Clojure.
 

### 2016-01-11 intro.tex
submission

@@
 \begin{figure}[t!]
-\inputminted[firstline=15]{clojure}{code/demo/src/demo/ex-intro.clj}
+\begin{verbatim}
+(ann pname [(U File String) -> (U nil String)])
+(defmulti pname class)  ; multimethod dispatching on class of argument
+(defmethod pname String [s] (pname (new File s))) ; String case 
+(defmethod pname File [f] (.getName f)) ; File case, with static null check
+(pname "STAINS/JELLY") ;=> "JELLY" :- (U nil Str)
+\end{verbatim}
 \caption{A simple Typed Clojure program}

### 2016-01-11 overview.tex
page break, cite core.typed
submission

@@
 
-\begin{minted}{clj}
+\begin{verbatim}
 (ann  greet [Str -> Str])
@@
 (greet "Grace") ;=> "Hello, Grace!" :- Str
-\end{minted}
+\end{verbatim}
 %
@@
 %
-\begin{minted}{clj}
+\begin{verbatim}
 (greet nil) ; Type Error: Expected Str, given nil
-\end{minted}
+\end{verbatim}
 
@@
 %
-\begin{minted}{clj}
+\begin{verbatim}
 (ann  greet-nil [(U nil Str) -> Str])
@@
 (greet-nil nil)      ;=> "Hello!"         :- Str
-\end{minted}
+\end{verbatim}
 %
@@
 %
-\begin{minted}{clj}
+\begin{verbatim}
 (ann  greetings [Str (U nil Int) -> Str])
@@
 (greetings "Grace" nil) ;=> "Hello, Grace!"                :- Str
-\end{minted}
+\end{verbatim}
 %
@@
 %
-\begin{minted}{clj}
+\begin{verbatim}
 (ann  greetings-bad [Str (U nil Int) -> Str])
@@
   (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
-\end{minted}
+\end{verbatim}
 
@@
 \begin{exmp}
-\begin{minted}{clj}
+\begin{verbatim}
 (.getParent (new File "a/b"))  ;=> "a" :- (U nil Str)
-\end{minted}
+\end{verbatim}
 \label{example:getparent-direct-constructor}
@@
 %
-\begin{minted}{clj}
+\begin{verbatim}
 (fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
-\end{minted}
+\end{verbatim}
 %
@@
 \begin{exmp}
-\begin{minted}{clojure}
+\begin{verbatim}
 (ann parent [(U nil File) -> (U nil Str)])
 (defn parent [f] (if f (.getParent f) nil))
-\end{minted}
+\end{verbatim}
 \label{example:parent-if}
@@
 it is a static type error, as typed code could possibly dereference \clj{nil}.
-\begin{minted}{clj}
+\begin{verbatim}
 (defn parent-bad-in [f :- (U nil File)]
   (.getParent f)) ; Type Error: Cannot call instance method on nil.
-\end{minted}
+\end{verbatim}
 
@@
 %
-\begin{minted}{clj}
+\begin{verbatim}
 (defn parent-bad-out [f :- File] :- Str
   (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
-\end{minted}
+\end{verbatim}
 Method targets can never be \clj{nil}.
@@
 \begin{exmp}
-\inputminted[firstline=15,lastline=15]{clojure}{code/demo/src/demo/parent3.clj}
+\begin{verbatim}
+(parent (new File s))
+\end{verbatim}
 \end{exmp}
@@
 \begin{exmp}
-\begin{minted}{clojure}
+\begin{verbatim}
 (ann hi [Kw -> Str]) ; multimethod type
@@
 (defmethod hi :default [_] "um...") ; default method
-\end{minted}
+\end{verbatim}
 \label{example:hi-multimethod}
@@
 %
-\begin{minted}{clojure}
+\begin{verbatim}
 (map hi [:en :fr :bocce]) ;=> ("hello" "bonjour" "um...")
-\end{minted}
+\end{verbatim}
 %
@@
 %(isa? (class (new File "JELLY")) String) ;=> false
-\begin{minted}{clojure}
+\begin{verbatim}
 (isa? :en :en)       ;=> true
 (isa? String Object) ;=> true
-\end{minted}
+\end{verbatim}
 %
@@
 \begin{exmp}
-\begin{minted}{clojure}
+\begin{verbatim}
 (def breakfast {:en "waffles" :fr "croissants"})
@@
 (:bocce breakfast) ;=> nil       :- nil
-\end{minted}
+\end{verbatim}
 \label{example:breakfastcomplete}
@@
 keyword-keyed maps.
-\begin{minted}{clojure}
+\begin{verbatim}
 breakfast ; :- (HMap :mandatory {:en Str, :fr Str}, :complete? true)
-\end{minted}
+\end{verbatim}
 This says
@@
 \begin{exmp}
-\begin{minted}{clojure}
+\begin{verbatim}
 (ann lunch '{:en Str :fr Str})
@@
 (:bocce lunch) ;=> nil :- Any ; less accurate type
-\end{minted}
+\end{verbatim}
 \label{example:lunchpartial}
@@
 \begin{exmp}
-\begin{minted}{clojure}
+\begin{verbatim}
 (defalias RawKeyPair ; extra keys disallowed
@@
   (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
-\end{minted}
+\end{verbatim}
 \label{example:circleci}
@@
 %
-\begin{minted}{clojure}
+\begin{verbatim}
 (defn enc-keypair-bad [kp] ; Type error: :priv disallowed
   (assoc kp :enc-priv (encrypt (:priv kp))))
-\end{minted}
+\end{verbatim}
 
@@
 
-\begin{minted}{clojure}
+\begin{verbatim}
 (defalias Order "A meal order, tracking dessert quantities."
@@
      '{:Meal ':combo :meal1 Order :meal2 Order}))
-\end{minted}
+\end{verbatim}
 
@@
 %
-\begin{exmp}
-\begin{minted}{clojure}
+\begin{exmp2} % use other environment to work around page break
+\begin{verbatim}
 (ann desserts [Order -> Int])
@@
            :meal2 {:Meal :dinner :desserts 2}}) ;=> 3
-\end{minted}
+\end{verbatim}
 \label{example:desserts-on-meal}
-\end{exmp}
+\end{exmp2}
 %
@@
 \begin{exmp}
-\begin{minted}{clojure}
+\begin{verbatim}
 (ann desserts' [Order -> Int])
@@
   (+ (desserts' (:meal1 o)) (desserts' (:meal2 o))))
-\end{minted}
+\end{verbatim}
 \label{example:desserts-on-class}
@@
 % Note: mintedinline doesn't work with hash characters #
-\begin{minted}{clojure}
+\begin{verbatim}
 (U '{:desserts nil} (HMap :absent-keys #{:desserts}))
-\end{minted}
+\end{verbatim}
 This eliminates non-\clj{:combo} meals

### 2016-01-13 intro.tex
WIP grayscale

--- 2016-01-07 intro.tex
@@
 
-\begin{figure}[t!]
-\inputminted[firstline=15]{clojure}{code/demo/src/demo/ex-intro.clj}
+\begin{figure*}[t!]
+\begin{lstlisting}
+(ann pname [(U File String) -> (U nil String)])
+(defmulti pname class)  ; multimethod dispatching on class of argument
+(defmethod pname String [s] (pname (new File s))) ; String case 
+(defmethod pname File [f] (.getName f)) ; File case, with static null check
+(pname "STAINS/JELLY") ;=> "JELLY" :- (U nil Str)
+\end{lstlisting}
 \caption{A simple Typed Clojure program}
 \label{fig:ex1}
-\end{figure}
+\end{figure*}
 

### 2016-01-13 formal-model.tex
defend exercise 9

@@
 which states that \e{} evaluates to answer \definedreduction{} in environment
-\openv{}. We chose to omit the core rules (see \figref{appendix:figure:opsem})
+\openv{}. We chose to omit the core rules (included in supplemental material)
 however a notable difference is \nil{} is a false value, which affects the
@@
 because a \makekw{combo} meal lacks a \makekw{desserts} entry.
+Adding a rule to handle \egref{example:desserts-on-class} is otherwise straightforward.
 

### 2016-01-13 overview.tex

@@
 
-\begin{verbatim}
+\begin{lstlisting}
 (ann  greet [Str -> Str])
@@
 (greet "Grace") ;=> "Hello, Grace!" :- Str
-\end{verbatim}
+\end{lstlisting}
 %
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (greet nil) ; Type Error: Expected Str, given nil
-\end{verbatim}
+\end{lstlisting}
 
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (ann  greet-nil [(U nil Str) -> Str])
@@
 (greet-nil nil)      ;=> "Hello!"         :- Str
-\end{verbatim}
+\end{lstlisting}
 %
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (ann  greetings [Str (U nil Int) -> Str])
@@
 (greetings "Grace" nil) ;=> "Hello, Grace!"                :- Str
-\end{verbatim}
+\end{lstlisting}
 %
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (ann  greetings-bad [Str (U nil Int) -> Str])
@@
   (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
-\end{verbatim}
+\end{lstlisting}
 
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (.getParent (new File "a/b"))  ;=> "a" :- (U nil Str)
-\end{verbatim}
+\end{lstlisting}
 \label{example:getparent-direct-constructor}
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
-\end{verbatim}
+\end{lstlisting}
 %
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (ann parent [(U nil File) -> (U nil Str)])
 (defn parent [f] (if f (.getParent f) nil))
-\end{verbatim}
+\end{lstlisting}
 \label{example:parent-if}
@@
 it is a static type error, as typed code could possibly dereference \clj{nil}.
-\begin{verbatim}
+\begin{lstlisting}
 (defn parent-bad-in [f :- (U nil File)]
   (.getParent f)) ; Type Error: Cannot call instance method on nil.
-\end{verbatim}
+\end{lstlisting}
 
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (defn parent-bad-out [f :- File] :- Str
   (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
-\end{verbatim}
+\end{lstlisting}
 Method targets can never be \clj{nil}.
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (parent (new File s))
-\end{verbatim}
+\end{lstlisting}
 \end{exmp}
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (ann hi [Kw -> Str]) ; multimethod type
@@
 (defmethod hi :default [_] "um...") ; default method
-\end{verbatim}
+\end{lstlisting}
 \label{example:hi-multimethod}
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (map hi [:en :fr :bocce]) ;=> ("hello" "bonjour" "um...")
-\end{verbatim}
+\end{lstlisting}
 %
@@
 %(isa? (class (new File "JELLY")) String) ;=> false
-\begin{verbatim}
+\begin{lstlisting}
 (isa? :en :en)       ;=> true
 (isa? String Object) ;=> true
-\end{verbatim}
+\end{lstlisting}
 %
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (def breakfast {:en "waffles" :fr "croissants"})
@@
 (:bocce breakfast) ;=> nil       :- nil
-\end{verbatim}
+\end{lstlisting}
 \label{example:breakfastcomplete}
@@
 keyword-keyed maps.
-\begin{verbatim}
+\begin{lstlisting}
 breakfast ; :- (HMap :mandatory {:en Str, :fr Str}, :complete? true)
-\end{verbatim}
+\end{lstlisting}
 This says
@@
 HMap types default to partial specification, with
-\clj{'{:en Str :fr Str}} abbreviating \clj{(HMap :mandatory {:en Str, :fr Str})}.
+\clj{\{:en Str :fr Str\}} abbreviating \clj{(HMap :mandatory \{:en Str, :fr Str\})}.
 %
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (ann lunch '{:en Str :fr Str})
@@
 (:bocce lunch) ;=> nil :- Any ; less accurate type
-\end{verbatim}
+\end{lstlisting}
 \label{example:lunchpartial}
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (defalias RawKeyPair ; extra keys disallowed
@@
   (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
-\end{verbatim}
+\end{lstlisting}
 \label{example:circleci}
@@
 %
-\begin{verbatim}
+\begin{lstlisting}
 (defn enc-keypair-bad [kp] ; Type error: :priv disallowed
   (assoc kp :enc-priv (encrypt (:priv kp))))
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 (defalias Order "A meal order, tracking dessert quantities."
@@
      '{:Meal ':combo :meal1 Order :meal2 Order}))
-\end{verbatim}
+\end{lstlisting}
 
@@
 \begin{exmp2} % use other environment to work around page break
-\begin{verbatim}
+\begin{lstlisting}
 (ann desserts [Order -> Int])
@@
            :meal2 {:Meal :dinner :desserts 2}}) ;=> 3
-\end{verbatim}
+\end{lstlisting}
 \label{example:desserts-on-meal}
@@
 \begin{exmp}
-\begin{verbatim}
+\begin{lstlisting}
 (ann desserts' [Order -> Int])
@@
   (+ (desserts' (:meal1 o)) (desserts' (:meal2 o))))
-\end{verbatim}
+\end{lstlisting}
 \label{example:desserts-on-class}
@@
 % Note: mintedinline doesn't work with hash characters #
-\begin{verbatim}
+\begin{lstlisting}
 (U '{:desserts nil} (HMap :absent-keys #{:desserts}))
-\end{verbatim}
+\end{lstlisting}
 This eliminates non-\clj{:combo} meals

### 2016-01-14 intro.tex
colors

--- 2016-01-07 intro.tex
@@
 
-\begin{figure}[t!]
-\inputminted[firstline=15]{clojure}{code/demo/src/demo/ex-intro.clj}
-\caption{A simple Typed Clojure program}
+\begin{figure*}[t!]
+  \normalsize
+\begin{lstlisting}
+(*typed ann pname [(U File String) -> (U nil String)] typed*)
+(defmulti pname class)  ; multimethod dispatching on class of argument
+(defmethod pname String [s] (*invoke pname (*interop new File s interop*) invoke*)) ; String case 
+(defmethod pname File [f] (*interop .getName f interop*)) ; File case, static null check
+(*invoke pname "STAINS/JELLY" invoke*) ;=> "JELLY" :- (U nil Str)
+\end{lstlisting}
+\caption{A simple Typed Clojure program (delimiter colors: {\color{interop}Java interoperation}, {\color{types}type annotation},
+  {\color{invoke}function invocation}, {\color{red}collection literals}, {\color{mygray}miscellaneous})}
 \label{fig:ex1}
-\end{figure}
+\end{figure*}
 

### 2016-01-14 overview.tex
colors

@@
 \begin{lstlisting}
-(ann  greet [Str -> Str])
-(defn greet [n] (str "Hello, " n "!"))
-(greet "Grace") ;=> "Hello, Grace!" :- Str
+(*typed ann  greet [Str -> Str] typed*)
+(defn greet [n] (*invoke str "Hello, " n "!" invoke*))
+(*invoke greet "Grace" invoke*) ;=> "Hello, Grace!" :- Str
 \end{lstlisting}
@@
 \begin{lstlisting}
-(greet nil) ; Type Error: Expected Str, given nil
+(*invoke greet nil invoke*) ; Type Error: Expected Str, given nil
 \end{lstlisting}
@@
 \begin{lstlisting}
-(ann  greet-nil [(U nil Str) -> Str])
-(defn greet-nil [n] (str "Hello" (when n (str ", " n)) "!"))
-(greet-nil "Donald") ;=> "Hello, Donald!" :- Str 
-(greet-nil nil)      ;=> "Hello!"         :- Str
+(*typed ann  greet-nil [(U nil Str) -> Str] typed*)
+(defn greet-nil [n] (*invoke str "Hello" (when n (*invoke str ", " n invoke*)) "!" invoke*))
+(*invoke greet-nil "Donald" invoke*) ;=> "Hello, Donald!" :- Str 
+(*invoke greet-nil nil invoke*)      ;=> "Hello!"         :- Str
 \end{lstlisting}
@@
 \begin{lstlisting}
-(ann  greetings [Str (U nil Int) -> Str])
+(*typed ann  greetings [Str (U nil Int) -> Str] typed*)
 (defn greetings [n i]
-  (str "Hello, " (when i (apply str (repeat i "hello, "))) n "!"))
-(greetings "Donald" 2)  ;=> "Hello, hello, hello, Donald!" :- Str
-(greetings "Grace" nil) ;=> "Hello, Grace!"                :- Str
+  (*invoke str "Hello, " (when i (*invoke apply str (*invoke repeat i "hello, " invoke*) invoke*)) n "!" invoke*))
+(*invoke greetings "Donald" 2 invoke*)  ;=> "Hello, hello, hello, Donald!" :- Str
+(*invoke greetings "Grace" nil invoke*) ;=> "Hello, Grace!"                :- Str
 \end{lstlisting}
@@
 \begin{lstlisting}
-(ann  greetings-bad [Str (U nil Int) -> Str])
+(*typed ann  greetings-bad [Str (U nil Int) -> Str] typed*)
 (defn greetings-bad [n i]           ; Expected Int, given (U nil Int)
-  (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
+  (*invoke str "Hello, " (*invoke apply str (*invoke repeat i "hello, " invoke*) invoke*) n "!" invoke*))
 \end{lstlisting}
@@
 \begin{lstlisting}
-(.getParent (new File "a/b"))  ;=> "a" :- (U nil Str)
+(*interop .getParent (*interop new File "a/b" interop*) interop*)  ;=> "a" :- (U nil Str)
 \end{lstlisting}
@@
 \begin{lstlisting}
-(fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
+(fn [f] (*interop .getParent f interop*)) ; Type Error: Unresolved interop: getParent
 \end{lstlisting}
@@
 \begin{lstlisting}
-(ann parent [(U nil File) -> (U nil Str)])
-(defn parent [f] (if f (.getParent f) nil))
+(*typed ann parent [(U nil File) -> (U nil Str)] typed*)
+(defn parent [f] (if f (*interop .getParent f interop*) nil))
 \end{lstlisting}
@@
 (defn parent-bad-in [f :- (U nil File)]
-  (.getParent f)) ; Type Error: Cannot call instance method on nil.
+  (*interop .getParent f interop*)) ; Type Error: Cannot call instance method on nil.
 \end{lstlisting}
@@
 (defn parent-bad-out [f :- File] :- Str
-  (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
+  (*interop .getParent f interop*)) ; Type Error: Expected Str, given (U nil Str).
 \end{lstlisting}
@@
 \begin{lstlisting}
-(parent (new File s))
+(*invoke parent (*interop new File s interop*) invoke*)
 \end{lstlisting}
@@
 \begin{lstlisting}
-(ann hi [Kw -> Str]) ; multimethod type
+(*typed ann hi [Kw -> Str] typed*) ; multimethod type
 (defmulti hi identity) ; dispatch function `identity`
@@
 \begin{lstlisting}
-(map hi [:en :fr :bocce]) ;=> ("hello" "bonjour" "um...")
+(*invoke map hi [*vec :en :fr :bocce vec*] invoke*) ;=> (*list "hello" "bonjour" "um..." list*)
 \end{lstlisting}
@@
 For example, 
-\clj{(hi :en)} evaluates to \clj{"hello"}---it executes
+\clj{(*invoke hi :en invoke*)} evaluates to \clj{"hello"}---it executes
 the \clj{:en} method
-because \clj{(= (identity :en) :en)} is true
-and \clj{(= (identity :en) :fr)} is false.
+because \clj{(*invoke = (*invoke identity :en invoke*) :en invoke*)} is true
+and \clj{(*invoke = (*invoke identity :en invoke*) :fr invoke*)} is false.
 
@@
 \begin{lstlisting}
-(isa? :en :en)       ;=> true
-(isa? String Object) ;=> true
+(*invoke isa? :en :en invoke*)       ;=> true
+(*invoke isa? String Object invoke*) ;=> true
 \end{lstlisting}
@@
 the call
-  \clj{(pname "STAINS/JELLY")}
-picks the \clj{String} method because \clj{(isa? String String)}
+  \clj{(*invoke pname "STAINS/JELLY" invoke*)}
+picks the \clj{String} method because \clj{(*invoke isa? String String invoke*)}
 is true, and
-\clj{(isa? String File)}
+\clj{(*invoke isa? String File invoke*)}
 is not.
@@
 \begin{lstlisting}
-(def breakfast {:en "waffles" :fr "croissants"})
-(:en breakfast)    ;=> "waffles" :- Str
-(:bocce breakfast) ;=> nil       :- nil
+(def breakfast {*map :en "waffles" :fr "croissants" map*})
+(*invoke :en breakfast invoke*)    ;=> "waffles" :- Str
+(*invoke :bocce breakfast invoke*) ;=> nil       :- nil
 \end{lstlisting}
@@
 HMap types default to partial specification, with
-\clj{\{:en Str :fr Str\}} abbreviating \clj{(HMap :mandatory \{:en Str, :fr Str\})}.
+\clj{'\{:en Str :fr Str\}} abbreviating \clj{(HMap :mandatory \{:en Str, :fr Str\})}.
 %
@@
 \begin{lstlisting}
-(ann lunch '{:en Str :fr Str})
-(def lunch {:en "muffin" :fr "baguette"})
-(:bocce lunch) ;=> nil :- Any ; less accurate type
+(*typed ann lunch '{:en Str :fr Str} typed*)
+(def lunch {*map :en "muffin" :fr "baguette" map*})
+(*invoke :bocce lunch invoke*) ;=> nil :- Any ; less accurate type
 \end{lstlisting}
@@
 \begin{lstlisting}
-(defalias RawKeyPair ; extra keys disallowed
+(*typed defalias RawKeyPair ; extra keys disallowed
   (HMap :mandatory {:pub RawKey, :priv RawKey}, 
-        :complete? true))
-(defalias EncKeyPair ; extra keys disallowed
-  (HMap :mandatory {:pub RawKey, :enc-priv EncKey}, :complete? true))
+        :complete? true) typed*)
+(*typed defalias EncKeyPair ; extra keys disallowed
+  (HMap :mandatory {:pub RawKey, :enc-priv EncKey}, :complete? true) typed*)
 
-(ann enc-keypair [RawKeyPair -> EncKeyPair])
+(*typed ann enc-keypair [RawKeyPair -> EncKeyPair] typed*)
 (defn enc-keypair [kp]
-  (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
+  (*invoke assoc (*invoke dissoc kp :priv invoke*) :enc-priv (*invoke encrypt (*invoke :priv kp invoke*) invoke*) invoke*))
 \end{lstlisting}
@@
 (defn enc-keypair-bad [kp] ; Type error: :priv disallowed
-  (assoc kp :enc-priv (encrypt (:priv kp))))
+  (*invoke assoc kp :enc-priv (*invoke encrypt (*invoke :priv kp invoke*) invoke*) invoke*))
 \end{lstlisting}
@@
 \begin{lstlisting}
-(defalias Order "A meal order, tracking dessert quantities."
+(*typed defalias Order "A meal order, tracking dessert quantities."
   (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner :desserts Int}
-     '{:Meal ':combo :meal1 Order :meal2 Order}))
+     '{:Meal ':combo :meal1 Order :meal2 Order}) typed*)
 \end{lstlisting}
@@
 \begin{lstlisting}
-(ann desserts [Order -> Int])
+(*typed ann desserts [Order -> Int] typed*)
 (defmulti desserts :Meal)  ; dispatch on :Meal entry
-(defmethod desserts :lunch [o] (:desserts d))
-(defmethod desserts :dinner [o] (:desserts d))
+(defmethod desserts :lunch [o] (*invoke :desserts d invoke*))
+(defmethod desserts :dinner [o] (*invoke :desserts d invoke*))
 (defmethod desserts :combo [o] 
-  (+ (desserts (:meal1 o)) (desserts (:meal2 o))))
+  (*invoke + (*invoke desserts (*invoke :meal1 o invoke*) invoke*) (*invoke desserts (*invoke :meal2 o invoke*) invoke*) invoke*))
 
-(desserts {:Meal :combo, :meal1 {:Meal :lunch :desserts 1}, 
-           :meal2 {:Meal :dinner :desserts 2}}) ;=> 3
+(*invoke desserts {*map :Meal :combo, :meal1 {*map :Meal :lunch :desserts 1 map*}, 
+           :meal2 {*map :Meal :dinner :desserts 2 map*} map*} invoke*) ;=> 3
 \end{lstlisting}
@@
 be of type
-\clj{'{:Meal :combo}}
+\clj{'\{:Meal :combo\}}
 since
@@
 orders, simplifying \clj{o} to
-\clj{'{:Meal ':combo :meal1 Order :meal2 Order}}
+\clj{'\{:Meal ':combo :meal1 Order :meal2 Order\}}
 which contains appropriate arguments for both recursive calls.
@@
 \begin{lstlisting}
-(ann desserts' [Order -> Int])
+(*typed ann desserts' [Order -> Int] typed*)
 (defmulti desserts' 
-  (fn [o :- Order] (class (:desserts o))))
+  (fn [o :- Order] (*invoke class (*invoke :desserts o invoke*) invoke*)))
 (defmethod desserts' Long [o] 
-; o :- (U '{:Meal (U ':dinner ':lunch), :desserts Int}
+;o :- (U '{:Meal (U ':dinner ':lunch), :desserts Int}
 ;       '{:Meal ':combo, :desserts Int, :meal1 Order, :meal2 Order})
-  (:desserts o))
+  (*invoke :desserts o invoke*))
 (defmethod desserts' nil [o]
   ; o :- '{:Meal ':combo, :meal1 Order, :meal2 Order}
-  (+ (desserts' (:meal1 o)) (desserts' (:meal2 o))))
+  (*invoke + (*invoke desserts' (*invoke :meal1 o invoke*) invoke*) (*invoke desserts' (*invoke :meal2 o invoke*) invoke*) invoke*))
 \end{lstlisting}
@@
 In the \clj{Long} method, Typed Clojure learns that
-its argument is at least of type \clj{'{:desserts Long}}---since
-\clj{(isa? (class (:desserts o)) Long)}
+its argument is at least of type \clj{'\{:desserts Long\}}---since
+\clj{(*invoke isa? (*invoke class (*invoke :desserts o invoke*) invoke*) Long invoke*)}
 must be true.
@@
 In the \clj{nil} method,
-\clj{(isa? (class (:desserts o)) nil)}
-must be true---which implies \clj{(class (:desserts o))} is \clj{nil}.
+\clj{(*invoke isa? (*invoke class (*invoke :desserts o invoke*) invoke*) nil invoke*)}
+must be true---which implies \clj{(*invoke class (*invoke :desserts o invoke*) invoke*)} is \clj{nil}.
 %
@@
 This eliminates non-\clj{:combo} meals
-since their \clj{'{:desserts Int}} type does not agree
+since their \clj{'\{:desserts Int\}} type does not agree
 with this new information (because \clj{:desserts}

### 2016-01-14 intro.tex
change colors, fix breaks

@@
 \caption{A simple Typed Clojure program (delimiter colors: {\color{interop}Java interoperation}, {\color{types}type annotation},
-  {\color{invoke}function invocation}, {\color{red}collection literals}, {\color{mygray}miscellaneous})}
+  {\color{invoke}function invocation (black)}, {\color{red}collection literals}, {\color{mygray}miscellaneous})}
 \label{fig:ex1}

### 2016-01-14 intro.tex
colors

--- 2016-01-14 intro.tex
@@
 \end{lstlisting}
-\caption{A simple Typed Clojure program (delimiter colors: {\color{interop}Java interoperation}, {\color{types}type annotation},
-  {\color{invoke}function invocation}, {\color{red}collection literals}, {\color{mygray}miscellaneous})}
+\caption{A simple Typed Clojure program (delimiter colors: {\color{interop}Java interoperation (green)}, 
+  {\color{types}type annotation (blue)},
+  {\color{invoke}function invocation (black)}, {\color{red}collection literal (red)}, {\color{mygray}miscellaneous (gray)})}
 \label{fig:ex1}
@@
 a \clj{File} constructor, returning a non-nil \clj{File} instance---the 
-\clj{.getName} method 
+\clj{getName} method 
 on \clj{File} requires a non-nil target, returning a nilable

### 2016-01-14 intro.tex
colors, footnote to examples

@@
 \end{lstlisting}
-\caption{A simple Typed Clojure program (delimiter colors: {\color{interop}Java interoperation (green)}, 
+\caption{A simple Typed Clojure program (colors: {\color{interop}Java interoperation (green)}, 
   {\color{types}type annotation (blue)},
-  {\color{invoke}function invocation (black)}, {\color{red}collection literal (red)}, {\color{mygray}miscellaneous (gray)})}
+  {\color{invoke}function invocation (black)}, {\color{red}collection literal (red)}, {\color{mygray}other (gray)})}
 \label{fig:ex1}

### 2016-01-14 overview.tex
change colors, fix breaks
colors, footnote to examples

@@
 uses the full Typed Clojure system to illustrate key type system
-ideas, before studying the core features in detail in
+ideas,\footnote{Full examples: \url{https://github.com/typedclojure/esop16}} before studying the core features in detail in
 \secref{sec:formal}.
@@
 
+\newpage
+
 \begin{exmp}
@@
 natural to dispatch on a known entry like \clj{:Meal}.
-%
-\begin{exmp2} % use other environment to work around page break
+
+\newpage
+
+\begin{exmp}
 \begin{lstlisting}
@@
 \label{example:desserts-on-meal}
-\end{exmp2}
+\end{exmp}
 %

### 2018-05-17 intro.tex
spelling

@@
 for Python among the optional systems, and Typed Racket~\cite{TF08}, Reticulated
-Python~\cite{Vitousek14}, and GradualTalk~\cite{gradualtalk} among gradually-typed systems.\footnote{We
+Python~\cite{Vitousek14}, and Gradualtalk~\cite{gradualtalk} among gradually-typed systems.\footnote{We
   use ``gradual typing'' for systems like Typed Racket with sound
@@
 One key lesson of these systems, indeed a lesson known to early
-developers of optional type systems such as StrongTalk, is that type
+developers of optional type systems such as Strongtalk, is that type
 systems for existing languages must be designed to work with the
@@
 \end{lstlisting}
-\caption{A simple Typed Clojure program (colors: {\color{interop}Java interoperation (green)}, 
+\caption{A simple Typed Clojure program (delimiters: {\color{interop}Java interoperation (green)}, 
   {\color{types}type annotation (blue)},

### 2018-05-18 overview.tex
typo

@@
 (defmulti desserts :Meal)  ; dispatch on :Meal entry
-(defmethod desserts :lunch [o] (*invoke :desserts d invoke*))
-(defmethod desserts :dinner [o] (*invoke :desserts d invoke*))
+(defmethod desserts :lunch [o] (*invoke :desserts o invoke*))
+(defmethod desserts :dinner [o] (*invoke :desserts o invoke*))
 (defmethod desserts :combo [o] 
@@
 be of type
-\clj{'\{:Meal :combo\}}
+\clj{'\{:Meal ':combo\}}
 since

### 2018-12-09 abstract.tex
abstract in own file

--- 2015-10-17 small.tex
@@
-% rubber: set arguments --shell-escape
-% arara: pdflatex: {shell: true}
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass{llncs}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\bibliographystyle{splncs03}
-
-\usepackage{common}
-\usepackage{minted}
-\include{bibliography.bib}
-\usepackage{chngcntr}
-\usepackage{url}
-% seems like a nightmare to use, but you get clickable links
-%\usepackage{hyperref}
-
-\newcounter{ex}
-\newenvironment{exmp}{
-\refstepcounter{ex}%
-\par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
-\\[-3.23em]\nopagebreak}
-{\relax}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-%
-%\setminted{%fontsize=\footnotesize
-%}
-
-\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
-\newcommand{\java}[1]{\mintinline{java}{#1}}
-\newcommand{\rkt}[1]{\mintinline{racket}{#1}}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-%\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Practical Optional Types for Clojure}
-
-%\authorinfo{}{}{\vspace*{-2mm}}
-\author{\ma{\text{Ambrose Bonnaire-Sergeant}^\dagger\text{, Rowan Davies*}, \text{Sam Tobin-Hochstadt}^\dagger}}
-\institute{Indiana University$\dagger$; Omnia Team, Commonwealth Bank of Australia*}
-%       {Indiana University}
-%       {abonnair@indiana.edu}
-%       {University of Western Australia}
-%       {Rowan.Davies@cba.com.au}
-%       {Indiana University}
-%       {samth@indiana.edu}
-
-\maketitle
-
-\begin{abstract}
 Typed Clojure is an optional type system for Clojure, a dynamic
@@
 %% for type checking non-trivial Clojure idioms.
-\end{abstract}
-
-%% \category{CR-number}{subcategory}{third-level}
-
-%% \keywords
-%% optional types, Clojure
-
-\input{intro}
-
-\input{overview}
-
-\input{formal-model}
-
-\input{metatheory}
-
-\input{experience}
-
-\input{related-work}
-
-\input{conclusion}
-
-%\input{acknowledgements}
-
-% We recommend abbrvnat bibliography style.
-
-% The bibliography should be embedded for final submission.
-\newpage
-
-\bibliography{bibliography}
-
-\clearpage
-
-\input{appendix}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2018-12-09 small.tex
page break, cite core.typed
email
submission
fix page length and bib
WIP grayscale
abstract in own file

@@
 \usepackage{common}
-\usepackage{minted}
 \include{bibliography.bib}
@@
 
+\newenvironment{exmp2}{
+\refstepcounter{ex}%
+\par\vspace{3em}\hfill\framebox{\small Example~\theex}\nopagebreak
+\\[-3.23em]\nopagebreak}
+{\relax}
+
+\newenvironment{mintedclj}{
+\begin{verbatim}%
+}
+{\end{verbatim}%
+}%
+
 \begin{document}
@@
 
-\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
-\newcommand{\java}[1]{\mintinline{java}{#1}}
-\newcommand{\rkt}[1]{\mintinline{racket}{#1}}
+\newcommand{\clj}[1]{\lstinline{#1}}
+\newcommand{\java}[1]{\lstinline{#1}}
+\newcommand{\rkt}[1]{\lstinline{#1}}
 
@@
 \author{\ma{\text{Ambrose Bonnaire-Sergeant}^\dagger\text{, Rowan Davies*}, \text{Sam Tobin-Hochstadt}^\dagger}}
-\institute{Indiana University$\dagger$; Omnia Team, Commonwealth Bank of Australia*}
+\institute{Indiana University$\dagger$; Omnia Team, Commonwealth Bank of Australia*
+  \{abonnair,samth\}@indiana.edu, Rowan.Davies@cba.com.au}
 %       {Indiana University}
@@
 \begin{abstract}
-Typed Clojure is an optional type system for Clojure, a dynamic
-language in the Lisp family that targets the JVM. Typed Clojure
-enables Clojure programmers to gain greater confidence in the
-correctness of their code via static type checking while remaining in
-the Clojure world, and has acquired significant adoption in the
-Clojure community. Typed Clojure repurposes Typed Racket's
-\emph{occurrence typing}, an approach to statically reasoning about
-predicate tests, and also includes several new type system features to
-handle existing Clojure idioms.
-
-In this paper, we describe Typed Clojure and present these type system
-extensions, focusing on three features widely used in Clojure. 
-%
- First, multimethods provide extensible
-operations, and their Clojure semantics turns out to have a surprising
-synergy with the underlying occurrence typing framework.
-%
-Second, Java
-interoperability is central to Clojure's mission but introduces
-challenges such as ubiquitous \texttt{null}; Typed Clojure handles
-Java interoperability while ensuring the absence of null-pointer
-exceptions in typed programs. 
-%
-Third, Clojure programmers
-idiomatically use immutable dictionaries for data structures; Typed
-Clojure handles this with multiple forms of
-heterogeneous dictionary types.
-%
-
-We provide a formal model of the Typed Clojure type system
-incorporating these and other features, with a proof of
-soundness. Additionally, Typed Clojure is now in use by numerous
-corporations and developers working with Clojure, and we present
-a quantitative analysis on the use of type system features
-in two substantial code bases.
-
-%% Clojure is a dynamically typed language hosted on the Java 
-%% Virtual Machine.
-%% Typed Racket is a valuable starting point for
-%% a gradual type system that targets Clojure.
-%% Building a similar type system for a new language gives the
-%% designer some flexibility to repurpose and extend features.
-%% This paper gives an overview of Typed Clojure, concentrating
-%% on the extensions and differences from Typed Racket. We also
-%% show where Typed Racket's features were particularly useful
-%% for type checking non-trivial Clojure idioms.
+  \input{abstract}
 \end{abstract}
@@
 
-\clearpage
-
-\input{appendix}
+%\clearpage
+%
+%\input{appendix}
 
